### 2012-06-28 doc/original/main.tex
Added original paper draft

\documentclass{llncs}
%%
%\setlength{\textheight}{23cm}
\setlength{\textwidth}{13.5cm}
\advance\evensidemargin by -.65cm
\advance\oddsidemargin by -.65cm
%\setlength{\hoffset}{-1cm}
%\renewcommand{\baselinestretch}{0.95}

%\setlength{\abovedisplayskip}{3pt plus 0pt minus 0pt}
%\setlength{\abovedisplayskip}{3pt}
%\setlength{\abovedisplayshortskip}{3pt}
%\setlength{\belowdisplayskip}{3pt}
%\setlength{\belowdisplayshortskip}{3pt}

\usepackage{fleqn}
\usepackage{listings}
\usepackage{math}
\usepackage{amsmath}
\usepackage{latexsym}
\usepackage{bcprules}
\usepackage[scaled=0.848971]{luximono} % This is for 11 pt Default font
\usepackage[T1]{fontenc}

% Prooftree formatting
\usepackage{prooftree}

% support for generating PDF files
%\newif\ifpdf
%    \ifx\pdfoutput\undefined
%    \pdffalse
%\else
%    \pdftrue
%    \pdfoutput=1
%\fi

%versions
% Use dependent function types
\newif\ifdep\depfalse

\lstset{
  literate=
  {=>}{$\Rightarrow\;$}{2}
  {<:}{$<:\;$}{1}
}

\lstdefinelanguage{scala}{% 
       morekeywords={% 
                try, catch, throw, private, public, protected, import, package, implicit, final, package, trait, type, class, val, def, var, if, for, this, else, extends, with, while, new, abstract, object, case, match, sealed,override},
         sensitive=t, % 
   morecomment=[s]{/*}{*/},morecomment=[l]{\//},% 
   mathescape,
%   escapeinside={/*\%}{*/},%
   rangeprefix= /*< ,rangesuffix= >*/,%
   morestring=[d]{"}% 
 }

\lstset{breaklines=true,language=scala}

\def\c{\lstinline}  % shorter version so you can write \c|String[Foo]|
                    % -- \def must be in same file as uses for this to
                    % work...
\newcommand{\lstref}[1]{Listing~\ref{#1}}
\newcommand{\Lstref}[1]{Listing~\ref{#1}} % only capitalise at beginning of sentence?
\newcommand{\secref}[1]{Section~\ref{#1}}
\newcommand{\Secref}[1]{Section~\ref{#1}} % only capitalise at beginning of sentence?

% \lstset{basicstyle=\footnotesize\ttfamily, breaklines=true, language=scala, tabsize=2, columns=fixed, mathescape=false,includerangemarker=false}
% thank you, Burak 
% (lstset tweaking stolen from
% http://lampsvn.epfl.ch/svn-repos/scala/scala/branches/typestate/docs/tstate-report/datasway.tex)
\lstset{
    xleftmargin=2em,%
    framesep=5pt,%
    frame=none,%
    captionpos=b,%
    fontadjust=true,%
    columns=[c]fixed,%
    keepspaces=false,%
    basewidth={0.56em, 0.52em},%
    tabsize=2,%
    basicstyle=\small\tt,% \small\tt
    commentstyle=\textit,%
    keywordstyle=\bfseries,%
    escapechar=\%,%
}

%% set latex/pdflatex specific stuff
%\ifpdf
    \usepackage[pdftex,
                hyperindex,
                plainpages=false,
                breaklinks,
                colorlinks,
                citecolor=black,
                filecolor=black,
                linkcolor=black,
                pagecolor=black,
                urlcolor=black]{hyperref}
    \usepackage[pdftex]{graphicx}
    \DeclareGraphicsExtensions{.jpg,.pdf}
    \pdfcatalog {
        /PageMode (/UseNone)
    }
    \usepackage{thumbpdf}
    \usepackage[pdftex]{color}
%\else
%    \usepackage[ps2pdf]{hyperref}
%    \usepackage{graphicx}
%    \DeclareGraphicsExtensions{.eps,.jpg}
%    \usepackage{color}
%\fi

%\setlength{\parindent}{0pt}
%\setlength{\parskip}{5pt}

% verbfilter stuff
\newcommand{\prog}[1]{{\sl #1}}
\newenvironment{program}[1][10.5]
  {\fontsize{#1}{13.6}\tt\begin{tabbing}\hspace*{0.5\parindent}\=\+\kill}
  {\end{tabbing}\noindent}
\newcommand{\blockcomment}[1]{{\color{grayPoint3}#1}}
\newcommand{\linecomment}{\color{grayPoint3}}
\newcommand{\grey}{\color{grey}}

%\newenvironment{program}{\ \ \ \ \begin{minipage}{\textwidth}\renewcommand{\baselinestretch}{1.0}\sl\begin{tabbing}}{\end{tabbing}\end{minipage}}
\newcommand{\vem}{\bfseries}
\newcommand{\quotedstring}[1]{{#1}}
\newcommand{\typename}[1]{{#1}}
\newcommand{\literal}[1]{{#1}}

% comments and notes
\newcommand{\comment}[1]{}
%\newcommand{\note}[1]{{\bf $\clubsuit$ #1 $\spadesuit$}}

% figures
\newcommand{\figurebox}[1]
        {\fbox{\begin{minipage}{\textwidth} #1 \medskip\end{minipage}}}
\newcommand{\twofig}[3]
        {\begin{figure*}[t]#3\ \hrulefill\ 
        \caption{\label{#1}#2}\end{figure*}}
\newcommand{\boxfig}[3]
        {\begin{figure*}\figurebox{#3\caption{\label{#1}#2}}\end{figure*}}
\newcommand{\figref}[1]
        {Figure~\ref{#1}}

% typing rules (not used here)
\newcommand{\ttag}[1]{\mbox{\textsc{\small(#1)}}}
\newcommand{\infer}[3]{\mbox{#1 }\ba{c} #2 \\ \hline #3 \ea}
\newcommand{\irule}[2]{{\renewcommand{\arraystretch}{1.2}\ba{c} #1 
                        \\ \hline #2 \ea}}
\newlength{\trulemargin}
\newlength{\trulewidth}
\newlength{\srulewidth}
\setlength{\trulemargin}{1.75cm}
\setlength{\trulewidth}{83.7mm}
\setlength{\srulewidth}{6.0cm}
\newenvironment{trules}{$\vspace{0.5em}\ba{p{\trulemargin}@{~}p{\trulewidth}@{~}p{\trulemargin}}}{\ea$}
\newenvironment{srules}{$\vspace{0.5em}\ba{p{\trulemargin}@{~}p{\srulewidth}}}{\ea$}
\newcommand{\laxiom}[2]{\ttag{#1} & $ #2 \hfill\ }
\newcommand{\raxiom}[2]{\hfill #2 $& \hfill \ttag{#1}}
\newcommand{\caxiom}[2]{\ttag{#1} & $\hfill #2 \hfill $& \ }
\newcommand{\lrule}[3]{\laxiom{#1}{\irule{#2}{#3}}}
\newcommand{\rrule}[3]{\raxiom{#1}{\irule{#2}{#3}}}
\newcommand{\crule}[3]{\caxiom{#1}{\irule{#2}{#3}}}
\newcommand{\lsrule}[3]{\lsaxiom{#1}{\irule{#2}{#3}}}
\newcommand{\rsrule}[3]{\rsaxiom{#1}{\irule{#2}{#3}}}
\newcommand{\nl}{\end{trules}\\[0.5em] \begin{trules}}
\newcommand{\snl}{\end{srules}\\[0.5em] \begin{srules}}

% commas and semicolons
\newcommand{\comma}{,\,}
\newcommand{\commadots}{\comma \ldots \comma}
\newcommand{\semi}{;\mbox{;};}
\newcommand{\semidots}{\semi \ldots \semi}

% spacing
\newcommand{\gap}{\quad\quad}
\newcommand{\biggap}{\quad\quad\quad}
\newcommand{\nextline}{\\ \\}
\newcommand{\htabwidth}{0.5cm}
\newcommand{\tabwidth}{1cm}
\newcommand{\htab}{\hspace{\htabwidth}}
\newcommand{\tab}{\hspace{\tabwidth}}
\newcommand{\linesep}{\ \hrulefill \ \smallskip}

% math stuff
\newenvironment{myproof}{{\em Proof:}}{$\Box$}
\newenvironment{proofsketch}{{\em Proof Sketch:}}{$\Box$}
\newcommand{\Case}{{\em Case\ }}

% make ; a delimiter in math mode
% \mathcode`\;="8000 % Makes ; active in math mode
% {\catcode`\;=\active \gdef;{\;}}
% \mathchardef\semicolon="003B

% reserved words
\newcommand{\mathem}{\bf}

% brackets
\newcommand{\set}[1]{\{#1\}}
\newcommand{\sbs}[1]{\lquote #1 \rquote}

% arrays
\newcommand{\ba}{\begin{array}}
\newcommand{\ea}{\end{array}}
\newcommand{\bda}{\[\ba}
\newcommand{\eda}{\ea\]}
\newcommand{\ei}{\end{array}}
\newcommand{\bcases}{\left\{\begin{array}{ll}}
\newcommand{\ecases}{\end{array}\right.}

% \cal ids
\renewcommand{\AA}{{\cal A}}
\newcommand{\BB}{{\cal B}}
\newcommand{\CC}{{\cal C}}
\newcommand{\DD}{{\cal D}}
\newcommand{\EE}{{\cal E}}
\newcommand{\FF}{{\cal F}}
\newcommand{\GG}{{\cal G}}
\newcommand{\HH}{{\cal H}}
\newcommand{\II}{{\cal I}}
\newcommand{\JJ}{{\cal J}}
\newcommand{\KK}{{\cal K}}
\newcommand{\LL}{{\cal L}}
\newcommand{\MM}{{\cal M}}
\newcommand{\NN}{{\cal N}}
\newcommand{\OO}{{\cal O}}
\newcommand{\PP}{{\cal P}}
\newcommand{\QQ}{{\cal Q}}
\newcommand{\RR}{{\cal R}}
\newcommand{\TT}{{\cal T}}
\newcommand{\UU}{{\cal U}}
\newcommand{\VV}{{\cal V}}
\newcommand{\WW}{{\cal W}}
\newcommand{\XX}{{\cal X}}
\newcommand{\YY}{{\cal Y}}
\newcommand{\ZZ}{{\cal Z}}

% misc symbols
\newcommand{\dhd}{\!\!\!\!\!\rightarrow}
\newcommand{\Dhd}{\!\!\!\!\!\Rightarrow}
\renewcommand{\ts}{\,\vdash\,}
\newcommand{\la}{\langle}
\newcommand{\ra}{\rangle}
\newcommand{\eg}{{\em e.g.}}

% misc identifiers
\newcommand{\dom}{\mbox{\sl dom}}
\newcommand{\fn}{\mbox{\sl fn}}
\newcommand{\bn}{\mbox{\sl bn}}
\newcommand{\sig}{\mbox{\sl sig}}
\newcommand{\IF}{\mbox{\mathem if}}
\newcommand{\OTHERWISE}{\mbox{\mathem otherwise}}
\newcommand{\expand}{\prec}
\newcommand{\weakexpand}{\prec^W}
\newcommand{\spcomma}{~,~}

%\newcommand{\inst}{\mbox{\mathem inst}}
\newcommand{\trans}[1]{\la\!\la#1\ra\!\ra}
\renewcommand{\remark}[1]{{\bf $\clubsuit$ #1 $\spadesuit$}}
\newcommand{\todo}[1]{\remark{to do: #1}}
%\newcommand{\J}{\justifies}
%\newcommand{\U}{\using}

% names
\newcommand{\Scala}{\mbox{\textsc{Scala}}}
\newcommand{\Java}{\mbox{\textsc{Java}}}

%\renewcommand\textfraction{.05}
%\renewcommand\floatpagefraction{.9}
%\renewcommand\topfraction{.8}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%   Language abstraction commands     %
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

%% Relations
% Subtype 
\newcommand{\sub}{<:}
% Type assignment
\newcommand{\typ}{:}
% reduction
\newcommand{\reduces}{\;\rightarrow\;}
% well-formedness
\newcommand{\wf}{\;\mbox{\textbf{wf}}}

%% Operators
% Type selection
\newcommand{\tsel}{\#}
% Function type
\newcommand{\tfun}{\rightarrow}
\newcommand{\dfun}[3]{(#1\!:\!#2) \Rightarrow #3}
% Conjunction
\newcommand{\tand}{\wedge}
% Disjunction
\newcommand{\tor}{\vee}
% Singleton type suffix
\newcommand{\sing}{.\textbf{type}}

%% Syntax
% Header for typing rules
\newcommand{\judgement}[2]{{\bf #1} \hfill #2}
% Refinement
\newcommand{\refine}[2]{\left\{#1 \Rightarrow #2 \right\}}
% Field definitions
\newcommand{\ldefs}[1]{\left\{#1\right\}}
% Member sequences
\newcommand{\seq}[1]{\overline{#1}}
% Lambda
\newcommand{\dabs}[3]{(#1\!:\!#2)\Rightarrow #3}
\newcommand{\abs}[3]{\lambda #1\!:\!#2.#3}
% Application
\newcommand{\app}[2]{#1\;#2}
% Substitution
\newcommand{\subst}[3]{[#1/#2]#3}
% Object creation
\newcommand{\new}[3]{\textbf{val }#1 = \textbf{new }#2 ;\; #3}
%\renewcommand{\new}[3]{#1 \leftarrow #2 \,\textbf{in}\, #3}
% Field declaration
\newcommand{\Ldecl}[3]{#1 : #2..#3}%{#1 \operatorname{>:} #2 \operatorname{<:} #3}
\newcommand{\ldecl}[2]{#1 : #2}
% Top and Bottom
\newcommand{\Top}{\top}%{\textbf{Top}}
\newcommand{\Bot}{\bot}%\textbf{Bot}}
% Environment extension
%\newcommand{\envplus}[1]{\uplus \{ #1 \}}
\newcommand{\envplus}[1]{, #1}
% Reduction
\newcommand{\reduction}[4]{#1 \operatorname{|} #2 \reduces #3 \operatorname{|} #4}

%%%%%%%%%%%%%%%%%%%%%%%%
%%%% BEGIN DOCUMENT %%%%
%%%%%%%%%%%%%%%%%%%%%%%%

\begin{document}
\thispagestyle{plain}
\pagestyle{plain}
\mainmatter

\title{Dependent Object Types - A foundation for Scala's type system
  \\ \ \\ \large Draft of \today\ -- Do Not Distrubute}
\author{Martin Odersky, Geoffrey Alan Washburn}
\institute{EPFL}

\maketitle
\sloppy
\newcommand{\lindent}{\hspace{-4mm}}

\begin{abstract}
\end{abstract}
\renewcommand{\abs}[3]{\ifdep \dabs #1 #2 #3 \else \lambda #1\!:\!#2.#3 \fi}

%\fontfamily{timesroman}
\section{Introduction}

This paper presents a proposal for a new type-theoretic foundation of
Scala and languages like it. The properties we are interested in
modelling are Scala's path-dependent types and abstract type members,
as well as its mixture of nominal and structural typing through the
use of refinement types. Compared to previous approaches (nuObj, FS),
we make no attempt to model inheritance or mixin composition. Indeed
we will argue that such concepts are better modelled in a different
setting.

The calculus does not precisely describe what's currently in Scala. It
is more normative than descriptive. The main point of deviation
concerns the difference between Scala's compound type formation using
{\bf with} and classical type intersection, as it is modelled in the
calculus. Scala, and the previous calculi attempting to model it
conflates the concepts of compound types (which inherit the members of
several parent types) and mixin composition (which build classes from
other classes and traits). At first glance, this offers an economy of
concepts. However, it is problematic because mixin composition and
intersection types have quite different properties. In the case of
several inherited members with the same name, mixin composition has to
pick one which overrides the others. It uses for that the concept of
linearization of a trait hierarchy. Typically, given two independent
traits $T_1$ and $T_2$ with a common method $m$, the mixin composition
\c@$T_1$ with $T_2$@ would pick the $m$ in $T_2$, whereas the member in
$T_1$ would be available via a super-call. All this makes sense from
an implementation standpoint. From a typing standpoint it is more
awkward, because it breaks commutativity and with it several
monotonicity properties.

In the present calculus, we replace Scala's compound types by
classical intersection types, which are commutative. We also
complement this by classical union types. Intersections and unions
form a lattice wrt subtyping. This addresses another problematic
feature of Scala: In Scala's current type system, least upper bounds
and greatest lower bounds do not always exist. Here is an example:
Given two traits
\begin{lstlisting}
  trait A { type T <: A }
  trait B { type T <: B }
\end{lstlisting}
The greatest lower bound of \c@A@ and \c@B@ is approximated by the
infinite sequence
\begin{lstlisting}
  A with B { type T <: A with B { type T <: A with B { type T < ... }}}
\end{lstlisting}
The limit of this sequence does not exist as a type in Scala.

This is problematic because glbs and lubs play a central role in
Scala's type inference. The absence of universal glbs and lubs makes
type inference more brittle and more unpredictable.

\subsection*{Why No Inheritance?}

In the calculus we made the deliberate choice not to model any form of
inheritance. This is, first and foremost, to keep the calculus simple.
Secondly, there are many different approaches to inheritance and
mixin composition, so that it looks advantageous not to tie the basic
calculus to a specific one. Finally, it seems that the modelization of
inheritance lends itself to a different approach than the basic
calculus. For the latter, we need to prove the standard theorems of
preservation and progress to establish soundness of the type system.
One might try to do this also for a calculus with inheritance, but our
experience suggests that this complicates the proofs considerably.  An
alternative approach to inheritance that might work better is to model
it as a form of code-reuse. Starting with an enriched type system with
inheritance, and a translation to the basic calculus, one needs to
show type preservation wrt the translation. This might be easier than
to prove type preservation wrt reduction.

\section{The DOT Calculus}

\begin{figure}
\figurebox{
%\renewcommand{\baselinestretch}{0.95}
{\bf Syntax}\medskip

$\ba{l@{\hspace{2mm}}|@{\hspace{2mm}}l}
\ba[t]{l@{\hspace{10mm}}l}
x, y, z    & \lindent{\mbox{Variable}} \\
l          & \lindent{\mbox{Value label}}\\[0.2em]
v ::=      & \lindent{\mbox{Value}} \\
\gap x     & \mbox{variable} \\
\gap \abs x T t & \mbox{function} \\[0.2em]
t ::=      & \lindent{\mbox{Term}} \\
\gap v     & \mbox{value} \\
\gap \app t t  & \mbox{application} \\
\gap \new x c t & \mbox{new instance} \\
\gap t.l  & \mbox{selection} \\[0.2em]
p ::= & \lindent \mbox{Path} \\
\gap x & \mbox{variable} \\
\gap p.l & \mbox{selection} \\
c ::= T_c \ldefs{\seq{l = v}} & \lindent{\mbox{Constructor}} \\[0.2em]
\Gamma ::= \seq{x \typ T} & \lindent\mbox{Environment} \\
s      ::= \seq{x \mapsto c} & \lindent\mbox{Store} \\
\ea
&
\ba[t]{l@{\hspace{10mm}}l}
L ::=      & \lindent{\mbox{Type label}} \\
\gap L_c   & \mbox{class label} \\
\gap L_a   & \mbox{abstract type label} \\[0.2em]
S,T,U,V  ::= & \lindent\mbox{Type}\\
\gap p.L & \mbox{type selection} \\
\gap T \refine z {\seq D} & \mbox{refinement} \\
\ifdep
  \gap \dfun x T T & \mbox{dependent function type} \\
\else
  \gap T \tfun T & \mbox{function type} \\
\fi
\gap T \tand T & \mbox{intersection type} \\
\gap T \tor T & \mbox{union type} \\
\gap \Top  & \mbox{top type} \\
\gap \Bot  & \mbox{bottom type} \\[0.2em]
S_c, T_c ::= & \lindent \mbox{Concrete type} \\
\multicolumn{2}{l}{\gap p.L_c ~|~ T_c \refine z {\seq D} ~|~ T_c \wedge T_c  ~|~ \Top} \\[0.2em]
D ::= & \lindent\mbox{Declaration} \\
\gap \Ldecl L S U & \mbox{type declaration} \\
\gap \ldecl l T   & \mbox{value declaration}
\ea
\ea$
\medskip

\linesep

\begin{multicols}{2}[\judgement{Reduction}{\fbox{$\reduction t s {t'} {s'}$}}]

\infax[$\beta_v$]
{\reduction {(\app{\abs x T t}){v}} s {\subst v x t} s}

\infrule[\textsc{sel}]
{x \mapsto T_c \ldefs{\seq{l = v}} \in s}
{\reduction {x.l_i} s {v_i} s}

\infax[\textsc{new}]
{\reduction {\new x c t} s t {s \envplus{x \mapsto c}}}

\infrule[\textsc{context}]
{\reduction t s {t'} {s'}}
{\reduction {e[t]} s {e[t']} s'}
\end{multicols}

\hfill {\bf where} evaluation context $\gap e ::= [\,] ~|~ \app e t ~|~ \app v e ~|~ e.l\hspace{2cm}$

\linesep

\begin{multicols}{2}[\judgement{Type Assignment}{\fbox{$\Gamma \ts t \typ T$}}]

\infrule[\textsc{var}]
{x \typ T \in \Gamma}
{\Gamma \ts x \typ T}

\ifdep
  \infrule[\textsc{app-term}]
  {x \not\in \fn(T) \\
  \Gamma \ts t \typ \dfun x {T'} T \spcomma t' \typ T'}
  {\Gamma \ts \app t {t'} \typ T}

  \infrule[\textsc{app-path}]
  {\Gamma \ts t \typ \dfun x {T'} T \spcomma p \typ T'}
  {\Gamma \ts \app t p \typ \subst p x T}

  \infrule[\textsc{abs}]
  {\Gamma \envplus{x \typ S} \ts t: T \\
   \Gamma \ts S \wf \spcomma \dfun x S T \sub T' \spcomma T' \wf}
  {\Gamma \ts \abs x S t \typ T'}
\else
  \infrule[\textsc{app}]
  {\Gamma \ts t \typ  S \tfun T \spcomma t' \typ T' \spcomma T' \sub S}
  {\Gamma \ts \app t {t'} \typ T}

  \infrule[\textsc{abs}]
  {x \notin \fn(T) \andalso \Gamma \ts S \wf \\
  \Gamma \envplus{x \typ S} \ts t: T}
  {\Gamma \ts \abs x S t \typ S \tfun T}
\fi

\infrule[\textsc{sel}]
{\Gamma \ts t \ni l \typ T'}
{\Gamma \ts t.l \typ T'}

\infrule[\textsc{new}]
{x \notin \fn(T') \\
\Gamma \ts T_c \wf \spcomma T_c \expand_x \seq{\Ldecl L S U},\seq{\ldecl l V} \\
 \Gamma \envplus{x: T_c} \ts \seq{S \sub U} \spcomma \seq{v \typ V'}
 \spcomma \seq{V' \sub V} \spcomma t \typ T'}
{\Gamma \ts \new x {T_c \ldefs{\seq{l = v}}} t \typ T'}

%\infrule[\textsc{sub}]
%{\Gamma \ts t: T \spcomma T \sub T'}
%{\Gamma \ts t: T'}

\end{multicols}

}

\caption{The DOT Calculus : Syntax, Reduction, Type Assignment}\label{dot-one}
\end{figure}

The DOT calculus is a simple system of dependent
object-types. Figure~\ref{dot-one} gives its syntax, reduction rules,
and type assignment rules.

\subsection*{Notation} We use standard notational conventions for
sets. The notation $\seq{X}$ denotes a set of elements $X$. Given a
such a set $\seq X$ in a typing rule, $X_i$ denotes an arbitrary
element of $X$. 
The $\uplus$ operator extends a set of bindings. It is required that the added binding does not
introduce a variable which is already bound in the base-set. 
We use an
abbreviation for preconditions in typing judgements. Given an
environment $\Gamma$ and some predicates $P$ and $Q$, the condition $\Gamma \ts P \spcomma Q$
is a shorthand for the two conditions $\Gamma \ts P$ and $\Gamma \ts Q$.

\subsection*{Syntax}

There are three alphabets: Variable names $x$, $y$, $z$ are freely
alpha-renamable. They occur as parameters of lambda abstractions, as
binders for objects created by \verb@new@-expressions, and as self
references in refinements. Value labels $l$ denote fields in objects,
which are bound to values at run-time. Type labels $L$ denote type
members of objects. Type labels are further separated into labels for
abstract types $L_a$ and labels for classes $L_c$. It is assumed that
in each program every class label $L_c$ is declared at most once.

We assume that the label alphabets $l$ and $L$ are finite. This is
not a restriction in practice, because one can include in these 
alphabets every label occurring in a given program.

The terms $t$ in DOT consist of variables $x$, $y$, $z$, lambda
abstractions $\abs x T t$, function applications $\app t {t'}$, field
selections $t.l$, and object creation expressions $\new x c t$ where
$c$ is a constructor $T_c \ldefs{\seq{l = v}}$. The latter binds a
variable $x$ to a new instance of type $T_c$ with fields $\seq l$ initialized to values $\seq v$.  The scope of $x$ extends through the term $t$.

Two subclasses of terms are values $v$, which consist of just
variables and lambda abstractions, and paths $v$ which consist of just
variables and selections.

The types in DOT are denoted by letters $S$, $T$, $U$, or $V$. They consist of the following:
\begin{itemize}
\item[-] Type selections $p.L$, which denote the type member $L$ of path $p$.
%         A shorthand notation applies to type selections on singleton types: Instead
%         of $p \sing \tsel L$ one can write just $p.L$.
\item[-] Refinement types $T \refine z {\seq D}$, which refine a type $T$ by a set of declarations $D$.
         The variable $z$ refers to the ``self''-reference of the type. Declarations can refer to
         other declarations in the same type by selecting from $z$.
\ifdep
  \item[-] Dependent function types $\dfun x T {T'}$.
\else
  \item[-] Function types $T \tfun T'$.
\fi
\item[-] Type intersections $T \tand T'$, which carry the declarations of members present in either $T$ or $T'$.
\item[-] Type unions $T \tor T'$, which carry only the declarations of members present in both $T$ and $T'$.
\item[-] A top type $\Top$, which corresponds to an empty object.
\item[-] A bottom type $\Bot$, which represents a non-terminating computation.
\end{itemize}
A subset of types $T_c$ are called {\em concrete types}. These are type selections
$p.L_c$ of class labels,
the top type $\Top$, intersections of concrete types, and refinements $T_c \refine z {\seq D}$ of concrete types. Only concrete types are allowed in constructors $c$.

%Types in DOT form a lattice where $\tand$ is join and $\tor$ is meet.

There are only two forms of declarations in DOT, which are both part
of refinement types.  A value declaration $\ldecl l T$ introduces a
field with type $T$.  A type declaration $\Ldecl L S U$ introduces a
type member $L$ with a lower bound type $S$ and an upper bound type
$U$. There are no type aliases, but a type alias can be simulated by a
type declaration $\Ldecl L T T$ where the lower bound and the upper
bound are the same type $T$.

%It is assumed that the lower bound of
%every class type is the bottom type and that the upper bound is a
%concrete type. That is, the only permissible form of a class
%declaration is $\Ldecl {L_c} \Bot T_c$.

Every field or type label can be declared only once in a set of
declarations $\seq D$. A set of declarations can hence be seen as a map from
labels to their declarations.  Meets $\wedge$ and joins $\vee$ on sets of
declarations are defined as follows.

\bda{lcl@{\gap}l}

      \dom(\seq D \tand \seq {D'}) &~=~& \dom(\seq{D}) \cup \dom(\seq{D'}) \\
      \dom(\seq D \tor \seq {D'}) &=& \dom(\seq{D}) \cap \dom(\seq{D'}) \\[0.5em]
      (D \tand D')(L) &=&
        \Ldecl L {(S \tor S')} {(U \tand U')} & \mbox{if~} (\Ldecl L S U) \in \seq{D} \;\mbox{and}\; (\Ldecl L {S'} {U'}) \in \seq{D'} \\
        &=& D(L) & \mbox{if~} L \notin \dom(\seq{D'}) \\
        &=& D'(L) & \mbox{if~} L \notin \dom(\seq{D}) \\
      (D \tand D')(l) &=&
        \ldecl l {T \tand T'} & \mbox{if~} (\ldecl l T) \in \seq{D} \;\mbox{and}\; (\ldecl l {T'}) \in \seq{D'} \\
        &=& D(l) & \mbox{if~} l \notin \dom(\seq{D'}) \\
        &=& D'(l) & \mbox{if~} l \notin \dom(\seq{D}) \\[0.5em]
      (D \tor D')(L) &=&
        \Ldecl L {(S \tand S')} {(U \tor U')} & \mbox{if~} (\Ldecl L S U) \in \seq{D} \;\mbox{and}\; (\Ldecl L {S'} {U'}) \in \seq{D'} \\
      (D \tor D')(l) &=&
        \ldecl l {T \tor T'} & \mbox{if~} (\ldecl l T) \in \seq{D} \;\mbox{and}\; (\ldecl l {T'}) \in \seq{D'} 
    \eda

Sets of declarations form a lattice with the given meet $\wedge$ and
join $\vee$, the empty set of declatations as the top element, and
the bottom element $\seq{D_\Bot}$, Here $\seq{D_\Bot}$ is the set of
declarations that contains for every term label $l$ the
declaration $\ldecl l \Bot$ and for every type label $L$ the
declaration $\Ldecl L \Top \Bot$.

\subsection*{Reduction rules}

Reduction rules $\reduction t s {t'} {s'}$ in DOT rewrite pairs of
terms $t$ and stores $s$, where stores map variables to constructors.
There are three main reduction rules: ($\beta_V$) is the standard beta
reduction of call-by-value lambda calculus. Rule (\textsc{new}) rewrites an
object creation $\new x c t$ by placing the binding of variable $x$ 
to constructor $c$ in the store and continuing with term $t$. Rule (\textsc{sel})
rewrites a field selection $x.l$ by retrieving the corresponding value from the store.
These reduction rules can be applied anywhere in a term where the hole $[\,]$ of 
an evaluation context $e$ can be situated.

\subsection*{Type assignment rules}

The last part of Figure~\ref{dot-one} presents rules for type
assignment.  
\ifdep
The four rules (\textsc{var}), (\textsc{abs}), (\textsc{app-term}) and
(\textsc{app-path}) correspond to dependently typed lambda calculus
with a subtyping discipline. There are two rules for
applications. Rule (\textsc{app-term}) implements the standard
application rule if the function's result type does not depend on its
argument type. Rule (\textsc{app-path}) applies to dependently typed
functions which can only be applied to paths.
Instead of adding a separate subsumption rule,
subtyping is expressed by preconditions in the application rules and
also in rule (\textsc{new}).  
\else
The three rules (\textsc{var}), (\textsc{abs}), and
(\textsc{app}) on the left correspond to simply typed lambda calculus
with a subtyping discipline (I believe it is straightforward to extend
this to full $F_{\sub}$). Rule (\textsc{abs}) has an additional precondition
stating that the bound variable $x$ may not appear in the function's
result type $T'$. It would be an interesting extension of the calculus
to drop this precondition, thereby allowing dependent function types.
Instead of adding a separate subsumption rule,
subtyping is expressed by preconditions in rules (\textsc{app}) and
(\textsc{new}).  
\fi
Rule rule (\textsc{sub}) is the standard subsumption rule for subtyping.

%The (\textsc{path}) rule gives the alternate type $p \sing$ to a well-typed path
%$p$ whose type $T$ has a non-empty expansion $D,\seq{D}$.  By the subtyping rule
%(\textsc{single-$\sub$}) in Figure~\ref{dot-sub}, the alternate type $p \sing$
%is guaranteed to be a subtype of the underlying type $T$ of $p$.

Rule (\textsc{sel}) types a field selection by means of an auxiliary
membership relation $\ni$, which determines whether a given term contains a given
declaration as one of its members. The membership relation is defined in
Figure~\ref{dot-two} and is further explained below.

The last rule, (\textsc{new}), assigns types to object creation
expressions. It is the most complex of DOT's typing rules.  To
type-check an object creation $\new x {T_c \ldefs {\seq{l = v}}} t$,
one verifies first that the type $T_c$ is well-formed (see
Figure~\ref{dot-wf} for a definition of well-formedness).  One then
determines the set of all declarations that this type carries, using
the expansion relation $\expand$ defined in Figure~\ref{dot-two}.
Every type declaration $\Ldecl L S U$ in this set must be realizable,
i.e.\ its lower bound $S$ must be a subtype of its upper bound $U$.
Every field declaration $\ldecl l V$ in this set must have a
corresponding initializing value of $v$ of type $V$.  These checks are
made in an environment which is extended by the binding $x: T_c$. In
particular this allows field values that recurse on ``self'' by
referring to the bound variable $x$. As in the (\textsc{abs}) rule, it
is required that the bound variable does not appear in the
expression's result type $T'$.

\begin{figure}
 \figurebox{

    \begin{multicols}{2}[\judgement{Membership}{\fbox{$\Gamma \ts t \ni D$}}]

      \infrule[\textsc{path-$\ni$}]
      {\Gamma \ts p \typ T \spcomma T \expand_z \seq D}
      {\Gamma \ts p \ni \subst p z {D_i}}

      \infrule[\textsc{term-$\ni$}]
      {z \not\in \fn(D_i) \andalso \Gamma \ts t \typ T \spcomma T \expand_z \seq D}
      {\Gamma \ts t \ni D_i}
   \end{multicols}

   \linesep

    \begin{multicols}{2}[\judgement{Expansion}{\fbox{$\Gamma \ts T_c \expand_z \seq D$}}]

      \infrule[\textsc{rfn-$\expand$}]
      {\Gamma \ts T \expand_z {\seq D'}}
      {\Gamma \ts T \refine z {\seq D} \expand_z \seq {D'} \tand \seq D}

      \infrule[\textsc{$\tand$-$\expand$}]
      {\Gamma \ts T_1 \expand_z {\seq D_1} \spcomma T_2 \expand_z {\seq D_2}}
      {\Gamma \ts T_1 \tand T_2 \expand_x {\seq D_1 \tand \seq D_2}}

      \infax[\textsc{$\Top$-$\expand$}]
      {\Gamma \ts \Top \expand_z \{\}}

      \infax[\textsc{$\tfun$-$\expand$}]
      {\Gamma \ts S \tfun T  \expand_z \{\}}

      \infrule[\textsc{tsel-$\expand$}]
      {\Gamma \ts p \ni \Ldecl L S U \spcomma U \expand_z \seq D}
      {\Gamma \ts p.L \expand_z \seq D}

      \infrule[\textsc{$\tor$-$\expand$}]
      {\Gamma \ts T_1 \expand_z {\seq D_1} \spcomma T_2 \expand_z {\seq D_2}}
      {\Gamma \ts T_1 \tor T_2 \expand_x {\seq D_1 \tor \seq D_2}}

      \infax[\textsc{$\Bot$-$\expand$}]
      {\Gamma \ts \Bot \expand_z \seq{D_\Bot}}
    \end{multicols}

}
\caption{The DOT Calculus : Membership and Expansion}\label{dot-two}
\end{figure}

\subsection*{Membership}

Figure~\ref{dot-two} presents typing rules for membership and
expansion.  The membership judgement $\Gamma \ts t \ni D$ states that
in environment $\Gamma$ a term $t$ has a declaration $D$ as a member.
There are different rules for paths and general terms.  Rule
(\textsc{path-$\ni$}) applies to paths $p$ that have a refinement type
$T \refine z {\seq D}$ as their type. Members of $p$ are then all
definitions $D$ in the refinement, where any use of the self reference
$z$ is replaced by the path $p$ itself.  Rule (\textsc{and-$\ni$})
allows to merge two member definitions (of the same label) by
conjoining them with $\tand$. Rule (\textsc{term-$\ni$}) establishes
the members of general terms $t$ of type $T$ by introducing a dummy
variable $z$ of type $T$ and then using the rules for path membership
on $z$. It must hold that $z$ itself does not form part of the
resulting member $D$.

\subsection*{Expansion}

The expansion relation $\expand$ is needed to typecheck the complete
set of declarations carried by a concrete type that is used in a
\textbf{new}-expression.  Since this is the only place where expansion
is needed, it is sufficient to define it on concrete types only. The
bottom part of Figure~\ref{dot-two} gives the typing rules for
expansion.

Rule (\textsc{rfn-$\expand$}) states that a refinement type $T_c
\expand_z {\seq D}$ expands to the conjunction of the expansion $D'$
of $T_c$ and the newly added declarations $D$. Rule
(\textsc{tsel-$\expand$}) states that a type selection $p.L$ carries
the same declarations as the upper bound $U$ of $L$ in $T$.  Rule
($\tand$-$\expand$) states that expansion distributes through meets.
Rule (\textsc{$\top$-$\expand$}) states that the top type $\top$
expands to the empty set.

\begin{figure}
  \figurebox{

    \begin{multicols}{2}[\judgement{Subtyping}{\fbox{$\Gamma \ts S \sub T$}}]

%      \infrule[\textsc{trans}]
%      {\Gamma \ts S \sub T \spcomma T \sub U}
%      {\Gamma \ts S \sub U}

      \infax[\textsc{refl}]
      {\Gamma \ts T \sub T}

      \infrule[\textsc{$\sub$-rfn}]
      {\Gamma \ts S \sub T \spcomma S \expand_z \seq{D'} \spcomma \seq{D' \sub D}}
      {\Gamma \ts S \sub T \refine z {\seq D}}

      \infrule[\textsc{$\sub$-tsel}]
      {\Gamma \ts p \ni \Ldecl L S U \spcomma S \sub U \spcomma S' \sub S }
      {\Gamma \ts S' \sub p.L}

      \infrule[\textsc{$\sub$-$\tand$}]
      {\Gamma \ts T \sub T_1 \spcomma T \sub T_2}
      {\Gamma \ts T \sub T_1 \tand T_2}

      \infrule[\textsc{$\sub$-$\tor$}]
      {\Gamma \ts T \sub T_i}
      {\Gamma \ts T \sub T_1 \tor T_2}

      \infax[\textsc{$\sub$-$\Top$}]
      {\Gamma \ts T \sub \Top}

      %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

\ifdep
      \infrule[\textsc{$\sub$-$\Rightarrow$}]
      {\Gamma \ts T \sub S \\ \Gamma \envplus {x: S} \ts S' \sub T'}
      {\Gamma \ts \dfun x S {S'} \sub \dfun x T {T'}}
\else
      \infrule[\textsc{$\sub$-$\tfun$}]
      {\Gamma \ts T \sub S \spcomma S' \sub T'}
      {\Gamma \ts S \tfun S' \sub T \tfun T'}
\fi
      \infrule[\textsc{rfn-$\sub$}]
      {\Gamma \ts T \sub T'}
      {\Gamma \ts T \refine z {\seq D} \sub T'}

      \infrule[\textsc{tsel-$\sub$}]
      {\Gamma \ts p \ni \Ldecl L S U \spcomma S \sub U \spcomma U \sub U'}
      {\Gamma \ts p.L \sub U'}

      \infrule[\textsc{$\tand$-$\sub$}]
      {\Gamma \ts T_i \sub T}
      {\Gamma \ts T_1 \tand T_2 \sub T}

      \infrule[\textsc{$\tor$-$\sub$}]
      {\Gamma \ts T_1 \sub T \spcomma T_2 \sub T}
      {\Gamma \ts T_1 \tor T_2 \sub T}

      \infax[\textsc{$\Bot$-$\sub$}]
      {\Gamma \ts \Bot \sub T}

      \end{multicols}

    \linesep

    \begin{multicols}{2}[\judgement{Declaration subsumption}{\fbox{$\Gamma \ts D \sub D'$}}]

    \infrule[\textsc{tdecl-$\sub$}]
            {\Gamma \ts S' \sub S \spcomma T \sub T'}
            {\Gamma \ts (\Ldecl L S T) \sub (\Ldecl L {S'} {T'})}

    \infrule[\textsc{vdecl-$\sub$}]
            {\Gamma \ts T \sub T'}
            {\Gamma \ts (\ldecl l T) \sub (\ldecl l {T'})}

    \end{multicols}
  }
\caption{The DOT Calculus : Subtyping and Declaration Subsumption}\label{dot-sub}
\end{figure}

\subsection*{Subtyping}

Figure~\ref{dot-sub} defines the subtyping judgement $\Gamma \ts S
\sub T$ which states that in environment $\Gamma$ type $S$ is a
subtype of type $T$.  Any term of type $S$ can then be regarded via
the subsumption rule (\textsc{sub}) as a term of type $T$.

As usual, subtyping is reflexive (\textsc{refl}) but a corresponding
rule for transitivity is missing.  However, most of the rules for
specific types in Figure~\ref{dot-sub} have transitivity built in. The
issue of transitivity is further discussed below.

To prove that a type $S$ is a subtype of a refinement type $T \refine
z {\seq D}$, it must hold that $S$ is a subtype of $T$ and that $S$
contains declarations that subsume the declarations $\seq D$
(\textsc{$\sub$-rfn}). The refinement type itself is a subtype of its
parent type $T$ and all its supertypes (\textsc{rfn-$\sub$}).

A type selection $p.L$ is a subtype of (all supertypes of) the upper bound $U$ of
$L$ in $p$ (\textsc{tsel-$\sub$}).
It is a supertype of (all subtypes of) its lower
bound $S$ (\textsc{$\sub$-tsel}).  Function types are subject to the
standard co-/contravariant subtyping rule 
\ifdep
 (\textsc{$\sub$-$\Rightarrow$}).
\else
  (\textsc{$\sub$-$\tfun$}).
\fi
%A singleton type $p\sing$ is a subtype of any underlying type of path
%$p$ (\textsc{single-$\sub$}). 
The final six rules in
Figure~\ref{dot-sub} turn the subtyping relation into a lattice with 
meets $\tand$, joins $\tor$, bottom element $\Bot$ and top element $\Top$.

One interesting aspect of the subtyping relation of DOT is that it is
reflexive (\textsc{refl}) but not transitive. Instead of a separate
rule that specifies global transitivity, we find transitivity encoded
in most of the inference rules of Figure~\ref{dot-sub}. There is only
a single situation where transitivity does not hold:  Given a type
declaration $\Ldecl L S U$ which is a member of some type $T$, and a
variable $x$ of type $T$, we have that $S \sub x.L$ by
(\textsc{$\sub$-tsel}) and $x.L \sub U$ by (\textsc{tsel-$\sub$}).
However, this does not imply that $S \sub U$. In fact, 
$T$ might be an {\em unrealizable} type, which admits no solution for
its $L$ member, precisely because the lower bound of $L$ is not a
subtype of its upper bound.

The type assignment rule (\textsc{new}) in Figure~\ref{dot-one}
has as one of its preconditions that the type of
the created object must be realizable: each lower bounds must be a subtype
of its corresponding upper bound. Subtyping can be shown to be transitive in DOT if the
least type of a subtype chain is realizable.

\begin{figure}
  \figurebox{
       \begin{multicols}{2}[\judgement{Well-formed types}{\fbox{$\Gamma \ts T \wf$}}]

      \infrule[\textsc{rfn-wf}]
      {\Gamma \ts T \wf \\ 
       \Gamma \envplus {z: T \refine z {\seq D}} \ts \seq {D \wf}}
      {\Gamma \ts T \refine z {\seq D} \wf}

      \infrule[\textsc{tsel-wf$_1$}]
      {\Gamma \ts p \ni \Ldecl L S U \spcomma S \wf \spcomma U \wf}
      {\Gamma \ts p.L \wf}

      \infrule[\textsc{$\tand$-wf}]
      {\Gamma \ts T \wf \spcomma T' \wf}
      {\Gamma \ts T \tand T' \wf}

      \infax[\textsc{$\Bot$-wf}]
      {\Gamma \ts \Bot \wf}

      %%%%%%%%%%%%%%%%%%%%%%%%%%%

\ifdep
      \infrule[\textsc{$\Rightarrow$-wf}]
      {\Gamma \ts T \wf \\ \Gamma \envplus{x: T} \ts T' \wf}
      {\Gamma \ts \dfun x T {T'} \wf}
\else
      \infrule[\textsc{$\tfun$-wf}]
      {\Gamma \ts T \wf \spcomma T' \wf}
      {\Gamma \ts T \tfun T' \wf}
\fi
      \infrule[\textsc{tsel-wf$_2$}]
      {\Gamma \ts p \ni \Ldecl L \Bot U}
      {\Gamma \ts p.L \wf}

      \infrule[\textsc{$\tor$-wf}]
      {\Gamma \ts T \wf \spcomma T' \wf}
      {\Gamma \ts T \tor T' \wf}

      \infax[\textsc{$\Top$-wf}]
      {\Gamma \ts \Top \wf}

    \end{multicols}

    \linesep

    \begin{multicols}{2}[\judgement{Well-formed declarations}{\fbox{$\Gamma \ts D \wf$}}]
      \infrule[\textsc{tdecl-wf}]
      {\Gamma \ts S \wf \spcomma U \wf}
      {\Gamma \ts \Ldecl L S U \wf}

      \infrule[\textsc{vdecl-wf}]
      {\Gamma \ts T \wf}
      {\Gamma \ts \ldecl l T \wf}

    \end{multicols}
  }

\caption{The DOT Calculus : Well-Formedness}\label{dot-wf}
\end{figure}

\subsection*{Declaration Subsumption}

The declaration subsumption judgement $\Gamma \ts D \sub D'$ in Figure~\ref{dot-sub}
states that in environment $\Gamma$
the declaration $D$ subsumes the declaration $D'$. There are two
rules, one for type declarations and one for value declarations. 
Rule (\textsc{tdecl-$\sub$})
states that a type declaration $\Ldecl L S U$ subsumes another type declaration $\Ldecl L {S'} {U'}$
if $S'$ is a subtype of $S$ and $U$ is a subtype of $U'$. In other words, the set of types
between $S$ and $U$ is contained in the set of types between $S'$ and $U'$.
Rule (\textsc{vdecl-$\sub$})
states that a value declaration $\ldecl l T$ subsumes another value declaration $\ldecl l T'$
if $T$ is a subtype of $T'$.

%Declaration subsumption is extended a binary relation between sequences of
%declarations as follows: $\seq{D} <: \seq{D'}$ iff $\forall D'_i \exists D_j. D_j \sub D'_i$.

\subsection*{Well-formedness}

The well-formedness judgement $\Gamma \ts T \wf$ in Figure~\ref{dot-wf}
states that in environment $\Gamma$ the type $T$ is well-formed.

%A singleton type $p \sing$ is well-formed if the path $p$ is
%well-typed (\textsc{single-wf}).

A refinement type $T \refine z {\seq D}$ is
well-formed if the parent type $T$ is well-formed and every
declaration in $\seq D$ is well-formed in an environment augmented by
the binding of the self-reference $z$ to the refinement type itself
(\textsc{rfn-wf}).

A type selection $p.L$ is
well-formed if $L$ is a member of $p$, and the lower bound of $L$
is also well-formed (\textsc{tsel-wf}). The latter condition
has the effect that the lower bound of a type $p.L$ may not
refer directly or indirectly to a type containing $p.L$ itself
--- if it would, the well-formedness judgement of $p.L$ would
not have a finite proof. No such restriction exists for the upper
bound of $L$. The upper bound may in fact refer back to the
type. Hence, recursive class types and F-bounded abstract types are
both expressible.

The other forms of types in DOT are all well-formed if their
constituent types are well-formed.

Well-formedness extends straightforwardly to
declarations with the judgement $\Gamma \ts D \wf$. All declarations are well-formed if their constituent types
are well-formed.

\end{document}

\section{Program Examples}

\begin{figure}
 \figurebox{\renewcommand{\baselinestretch}{0.6}
\begin{lstlisting}
val root = new Any { rootThis =>
  trait Unit extends Any {}
  val unit: Any => rootThis.Unit
  trait Boolean extends Any {
    val ifNat: (rootThis.Unit => rootThis.Nat) => (rootThis.Unit => rootThis.Nat) => rootThis.Nat
  }
  val false: rootThis.Unit => rootThis.Boolean
  val true: rootThis.Unit => rootThis.Boolean
  trait Nat extends Any {
    val isZero: rootThis.Unit => rootThis.Boolean
    val pred: rootThis.Unit => rootThis.Nat
    val succ: rootThis.Unit => rootThis.Nat
    val add: rootThis.Nat => rootThis.Nat
  }  
  val zero: rootThis.Unit => rootThis.Nat
  val successor: rootThis.Nat => rootThis.Nat
  val add2: rootThis.Nat => rootThis.Nat => rootThis.Nat
  val error: rootThis.Unit => Bot
} {
  val unit = (x: Any) => val u = new root.Unit; u
  val false = (u: root.Unit) => {
    val ff = new root.Boolean {
      val ifNat = (t: root.Unit => root.Nat) => (e: root.Unit => root.Nat) => e(root.unit)
    }
    ff
  }
  val true = (u: root.Unit) => {
    val tt = new root.Boolean {
      val ifNat = (t: root.Unit => root.Nat) => (e: root.Unit => root.Nat) => t(root.unit)
    }
    tt
  }
  val zero = (u: root.Unit) => {
    val zz = new root.Nat { 
      val isZero = (u: root.Unit) => root.false(root.unit)
      val succ = (u: root.Unit) => root.successor(zz)
      val pred = (u: root.Unit) => error(root.unit)
      val add = (other: root.Nat) => add2(other, zz)
    }
    zz
  }
  val successor = (n: root.Nat) => {
    val ss = new root.Nat {
      val isZero = (u: root.Unit) => root.true(root.unit)
      val succ = (u: root.Unit) => root.successor(ss)
      val pred = (u: root.Unit) => n
      val add = (other: root.Nat) => add2(other, ss)
    }
    ss
  }
  val add2 = (n1: root.Nat) => (n2: root.Nat) =>
    n1.isZero(root.unit).ifNat
     ((u: root.Unit) => n2)
     ((u: root.Unit) => root.add2(n1.pred(root.unit))(n2.succ(root.unit)))
  val error = (u: root.Unit) => error(x)
}
\end{lstlisting}
}
\caption{Some root classes}\label{code:root}
\end{figure}

\begin{figure}
 \figurebox{
\begin{lstlisting}
val genLists = new Any {  
  trait ListPackage extends Any { thisListPackage =>
    type Elem
    type ListOfElem = List { type Elem = thisList.Elem }
    trait List extends Any {
      val isEmpty: root.Unit => root.Boolean
      val head: root.Unit => root.Boolean
      val tail: root.Unit => thisListPackage.ListOfElem
      val prepend: Elem => ListOfElem
    }
    val nil: root.Unit => ListOfElem
    val cons: Elem => ListOfElem => ListOfElem
  }
}
val natLists = new ListPackage { 
  type Elem = Nat
} {
  val nil = (u: root.Unit) => {
    val nn = new natLists.List {
      val isEmpty = (u: root.Unit) => root.true(root.unit)
      val head = (u: root.Unit) => root.error(root.unit)
      val tail = (u: root.Unit) => root.error(root.unit)
      val prepend = (x: root.Nat) => natLists.cons(x, nn)
    }
    nn
  }
  val cons = (x: natLists.Elem) => (xs: natLists.ListOfElem) => {
    val cc = new natLists.List {
      val isEmpty = (u: root.Unit) => root.false(root.unit)
      val head = (u: root.Unit) => x
      val tail = (u: root.Unit) => xs
      val prepend = (x: root.Nat) => natLists.cons(x, cc)
    }
    cc
  }
}
\end{lstlisting}
}
\caption{List classes}\label{code:more}
\end{figure}

\end{document}

### 2012-06-29 doc/semester/report.tex
template for report

\documentclass[preprint]{sigplanconf}

% The following \documentclass options may be useful:
%
% 10pt          To set in 10-point type instead of 9-point.
% 11pt          To set in 11-point type instead of 9-point.
% authoryear    To obtain author/year citation style instead of numeric.

\usepackage{amsmath}
\usepackage{listings}
\usepackage{alltt}

\lstdefinelanguage{Scala}%
{morekeywords={abstract,%
  case,catch,char,class,%
  def,else,extends,final,for,%
  if,import,implicit,%
  match,module,%
  new,null,%
  object,override,%
  package,private,protected,public,%
  for,public,return,super,%
  this,throw,trait,try,type,%
  val,var,%
  with,%
  yield,%
  lazy%
  },%
  sensitive,%
  morecomment=[l]//,%
  morecomment=[s]{/*}{*/},%
  morestring=[b]",%
  morestring=[b]',%
  showstringspaces=false%
}[keywords,comments,strings]%

\lstset{language=Scala,
  basicstyle=\scriptsize,
  columns=fullflexible
}

\newcommand{\code}[1]{\lstinline[language=Scala,columns=fixed,basicstyle=\ttfamily\small]|#1|}

\begin{document}

\conferenceinfo{EDIC}{July 2012, EPFL.} 
\copyrightyear{2012} 
\copyrightdata{[to be supplied]}

\titlebanner{Dependent Object Types}                           % These are ignored unless
\preprintfooter{Spring 2012 Semester Project -- Nada Amin}     % 'preprint' option specified.

\title{Dependent Object Types}
\subtitle{Spring 2012 Semester Project}

\authorinfo{Nada Amin}
           {EPFL}
           {nada.amin@epfl.ch}

\maketitle

\begin{abstract}
% TODO
\end{abstract}

\acks

I thank Adriaan Moors and Martin Odersky for fruitful discussions and
guidance.

%\bibliographystyle{abbrvnat}
%\bibliography{bib}

\end{document}

### 2012-06-29 doc/semester/report.tex
more templating

@@
 
-% The following \documentclass options may be useful:
-%
-% 10pt          To set in 10-point type instead of 9-point.
-% 11pt          To set in 11-point type instead of 9-point.
-% authoryear    To obtain author/year citation style instead of numeric.
+\input{dot_macros}
 
-\usepackage{amsmath}
-\usepackage{listings}
-\usepackage{alltt}
-
-\lstdefinelanguage{Scala}%
-{morekeywords={abstract,%
-  case,catch,char,class,%
-  def,else,extends,final,for,%
-  if,import,implicit,%
-  match,module,%
-  new,null,%
-  object,override,%
-  package,private,protected,public,%
-  for,public,return,super,%
-  this,throw,trait,try,type,%
-  val,var,%
-  with,%
-  yield,%
-  lazy%
-  },%
-  sensitive,%
-  morecomment=[l]//,%
-  morecomment=[s]{/*}{*/},%
-  morestring=[b]",%
-  morestring=[b]',%
-  showstringspaces=false%
-}[keywords,comments,strings]%
-
-\lstset{language=Scala,
-  basicstyle=\scriptsize,
-  columns=fullflexible
+    \usepackage[pdftex,
+                hyperindex,
+                plainpages=false,
+                breaklinks,
+                colorlinks,
+                citecolor=black,
+                filecolor=black,
+                linkcolor=black,
+                pagecolor=black,
+                urlcolor=black]{hyperref}
+    \usepackage[pdftex]{graphicx}
+    \DeclareGraphicsExtensions{.jpg,.pdf}
+    \pdfcatalog {
+        /PageMode (/UseNone)
     }
-
-\newcommand{\code}[1]{\lstinline[language=Scala,columns=fixed,basicstyle=\ttfamily\small]|#1|}
+    \usepackage{thumbpdf}
+    \usepackage[pdftex]{color}
 
@@
 \begin{abstract}
+In my semester project, I have studied a new proposed type-theoretic
+foundation of Scala and languages like it: the Dependent Object Types
+calculus (DOT). DOT models Scala's path-dependent types and abstract
+type members, as well as its mixture of nominal and structural typing
+through the use of refinement types. My ultimate goal was to prove DOT
+sound, but instead, I found lots of counterexamples to soundness, and
+have explored patches to the calculus.
+\end{abstract}
+
+\section{Introduction}
+
 % TODO
-\end{abstract}
+
+\section{Counterexamples}
+
+\section{Non-Expanding Types and Subtyping Transitivity}
 
@@
 
-I thank Adriaan Moors and Martin Odersky for fruitful discussions and
-guidance.
+I thank Adriaan Moors and Martin Odersky for sharing previous work,
+fruitful discussions and guidance.
 

### 2012-06-30 doc/semester/report.tex
blurb

@@
-\documentclass[preprint]{sigplanconf}
+\documentclass[preprint,9pt]{sigplanconf}
 
@@
 
-\section{Non-Expanding Types and Subtyping Transitivity}
+\subsection{Subtyping Transitivity and Preservation}
+
+Some subtyping transitivity is essential for soundness. Indeed, I show
+how to construct a counterexample to preservation from any
+counterexample to subtyping transitivity where the three involved
+types $S, T, U$ are expressible within a universe $u$ but not
+necessarily realizable:
+
+\begin{align*}
+\mlnew u \ldots {
+\ \mlapp{\abs x \Top x}{
+\ \ \mlapp{\abs f {S \tfun U} f}{
+\ \ \ \mlapp{\abs f {S \tfun T} f}{
+\ \ \ \ \mlapp{\abs f {S \tfun S} f}{
+\ \ \  \ \ \abs x S x}}}}}
+\end{align*}
+
+The idea is to start with a function from $S \tfun S$ and cast it
+successively to $S \tfun T$ then $S \tfun U$. To typecheck the
+expression initially, we need to check $S \sub T$ and $T \sub
+U$. After some reduction steps, the first few casts vanish, and the
+reduced expression casts directly from $S \tfun S$ to $S \tfun U$, so
+we need to check $S \sub U$. Thus, we need subtyping transitivity: $S
+\sub T$ and $T \sub U$ implies $S \sub U$.
+
+\subsection{Non-Expanding Types and Subtyping Transitivity}
+
+The well-formedness rule $\textsc{tsel-wf$_2$}$ allow a type selection
+$p.L$ to refer back to itself in its upper bound. This is useful for
+expressing recursive class types and F-bounded abstract types.
+
+Consider the simplest possible type that refers to itself: $T = \Top
+\refine z {\Ldecl L \Bot z.L}$. This type is well-formed but
+non-expanding.
+
+Non-expanding types are problematic for subtyping transitivity.
 

### 2012-06-30 doc/semester/report.tex
non-expanding types and subtyping transitivity

@@
 
-\subsection{Subtyping Transitivity and Preservation}
+\subsection{Subtyping Transitivity and Preservation}\label{subpres}
 
@@
 counterexample to subtyping transitivity where the three involved
-types $S, T, U$ are expressible within a universe $u$ but not
-necessarily realizable:
+types $S, T, U$ are expressible within a realizable universe, though
+the types themselves don't need to be realizable:
 
@@
 
-Consider the simplest possible type that refers to itself: $T = \Top
-\refine z {\Ldecl L \Bot z.L}$. This type is well-formed but
-non-expanding.
+Consider the simplest possible type that refers to itself: $p \typ T =
+\Top \refine z {\Ldecl L \Bot z.L}$. Now, $p.L$ is well-formed but
+non-expanding. Indeed, the only expansion rule that applies to $p.L$
+is $\textsc{tsel-$\expand$}$, but it requires expanding the upper
+bound... $p.L$! Thus, there is no finite derivation for the expansion
+of $p.L$.
 
-Non-expanding types are problematic for subtyping transitivity.
+Non-expanding types are problematic for subtyping
+transitivity. Consider: $p.L \sub \Top$ by \textsc{$\sub$-$\Top$} and
+$\Top \sub \Top \refine z {}$ by \textsc{$\sub$-rfn}. By transitivity
+on $\Top$, we expect $p.L \sub \Top \refine z {}$, but we cannot infer
+this from the rules. The only rules that apply are \textsc{$\sub$-rfn}
+and \textsc{tsel-$\sub$}. \textsc{$\sub$-rfn} gets stuck expanding
+$p.L$. \textsc{tsel-$\sub$} gets stuck because the conclusion is also
+a premise as $p \ni \Ldecl L \Bot p.L$.
+
+For this counterexample, it seems enough to just add transitivity on
+$\Top$ as a built-in rule. But this doesn't solve the general issue
+that non-expanding types don't play well with subtyping. The problem
+is deep, as it's possible to construct a realizable universe with
+three non-trivial types that fail subtyping transitivity.
+
+Consider an environment where $u$ is bound to:
+\begin{align*}
+\Top & \mlrefine u {\\
+&\ \Ldecl {\mathit{Bad}} {\Bot} {u.\mathit{Bad}}\\
+&\ \Ldecl {\mathit{Good}} {\Top \refine z {\Ldecl L \Bot \Top}} {\Top \refine z {\Ldecl L \Bot \Top}}\\
+&\ \Ldecl {\mathit{Lower}} {u.\mathit{Bad} \tand u.\mathit{Good}} {u.\mathit{Good}}\\
+&\ \Ldecl {\mathit{Upper}} {u.\mathit{Good}} {u.\mathit{Bad} \tor u.\mathit{Good}}\\
+&\ \Ldecl X {u.\mathit{Lower}} {u.\mathit{Upper}}\\
+}&
+\end{align*}
+Note that each lower bound is a subtype of its upper bound, so $u$ is
+realizable. Hence, if we find a counterexample to subtyping
+transitivity with types expressible within $u$, we can apply the trick
+from section~\ref{subpres} to create a counterexample to preservation.
+
+Indeed, here is such a counterexample to subtyping transitivity:
+\begin{align*}
+S &= u.\mathit{Bad} \tand u.\mathit{Good}\\
+T &= u.\mathit{Lower}\\
+U &= u.X \refine z {\Ldecl L \Bot \Top}
+\end{align*}
+We have $S <: T$ and $T <: U$, but we cannot derive $S <: U$ because
+$S$ doesn't expand.
 

### 2012-06-30 doc/semester/report.tex
functions as objects

@@
 \begin{align*}
-\mlnew u \ldots {
+&\mlnew u \ldots {
 \ \mlapp{\abs x \Top x}{
@@
 
-\subsection{Non-Expanding Types and Subtyping Transitivity}
+\subsection{Non-Expanding Types and Subtyping Transitivity}\label{nonexp}
 
@@
 
+\subsection{Narrowing}\label{narrowing}
+
+\subsubsection{Functions as Objects}\label{fun}
+
+The original DOT calculus has both object and function types. It
+doesn't have built-in methods, but they can be expressed as functional
+labels. On the other hand, Scala doesn't have a built-in function
+type. Instead, it has both methods and fields, and encodes functions
+as objects with an apply method. Here, I show that the DOT model as
+originally designed is problematic.
+
+A concrete object can be a subtype of a function type without a
+function ever being defined. Consider:
+\begin{align*}
+&\mlnew u {\Top \refine z {\Ldecl C {\Top \tfun \Top} {\Top \tfun \Top}} \ldefs{}} {
+\mlnew f {u.C \ldefs{}} {
+\ldots
+}}
+\end{align*}
+
+Now, $f$ is a subtype of $\Top \tfun \Top$, but $\app f {(\abs x \Top
+  x)}$ is stuck (and, rightfully, doesn't typecheck). But we can use
+narrowing to create a counterexample to preservation: $\app {(\abs g
+  {\Top \tfun \Top} {g (\abs x \Top x)})} f$.
+
 \acks

### 2012-06-30 doc/semester/report.tex
toc
term-mem restriction

@@
 
+\subsubsection{\texorpdfstring{\textsc{term-$\ni$}}{Term-Mem} Restriction}\label{term_mem}
+
+There are two membership ($t \ni D$) rules: one for when the term $t$
+is a path, and one for an arbitrary term $t$. For paths, we can
+substitute the self-references in the declarations, but we cannot do
+so for arbitrary terms as the resulting types wouldn't be well-formed
+syntactically. Hence, the $\textsc{term-$\ni$}$ has the restriction
+that self-ocurrences are not allowed. Here is a counterexample related
+to this restriction.
+
+Let $X$ be a shorthand for the type:
+\begin{align*}
+\Top & \mlrefine z {\\
+&\ \Ldecl {L_a} \Top \Top\\
+&\ \ldecl l {z.L_a}\\
+}&
+\end{align*}
+
+Let $Y$ be a shorthand for the type:
+\begin{align*}
+\Top & \mlrefine z {\\
+&\ \ldecl l \Top\\
+}&
+\end{align*}
+
+Now, consider the term
+\begin{align*}
+&\mlnew u {X \ldefs{ l = u }}{
+(\app {(\abs y {\Top \tfun Y} {\app y u})} {(\abs d \Top {(\app {(\abs x X x)} u)})}).l
+}
+\end{align*}
+
+The term type-checks because the $t=(\app {(\abs y {\Top \tfun Y}
+  {\app y u})} {(\abs d \Top {(\app {(\abs x X x)} u)})})$ has type
+$Y$, so we can apply $\textsc{term-$\ni$}$ for $l$. However, the term
+$t$ eventually steps to $(\app {(\abs x X x)} u)$ which has type $X$,
+so we cannot apply $\textsc{term-$\ni$}$ for $l$ because of the
+self-reference ($z.L_a$).
+
+\subsubsection{Expansion Lost}\label{narrowing_exp}
+
+\subsubsection{Well-Formedness Lost}\label{narrowing_wf}
+
+\subsection{Path Equality}\label{patheq}
+
 \acks

### 2012-06-30 doc/semester/report.tex
expansion lost
well-formedness counterexample

@@
-\documentclass[preprint,9pt]{sigplanconf}
+\documentclass[preprint]{sigplanconf}
 
@@
 
+Expansion is not preserved by narrowing.
+
+\begin{align*}
+&\mlnew {x_0} {\Top \mlrefine z { \Ldecl {A_1} \Bot {\Top \mlrefine z {\\
+&\gap\Ldecl {A_2} \Bot \Top\\
+&\gap\Ldecl {A_3} \Bot \Top\\
+&\gap\Ldecl {C_2} \Bot {z.A_2}}}}\ldefs{}}{
+\mlnew {x_1} {\Top \refine z {\Ldecl {C_1} \Bot {\Top \refine z {\Ldecl {A_1} \Bot {x_0.A_1}}}}\ldefs{}}}{
+\mlnew {x_2} {{x_1.C_1} \refine z {\Ldecl {A_1} \Bot {{x_0.A_1} \refine z {\Ldecl {A_2} \Bot {z.A_3}}}}\ldefs{}}}{
+\mlnew {x_3} {{x_1.C_1} \refine z {\Ldecl {A_1} \Bot {{x_0.A_1} \refine z {\Ldecl {A_3} \Bot {z.A_2}}}}\ldefs{}}}{
+\mlapp {\abs x {x_1.C_1} {(\abs {z_0} {x.A_1 \tand x_3.A_1} {\new z {z_0.C_2} {\app {(\abs x \Top x)} z}})}} {\ x_2}}
+\end{align*}
+
 \subsubsection{Well-Formedness Lost}\label{narrowing_wf}
+
+Even well-formedness is not preserved by narrowing. The trick is that
+if the lower bound of a type selection is not $\Bot$, then the upper
+bound needs to be checked for well-formedness. Here, we create two
+type selections that are mutually recursive in their upper bounds
+after narrowing.
+
+\begin{align*}
+&\mlnew v {\Top \refine z {\Ldecl L \Bot {\Top \refine z {\Ldecl A \Bot \Top, \Ldecl B {z.A} {z.A}}}} \ldefs{}}{
+\mlapp {\abs x {\Top \refine z {\Ldecl L \Bot {\Top \refine z {\Ldecl A \Bot \Top, \Ldecl B \Bot \Top}}}} {\\&\gap
+\mlnew z {\Top \refine z {\ldecl l {\Bot \tfun \Top}}\mlldefs{\\&\gap\gap l = \abs y {x.L \tand {\Top \refine z {\Ldecl A {z.B} {z.B}, \Ldecl B \Bot \Top}}} {\\&\gap\gap\gap\abs a {y.A} {\app {(\abs x \Top x)} a}}}}{
+\ \app {(\abs x \Top x)} z
+}}}{\ v}}
+\end{align*}
 

### 2012-06-30 doc/semester/report.tex
path equality

@@
 
+We need to be able to relate path-dependent types after reduction. The
+original DOT calculus doesn't have any rules for dealing with
+path-equality. Here is an example which motivates the need for
+path-equality provisions.
+
+\begin{align*}
+&\mlnew b {\Top \mlrefine z {&&\Ldecl X \Top \Top&\\
+&&&\ \ldecl l {z.X}&}\ldefs{l = b}}{
+\mlnew a {\Top \mlrefine z {\ldecl i {\Top \mlrefine z {&&&\\
+&&&\Ldecl X \Bot \Top\\
+&&&\ldecl l {z.X}}}&}\ldefs{i = b}}}{
+\app {(\abs x \Top x)} {(\app {(\abs x {a.i.X} x)} {a.i.l})}}
+\end{align*}
+
+$a.i.l$ reduces to $b.l$. $b.l$ has type $b.X$, so we need $b.X <:
+a.i.X$. This cannot be established with the current rules: it is not
+true in general, but true here because $a.i$ reduces to $b$. Hence,
+the need for acknowledging path equality.
+
 \acks

### 2012-06-30 doc/semester/report.tex
rev

@@
 
-The term type-checks because the $t=(\app {(\abs y {\Top \tfun Y}
+The term type-checks because the term $t=(\app {(\abs y {\Top \tfun Y}
   {\app y u})} {(\abs d \Top {(\app {(\abs x X x)} u)})})$ has type
@@
 
-Expansion is not preserved by narrowing.
+Expansion is not preserved by narrowing. Here, we create two type
+selections that are mutually recursive in their upper bounds after
+narrowing: $z_0.C2$ initially expands, but after narrowing, $z_0.C_2$
+expands to what $z_0.A_2$ expands to, which expands to what $z_0.A_1$
+expands to, which expands to what $z_0.A_2$ expands to, and thus we
+have an infinite expansion. Thus, the last new expression initially
+type-checks, but after narrowing, it doesn't because the precise
+expansion needed by $\textsc{new}$ cannot be inferred.
 
@@
 type selections that are mutually recursive in their upper bounds
-after narrowing.
+after narrowing. $y.A$ is initially well-formed, but after narrowing,
+it isn't because we run into an infinite derivation trying to prove
+the well-formedness of its upper bound.
 

### 2012-06-30 doc/semester/report.tex
wfe

@@
 
-Some subtyping transitivity is essential for soundness. Indeed, I show
+Some subtyping transitivity is essential for soundness. Indeed, we show
 how to construct a counterexample to preservation from any
@@
 type. Instead, it has both methods and fields, and encodes functions
-as objects with an apply method. Here, I show that the DOT model as
+as objects with an apply method. Here, we show that the DOT model as
 originally designed is problematic.
@@
 
+\section{Patches}
+
+We have explored several patches to the calculus to deal with the
+counterexamples presented above.
+
+\subsection{Well-Formed and Expanding Types}\label{wfe}
+
+We introduce a new judgement form for whether a type is well-formed
+and expanding: $\Gamma \ts T \wfe$ if and only if $\Gamma \ts T \wf$
+and $\exists \seq D$ such that $\Gamma \ts T \expand_z \seq D$. Then,
+we replace all other uses of the $\wf$ judgment, including those
+within the $\wf$ judgment inference rules, with uses of the $\wfe$
+judgment.
+
+We limit subtyping to $\wfe$ types. In fact, we make subtyping regular
+with respect to $\wfe$: i.e., if $\Gamma \ts S <: T$, then $\Gamma \ts
+S \wfe$ and $\Gamma \ts T \wfe$. Assuming a few lemmas, we are able to
+formally prove that this modified subtyping judgment is
+transitive. The most notable lemma we assume is the ``Galois
+connection'' between subtyping and expansion: if $\Gamma \ts S \sub T
+\spcomma S \expand_z \seq D_S \spcomma T \expand_z\seq D_T$, then
+$\Gamma \envplus{z: S} \ts \seq{D_s \sub D_T}$.
+
 \acks

### 2012-06-30 doc/semester/report.tex
more patches

@@
 
+\subsection{Functions as Sugar}\label{funsug}
+
+We adopt Scala's model of treating functions as syntactic sugar for an
+object with a special method. In order to do so, we introduce a new
+kind of label for methods with one parameter: $m : S \tfun U$.
+
+A difference in expressivity between this model and the original one
+is that we now have to explicitly provide a return type of the method,
+while the return type was inferred for functions. In practice, this is
+cumbersome but not fundamentally limiting.
+
+\subsection{Explicit Widening}\label{widening}
+
+In the original DOT calculus, the \textsc{app} and \textsc{new} typing
+rules have implicit relaxations. For instance, in \textsc{app}, the
+argument type may be a subtypine of the declared parameter type. In
+order to deal with all the soundness problems due to narrowing, we
+make widening an explicit operation and changes those rules to be
+strict by replacing those relaxed subtyping judgments with equality
+judgments. Two types $S$ and $T$ are judged to be equal if $S <: T$
+and $T <: S$.
+
 \acks

### 2012-06-30 doc/semester/report.tex
first draft of patches sec

@@
 rules have implicit relaxations. For instance, in \textsc{app}, the
-argument type may be a subtypine of the declared parameter type. In
+argument type may be a subtype of the declared parameter type. In
 order to deal with all the soundness problems due to narrowing, we
-make widening an explicit operation and changes those rules to be
+make widening an explicit operation and change those rules to be
 strict by replacing those relaxed subtyping judgments with equality
@@
 and $T <: S$.
+
+Syntactically, we add a widening term: $\wid t T$, and extend values
+with a case for widening: $\wid v T$. The typing rule for widening,
+\textsc{wid}, is the only one admitting a subtyping relaxation:
+$\Gamma \ts (\wid t T) \typ T$ if $\Gamma \ts t : T'$ and $\Gamma \ts T'
+\sub T$.
+
+The reduction rules become more complicated because the type
+information in the widening needs to be propagated correctly. We will
+motivate this informally with examples.
+
+\begin{align*}
+&\mlnew v {\Top \refine z {\Ldecl {L_a} \Bot \Top, \ldecl l {\Top \refine z {\Ldecl {L_a} \Bot \Top}}}\\
+&\gap\gap\ldefs{l = \wid v {\Top \refine z {\Ldecl {L_a} \Bot \Top}}}}{
+\app {(\abs x \Top x)} {(\wid v {\Top \refine z {\ldecl l \Top}}).l}
+}
+\end{align*}
+
+The term $(\wid v {\Top \refine z {\ldecl l \Top}}).l$ first widens
+$v$ so that the label has type $\Top$ instead of $\Top \refine z
+{\Ldecl {L_a} \Bot \Top}$.
+
+How should reduction proceed? We cannot just strip the widening and
+then reduce, because then the strict function application would not
+accept the reduced term. In short, we need to do some type
+manipulations during reduction, by using the membership and expansion
+judgments. This is a bit unfortunate, because it means that reduction
+now needs to know about typing.
+
+Next, we look at path equality provisions. These are even more
+essential now in the presence of explicit widening. Consider this
+example:
+
+\begin{align*}
+&\mlnew b {\Top \refine z {\Ldecl X \Top \Top, \ldecl l {z.X}} \ldefs{l = \wid b {b.X}}}{
+\mlnew a {\Top \refine z {\ldecl i {\Top \refine z {\Ldecl X \Bot \Top, \ldecl l {z.X}}}}\\
+&\gap\gap\ldefs{i = \wid b {\Top \refine z {\Ldecl X \Bot \Top, \ldecl l {z.X}}}}}}{
+\wid {a.i.l} \Top
+}
+\end{align*}
+
+$a.i.l$ reduces to $\wid b {\Top \refine z {\Ldecl X \Bot \Top, \ldecl
+    l {z.X}}}$. Now, how can we continue? $b.l$ reduces to $b : b.X$
+which has bounds $\Top..\Top$, but $(\wid b {\Top \refine z {\Ldecl X
+    \Bot Top, \ldecl l {z.X}}}).l$ has bounds $\Bot..\Top$, so without
+some provision for path equality, we cannot widen $b.l$ to $(\wid b
+{\Top \refine z {\Ldecl X \Bot \Top, \ldecl l {z.X}}}).l$.
+
+\subsection{Path Equality Provisions}
+
+We add the path equality provisions to the subtyping rules.
+
+Let's first ignore the extension of the calculus requiring explicit
+widenings introduced in~\ref{widening}. Then, we need to add one
+intuitive rule to the subtyping judgment: $\textsc{$\sub$-path}$. If
+$p$ (path-)reduces to $q$, and $T \sub q.L$, then $T \sub p.L$. Path
+reduction is a simplified form of reduction involving only
+paths. However, this means that the subtyping judgment, and
+indirectly, all the typing-related judgments, now need to carry the
+store in addition to the context so that path reductions can be
+calculated.
+
+Now, let's see how path equality provisions and explicit widening can
+fit together.
+
+First, path reduction is not isomorphic to reduction anymore, since we
+actually want to actually skip over widenings, as motivated by the
+last example in~\ref{widening}.
+
+In addition, we now also need a dual rule, $\textsc{path-$\sub$}$: if
+$p$ (path-dually)-reduces to $q$, and $q.L \sub T$ then $p.L \sub
+T$. This is because when we have a widening on an object on which a
+method is called, we have to upcast the argument to the parameter type
+expected by the original method. Here is a motivating example.
+
+Let $T_c$ be a shorthand for the type:
+\begin{align*}
+\Top &\mlrefine z {\\
+&\ \Ldecl A {\Top \refine z {\mdecl m \Bot \Top}} \Top\\
+&\ \Ldecl B \Top \Top\\
+&\ \mdecl m {z.A} \Top\\
+}&
+\end{align*}
+
+Let $T$ be a shorthand for the type:
+\begin{align*}
+\Top &\mlrefine z {\\
+&\ \Ldecl A {\Top \refine z {\mdecl m \Bot \Top}} \Top\\
+&\ \Ldecl B \Top \Top\\
+&\ \mdecl m {\Top \refine z {\Ldecl B \Top \Top}} \Top\\
+}&
+\end{align*}
+
+Now, consider the term:
+\begin{align*}
+&\mlnew v {Tc \ldefs{m(x)=\wid x \Top}}{
+(\wid v T).m(\wid v {(\wid v {A \refine z {\Ldecl B \Top \Top}})})
+}
+\end{align*}
+
+When we evaluate the method invocation, we need to cast $\wid v {(\wid
+  v {A \refine z {\Ldecl B \Top \Top}})}$ to $v.A$, and for this, we
+need the newly introduced $\textsc{path-$\sub$}$ rule.
+
+Note that the path dual reduction can be a bit stricter with casts
+than the path reduction. In any case, introducing this
+$\textsc{path-$\sub$}$ rule into the subtyping judgment is
+problematic: it is now possible to say $p.L \sub T$, even though $T$
+can do more than what $p.L$ is defined to do. Here is an example,
+where we construct an object, with $T = \Bot$. (The convolution in the
+example is due to the requirement that concrete types be only
+mentioned once.)
+
+\begin{align*}
+&\mlnew a {\Top \refine z {\Ldecl C \Bot {\Top \refine z {\Ldecl D \Bot {z.X}, \Ldecl X \Bot \Top}}}}{
+\mlnew b {{a.C} \refine z {\Ldecl X \Bot \Bot}}}{
+\mlnew c {a.C}}{
+\mlnew d {(\wid b {a.C}).D}}{
+\app {(\abs x \Bot {x.\mathit{foo}})} d
+}
+\end{align*}
+
+Notice that $d$ has type $\Bot$ if you ginore the cast on $b$. This
+example doesn't typecheck initially because $\textsc{path-$\sub$}$
+only applies when objects are in the store, so the application is not
+well-typed. But if we start preservation in a store which has $a$,
+$b$, $c$ and $d$ then the application type-checks, because, through
+$\textsc{path-$\sub$}$, we can find that the type of $d$ is a subtype
+of $\Bot$. Now, of course, when we get to $d.\mathit{foo}$, reduction
+fails.
+
+So the preservation theorem as defined on a small-step semantics
+(where we start with an arbitrary well-formed environment) fails when
+we add the $\textsc{path-$\sub$}$ rule.
+
+\section{Conclucion and Future Work}
 

### 2012-06-30 doc/semester/report.tex
wip: example

@@
 
-% TODO
+\subsection{Program Examples}
+
+\begin{align*}
+&\mlnew {\mi{root}} {\Top \mlrefine r {\\
+&\ \Ldecl {\mi{Unit}} \Bot \Top\\
+&\ \ldecl {\mi{unit}} {\Top \tfun {.\mi{Unit}}}\\
+&\ \Ldecl {\mi{Boolean}} \Bot {\Top \mlrefine z {\\
+&\ \gap \ldecl {\mi{ifNat}} {(r.\mi{Unit} \tfun r.\mi{Nat}) \tfun (r.\mi{Unit} \tfun r.\mi{Nat}) \tfun r.\mi{Nat}}\\
+&\gap}}\\
+&\ \ldecl {\mi{false}} {{r.\mi{Unit}} \tfun {r.\mi{Boolean}}}\\
+&\ \ldecl {\mi{true}} {{r.\mi{Unit}} \tfun {r.\mi{Boolean}}}\\
+&\ \Ldecl {\mi{Nat}} \Bot {\Top \mlrefine z {\\
+&\ \gap \ldecl {\mi{isZero}} {r.\mi{Unit} \tfun r.\mi{Boolean}}\\
+&\ \gap \ldecl {\mi{pred}} {r.\mi{Unit} \tfun r.\mi{Nat}}\\
+&\ \gap \ldecl {\mi{succ}} {r.\mi{Unit} \tfun r.\mi{Nat}}\\
+&\gap}}\\
+&\ \ldecl {\mi{zero}} {{r.\mi{Unit}} \tfun {r.\mi{Nat}}}\\
+&\ \ldecl {\mi{successor}} {{r.\mi{Nat}} \tfun {r.\mi{Nat}}}\\
+&\ \ldecl {\mi{error}} {{r.\mi{Unit}} \tfun \Bot}\\
+&}\mlldefs{\\\displaybreak[3]
+&\ \mi{unit}  = \abs x \Top {\new u {\mi{root}.\mi{Unit}} u}\\
+%
+&\ \mi{false} = \abs u {\mi{root}.\mi{Unit}} {\\
+&\ \gap \mlnew {\mi{ff}} {\mi{root}.\mi{Boolean} \mlldefs{\\
+&\ \gap\gap {\mi{ifNat}} = \abs t {\mi{root}.\mi{Unit} \tfun \mi{root}.\mi{Nat}} {\abs e {\mi{root}.\mi{Unit} \tfun \mi{root}.\mi{Nat}}\\&\gap\gap\gap\app e {\mi{root}.\mi{unit}}}\\
+&\ \gap}}{
+\ \gap \mi{ff}}
+&}\\
+%
+&\ \mi{true} = \abs u {\mi{root}.\mi{Unit}} {\\
+&\ \gap \mlnew {\mi{t}} {\mi{root}.\mi{Boolean} \mlldefs{\\
+&\ \gap\gap {\mi{ifNat}} = \abs t {\mi{root}.\mi{Unit} \tfun \mi{root}.\mi{Nat}} {\abs e {\mi{root}.\mi{Unit} \tfun \mi{root}.\mi{Nat}}\\&\gap\gap\gap\app t {\mi{root}.\mi{unit}}}\\
+&\ \gap}}{
+\ \gap \mi{tt}}
+&}\\
+%
+&}}{
+foo\\
+&foo
+}
+\end{align*}
 

### 2012-07-01 doc/semester/report.tex
program excerpts

--- 2012-06-30 doc/semester/report.tex
@@
 
-% TODO
+\subsection{Program Examples}
+
+\begin{align*}
+&\mlnew {\mi{root}} {\Top \mlrefine r {\\
+&\ \Ldecl {\mi{Unit}} \Bot \Top\\
+&\ \ldecl {\mi{unit}} {\Top \tfun {r.\mi{Unit}}}\\
+&\ \Ldecl {\mi{Boolean}} \Bot {\Top \mlrefine z {\\
+&\ \gap \ldecl {\mi{ifNat}} {(r.\mi{Unit} \tfun r.\mi{Nat}) \tfun (r.\mi{Unit} \tfun r.\mi{Nat}) \tfun r.\mi{Nat}}\\
+&\gap}}\\
+&\ \ldecl {\mi{false}} {{r.\mi{Unit}} \tfun {r.\mi{Boolean}}}\\
+&\ \ldecl {\mi{true}} {{r.\mi{Unit}} \tfun {r.\mi{Boolean}}}\\
+&\ \ldots\\
+&\ \ldecl {\mi{error}} {{r.\mi{Unit}} \tfun \Bot}\\
+&}\mlldefs{\\\displaybreak[3]
+&\ \mi{unit}  = \abs x \Top {\new u {\mi{root}.\mi{Unit}} u}\\
+%
+&\ \mi{false} = \abs u {\mi{root}.\mi{Unit}} {\\
+&\ \gap \mlnew {\mi{ff}} {\mi{root}.\mi{Boolean} \mlldefs{\\
+&\ \gap\gap {\mi{ifNat}} = \abs t {\mi{root}.\mi{Unit} \tfun \mi{root}.\mi{Nat}} {\\&\gap\gap\gap\abs e {\mi{root}.\mi{Unit} \tfun \mi{root}.\mi{Nat}}\\&\gap\gap\gap\app e {\mi{root}.\mi{unit}}}\\
+&\ \gap}}{
+\ \gap \mi{ff}}
+&}\\
+%
+&\ \mi{true} = \ldots\\
+%
+&\ \mi{error} = \abs u {\mi{root}.\mi{Unit}} {\app {\mi{root}.\mi{error}} u}\\
+&}}{
+\ldots
+}
+\end{align*}
+
+\begin{align*}
+&\mlnew {\mi{genLists}} {\Top \mlrefine g {\Ldecl {\mi{ListPackage}} \Bot {\Top \mlrefine p {\\
+&\ \Ldecl {\mi{Elem}} \Bot \Top\\
+&\ \Ldecl {\mi{ListOfElement}} {g.\mi{List} \refine z {\Ldecl {\mi{Elem}} {p.\mi{Elem}} {p.\mi{Elem}}}} {\\&\gap\gap\gap\gap g.\mi{List} \refine z {\Ldecl {\mi{Elem}} {p.\mi{Elem}} {p.\mi{Elem}}}}\\
+&\ \Ldecl {\mi{List}} \Bot {\Top \mlrefine z {\\
+&\ \gap \ldecl {\mi{isEmpty}} {\mi{root}.\mi{Unit} \tfun \mi{root}.\mi{Boolean}}\\
+&\ \gap \ldecl {\mi{head}} {\mi{root}.\mi{Unit} \tfun g.\mi{Elem}}\\
+&\ \gap \ldecl {\mi{tail}} {\mi{root}.\mi{Unit} \tfun g.\mi{ListOfElem}}\\
+&\gap}}\\
+&\ \ldecl {\mi{nil}} {\mi{root}.\mi{Unit} \tfun g.\mi{ListOfElem}}\\
+&\ \ldecl {\mi{cons}} {g.\mi{Elem} \tfun g.\mi{ListOfElem}}\\
+&}}}}{
+\mlnew {\mi{natLists}} {{\mi{genLists}.\mi{ListPacakge}} \mlrefine p { \Ldecl {\mi{Elem}} {\mi{Nat}} {\mi{Nat}} }\mlldefs{\\
+&\ {\mi{nil}} = \ldots\\
+&\ {\mi{cons}} = \ldots\\
+&}}}{
+\ldots
+}
+\end{align*}
 

### 2012-07-01 doc/semester/report.tex
ex section

--- 2012-06-30 doc/semester/report.tex
@@
 
-% TODO
+We briefly summarize the salient features of the original DOT
+calculus. The reader is encouraged to consult the appendices for the
+full description. We will often refer to the rules of the calculus
+in the remainder of this report.
+
+The terms in DOT consists of variables $x$, $y$, $z$, lambda
+abstractions $\abs x T t$, function application $\app t t'$, field
+selections $t.l$ and object creation expressions $\new x c t$, where
+$c$ is a constructor $T_c \ldefs{\seq{l = v}}$.
+
+The types in DOT consists of type selections $p.L$, refinement types
+$T \refine z {\seq{D}}$, function types $T \tfun T'$, type intersections
+$T \tand T'$, type unions $T \tor T'$, a top type $\Top$, a bottom
+type $\Bot$.
+
+The DOT calculus is expressed using a small-step operational
+semantics. The typing judgement directly or indirectly uses the
+following other judgements: membership, expansion, well-formedness and
+subtyping. Expansion ``flattens'' all the declarations of a type, and
+membership selects a particular flattened declaration by
+label. Expansion is precise. The \textsc{new} typing judgement relies
+on the preciseness of expansion to check that all bounds of types
+declared within the type of the constructor are consistent (i.e. that
+the lower bound is a subtype of the upper bound), and that all values
+declared within the type of the constructor are initialized.
+
+\subsection{Program Examples}
+
+\subsubsection{Basics: Booleans, Error, \ldots}
+This program defines a $\mi{root}$ object with basic types
+($\mi{Unit}$, $\mi{Boolean}$, $\mi{Nat}$) and values ($\mi{true}$,
+$\mi{false}$, $\mi{error}$, $\mi{zero}$). For brevity, I've omitted
+the code related to natural numbers. During the object creation, the
+value labels such as $\mi{false}$ are all initialized.
+
+\begin{align*}
+&\mlnew {\mi{root}} {\Top \mlrefine r {\\\displaybreak[3]
+&\ \Ldecl {\mi{Unit}} \Bot \Top\\
+&\ \ldecl {\mi{unit}} {\Top \tfun {r.\mi{Unit}}}\\
+&\ \Ldecl {\mi{Boolean}} \Bot {\Top \mlrefine z {\\
+&\ \gap \ldecl {\mi{ifNat}} {(r.\mi{Unit} \tfun r.\mi{Nat}) \tfun (r.\mi{Unit} \tfun r.\mi{Nat}) \tfun r.\mi{Nat}}\\
+&\gap}}\\
+&\ \ldecl {\mi{false}} {{r.\mi{Unit}} \tfun {r.\mi{Boolean}}}\\
+&\ \ldecl {\mi{true}} {{r.\mi{Unit}} \tfun {r.\mi{Boolean}}}\\
+&\ \ldots\\\displaybreak[3]
+&\ \ldecl {\mi{error}} {{r.\mi{Unit}} \tfun \Bot}\\
+&}\mlldefs{\\\displaybreak[3]
+&\ \mi{unit}  = \abs x \Top {\new u {\mi{root}.\mi{Unit}} u}\\
+%
+&\ \mi{false} = \abs u {\mi{root}.\mi{Unit}} {\\
+&\ \gap \mlnew {\mi{ff}} {\mi{root}.\mi{Boolean} \mlldefs{\\\displaybreak[3]
+&\ \gap\gap {\mi{ifNat}} = \abs t {\mi{root}.\mi{Unit} \tfun \mi{root}.\mi{Nat}} {\\&\gap\gap\gap\abs e {\mi{root}.\mi{Unit} \tfun \mi{root}.\mi{Nat}}\\&\gap\gap\gap\app e {\mi{root}.\mi{unit}}}\\
+&\ \gap}}{
+\ \gap \mi{ff}}
+&}\\
+%
+&\ \mi{true} = \ldots\\
+%
+&\ \mi{error} = \abs u {\mi{root}.\mi{Unit}} {\app {\mi{root}.\mi{error}} u}\\
+&\ \ldots\\
+&}}{
+\ldots
+}
+\end{align*}
+
+\subsubsection{Polymorphic Lists}
+Polymorphic lists can be expressed using an abstract type member for
+the element type ($\mi{Elem}$). We can instantiate a refinement of the
+list package to manipulate lists with a particular element type. We
+can choose to make the lists invariant or covariant in the element
+type by fixing the lower bound of the declaration to the upper bound
+or to $\Bot$, respectively.
+
+\begin{align*}
+&\mlnew {\mi{genLists}} {\Top \mlrefine g {\Ldecl {\mi{ListPackage}} \Bot {\Top \mlrefine p {\\\displaybreak[3]
+&\ \Ldecl {\mi{Elem}} \Bot \Top\\
+&\ \Ldecl {\mi{ListOfElement}} {g.\mi{List} \refine z {\Ldecl {\mi{Elem}} {p.\mi{Elem}} {p.\mi{Elem}}}} {\\&\gap\gap\gap\gap g.\mi{List} \refine z {\Ldecl {\mi{Elem}} {p.\mi{Elem}} {p.\mi{Elem}}}}\\
+&\ \Ldecl {\mi{List}} \Bot {\Top \mlrefine z {\\\displaybreak[3]
+&\ \gap \ldecl {\mi{isEmpty}} {\mi{root}.\mi{Unit} \tfun \mi{root}.\mi{Boolean}}\\
+&\ \gap \ldecl {\mi{head}} {\mi{root}.\mi{Unit} \tfun g.\mi{Elem}}\\
+&\ \gap \ldecl {\mi{tail}} {\mi{root}.\mi{Unit} \tfun g.\mi{ListOfElem}}\\
+&\gap}}\\
+&\ \ldecl {\mi{nil}} {\mi{root}.\mi{Unit} \tfun g.\mi{ListOfElem}}\\
+&\ \ldecl {\mi{cons}} {g.\mi{Elem} \tfun g.\mi{ListOfElem}}\\
+&}}}}{
+\mlnew {\mi{natLists}} {{\mi{genLists}.\mi{ListPackage}} \mlrefine p { \Ldecl {\mi{Elem}} {\mi{Nat}} {\mi{Nat}} }\mlldefs{\\\displaybreak[3]
+&\ {\mi{nil}} = \ldots\\
+&\ {\mi{cons}} = \ldots\\
+&}}}{
+\ldots
+}
+\end{align*}
 

### 2012-07-01 doc/semester/report.tex
rev

@@
 
+DOT models Scala's path-dependent types and abstract type members, as
+well as its mixture of nominal and structural typing through the use
+of refinement types. It makes no attempt to model inheritance or
+mixing composition. The calculus does not model what's currently in
+Scala: it is more normative than descriptive.
+
 The terms in DOT consists of variables $x$, $y$, $z$, lambda
@@
 declared within the type of the constructor are initialized.
+
+In the rest of this introduction, we present some program examples. We
+summarize our findings of counterexamples to soundness in
+section~\ref{counterexs}. We explore some patches to the calculus in
+section~\ref{patches}. We discuss opportunities for future work and
+conclude in section~\ref{conclusion}.
 
@@
 
-\section{Counterexamples}
+\section{Counterexamples}\label{counterexs}
 
@@
 
-\section{Patches}
+\section{Patches}\label{patches}
 
@@
 First, path reduction is not isomorphic to reduction anymore, since we
-actually want to actually skip over widenings, as motivated by the
-last example in~\ref{widening}.
+want to actually skip over widenings, as motivated by the last example
+in~\ref{widening}.
 
@@
 
-Notice that $d$ has type $\Bot$ if you ginore the cast on $b$. This
+Notice that $d$ has type $\Bot$ if you ignore the cast on $b$. This
 example doesn't typecheck initially because $\textsc{path-$\sub$}$
@@
 
-\section{Conclucion and Future Work}
+\section{Conclucion and Future Work}\label{conclusion}
+This semester's work was instructive in pointing out the problematic
+aspects of the original DOT calculus, and exploring some possible
+fixes. However, I haven't been able to reach a revised calculus
+without soundness hole. Furthermore, I believe that the revised
+calculus with all the patches is not as elegant as the original
+calculus. Going forward, I want to take a broader view by by studying
+and incorporating ideas from other calculi which models dependent
+types, and path-dependent types in particular.
 

### 2012-07-01 doc/semester/report.tex
formatting, typos

@@
 The DOT calculus is expressed using a small-step operational
-semantics. The typing judgement directly or indirectly uses the
-following other judgements: membership, expansion, well-formedness and
+semantics. The typing judgment directly or indirectly uses the
+following other judgments: membership, expansion, well-formedness and
 subtyping. Expansion ``flattens'' all the declarations of a type, and
 membership selects a particular flattened declaration by
-label. Expansion is precise. The \textsc{new} typing judgement relies
+label. Expansion is precise. The \textsc{new} typing judgment relies
 on the preciseness of expansion to check that all bounds of types
@@
 syntactically. Hence, the $\textsc{term-$\ni$}$ has the restriction
-that self-ocurrences are not allowed. Here is a counterexample related
+that self-occurrences are not allowed. Here is a counterexample related
 to this restriction.
@@
 \mlnew {x_3} {{x_1.C_1} \refine z {\Ldecl {A_1} \Bot {{x_0.A_1} \refine z {\Ldecl {A_3} \Bot {z.A_2}}}}\ldefs{}}}{
-\mlapp {\abs x {x_1.C_1} {(\abs {z_0} {x.A_1 \tand x_3.A_1} {\new z {z_0.C_2} {\app {(\abs x \Top x)} z}})}} {\ x_2}}
+\mlapp {\abs x {x_1.C_1} {(\abs {z_0} {x.A_1 \tand x_3.A_1} {\\&\gap\gap\gap\new z {z_0.C_2} {\app {(\abs x \Top x)} z}})}} {\ x_2}}
 \end{align*}
@@
 
-We introduce a new judgement form for whether a type is well-formed
+We introduce a new judgment form for whether a type is well-formed
 and expanding: $\Gamma \ts T \wfe$ if and only if $\Gamma \ts T \wf$
@@
 
-\section{Conclucion and Future Work}\label{conclusion}
+\section{Conclusion and Future Work}\label{conclusion}
 This semester's work was instructive in pointing out the problematic

### 2012-07-01 doc/semester/report.tex
clarif

@@
 can choose to make the lists invariant or covariant in the element
-type by fixing the lower bound of the declaration to the upper bound
-or to $\Bot$, respectively.
+type by fixing the lower bound of the $\mi{Elem}$ re-declaration to
+the upper bound or to $\Bot$, respectively.
 

### 2012-07-01 doc/semester/report.tex
typo

@@
 calculus. Going forward, I want to take a broader view by by studying
-and incorporating ideas from other calculi which models dependent
+and incorporating ideas from other calculi which model dependent
 types, and path-dependent types in particular.

### 2012-07-02 doc/semester/report.tex
slide progress

@@
 The terms in DOT consists of variables $x$, $y$, $z$, lambda
-abstractions $\abs x T t$, function application $\app t t'$, field
+abstractions $\abs x T t$, function application $\app t {t'}$, field
 selections $t.l$ and object creation expressions $\new x c t$, where

### 2012-07-03 doc/semester/report.tex
slides up to all counterexs

@@
 narrowing to create a counterexample to preservation: $\app {(\abs g
-  {\Top \tfun \Top} {g (\abs x \Top x)})} f$.
+  {\Top \tfun \Top} {\app g (\abs x \Top x)})} f$.
 
@@
 Even well-formedness is not preserved by narrowing. The trick is that
-if the lower bound of a type selection is not $\Bot$, then the upper
-bound needs to be checked for well-formedness. Here, we create two
-type selections that are mutually recursive in their upper bounds
+if the lower bound of a type selection is not $\Bot$, then the
+bounds needs to be checked for well-formedness. Here, we create two
+type selections that are mutually recursive in their bounds
 after narrowing. $y.A$ is initially well-formed, but after narrowing,
 it isn't because we run into an infinite derivation trying to prove
-the well-formedness of its upper bound.
+the well-formedness of its bounds.
 

### 2012-07-04 doc/semester/report.tex
minor
fix

@@
 \textsc{wid}, is the only one admitting a subtyping relaxation:
-$\Gamma \ts (\wid t T) \typ T$ if $\Gamma \ts t : T'$ and $\Gamma \ts T'
+$\Gamma \ts (\wid t T) \typ T$ if $\Gamma \ts t \typ T'$ and $\Gamma \ts T'
 \sub T$.
@@
 \begin{align*}
-&\mlnew v {Tc \ldefs{m(x)=\wid x \Top}}{
-(\wid v T).m(\wid v {(\wid v {A \refine z {\Ldecl B \Top \Top}})})
+&\mlnew v {T_c \ldefs{m(x)=\wid x \Top}}{
+(\wid v T).m(\wid v {((\wid v T).{A \refine z {\Ldecl B \Top \Top}})})
 }
@@
 
-When we evaluate the method invocation, we need to cast $\wid v {(\wid
-  v {A \refine z {\Ldecl B \Top \Top}})}$ to $v.A$, and for this, we
-need the newly introduced $\textsc{path-$\sub$}$ rule.
+When we evaluate the method invocation, we need to cast $\wid v
+{((\wid v T).{A \refine z {\Ldecl B \Top \Top}})}$ to $v.A$, and for
+this, we need the newly introduced $\textsc{path-$\sub$}$ rule.
 

### 2012-07-06 doc/semester/report.tex
fix semantic typo
minor typoes

@@
 The terms in DOT consists of variables $x$, $y$, $z$, lambda
-abstractions $\abs x T t$, function application $\app t {t'}$, field
+abstractions $\abs x T t$, function applications $\app t {t'}$, field
 selections $t.l$ and object creation expressions $\new x c t$, where
@@
 $T \refine z {\seq{D}}$, function types $T \tfun T'$, type intersections
-$T \tand T'$, type unions $T \tor T'$, a top type $\Top$, a bottom
+$T \tand T'$, type unions $T \tor T'$, a top type $\Top$, and a bottom
 type $\Bot$.
@@
 selections that are mutually recursive in their upper bounds after
-narrowing: $z_0.C2$ initially expands, but after narrowing, $z_0.C_2$
+narrowing: $z_0.C_2$ initially expands, but after narrowing, $z_0.C_2$
 expands to what $z_0.A_2$ expands to, which expands to what $z_0.A_1$
@@
 &\ \Ldecl B \Top \Top\\
-&\ \mdecl m {\Top \refine z {\Ldecl B \Top \Top}} \Top\\
+&\ \mdecl m {{z.A} \refine z {\Ldecl B \Top \Top}} \Top\\
 }&

### 2012-08-03 article.tex
template for DOT FOOL submission

%-----------------------------------------------------------------------------
%
%               Template for sigplanconf LaTeX Class
%
% Name:         sigplanconf-template.tex
%
% Purpose:      A template for sigplanconf.cls, which is a LaTeX 2e class
%               file for SIGPLAN conference proceedings.
%
% Guide:        Refer to "Author's Guide to the ACM SIGPLAN Class,"
%               sigplanconf-guide.pdf
%
% Author:       Paul C. Anagnostopoulos
%               Windfall Software
%               978 371-2316
%               paul@windfall.com
%
% Created:      15 February 2005
%
%-----------------------------------------------------------------------------

\documentclass[9pt]{sigplanconf}

% The following \documentclass options may be useful:
%
% 10pt          To set in 10-point type instead of 9-point.
% 11pt          To set in 11-point type instead of 9-point.
% authoryear    To obtain author/year citation style instead of numeric.

\usepackage{amsmath}

\begin{document}

\conferenceinfo{FOOL '12}{October 22, 2012, Tucson, AZ, USA.} 
\copyrightyear{2012} 
\copyrightdata{[to be supplied]}

\title{Dependent Object Types}
\subtitle{A foundations for Scala's type system} % TODO: ask

\authorinfo{Nada Amin \and Adriaan Moors \and Martin Odersky}
           {EPFL}
           {first.last@epfl.ch}

\maketitle

\begin{abstract}
We propose a new type-theoretic foundation of Scala and languages like
it: the Dependent Object Types calculus (DOT). DOT models Scala's
path-dependent types and abstract type members, as well as its mixture
of nominal and structural typing through the use of reﬁnement
types. It makes no attempt to model inheritance or mixing
composition. The calculus does not model what's currently in Scala: it
is more normative than descriptive.

We show that DOT and its patched-up variants are not syntactically
sound, by exhibiting counterexamples to preservation. Nevertheless, we
prove type-safety of the calculus via step-indexed logical relations.
\end{abstract}

\category{D.3.3}{Language Constructs and Features}{Abstract data types, Classes and objects, polymorphism}
\category{D.3.1}{Formal Deﬁnitions and Theory}{Syntax, Semantics}
\category{F.3.1}{Specifying and Verifying and Reasoning about Programs}{}
\category{F.3.3}{Studies of Program Constructs}{Object-oriented constructs, type structure}
\category{F.3.2}{Semantics or Programming Languages}{Operational semantics}

\terms
Languages, Theory, Verification

\keywords
calculus, objects, dependent types, step-indexed logical relations

%\section{Introduction}
%\appendix
%\section{Appendix Title}
%\acks

\bibliographystyle{abbrvnat}
\bibliography{dot}

\end{document}

### 2012-08-05 article.tex
added all rules

@@
 \usepackage{amsmath}
+
+\usepackage{fleqn}
+\usepackage{listings}
+\usepackage{math}
+\usepackage{amsmath}
+\usepackage{latexsym}
+\usepackage{bcprules}
+\usepackage[scaled=0.848971]{luximono} % This is for 11 pt Default font
+\usepackage[T1]{fontenc}
+
+% Prooftree formatting
+\usepackage{prooftree}
+
+\usepackage{multicol}
+\usepackage{framed}
+
+%\usepackage{float}
+%\floatstyle{boxed} 
+%\restylefloat{figure}
+
+% support for generating PDF files
+%\newif\ifpdf
+%    \ifx\pdfoutput\undefined
+%    \pdffalse
+%\else
+%    \pdftrue
+%    \pdfoutput=1
+%\fi
+
+%versions
+% Use dependent function types
+\newif\ifdep\depfalse
+
+\lstset{
+  literate=
+  {=>}{$\Rightarrow\;$}{2}
+  {<:}{$<:\;$}{1}
+}
+
+\lstdefinelanguage{scala}{% 
+       morekeywords={% 
+                try, catch, throw, private, public, protected, import, package, implicit, final, package, trait, type, class, val, def, var, if, for, this, else, extends, with, while, new, abstract, object, case, match, sealed,override},
+         sensitive=t, % 
+   morecomment=[s]{/*}{*/},morecomment=[l]{\//},% 
+   mathescape,
+%   escapeinside={/*\%}{*/},%
+   rangeprefix= /*< ,rangesuffix= >*/,%
+   morestring=[d]{"}% 
+ }
+
+\lstset{breaklines=true,language=scala}
+
+\def\c{\lstinline}  % shorter version so you can write \c|String[Foo]|
+                    % -- \def must be in same file as uses for this to
+                    % work...
+\newcommand{\lstref}[1]{Listing~\ref{#1}}
+\newcommand{\Lstref}[1]{Listing~\ref{#1}} % only capitalise at beginning of sentence?
+\newcommand{\secref}[1]{Section~\ref{#1}}
+\newcommand{\Secref}[1]{Section~\ref{#1}} % only capitalise at beginning of sentence?
+
+% \lstset{basicstyle=\footnotesize\ttfamily, breaklines=true, language=scala, tabsize=2, columns=fixed, mathescape=false,includerangemarker=false}
+% thank you, Burak 
+% (lstset tweaking stolen from
+% http://lampsvn.epfl.ch/svn-repos/scala/scala/branches/typestate/docs/tstate-report/datasway.tex)
+\lstset{
+    xleftmargin=2em,%
+    framesep=5pt,%
+    frame=none,%
+    captionpos=b,%
+    fontadjust=true,%
+    columns=[c]fixed,%
+    keepspaces=false,%
+    basewidth={0.56em, 0.52em},%
+    tabsize=2,%
+    basicstyle=\small\tt,% \small\tt
+    commentstyle=\textit,%
+    keywordstyle=\bfseries,%
+    escapechar=\%,%
+}
+
+%% set latex/pdflatex specific stuff
+%\ifpdf
+    \usepackage[pdftex,
+                hyperindex,
+                plainpages=false,
+                breaklinks,
+                colorlinks,
+                citecolor=black,
+                filecolor=black,
+                linkcolor=black,
+                pagecolor=black,
+                urlcolor=black]{hyperref}
+    \usepackage[pdftex]{graphicx}
+    \DeclareGraphicsExtensions{.jpg,.pdf}
+    \pdfcatalog {
+        /PageMode (/UseNone)
+    }
+    \usepackage{thumbpdf}
+    \usepackage[pdftex]{color}
+%\else
+%    \usepackage[ps2pdf]{hyperref}
+%    \usepackage{graphicx}
+%    \DeclareGraphicsExtensions{.eps,.jpg}
+%    \usepackage{color}
+%\fi
+
+%\setlength{\parindent}{0pt}
+%\setlength{\parskip}{5pt}
+
+% verbfilter stuff
+\newcommand{\prog}[1]{{\sl #1}}
+\newenvironment{program}[1][10.5]
+  {\fontsize{#1}{13.6}\tt\begin{tabbing}\hspace*{0.5\parindent}\=\+\kill}
+  {\end{tabbing}\noindent}
+\newcommand{\blockcomment}[1]{{\color{grayPoint3}#1}}
+\newcommand{\linecomment}{\color{grayPoint3}}
+\newcommand{\grey}{\color{grey}}
+
+%\newenvironment{program}{\ \ \ \ \begin{minipage}{\textwidth}\renewcommand{\baselinestretch}{1.0}\sl\begin{tabbing}}{\end{tabbing}\end{minipage}}
+\newcommand{\vem}{\bfseries}
+\newcommand{\quotedstring}[1]{{#1}}
+\newcommand{\typename}[1]{{#1}}
+\newcommand{\literal}[1]{{#1}}
+
+% comments and notes
+\newcommand{\comment}[1]{}
+%\newcommand{\note}[1]{{\bf $\clubsuit$ #1 $\spadesuit$}}
+
+% figures
+\newcommand{\figurebox}[1]
+        {\fbox{\begin{minipage}{\textwidth} #1 \medskip\end{minipage}}}
+%        {\fbox{\begin{minipage}{\textwidth}\begin{center} #1 \end{center}\medskip\end{minipage}}}
+\newcommand{\boxfig}[3]
+        {\begin{figure*}\figurebox{#3\caption{\label{#1}#2}}\end{figure*}}
+\newcommand{\figref}[1]
+        {Figure~\ref{#1}}
+
+% typing rules (not used here)
+\newcommand{\ttag}[1]{\mbox{\textsc{\small(#1)}}}
+\newcommand{\infer}[3]{\mbox{#1 }\ba{c} #2 \\ \hline #3 \ea}
+\newcommand{\irule}[2]{{\renewcommand{\arraystretch}{1.2}\ba{c} #1 
+                        \\ \hline #2 \ea}}
+\newlength{\trulemargin}
+\newlength{\trulewidth}
+\newlength{\srulewidth}
+\setlength{\trulemargin}{0.80cm}
+\setlength{\trulewidth}{40.0mm}
+\setlength{\srulewidth}{3.0cm}
+\newenvironment{trules}{$\vspace{0.5em}\ba{p{\trulemargin}@{~}p{\trulewidth}@{~}p{\trulemargin}}}{\ea$}
+\newenvironment{srules}{$\vspace{0.5em}\ba{p{\trulemargin}@{~}p{\srulewidth}}}{\ea$}
+\newcommand{\laxiom}[2]{\ttag{#1} & $ #2 \hfill\ }
+\newcommand{\raxiom}[2]{\hfill #2 $& \hfill \ttag{#1}}
+\newcommand{\caxiom}[2]{\ttag{#1} & $\hfill #2 \hfill $& \ }
+\newcommand{\lrule}[3]{\laxiom{#1}{\irule{#2}{#3}}}
+\newcommand{\rrule}[3]{\raxiom{#1}{\irule{#2}{#3}}}
+\newcommand{\crule}[3]{\caxiom{#1}{\irule{#2}{#3}}}
+\newcommand{\lsrule}[3]{\lsaxiom{#1}{\irule{#2}{#3}}}
+\newcommand{\rsrule}[3]{\rsaxiom{#1}{\irule{#2}{#3}}}
+\newcommand{\nl}{\end{trules}\\[0.5em] \begin{trules}}
+\newcommand{\snl}{\end{srules}\\[0.5em] \begin{srules}}
+
+% commas and semicolons
+\newcommand{\comma}{,\,}
+\newcommand{\commadots}{\comma \ldots \comma}
+\newcommand{\semi}{;\mbox{;};}
+\newcommand{\semidots}{\semi \ldots \semi}
+
+% spacing
+\newcommand{\gap}{\quad\quad}
+\newcommand{\biggap}{\quad\quad\quad}
+\newcommand{\nextline}{\\ \\}
+\newcommand{\htabwidth}{0.5cm}
+\newcommand{\tabwidth}{1cm}
+\newcommand{\htab}{\hspace{\htabwidth}}
+\newcommand{\tab}{\hspace{\tabwidth}}
+\newcommand{\linesep}{\ \hrulefill \ \smallskip}
+
+% math stuff
+\newenvironment{myproof}{{\em Proof:}}{$\Box$}
+\newenvironment{proofsketch}{{\em Proof Sketch:}}{$\Box$}
+\newcommand{\Case}{{\em Case\ }}
+
+% make ; a delimiter in math mode
+% \mathcode`\;="8000 % Makes ; active in math mode
+% {\catcode`\;=\active \gdef;{\;}}
+% \mathchardef\semicolon="003B
+
+% reserved words
+\newcommand{\mathem}{\bf}
+
+% brackets
+\newcommand{\set}[1]{\{#1\}}
+\newcommand{\sbs}[1]{\lquote #1 \rquote}
+
+% arrays
+\newcommand{\ba}{\begin{array}}
+\newcommand{\ea}{\end{array}}
+\newcommand{\bda}{\[\ba}
+\newcommand{\eda}{\ea\]}
+\newcommand{\ei}{\end{array}}
+\newcommand{\bcases}{\left\{\begin{array}{ll}}
+\newcommand{\ecases}{\end{array}\right.}
+
+% \cal ids
+\renewcommand{\AA}{{\cal A}}
+\newcommand{\BB}{{\cal B}}
+\newcommand{\CC}{{\cal C}}
+\newcommand{\DD}{{\cal D}}
+\newcommand{\EE}{{\cal E}}
+\newcommand{\FF}{{\cal F}}
+\newcommand{\GG}{{\cal G}}
+\newcommand{\HH}{{\cal H}}
+\newcommand{\II}{{\cal I}}
+\newcommand{\JJ}{{\cal J}}
+\newcommand{\KK}{{\cal K}}
+\newcommand{\LL}{{\cal L}}
+\newcommand{\MM}{{\cal M}}
+\newcommand{\NN}{{\cal N}}
+\newcommand{\OO}{{\cal O}}
+\newcommand{\PP}{{\cal P}}
+\newcommand{\QQ}{{\cal Q}}
+\newcommand{\RR}{{\cal R}}
+\newcommand{\TT}{{\cal T}}
+\newcommand{\UU}{{\cal U}}
+\newcommand{\VV}{{\cal V}}
+\newcommand{\WW}{{\cal W}}
+\newcommand{\XX}{{\cal X}}
+\newcommand{\YY}{{\cal Y}}
+\newcommand{\ZZ}{{\cal Z}}
+
+% misc symbols
+\newcommand{\dhd}{\!\!\!\!\!\rightarrow}
+\newcommand{\Dhd}{\!\!\!\!\!\Rightarrow}
+\newcommand{\ts}{\,\vdash\,}
+\newcommand{\la}{\langle}
+\newcommand{\ra}{\rangle}
+\newcommand{\eg}{{\em e.g.}}
+
+% misc identifiers
+\newcommand{\dom}{\mbox{\sl dom}}
+\newcommand{\fn}{\mbox{\sl fn}}
+\newcommand{\bn}{\mbox{\sl bn}}
+\newcommand{\sig}{\mbox{\sl sig}}
+\newcommand{\IF}{\mbox{\mathem if}}
+\newcommand{\OTHERWISE}{\mbox{\mathem otherwise}}
+\newcommand{\expand}{\prec}
+\newcommand{\weakexpand}{\prec^W}
+\newcommand{\spcomma}{~,~}
+
+%\newcommand{\inst}{\mbox{\mathem inst}}
+\newcommand{\trans}[1]{\la\!\la#1\ra\!\ra}
+\newcommand{\remark}[1]{{\bf $\clubsuit$ #1 $\spadesuit$}}
+\newcommand{\todo}[1]{\remark{to do: #1}}
+%\newcommand{\J}{\justifies}
+%\newcommand{\U}{\using}
+
+% names
+\newcommand{\Scala}{\mbox{\textsc{Scala}}}
+\newcommand{\Java}{\mbox{\textsc{Java}}}
+
+%\renewcommand\textfraction{.05}
+%\renewcommand\floatpagefraction{.9}
+%\renewcommand\topfraction{.8}
+
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+%   Language abstraction commands     %
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+
+%% Relations
+% Subtype 
+\newcommand{\sub}{<:}
+% Type assignment
+\newcommand{\typ}{:}
+% reduction
+\newcommand{\reduces}{\;\rightarrow\;}
+% well-formedness
+\newcommand{\wf}{\;\mbox{\textbf{wf}}}
+\newcommand{\wfe}{\;\mbox{\textbf{wfe}}}
+
+%% Operators
+% Type selection
+\newcommand{\tsel}{\#}
+% Function type
+\newcommand{\tfun}{\rightarrow}
+\newcommand{\dfun}[3]{(#1\!:\!#2) \Rightarrow #3}
+% Conjunction
+\newcommand{\tand}{\wedge}
+% Disjunction
+\newcommand{\tor}{\vee}
+% Singleton type suffix
+\newcommand{\sing}{.\textbf{type}}
+
+%% Syntax
+% Header for typing rules
+\newcommand{\judgement}[2]{{\bf #1} \hfill #2}
+% Refinement
+\newcommand{\refine}[2]{\left\{#1 \Rightarrow #2 \right\}}
+% Field definitions
+\newcommand{\ldefs}[1]{\left\{#1\right\}}
+% Member sequences
+\newcommand{\seq}[1]{\overline{#1}}
+% Lambda
+\newcommand{\dabs}[3]{(#1\!:\!#2)\Rightarrow #3}
+\newcommand{\abs}[3]{\lambda #1\!:\!#2.#3}
+% Application
+\newcommand{\app}[2]{#1\;#2}
+% Method Application
+\newcommand{\mapp}[3]{#1.#2(#3)}
+% Substitution
+\newcommand{\subst}[3]{[#1/#2]#3}
+% Object creation
+\newcommand{\new}[3]{\textbf{val }#1 = \textbf{new }#2 ;\; #3}
+%\renewcommand{\new}[3]{#1 \leftarrow #2 \,\textbf{in}\, #3}
+% Field declaration
+\newcommand{\Ldecl}[3]{#1 : #2..#3}%{#1 \operatorname{>:} #2 \operatorname{<:} #3}
+\newcommand{\ldecl}[2]{#1 : #2}
+\newcommand{\mdecl}[3]{#1 : #2 \tfun #3}
+% Top and Bottom
+\newcommand{\Top}{\top}%{\textbf{Top}}
+\newcommand{\Bot}{\bot}%\textbf{Bot}}
+% Environment extension
+%\newcommand{\envplus}[1]{\uplus \{ #1 \}}
+\newcommand{\envplus}[1]{, #1}
+% Reduction
+\newcommand{\reduction}[4]{#1 \operatorname{|} #2 \reduces #3 \operatorname{|} #4}
+
+\newcommand{\lindent}{\hspace{-4mm}}
 
@@
 
-%\section{Introduction}
+\section{Introduction}
+% TODO
+
+\section{The DOT Calculus}
+
+\boxfig{fig:dot-one}{The DOT Calculus : Syntax, Reduction, Type Assignment}{
+{\bf Syntax}\medskip
+\begin{center}    
+$\ba{l@{\hspace{0.2mm}}|@{\hspace{0.2mm}}l}
+\ba[t]{l@{\hspace{10mm}}l}
+x, y, z    & \lindent{\mbox{Variable}} \\
+l          & \lindent{\mbox{Value label}}\\
+m          & \lindent{\mbox{Method label}}\\[0.2em]
+v ::=      & \lindent{\mbox{Value}} \\
+\gap x     & \mbox{variable} \\[0.2em]
+t ::=      & \lindent{\mbox{Term}} \\
+\gap v     & \mbox{value} \\
+\gap \new x c t & \mbox{new instance} \\
+\gap t.l  & \mbox{field selection} \\
+\gap \mapp t m t  & \mbox{method invocation} \\[0.2em]
+p ::= & \lindent \mbox{Path} \\
+\gap x & \mbox{variable} \\
+\gap p.l & \mbox{selection} \\
+c ::= T_c \ldefs{\seq{l = v}\;\seq{m(x) = t}} & \lindent{\mbox{Constructor}} \\[0.2em]
+\Gamma ::= \seq{x \typ T} & \lindent\mbox{Environment} \\
+s      ::= \seq{x \mapsto c} & \lindent\mbox{Store} \\
+\ea
+&
+\ba[t]{l@{\hspace{10mm}}l}
+L ::=      & \lindent{\mbox{Type label}} \\
+\gap L_c   & \mbox{class label} \\
+\gap L_a   & \mbox{abstract type label} \\[0.2em]
+S,T,U,V,W ::= & \lindent\mbox{Type}\\
+\gap p.L & \mbox{type selection} \\
+\gap T \refine z {\seq D} & \mbox{refinement} \\
+\gap T \tand T & \mbox{intersection type} \\
+\gap T \tor T & \mbox{union type} \\
+\gap \Top  & \mbox{top type} \\
+\gap \Bot  & \mbox{bottom type} \\[0.2em]
+S_c, T_c ::= & \lindent \mbox{Concrete type} \\
+\multicolumn{2}{l}{\gap p.L_c ~|~ T_c \refine z {\seq D} ~|~ T_c \wedge T_c  ~|~ \Top} \\[0.2em]
+D ::= & \lindent\mbox{Declaration} \\
+\gap \Ldecl L S U & \mbox{type declaration} \\
+\gap \ldecl l T   & \mbox{value declaration} \\
+\gap \mdecl m S U & \mbox{method declaration}
+\ea
+\ea$
+\end{center}
+\medskip
+
+\linesep
+
+\begin{multicols}{2}[\judgement{Reduction}{\fbox{$\reduction t s {t'} {s'}$}}]
+
+\infrule[\textsc{msel}]
+{y \mapsto T_c \ldefs{\seq{l = v'}\;\seq{m(x)=t}} \in s}
+{\reduction {\mapp y {m_i} v} s {\subst v {x_i} {t_i}} s}
+
+\infrule[\textsc{sel}]
+{y \mapsto T_c \ldefs{\seq{l = v}\;\seq{m(x)=t}} \in s}
+{\reduction {y.l_i} s {v_i} s}
+
+\infax[\textsc{new}]
+{\reduction {\new x c t} s t {s \envplus{x \mapsto c}}}
+
+\infrule[\textsc{context}]
+{\reduction t s {t'} {s'}}
+{\reduction {e[t]} s {e[t']} s'}
+\end{multicols}
+
+\hfill {\bf where} evaluation context $\gap e ::= [\,] ~|~ \mapp e m t ~|~ \mapp v m e ~|~ e.l\hspace{2cm}$
+
+\linesep
+
+\begin{multicols}{2}[\judgement{Type Assignment}{\fbox{$\Gamma \ts t \typ T$}}]
+
+\infrule[\textsc{var}]
+{x \typ T \in \Gamma}
+{\Gamma \ts x \typ T}
+
+\infrule[\textsc{msel}]
+{\Gamma \ts t \ni m \typ {S \tfun T} \\
+ \Gamma \ts t' \typ T' \spcomma T' \sub S}
+{\Gamma \ts \mapp t m {t'} \typ T}
+
+\infrule[\textsc{sel}]
+{\Gamma \ts t \ni l \typ T'}
+{\Gamma \ts t.l \typ T'}
+
+\infrule[\textsc{new}]
+{
+c = {T_c \ldefs{\seq{l = v}\;\seq{m(x) = t}}}\\
+y \notin \fn(T') \\
+\Gamma \ts T_c \wf \\
+\Gamma \ts T_c \expand_y \seq{\Ldecl L S U},\seq{\ldecl l V},\seq{\mdecl m T W} \\
+\Gamma \envplus{y: T_c} \ts \seq{S \sub U}\\
+\Gamma \envplus{y: T_c} \ts \seq{v \typ V'} \spcomma \seq{V' \sub V}\\
+\Gamma \envplus{y: T_c} \ts \seq{{T_i} \wfe}\\
+\seq{\Gamma \envplus{y: T_c} \envplus{x_i: T_i} \ts t_i \typ {W_i}' \spcomma {W_i}' \sub {W_i}}\\
+\Gamma \envplus{y: T_c} \ts  {t'} \typ T'}
+{\Gamma \ts \new y c {t'} \typ T'}
+
+\end{multicols}
+}
+
+\boxfig{fig:dot-mem-exp}{The DOT Calculus : Membership and Expansion}{
+    \begin{multicols}{2}[\judgement{Membership}{\fbox{$\Gamma \ts t \ni D$}}]
+
+      \infrule[\textsc{path-$\ni$}]
+      {\Gamma \ts p \typ T \spcomma T \expand_z \seq D}
+      {\Gamma \ts p \ni \subst p z {D_i}}
+
+      \infrule[\textsc{term-$\ni$}]
+      {z \not\in \fn(D_i) \andalso \Gamma \ts t \typ T \spcomma T \expand_z \seq D}
+      {\Gamma \ts t \ni D_i}
+   \end{multicols}
+
+   \linesep
+
+    \begin{multicols}{2}[\judgement{Expansion}{\fbox{$\Gamma \ts T \expand_z \seq D$}}]
+
+      \infrule[\textsc{rfn-$\expand$}]
+      {\Gamma \ts T \expand_z {\seq D'}}
+      {\Gamma \ts T \refine z {\seq D} \expand_z \seq {D'} \tand \seq D}
+
+      \infrule[\textsc{$\tand$-$\expand$}]
+      {\Gamma \ts T_1 \expand_z {\seq D_1} \spcomma T_2 \expand_z {\seq D_2}}
+      {\Gamma \ts T_1 \tand T_2 \expand_z {\seq D_1 \tand \seq D_2}}
+
+      \infax[\textsc{$\Top$-$\expand$}]
+      {\Gamma \ts \Top \expand_z \{\}}
+
+      \infrule[\textsc{tsel-$\expand$}]
+      {\Gamma \ts p \ni \Ldecl L S U \spcomma U \expand_z \seq D}
+      {\Gamma \ts p.L \expand_z \seq D}
+
+      \infrule[\textsc{$\tor$-$\expand$}]
+      {\Gamma \ts T_1 \expand_z {\seq D_1} \spcomma T_2 \expand_z {\seq D_2}}
+      {\Gamma \ts T_1 \tor T_2 \expand_z {\seq D_1 \tor \seq D_2}}
+
+      \infax[\textsc{$\Bot$-$\expand$}]
+      {\Gamma \ts \Bot \expand_z \seq{D_\Bot}}
+    \end{multicols}
+
+}
+
+\boxfig{dot-sub}{The DOT Calculus : Subtyping and Declaration Subsumption}{
+
+    \begin{multicols}{2}[\judgement{Subtyping}{\fbox{$\Gamma \ts S \sub T$}}]
+
+%      \infrule[\textsc{trans}]
+%      {\Gamma \ts S \sub T \spcomma T \sub U}
+%      {\Gamma \ts S \sub U}
+
+      \infrule[\textsc{refl}]
+      {\Gamma \ts T \wfe}
+      {\Gamma \ts T \sub T}
+
+      \infrule[\textsc{$\sub$-rfn}]
+      {\Gamma \ts {T \refine z {\seq D}} \wfe
+       \spcomma S \sub T \spcomma S \expand_z \seq{D'} \\
+       \Gamma \envplus{z: S} \ts \seq{D' \sub D}}
+      {\Gamma \ts S \sub T \refine z {\seq D}}
+
+      \infrule[\textsc{$\sub$-tsel}]
+      {\Gamma \ts p \ni \Ldecl L S U \spcomma S \sub U \spcomma S' \sub S }
+      {\Gamma \ts S' \sub p.L}
+
+      \infrule[\textsc{$\sub$-$\tand$}]
+      {\Gamma \ts T \sub T_1 \spcomma T \sub T_2}
+      {\Gamma \ts T \sub T_1 \tand T_2}
+
+      \infrule[\textsc{$\sub$-$\tor_1$}]
+      {\Gamma \ts T_2 \wfe \spcomma T \sub T_1}
+      {\Gamma \ts T \sub T_1 \tor T_2}
+
+      \infrule[\textsc{$\sub$-$\tor_2$}]
+      {\Gamma \ts T_1 \wfe \spcomma T \sub T_2}
+      {\Gamma \ts T \sub T_1 \tor T_2}
+
+      %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+
+      \infrule[\textsc{$\sub$-$\Top$}]
+      {\Gamma \ts T \wfe}
+      {\Gamma \ts T \sub \Top}
+
+      \infrule[\textsc{$\Bot$-$\sub$}]
+      {\Gamma \ts T \wfe}
+      {\Gamma \ts \Bot \sub T}
+
+      \infrule[\textsc{rfn-$\sub$}]
+      {\Gamma \ts {T \refine z {\seq D}} \wfe
+       \spcomma T \sub T'}
+      {\Gamma \ts T \refine z {\seq D} \sub T'}
+
+      \infrule[\textsc{tsel-$\sub$}]
+      {\Gamma \ts p \ni \Ldecl L S U \spcomma S \sub U \spcomma U \sub U'}
+      {\Gamma \ts p.L \sub U'}
+
+      \infrule[\textsc{$\tor$-$\sub$}]
+      {\Gamma \ts T_1 \sub T \spcomma T_2 \sub T}
+      {\Gamma \ts T_1 \tor T_2 \sub T}
+
+      \infrule[\textsc{$\tand_1$-$\sub$}]
+      {\Gamma \ts T_2 \wfe \spcomma T_1 \sub T}
+      {\Gamma \ts T_1 \tand T_2 \sub T}
+
+      \infrule[\textsc{$\tand_2$-$\sub$}]
+      {\Gamma \ts T_1 \wfe \spcomma T_2 \sub T}
+      {\Gamma \ts T_1 \tand T_2 \sub T}
+
+      \end{multicols}
+
+    \linesep
+
+    \begin{multicols}{2}[\judgement{Declaration subsumption}{\fbox{$\Gamma \ts D \sub D'$}}]
+
+    \infrule[\textsc{tdecl-$\sub$}]
+            {\Gamma \ts S' \sub S \spcomma T \sub T'}
+            {\Gamma \ts (\Ldecl L S T) \sub (\Ldecl L {S'} {T'})}
+
+    \infrule[\textsc{vdecl-$\sub$}]
+            {\Gamma \ts T \sub T'}
+            {\Gamma \ts (\ldecl l T) \sub (\ldecl l {T'})}
+
+    \infrule[\textsc{mdecl-$\sub$}]
+            {\Gamma \ts S' \sub S \spcomma T \sub T'}
+            {\Gamma \ts (\mdecl m S T) \sub (\mdecl m {S'} {T'})}
+
+    \end{multicols}
+}
+
+\boxfig{fig:dot-wf}{The DOT Calculus : Well-Formedness}{
+
+       \begin{multicols}{2}[\judgement{Well-formed types}{\fbox{$\Gamma \ts T \wf$}}]
+
+      \infrule[\textsc{rfn-wf}]
+      {\Gamma \ts T \wf \\ 
+       \Gamma \envplus {z: T \refine z {\seq D}} \ts \seq {D \wf}}
+      {\Gamma \ts T \refine z {\seq D} \wf}
+
+      \infrule[\textsc{tsel-wf$_1$}]
+      {\Gamma \ts p \ni \Ldecl L S U \spcomma S \wf \spcomma U \wf}
+      {\Gamma \ts p.L \wf}
+
+      \infrule[\textsc{$\tand$-wf}]
+      {\Gamma \ts T \wf \spcomma T' \wf}
+      {\Gamma \ts T \tand T' \wf}
+
+      %%%%%%%%%%%%%%%%%%%%%%%%%%%
+
+      \infax[\textsc{$\Top$-wf}]
+      {\Gamma \ts \Top \wf}
+
+      \infax[\textsc{$\Bot$-wf}]
+      {\Gamma \ts \Bot \wf}
+
+      \infrule[\textsc{tsel-wf$_2$}]
+      {\Gamma \ts p \ni \Ldecl L \Bot U}
+      {\Gamma \ts p.L \wf}
+
+      \infrule[\textsc{$\tor$-wf}]
+      {\Gamma \ts T \wf \spcomma T' \wf}
+      {\Gamma \ts T \tor T' \wf}
+
+    \end{multicols}
+
+    \linesep
+
+    \begin{multicols}{2}[\judgement{Well-formed declarations}{\fbox{$\Gamma \ts D \wf$}}]
+      \infrule[\textsc{tdecl-wf}]
+      {\Gamma \ts S \wf \spcomma U \wf}
+      {\Gamma \ts \Ldecl L S U \wf}
+
+      \infrule[\textsc{vdecl-wf}]
+      {\Gamma \ts T \wf}
+      {\Gamma \ts \ldecl l T \wf}
+
+      \infrule[\textsc{mdecl-wf}]
+      {\Gamma \ts S \wf \spcomma U \wf}
+      {\Gamma \ts \mdecl m S U \wf}
+
+    \end{multicols}
+
+    \linesep
+
+    \begin{multicols}{2}[\judgement{Well-formed and expanding types}{\fbox{$\Gamma \ts T \wfe$}}]
+      \infrule[\textsc{wfe}]
+      {\Gamma \ts T \wf \spcomma T \expand_z \seq{D}}
+      {\Gamma \ts T \wfe}
+    \end{multicols}
+}
+
+\boxfig{fig:dot-decls}{The DOT Calculus : Declaration Lattice}{
+
+\bda{lcl@{\gap}l}
+
+      \dom(\seq D \tand \seq {D'}) &~=~& \dom(\seq{D}) \cup \dom(\seq{D'}) \\
+      \dom(\seq D \tor \seq {D'}) &=& \dom(\seq{D}) \cap \dom(\seq{D'}) \\[0.5em]
+      (D \tand D')(L) &=&
+        \Ldecl L {(S \tor S')} {(U \tand U')} & \mbox{if~} (\Ldecl L S U) \in \seq{D} \;\mbox{and}\; (\Ldecl L {S'} {U'}) \in \seq{D'} \\
+        &=& D(L) & \mbox{if~} L \notin \dom(\seq{D'}) \\
+        &=& D'(L) & \mbox{if~} L \notin \dom(\seq{D}) \\
+      (D \tand D')(m) &=&
+        \mdecl m {(S \tor S')} {(U \tand U')} & \mbox{if~} (\mdecl m S U) \in \seq{D} \;\mbox{and}\; (\mdecl m {S'} {U'}) \in \seq{D'} \\
+        &=& D(m) & \mbox{if~} m \notin \dom(\seq{D'}) \\
+        &=& D'(m) & \mbox{if~} m \notin \dom(\seq{D}) \\
+      (D \tand D')(l) &=&
+        \ldecl l {T \tand T'} & \mbox{if~} (\ldecl l T) \in \seq{D} \;\mbox{and}\; (\ldecl l {T'}) \in \seq{D'} \\
+        &=& D(l) & \mbox{if~} l \notin \dom(\seq{D'}) \\
+        &=& D'(l) & \mbox{if~} l \notin \dom(\seq{D}) \\[0.5em]
+      (D \tor D')(L) &=&
+        \Ldecl L {(S \tand S')} {(U \tor U')} & \mbox{if~} (\Ldecl L S U) \in \seq{D} \;\mbox{and}\; (\Ldecl L {S'} {U'}) \in \seq{D'} \\
+      (D \tor D')(m) &=&
+        \mdecl m {(S \tand S')} {(U \tor U')} & \mbox{if~} (\mdecl m S U) \in \seq{D} \;\mbox{and}\; (\mdecl m {S'} {U'}) \in \seq{D'} \\
+      (D \tor D')(l) &=&
+        \ldecl l {T \tor T'} & \mbox{if~} (\ldecl l T) \in \seq{D} \;\mbox{and}\; (\ldecl l {T'}) \in \seq{D'} 
+    \eda
+
+Sets of declarations form a lattice with the given meet $\wedge$ and
+join $\vee$, the empty set of declatations as the top element, and the
+bottom element $\seq{D_\Bot}$, Here $\seq{D_\Bot}$ is the set of
+declarations that contains for every term label $l$ the declaration
+$\ldecl l \Bot$, for every type label $L$ the declaration $\Ldecl L
+\Top \Bot$ and for every method label $m$ the declaration $\mdecl m
+\Top \Bot$.
+}
+
 %\appendix

### 2012-08-05 article.tex
original intro

@@
 
-\def\c{\lstinline}  % shorter version so you can write \c|String[Foo]|
+\def\code{\lstinline}  % shorter version so you can write \code|String[Foo]|
                        % -- \def must be in same file as uses for this to
@@
 \section{Introduction}
-% TODO
+
+This paper presents a proposal for a new type-theoretic foundation of
+Scala and languages like it. The properties we are interested in
+modelling are Scala's path-dependent types and abstract type members,
+as well as its mixture of nominal and structural typing through the
+use of refinement types. Compared to previous approaches (nuObj, FS),
+we make no attempt to model inheritance or mixin composition. Indeed
+we will argue that such concepts are better modelled in a different
+setting.
+
+The calculus does not precisely describe what's currently in Scala. It
+is more normative than descriptive. The main point of deviation
+concerns the difference between Scala's compound type formation using
+{\bf with} and classical type intersection, as it is modelled in the
+calculus. Scala, and the previous calculi attempting to model it
+conflates the concepts of compound types (which inherit the members of
+several parent types) and mixin composition (which build classes from
+other classes and traits). At first glance, this offers an economy of
+concepts. However, it is problematic because mixin composition and
+intersection types have quite different properties. In the case of
+several inherited members with the same name, mixin composition has to
+pick one which overrides the others. It uses for that the concept of
+linearization of a trait hierarchy. Typically, given two independent
+traits $T_1$ and $T_2$ with a common method $m$, the mixin composition
+\code@$T_1$ with $T_2$@ would pick the $m$ in $T_2$, whereas the member in
+$T_1$ would be available via a super-call. All this makes sense from
+an implementation standpoint. From a typing standpoint it is more
+awkward, because it breaks commutativity and with it several
+monotonicity properties.
+
+In the present calculus, we replace Scala's compound types by
+classical intersection types, which are commutative. We also
+complement this by classical union types. Intersections and unions
+form a lattice wrt subtyping. This addresses another problematic
+feature of Scala: In Scala's current type system, least upper bounds
+and greatest lower bounds do not always exist. Here is an example:
+Given two traits
+\begin{lstlisting}
+  trait A { type T <: A }
+  trait B { type T <: B }
+\end{lstlisting}
+The greatest lower bound of \code@A@ and \code@B@ is approximated by the
+infinite sequence
+\begin{lstlisting}
+  A with B { type T <: A with B { type T <: A with B {
+    type T < ...
+  }}}
+\end{lstlisting}
+The limit of this sequence does not exist as a type in Scala.
+
+This is problematic because glbs and lubs play a central role in
+Scala's type inference. The absence of universal glbs and lubs makes
+type inference more brittle and more unpredictable.
+
+\subsection*{Why No Inheritance?}
+
+In the calculus we made the deliberate choice not to model any form of
+inheritance. This is, first and foremost, to keep the calculus simple.
+Secondly, there are many different approaches to inheritance and
+mixin composition, so that it looks advantageous not to tie the basic
+calculus to a specific one. Finally, it seems that the modelization of
+inheritance lends itself to a different approach than the basic
+calculus. For the latter, we need to prove the standard theorems of
+preservation and progress to establish soundness of the type system.
+One might try to do this also for a calculus with inheritance, but our
+experience suggests that this complicates the proofs considerably.  An
+alternative approach to inheritance that might work better is to model
+it as a form of code-reuse. Starting with an enriched type system with
+inheritance, and a translation to the basic calculus, one needs to
+show type preservation wrt the translation. This might be easier than
+to prove type preservation wrt reduction.
 
 \section{The DOT Calculus}
+
+%%% BEGIN RULES %%%
 
@@
 
+%%% END RULES %%%
+
 %\appendix

### 2012-08-05 article.tex
revise intro

@@
 inheritance lends itself to a different approach than the basic
-calculus. For the latter, we need to prove the standard theorems of
-preservation and progress to establish soundness of the type system.
+calculus. For the latter, we need to prove type safety of the calculus.
 One might try to do this also for a calculus with inheritance, but our
@@
 inheritance, and a translation to the basic calculus, one needs to
-show type preservation wrt the translation. This might be easier than
-to prove type preservation wrt reduction.
+show type safety wrt the translation. This might be easier than
+to prove type safety wrt reduction.
+
+% TODO: ask: does this last point still makes sense wrt to type safety via logical relations instead of via the standard theorems of preservation and progress?
 

### 2012-08-05 article.tex
original dot-one

@@
 
-%%% BEGIN RULES %%%
-
 \boxfig{fig:dot-one}{The DOT Calculus : Syntax, Reduction, Type Assignment}{
@@
 \end{multicols}
-}
+} % END dot-one
+
+The DOT calculus is a simple system of dependent
+object-types. Figure~\ref{dot-one} gives its syntax, reduction rules,
+and type assignment rules.
+
+\subsection*{Notation} We use standard notational conventions for
+sets. The notation $\seq{X}$ denotes a set of elements $X$. Given a
+such a set $\seq X$ in a typing rule, $X_i$ denotes an arbitrary
+element of $X$. 
+The $\uplus$ operator extends a set of bindings. It is required that the added binding does not
+introduce a variable which is already bound in the base-set. 
+We use an
+abbreviation for preconditions in typing judgements. Given an
+environment $\Gamma$ and some predicates $P$ and $Q$, the condition $\Gamma \ts P \spcomma Q$
+is a shorthand for the two conditions $\Gamma \ts P$ and $\Gamma \ts Q$.
+
+\subsection*{Syntax}
+
+There are three alphabets: Variable names $x$, $y$, $z$ are freely
+alpha-renamable. They occur as parameters of lambda abstractions, as
+binders for objects created by \verb@new@-expressions, and as self
+references in refinements. Value labels $l$ denote fields in objects,
+which are bound to values at run-time. Type labels $L$ denote type
+members of objects. Type labels are further separated into labels for
+abstract types $L_a$ and labels for classes $L_c$. It is assumed that
+in each program every class label $L_c$ is declared at most once.
+
+We assume that the label alphabets $l$ and $L$ are finite. This is
+not a restriction in practice, because one can include in these 
+alphabets every label occurring in a given program.
+
+The terms $t$ in DOT consist of variables $x$, $y$, $z$, lambda
+abstractions $\abs x T t$, function applications $\app t {t'}$, field
+selections $t.l$, and object creation expressions $\new x c t$ where
+$c$ is a constructor $T_c \ldefs{\seq{l = v}}$. The latter binds a
+variable $x$ to a new instance of type $T_c$ with fields $\seq l$ initialized to values $\seq v$.  The scope of $x$ extends through the term $t$.
+
+Two subclasses of terms are values $v$, which consist of just
+variables and lambda abstractions, and paths $v$ which consist of just
+variables and selections.
+
+The types in DOT are denoted by letters $S$, $T$, $U$, or $V$. They consist of the following:
+\begin{itemize}
+\item[-] Type selections $p.L$, which denote the type member $L$ of path $p$.
+%         A shorthand notation applies to type selections on singleton types: Instead
+%         of $p \sing \tsel L$ one can write just $p.L$.
+\item[-] Refinement types $T \refine z {\seq D}$, which refine a type $T$ by a set of declarations $D$.
+         The variable $z$ refers to the ``self''-reference of the type. Declarations can refer to
+         other declarations in the same type by selecting from $z$.
+\ifdep
+  \item[-] Dependent function types $\dfun x T {T'}$.
+\else
+  \item[-] Function types $T \tfun T'$.
+\fi
+\item[-] Type intersections $T \tand T'$, which carry the declarations of members present in either $T$ or $T'$.
+\item[-] Type unions $T \tor T'$, which carry only the declarations of members present in both $T$ and $T'$.
+\item[-] A top type $\Top$, which corresponds to an empty object.
+\item[-] A bottom type $\Bot$, which represents a non-terminating computation.
+\end{itemize}
+A subset of types $T_c$ are called {\em concrete types}. These are type selections
+$p.L_c$ of class labels,
+the top type $\Top$, intersections of concrete types, and refinements $T_c \refine z {\seq D}$ of concrete types. Only concrete types are allowed in constructors $c$.
+
+%Types in DOT form a lattice where $\tand$ is join and $\tor$ is meet.
+
+There are only two forms of declarations in DOT, which are both part
+of refinement types.  A value declaration $\ldecl l T$ introduces a
+field with type $T$.  A type declaration $\Ldecl L S U$ introduces a
+type member $L$ with a lower bound type $S$ and an upper bound type
+$U$. There are no type aliases, but a type alias can be simulated by a
+type declaration $\Ldecl L T T$ where the lower bound and the upper
+bound are the same type $T$.
+
+%It is assumed that the lower bound of
+%every class type is the bottom type and that the upper bound is a
+%concrete type. That is, the only permissible form of a class
+%declaration is $\Ldecl {L_c} \Bot T_c$.
+
+Every field or type label can be declared only once in a set of
+declarations $\seq D$. A set of declarations can hence be seen as a map from
+labels to their declarations.  Meets $\wedge$ and joins $\vee$ on sets of
+declarations are defined as follows.
+
+\bda{lcl@{\gap}l}
+
+      \dom(\seq D \tand \seq {D'}) &~=~& \dom(\seq{D}) \cup \dom(\seq{D'}) \\
+      \dom(\seq D \tor \seq {D'}) &=& \dom(\seq{D}) \cap \dom(\seq{D'}) \\[0.5em]
+      (D \tand D')(L) &=&
+        \Ldecl L {(S \tor S')} {(U \tand U')} & \mbox{if~} (\Ldecl L S U) \in \seq{D} \;\mbox{and}\; (\Ldecl L {S'} {U'}) \in \seq{D'} \\
+        &=& D(L) & \mbox{if~} L \notin \dom(\seq{D'}) \\
+        &=& D'(L) & \mbox{if~} L \notin \dom(\seq{D}) \\
+      (D \tand D')(l) &=&
+        \ldecl l {T \tand T'} & \mbox{if~} (\ldecl l T) \in \seq{D} \;\mbox{and}\; (\ldecl l {T'}) \in \seq{D'} \\
+        &=& D(l) & \mbox{if~} l \notin \dom(\seq{D'}) \\
+        &=& D'(l) & \mbox{if~} l \notin \dom(\seq{D}) \\[0.5em]
+      (D \tor D')(L) &=&
+        \Ldecl L {(S \tand S')} {(U \tor U')} & \mbox{if~} (\Ldecl L S U) \in \seq{D} \;\mbox{and}\; (\Ldecl L {S'} {U'}) \in \seq{D'} \\
+      (D \tor D')(l) &=&
+        \ldecl l {T \tor T'} & \mbox{if~} (\ldecl l T) \in \seq{D} \;\mbox{and}\; (\ldecl l {T'}) \in \seq{D'} 
+    \eda
+
+Sets of declarations form a lattice with the given meet $\wedge$ and
+join $\vee$, the empty set of declatations as the top element, and
+the bottom element $\seq{D_\Bot}$, Here $\seq{D_\Bot}$ is the set of
+declarations that contains for every term label $l$ the
+declaration $\ldecl l \Bot$ and for every type label $L$ the
+declaration $\Ldecl L \Top \Bot$.
+
+\subsection*{Reduction rules}
+
+Reduction rules $\reduction t s {t'} {s'}$ in DOT rewrite pairs of
+terms $t$ and stores $s$, where stores map variables to constructors.
+There are three main reduction rules: ($\beta_V$) is the standard beta
+reduction of call-by-value lambda calculus. Rule (\textsc{new}) rewrites an
+object creation $\new x c t$ by placing the binding of variable $x$ 
+to constructor $c$ in the store and continuing with term $t$. Rule (\textsc{sel})
+rewrites a field selection $x.l$ by retrieving the corresponding value from the store.
+These reduction rules can be applied anywhere in a term where the hole $[\,]$ of 
+an evaluation context $e$ can be situated.
+
+\subsection*{Type assignment rules}
+
+The last part of Figure~\ref{dot-one} presents rules for type
+assignment.  
+\ifdep
+The four rules (\textsc{var}), (\textsc{abs}), (\textsc{app-term}) and
+(\textsc{app-path}) correspond to dependently typed lambda calculus
+with a subtyping discipline. There are two rules for
+applications. Rule (\textsc{app-term}) implements the standard
+application rule if the function's result type does not depend on its
+argument type. Rule (\textsc{app-path}) applies to dependently typed
+functions which can only be applied to paths.
+Instead of adding a separate subsumption rule,
+subtyping is expressed by preconditions in the application rules and
+also in rule (\textsc{new}).  
+\else
+The three rules (\textsc{var}), (\textsc{abs}), and
+(\textsc{app}) on the left correspond to simply typed lambda calculus
+with a subtyping discipline (I believe it is straightforward to extend
+this to full $F_{\sub}$). Rule (\textsc{abs}) has an additional precondition
+stating that the bound variable $x$ may not appear in the function's
+result type $T'$. It would be an interesting extension of the calculus
+to drop this precondition, thereby allowing dependent function types.
+Instead of adding a separate subsumption rule,
+subtyping is expressed by preconditions in rules (\textsc{app}) and
+(\textsc{new}).  
+\fi
+Rule rule (\textsc{sub}) is the standard subsumption rule for subtyping.
+
+%The (\textsc{path}) rule gives the alternate type $p \sing$ to a well-typed path
+%$p$ whose type $T$ has a non-empty expansion $D,\seq{D}$.  By the subtyping rule
+%(\textsc{single-$\sub$}) in Figure~\ref{dot-sub}, the alternate type $p \sing$
+%is guaranteed to be a subtype of the underlying type $T$ of $p$.
+
+Rule (\textsc{sel}) types a field selection by means of an auxiliary
+membership relation $\ni$, which determines whether a given term contains a given
+declaration as one of its members. The membership relation is defined in
+Figure~\ref{dot-two} and is further explained below.
+
+The last rule, (\textsc{new}), assigns types to object creation
+expressions. It is the most complex of DOT's typing rules.  To
+type-check an object creation $\new x {T_c \ldefs {\seq{l = v}}} t$,
+one verifies first that the type $T_c$ is well-formed (see
+Figure~\ref{dot-wf} for a definition of well-formedness).  One then
+determines the set of all declarations that this type carries, using
+the expansion relation $\expand$ defined in Figure~\ref{dot-two}.
+Every type declaration $\Ldecl L S U$ in this set must be realizable,
+i.e.\ its lower bound $S$ must be a subtype of its upper bound $U$.
+Every field declaration $\ldecl l V$ in this set must have a
+corresponding initializing value of $v$ of type $V$.  These checks are
+made in an environment which is extended by the binding $x: T_c$. In
+particular this allows field values that recurse on ``self'' by
+referring to the bound variable $x$. As in the (\textsc{abs}) rule, it
+is required that the bound variable does not appear in the
+expression's result type $T'$.
 
@@
 
-%%% END RULES %%%
-
 %\appendix

### 2012-08-05 article.tex
revise dot-one

--- 2012-08-05 article.tex
@@
 \end{multicols}
+} % END dot-one
+
+The DOT calculus is a simple system of dependent
+object-types. \figref{fig:dot-one} gives its syntax, reduction rules,
+and type assignment rules.
+
+\subsection*{Notation} We use standard notational conventions for
+sets. The notation $\seq{X}$ denotes a set of elements $X$. Given a
+such a set $\seq X$ in a typing rule, $X_i$ denotes an arbitrary
+element of $X$. 
+%The $\uplus$ operator extends a set of bindings. It is required that the added binding does not
+%introduce a variable which is already bound in the base-set.
+We use an
+abbreviation for preconditions in typing judgements. Given an
+environment $\Gamma$ and some predicates $P$ and $Q$, the condition $\Gamma \ts P \spcomma Q$
+is a shorthand for the two conditions $\Gamma \ts P$ and $\Gamma \ts Q$.
+
+\subsection*{Syntax}
+
+There are four alphabets: Variable names $x$, $y$, $z$ are freely
+alpha-renamable. They occur as parameters of lambda abstractions, as
+binders for objects created by \verb@new@-expressions, and as self
+references in refinements. Value labels $l$ denote fields in objects,
+which are bound to values at run-time. Similarly, method labels $m$
+denote methods in objects. Type labels $L$ denote type members of
+objects. Type labels are further separated into labels for abstract
+types $L_a$ and labels for classes $L_c$. It is assumed that in each
+program every class label $L_c$ is declared at most once.
+
+We assume that the label alphabets $l$, $m$ and $L$ are finite. This is
+not a restriction in practice, because one can include in these 
+alphabets every label occurring in a given program.
+
+The terms $t$ in DOT consist of variables $x$, $y$, $z$, field
+selections $t.l$, method invocations $t.m(t)$ and object creation
+expressions $\new y c {t'}$ where $c$ is a constructor $T_c \ldefs{\seq{l
+    = v}\;\seq{m(x) = t}}$. The latter binds a variable $y$ to a new
+instance of type $T_c$ with fields $\seq l$ initialized to values
+$\seq v$ and methods $\seq m$ initialized to methods of one parameter
+$\seq{x}$ and body $\seq{t}$.  The scope of $y$ extends through the term
+${t'}$.
+
+Two subclasses of terms are values $v$, which consist of just
+variables, and paths $v$ which consist of just variables and field
+selections.
+
+The types in DOT are denoted by letters $S$, $T$, $U$, or $V$. They consist of the following:
+\begin{itemize}
+\item[-] Type selections $p.L$, which denote the type member $L$ of path $p$.
+\item[-] Refinement types $T \refine z {\seq D}$, which refine a type $T$ by a set of declarations $D$.
+         The variable $z$ refers to the ``self''-reference of the type. Declarations can refer to
+         other declarations in the same type by selecting from $z$.
+\item[-] Type intersections $T \tand T'$, which carry the declarations of members present in either $T$ or $T'$.
+\item[-] Type unions $T \tor T'$, which carry only the declarations of members present in both $T$ and $T'$.
+\item[-] A top type $\Top$, which corresponds to an empty object.
+\item[-] A bottom type $\Bot$, which represents a non-terminating computation.
+\end{itemize}
+A subset of types $T_c$ are called {\em concrete types}. These are type selections
+$p.L_c$ of class labels,
+the top type $\Top$, intersections of concrete types, and refinements $T_c \refine z {\seq D}$ of concrete types. Only concrete types are allowed in constructors $c$.
+
+There are only three forms of declarations in DOT, which are both part
+of refinement types.  A value declaration $\ldecl l T$ introduces a
+field with type $T$.  A method declaration $\mdecl m S U$ introduces a
+method with parameter of type $S$ and result of type $U$. A type
+declaration $\Ldecl L S U$ introduces a type member $L$ with a lower
+bound type $S$ and an upper bound type $U$. There are no type aliases,
+but a type alias can be simulated by a type declaration $\Ldecl L T T$
+where the lower bound and the upper bound are the same type $T$.
+
+\boxfig{fig:dot-decls}{The DOT Calculus : Declaration Lattice}{
+
+\bda{lcl@{\gap}l}
+
+      \dom(\seq D \tand \seq {D'}) &~=~& \dom(\seq{D}) \cup \dom(\seq{D'}) \\
+      \dom(\seq D \tor \seq {D'}) &=& \dom(\seq{D}) \cap \dom(\seq{D'}) \\[0.5em]
+      (D \tand D')(L) &=&
+        \Ldecl L {(S \tor S')} {(U \tand U')} & \mbox{if~} (\Ldecl L S U) \in \seq{D} \;\mbox{and}\; (\Ldecl L {S'} {U'}) \in \seq{D'} \\
+        &=& D(L) & \mbox{if~} L \notin \dom(\seq{D'}) \\
+        &=& D'(L) & \mbox{if~} L \notin \dom(\seq{D}) \\
+      (D \tand D')(m) &=&
+        \mdecl m {(S \tor S')} {(U \tand U')} & \mbox{if~} (\mdecl m S U) \in \seq{D} \;\mbox{and}\; (\mdecl m {S'} {U'}) \in \seq{D'} \\
+        &=& D(m) & \mbox{if~} m \notin \dom(\seq{D'}) \\
+        &=& D'(m) & \mbox{if~} m \notin \dom(\seq{D}) \\
+      (D \tand D')(l) &=&
+        \ldecl l {T \tand T'} & \mbox{if~} (\ldecl l T) \in \seq{D} \;\mbox{and}\; (\ldecl l {T'}) \in \seq{D'} \\
+        &=& D(l) & \mbox{if~} l \notin \dom(\seq{D'}) \\
+        &=& D'(l) & \mbox{if~} l \notin \dom(\seq{D}) \\[0.5em]
+      (D \tor D')(L) &=&
+        \Ldecl L {(S \tand S')} {(U \tor U')} & \mbox{if~} (\Ldecl L S U) \in \seq{D} \;\mbox{and}\; (\Ldecl L {S'} {U'}) \in \seq{D'} \\
+      (D \tor D')(m) &=&
+        \mdecl m {(S \tand S')} {(U \tor U')} & \mbox{if~} (\mdecl m S U) \in \seq{D} \;\mbox{and}\; (\mdecl m {S'} {U'}) \in \seq{D'} \\
+      (D \tor D')(l) &=&
+        \ldecl l {T \tor T'} & \mbox{if~} (\ldecl l T) \in \seq{D} \;\mbox{and}\; (\ldecl l {T'}) \in \seq{D'} 
+    \eda
+
+Sets of declarations form a lattice with the given meet $\wedge$ and
+join $\vee$, the empty set of declatations as the top element, and the
+bottom element $\seq{D_\Bot}$, Here $\seq{D_\Bot}$ is the set of
+declarations that contains for every term label $l$ the declaration
+$\ldecl l \Bot$, for every type label $L$ the declaration $\Ldecl L
+\Top \Bot$ and for every method label $m$ the declaration $\mdecl m
+\Top \Bot$.
 }
 
-\boxfig{fig:dot-mem-exp}{The DOT Calculus : Membership and Expansion}{
+Every field or type label can be declared only once in a set of
+declarations $\seq D$. A set of declarations can hence be seen as a map from
+labels to their declarations.  Meets $\wedge$ and joins $\vee$ on sets of
+declarations are defined in ~\figref{fig:dot-decls}.
+
+\subsection*{Reduction rules}
+
+Reduction rules $\reduction t s {t'} {s'}$ in DOT rewrite pairs of
+terms $t$ and stores $s$, where stores map variables to constructors.
+There are three main reduction rules: Rule (\textsc{msel}) rewrites a
+method invocation $y {m_i} v$ by retrieving the corresponding method
+definition from the store, and performing a substitution of the
+argument for the parameter in the body. Rule (\textsc{sel}) rewrites a
+field selection $x.l$ by retrieving the corresponding value from the
+store. Rule (\textsc{new}) rewrites an object creation $\new x c t$ by
+placing the binding of variable $x$ to constructor $c$ in the store
+and continuing with term $t$.  These reduction rules can be applied
+anywhere in a term where the hole $[\,]$ of an evaluation context $e$
+can be situated.
+
+\subsection*{Type assignment rules}
+
+The last part of \figref{fig:dot-one} presents rules for type
+assignment.
+
+Rules (\textsc{sel}) and (\textsc{msel}) type field selections and
+method invocations by means of an auxiliary membership relation $\ni$,
+which determines whether a given term contains a given declaration as
+one of its members. The membership relation is defined in
+\figref{fig:dot-two} and is further explained below.
+
+The last rule, (\textsc{new}), assigns types to object creation
+expressions. It is the most complex of DOT's typing rules.  To
+type-check an object creation $\new y {T_c \ldefs {\seq{l =
+      v}\;\seq{m(x) = t}}} t'$, one verifies first that the type $T_c$
+is well-formed (see \figref{fig:dot-wf} for a definition of
+well-formedness).  One then determines the set of all declarations
+that this type carries, using the expansion relation $\expand$ defined
+in \figref{fig:dot-two}.  Every type declaration $\Ldecl L S U$ in
+this set must be realizable, i.e.\ its lower bound $S$ must be a
+subtype of its upper bound $U$.  Every field declaration $\ldecl l V$
+in this set must have a corresponding initializing value of $v$ of
+type $V$.  These checks are made in an environment which is extended
+by the binding $y: T_c$. In particular this allows field values that
+recurse on ``self'' by referring to the bound variable $x$. Similarly,
+every method declaration $\mdecl m T W$ must have a corresponding
+initializing method definition $m(x) = t$. The parameter type $T$ must
+be $\wfe$ (well-formed and expanding; see \figref{fig:dot-wf}), and
+the body $t$ must type check to $W$ in an environment extended by the
+bindings $y : T_c$ and $x : T$.
+
+Instead of adding a separate subsumption rule, subtyping is expressed
+by preconditions in rules (\textsc{msel}) and (\textsc{new}).
+
+\boxfig{fig:dot-two}{The DOT Calculus : Membership and Expansion}{
     \begin{multicols}{2}[\judgement{Membership}{\fbox{$\Gamma \ts t \ni D$}}]
@@
 
-\boxfig{fig:dot-decls}{The DOT Calculus : Declaration Lattice}{
-
-\bda{lcl@{\gap}l}
-
-      \dom(\seq D \tand \seq {D'}) &~=~& \dom(\seq{D}) \cup \dom(\seq{D'}) \\
-      \dom(\seq D \tor \seq {D'}) &=& \dom(\seq{D}) \cap \dom(\seq{D'}) \\[0.5em]
-      (D \tand D')(L) &=&
-        \Ldecl L {(S \tor S')} {(U \tand U')} & \mbox{if~} (\Ldecl L S U) \in \seq{D} \;\mbox{and}\; (\Ldecl L {S'} {U'}) \in \seq{D'} \\
-        &=& D(L) & \mbox{if~} L \notin \dom(\seq{D'}) \\
-        &=& D'(L) & \mbox{if~} L \notin \dom(\seq{D}) \\
-      (D \tand D')(m) &=&
-        \mdecl m {(S \tor S')} {(U \tand U')} & \mbox{if~} (\mdecl m S U) \in \seq{D} \;\mbox{and}\; (\mdecl m {S'} {U'}) \in \seq{D'} \\
-        &=& D(m) & \mbox{if~} m \notin \dom(\seq{D'}) \\
-        &=& D'(m) & \mbox{if~} m \notin \dom(\seq{D}) \\
-      (D \tand D')(l) &=&
-        \ldecl l {T \tand T'} & \mbox{if~} (\ldecl l T) \in \seq{D} \;\mbox{and}\; (\ldecl l {T'}) \in \seq{D'} \\
-        &=& D(l) & \mbox{if~} l \notin \dom(\seq{D'}) \\
-        &=& D'(l) & \mbox{if~} l \notin \dom(\seq{D}) \\[0.5em]
-      (D \tor D')(L) &=&
-        \Ldecl L {(S \tand S')} {(U \tor U')} & \mbox{if~} (\Ldecl L S U) \in \seq{D} \;\mbox{and}\; (\Ldecl L {S'} {U'}) \in \seq{D'} \\
-      (D \tor D')(m) &=&
-        \mdecl m {(S \tand S')} {(U \tor U')} & \mbox{if~} (\mdecl m S U) \in \seq{D} \;\mbox{and}\; (\mdecl m {S'} {U'}) \in \seq{D'} \\
-      (D \tor D')(l) &=&
-        \ldecl l {T \tor T'} & \mbox{if~} (\ldecl l T) \in \seq{D} \;\mbox{and}\; (\ldecl l {T'}) \in \seq{D'} 
-    \eda
-
-Sets of declarations form a lattice with the given meet $\wedge$ and
-join $\vee$, the empty set of declatations as the top element, and the
-bottom element $\seq{D_\Bot}$, Here $\seq{D_\Bot}$ is the set of
-declarations that contains for every term label $l$ the declaration
-$\ldecl l \Bot$, for every type label $L$ the declaration $\Ldecl L
-\Top \Bot$ and for every method label $m$ the declaration $\mdecl m
-\Top \Bot$.
-}
-
-%%% END RULES %%%
-
 %\appendix

### 2012-08-05 article.tex
original dot-two

@@
 \newcommand{\boxfig}[3]
-        {\begin{figure*}\figurebox{#3\caption{\label{#1}#2}}\end{figure*}}
+        {\begin{figure*}\figurebox{#3\caption{\label{fig:#1}#2}}\end{figure*}}
 \newcommand{\figref}[1]
-        {Figure~\ref{#1}}
+        {Figure~\ref{fig:#1}}
 
@@
 
-\boxfig{fig:dot-one}{The DOT Calculus : Syntax, Reduction, Type Assignment}{
+\boxfig{dot-one}{The DOT Calculus : Syntax, Reduction, Type Assignment}{
 {\bf Syntax}\medskip
@@
 The DOT calculus is a simple system of dependent
-object-types. \figref{fig:dot-one} gives its syntax, reduction rules,
+object-types. \figref{dot-one} gives its syntax, reduction rules,
 and type assignment rules.
@@
 
-\boxfig{fig:dot-decls}{The DOT Calculus : Declaration Lattice}{
+\boxfig{dot-decls}{The DOT Calculus : Declaration Lattice}{
 
@@
 labels to their declarations.  Meets $\wedge$ and joins $\vee$ on sets of
-declarations are defined in ~\figref{fig:dot-decls}.
+declarations are defined in ~\figref{dot-decls}.
 
@@
 
-The last part of \figref{fig:dot-one} presents rules for type
+The last part of \figref{dot-one} presents rules for type
 assignment.
@@
 one of its members. The membership relation is defined in
-\figref{fig:dot-two} and is further explained below.
+\figref{dot-two} and is further explained below.
 
@@
       v}\;\seq{m(x) = t}}} t'$, one verifies first that the type $T_c$
-is well-formed (see \figref{fig:dot-wf} for a definition of
+is well-formed (see \figref{dot-wf} for a definition of
 well-formedness).  One then determines the set of all declarations
 that this type carries, using the expansion relation $\expand$ defined
-in \figref{fig:dot-two}.  Every type declaration $\Ldecl L S U$ in
+in \figref{dot-two}.  Every type declaration $\Ldecl L S U$ in
 this set must be realizable, i.e.\ its lower bound $S$ must be a
@@
 initializing method definition $m(x) = t$. The parameter type $T$ must
-be $\wfe$ (well-formed and expanding; see \figref{fig:dot-wf}), and
+be $\wfe$ (well-formed and expanding; see \figref{dot-wf}), and
 the body $t$ must type check to $W$ in an environment extended by the
@@
 
-\boxfig{fig:dot-two}{The DOT Calculus : Membership and Expansion}{
+\boxfig{dot-two}{The DOT Calculus : Membership and Expansion}{
     \begin{multicols}{2}[\judgement{Membership}{\fbox{$\Gamma \ts t \ni D$}}]
@@
 
-}
+} % END dot-two
+
+\subsection*{Membership}
+
+\figref{dot-two} presents typing rules for membership and
+expansion.  The membership judgement $\Gamma \ts t \ni D$ states that
+in environment $\Gamma$ a term $t$ has a declaration $D$ as a member.
+There are different rules for paths and general terms.  Rule
+(\textsc{path-$\ni$}) applies to paths $p$ that have a refinement type
+$T \refine z {\seq D}$ as their type. Members of $p$ are then all
+definitions $D$ in the refinement, where any use of the self reference
+$z$ is replaced by the path $p$ itself.  Rule (\textsc{and-$\ni$})
+allows to merge two member definitions (of the same label) by
+conjoining them with $\tand$. Rule (\textsc{term-$\ni$}) establishes
+the members of general terms $t$ of type $T$ by introducing a dummy
+variable $z$ of type $T$ and then using the rules for path membership
+on $z$. It must hold that $z$ itself does not form part of the
+resulting member $D$.
+
+\subsection*{Expansion}
+
+The expansion relation $\expand$ is needed to typecheck the complete
+set of declarations carried by a concrete type that is used in a
+\textbf{new}-expression.  Since this is the only place where expansion
+is needed, it is sufficient to define it on concrete types only. The
+bottom part of Figure~\ref{dot-two} gives the typing rules for
+expansion.
+
+Rule (\textsc{rfn-$\expand$}) states that a refinement type $T_c
+\expand_z {\seq D}$ expands to the conjunction of the expansion $D'$
+of $T_c$ and the newly added declarations $D$. Rule
+(\textsc{tsel-$\expand$}) states that a type selection $p.L$ carries
+the same declarations as the upper bound $U$ of $L$ in $T$.  Rule
+($\tand$-$\expand$) states that expansion distributes through meets.
+Rule (\textsc{$\top$-$\expand$}) states that the top type $\top$
+expands to the empty set.
 
@@
 
-\boxfig{fig:dot-wf}{The DOT Calculus : Well-Formedness}{
+\boxfig{dot-wf}{The DOT Calculus : Well-Formedness}{
 

### 2012-08-05 article.tex
revised dot-two

@@
 set of declarations carried by a concrete type that is used in a
-\textbf{new}-expression.  Since this is the only place where expansion
-is needed, it is sufficient to define it on concrete types only. The
-bottom part of Figure~\ref{dot-two} gives the typing rules for
-expansion.
+\textbf{new}-expression. Expansion is also used in subtyping
+refinements on the right (see \figref{dot-sub}).
 
-Rule (\textsc{rfn-$\expand$}) states that a refinement type $T_c
+Rule (\textsc{rfn-$\expand$}) states that a refinement type $T
 \expand_z {\seq D}$ expands to the conjunction of the expansion $D'$
-of $T_c$ and the newly added declarations $D$. Rule
+of $T$ and the newly added declarations $D$. Rule
 (\textsc{tsel-$\expand$}) states that a type selection $p.L$ carries
-the same declarations as the upper bound $U$ of $L$ in $T$.  Rule
-($\tand$-$\expand$) states that expansion distributes through meets.
-Rule (\textsc{$\top$-$\expand$}) states that the top type $\top$
-expands to the empty set.
+the same declarations as the upper bound $U$ of $L$ in $T$.  Rules
+($\tand$-$\expand$) and ($\tor$-$\expand$) states that expansion
+distributes through meets and joins.  Rule (\textsc{$\top$-$\expand$})
+states that the top type $\top$ expands to the empty set. Rule
+(\textsc{$\bot$-$\expand$}) states that the bottom type $\bot$ expands
+to the bottom element $\seq{D_\Bot}$ of the lattice of sets of
+declarations (recall \figref{dot-decls}).
 

### 2012-08-05 article.tex
original dot-sub

@@
     \end{multicols}
-}
+} % END dot-sub
+
+\subsection*{Subtyping}
+
+Figure~\ref{dot-sub} defines the subtyping judgement $\Gamma \ts S
+\sub T$ which states that in environment $\Gamma$ type $S$ is a
+subtype of type $T$.  Any term of type $S$ can then be regarded via
+the subsumption rule (\textsc{sub}) as a term of type $T$.
+
+As usual, subtyping is reflexive (\textsc{refl}) but a corresponding
+rule for transitivity is missing.  However, most of the rules for
+specific types in Figure~\ref{dot-sub} have transitivity built in. The
+issue of transitivity is further discussed below.
+
+To prove that a type $S$ is a subtype of a refinement type $T \refine
+z {\seq D}$, it must hold that $S$ is a subtype of $T$ and that $S$
+contains declarations that subsume the declarations $\seq D$
+(\textsc{$\sub$-rfn}). The refinement type itself is a subtype of its
+parent type $T$ and all its supertypes (\textsc{rfn-$\sub$}).
+
+A type selection $p.L$ is a subtype of (all supertypes of) the upper bound $U$ of
+$L$ in $p$ (\textsc{tsel-$\sub$}).
+It is a supertype of (all subtypes of) its lower
+bound $S$ (\textsc{$\sub$-tsel}).  Function types are subject to the
+standard co-/contravariant subtyping rule 
+\ifdep
+ (\textsc{$\sub$-$\Rightarrow$}).
+\else
+  (\textsc{$\sub$-$\tfun$}).
+\fi
+%A singleton type $p\sing$ is a subtype of any underlying type of path
+%$p$ (\textsc{single-$\sub$}). 
+The final six rules in
+Figure~\ref{dot-sub} turn the subtyping relation into a lattice with 
+meets $\tand$, joins $\tor$, bottom element $\Bot$ and top element $\Top$.
+
+One interesting aspect of the subtyping relation of DOT is that it is
+reflexive (\textsc{refl}) but not transitive. Instead of a separate
+rule that specifies global transitivity, we find transitivity encoded
+in most of the inference rules of Figure~\ref{dot-sub}. There is only
+a single situation where transitivity does not hold:  Given a type
+declaration $\Ldecl L S U$ which is a member of some type $T$, and a
+variable $x$ of type $T$, we have that $S \sub x.L$ by
+(\textsc{$\sub$-tsel}) and $x.L \sub U$ by (\textsc{tsel-$\sub$}).
+However, this does not imply that $S \sub U$. In fact, 
+$T$ might be an {\em unrealizable} type, which admits no solution for
+its $L$ member, precisely because the lower bound of $L$ is not a
+subtype of its upper bound.
+
+The type assignment rule (\textsc{new}) in Figure~\ref{dot-one}
+has as one of its preconditions that the type of
+the created object must be realizable: each lower bounds must be a subtype
+of its corresponding upper bound. Subtyping can be shown to be transitive in DOT if the
+least type of a subtype chain is realizable.
 

### 2012-08-05 article.tex
revised dot-sub

--- 2012-08-05 article.tex
@@
     \end{multicols}
-}
+} % END dot-sub
+
+\subsection*{Subtyping}
+
+\figref{dot-sub} defines the subtyping judgement $\Gamma \ts S
+\sub T$ which states that in environment $\Gamma$ type $S$ is a
+subtype of type $T$. Subtyping is regular with respect to $\wfe$: if
+type $S$ is a subtype of type $T$, then $S$ and $T$ are well-formed
+and expanding. Though this regularity limits our calculus to
+$\wfe$-types, this limitation allows us to show that subtyping is
+transitive, as discussed in section~\ref{dot-preservation}.
 
@@
 
+\section{Counterexamples to Preservation}\label{dot-preservation}
+
+\section{Type Safety via Logical Relations}\label{dot-type-safety}
+
+\section{Conclusion}\label{conclusion}
+
 %\appendix

### 2012-08-05 article.tex
decl-sub original

@@
 
+\subsection*{Declaration Subsumption}
+
+The declaration subsumption judgement $\Gamma \ts D \sub D'$ in
+\figref{dot-sub} states that in environment $\Gamma$ the
+declaration $D$ subsumes the declaration $D'$. There are two rules,
+one for type declarations and one for value declarations.  Rule
+(\textsc{tdecl-$\sub$}) states that a type declaration $\Ldecl L S U$
+subsumes another type declaration $\Ldecl L {S'} {U'}$ if $S'$ is a
+subtype of $S$ and $U$ is a subtype of $U'$. In other words, the set
+of types between $S$ and $U$ is contained in the set of types between
+$S'$ and $U'$.  Rule (\textsc{vdecl-$\sub$}) states that a value
+declaration $\ldecl l T$ subsumes another value declaration $\ldecl l
+T'$ if $T$ is a subtype of $T'$.
+
+Declaration subsumption is extended a binary relation between
+sequences of declarations as follows: $\seq{D} <: \seq{D'}$ iff
+$\forall D'_i \exists D_j. D_j \sub D'_i$.
+
 \boxfig{dot-wf}{The DOT Calculus : Well-Formedness}{

### 2012-08-05 article.tex
revised decl sub

@@
 The declaration subsumption judgement $\Gamma \ts D \sub D'$ in
-\figref{dot-sub} states that in environment $\Gamma$ the
-declaration $D$ subsumes the declaration $D'$. There are two rules,
-one for type declarations and one for value declarations.  Rule
+\figref{dot-sub} states that in environment $\Gamma$ the declaration
+$D$ subsumes the declaration $D'$. There are three rules, one for each
+kind (type, value, method) of declarations.  Rule
 (\textsc{tdecl-$\sub$}) states that a type declaration $\Ldecl L S U$
@@
 declaration $\ldecl l T$ subsumes another value declaration $\ldecl l
-T'$ if $T$ is a subtype of $T'$.
+T'$ if $T$ is a subtype of $T'$. Rule (\textsc{mdecl-$\sub$}) is
+similar to (\textsc{tdecl-$\sub$}), as the parameter type varies
+contravariantly and the return type covariantly.
 
-Declaration subsumption is extended a binary relation between
+Declaration subsumption is extended to a binary relation between
 sequences of declarations as follows: $\seq{D} <: \seq{D'}$ iff

### 2012-08-05 article.tex
wf

@@
       {\Gamma \ts p \ni \Ldecl L S U \spcomma S \wf \spcomma U \wf}
-      {\Gamma \ts p.L \wf}
+      {\Gamma \ts p.L \wf \spcomma p.U \wf}
 
@@
     \end{multicols}
-}
+} % END dot-wf
+
+\subsection*{Well-formedness}
+
+The well-formedness judgement $\Gamma \ts T \wf$ in \figref{dot-wf}
+states that in environment $\Gamma$ the type $T$ is
+well-formed.
+
+A refinement type $T \refine z {\seq D}$ is
+well-formed if the parent type $T$ is well-formed and every
+declaration in $\seq D$ is well-formed in an environment augmented by
+the binding of the self-reference $z$ to the refinement type itself
+(\textsc{rfn-wf}).
+
+A type selection $p.L$ is well-formed if $L$ is a member of $p$, and
+the lower bound of $L$ is also well-formed (\textsc{tsel-wf$_1$} and
+\textsc{tsel-wf$_2$}). The latter condition has the effect that the
+lower bound of a type $p.L$ may not refer directly or indirectly to a
+type containing $p.L$ itself --- if it would, the well-formedness
+judgement of $p.L$ would not have a finite proof. No such restriction
+exists for the upper bound of $L$ if the lower bound is $\bot$
+(\textsc{tsel-wf$_2$}). The upper bound may in fact refer back to the
+type. Hence, recursive class types and F-bounded abstract types are
+both expressible.
+
+The other forms of types in DOT are all well-formed if their
+constituent types are well-formed.
+
+Well-formedness extends straightforwardly to declarations with the
+judgement $\Gamma \ts D \wf$. All declarations are well-formed if
+their constituent types are well-formed.
 

### 2012-08-05 article.tex
example from semester report

@@
 \newcommand{\literal}[1]{{#1}}
+\newcommand{\mi}[1]{\mathit{#1}}
 
@@
 \newcommand{\refine}[2]{\left\{#1 \Rightarrow #2 \right\}}
+\newcommand{\mlrefine}[2]{\{#1 \Rightarrow #2 \}}
 % Field definitions
 \newcommand{\ldefs}[1]{\left\{#1\right\}}
+\newcommand{\mlldefs}[1]{\{#1\}}
 % Member sequences
@@
 \newcommand{\new}[3]{\textbf{val }#1 = \textbf{new }#2 ;\; #3}
+\newcommand{\mlnew}[3]{\textbf{val }#1 = \textbf{new }#2 ;\;\\&#3}
 %\renewcommand{\new}[3]{#1 \leftarrow #2 \,\textbf{in}\, #3}
@@
 
+\section{Examples}\label{dot-examples}
+
+\subsection{Sugar: Functions}
+
+\subsection{Basics: Booleans, Error, \ldots}
+This program defines a $\mi{root}$ object with basic types
+($\mi{Unit}$, $\mi{Boolean}$, $\mi{Nat}$) and values ($\mi{true}$,
+$\mi{false}$, $\mi{error}$, $\mi{zero}$). For brevity, I've omitted
+the code related to natural numbers. During the object creation, the
+value labels such as $\mi{false}$ are all initialized.
+\begin{align*}
+&\mlnew {\mi{root}} {\Top \mlrefine r {\\\displaybreak[3]
+&\ \Ldecl {\mi{Unit}} \Bot \Top\\
+&\ \ldecl {\mi{unit}} {\Top \tfun {r.\mi{Unit}}}\\
+&\ \Ldecl {\mi{Boolean}} \Bot {\Top \mlrefine z {\\
+&\ \gap \ldecl {\mi{ifNat}} {(r.\mi{Unit} \tfun r.\mi{Nat}) \tfun (r.\mi{Unit} \tfun r.\mi{Nat}) \tfun r.\mi{Nat}}\\
+&\gap}}\\
+&\ \ldecl {\mi{false}} {{r.\mi{Unit}} \tfun {r.\mi{Boolean}}}\\
+&\ \ldecl {\mi{true}} {{r.\mi{Unit}} \tfun {r.\mi{Boolean}}}\\
+&\ \ldots\\\displaybreak[3]
+&\ \ldecl {\mi{error}} {{r.\mi{Unit}} \tfun \Bot}\\
+&}\mlldefs{\\\displaybreak[3]
+&\ \mi{unit}  = \abs x \Top {\new u {\mi{root}.\mi{Unit}} u}\\
+%
+&\ \mi{false} = \abs u {\mi{root}.\mi{Unit}} {\\
+&\ \gap \mlnew {\mi{ff}} {\mi{root}.\mi{Boolean} \mlldefs{\\\displaybreak[3]
+&\ \gap\gap {\mi{ifNat}} = \abs t {\mi{root}.\mi{Unit} \tfun \mi{root}.\mi{Nat}} {\\&\gap\gap\gap\abs e {\mi{root}.\mi{Unit} \tfun \mi{root}.\mi{Nat}}\\&\gap\gap\gap\app e {\mi{root}.\mi{unit}}}\\
+&\ \gap}}{
+\ \gap \mi{ff}}
+&}\\
+%
+&\ \mi{true} = \ldots\\
+%
+&\ \mi{error} = \abs u {\mi{root}.\mi{Unit}} {\app {\mi{root}.\mi{error}} u}\\
+&\ \ldots\\
+&}}{
+\ldots
+}
+\end{align*}
+
+\subsection{Polymorphic Lists}
+Polymorphic lists can be expressed using an abstract type member for
+the element type ($\mi{Elem}$). We can instantiate a refinement of the
+list package to manipulate lists with a particular element type. We
+can choose to make the lists invariant or covariant in the element
+type by fixing the lower bound of the $\mi{Elem}$ re-declaration to
+the upper bound or to $\Bot$, respectively.
+\begin{align*}
+&\mlnew {\mi{genLists}} {\Top \mlrefine g {\Ldecl {\mi{ListPackage}} \Bot {\Top \mlrefine p {\\\displaybreak[3]
+&\ \Ldecl {\mi{Elem}} \Bot \Top\\
+&\ \Ldecl {\mi{ListOfElement}} {g.\mi{List} \refine z {\Ldecl {\mi{Elem}} {p.\mi{Elem}} {p.\mi{Elem}}}} {\\&\gap\gap\gap\gap g.\mi{List} \refine z {\Ldecl {\mi{Elem}} {p.\mi{Elem}} {p.\mi{Elem}}}}\\
+&\ \Ldecl {\mi{List}} \Bot {\Top \mlrefine z {\\\displaybreak[3]
+&\ \gap \ldecl {\mi{isEmpty}} {\mi{root}.\mi{Unit} \tfun \mi{root}.\mi{Boolean}}\\
+&\ \gap \ldecl {\mi{head}} {\mi{root}.\mi{Unit} \tfun g.\mi{Elem}}\\
+&\ \gap \ldecl {\mi{tail}} {\mi{root}.\mi{Unit} \tfun g.\mi{ListOfElem}}\\
+&\gap}}\\
+&\ \ldecl {\mi{nil}} {\mi{root}.\mi{Unit} \tfun g.\mi{ListOfElem}}\\
+&\ \ldecl {\mi{cons}} {g.\mi{Elem} \tfun g.\mi{ListOfElem}}\\
+&}}}}{
+\mlnew {\mi{natLists}} {{\mi{genLists}.\mi{ListPackage}} \mlrefine p { \Ldecl {\mi{Elem}} {\mi{Nat}} {\mi{Nat}} }\mlldefs{\\\displaybreak[3]
+&\ {\mi{nil}} = \ldots\\
+&\ {\mi{cons}} = \ldots\\
+&}}}{
+\ldots
+}
+\end{align*}
+
 \section{Counterexamples to Preservation}\label{dot-preservation}

### 2012-08-05 article.tex
functions as sugar
basic ex

@@
 \newcommand{\abs}[3]{\lambda #1\!:\!#2.#3}
-% Application
-\newcommand{\app}[2]{#1\;#2}
 % Method Application
@@
 \newcommand{\reduction}[4]{#1 \operatorname{|} #2 \reduces #3 \operatorname{|} #4}
+
+% Sugar
+\newcommand{\arrow}[2]{#1\rightarrow_s#2}
+\newcommand{\fun}[4]{\textbf{fun } (#1:#2)\;#3\;#4}
+\newcommand{\app}[2]{(\textbf{app }#1\;#2)}
+\newcommand{\cast}[2]{(\textbf{cast }#1\;#2)}
 
@@
 \subsection{Sugar: Functions}
+Like in Scala, we can encode functions as objects with a special
+method. We will freely use the following sugar in the remaining of
+this paper. Note that the variable $z$ must be fresh.
+\begin{align*}
+\arrow S T &\iff {\Top \refine z {\mdecl {\mi{apply}} S T}}\\
+\fun x S T t &\iff \new z {\arrow S T} \ldefs{\mi{apply}(x) = t}\\
+\app f x &\iff \mapp f {\mi{apply}} x\\
+\cast T t&\iff \app {\fun x T T x} t
+\end{align*}
+
+In the remaining examples, for brevity, we will also use $\abs x S t$
+for $\fun x S \_ t$ where the return type can be somewhat inferred
+from the context.
 
@@
 This program defines a $\mi{root}$ object with basic types
-($\mi{Unit}$, $\mi{Boolean}$, $\mi{Nat}$) and values ($\mi{true}$,
-$\mi{false}$, $\mi{error}$, $\mi{zero}$). For brevity, I've omitted
+($\mi{Unit}$, $\mi{Boolean}$, $\mi{Nat}$) and methods ($\mi{true}$,
+$\mi{false}$, $\mi{error}$, $\mi{zero}$). For brevity, we've omitted
 the code related to natural numbers. During the object creation, the
-value labels such as $\mi{false}$ are all initialized.
+method labels such as $\mi{false}$ are all initialized.
 \begin{align*}
 &\mlnew {\mi{root}} {\Top \mlrefine r {\\\displaybreak[3]
-&\ \Ldecl {\mi{Unit}} \Bot \Top\\
-&\ \ldecl {\mi{unit}} {\Top \tfun {r.\mi{Unit}}}\\
-&\ \Ldecl {\mi{Boolean}} \Bot {\Top \mlrefine z {\\
-&\ \gap \ldecl {\mi{ifNat}} {(r.\mi{Unit} \tfun r.\mi{Nat}) \tfun (r.\mi{Unit} \tfun r.\mi{Nat}) \tfun r.\mi{Nat}}\\
-&\gap}}\\
-&\ \ldecl {\mi{false}} {{r.\mi{Unit}} \tfun {r.\mi{Boolean}}}\\
-&\ \ldecl {\mi{true}} {{r.\mi{Unit}} \tfun {r.\mi{Boolean}}}\\
+&\ \Ldecl {\mi{Unit}} \Bot \Top\\\displaybreak[3]
+&\ \mdecl {\mi{unit}} \Top {r.\mi{Unit}}\\\displaybreak[3]
+&\ \Ldecl {\mi{Boolean}} \Bot {\Top \mlrefine z {\\\displaybreak[3]
+&\ \gap \ldecl {\mi{ifNat}} \arrow {(\arrow {r.\mi{Unit}} {r.\mi{Nat}})} {\arrow {(\arrow {r.\mi{Unit}} {r.\mi{Nat}})} {r.\mi{Nat}}}\\\displaybreak[3]
+&\gap}}\\\displaybreak[3]
+&\ \mdecl {\mi{false}} {r.\mi{Unit}} {r.\mi{Boolean}}\\\displaybreak[3]
+&\ \mdecl {\mi{true}} {r.\mi{Unit}} {r.\mi{Boolean}}\\\displaybreak[3]
+&\ \mdecl {\mi{error}} {r.\mi{Unit}} \Bot\\\displaybreak[3]
+& \ \ldots\\\displaybreak[3]
+&}\mlldefs{\\\displaybreak[3]
+&\ \mi{unit}(x) = {\new u {\mi{root}.\mi{Unit}} u}\\\displaybreak[3]
+%
+&\ \mi{false}(u) =\\\displaybreak[3]
+&\ \gap \mlnew {\mi{ff}} {\mi{root}.\mi{Boolean} \mlldefs{\\\displaybreak[3]
+&\ \gap\gap {\mi{ifNat}} = \abs t {\arrow {\mi{root}.\mi{Unit}} {\mi{root}.\mi{Nat}}} {\\&\gap\gap\gap\abs e {\arrow {\mi{root}.\mi{Unit}} {\mi{root}.\mi{Nat}}}\\&\gap\gap\gap\app e {\mi{root}.\mi{unit}}}\\\displaybreak[3]
+&\ \gap}}{
+\ \gap \mi{ff}}\\\displaybreak[3]
+%
+&\ \mi{true}(u) = \ldots\\\displaybreak[3]
+%
+&\ \mi{error}(u) = {\app {\mi{root}.\mi{error}} u}\\\displaybreak[3]
 &\ \ldots\\\displaybreak[3]
-&\ \ldecl {\mi{error}} {{r.\mi{Unit}} \tfun \Bot}\\
-&}\mlldefs{\\\displaybreak[3]
-&\ \mi{unit}  = \abs x \Top {\new u {\mi{root}.\mi{Unit}} u}\\
-%
-&\ \mi{false} = \abs u {\mi{root}.\mi{Unit}} {\\
-&\ \gap \mlnew {\mi{ff}} {\mi{root}.\mi{Boolean} \mlldefs{\\\displaybreak[3]
-&\ \gap\gap {\mi{ifNat}} = \abs t {\mi{root}.\mi{Unit} \tfun \mi{root}.\mi{Nat}} {\\&\gap\gap\gap\abs e {\mi{root}.\mi{Unit} \tfun \mi{root}.\mi{Nat}}\\&\gap\gap\gap\app e {\mi{root}.\mi{unit}}}\\
-&\ \gap}}{
-\ \gap \mi{ff}}
-&}\\
-%
-&\ \mi{true} = \ldots\\
-%
-&\ \mi{error} = \abs u {\mi{root}.\mi{Unit}} {\app {\mi{root}.\mi{error}} u}\\
-&\ \ldots\\
 &}}{

### 2012-08-05 article.tex
list ex

@@
 &\ \Ldecl {\mi{List}} \Bot {\Top \mlrefine z {\\\displaybreak[3]
-&\ \gap \ldecl {\mi{isEmpty}} {\mi{root}.\mi{Unit} \tfun \mi{root}.\mi{Boolean}}\\
-&\ \gap \ldecl {\mi{head}} {\mi{root}.\mi{Unit} \tfun g.\mi{Elem}}\\
-&\ \gap \ldecl {\mi{tail}} {\mi{root}.\mi{Unit} \tfun g.\mi{ListOfElem}}\\
+&\ \gap \mdecl {\mi{isEmpty}} {\mi{root}.\mi{Unit}} {\mi{root}.\mi{Boolean}}\\
+&\ \gap \mdecl {\mi{head}} {\mi{root}.\mi{Unit}} {g.\mi{Elem}}\\
+&\ \gap \mdecl {\mi{tail}} {\mi{root}.\mi{Unit}} {g.\mi{ListOfElem}}\\
 &\gap}}\\
-&\ \ldecl {\mi{nil}} {\mi{root}.\mi{Unit} \tfun g.\mi{ListOfElem}}\\
-&\ \ldecl {\mi{cons}} {g.\mi{Elem} \tfun g.\mi{ListOfElem}}\\
+&\ \mdecl {\mi{nil}} {\mi{root}.\mi{Unit}} {g.\mi{ListOfElem}}\\
+&\ \mdecl {\mi{cons}} {g.\mi{Elem}} {g.\mi{ListOfElem}}\\
 &}}}}{
 \mlnew {\mi{natLists}} {{\mi{genLists}.\mi{ListPackage}} \mlrefine p { \Ldecl {\mi{Elem}} {\mi{Nat}} {\mi{Nat}} }\mlldefs{\\\displaybreak[3]
-&\ {\mi{nil}} = \ldots\\
-&\ {\mi{cons}} = \ldots\\
+&\ {\mi{nil}}(u) = \ldots\\
+&\ {\mi{cons}}(u) = \ldots\\
 &}}}{

### 2012-08-05 article.tex
some counterexs from report

@@
 \newcommand{\app}[2]{(\textbf{app }#1\;#2)}
+\newcommand{\mlapp}[2]{(\textbf{app }#1\;\\&#2)}
 \newcommand{\cast}[2]{(\textbf{cast }#1\;#2)}
@@
 
+We prove that the DOT calculus is type-safe using logical relations in
+section~\ref{dot-type-safety}. However, we first tried to prove the
+calculus type-safe using the standard theorems of preservation and
+progress. Unfortunately, for the calculus as presented, and any of its
+variants that we devised, preservation doesn't hold. In this section,
+we review some of the most salient counterexamples to preservation
+that we found. These counterexamples have been checked with PLT Redex.
+
+\subsection{\texorpdfstring{\textsc{term-$\ni$}}{Term-Mem} Restriction}\label{term_mem}
+
+There are two membership ($t \ni D$) rules: one for when the term $t$
+is a path, and one for an arbitrary term $t$. For paths, we can
+substitute the self-references in the declarations, but we cannot do
+so for arbitrary terms as the resulting types wouldn't be well-formed
+syntactically. Hence, the $\textsc{term-$\ni$}$ has the restriction
+that self-occurrences are not allowed. Here is a counterexample related
+to this restriction.
+
+Let $X$ be a shorthand for the type:
+\begin{align*}
+\Top & \mlrefine z {\\
+&\ \Ldecl {L_a} \Top \Top\\
+&\ \ldecl l {z.L_a}\\
+}&
+\end{align*}
+
+Let $Y$ be a shorthand for the type:
+\begin{align*}
+\Top & \mlrefine z {\\
+&\ \ldecl l \Top\\
+}&
+\end{align*}
+
+Now, consider the term
+\begin{align*}
+&\mlnew u {X \ldefs{ l = u }}{
+(\app {(\abs y {\Top \tfun Y} {\app y u})} {(\abs d \Top {(\app {(\abs x X x)} u)})}).l
+}
+\end{align*}
+
+The term type-checks because the term $t=(\app {(\abs y {\Top \tfun Y}
+  {\app y u})} {(\abs d \Top {(\app {(\abs x X x)} u)})})$ has type
+$Y$, so we can apply $\textsc{term-$\ni$}$ for $l$. However, the term
+$t$ eventually steps to $(\app {(\abs x X x)} u)$ which has type $X$,
+so we cannot apply $\textsc{term-$\ni$}$ for $l$ because of the
+self-reference ($z.L_a$).
+
+\subsection{Expansion Lost}\label{narrowing_exp}
+
+Expansion is not preserved by narrowing. Here, we create two type
+selections that are mutually recursive in their upper bounds after
+narrowing: $z_0.C_2$ initially expands, but after narrowing, $z_0.C_2$
+expands to what $z_0.A_2$ expands to, which expands to what $z_0.A_1$
+expands to, which expands to what $z_0.A_2$ expands to, and thus we
+have an infinite expansion. Thus, the last new expression initially
+type-checks, but after narrowing, it doesn't because the precise
+expansion needed by $\textsc{new}$ cannot be inferred.
+
+\begin{align*}
+&\mlnew {x_0} {\Top \mlrefine z { \Ldecl {A_1} \Bot {\Top \mlrefine z {\\
+&\gap\Ldecl {A_2} \Bot \Top\\
+&\gap\Ldecl {A_3} \Bot \Top\\
+&\gap\Ldecl {C_2} \Bot {z.A_2}}}}\ldefs{}}{
+\mlnew {x_1} {\Top \refine z {\Ldecl {C_1} \Bot {\Top \refine z {\Ldecl {A_1} \Bot {x_0.A_1}}}}\ldefs{}}}{
+\mlnew {x_2} {{x_1.C_1} \refine z {\Ldecl {A_1} \Bot {{x_0.A_1} \refine z {\Ldecl {A_2} \Bot {z.A_3}}}}\ldefs{}}}{
+\mlnew {x_3} {{x_1.C_1} \refine z {\Ldecl {A_1} \Bot {{x_0.A_1} \refine z {\Ldecl {A_3} \Bot {z.A_2}}}}\ldefs{}}}{
+\mlapp {\abs x {x_1.C_1} {(\abs {z_0} {x.A_1 \tand x_3.A_1} {\\&\gap\gap\gap\new z {z_0.C_2} {\app {(\abs x \Top x)} z}})}} {\ x_2}}
+\end{align*}
+
+\subsection{Well-Formedness Lost}\label{narrowing_wf}
+
+Even well-formedness is not preserved by narrowing. The trick is that
+if the lower bound of a type selection is not $\Bot$, then the
+bounds needs to be checked for well-formedness. Here, we create two
+type selections that are mutually recursive in their bounds
+after narrowing. $y.A$ is initially well-formed, but after narrowing,
+it isn't because we run into an infinite derivation trying to prove
+the well-formedness of its bounds.
+
+\begin{align*}
+&\mlnew v {\Top \refine z {\Ldecl L \Bot {\Top \refine z {\Ldecl A \Bot \Top, \Ldecl B {z.A} {z.A}}}} \ldefs{}}{
+\mlapp {\abs x {\Top \refine z {\Ldecl L \Bot {\Top \refine z {\Ldecl A \Bot \Top, \Ldecl B \Bot \Top}}}} {\\&\gap
+\mlnew z {\Top \refine z {\ldecl l {\Bot \tfun \Top}}\mlldefs{\\&\gap\gap l = \abs y {x.L \tand {\Top \refine z {\Ldecl A {z.B} {z.B}, \Ldecl B \Bot \Top}}} {\\&\gap\gap\gap\abs a {y.A} {\app {(\abs x \Top x)} a}}}}{
+\ \app {(\abs x \Top x)} z
+}}}{\ v}}
+\end{align*}
+
+\subsection{Path Equality}\label{patheq}
+
+We need to be able to relate path-dependent types after reduction. The
+original DOT calculus doesn't have any rules for dealing with
+path-equality. Here is an example which motivates the need for
+path-equality provisions.
+
+\begin{align*}
+&\mlnew b {\Top \mlrefine z {&&\Ldecl X \Top \Top&\\
+&&&\ \ldecl l {z.X}&}\ldefs{l = b}}{
+\mlnew a {\Top \mlrefine z {\ldecl i {\Top \mlrefine z {&&&\\
+&&&\Ldecl X \Bot \Top\\
+&&&\ldecl l {z.X}}}&}\ldefs{i = b}}}{
+\app {(\abs x \Top x)} {(\app {(\abs x {a.i.X} x)} {a.i.l})}}
+\end{align*}
+
+$a.i.l$ reduces to $b.l$. $b.l$ has type $b.X$, so we need $b.X <:
+a.i.X$. This cannot be established with the current rules: it is not
+true in general, but true here because $a.i$ reduces to $b$. Hence,
+the need for acknowledging path equality.
+
 \section{Type Safety via Logical Relations}\label{dot-type-safety}
+
+\section{Discussion}\label{discussion}
 

### 2012-08-05 article.tex
simplify apps

@@
 &\mlnew u {X \ldefs{ l = u }}{
-(\app {(\abs y {\Top \tfun Y} {\app y u})} {(\abs d \Top {(\app {(\abs x X x)} u)})}).l
+\app {(\abs y {\Top \tfun Y} {\app y u})} {(\abs d \Top {\app {(\abs x X x)} u})}.l
 }
@@
 
-The term type-checks because the term $t=(\app {(\abs y {\Top \tfun Y}
-  {\app y u})} {(\abs d \Top {(\app {(\abs x X x)} u)})})$ has type
+The term type-checks because the term $t=\app {(\abs y {\Top \tfun Y}
+  {\app y u})} {(\abs d \Top {\app {(\abs x X x)} u)}}$ has type
 $Y$, so we can apply $\textsc{term-$\ni$}$ for $l$. However, the term
-$t$ eventually steps to $(\app {(\abs x X x)} u)$ which has type $X$,
+$t$ eventually steps to $\app {(\abs x X x)} u$ which has type $X$,
 so we cannot apply $\textsc{term-$\ni$}$ for $l$ because of the
@@
 &&&\ldecl l {z.X}}}&}\ldefs{i = b}}}{
-\app {(\abs x \Top x)} {(\app {(\abs x {a.i.X} x)} {a.i.l})}}
+\app {(\abs x \Top x)} {\app {(\abs x {a.i.X} x)} {a.i.l}}}
 \end{align*}

### 2012-08-05 article.tex
rewording

@@
 
-We need to be able to relate path-dependent types after reduction. The
-original DOT calculus doesn't have any rules for dealing with
-path-equality. Here is an example which motivates the need for
-path-equality provisions.
+For preservation, we need to be able to relate path-dependent types
+after reduction. Here is a motivating example:
 
@@
 true in general, but true here because $a.i$ reduces to $b$. Hence,
-the need for acknowledging path equality.
+the need for acknowledging path equality for preservation to hold.
 

### 2012-08-05 article.tex
update counterex to latest calculus
formatting
add numbered subsections throughout
a bit of discussion

@@
 
-\subsection*{Why No Inheritance?}
+\subsection{Why No Inheritance?}
 
@@
 
-\subsection*{Notation} We use standard notational conventions for
+\subsection{Notation} We use standard notational conventions for
 sets. The notation $\seq{X}$ denotes a set of elements $X$. Given a
@@
 
-\subsection*{Syntax}
+\subsection{Syntax}
 
@@
 
-\subsection*{Reduction rules}
+\subsection{Reduction rules}
 
@@
 
-\subsection*{Type assignment rules}
+\subsection{Type assignment rules}
 
@@
 
-\subsection*{Membership}
+\subsection{Membership}
 
@@
 
-\subsection*{Expansion}
+\subsection{Expansion}
 
@@
 
-\subsection*{Subtyping}
+\subsection{Subtyping}
 
@@
 
-\subsection*{Declaration Subsumption}
+\subsection{Declaration Subsumption}
 
@@
 
-\subsection*{Well-formedness}
+\subsection{Well-formedness}
 
@@
 &\mlnew v {\Top \refine z {\Ldecl L \Bot {\Top \refine z {\Ldecl A \Bot \Top, \Ldecl B {z.A} {z.A}}}} \ldefs{}}{
-\mlapp {\abs x {\Top \refine z {\Ldecl L \Bot {\Top \refine z {\Ldecl A \Bot \Top, \Ldecl B \Bot \Top}}}} {\\&\gap
-\mlnew z {\Top \refine z {\ldecl l {\Bot \tfun \Top}}\mlldefs{\\&\gap\gap l = \abs y {x.L \tand {\Top \refine z {\Ldecl A {z.B} {z.B}, \Ldecl B \Bot \Top}}} {\\&\gap\gap\gap\abs a {y.A} {\app {(\abs x \Top x)} a}}}}{
-\ \app {(\abs x \Top x)} z
-}}}{\ v}}
+\mlapp {(\abs x {\Top \refine z {\Ldecl L \Bot {\Top \refine z {\Ldecl A \Bot \Top, \Ldecl B \Bot \Top}}}} {\\&\gap
+\mlnew z {\Top \mlrefine z {\\&\gap\gap\gap
+\mdecl l {x.L \tand {\Top \refine z {\Ldecl A {z.B} {z.B}, \Ldecl B \Bot \Top}}} \Top}\mlldefs{\\&\gap\gap\gap
+l(y) = \fun a {y.A} \Top a}}{
+\gap \cast {\top} z
+}})}{\ v}}
 \end{align*}
@@
 
+\subsection{Variants of the DOT Calculus}\label{dot-variants}
+
+\subsubsection{Why limit the calculus to $\wfe$-types?}
+
+Currently, the proof of type-safety via logical relations
+fundamentally relies on types having an expansion. However, this was
+not our original motivation for limiting the calculus to $\wfe$-types.
+
+Originally, subtyping was not regular with respect to $\wfe$. Roughly,
+all the $\wfe$ preconditions in subtyping were dropped. In this
+broader calculus, subtyping transivitiy doesn't hold, because of the
+rule (\textsc{$\sub$-rfn}) which requires expansion of the left type.
+
+The problem is deep, as attested by this elaborate counterexample
+that is not so easily patched, and directly leads to a counterexample
+to preservation.
+
+Consider an environment where $u$ is bound to:
+\begin{align*}
+\Top & \mlrefine u {\\
+&\ \Ldecl {\mathit{Bad}} {\Bot} {u.\mathit{Bad}}\\
+&\ \Ldecl {\mathit{Good}} {\Top \refine z {\Ldecl L \Bot \Top}} {\Top \refine z {\Ldecl L \Bot \Top}}\\
+&\ \Ldecl {\mathit{Lower}} {u.\mathit{Bad} \tand u.\mathit{Good}} {u.\mathit{Good}}\\
+&\ \Ldecl {\mathit{Upper}} {u.\mathit{Good}} {u.\mathit{Bad} \tor u.\mathit{Good}}\\
+&\ \Ldecl X {u.\mathit{Lower}} {u.\mathit{Upper}}\\
+}&
+\end{align*}
+
+Now, consider the types $S$, $T$, $U$ defined in terms of $u$:
+\begin{align*}
+S &= u.\mathit{Bad} \tand u.\mathit{Good}\\
+T &= u.\mathit{Lower}\\
+U &= u.X \refine z {\Ldecl L \Bot \Top}
+\end{align*}
+We have $S \sub T$ and $T \sub U$, but we cannot derive $S \sub U$ because
+$S$ doesn't expand.
+
+Note that $u$ is realizable, since each lower bound is a subtype of
+its upper bound. So it is straightforward to turn this counterexample
+to subtyping transivity into a counterexample to preservation:
+\begin{align*}
+&\mlnew u \ldots {
+\ \mlapp{\abs x \Top x}{
+\ \ \mlapp{\abs f {S \tfun U} f}{
+\ \ \ \mlapp{\abs f {S \tfun T} f}{
+\ \ \ \ \mlapp{\abs f {S \tfun S} f}{
+\ \ \  \ \ \abs x S x}}}}}
+\end{align*}
+
+The idea is to start with a function from $S \tfun S$ and cast it
+successively to $S \tfun T$ then $S \tfun U$. To typecheck the
+expression initially, we need to check $S \sub T$ and $T \sub
+U$. After some reduction steps, the first few casts vanish, and the
+reduced expression casts directly from $S \tfun S$ to $S \tfun U$, so
+we need to check $S \sub U$. Thus, we need subtyping transitivity: $S
+\sub T$ and $T \sub U$ implies $S \sub U$.
+
 \section{Conclusion}\label{conclusion}

### 2012-08-05 article.tex
discuss methods vs lambda

@@
 
+\subsubsection{Why not include the lambda-calculus instead of methods?}
+
+Originally, the DOT calculus included the lambda-calculus, and
+explicit methods were not needed since they could be represented by a
+value label with a function type. However, the expansion of the
+function type was defined to be the empty set of declarations (like
+for $\Top$), which caused a real breach of type-safety.
+
+A concrete object could be a subtype of a function type without a
+function ever being defined. Consider:
+\begin{align*}
+&\mlnew u {\Top \refine z {\Ldecl C {\Top \tfun \Top} {\Top \tfun \Top}} \ldefs{}} {
+\mlnew f {u.C \ldefs{}} {
+\ldots
+}}
+\end{align*}
+
+Now, $f$ is a subtype of $\Top \tfun \Top$, but $\app f {(\abs x \Top
+  x)}$ is stuck (and, rightfully, doesn't typecheck). But we can use
+narrowing to create a counterexample to type safety: $\app {(\abs g
+  {\Top \tfun \Top} {\app g (\abs x \Top x)})} f$.
+
+Because of this complication, we decided to drop the lambda-calculus
+from DOT, and instead introduce methods with one parameter. Like in
+Scala, functions are then just sugar for objects with a special
+method.
+
+An alternative design would have been to change the expansion of the
+function type to have a special marker value label.
+
 \section{Conclusion}\label{conclusion}

### 2012-08-05 article.tex
why not make preservation hold

@@
 
+\subsubsection{Why not patch the DOT calculus for preservation to hold?}
+
+We tried! However, the resulting calculi were not elegant, and
+furthermore, we still found issues with preservation. Below, we give a
+brief summary of one failed attempt to patch the calculus for
+preservation to hold.
+
+Because many of the counterexamples to preservation are related to
+narrowing, we tried to make widening an explicit operation and change
+rules with implicit relaxations (\textsc{msel} and \textsc{new}) to be
+strict. From a typing perspective, the change was straightforward, but
+reduction became more complicated and dependent on typing because the
+type information in widenings needed to be propagated correctly.
+
+We added path equality provisions in the subtyping rules, in the same
+spirit as the Tribe calculus.
+
+Unfortunately, these two patches interacted badly, and we were left
+with a disturbing counterexample to type safety.
+
+\begin{align*}
+&\mlnew a {\Top \refine z {\Ldecl C \Bot {\Top \refine z {\Ldecl D \Bot {z.X}, \Ldecl X \Bot \Top}}}}{
+\mlnew b {{a.C} \refine z {\Ldecl X \Bot \Bot}}}{
+\mlnew c {a.C}}{
+\mlnew d {{\cast {a.C} b}.D}}{
+\app {(\abs x \Bot {x.\mathit{foo}})} d
+}
+\end{align*}
+
+Notice that $d$ has type $\Bot$ if you ignore the cast on $b$. This
+example doesn't typecheck initially because the path-equality
+provisions only applies when objects are in the store, so the
+application is not well-typed. But if we start preservation in a store
+which has $a$, $b$, $c$ and $d$ then the application type-checks,
+because, through one of the path-equality provision, we can find that
+the type of $d$ is a subtype of $\Bot$. Now, of course, when we get to
+$d.\mathit{foo}$, reduction fails.
+
 \section{Conclusion}\label{conclusion}

### 2012-08-05 article.tex
bib

@@
 as well as its mixture of nominal and structural typing through the
-use of refinement types. Compared to previous approaches (nuObj, FS),
+use of refinement types. Compared to previous approaches~\cite{nuObj,FS},
 we make no attempt to model inheritance or mixin composition. Indeed
@@
 calculus type-safe using the standard theorems of preservation and
-progress. Unfortunately, for the calculus as presented, and any of its
-variants that we devised, preservation doesn't hold. In this section,
-we review some of the most salient counterexamples to preservation
-that we found. These counterexamples have been checked with PLT Redex.
+progress~\cite{soundness,tapl}. Unfortunately, for the calculus as
+presented, and any of its variants that we devised, preservation
+doesn't hold. In this section, we review some of the most salient
+counterexamples to preservation that we found. These counterexamples
+have been checked with PLT Redex.
 
@@
 \section{Type Safety via Logical Relations}\label{dot-type-safety}
+We prove type-safety of the DOT calculus via step-indexed logical
+relations~\cite{ahmed04,ahmed06,step_indexed_obj}.
 
@@
 We added path equality provisions in the subtyping rules, in the same
-spirit as the Tribe calculus.
+spirit as the Tribe calculus~\cite{tribe}.
 

### 2012-08-05 article.tex
revision: one pass through

@@
 path-dependent types and abstract type members, as well as its mixture
-of nominal and structural typing through the use of reﬁnement
+of nominal and structural typing through the use of refinement
 types. It makes no attempt to model inheritance or mixing
@@
 \category{D.3.3}{Language Constructs and Features}{Abstract data types, Classes and objects, polymorphism}
-\category{D.3.1}{Formal Deﬁnitions and Theory}{Syntax, Semantics}
+\category{D.3.1}{Formal Definitions and Theory}{Syntax, Semantics}
 \category{F.3.1}{Specifying and Verifying and Reasoning about Programs}{}
@@
 
-The types in DOT are denoted by letters $S$, $T$, $U$, or $V$. They consist of the following:
+The types in DOT are denoted by letters $S$, $T$, $U$, $V$, or $W$. They consist of the following:
 \begin{itemize}
@@
 
-There are only three forms of declarations in DOT, which are both part
+There are only three forms of declarations in DOT, which are all part
 of refinement types.  A value declaration $\ldecl l T$ introduces a
@@
 There are three main reduction rules: Rule (\textsc{msel}) rewrites a
-method invocation $y {m_i} v$ by retrieving the corresponding method
+method invocation $\mapp y {m_i} v$ by retrieving the corresponding method
 definition from the store, and performing a substitution of the
@@
 one of its members. The membership relation is defined in
-\figref{dot-two} and is further explained below.
+\figref{dot-two} and is further explained in section~\ref{mem}.
 
@@
 
-\subsection{Membership}
-
-\figref{dot-two} presents typing rules for membership and
-expansion.  The membership judgement $\Gamma \ts t \ni D$ states that
-in environment $\Gamma$ a term $t$ has a declaration $D$ as a member.
-There are different rules for paths and general terms.  Rule
-(\textsc{path-$\ni$}) applies to paths $p$ that have a refinement type
-$T \refine z {\seq D}$ as their type. Members of $p$ are then all
-definitions $D$ in the refinement, where any use of the self reference
-$z$ is replaced by the path $p$ itself.  Rule (\textsc{and-$\ni$})
-allows to merge two member definitions (of the same label) by
-conjoining them with $\tand$. Rule (\textsc{term-$\ni$}) establishes
-the members of general terms $t$ of type $T$ by introducing a dummy
-variable $z$ of type $T$ and then using the rules for path membership
-on $z$. It must hold that $z$ itself does not form part of the
-resulting member $D$.
+\subsection{Membership}\label{mem}
+
+\figref{dot-two} presents typing rules for membership and expansion.
+The membership judgement $\Gamma \ts t \ni D$ states that in
+environment $\Gamma$ a term $t$ has a declaration $D$ as a member. The
+membership rules rely on expansion. There are two rules, one for paths
+(\textsc{path-$\ni$}) and one for general terms
+(\textsc{term-$\ni$}). For general terms, the ``self''-reference of
+the type must not occur in the resulting declaration $D$, since, to
+guarantee syntactic validity, we can only substitute a path for the
+``self''-reference.
 
@@
 set of declarations carried by a concrete type that is used in a
-\textbf{new}-expression. Expansion is also used in subtyping
-refinements on the right (see \figref{dot-sub}).
+\textbf{new}-expression. Expansion is also used by the membership
+rules and in subtyping refinements on the right (see
+\figref{dot-sub}).
 
@@
 \sub T$ which states that in environment $\Gamma$ type $S$ is a
-subtype of type $T$. Subtyping is regular with respect to $\wfe$: if
+subtype of type $T$. Subtyping is regular wrt $\wfe$: if
 type $S$ is a subtype of type $T$, then $S$ and $T$ are well-formed
@@
 counterexamples to preservation that we found. These counterexamples
-have been checked with PLT Redex.
+have been checked with PLT Redex~\cite{plt_redex}.
 
@@
 &\mlnew u {X \ldefs{ l = u }}{
-\app {(\abs y {\Top \tfun Y} {\app y u})} {(\abs d \Top {\app {(\abs x X x)} u})}.l
+\app {(\abs y {\Top \tfun Y} {\app y u})} {(\abs d \Top {\cast X u})}.l
 }
@@
 The term type-checks because the term $t=\app {(\abs y {\Top \tfun Y}
-  {\app y u})} {(\abs d \Top {\app {(\abs x X x)} u)}}$ has type
+  {\app y u})} {(\abs d \Top {\cast X u})}$ has type
 $Y$, so we can apply $\textsc{term-$\ni$}$ for $l$. However, the term
-$t$ eventually steps to $\app {(\abs x X x)} u$ which has type $X$,
+$t$ eventually steps to ${\cast X u}$ which has type $X$,
 so we cannot apply $\textsc{term-$\ni$}$ for $l$ because of the
@@
 
+% TODO: section
+
 \section{Discussion}\label{discussion}
@@
 
-Originally, subtyping was not regular with respect to $\wfe$. Roughly,
+Originally, subtyping was not regular wrt $\wfe$. Roughly,
 all the $\wfe$ preconditions in subtyping were dropped. In this
@@
 reduced expression casts directly from $S \tfun S$ to $S \tfun U$, so
-we need to check $S \sub U$. Thus, we need subtyping transitivity: $S
-\sub T$ and $T \sub U$ implies $S \sub U$.
+we need to check $S \sub U$.
 
@@
 
-Now, $f$ is a subtype of $\Top \tfun \Top$, but $\app f {(\abs x \Top
-  x)}$ is stuck (and, rightfully, doesn't typecheck). But we can use
+Now, $f$ was a subtype of $\Top \tfun \Top$, but $\app f {(\abs x \Top
+  x)}$ was stuck (and, rightfully, didn't typecheck). But we could use
 narrowing to create a counterexample to type safety: $\app {(\abs g
@@
 An alternative design would have been to change the expansion of the
-function type to have a special marker value label.
+function type to have a declaration for a special marker value label.
 
@@
 
-Notice that $d$ has type $\Bot$ if you ignore the cast on $b$. This
-example doesn't typecheck initially because the path-equality
-provisions only applies when objects are in the store, so the
-application is not well-typed. But if we start preservation in a store
-which has $a$, $b$, $c$ and $d$ then the application type-checks,
-because, through one of the path-equality provision, we can find that
-the type of $d$ is a subtype of $\Bot$. Now, of course, when we get to
-$d.\mathit{foo}$, reduction fails.
+Notice that $d$ has type $\Bot$ if the cast on $b$ is ignored. This
+example didn't typecheck initially because the path-equality
+provisions only applied when objects are in the store, so the
+application was not well-typed. But if we started preservation in a store
+which has $a$, $b$, $c$ and $d$ then the application type-checked,
+because, through one of the path-equality provision, we could find that
+the type of $d$ was a subtype of $\Bot$. Now, of course, when we got to
+$d.\mathit{foo}$, reduction failed.
+
+% TODO: more discussion subsections?
 
 \section{Conclusion}\label{conclusion}
+% TODO: section
 

### 2012-08-06 article.tex
remote template blurb
fix fun sugar
make todos and remarks explicit
grammar
sketch instead of prove

@@
-%-----------------------------------------------------------------------------
-%
-%               Template for sigplanconf LaTeX Class
-%
-% Name:         sigplanconf-template.tex
-%
-% Purpose:      A template for sigplanconf.cls, which is a LaTeX 2e class
-%               file for SIGPLAN conference proceedings.
-%
-% Guide:        Refer to "Author's Guide to the ACM SIGPLAN Class,"
-%               sigplanconf-guide.pdf
-%
-% Author:       Paul C. Anagnostopoulos
-%               Windfall Software
-%               978 371-2316
-%               paul@windfall.com
-%
-% Created:      15 February 2005
-%
-%-----------------------------------------------------------------------------
-
 \documentclass[9pt]{sigplanconf}
@@
 \title{Dependent Object Types}
-\subtitle{A foundations for Scala's type system} % TODO: ask
+\subtitle{Towards a foundation for Scala's type system}
 
@@
 sound, by exhibiting counterexamples to preservation. Nevertheless, we
-prove type-safety of the calculus via step-indexed logical relations.
+sketch a proof of type-safety of the calculus via step-indexed logical
+relations.
 \end{abstract}
@@
 
-% TODO: ask: does this last point still makes sense wrt to type safety via logical relations instead of via the standard theorems of preservation and progress?
+\remark{Does this last point still makes sense wrt to type safety via
+  logical relations instead of via the standard theorems of
+  preservation and progress?}
 
@@
 \arrow S T &\iff {\Top \refine z {\mdecl {\mi{apply}} S T}}\\
-\fun x S T t &\iff \new z {\arrow S T} \ldefs{\mi{apply}(x) = t}\\
+\fun x S T t &\iff \new z {{\arrow S T}\ldefs{\mi{apply}(x) = t}} z\\
 \app f x &\iff \mapp f {\mi{apply}} x\\
@@
 
-We prove that the DOT calculus is type-safe using logical relations in
-section~\ref{dot-type-safety}. However, we first tried to prove the
-calculus type-safe using the standard theorems of preservation and
-progress~\cite{soundness,tapl}. Unfortunately, for the calculus as
-presented, and any of its variants that we devised, preservation
-doesn't hold. In this section, we review some of the most salient
-counterexamples to preservation that we found. These counterexamples
-have been checked with PLT Redex~\cite{plt_redex}.
+We sketch a proof that the DOT calculus is type-safe using logical
+relations in section~\ref{dot-type-safety}. However, we first tried to
+prove the calculus type-safe using the standard theorems of
+preservation and progress~\cite{soundness,tapl}. Unfortunately, for
+the calculus as presented, and any of its variants that we devised,
+preservation doesn't hold. In this section, we review some of the most
+salient counterexamples to preservation that we found. These
+counterexamples have been checked with PLT Redex~\cite{plt_redex}.
 
@@
 true in general, but true here because $a.i$ reduces to $b$. Hence,
-the need for acknowledging path equality for preservation to hold.
+we need to acknowledge path equality for preservation to hold.
 
 \section{Type Safety via Logical Relations}\label{dot-type-safety}
-We prove type-safety of the DOT calculus via step-indexed logical
-relations~\cite{ahmed04,ahmed06,step_indexed_obj}.
-
-% TODO: section
+We sketch a proof of type-safety of the DOT calculus via step-indexed
+logical relations~\cite{ahmed04,ahmed06,step_indexed_obj}.
+
+\todo{complete proof!}
 
@@
 
-% TODO: more discussion subsections?
+\todo{more discussion subsections?}
 
 \section{Conclusion}\label{conclusion}
-% TODO: section
+\todo{conclusion}
 

### 2012-08-06 article.tex
defs of logical relations
fix indices
rel s in addition to g, do not overload, acks todo
logical rels

@@
 \usepackage{amsmath}
+\usepackage{amssymb}
 
@@
 \usepackage{framed}
+
+\usepackage{stmaryrd}
 
@@
 \newcommand{\lindent}{\hspace{-4mm}}
+
+% Logical relations
+\newcommand{\relv}[4]{\mathcal{V}_{#1;#2;#3}\llbracket#4\rrbracket}
+\newcommand{\rele}[4]{\mathcal{E}_{#1;#2;#3}\llbracket#4\rrbracket}
+\newcommand{\rels}[3]{\mathcal{S}_{#1;#2}\llbracket#3\rrbracket}
+\newcommand{\relg}[3]{\mathcal{G}_{#1;#2}\llbracket#3\rrbracket}
+\newcommand{\irred}[2]{\text{irred }(#1,#2)}
+\newcommand{\andl}{\;\wedge}
+\newcommand{\orl}{\vee}
+\newcommand{\impliesl}{\rightarrow}
+\newcommand{\reductionl}[5]{#1 \operatorname{|} #2 \;\rightarrow^{#5}\; #3 \operatorname{|} #4}
+\newcommand{\ds}{\,\vDash\,}
 
@@
 
-\todo{complete proof!}
+\subsection{Type Safety}
+Type-safety states that a well-typed program doesn't get stuck. More
+formally: If $- \ts t \typ T$ and $\reductionl t - {t'} {s'} *$ then
+either $t'$ is a value or $\exists t'', s'', \reduction {t'} {s'}
+{t''} {s''}$.
+
+Our strategy is to define a logical relation $\Gamma \ds t : T$, such
+that $\Gamma \ts t : T$ implies $\Gamma \ds t : T$ implies
+type-safety.
+
+\subsection{Logical Relations}
+
+\subsubsection{$\mathcal{V}$}
+$\relv k \Gamma s T$ defines the set of values that appear to have
+type $T$ when taking at most $k$ steps. $\Gamma$ and $s$ must agree:
+$\dom(\Gamma) = \dom(s)$ and $\forall (x : T) \in \Gamma, x \in \relv
+k \Gamma s T$. A variable $y$ belongs to $\relv 0 \Gamma s T$ simply
+by being in the store. In addition, it belongs to $\relv k \Gamma s T$
+for $k > 0$, if it defines all method and value labels in the
+expansion of $T$ appropriately for $j < k$ steps.
+\begin{align*}
+&\relv k \Gamma s T = \{ y \ |\\
+&\gap \Gamma \ts T \wfe \andl\\
+&\gap y \in \dom(s) \andl\\
+&\gap \forall j < k,\\
+&\gap y \mapsto T_c \ldefs{\seq{l = v}\;\seq{m(x)=t}} \in s,\\
+&\gap \Gamma \ts T \expand_y \seq{D},\\
+&\gap (\forall m_i : S \rightarrow U \in \seq{D},\\
+&\gap\gap t_i \in \rele j {\Gamma \envplus{x : S}} s U)\\
+&\gap \andl\\
+&\gap (\forall l_i : V \in \seq{D},\\
+&\gap\gap v_i \in \relv j \Gamma s V)\\
+&\}
+\end{align*}
+
+\subsubsection{$\mathcal{E}$}
+$\rele k \Gamma s T$ defines the set of terms that appear to have type
+$T$ when taking at most $k$ steps. $s$ must agree with $\Gamma$, but
+$\Gamma$ can additionaly contain variables not in $s$. This is needed
+for checking methods in $\mathcal{V}$ above, and for relating open
+terms. If $k > 0$, $\mathcal{E}$ completes the store to agree with
+$\Gamma$. It then states that if it can reduce $t$ in the completed
+store to an irreducible term in $j < k$ steps, then this term must be
+in a corresponding $\mathcal{V}$ set.
+\begin{align*}
+&\rele k \Gamma s T = \{ t \ |\\
+&\gap \Gamma \ts T \wfe \andl\\
+&\gap k = 0 \orl (\forall j \leq k,\\
+&\gap\gap \forall s' \in \rels k \Gamma s ,\\
+&\gap\gap \reductionl t {s'} {t'} {s''} j \andl\\
+&\gap\gap \irred {t'} {s''} \impliesl\\
+&\gap\gap \forall \Gamma' \in \relg k {s''} \Gamma ,\\
+&\gap\gap {t'} \in \relv {k-j-1} {\Gamma'} {s''} T)\\
+&\}
+\end{align*}
+
+\subsubsection{$\mathcal{S}$}
+$\rels k \Gamma s$ for $k > 0$ defines the set of completed stores
+that agree with $\Gamma$ on each variable when taking at most $k-1$
+steps.
+\begin{align*}
+&\rels k \Gamma s = \{ s, \seq{x_i \mapsto c_i}^{m \leq i < n} \ |\\
+&\gap s = \seq{x \mapsto c}^m \andl \Gamma = \seq{x : T}^n \andl \\
+&\gap m\leq n \andl \forall i, m \leq i < n,\\
+&\gap c_i \in \relv {k-1} \Gamma {s, \seq{x_j \mapsto c_j}^{m \leq j < i}} {T_i}\\
+&\}
+\end{align*}
+
+\subsubsection{$\mathcal{G}$}
+$\relg k s \Gamma$ defines a singleton set of a completed environment
+that agrees with a store $s$ by simply copying the constructor type
+from the store for each missig variable.
+\begin{align*}
+&\relg k s \Gamma = \{ \Gamma, \seq{x_i : T_{c_i}}^{m \leq i < n} \ |\\
+&\gap \Gamma = \seq{x : T}^m \andl s = \seq{x \mapsto c}^n \andl\\
+&\}
+\end{align*}
+
+\subsubsection{$\Gamma \ds t : T$}
+
+$\Gamma \ds t : T$ is simplfy defined as $t \in \rele k \Gamma - T,
+\forall k$.
+
+\subsection{Proof of the Fundamental Theorem}
+
+The fundamental theorem is the implication from $\Gamma \ts t : T$ to
+$\Gamma \ds t : T$. Type safety is a straightforward corollary of this
+theorem.
 
@@
 %\section{Appendix Title}
-%\acks
+
+\acks
+\todo{acks}
 

### 2012-08-06 article.tex
some cases of proof

@@
 \newcommand{\irred}[2]{\text{irred }(#1,#2)}
-\newcommand{\andl}{\;\wedge}
+\newcommand{\andl}{\;\wedge\;}
 \newcommand{\orl}{\vee}
@@
 
+\begin{myproof}
+The proof is on induction on the derivation of $\Gamma \ts t : T$.
+
+\Case \textsc{var}: $\Gamma \ts x : T$ knowing $(x : T) \in
+\Gamma$.
+
+Need to show $x \in \rele k \Gamma - T$. Suppose $k > 0$ and $s \in
+\rels k \Gamma -$. Remains to show $x \in \relv {k-1} \Gamma s
+T$. This follows from the definition of $\mathcal{G}$.
+
+\Case \textsc{sel}: $\Gamma \ts t_1.l_i : T$ knowing $\Gamma \ts t_1 :
+T_1$, $\Gamma \ts T_1 \expand_z \seq{D}$, $l_i : V_i \in \seq{D}$ and
+knowing either that $t_1 = p_1 \andl T = \subst p z {V_i}$ or that $z
+\not\in \fn(V_i) \andl T = V_i$.
+
+Need to show $t_1.l_i \in \rele k \Gamma - T$. Suppose $k > 0$ and $s
+\in \rels k \Gamma -$. Suppose $\reductionl {t_1.l_i} s {t'} {s'}
+\andl \irred {t'} {s'} \andl \Gamma' \in \relg \Gamma k {s'}
+\Gamma$. Remains to show ${t'} \in \relv {k-j-1} {\Gamma'} {s'} T$.
+
+By operational semantics, $\reductionl {t_1} s {t_1'} {s'} {j-1}$ and
+$\irred {t_1'} {s'}$. By induction hypothesis, ${t_1'} \in \relv
+{k-j+1-1} {\Gamma'} {s'} {T_1}$.
+
+By operational semantics, $\reductionl {t_1'.l_i} {s'} {t'} {s'}
+1$. Since ${t_1'} \in \relv {k-j} {\Gamma'} {s'} {T_1}$, we can
+conclude $t' \in \relv {k-j-1} {\Gamma'} {s'} T$ from the clause for
+value labels.
+
+\Case \textsc{msel}: \todo{case}.
+
+\Case \textsc{new}: \todo{case}.
+
+\end{myproof}
+
 \section{Discussion}\label{discussion}

### 2012-08-06 article.tex
case msel, not so convincing

@@
 \Case \textsc{sel}: $\Gamma \ts t_1.l_i : T$ knowing $\Gamma \ts t_1 :
-T_1$, $\Gamma \ts T_1 \expand_z \seq{D}$, $l_i : V_i \in \seq{D}$ and
-knowing either that $t_1 = p_1 \andl T = \subst p z {V_i}$ or that $z
-\not\in \fn(V_i) \andl T = V_i$.
+T_1$, $\Gamma \ts T_1 \expand_z \seq{D}$, $\ldecl {l_i} {V_i} \in
+\seq{D}$ and knowing either that $t_1 = p_1 \andl T = \subst p z
+    {V_i}$ or that $z \not\in \fn(V_i) \andl T = V_i$.
 
 Need to show $t_1.l_i \in \rele k \Gamma - T$. Suppose $k > 0$ and $s
-\in \rels k \Gamma -$. Suppose $\reductionl {t_1.l_i} s {t'} {s'}
-\andl \irred {t'} {s'} \andl \Gamma' \in \relg \Gamma k {s'}
-\Gamma$. Remains to show ${t'} \in \relv {k-j-1} {\Gamma'} {s'} T$.
-
-By operational semantics, $\reductionl {t_1} s {t_1'} {s'} {j-1}$ and
-$\irred {t_1'} {s'}$. By induction hypothesis, ${t_1'} \in \relv
+\in \rels k \Gamma -$. Suppose, for some $j \leq k$, $\reductionl
+    {t_1.l_i} s {t'} {s'} j \andl \irred {t'} {s'} \andl \Gamma' \in
+    \relg \Gamma k {s'} \Gamma$. Remains to show ${t'} \in \relv
+          {k-j-1} {\Gamma'} {s'} T$.
+
+By operational semantics and induction hypothesis, $\reductionl {t_1}
+s {t_1'} {s'} {j-1}$ and $\irred {t_1'} {s'}$ and ${t_1'} \in \relv
 {k-j+1-1} {\Gamma'} {s'} {T_1}$.
@@
 
-\Case \textsc{msel}: \todo{case}.
+\Case \textsc{msel}: $\Gamma \ts t_1.m_i(t_2) : T$ knowing $\Gamma \ts
+t_1 : T_1$, $\Gamma \ts t_2 : T_2$, $\Gamma \ts T_1 \expand_z
+\seq{D}$, $\mdecl {m_i} {S_i} {U_i} \in \seq{D}$ and knowing either
+that $t_1 = p_1 \andl S = \subst p z {S_i} \andl T = \subst p z {U_i}$
+or that $z \not\in \fn(S_i) \andl z \not\in \fn(U_i) \andl S = S_i \andl
+T = U_i$.
+
+Need to show $t_1.m_i(t_2) \in \rele k \Gamma - T$. Suppose $k > 0$
+and $s \in \rels k \Gamma -$. Suppose, for some $j \leq k$,
+$\reductionl {t_1.m_i(t_2)} s {t'} {s'} j \andl \irred {t'} {s'} \andl
+\Gamma' \in \relg \Gamma k {s'} \Gamma$. Remains to show ${t'} \in
+\relv {k-j-1} {\Gamma'} {s'} T$.
+
+By operational semantics and induction hypotheses, $\reductionl {t_1}
+s {t_1'} {s_1} {j_1}$ and $\irred {t_1'} {s_1}$ and $\reductionl {t_2}
+{s_1} {t_2'} {s_2} {j_2}$ and $\irred {t_2'} {s_2}$ and ${t_1'} \in \relv
+{k-j_1-1} {\Gamma_1} {s_1} {T_1}$ and ${t_2'} \in \relv {k-j_1-j_2-1}
+{\Gamma_2} {s_2} {T_2}$.
+
+By operational semantics, $\reductionl {t_1'.m_i(t_2')} {s_2} {\subst
+  {t_2'} x_i t_i} {s_2} 1$. By substitution lemma, ${\subst {t_2'} x_i
+  t_i} \in \rele {k-j_1-j_2-1} {\Gamma_2} {s_2} T$. So, $\reductionl
+{\subst {t_2'} x_i t_i} {s_2} {t'} {s'} {j_3}$, with $j_1 + j_2 + j_3 +
+1 = j$ implies what remains to be shown.
 

### 2012-08-06 article.tex
all cases
air
cleanup of proof

--- 2012-08-06 article.tex
@@
 
+\begin{myproof}
+The proof is on induction on the derivation of $\Gamma \ts t : T$. For
+each case, we need to show $t \in \rele k \Gamma - T, \forall k$. The
+non-trivial case is to need to show ${t'} \in \relv {k-j-1} {\Gamma'}
+{s'} T$ by supposing $k > 0$, $s \in \rels k \Gamma -$, and, for some
+$j \leq k$, $\reductionl {t} s {t'} {s'} j \andl \irred {t'}
+{s'} \andl \Gamma' \in \relg \Gamma k {s'} \Gamma$.
+
+\ \\
+
+\Case \textsc{var}: $\Gamma \ts x : T$ knowing $(x : T) \in
+\Gamma$. $x \in \relv {k-1} \Gamma s T$ follows from the definition of
+$\mathcal{G}$.
+
+\ \\
+
+\Case \textsc{sel}: $\Gamma \ts t_1.l_i : T$ knowing $\Gamma \ts t_1 :
+T_1$, $\Gamma \ts T_1 \expand_z \seq{D}$, $\ldecl {l_i} {V_i} \in
+\seq{D}$ and knowing either that $t_1 = p_1 \andl T = \subst p z
+    {V_i}$ or that $z \not\in \fn(V_i) \andl T = V_i$.
+
+%Need to show $t_1.l_i \in \rele k \Gamma - T$. Suppose $k > 0$ and $s
+%\in \rels k \Gamma -$. Suppose, for some $j \leq k$, $\reductionl
+%    {t_1.l_i} s {t'} {s'} j \andl \irred {t'} {s'} \andl \Gamma' \in
+%    \relg \Gamma k {s'} \Gamma$. Remains to show ${t'} \in \relv
+%          {k-j-1} {\Gamma'} {s'} T$.
+
+By operational semantics and induction hypothesis, $\reductionl {t_1}
+s {t_1'} {s'} {j-1}$ and $\irred {t_1'} {s'}$ and ${t_1'} \in \relv
+{k-j+1-1} {\Gamma'} {s'} {T_1}$.
+
+By operational semantics, $\reductionl {t_1'.l_i} {s'} {t'} {s'}
+1$. Since ${t_1'} \in \relv {k-j} {\Gamma'} {s'} {T_1}$, we can
+conclude $t' \in \relv {k-j-1} {\Gamma'} {s'} T$ from the clause for
+value labels.
+
+\ \\
+
+\Case \textsc{msel}: $\Gamma \ts t_1.m_i(t_2) : T$ knowing $\Gamma \ts
+t_1 : T_1$, $\Gamma \ts t_2 : T_2$, $\Gamma \ts T_1 \expand_z
+\seq{D}$, $\mdecl {m_i} {S_i} {U_i} \in \seq{D}$ and knowing either
+that $t_1 = p_1 \andl S = \subst p z {S_i} \andl T = \subst p z {U_i}$
+or that $z \not\in \fn(S_i) \andl z \not\in \fn(U_i) \andl S = S_i \andl
+T = U_i$.
+
+%Need to show $t_1.m_i(t_2) \in \rele k \Gamma - T$. Suppose $k > 0$
+%and $s \in \rels k \Gamma -$. Suppose, for some $j \leq k$,
+%$\reductionl {t_1.m_i(t_2)} s {t'} {s'} j \andl \irred {t'} {s'} \andl
+%\Gamma' \in \relg \Gamma k {s'} \Gamma$. Remains to show ${t'} \in
+%\relv {k-j-1} {\Gamma'} {s'} T$.
+
+By operational semantics and induction hypotheses, $\reductionl {t_1}
+s {t_1'} {s_1} {j_1}$ and $\irred {t_1'} {s_1}$ and $\reductionl {t_2}
+{s} {t_2'} {s_2} {j_2}$ and $\irred {t_2'} {s_2}$ and ${t_1'} \in
+\relv {k-j_1-1} {\Gamma_1} {s_1} {T_1}$ and ${t_2'} \in \relv
+{k-j_2-1} {\Gamma_2} {s_2} {T_2}$.
+
+Because $t_2$ reduces to a value $t_2'$ starting in store $s$, it
+should also reduce to a value $v_2$ in the same number of steps
+starting in store $s_1$, since $s_1$ includes $s$. So let $\reductionl
+{t_2} {s_1} {v_2} {s_{12}} {j_2}$ with $v_2 \in \relv {k-j_2-1}
+{\Gamma_{12}} s_{12}$.
+
+By operational semantics, $\reductionl {t_1'.m_i(v_2)} {s_{12}}
+{\subst {v_2} {x_i} {t_i}} {s_{12}} 1$.
+
+It remains to show that ${\subst {v_2} {x_i} {t_i}} \in \rele
+{k-j_2-j_2-1} {\Gamma_{12}} {s_{12}} {T}$ to complete the case. We
+defer this to the substitution lemma.
+
+\ \\
+
+\Case \textsc{new}: $\Gamma \ts \new y c {t_b} : T$ knowing ...
+
+%Need to show $\new y c {t'} \in \rele k \Gamma - T$. Suppose $k > 0$
+%and $s \in \rels k \Gamma -$. Suppose, for some $j \leq k$,
+%$\reductionl {\new y c {t'}} s {t''} {s''} j \andl \irred {t''} {s''} \andl
+%\Gamma'' \in \relg \Gamma k {s''} \Gamma$. Remains to show ${t''} \in
+%\relv {k-j-1} {\Gamma''} {s''} T$.
+
+By operational semantics, $\reductionl {\new y c {t_b}} s {t_b} {s_b} 1$
+where $s_b = s \envplus{y \mapsto c}$. So $\reductionl {t_b} {s_b} {t'}
+{s'} {j-1}$.
+
+By induction hypotheses, $y \in \relv k {\Gamma_b} {s_b} {T_c}$ and $t_b \in
+\rele e k {\Gamma_b} {s_b}$.
+
+Result follows by monotonicity of $\mathcal{V}$.
+
+\end{myproof}
+
+\subsection{Proof of the Substitution Lemma}
+
+\todo{statement and proof}
+
 \section{Discussion}\label{discussion}

### 2012-08-06 article.tex
minor rewording

@@
 
-By operational semantics, $\reductionl {t_1'.l_i} {s'} {t'} {s'}
-1$. Since ${t_1'} \in \relv {k-j} {\Gamma'} {s'} {T_1}$, we can
-conclude $t' \in \relv {k-j-1} {\Gamma'} {s'} T$ from the clause for
-value labels.
+By operational semantics and the above, $\reductionl {t_1'.l_i} {s'}
+{t'} {s'} 1$, and we can conclude $t' \in \relv {k-j-1} {\Gamma'} {s'}
+T$ from the clause for value labels of ${t_1'} \in \relv {k-j}
+{\Gamma'} {s'} {T_1}$.
 
@@
 
-By operational semantics, $\reductionl {t_1'.m_i(v_2)} {s_{12}}
-{\subst {v_2} {x_i} {t_i}} {s_{12}} 1$.
+By the above and operational semantics, $\reductionl {t_1'.m_i(v_2)}
+{s_{12}} {\subst {v_2} {x_i} {t_i}} {s_{12}} 1$.
 

### 2012-08-06 article.tex
a little more

@@
 or that $z \not\in \fn(S_i) \andl z \not\in \fn(U_i) \andl S = S_i \andl
-T = U_i$.
+T = U_i$, and knowing that $\Gamma \ts T_2 \sub S$.
 
@@
 
-It remains to show that ${\subst {v_2} {x_i} {t_i}} \in \rele
-{k-j_2-j_2-1} {\Gamma_{12}} {s_{12}} {T}$ to complete the case. We
-defer this to the substitution lemma.
+By the substitution lemma, ${\subst {v_2} {x_i} {t_i}} \in \rele
+{k-\max(j_1,j_2)-1} {\Gamma_{12}} {s_{12}} {T}$. Supposing,
+$\reductionl {\subst {v_2} {x_i} {t_i}} {s_{12}} {t'} {s'} {j_3}$,
+with $j_1 + j_2 + j_3 + 1 = k$, this completes the case by
+monotonicity of $\mathcal{V}$.
 

### 2012-08-06 article.tex
fix typo
statement of subst lemma

@@
 {t_2} {s_1} {v_2} {s_{12}} {j_2}$ with $v_2 \in \relv {k-j_2-1}
-{\Gamma_{12}} s_{12}$.
+{\Gamma_{12}} {s_{12}} {T_2}$.
 
@@
 By the substitution lemma, ${\subst {v_2} {x_i} {t_i}} \in \rele
-{k-\max(j_1,j_2)-1} {\Gamma_{12}} {s_{12}} {T}$. Supposing,
+{k-\max(j_1,j_2)-2} {\Gamma_{12}} {s_{12}} {T}$. Supposing,
 $\reductionl {\subst {v_2} {x_i} {t_i}} {s_{12}} {t'} {s'} {j_3}$,
-with $j_1 + j_2 + j_3 + 1 = k$, this completes the case by
+with $j_1 + j_2 + j_3 + 1 = j$, this completes the case, eventually by
 monotonicity of $\mathcal{V}$.
@@
 
-\todo{statement and proof}
+The substitution lemma states that if $v_2 \in \relv {k_2}
+{\Gamma_{12}} {s_{12}} {T_2}$ and $t \in \rele {k_1} {\Gamma_1
+  \envplus{{x_i} : S}} {s_1} T$ and $\Gamma \ts {T_2} \sub S$ with
+$\Gamma$ included in $\Gamma_1$ and $\Gamma_{12}$ included in
+$\Gamma_1$ and $s_{12}$ included in $s_1$, then $\subst {v_2} {x_i}
+{t_i} \in \rele {\min(k_1,k_2)-1} {\Gamma_{12}} {s_{12}} T$.
+
+\todo{proof}
 

### 2012-08-06 article.tex
really rough sketch of the subst lemma

--- 2012-08-06 article.tex
@@
 $\reductionl {\subst {v_2} {x_i} {t_i}} {s_{12}} {t'} {s'} {j_3}$,
-with $j_1 + j_2 + j_3 + 1 = k$, this completes the case by
+with $j_1 + j_2 + j_3 + 1 = j$, this completes the case, eventually by
 monotonicity of $\mathcal{V}$.
@@
 
-\todo{statement and proof}
+The substitution lemma states that if (1) $v_2 \in \relv {k_2}
+{\Gamma_{12}} {s_{12}} {T_2}$ and (2) $t \in \rele {k_1} {\Gamma_1
+  \envplus{{x_i} : S}} {s_1} T$ and (3) $\Gamma \ts {T_2} \sub S$ with
+$\Gamma$ included in $\Gamma_1$ and $\Gamma_{12}$ included in
+$\Gamma_1$ and $s_{12}$ included in $s_1$, then $\subst {v_2} {x_i}
+{t_i} \in \rele {\min(k_1,k_2)} {\Gamma_{12}} {s_{12}} T$.
+
+\begin{proofsketch}
+By (1) and (3), it should hold that (4) $v_2 \in \relv {k_2}
+{\Gamma_{12}} {s_{12}} {S}$. Since (2) holds, it should also hold that
+${t_i} \in \rele {\min(k_1,k_2)} {\Gamma_{12} \envplus{{x_i} : S}}
+{s_{12}} T$ by monotonicity of the logical relation on all
+subscripts. Then, we can instantiate $x_i$ in the complete store to
+map to what $v_2$ maps to. This should be fine by (4) and
+monotonicity. Thus, ${t_i} \in \rele {\min(k_1,k_2)} {\Gamma_{12}
+  \envplus{{x_i} : S}} {s_{12} \envplus{{x_i} \mapsto s_{12}(v)}}
+T$. This implies what needs to be shown.
+\end{proofsketch}
 

### 2012-08-06 article.tex
minor rev

@@
 
-$\Gamma \ds t : T$ is simplfy defined as $t \in \rele k \Gamma - T,
+$\Gamma \ds t : T$ is simply defined as $t \in \rele k \Gamma - T,
 \forall k$.
@@
 $\reductionl {\subst {v_2} {x_i} {t_i}} {s_{12}} {t'} {s'} {j_3}$,
-with $j_1 + j_2 + j_3 + 1 = j$, this completes the case, eventually by
+with $j_1 + j_2 + j_3 + 1 = j$, this completes the case, by
 monotonicity of $\mathcal{V}$.
@@
 By induction hypotheses, $y \in \relv k {\Gamma_b} {s_b} {T_c}$ and $t_b \in
-\rele e k {\Gamma_b} {s_b}$.
+\rele k {\Gamma_b} {s_b} {T}$.
 

### 2012-08-06 article.tex
grammar

@@
 application was not well-typed. But if we started preservation in a store
-which has $a$, $b$, $c$ and $d$ then the application type-checked,
+which had $a$, $b$, $c$ and $d$ then the application type-checked,
 because, through one of the path-equality provision, we could find that

### 2012-08-06 article.tex
a bit more details on subst lemma

@@
   \envplus{{x_i} : S}} {s_1} T$ and (3) $\Gamma \ts {T_2} \sub S$ with
-$\Gamma$ included in $\Gamma_1$ and $\Gamma_{12}$ included in
-$\Gamma_1$ and $s_{12}$ included in $s_1$, then $\subst {v_2} {x_i}
-{t_i} \in \rele {\min(k_1,k_2)} {\Gamma_{12}} {s_{12}} T$.
+(4) $x_i \not\in \fn(T)$ and $\Gamma$ included in $\Gamma_1$ and
+$\Gamma_{12}$ included in $\Gamma_1$ and $s_{12}$ included in $s_1$,
+then $\subst {v_2} {x_i} {t_i} \in \rele {\min(k_1,k_2)} {\Gamma_{12}}
+{s_{12}} T$.
 
 \begin{proofsketch}
-By (1) and (3), it should hold that (4) $v_2 \in \relv {k_2}
-{\Gamma_{12}} {s_{12}} {S}$. Since (2) holds, it should also hold that
-${t_i} \in \rele {\min(k_1,k_2)} {\Gamma_{12} \envplus{{x_i} : S}}
-{s_{12}} T$ by monotonicity of the logical relation on all
+By (1) and (3), it should hold that (5) $v_2 \in \relv {k_2}
+{\Gamma_{12}} {s_{12}} {S}$.  Since (2) holds, it should also hold
+that ${t_i} \in \rele {\min(k_1,k_2)} {\Gamma_{12} \envplus{{x_i} :
+    S}} {s_{12}} T$ by monotonicity of the logical relation on all
 subscripts. Then, we can instantiate $x_i$ in the complete store to
-map to what $v_2$ maps to. This should be fine by (4) and
+map to what $v_2$ maps to. This should be fine by (5) and
 monotonicity. Thus, ${t_i} \in \rele {\min(k_1,k_2)} {\Gamma_{12}
   \envplus{{x_i} : S}} {s_{12} \envplus{{x_i} \mapsto s_{12}(v)}}
-T$. This implies what needs to be shown.
+T$. Thanks to (4), we don't actually need $x_i$ to be held abstract in
+the environment, because it won't occur in $T$ or its expansion, so
+we can use the actual type constructor of $v_2$, say $T_{v_2}$, instead
+of $S$ for $x_i$: ${t_i} \in \rele {\min(k_1,k_2)} {\Gamma_{12}
+  \envplus{{x_i} : {T_{v_2}}}} {s_{12} \envplus{{x_i} \mapsto
+      s_{12}(v)}} T$. This implies what needs to be shown.
+
 \end{proofsketch}

### 2012-08-07 article.tex
fix
fix tsel-wf1 rule
typo

@@
 There are four alphabets: Variable names $x$, $y$, $z$ are freely
-alpha-renamable. They occur as parameters of lambda abstractions, as
+alpha-renamable. They occur as parameters of methods, as
 binders for objects created by \verb@new@-expressions, and as self
@@
       {\Gamma \ts p \ni \Ldecl L S U \spcomma S \wf \spcomma U \wf}
-      {\Gamma \ts p.L \wf \spcomma p.U \wf}
+      {\Gamma \ts p.L \wf}
 
@@
 that agrees with a store $s$ by simply copying the constructor type
-from the store for each missig variable.
+from the store for each missing variable.
 \begin{align*}

### 2012-08-08 article.tex
paths p not v
- to \emptyset for clarity

@@
 Two subclasses of terms are values $v$, which consist of just
-variables, and paths $v$ which consist of just variables and field
+variables, and paths $p$ which consist of just variables and field
 selections.
@@
 Type-safety states that a well-typed program doesn't get stuck. More
-formally: If $- \ts t \typ T$ and $\reductionl t - {t'} {s'} *$ then
+formally: If $\emptyset \ts t \typ T$ and $\reductionl t \emptyset {t'} {s'} *$ then
 either $t'$ is a value or $\exists t'', s'', \reduction {t'} {s'}
@@
 
-$\Gamma \ds t : T$ is simply defined as $t \in \rele k \Gamma - T,
+$\Gamma \ds t : T$ is simply defined as $t \in \rele k \Gamma \emptyset T,
 \forall k$.
@@
 The proof is on induction on the derivation of $\Gamma \ts t : T$. For
-each case, we need to show $t \in \rele k \Gamma - T, \forall k$. The
+each case, we need to show $t \in \rele k \Gamma \emptyset T, \forall k$. The
 non-trivial case is to need to show ${t'} \in \relv {k-j-1} {\Gamma'}
-{s'} T$ by supposing $k > 0$, $s \in \rels k \Gamma -$, and, for some
+{s'} T$ by supposing $k > 0$, $s \in \rels k \Gamma \emptyset$, and, for some
 $j \leq k$, $\reductionl {t} s {t'} {s'} j \andl \irred {t'}
@@
 
-%Need to show $t_1.l_i \in \rele k \Gamma - T$. Suppose $k > 0$ and $s
-%\in \rels k \Gamma -$. Suppose, for some $j \leq k$, $\reductionl
+%Need to show $t_1.l_i \in \rele k \Gamma \emptyset T$. Suppose $k > 0$ and $s
+%\in \rels k \Gamma \emptyset$. Suppose, for some $j \leq k$, $\reductionl
 %    {t_1.l_i} s {t'} {s'} j \andl \irred {t'} {s'} \andl \Gamma' \in
@@
 
-%Need to show $t_1.m_i(t_2) \in \rele k \Gamma - T$. Suppose $k > 0$
-%and $s \in \rels k \Gamma -$. Suppose, for some $j \leq k$,
+%Need to show $t_1.m_i(t_2) \in \rele k \Gamma \emptyset T$. Suppose $k > 0$
+%and $s \in \rels k \Gamma \emptyset$. Suppose, for some $j \leq k$,
 %$\reductionl {t_1.m_i(t_2)} s {t'} {s'} j \andl \irred {t'} {s'} \andl
@@
 
-%Need to show $\new y c {t'} \in \rele k \Gamma - T$. Suppose $k > 0$
-%and $s \in \rels k \Gamma -$. Suppose, for some $j \leq k$,
+%Need to show $\new y c {t'} \in \rele k \Gamma \emptyset T$. Suppose $k > 0$
+%and $s \in \rels k \Gamma \emptyset$. Suppose, for some $j \leq k$,
 %$\reductionl {\new y c {t'}} s {t''} {s''} j \andl \irred {t''} {s''} \andl

### 2012-08-08 article.tex
reorg
proof

@@
 \newcommand{\rele}[4]{\mathcal{E}_{#1;#2;#3}\llbracket#4\rrbracket}
-\newcommand{\rels}[3]{\mathcal{S}_{#1;#2}\llbracket#3\rrbracket}
-\newcommand{\relg}[3]{\mathcal{G}_{#1;#2}\llbracket#3\rrbracket}
+\newcommand{\rels}[3]{\mathcal{\supseteq}_{#1}\llbracket#2;#3\rrbracket}
+\newcommand{\relg}[3]{\mathcal{\supseteq}_{#1;#2}\llbracket!#3\rrbracket}
 \newcommand{\irred}[2]{\text{irred }(#1,#2)}
@@
 
-\subsubsection{$\mathcal{V}$}
+\subsubsection{Set of Values}
 $\relv k \Gamma s T$ defines the set of values that appear to have
 type $T$ when taking at most $k$ steps. $\Gamma$ and $s$ must agree:
-$\dom(\Gamma) = \dom(s)$ and $\forall (x : T) \in \Gamma, x \in \relv
-k \Gamma s T$. A variable $y$ belongs to $\relv 0 \Gamma s T$ simply
-by being in the store. In addition, it belongs to $\relv k \Gamma s T$
-for $k > 0$, if it defines all method and value labels in the
-expansion of $T$ appropriately for $j < k$ steps.
+$\dom(\Gamma) = \dom(s)$ (ordered) and $\forall (x : T) \in \Gamma, x
+\in \relv k \Gamma s T$. A variable $y$ belongs to $\relv 0 \Gamma s
+T$ simply by being in the store. In addition, it belongs to $\relv k
+\Gamma s T$ for $k > 0$, if it defines all method and value labels in
+the expansion of $T$ appropriately for $j < k$ steps.
 \begin{align*}
@@
 
-\subsubsection{$\mathcal{E}$}
+\subsubsection{Set of Terms}
 $\rele k \Gamma s T$ defines the set of terms that appear to have type
-$T$ when taking at most $k$ steps. $s$ must agree with $\Gamma$, but
-$\Gamma$ can additionaly contain variables not in $s$. This is needed
-for checking methods in $\mathcal{V}$ above, and for relating open
-terms. If $k > 0$, $\mathcal{E}$ completes the store to agree with
-$\Gamma$. It then states that if it can reduce $t$ in the completed
-store to an irreducible term in $j < k$ steps, then this term must be
-in a corresponding $\mathcal{V}$ set.
+$T$ when taking at most $k$ steps. $s$ must agree with a {\it prefix}
+of $\Gamma$, so $\Gamma$ can additionaly contain variables not in
+$s$. This is needed for checking methods in $\mathcal{V}$ above, and
+for relating open terms. If $k > 0$, $\mathcal{E}$ extends $\Gamma$
+and $s$ so that they agree. It then states that if it can reduce $t$
+in the extended store to an irreducible term in $j < k$ steps, then
+this term must be in a corresponding $\mathcal{V}$ set.
 \begin{align*}
@@
 &\gap k = 0 \orl (\forall j \leq k,\\
-&\gap\gap \forall s' \in \rels k \Gamma s ,\\
+&\gap\gap \forall (\Gamma'; s') \in \rels k \Gamma s ,\\
 &\gap\gap \reductionl t {s'} {t'} {s''} j \andl\\
 &\gap\gap \irred {t'} {s''} \impliesl\\
-&\gap\gap \forall \Gamma' \in \relg k {s''} \Gamma ,\\
-&\gap\gap {t'} \in \relv {k-j-1} {\Gamma'} {s''} T)\\
+&\gap\gap \forall \Gamma'' \in \relg k {s''} {\Gamma'} ,\\
+&\gap\gap {t'} \in \relv {k-j-1} {\Gamma''} {s''} T)\\
 &\}
@@
 
-\subsubsection{$\mathcal{S}$}
-$\rels k \Gamma s$ for $k > 0$ defines the set of completed stores
-that agree with $\Gamma$ on each variable when taking at most $k-1$
-steps.
-\begin{align*}
-&\rels k \Gamma s = \{ s, \seq{x_i \mapsto c_i}^{m \leq i < n} \ |\\
-&\gap s = \seq{x \mapsto c}^m \andl \Gamma = \seq{x : T}^n \andl \\
-&\gap m\leq n \andl \forall i, m \leq i < n,\\
-&\gap c_i \in \relv {k-1} \Gamma {s, \seq{x_j \mapsto c_j}^{m \leq j < i}} {T_i}\\
+\subsubsection{Extending the environment and the store}
+$\rels k \Gamma s$ for $k > 0$ defines the set of completed
+environment and stores that agree on $k-1$ steps, and that extend
+$\Gamma$ and $s$. $s$ must agree with a {\it prefix} of $\Gamma$.
+Both $\Gamma$ and $s$ are ordered maps. For $s$, $s'$ extends $s$ if
+$s$ is a prefix of $s'$. For $\Gamma$, $\Gamma'$ extends $\Gamma$ if
+we get back $\Gamma$ by keeping only the elements of $\Gamma'$ that
+belong to $\Gamma$. Furthermore, a prefix of $\Gamma'$ agrees with
+$s$.
+\begin{align*}
+&\rels k \Gamma s = \{\\
+&\ (\seq{x : T}^m, \seq{x_{ij} : T_{ij}}^{m \leq i < n; 0 \leq j < i_n}; s, \seq{x_{ij} \mapsto c_{ij}}^{m \leq i < n; 0 \leq j < i_n}) |\\
+&\ s = \seq{x \mapsto c}^m \andl \Gamma = \seq{x : T}^n \andl \\
+&\ m\leq n \andl \forall i, m \leq i < n, \forall {i_n}, j, 0 \leq j < i_n,\\
+&\ \forall T_{ij}, c_{ij}, T_{i{(i_n-1)}} = T_i, \forall n' \leq n, i_n' \leq i_n,\\
+&\ c_{ij} \in \relv {k-1} {\seq{x : T}^m, \seq{x_{ij} : T_{ij}}^{m \leq i' < n; 0 \leq j < i_n'}} {s, \seq{x_{ij} \mapsto c_{ij}}^{m \leq i < n'; 0 \leq j < i_n'}} {T_{ij}}\\
 &\}
@@
 
-\subsubsection{$\mathcal{G}$}
+\subsubsection{Completing the environment to agree with the store}
 $\relg k s \Gamma$ defines a singleton set of a completed environment
@@
 
-\subsection{Proof of the Fundamental Theorem}
+\subsection{Statements and Proofs}
+
+\subsubsection{Fundamental Theorem}
 
@@
 each case, we need to show $t \in \rele k \Gamma \emptyset T, \forall k$. The
-non-trivial case is to need to show ${t'} \in \relv {k-j-1} {\Gamma'}
-{s'} T$ by supposing $k > 0$, $s \in \rels k \Gamma \emptyset$, and, for some
+non-trivial case is to need to show ${t'} \in \relv {k-j-1} {\Gamma''}
+{s'} T$ by supposing $k > 0$, $(\Gamma'; s) \in \rels k \Gamma \emptyset$, and, for some
 $j \leq k$, $\reductionl {t} s {t'} {s'} j \andl \irred {t'}
-{s'} \andl \Gamma' \in \relg \Gamma k {s'} \Gamma$.
+{s'} \andl {\Gamma''} \in \relg \Gamma k {s'} {\Gamma'}$.
 
@@
 \Case \textsc{var}: $\Gamma \ts x : T$ knowing $(x : T) \in
-\Gamma$. $x \in \relv {k-1} \Gamma s T$ follows from the definition of
-$\mathcal{G}$.
+\Gamma$. $x \in \relv {k-1} {\Gamma'} s T$ follows from the definition of
+$\rels k \Gamma \emptyset$.
 
@@
 
-%Need to show $t_1.l_i \in \rele k \Gamma \emptyset T$. Suppose $k > 0$ and $s
-%\in \rels k \Gamma \emptyset$. Suppose, for some $j \leq k$, $\reductionl
-%    {t_1.l_i} s {t'} {s'} j \andl \irred {t'} {s'} \andl \Gamma' \in
-%    \relg \Gamma k {s'} \Gamma$. Remains to show ${t'} \in \relv
-%          {k-j-1} {\Gamma'} {s'} T$.
-
 By operational semantics and induction hypothesis, $\reductionl {t_1}
@@
 By operational semantics and the above, $\reductionl {t_1'.l_i} {s'}
-{t'} {s'} 1$, and we can conclude $t' \in \relv {k-j-1} {\Gamma'} {s'}
+{t'} {s'} 1$, and we can conclude $t' \in \relv {k-j-1} {\Gamma''} {s'}
 T$ from the clause for value labels of ${t_1'} \in \relv {k-j}
-{\Gamma'} {s'} {T_1}$.
+{\Gamma''} {s'} {T_1}$.
 
@@
 
-%Need to show $t_1.m_i(t_2) \in \rele k \Gamma \emptyset T$. Suppose $k > 0$
-%and $s \in \rels k \Gamma \emptyset$. Suppose, for some $j \leq k$,
-%$\reductionl {t_1.m_i(t_2)} s {t'} {s'} j \andl \irred {t'} {s'} \andl
-%\Gamma' \in \relg \Gamma k {s'} \Gamma$. Remains to show ${t'} \in
-%\relv {k-j-1} {\Gamma'} {s'} T$.
-
 By operational semantics and induction hypotheses, $\reductionl {t_1}
@@
 should also reduce to a value $v_2$ in the same number of steps
-starting in store $s_1$, since $s_1$ includes $s$. So let $\reductionl
+starting in store $s_1$, since $s_1$ extends $s$. So let $\reductionl
 {t_2} {s_1} {v_2} {s_{12}} {j_2}$ with $v_2 \in \relv {k-j_2-1}
@@
 
-%Need to show $\new y c {t'} \in \rele k \Gamma \emptyset T$. Suppose $k > 0$
-%and $s \in \rels k \Gamma \emptyset$. Suppose, for some $j \leq k$,
-%$\reductionl {\new y c {t'}} s {t''} {s''} j \andl \irred {t''} {s''} \andl
-%\Gamma'' \in \relg \Gamma k {s''} \Gamma$. Remains to show ${t''} \in
-%\relv {k-j-1} {\Gamma''} {s''} T$.
-
 By operational semantics, $\reductionl {\new y c {t_b}} s {t_b} {s_b} 1$
@@
 
-\subsection{Proof of the Substitution Lemma}
-
-The substitution lemma states that if (1) $v_2 \in \relv {k_2}
+\subsubsection{Substitution Lemma}
+
+The substitution lemma states that if (1) $v \in \relv {k_2}
 {\Gamma_{12}} {s_{12}} {T_2}$ and (2) $t \in \rele {k_1} {\Gamma_1
-  \envplus{{x_i} : S}} {s_1} T$ and (3) $\Gamma \ts {T_2} \sub S$ with
-(4) $x_i \not\in \fn(T)$ and $\Gamma$ included in $\Gamma_1$ and
-$\Gamma_{12}$ included in $\Gamma_1$ and $s_{12}$ included in $s_1$,
-then $\subst {v_2} {x_i} {t_i} \in \rele {\min(k_1,k_2)} {\Gamma_{12}}
-{s_{12}} T$.
+  \envplus{{x} : S}} {s_1} T$ and (3) $\Gamma \ts {T_2} \sub S$ with
+(4) $x \not\in \fn(T)$ and $\Gamma_1$ extends $\Gamma$ and
+$\Gamma_{12}$ included in $\Gamma_1$ and $s_{12}$ extends $s_1$ and
+$\Gamma_1$ agrees with $s_1$ and $\Gamma_{12}$ agrees with $s_{12}$
+and a prefix of $\Gamma_{12}$ agrees with $s_1$, then $\subst {v} {x}
+{t} \in \rele {\min(k_1,k_2)} {\Gamma_{12}} {s_{12}} T$.
 
 \begin{proofsketch}
-By (1) and (3), it should hold that (5) $v_2 \in \relv {k_2}
-{\Gamma_{12}} {s_{12}} {S}$.  Since (2) holds, it should also hold
-that ${t_i} \in \rele {\min(k_1,k_2)} {\Gamma_{12} \envplus{{x_i} :
-    S}} {s_{12}} T$ by monotonicity of the logical relation on all
-subscripts. Then, we can instantiate $x_i$ in the complete store to
-map to what $v_2$ maps to. This should be fine by (5) and
-monotonicity. Thus, ${t_i} \in \rele {\min(k_1,k_2)} {\Gamma_{12}
-  \envplus{{x_i} : S}} {s_{12} \envplus{{x_i} \mapsto s_{12}(v)}}
-T$. Thanks to (4), we don't actually need $x_i$ to be held abstract in
-the environment, because it won't occur in $T$ or its expansion, so
-we can use the actual type constructor of $v_2$, say $T_{v_2}$, instead
-of $S$ for $x_i$: ${t_i} \in \rele {\min(k_1,k_2)} {\Gamma_{12}
-  \envplus{{x_i} : {T_{v_2}}}} {s_{12} \envplus{{x_i} \mapsto
-      s_{12}(v)}} T$. This implies what needs to be shown.
-
+By (1) and (3), it should hold that (5) $v \in \relv {k_2}
+{\Gamma_{12}} {s_{12}} {S}$ by the subset semantics lemma.  Since (2)
+holds, it should also hold that ${t} \in \rele {\min(k_1,k_2)}
+{\Gamma_{12} \envplus{{x} : S}} {s_{12}} T$ by the extended
+monotonicity lemma. Then, we can instantiate $x$ in the complete store
+to map to what $v$ maps to. This should be fine by (5) and
+monotonicity. Thus, ${t} \in \rele {\min(k_1,k_2)} {\Gamma_{12}
+  \envplus{{x} : S}} {s_{12} \envplus{{x} \mapsto s_{12}(v)}}
+T$. Thanks to (4), we don't actually need $x$ to be held abstract in
+the environment, because it won't occur in $T$ or its expansion, so we
+can use the actual type constructor of $v$, say $T_{v}$, instead of
+$S$ for $x$: ${t} \in \rele {\min(k_1,k_2)} {\Gamma_{12} \envplus{{x}
+    : {T_{v_2}}}} {s_{12} \envplus{{x} \mapsto s_{12}(v)}} T$. This
+implies what needs to be shown.
+
+\end{proofsketch}
+
+\subsubsection{Subset Semantics Lemma}
+
+The subset semantics lemma states that if (1) $v \in \relv k \Gamma s S$
+and (2) $\Gamma \ts S <: U$, then $v \in \relv k \Gamma s U$.
+
+\begin{proofsketch}
+Because $S$ is a subtype of $U$, it should hold that the expansion of
+$S$ subsumes the expansion of $U$. Therefore, for $v \in \relv k
+\Gamma s U$, we have fewer and ``simpler'' definitions to check than
+for $v \in \relv k \Gamma s S$. One worry is that for methods, because
+the parameter type is contravariant, $U$ could require a more precise
+type than $S$. This shouldn't lead to narrowing issues, because if the
+parameter type becomes ill-typed, then the check is vacuously true.
+\end{proofsketch}
+
+\subsubsection{Extended Monotonicity Lemma}
+
+The extended monotonicity lemma states that if $t \in \rele k {\Gamma
+  \envplus{x : S}} s T$ then $t \in \rele j {\Gamma' \envplus{x : S}}
+{s'} T$ for $j \leq k$, $\Gamma'$ extends $\Gamma$, $s'$
+extends $s$, and $\Gamma$ agrees with $s$ and a prefix of $\Gamma'$
+agrees with $s$.
+
+\begin{proofsketch}
+For the monotonicty with regards to the step-index, this follows
+directly from the definitions of $\mathcal{E}$ and $\mathcal{V}$. For the
+environment and the store, this follows from the definition of $\rels
+k {\Gamma \envplus{x : S}} s$. To extend the environment and the store
+for $x : S$, we can append as much as we want to $\Gamma$ and $s$, to
+get $\Gamma'$ and $s'$, and then ignore the last element which is for
+$x : S$.
 \end{proofsketch}

### 2012-08-08 article.tex
todos

@@
 experience suggests that this complicates the proofs considerably.  An
-alternative approach to inheritance that might work better is to model
+alternative approach that might work better is to model inheritance
 it as a form of code-reuse. Starting with an enriched type system with
@@
 to prove type safety wrt reduction.
-
-\remark{Does this last point still makes sense wrt to type safety via
-  logical relations instead of via the standard theorems of
-  preservation and progress?}
 
@@
 
-\todo{more discussion subsections?}
-
 \section{Conclusion}\label{conclusion}
-\todo{conclusion}
 
@@
 \acks
-\todo{acks}
+
+We thank Amal Ahmed for her many discussions and insights about
+applying logical relations to DOT. We thank Donna Malayeri and
+Geoffrey Washburn for preliminary work on DOT.
 

### 2012-08-08 article.tex
spellcheck

@@
 Scala and languages like it. The properties we are interested in
-modelling are Scala's path-dependent types and abstract type members,
+modeling are Scala's path-dependent types and abstract type members,
 as well as its mixture of nominal and structural typing through the
@@
 we make no attempt to model inheritance or mixin composition. Indeed
-we will argue that such concepts are better modelled in a different
+we will argue that such concepts are better modeled in a different
 setting.
@@
 concerns the difference between Scala's compound type formation using
-{\bf with} and classical type intersection, as it is modelled in the
+{\bf with} and classical type intersection, as it is modeled in the
 calculus. Scala, and the previous calculi attempting to model it
@@
 Sets of declarations form a lattice with the given meet $\wedge$ and
-join $\vee$, the empty set of declatations as the top element, and the
+join $\vee$, the empty set of declarations as the top element, and the
 bottom element $\seq{D_\Bot}$, Here $\seq{D_\Bot}$ is the set of
@@
 $T$ when taking at most $k$ steps. $s$ must agree with a {\it prefix}
-of $\Gamma$, so $\Gamma$ can additionaly contain variables not in
+of $\Gamma$, so $\Gamma$ can additionally contain variables not in
 $s$. This is needed for checking methods in $\mathcal{V}$ above, and
@@
 \begin{proofsketch}
-For the monotonicty with regards to the step-index, this follows
+For the monotonicity with regards to the step-index, this follows
 directly from the definitions of $\mathcal{E}$ and $\mathcal{V}$. For the
@@
 all the $\wfe$ preconditions in subtyping were dropped. In this
-broader calculus, subtyping transivitiy doesn't hold, because of the
+broader calculus, subtyping transitivity doesn't hold, because of the
 rule (\textsc{$\sub$-rfn}) which requires expansion of the left type.
@@
 its upper bound. So it is straightforward to turn this counterexample
-to subtyping transivity into a counterexample to preservation:
+to subtyping transitivity into a counterexample to preservation:
 \begin{align*}

### 2012-08-08 article.tex
refactored new rule
typo

@@
 \newcommand{\spcomma}{~,~}
+\newcommand{\sat}{{\bf\;sat\;}}
 
@@
 \gap p.l & \mbox{selection} \\
-c ::= T_c \ldefs{\seq{l = v}\;\seq{m(x) = t}} & \lindent{\mbox{Constructor}} \\[0.2em]
-\Gamma ::= \seq{x \typ T} & \lindent\mbox{Environment} \\
-s      ::= \seq{x \mapsto c} & \lindent\mbox{Store} \\
+c ::= T_c \ldefs{\seq{d}} & \lindent{\mbox{Constructor}} \\[0.2em]
+d ::= & \lindent{\mbox{Initialization}}\\
+\gap l = v & \mbox{field initialization}\\
+\gap m(x) = t & \mbox{method initialization}\\
+s      ::= \seq{x \mapsto c} & \lindent\mbox{Store}
 \ea
@@
 \gap \ldecl l T   & \mbox{value declaration} \\
-\gap \mdecl m S U & \mbox{method declaration}
+\gap \mdecl m S U & \mbox{method declaration}\\
+\ \\
+\Gamma ::= \seq{x \typ T} & \lindent\mbox{Environment}
 \ea
@@
 {
-c = {T_c \ldefs{\seq{l = v}\;\seq{m(x) = t}}}\\
 y \notin \fn(T') \\
-\Gamma \ts T_c \wf \\
-\Gamma \ts T_c \expand_y \seq{\Ldecl L S U},\seq{\ldecl l V},\seq{\mdecl m T W} \\
-\Gamma \envplus{y: T_c} \ts \seq{S \sub U}\\
-\Gamma \envplus{y: T_c} \ts \seq{v \typ V'} \spcomma \seq{V' \sub V}\\
-\Gamma \envplus{y: T_c} \ts \seq{{T_i} \wfe}\\
-\seq{\Gamma \envplus{y: T_c} \envplus{x_i: T_i} \ts t_i \typ {W_i}' \spcomma {W_i}' \sub {W_i}}\\
-\Gamma \envplus{y: T_c} \ts  {t'} \typ T'}
-{\Gamma \ts \new y c {t'} \typ T'}
+\Gamma \ts T_c \wfe \spcomma T_c \expand_y \seq{D} \\
+\Gamma \envplus{y: T_c} \ts \seq{D \sat \seq{d}} \spcomma {t'} \typ {T'}}
+{\Gamma \ts \new y {T_c \ldefs{\seq{d}}} {t'} \typ T'}
 
 \end{multicols}
+
+\linesep
+
+\begin{multicols}{2}[\judgement{Satisfied Declaration}{\fbox{$\Gamma \ts D \sat \seq{d}$}}]
+
+\infrule[\textsc{tdecl-sat}]
+{\Gamma \ts S \sub U}
+{\Gamma \ts \Ldecl L S U \sat \seq{d}}
+
+\infrule[\textsc{vdecl-sat}]
+{\Gamma \ts v \typ {V'} \spcomma {V'} \sub V}
+{\Gamma \ts \ldecl l V \sat l = v,\seq{d}}
+
+\infrule[\textsc{mdecl-sat}]
+{\Gamma \ts S \wfe\\
+ \Gamma \envplus{x : S} \ts t \typ {T'} \spcomma {T'} \sub T}
+{\Gamma \ts \mdecl m S T \sat m(x) = t,\seq{d}}
+
+\end{multicols}
+
 } % END dot-one
@@
 
+We have presented DOT, a calculus aimed as a new foundation of Scala
+and languages like it. DOT features path-dependent types, refinement
+types, and abstract type members.
+
+Proving the DOT calculus type-safe has been an interesting
+adventure. We have shown that DOT does not satisfy preservation (also
+known as subject-reduction). However, the standard theorems of
+preservation and progress are just one way to prove type-safety. We
+have sketched a plausible proof of type safety by using the powerful
+method of logical relations.
+
 %\appendix
@@
 
-We thank Amal Ahmed for her many discussions and insights about
-applying logical relations to DOT. We thank Donna Malayeri and
-Geoffrey Washburn for preliminary work on DOT.
+We thank Amal Ahmed for many discussions and insights about applying
+logical relations to DOT. We thank Donna Malayeri and Geoffrey
+Washburn for preliminary work on DOT.
 

### 2012-08-08 article.tex
typo
rev

@@
 known as subject-reduction). However, the standard theorems of
-preservation and progress are just one way to prove type-safety. We
-have sketched a plausible proof of type safety by using the powerful
+preservation and progress are just one way to prove type safety. We
+have sketched a plausible proof of type safety using the powerful
 method of logical relations.

### 2012-08-08 article.tex
root.Nat  instead of Nat

@@
 &\mlnew {\mi{genLists}} {\Top \mlrefine g {\Ldecl {\mi{ListPackage}} \Bot {\Top \mlrefine p {\\\displaybreak[3]
-&\ \Ldecl {\mi{Elem}} \Bot \Top\\
-&\ \Ldecl {\mi{ListOfElement}} {g.\mi{List} \refine z {\Ldecl {\mi{Elem}} {p.\mi{Elem}} {p.\mi{Elem}}}} {\\&\gap\gap\gap\gap g.\mi{List} \refine z {\Ldecl {\mi{Elem}} {p.\mi{Elem}} {p.\mi{Elem}}}}\\
+&\ \Ldecl {\mi{Elem}} \Bot \Top\\\displaybreak[3]
+&\ \Ldecl {\mi{ListOfElement}} {g.\mi{List} \refine z {\Ldecl {\mi{Elem}} {p.\mi{Elem}} {p.\mi{Elem}}}} {\\&\gap\gap\gap\gap g.\mi{List} \refine z {\Ldecl {\mi{Elem}} {p.\mi{Elem}} {p.\mi{Elem}}}}\\\displaybreak[3]
 &\ \Ldecl {\mi{List}} \Bot {\Top \mlrefine z {\\\displaybreak[3]
-&\ \gap \mdecl {\mi{isEmpty}} {\mi{root}.\mi{Unit}} {\mi{root}.\mi{Boolean}}\\
-&\ \gap \mdecl {\mi{head}} {\mi{root}.\mi{Unit}} {g.\mi{Elem}}\\
-&\ \gap \mdecl {\mi{tail}} {\mi{root}.\mi{Unit}} {g.\mi{ListOfElem}}\\
-&\gap}}\\
-&\ \mdecl {\mi{nil}} {\mi{root}.\mi{Unit}} {g.\mi{ListOfElem}}\\
-&\ \mdecl {\mi{cons}} {g.\mi{Elem}} {g.\mi{ListOfElem}}\\
+&\ \gap \mdecl {\mi{isEmpty}} {\mi{root}.\mi{Unit}} {\mi{root}.\mi{Boolean}}\\\displaybreak[3]
+&\ \gap \mdecl {\mi{head}} {\mi{root}.\mi{Unit}} {g.\mi{Elem}}\\\displaybreak[3]
+&\ \gap \mdecl {\mi{tail}} {\mi{root}.\mi{Unit}} {g.\mi{ListOfElem}}\\\displaybreak[3]
+&\gap}}\\\displaybreak[3]
+&\ \mdecl {\mi{nil}} {\mi{root}.\mi{Unit}} {g.\mi{ListOfElem}}\\\displaybreak[3]
+&\ \mdecl {\mi{cons}} {g.\mi{Elem}} {g.\mi{ListOfElem}}\\\displaybreak[3]
 &}}}}{
-\mlnew {\mi{natLists}} {{\mi{genLists}.\mi{ListPackage}} \mlrefine p { \Ldecl {\mi{Elem}} {\mi{Nat}} {\mi{Nat}} }\mlldefs{\\\displaybreak[3]
-&\ {\mi{nil}}(u) = \ldots\\
-&\ {\mi{cons}}(u) = \ldots\\
+\mlnew {\mi{natLists}} {{\mi{genLists}.\mi{ListPackage}} \mlrefine p {\\\displaybreak[3]
+&\gap \Ldecl {\mi{Elem}} {\mi{root}.\mi{Nat}} {\mi{root}.\mi{Nat}} }\mlldefs{\\\displaybreak[3]
+&\ {\mi{nil}}(u) = \ldots\\\displaybreak[3]
+&\ {\mi{cons}}(u) = \ldots\\\displaybreak[3]
 &}}}{

### 2012-08-08 article.tex
align
playing with the -1

@@
 Declaration subsumption is extended to a binary relation between
-sequences of declarations as follows: $\seq{D} <: \seq{D'}$ iff
-$\forall D'_i \exists D_j. D_j \sub D'_i$.
+sequences of declarations: $\seq{D} <: \seq{D'}$ iff $\forall D'_i,
+\exists D_j. {D_j \sub D'_i}$.
 
@@
 &\gap \Gamma \ts T \wfe \andl\\
-&\gap k = 0 \orl (\forall j \leq k,\\
+&\gap k = 0 \orl (\forall j < k,\\
 &\gap\gap \forall (\Gamma'; s') \in \rels k \Gamma s ,\\
@@
 {s'} T$ by supposing $k > 0$, $(\Gamma'; s) \in \rels k \Gamma \emptyset$, and, for some
-$j \leq k$, $\reductionl {t} s {t'} {s'} j \andl \irred {t'}
+$j < k$, $\reductionl {t} s {t'} {s'} j \andl \irred {t'}
 {s'} \andl {\Gamma''} \in \relg \Gamma k {s'} {\Gamma'}$.

### 2012-08-08 article.tex
subset sem lemma

@@
 
-The subset semantics lemma states that if (1) $v \in \relv k \Gamma s S$
-and (2) $\Gamma \ts S <: U$, then $v \in \relv k \Gamma s U$.
+The subset semantics lemma states that if $v \in \relv k \Gamma s S$
+and $\Gamma \ts S <: U$, then $v \in \relv k \Gamma s U$.
 
@@
 \Gamma s U$, we have fewer and ``simpler'' definitions to check than
-for $v \in \relv k \Gamma s S$. One worry is that for methods, because
-the parameter type is contravariant, $U$ could require a more precise
-type than $S$. This shouldn't lead to narrowing issues, because if the
-parameter type becomes ill-typed, then the check is vacuously true.
+for $v \in \relv k \Gamma s S$. One worry is potential narrowing
+issues for methods: because the parameter type is contravariant, $U$
+can require a more precise type than $S$. This should be fine, because
+if the parameter type becomes ill-typed, then the check is vacuously
+true.
 \end{proofsketch}

### 2012-08-09 article.tex
ack tiark
move why no inheritance sec to discussion
outline / contrib

@@
 
-\subsection{Why No Inheritance?}
-
-In the calculus we made the deliberate choice not to model any form of
-inheritance. This is, first and foremost, to keep the calculus simple.
-Secondly, there are many different approaches to inheritance and
-mixin composition, so that it looks advantageous not to tie the basic
-calculus to a specific one. Finally, it seems that the modelization of
-inheritance lends itself to a different approach than the basic
-calculus. For the latter, we need to prove type safety of the calculus.
-One might try to do this also for a calculus with inheritance, but our
-experience suggests that this complicates the proofs considerably.  An
-alternative approach that might work better is to model inheritance
-it as a form of code-reuse. Starting with an enriched type system with
-inheritance, and a translation to the basic calculus, one needs to
-show type safety wrt the translation. This might be easier than
-to prove type safety wrt reduction.
-
-\section{The DOT Calculus}
+We propose DOT as a core calculus for path-dependent types. We present
+the calculus formally in section \ref{pres} and through examples in
+section~\ref{dot-examples}. Though we show that the calculus does not
+satisfy the standard theorem of preservation in
+section~\ref{dot-preservation}, we contribute a proof sketch of type
+safety using logical relations in section~\ref{dot-type-safety}. In
+section~\ref{discussion}, we discuss choices and variants of the
+calculus, as well as related work, and conclude in
+section~\ref{conclusion}.
+
+\section{The DOT Calculus}\label{pres}
 
@@
 
+\subsection{Why No Inheritance?}\label{why-no-inheritance}
+
+In the calculus we made the deliberate choice not to model any form of
+inheritance. This is, first and foremost, to keep the calculus simple.
+Secondly, there are many different approaches to inheritance and
+mixin composition, so that it looks advantageous not to tie the basic
+calculus to a specific one. Finally, it seems that the modelization of
+inheritance lends itself to a different approach than the basic
+calculus. For the latter, we need to prove type safety of the calculus.
+One might try to do this also for a calculus with inheritance, but our
+experience suggests that this complicates the proofs considerably.  An
+alternative approach that might work better is to model inheritance
+as a form of code-reuse. Starting with an enriched type system with
+inheritance, and a translation to the basic calculus, one needs to
+show type safety wrt the translation. This might be easier than
+to prove type safety wrt reduction.
+
 \subsection{Variants of the DOT Calculus}\label{dot-variants}
@@
 logical relations to DOT. We thank Donna Malayeri and Geoffrey
-Washburn for preliminary work on DOT.
+Washburn for preliminary work on DOT. We thank Tiark Rompf for helpful
+comments.
 

### 2012-08-09 article.tex
typog of wfe

@@
 \newcommand{\wfe}{\;\mbox{\textbf{wfe}}}
+\newcommand{\nswfe}{\mbox{\textbf{wfe}}}
 
@@
 initializing method definition $m(x) = t$. The parameter type $T$ must
-be $\wfe$ (well-formed and expanding; see \figref{dot-wf}), and
+be $\nswfe$ (well-formed and expanding; see \figref{dot-wf}), and
 the body $t$ must type check to $W$ in an environment extended by the
@@
 \sub T$ which states that in environment $\Gamma$ type $S$ is a
-subtype of type $T$. Subtyping is regular wrt $\wfe$: if
+subtype of type $T$. Subtyping is regular wrt $\nswfe$: if
 type $S$ is a subtype of type $T$, then $S$ and $T$ are well-formed
 and expanding. Though this regularity limits our calculus to
-$\wfe$-types, this limitation allows us to show that subtyping is
+$\nswfe$-types, this limitation allows us to show that subtyping is
 transitive, as discussed in section~\ref{dot-preservation}.
@@
 
-\subsubsection{Why limit the calculus to $\wfe$-types?}
+\subsubsection{Why limit the calculus to $\nswfe$-types?}
 
@@
 fundamentally relies on types having an expansion. However, this was
-not our original motivation for limiting the calculus to $\wfe$-types.
-
-Originally, subtyping was not regular wrt $\wfe$. Roughly,
-all the $\wfe$ preconditions in subtyping were dropped. In this
+not our original motivation for limiting the calculus to $\nswfe$-types.
+
+Originally, subtyping was not regular wrt $\nswfe$. Roughly,
+all the $\nswfe$ preconditions in subtyping were dropped. In this
 broader calculus, subtyping transitivity doesn't hold, because of the

### 2012-08-09 article.tex
forward ref to pres-patch
references

@@
 
+%The primary contributions of this paper are:
+%\begin{itemize}
+%\item Development of DOT -- a core calculus which models
+%  path-dependent types. We formally specify a small-step operational
+%  semantics and a static typing assignment in
+%  section~\ref{dot-calculus} and show some examples in
+%  section~\ref{dot-examples}.
+%\item Counterexamples to the standard theorem of preservation --
+%  section~\ref{dot-preservation}.
+%\item A sketch of a type-safety proof based on logical relations --
+%  section~\ref{dot-type-safety}.
+%\end{itemize}
+
 We propose DOT as a core calculus for path-dependent types. We present
@@
 
-\section{The DOT Calculus}\label{pres}
+\section{The DOT Calculus}\label{dot-calculus}
 
@@
 
+In section~\ref{pres-patch}, we discuss our failure to patch the
+calculus for preservation to hold.
+
 \section{Type Safety via Logical Relations}\label{dot-type-safety}
@@
 
-\subsubsection{Why not patch the DOT calculus for preservation to hold?}
+\subsubsection{Why not patch the DOT calculus for preservation to hold?}\label{pres-patch}
 
@@
 
+\subsection{Related Work}
+
+In addition to Scala's previous models~\cite{nuObj,FS}, several
+calculi present some form of path-dependent types.
+
+The {\it vc} calculus~\cite{vc} models virtual classes with
+path-dependent types. {\it vc} restricts paths to start with ``this'',
+though it provides a way (``out'') to refer to the enclosing object.
+
+The Tribe calculus~\cite{tribe} builds an ownership types
+system~\cite{tribalo} on top of a core calculus which models virtual
+classes. The soundness proof for the core calculus seems to be tied to
+the ownership types system.
+
+Some ML-style module systems~\cite{homl,mixinml} have a form of
+stratified path-dependent types. Because of the stratification,
+recursion is not allowed. In MixML~\cite{mixml} like in Scala, this
+restriction is lifted.
+
 \section{Conclusion}\label{conclusion}

### 2012-08-09 article.tex
fix ref
revised new

@@
 \newcommand{\spcomma}{~,~}
-\newcommand{\sat}{{\bf\;sat\;}}
 
@@
 We propose DOT as a core calculus for path-dependent types. We present
-the calculus formally in section \ref{pres} and through examples in
+the calculus formally in section \ref{dot-calculus} and through examples in
 section~\ref{dot-examples}. Though we show that the calculus does not
@@
 
-\boxfig{dot-one}{The DOT Calculus : Syntax, Reduction, Type Assignment}{
+\boxfig{dot-one}{The DOT Calculus : Syntax, Reduction, Type / Declaration Assignment}{
 {\bf Syntax}\medskip
@@
 \Gamma \ts T_c \wfe \spcomma T_c \expand_y \seq{D} \\
-\Gamma \envplus{y: T_c} \ts \seq{D \sat \seq{d}} \spcomma {t'} \typ {T'}}
+\Gamma \envplus{y: T_c} \ts \seq{d} \typ \seq{D} \spcomma {t'} \typ {T'}}
 {\Gamma \ts \new y {T_c \ldefs{\seq{d}}} {t'} \typ T'}
@@
 
-\begin{multicols}{2}[\judgement{Satisfied Declaration}{\fbox{$\Gamma \ts D \sat \seq{d}$}}]
-
-\infrule[\textsc{tdecl-sat}]
-{\Gamma \ts S \sub U}
-{\Gamma \ts \Ldecl L S U \sat \seq{d}}
-
-\infrule[\textsc{vdecl-sat}]
+\begin{multicols}{2}[\judgement{Declaration Assignment}{\fbox{$\Gamma \ts d \typ D$}}]
+
+\infrule[\textsc{vdecl}]
 {\Gamma \ts v \typ {V'} \spcomma {V'} \sub V}
-{\Gamma \ts \ldecl l V \sat l = v,\seq{d}}
-
-\infrule[\textsc{mdecl-sat}]
+{\Gamma \ts l = v \typ \ldecl l V}
+
+\infrule[\textsc{mdecl}]
 {\Gamma \ts S \wfe\\
  \Gamma \envplus{x : S} \ts t \typ {T'} \spcomma {T'} \sub T}
-{\Gamma \ts \mdecl m S T \sat m(x) = t,\seq{d}}
+{\Gamma \ts m(x) = t \typ \mdecl m S T }
+
+\end{multicols}
+
+\begin{multicols}{2}[\judgement{}{\fbox{$\Gamma \ts \seq{d} \typ \seq{D}$}}]
+
+\infrule[\textsc{decls}]
+{\Gamma \ts \seq{d_i \typ D_i} \spcomma \seq{S \sub U}}
+{\Gamma \ts \seq{d} \typ \seq{D}, \seq{\Ldecl L S U}}
 

### 2012-08-09 article.tex
sprinkled todos in logical relation section

@@
 
+\todo{quick intro that explains the general approach of step-indexed
+  logical relations and the strategy here of setting up $\ds$ in terms
+  of $\mathcal{E}$ -- give an overview of what's coming}
+
 \subsubsection{Set of Values}
@@
 
+\todo{mention possible pitfall: we are not checking type labels of a
+  semantic value, yet because we're substituing the value for the
+  ``self'' in the expansion, the value and its type labels show up and
+  are constrained by T.}
+
 \subsubsection{Set of Terms}
@@
 
+\todo{irred is never formally defined (easy one!)}
+\todo{explain the strange ! notation, or make it more intuitive}
+\todo{briefly introduce the extensions to env and store before they are used}
+
 \subsubsection{Extending the environment and the store}
@@
 {s'} \andl {\Gamma''} \in \relg \Gamma k {s'} {\Gamma'}$.
+
+\todo{expand on why this non-trivial subcase is all that is needed for each case.}
 
@@
 
+\todo{possible pitfall with narrowing needs more cautiousness}
+
 \subsubsection{Extended Monotonicity Lemma}
@@
 agrees with $s$.
+
+\todo{explain that this is the reason why extending the store is so
+  flexible}
 

### 2012-08-10 article.tex
cosmetic change to new typing rule
add the condition that parameter type is wf
sub heading
clarifications in lr

--- 2012-08-09 article.tex
@@
 \newcommand{\rels}[3]{\mathcal{\supseteq}_{#1}\llbracket#2;#3\rrbracket}
-\newcommand{\relg}[3]{\mathcal{\supseteq}_{#1;#2}\llbracket!#3\rrbracket}
+\newcommand{\relg}[3]{\mathcal{\supseteq^!}_{#1;#2}\llbracket#3\rrbracket}
 \newcommand{\irred}[2]{\text{irred }(#1,#2)}
@@
 y \notin \fn(T') \\
-\Gamma \ts T_c \wfe \spcomma T_c \expand_y \seq{D} \\
-\Gamma \envplus{y: T_c} \ts \seq{d} \typ \seq{D} \spcomma {t'} \typ {T'}}
+\Gamma \ts T_c \wfe \spcomma T_c \expand_y \seq{\Ldecl L S U}, \seq{D} \\
+\Gamma \envplus{y: T_c} \ts \seq{S <: U} \spcomma \seq{d} \typ \seq{D} \spcomma {t'} \typ {T'}}
 {\Gamma \ts \new y {T_c \ldefs{\seq{d}}} {t'} \typ T'}
@@
 {\Gamma \ts m(x) = t \typ \mdecl m S T }
-
-\end{multicols}
-
-\begin{multicols}{2}[\judgement{}{\fbox{$\Gamma \ts \seq{d} \typ \seq{D}$}}]
-
-\infrule[\textsc{decls}]
-{\Gamma \ts \seq{d_i \typ D_i} \spcomma \seq{S \sub U}}
-{\Gamma \ts \seq{d} \typ \seq{D}, \seq{\Ldecl L S U}}
 
@@
 
-\subsection{Logical Relations}
+\subsection{Step-Indexed Logical Relations}
+
+In order to ensure that our logical relation is well-founded, we use a
+step index. For each step index $k$, we define the set of values and
+the set of terms that appear to belong to a given type, when taking at
+most $k$ steps. $\Gamma \ds t : T$ is then defined in terms of the
+step-indexed logical relation by requiring it to hold $\forall k$.
 
@@
 &\gap (\forall m_i : S \rightarrow U \in \seq{D},\\
-&\gap\gap t_i \in \rele j {\Gamma \envplus{x : S}} s U)\\
+&\gap\gap \Gamma \ts S \wf \andl t_i \in \rele j {\Gamma \envplus{x : S}} s U)\\
 &\gap \andl\\
@@
 \end{align*}
+
+This relation captures the observation that the only ways for a term
+to get stuck is to have a field selection on an uninitialized field or
+a method invocation on uninitialized method.
+
+Note that this relation doesn't directly impose constraints on the
+type labels. In particular, since we're just enforcing appropriate
+initializations of all fields and methods by looking at an expansion,
+lower bounds of type selections are completely ignored. This is by
+design, because a term doesn't get stuck in a method invocation if
+its argument type doesn't satisfy the lower bound required by a type
+selection -- the upper bound is what matters operationally.
+
+However, a {\it potential pitfall} is that the value itself might
+occur in the types $S$, $U$, $V$, because we substitute it for the
+``self'' occurrences in the expansion.
 
@@
 in the extended store to an irreducible term in $j < k$ steps, then
-this term must be in a corresponding $\mathcal{V}$ set.
+this term must be in a corresponding $\mathcal{V}$ set with $\Gamma$
+now extended to agree with the store resulting from the reduction
+steps.
+
+$\irred {t} {s}$ is a shorthand for $\neg\exists t', s'. \reduction t s
+{t'} {s'}$. $\mathcal{\supseteq}$ is used initially for the possibly
+shorter store to agree with the environment, and can extend both in
+many different ways. $\mathcal{\supseteq}^!$ is used finally for the
+possibly shorter environment to agree with the store, and just extends
+the environment in one straightforward way: hence, it defines
+singleton sets.
+
 \begin{align*}
@@
 
-\subsubsection{$\Gamma \ds t : T$}
+\subsubsection{Terms in the Logical Relation}
 

### 2012-08-10 article.tex
ouch

@@
 The proof is on induction on the derivation of $\Gamma \ts t : T$. For
-each case, we need to show $t \in \rele k \Gamma \emptyset T, \forall k$. The
-non-trivial case is to need to show ${t'} \in \relv {k-j-1} {\Gamma''}
-{s'} T$ by supposing $k > 0$, $(\Gamma'; s) \in \rels k \Gamma \emptyset$, and, for some
-$j < k$, $\reductionl {t} s {t'} {s'} j \andl \irred {t'}
-{s'} \andl {\Gamma''} \in \relg \Gamma k {s'} {\Gamma'}$.
-
-\todo{expand on why this non-trivial subcase is all that is needed for each case.}
+each case, we need to show $t \in \rele k \Gamma \emptyset T, \forall
+k$. The non-trivial case is when $k > 0$ and for some $j < k$,
+$\reductionl {t} s {t'} {s'} j \andl \irred {t'} {s'}$. Then, we need
+to show ${t'} \in \relv {k-j-1} {\Gamma''} {s'} T$ for ${\Gamma''} \in
+\relg \Gamma k {s'} {\Gamma'}$.
 
@@
 $S$ subsumes the expansion of $U$. Therefore, for $v \in \relv k
-\Gamma s U$, we have fewer and ``simpler'' definitions to check than
-for $v \in \relv k \Gamma s S$. One worry is potential narrowing
-issues for methods: because the parameter type is contravariant, $U$
-can require a more precise type than $S$. This should be fine, because
-if the parameter type becomes ill-typed, then the check is vacuously
-true.
+\Gamma s U$, we have fewer definitions to check than for $v \in \relv
+k \Gamma s S$.
+
+There are two {\it potential pitfalls}.
+
+First, because we don't check for type labels, we can end up in a
+situation where a type label defined in both $S$ and $U$ is actually
+only used meaningfully in $U$, and so $v \in \relv k \Gamma s S$ never
+enforces it. Consider $v$ to map to $\Top \refine z {\ldecl l \Top}
+\ldefs{l = v}$, $S$ is $\Top \refine z {\Ldecl L \Top \Top , \ldecl l
+  \Top}$ and $U$ is $\Top \refine z {\Ldecl L \Top \Top , \ldecl l
+  {z.\Top}}$. Now, $v \in \relv k \Gamma s S$ even though it doesn't
+define the type label $L$. But $v \not\in \relv k \Gamma s U$ because
+$v.L$ is not well-formed. \remark{Counterexample!}
+
+Another worry is potential narrowing issues for methods: because the
+parameter type is contravariant, $U$ can require a more precise type
+than $S$. This should be fine, because if the parameter type becomes
+non-expanding, then the check is vacuously true.
 \end{proofsketch}
-
-\todo{possible pitfall with narrowing needs more cautiousness}
 
@@
 
-\todo{explain that this is the reason why extending the store is so
-  flexible}
-
 \begin{proofsketch}
-For the monotonicity with regards to the step-index, this follows
-directly from the definitions of $\mathcal{E}$ and $\mathcal{V}$. For the
-environment and the store, this follows from the definition of $\rels
-k {\Gamma \envplus{x : S}} s$. To extend the environment and the store
-for $x : S$, we can append as much as we want to $\Gamma$ and $s$, to
-get $\Gamma'$ and $s'$, and then ignore the last element which is for
-$x : S$.
+For the monotonicity with regards to the step index, this follows
+directly from the definitions of $\mathcal{E}$ and $\mathcal{V}$. For
+the environment and the store, this follows by design from the
+definition of $\rels k {\Gamma \envplus{x : S}} s$. To extend the
+environment and the store for $x : S$, we can append as much as we
+want to $\Gamma$ and $s$, to get $\Gamma'$ and $s'$, and then ignore
+the last element which is for $x : S$.
 \end{proofsketch}

### 2012-08-10 article.tex
almost there

@@
 T$ simply by being in the store. In addition, it belongs to $\relv k
-\Gamma s T$ for $k > 0$, if it defines all method and value labels in
+\Gamma s T$ for $k > 0$, if it defines all type, method and value labels in
 the expansion of $T$ appropriately for $j < k$ steps.
@@
 &\gap \Gamma \ts T \expand_y \seq{D},\\
+&\gap (\forall L_i : S \rightarrow U \in \seq{D},\\
+&\gap\gap \Gamma \ts v.L_i \wfe \spcomma S \wf \spcomma U \wf \spcomma S \sub U\\
 &\gap (\forall m_i : S \rightarrow U \in \seq{D},\\
-&\gap\gap \Gamma \ts S \wf \andl t_i \in \rele j {\Gamma \envplus{x : S}} s U)\\
+&\gap\gap \Gamma \ts S \wf \spcomma U \wf \andl t_i \in \rele j {\Gamma \envplus{x : S}} s U)\\
 &\gap \andl\\
 &\gap (\forall l_i : V \in \seq{D},\\
-&\gap\gap v_i \in \relv j \Gamma s V)\\
+&\gap\gap \Gamma \ts V \wf \andl v_i \in \relv j \Gamma s V)\\
 &\}
@@
 to get stuck is to have a field selection on an uninitialized field or
-a method invocation on uninitialized method.
-
-Note that this relation doesn't directly impose constraints on the
-type labels. In particular, since we're just enforcing appropriate
-initializations of all fields and methods by looking at an expansion,
-lower bounds of type selections are completely ignored. This is by
-design, because a term doesn't get stuck in a method invocation if
-its argument type doesn't satisfy the lower bound required by a type
-selection -- the upper bound is what matters operationally.
-
-However, a {\it potential pitfall} is that the value itself might
-occur in the types $S$, $U$, $V$, because we substitute it for the
-``self'' occurrences in the expansion.
+a method invocation on uninitialized method. However, a {\it potential
+  pitfall} is that the value itself might occur in the types $S$, $U$,
+$V$, because we substitute it for the ``self'' occurrences in the
+expansion, so the relation makes sure that the type labels exist.
 
@@
 Because $S$ is a subtype of $U$, it should hold that the expansion of
-$S$ subsumes the expansion of $U$. Therefore, for $v \in \relv k
-\Gamma s U$, we have fewer definitions to check than for $v \in \relv
-k \Gamma s S$.
-
-There are two {\it potential pitfalls}.
-
-First, because we don't check for type labels, we can end up in a
-situation where a type label defined in both $S$ and $U$ is actually
-only used meaningfully in $U$, and so $v \in \relv k \Gamma s S$ never
-enforces it. Consider $v$ to map to $\Top \refine z {\ldecl l \Top}
-\ldefs{l = v}$, $S$ is $\Top \refine z {\Ldecl L \Top \Top , \ldecl l
-  \Top}$ and $U$ is $\Top \refine z {\Ldecl L \Top \Top , \ldecl l
-  {z.\Top}}$. Now, $v \in \relv k \Gamma s S$ even though it doesn't
-define the type label $L$. But $v \not\in \relv k \Gamma s U$ because
-$v.L$ is not well-formed. \remark{Counterexample!}
-
-Another worry is potential narrowing issues for methods: because the
-parameter type is contravariant, $U$ can require a more precise type
-than $S$. This should be fine, because if the parameter type becomes
-non-expanding, then the check is vacuously true.
+$S$ subsumes the expansion of $U$, when the ``self'' occurrences are
+of type $S$. Therefore, for $v \in \relv k \Gamma s U$, we have fewer
+declarations to check than for $v \in \relv k \Gamma s S$.
+
+A {\it potential pitfall} is whether some types of the expansion of
+$U$ can become non-expanding when the ``self'' occurrences are
+actually $v$ instead of just abstractly of type $S$, causing a check
+to fail. Another worry is that such a non-expanding type results from
+narrowing of a parameter type. However, we believe this is fine by
+subtyping regularity wtr to $\nswfe$ and $v \in \relv k \Gamma s S$
+which cheks bounds and $\nswfe$ of the type labels.
 \end{proofsketch}

### 2012-08-10 article.tex
ok

@@
 &\gap (\forall L_i : S \rightarrow U \in \seq{D},\\
-&\gap\gap \Gamma \ts v.L_i \wfe \spcomma S \wf \spcomma U \wf \spcomma S \sub U\\
+&\gap\gap \Gamma \ts y.L_i \wfe \spcomma S \wf \spcomma U \wf \spcomma S \sub U \andl\\
+&\gap\gap \Gamma \ts y \ni L_i \typ {S'..U'} \spcomma S \sub S' \spcomma U' \sub U\\
 &\gap (\forall m_i : S \rightarrow U \in \seq{D},\\
-&\gap\gap \Gamma \ts S \wf \spcomma U \wf \andl t_i \in \rele j {\Gamma \envplus{x : S}} s U)\\
+&\gap\gap \Gamma \ts S \wf \spcomma U \wf \andl t_i \in \rele j {\Gamma \envplus{x_i : S}} s U)\\
 &\gap \andl\\
@@
 $V$, because we substitute it for the ``self'' occurrences in the
-expansion, so the relation makes sure that the type labels exist.
+expansion, so the relation makes sure that the type labels are appropriate.
 

### 2012-08-10 article.tex
lil things

@@
 each case, we need to show $t \in \rele k \Gamma \emptyset T, \forall
-k$. The non-trivial case is when $k > 0$ and for some $j < k$,
-$\reductionl {t} s {t'} {s'} j \andl \irred {t'} {s'}$. Then, we need
-to show ${t'} \in \relv {k-j-1} {\Gamma''} {s'} T$ for ${\Gamma''} \in
-\relg \Gamma k {s'} {\Gamma'}$.
+k$. The non-trivial case is when $k > 0$ and for $(\Gamma'; s') \in
+\rels k \Gamma s$ and some $j < k$, $\reductionl {t} s {t'} {s'} j
+\andl \irred {t'} {s'}$. Then, we need to show ${t'} \in \relv {k-j-1}
+      {\Gamma''} {s'} T$ for ${\Gamma''} \in \relg \Gamma k {s'}
+      {\Gamma'}$.
 
@@
 (4) $x \not\in \fn(T)$ and $\Gamma_1$ extends $\Gamma$ and
-$\Gamma_{12}$ included in $\Gamma_1$ and $s_{12}$ extends $s_1$ and
+$\Gamma_{12}$ extends $\Gamma_1$ and $s_{12}$ extends $s_1$ and
 $\Gamma_1$ agrees with $s_1$ and $\Gamma_{12}$ agrees with $s_{12}$

### 2012-08-10 article.tex
minor rev

@@
 formally: If $\emptyset \ts t \typ T$ and $\reductionl t \emptyset {t'} {s'} *$ then
-either $t'$ is a value or $\exists t'', s'', \reduction {t'} {s'}
+either $t'$ is a value or $\exists t'', s''. \reduction {t'} {s'}
 {t''} {s''}$.
@@
 to get stuck is to have a field selection on an uninitialized field or
-a method invocation on uninitialized method. However, a {\it potential
+a method invocation on an uninitialized method. However, a {\it potential
   pitfall} is that the value itself might occur in the types $S$, $U$,

### 2012-08-10 article.tex
subst lemma

@@
 the environment, because it won't occur in $T$ or its expansion, so we
-can use the actual type constructor of $v$, say $T_{v}$, instead of
+can use the type of $v$ in the environment instead of
 $S$ for $x$: ${t} \in \rele {\min(k_1,k_2)} {\Gamma_{12} \envplus{{x}
-    : {T_{v_2}}}} {s_{12} \envplus{{x} \mapsto s_{12}(v)}} T$. This
+    : {\Gamma_{12}(v)}}} {s_{12} \envplus{{x} \mapsto s_{12}(v)}} T$. This
 implies what needs to be shown.

### 2012-08-10 article.tex
compromise

@@
 &\gap (\forall L_i : S \rightarrow U \in \seq{D},\\
-&\gap\gap \Gamma \ts y.L_i \wfe \spcomma S \wf \spcomma U \wf \spcomma S \sub U \andl\\
-&\gap\gap \Gamma \ts y \ni L_i \typ {S'..U'} \spcomma S \sub S' \spcomma U' \sub U\\
+&\gap\gap \Gamma \ts y \ni L_i \typ {S'..U'}) \andl\\
 &\gap (\forall m_i : S \rightarrow U \in \seq{D},\\
-&\gap\gap \Gamma \ts S \wf \spcomma U \wf \andl t_i \in \rele j {\Gamma \envplus{x_i : S}} s U)\\
-&\gap \andl\\
+&\gap\gap t_i \in \rele j {\Gamma \envplus{x_i : S}} s U) \andl\\
 &\gap (\forall l_i : V \in \seq{D},\\
-&\gap\gap \Gamma \ts V \wf \andl v_i \in \relv j \Gamma s V)\\
+&\gap\gap v_i \in \relv j \Gamma s V)\\
 &\}
@@
 $V$, because we substitute it for the ``self'' occurrences in the
-expansion, so the relation makes sure that the type labels are appropriate.
+expansion, so the relation makes sure that the type labels exist.
 
@@
 T$. Thanks to (4), we don't actually need $x$ to be held abstract in
-the environment, because it won't occur in $T$ or its expansion, so we
-can use the type of $v$ in the environment instead of
-$S$ for $x$: ${t} \in \rele {\min(k_1,k_2)} {\Gamma_{12} \envplus{{x}
-    : {\Gamma_{12}(v)}}} {s_{12} \envplus{{x} \mapsto s_{12}(v)}} T$. This
-implies what needs to be shown.
-
+the environment, because it won't occur in $T$ or its expansion (a
+{\it potential pitfall} is whether its occurrences in $t_i$ could
+still cause a check to fail through narrowing issues), so we can use
+the type of $v$ in the environment instead of $S$ for $x$: ${t} \in
+\rele {\min(k_1,k_2)} {\Gamma_{12} \envplus{{x} : {\Gamma_{12}(v)}}}
+      {s_{12} \envplus{{x} \mapsto s_{12}(v)}} T$. This implies what
+      needs to be shown.
 \end{proofsketch}
@@
 to fail. Another worry is that such a non-expanding type results from
-narrowing of a parameter type. However, we believe this is fine by
-subtyping regularity wtr to $\nswfe$ and $v \in \relv k \Gamma s S$
-which cheks bounds and $\nswfe$ of the type labels.
+narrowing of a parameter type.
+%However, we believe this is fine by
+%subtyping regularity wtr to $\nswfe$ and $v \in \relv k \Gamma s S$
+%which cheks bounds and $\nswfe$ of the type labels.
 \end{proofsketch}

### 2012-08-10 article.tex
getting rid of known counterexamples

@@
 &\relv k \Gamma s T = \{ y \ |\\
-&\gap \Gamma \ts T \wfe \andl\\
-&\gap y \in \dom(s) \andl\\
+&\gap y \in \dom(s) \andl (\\
+&(\Gamma \ts T \wfe \andl\\
 &\gap \forall j < k,\\
@@
 &\gap \Gamma \ts T \expand_y \seq{D},\\
-&\gap (\forall L_i : S \rightarrow U \in \seq{D},\\
-&\gap\gap \Gamma \ts y \ni L_i \typ {S'..U'}) \andl\\
 &\gap (\forall m_i : S \rightarrow U \in \seq{D},\\
@@
 &\gap (\forall l_i : V \in \seq{D},\\
-&\gap\gap v_i \in \relv j \Gamma s V)\\
+&\gap\gap v_i \in \relv j \Gamma s V)) \orl\\
+& (T = T_1 \tand T_2 \andl y \in \relv k \Gamma s {T_1} \andl y \in \relv k \Gamma s {T_2}) \orl\\
+& (T = T_1 \tor T_2 \andl  (y \in \relv k \Gamma s {T_1} \orl y \in \relv k \Gamma s {T_2}))\\
 &\}
@@
   pitfall} is that the value itself might occur in the types $S$, $U$,
-$V$, because we substitute it for the ``self'' occurrences in the
-expansion, so the relation makes sure that the type labels exist.
+$V$.
 

### 2012-08-11 article.tex
relax rels by providing alts

@@
 \begin{align*}
-&\relv k \Gamma s T = \{ y \ |\\
-&\gap y \in \dom(s) \andl (\\
-&(\Gamma \ts T \wfe \andl\\
+&\relv k \Gamma s T = \{ y \ | y \in \dom(s) \andl (\\
+& (\Gamma \ts T \wfe \andl\\
 &\gap \forall j < k,\\
@@
 &\gap \Gamma \ts T \expand_y \seq{D},\\
+&\gap (\forall L_i : S \rightarrow U \in \seq{D},\\
+&\gap\gap \Gamma \ts y \ni L_i \typ {S'..U'}) \andl\\
 &\gap (\forall m_i : S \rightarrow U \in \seq{D},\\
@@
 & (T = T_1 \tand T_2 \andl y \in \relv k \Gamma s {T_1} \andl y \in \relv k \Gamma s {T_2}) \orl\\
-& (T = T_1 \tor T_2 \andl  (y \in \relv k \Gamma s {T_1} \orl y \in \relv k \Gamma s {T_2}))\\
-&\}
+& (T = T_1 \tor T_2 \andl  (y \in \relv k \Gamma s {T_1} \orl y \in \relv k \Gamma s {T_2})) \orl\\
+& (T = p.L \andl \Gamma \ts p \ni L \typ {S'..U'} \andl \forall j < k, y \in \relv j \Gamma s {U'})\\
+&)\}
 \end{align*}
@@
   pitfall} is that the value itself might occur in the types $S$, $U$,
-$V$.
+$V$, because we substitute it for the ``self'' occurrences in the
+expansion, so the relation makes sure that the required type labels exist.
 
@@
 &\rele k \Gamma s T = \{ t \ |\\
-&\gap \Gamma \ts T \wfe \andl\\
 &\gap k = 0 \orl (\forall j < k,\\

### 2012-08-27 doc/candidacy/report.tex
template for candidacy exam write-up

%% bare_adv.tex
%% V1.3
%% 2007/01/11
%% by Michael Shell
%% See: 
%% http://www.michaelshell.org/
%% for current contact information.
%%
%% This is a skeleton file demonstrating the advanced use of IEEEtran.cls
%% (requires IEEEtran.cls version 1.7 or later) with an IEEE Computer
%% Society journal paper.
%%
%% Support sites:
%% http://www.michaelshell.org/tex/ieeetran/
%% http://www.ctan.org/tex-archive/macros/latex/contrib/IEEEtran/
%% and
%% http://www.ieee.org/

%%*************************************************************************
%% Legal Notice:
%% This code is offered as-is without any warranty either expressed or
%% implied; without even the implied warranty of MERCHANTABILITY or
%% FITNESS FOR A PARTICULAR PURPOSE! 
%% User assumes all risk.
%% In no event shall IEEE or any contributor to this code be liable for
%% any damages or losses, including, but not limited to, incidental,
%% consequential, or any other damages, resulting from the use or misuse
%% of any information contained here.
%%
%% All comments are the opinions of their respective authors and are not
%% necessarily endorsed by the IEEE.
%%
%% This work is distributed under the LaTeX Project Public License (LPPL)
%% ( http://www.latex-project.org/ ) version 1.3, and may be freely used,
%% distributed and modified. A copy of the LPPL, version 1.3, is included
%% in the base LaTeX documentation of all distributions of LaTeX released
%% 2003/12/01 or later.
%% Retain all contribution notices and credits.
%% ** Modified files should be clearly indicated as such, including  **
%% ** renaming them and changing author support contact information. **
%%
%% File list of work: IEEEtran.cls, IEEEtran_HOWTO.pdf, bare_adv.tex,
%%                    bare_conf.tex, bare_jrnl.tex, bare_jrnl_compsoc.tex
%%*************************************************************************

% *** Authors should verify (and, if needed, correct) their LaTeX system  ***
% *** with the testflow diagnostic prior to trusting their LaTeX platform ***
% *** with production work. IEEE's font choices can trigger bugs that do  ***
% *** not appear when using other class files.                            ***
% The testflow support page is at:
% http://www.michaelshell.org/tex/testflow/

% IEEEtran V1.7 and later provides for these CLASSINPUT macros to allow the
% user to reprogram some IEEEtran.cls defaults if needed. These settings
% override the internal defaults of IEEEtran.cls regardless of which class
% options are used. Do not use these unless you have good reason to do so as
% they can result in nonIEEE compliant documents. User beware. ;)
%
%\newcommand{\CLASSINPUTbaselinestretch}{1.0} % baselinestretch
%\newcommand{\CLASSINPUTinnersidemargin}{1in} % inner side margin
%\newcommand{\CLASSINPUToutersidemargin}{1in} % outer side margin
%\newcommand{\CLASSINPUTtoptextmargin}{1in}   % top text margin
%\newcommand{\CLASSINPUTbottomtextmargin}{1in}% bottom text margin

% Note that the a4paper option is mainly intended so that authors in
% countries using A4 can easily print to A4 and see how their papers will
% look in print - the typesetting of the document will not typically be
% affected with changes in paper size (but the bottom and side margins will).
% Use the testflow package mentioned above to verify correct handling of
% both paper sizes by the user's LaTeX system.
%
% Also note that the "draftcls" or "draftclsnofoot", not "draft", option
% should be used if it is desired that the figures are to be displayed in
% draft mode.
%
\documentclass[10pt,journal,a4paper]{IEEEtran}
% If IEEEtran.cls has not been installed into the LaTeX system files,
% manually specify the path to it like:
% \documentclass[10pt,journal,compsoc]{../sty/IEEEtran}

% For Computer Society journals, IEEEtran defaults to the use of 
% Palatino/Palladio as is done in IEEE Computer Society journals.
% To go back to Times Roman, you can use this code:
%\renewcommand{\rmdefault}{ptm}\selectfont

% Some very useful LaTeX packages include:
% (uncomment the ones you want to load)

% *** MISC UTILITY PACKAGES ***
%
%\usepackage{ifpdf}
% Heiko Oberdiek's ifpdf.sty is very useful if you need conditional
% compilation based on whether the output is pdf or dvi.
% usage:
% \ifpdf
%   % pdf code
% \else
%   % dvi code
% \fi
% The latest version of ifpdf.sty can be obtained from:
% http://www.ctan.org/tex-archive/macros/latex/contrib/oberdiek/
% Also, note that IEEEtran.cls V1.7 and later provides a builtin
% \ifCLASSINFOpdf conditional that works the same way.
% When switching from latex to pdflatex and vice-versa, the compiler may
% have to be run twice to clear warning/error messages.

% *** CITATION PACKAGES ***
%
\ifCLASSOPTIONcompsoc
  % IEEE Computer Society needs nocompress option
  % requires cite.sty v4.0 or later (November 2003)
  % \usepackage[nocompress]{cite}
\else
  % normal IEEE
  % \usepackage{cite}
\fi
% cite.sty was written by Donald Arseneau
% V1.6 and later of IEEEtran pre-defines the format of the cite.sty package
% \cite{} output to follow that of IEEE. Loading the cite package will
% result in citation numbers being automatically sorted and properly
% "compressed/ranged". e.g., [1], [9], [2], [7], [5], [6] without using
% cite.sty will become [1], [2], [5]--[7], [9] using cite.sty. cite.sty's
% \cite will automatically add leading space, if needed. Use cite.sty's
% noadjust option (cite.sty V3.8 and later) if you want to turn this off.
% cite.sty is already installed on most LaTeX systems. Be sure and use
% version 4.0 (2003-05-27) and later if using hyperref.sty. cite.sty does
% not currently provide for hyperlinked citations.
% The latest version can be obtained at:
% http://www.ctan.org/tex-archive/macros/latex/contrib/cite/
% The documentation is contained in the cite.sty file itself.
%
% Note that some packages require special options to format as the Computer
% Society requires. In particular, Computer Society  papers do not use
% compressed citation ranges as is done in typical IEEE papers
% (e.g., [1]-[4]). Instead, they list every citation separately in order
% (e.g., [1], [2], [3], [4]). To get the latter we need to load the cite
% package with the nocompress option which is supported by cite.sty v4.0
% and later. Note also the use of a CLASSOPTION conditional provided by
% IEEEtran.cls V1.7 and later.

% *** GRAPHICS RELATED PACKAGES ***
%
\ifCLASSINFOpdf
  % \usepackage[pdftex]{graphicx}
  % declare the path(s) where your graphic files are
  % \graphicspath{{../pdf/}{../jpeg/}}
  % and their extensions so you won't have to specify these with
  % every instance of \includegraphics
  % \DeclareGraphicsExtensions{.pdf,.jpeg,.png}
\else
  % or other class option (dvipsone, dvipdf, if not using dvips). graphicx
  % will default to the driver specified in the system graphics.cfg if no
  % driver is specified.
  % \usepackage[dvips]{graphicx}
  % declare the path(s) where your graphic files are
  % \graphicspath{{../eps/}}
  % and their extensions so you won't have to specify these with
  % every instance of \includegraphics
  % \DeclareGraphicsExtensions{.eps}
\fi
% graphicx was written by David Carlisle and Sebastian Rahtz. It is
% required if you want graphics, photos, etc. graphicx.sty is already
% installed on most LaTeX systems. The latest version and documentation can
% be obtained at: 
% http://www.ctan.org/tex-archive/macros/latex/required/graphics/
% Another good source of documentation is "Using Imported Graphics in
% LaTeX2e" by Keith Reckdahl which can be found as epslatex.ps or
% epslatex.pdf at: http://www.ctan.org/tex-archive/info/
%
% latex, and pdflatex in dvi mode, support graphics in encapsulated
% postscript (.eps) format. pdflatex in pdf mode supports graphics
% in .pdf, .jpeg, .png and .mps (metapost) formats. Users should ensure
% that all non-photo figures use a vector format (.eps, .pdf, .mps) and
% not a bitmapped formats (.jpeg, .png). IEEE frowns on bitmapped formats
% which can result in "jaggedy"/blurry rendering of lines and letters as
% well as large increases in file sizes.
%
% You can find documentation about the pdfTeX application at:
% http://www.tug.org/applications/pdftex

%\usepackage{ps4pdf}
% dvi->ps workflow is required to use such packages as psfrag.sty and
% pstricks.sty. However, Rolf Niepraschk's ps4pdf.sty provides a way to
% apply psfrag/pstricks effects to .eps figures and then get the resultant
% figures in .pdf form. Thus, providing an easier way for migrating from
% .eps to .pdf figures. After ps4pdf.sty loads, if:
% 1. producing .dvi output: the output file will consist ONLY of the
%    figures (or other constructs encased within \PSforPDF commands)
% 2. producing .pdf output: pdflatex will look in the filename-pics.pdf
%    file, where filename is the basename of the tex document, for the
%    graphics (or other constructs encased within \PSforPDF commands).
%    NOTE: If you ever change your figures, you must remember to remake
%    the filename-pics.pdf file.
%
% This way you can do a:
% 
% latex filename
% dvips -Ppdf -o filename-pics.ps filename.dvi
% ps2pdf filename-pics.ps filename-pics.pdf
% 
% to produce a filename-pics.pdf graphics container that contains
% .pdf versions of the graphics with psfrag, pstricks, etc. features.
% Note that you will not typically be able to view the figures in 
% filename-pics.ps because of an offset. However, you will be able to
% view them in filename-pics.pdf. Also, note that when ps4pdf is in effect
% with .dvi output, you may get harmless over/under full box warnings - 
% ignore them. 
% Then, run pdflatex:
% 
% pdflatex filename
% 
% to use pdflatex to make PDF output, automatically using the figures in
% filename-pics.pdf. Alternatively, you could use dvips -i option to
% obtain separate .pdf files for each figure:
%
% dvips -Ppdf -i -E -o fig filename
%
% then convert each figure to pdf via a command such as epstopdf and then
% use pdflatex with these pdf figures and then to dispense with ps4pdf.
%
% Remember to rerun through latex/dvips/ps2pdf if you ever change your
% figures so that filename-pics.pdf gets updated.
% ps4pdf requires David Kastrup's preview-latex and a recent LaTeX system
% (circa 2001 or later). The ps4pdf package and documentation can be
% obtained at: http://www.ctan.org/tex-archive/macros/latex/contrib/ps4pdf/
% The preview-latex package and documentation can be obtained at:
% http://www.ctan.org/tex-archive/macros/latex/contrib/preview/
%
% provide a bogus \PSforPDF, even when not loading pd4pdf. This way we can
% stop loading ps4pdf.sty if we choose to make separate .pdf versions of
% each of our figures.
\providecommand{\PSforPDF}[1]{#1}
% Note that in order for ps4pdf to work, all commands related to psfrag,
% pstricks, etc. must be called within the PSforPDF command. This applies
% even when *loading* via \usepackage psfrag.sty, etc.

%\PSforPDF{\usepackage{psfrag}}
% psfrag.sty was written by Craig Barratt, Michael C. Grant, and
% David Carlisle. It allows you to substitute LaTeX commands for text in
% imported EPS graphic files. In this way, LaTeX symbols can be placed into
% graphics that have been generated by other applications. You must use
% latex->dvips->ps2pdf workflow (not direct pdf output from pdflatex) if
% you wish to use this capability because it works via some PostScript
% tricks. Alternatively, the graphics could be processed as separate files
% via psfrag and dvips, then converted to PDF for inclusion in the main file
% which uses pdflatex. ps4pdf.sty (above) provides a way of doing this all
% at once within the main file.
% Docs are in "The PSfrag System" by Michael C. Grant and David Carlisle.
% There is also some information about using psfrag in "Using Imported
% Graphics in LaTeX2e" by Keith Reckdahl which documents the graphicx
% package (see above). The psfrag package and documentation can be obtained
% at: http://www.ctan.org/tex-archive/macros/latex/contrib/psfrag/
% 
% Note that the current version of psfrag does not "turn itself off" when
% running under pdf output. This will result in a harmless warning
% about a non-PDF \special. However, to silence this, a bogus psfrag
% command can be provided instead of loading psfrag.sty when PDF output
% is being used. Thus, a more complex alternative conditional loading scheme
% can be employed instead of the straightforword way above:
%
%\ifCLASSINFOpdf
% if outputting PDF, do not use or load psfrag.sty as current versions
% output a non-PDF special that generates a harmless, but annoying warning.
% Instead, we provide a bogus \psfrag command that does nothing with
% its arguments. This is a tad tricky because \psfrag can have up to six
% arguments four of which are optional: \psfrag{}[][][][]{}
% Code based on that in psfrag.sty
%\makeatletter
%\def\psfrag{\@ifstar{\@BOGUSpsfraga}{\@BOGUSpsfraga}}
%\def\@BOGUSpsfraga{\begingroup
%   \@makeother\"\@makeother\*\@makeother\!\@makeother\~%
%   \@makeother\:\@makeother\\\@makeother\%\@makeother\#%
%   \@makeother\ \@BOGUSpsfragb}
%\def\@BOGUSpsfragb#1{\endgroup
%                \@ifnextchar [{\@BOGUSpsfragc}%
%                              {\@BOGUSpsfrag}}
%\def\@BOGUSpsfragc[#1]{\@ifnextchar [{\@BOGUSpsfragd}%
%                                     {\@BOGUSpsfrag}}
%\def\@BOGUSpsfragd[#1]{\@ifnextchar [{\@BOGUSpsfrage}%
%                                     {\@BOGUSpsfrag}}
%\def\@BOGUSpsfrage[#1]{\@ifnextchar [{\@BOGUSpsfragf}%
%                                     {\@BOGUSpsfrag}}
%\def\@BOGUSpsfragf[#1]{\@BOGUSpsfrag}
%\def\@BOGUSpsfrag#1{\ignorespaces}
%\makeatother
%\else
% using dvi output, load psfrag, but funnel it through PSforPDF
% as required by ps4pdf.sty
%\PSforPDF{\usepackage{psfrag}}
%\fi

% *** MATH PACKAGES ***
%
%\usepackage[cmex10]{amsmath}
% A popular package from the American Mathematical Society that provides
% many useful and powerful commands for dealing with mathematics. If using
% it, be sure to load this package with the cmex10 option to ensure that
% only type 1 fonts will utilized at all point sizes. Without this option,
% it is possible that some math symbols, particularly those within
% footnotes, will be rendered in bitmap form which will result in a
% document that can not be IEEE Xplore compliant!
%
% Also, note that the amsmath package sets \interdisplaylinepenalty to 10000
% thus preventing page breaks from occurring within multiline equations. Use:
%\interdisplaylinepenalty=2500
% after loading amsmath to restore such page breaks as IEEEtran.cls normally
% does. amsmath.sty is already installed on most LaTeX systems. The latest
% version and documentation can be obtained at:
% http://www.ctan.org/tex-archive/macros/latex/required/amslatex/math/

% *** SPECIALIZED LIST PACKAGES ***
%\usepackage{acronym}
% acronym.sty was written by Tobias Oetiker. This package provides tools for
% managing documents with large numbers of acronyms. (You don't *have* to
% use this package - unless you have a lot of acronyms, you may feel that
% such package management of them is bit of an overkill.)
% Do note that the acronym environment (which lists acronyms) will have a
% problem when used under IEEEtran.cls because acronym.sty relies on the
% description list environment - which IEEEtran.cls has customized for
% producing IEEE style lists. A workaround is to declared the longest
% label width via the IEEEtran.cls \IEEEiedlistdecl global control:
%
% \renewcommand{\IEEEiedlistdecl}{\IEEEsetlabelwidth{SONET}}
% \begin{acronym}
%
% \end{acronym}
% \renewcommand{\IEEEiedlistdecl}{\relax}% remember to reset \IEEEiedlistdecl
%
% instead of using the acronym environment's optional argument.
% The latest version and documentation can be obtained at:
% http://www.ctan.org/tex-archive/macros/latex/contrib/acronym/

%\usepackage{algorithmic}
% algorithmic.sty was written by Peter Williams and Rogerio Brito.
% This package provides an algorithmic environment fo describing algorithms.
% You can use the algorithmic environment in-text or within a figure
% environment to provide for a floating algorithm. Do NOT use the algorithm
% floating environment provided by algorithm.sty (by the same authors) or
% algorithm2e.sty (by Christophe Fiorio) as IEEE does not use dedicated
% algorithm float types and packages that provide these will not provide
% correct IEEE style captions. The latest version and documentation of
% algorithmic.sty can be obtained at:
% http://www.ctan.org/tex-archive/macros/latex/contrib/algorithms/
% There is also a support site at:
% http://algorithms.berlios.de/index.html
% Also of interest may be the (relatively newer and more customizable)
% algorithmicx.sty package by Szasz Janos:
% http://www.ctan.org/tex-archive/macros/latex/contrib/algorithmicx/

% *** ALIGNMENT PACKAGES ***
%
%\usepackage{array}
% Frank Mittelbach's and David Carlisle's array.sty patches and improves
% the standard LaTeX2e array and tabular environments to provide better
% appearance and additional user controls. As the default LaTeX2e table
% generation code is lacking to the point of almost being broken with
% respect to the quality of the end results, all users are strongly
% advised to use an enhanced (at the very least that provided by array.sty)
% set of table tools. array.sty is already installed on most systems. The
% latest version and documentation can be obtained at:
% http://www.ctan.org/tex-archive/macros/latex/required/tools/

%\usepackage{mdwmath}
%\usepackage{mdwtab}
% Also highly recommended is Mark Wooding's extremely powerful MDW tools,
% especially mdwmath.sty and mdwtab.sty which are used to format equations
% and tables, respectively. The MDWtools set is already installed on most
% LaTeX systems. The lastest version and documentation is available at:
% http://www.ctan.org/tex-archive/macros/latex/contrib/mdwtools/

% IEEEtran contains the IEEEeqnarray family of commands that can be used to
% generate multiline equations as well as matrices, tables, etc., of high
% quality.

%\usepackage{eqparbox}
% Also of notable interest is Scott Pakin's eqparbox package for creating
% (automatically sized) equal width boxes - aka "natural width parboxes".
% Available at:
% http://www.ctan.org/tex-archive/macros/latex/contrib/eqparbox/

% *** SUBFIGURE PACKAGES ***
%\ifCLASSOPTIONcompsoc
%\usepackage[tight,normalsize,sf,SF]{subfigure}
%\else
%\usepackage[tight,footnotesize]{subfigure}
%\fi
% subfigure.sty was written by Steven Douglas Cochran. This package makes it
% easy to put subfigures in your figures. e.g., "Figure 1a and 1b". For IEEE
% work, it is a good idea to load it with the tight package option to reduce
% the amount of white space around the subfigures. Computer Society papers
% use a larger font and \sffamily font for their captions, hence the
% additional options needed under compsoc mode. subfigure.sty is already
% installed on most LaTeX systems. The latest version and documentation can
% be obtained at:
% http://www.ctan.org/tex-archive/obsolete/macros/latex/contrib/subfigure/
% subfigure.sty has been superceeded by subfig.sty.

%\ifCLASSOPTIONcompsoc
%  \usepackage[caption=false]{caption}
%  \usepackage[font=normalsize,labelfont=sf,textfont=sf]{subfig}
%\else
%  \usepackage[caption=false]{caption}
%  \usepackage[font=footnotesize]{subfig}
%\fi
% subfig.sty, also written by Steven Douglas Cochran, is the modern
% replacement for subfigure.sty. However, subfig.sty requires and
% automatically loads Axel Sommerfeldt's caption.sty which will override
% IEEEtran.cls handling of captions and this will result in nonIEEE style
% figure/table captions. To prevent this problem, be sure and preload
% caption.sty with its "caption=false" package option. This is will preserve
% IEEEtran.cls handing of captions. Version 1.3 (2005/06/28) and later 
% (recommended due to many improvements over 1.2) of subfig.sty supports
% the caption=false option directly:
%\ifCLASSOPTIONcompsoc
%  \usepackage[caption=false,font=normalsize,labelfont=sf,textfont=sf]{subfig}
%\else
%  \usepackage[caption=false,font=footnotesize]{subfig}
%\fi
%
% The latest version and documentation can be obtained at:
% http://www.ctan.org/tex-archive/macros/latex/contrib/subfig/
% The latest version and documentation of caption.sty can be obtained at:
% http://www.ctan.org/tex-archive/macros/latex/contrib/caption/

% *** FLOAT PACKAGES ***
%
%\usepackage{fixltx2e}
% fixltx2e, the successor to the earlier fix2col.sty, was written by
% Frank Mittelbach and David Carlisle. This package corrects a few problems
% in the LaTeX2e kernel, the most notable of which is that in current
% LaTeX2e releases, the ordering of single and double column floats is not
% guaranteed to be preserved. Thus, an unpatched LaTeX2e can allow a
% single column figure to be placed prior to an earlier double column
% figure. The latest version and documentation can be found at:
% http://www.ctan.org/tex-archive/macros/latex/base/

%\usepackage{stfloats}
% stfloats.sty was written by Sigitas Tolusis. This package gives LaTeX2e
% the ability to do double column floats at the bottom of the page as well
% as the top. (e.g., "\begin{figure*}[!b]" is not normally possible in
% LaTeX2e). It also provides a command:
%\fnbelowfloat
% to enable the placement of footnotes below bottom floats (the standard
% LaTeX2e kernel puts them above bottom floats). This is an invasive package
% which rewrites many portions of the LaTeX2e float routines. It may not work
% with other packages that modify the LaTeX2e float routines. The latest
% version and documentation can be obtained at:
% http://www.ctan.org/tex-archive/macros/latex/contrib/sttools/
% Documentation is contained in the stfloats.sty comments as well as in the
% presfull.pdf file. Do not use the stfloats baselinefloat ability as IEEE
% does not allow \baselineskip to stretch. Authors submitting work to the
% IEEE should note that IEEE rarely uses double column equations and
% that authors should try to avoid such use. Do not be tempted to use the
% cuted.sty or midfloat.sty packages (also by Sigitas Tolusis) as IEEE does
% not format its papers in such ways.

%\ifCLASSOPTIONcaptionsoff
%  \usepackage[nomarkers]{endfloat}
% \let\MYoriglatexcaption\caption
% \renewcommand{\caption}[2][\relax]{\MYoriglatexcaption[#2]{#2}}
%\fi
% endfloat.sty was written by James Darrell McCauley and Jeff Goldberg.
% This package may be useful when used in conjunction with IEEEtran.cls'
% captionsoff option. Some IEEE journals/societies require that submissions
% have lists of figures/tables at the end of the paper and that
% figures/tables without any captions are placed on a page by themselves at
% the end of the document. If needed, the draftcls IEEEtran class option or
% \CLASSINPUTbaselinestretch interface can be used to increase the line
% spacing as well. Be sure and use the nomarkers option of endfloat to
% prevent endfloat from "marking" where the figures would have been placed
% in the text. The two hack lines of code above are a slight modification of
% that suggested by in the endfloat docs (section 8.3.1) to ensure that
% the full captions always appear in the list of figures/tables - even if
% the user used the short optional argument of \caption[]{}.
% IEEE papers do not typically make use of \caption[]'s optional argument,
% so this should not be an issue. A similar trick can be used to disable
% captions of packages such as subfig.sty that lack options to turn off
% the subcaptions:
% For subfig.sty:
% \let\MYorigsubfloat\subfloat
% \renewcommand{\subfloat}[2][\relax]{\MYorigsubfloat[]{#2}}
% For subfigure.sty:
% \let\MYorigsubfigure\subfigure
% \renewcommand{\subfigure}[2][\relax]{\MYorigsubfigure[]{#2}}
% However, the above trick will not work if both optional arguments of
% the \subfloat/subfig command are used. Furthermore, there needs to be a
% description of each subfigure *somewhere* and endfloat does not add
% subfigure captions to its list of figures. Thus, the best approach is to
% avoid the use of subfigure captions (many IEEE journals avoid them anyway)
% and instead reference/explain all the subfigures within the main caption.
% The latest version of endfloat.sty and its documentation can obtained at:
% http://www.ctan.org/tex-archive/macros/latex/contrib/endfloat/
%
% The IEEEtran \ifCLASSOPTIONcaptionsoff conditional can also be used
% later in the document, say, to conditionally put the References on a 
% page by themselves.

% *** PDF, URL AND HYPERLINK PACKAGES ***
%
%\usepackage{url}
% url.sty was written by Donald Arseneau. It provides better support for
% handling and breaking URLs. url.sty is already installed on most LaTeX
% systems. The latest version can be obtained at:
% http://www.ctan.org/tex-archive/macros/latex/contrib/misc/
% Read the url.sty source comments for usage information. Basically,
% \url{my_url_here}.

% NOTE: PDF thumbnail features are not required in IEEE papers
%       and their use requires extra complexity and work.
%\ifCLASSINFOpdf
%  \usepackage[pdftex]{thumbpdf}
%\else
%  \usepackage[dvips]{thumbpdf}
%\fi
% thumbpdf.sty and its companion Perl utility were written by Heiko Oberdiek.
% It allows the user a way to produce PDF documents that contain fancy
% thumbnail images of each of the pages (which tools like acrobat reader can
% utilize). This is possible even when using dvi->ps->pdf workflow if the
% correct thumbpdf driver options are used. thumbpdf.sty incorporates the
% file containing the PDF thumbnail information (filename.tpm is used with
% dvips, filename.tpt is used with pdftex, where filename is the base name of
% your tex document) into the final ps or pdf output document. An external
% utility, the thumbpdf *Perl script* is needed to make these .tpm or .tpt
% thumbnail files from a .ps or .pdf version of the document (which obviously
% does not yet contain pdf thumbnails). Thus, one does a:
% 
% thumbpdf filename.pdf 
%
% to make a filename.tpt, and:
%
% thumbpdf --mode dvips filename.ps
%
% to make a filename.tpm which will then be loaded into the document by
% thumbpdf.sty the NEXT time the document is compiled (by pdflatex or
% latex->dvips->ps2pdf). Users must be careful to regenerate the .tpt and/or
% .tpm files if the main document changes and then to recompile the
% document to incorporate the revised thumbnails to ensure that thumbnails
% match the actual pages. It is easy to forget to do this!
% 
% Unix systems come with a Perl interpreter. However, MS Windows users
% will usually have to install a Perl interpreter so that the thumbpdf
% script can be run. The Ghostscript PS/PDF interpreter is also required.
% See the thumbpdf docs for details. The latest version and documentation
% can be obtained at.
% http://www.ctan.org/tex-archive/support/thumbpdf/
% Be sure and use only version 3.8 (2005/07/06) or later of thumbpdf as
% earlier versions will not work properly with recent versions of pdfTeX
% (1.20a and later).

% NOTE: PDF hyperlink and bookmark features are not required in IEEE
%       papers and their use requires extra complexity and work.
% *** IF USING HYPERREF BE SURE AND CHANGE THE EXAMPLE PDF ***
% *** TITLE/SUBJECT/AUTHOR/KEYWORDS INFO BELOW!!           ***
\newcommand\MYhyperrefoptions{bookmarks=true,bookmarksnumbered=true,
pdfpagemode={UseOutlines},plainpages=false,pdfpagelabels=true,
colorlinks=true,linkcolor={black},citecolor={black},pagecolor={black},
urlcolor={black},
pdftitle={How to write an EDIC thesis proposal},%<!CHANGE!
pdfsubject={Typesetting},%<!CHANGE!
pdfauthor={EDIC},%<!CHANGE!
pdfkeywords={EDIC, candidacy exam, EPFL}}%<^!CHANGE!
%\ifCLASSINFOpdf
%\usepackage[\MYhyperrefoptions,pdftex]{hyperref}
%\else
%\usepackage[\MYhyperrefoptions,breaklinks=true,dvips]{hyperref}
%\usepackage{breakurl}
%\fi
% One significant drawback of using hyperref under DVI output is that the
% LaTeX compiler cannot break URLs across lines or pages as can be done
% under pdfLaTeX's PDF output via the hyperref pdftex driver. This is
% probably the single most important capability distinction between the
% DVI and PDF output. Perhaps surprisingly, all the other PDF features
% (PDF bookmarks, thumbnails, etc.) can be preserved in
% .tex->.dvi->.ps->.pdf workflow if the respective packages/scripts are
% loaded/invoked with the correct driver options (dvips, etc.). 
% As most IEEE papers use URLs sparingly (mainly in the references), this
% may not be as big an issue as with other publications.
%
% That said, recently Vilar Camara Neto introduced his breakurl.sty
% package which permits hyperref to easily break URLs even in dvi
% mode. Note that breakurl, unlike most other packages, must be loaded
% AFTER hyperref. The latest version of breakurl and its documentation can
% be obtained at:
% http://www.ctan.org/tex-archive/macros/latex/contrib/breakurl/
% breakurl.sty is not for use under pdflatex pdf mode. Versions 1.10 
% (September 23, 2005) and later are recommened to avoid bugs in earlier
% releases.
%
% The advanced features offer by hyperref.sty are not required for IEEE
% submission, so users should weigh these features against the added
% complexity of use. Users who wish to use hyperref *must* ensure that
% their hyperref version is 6.72u or later *and* IEEEtran.cls is version
% 1.6b or later.
% The package options above demonstrate how to enable PDF bookmarks
% (a type of table of contents viewable in Acrobat Reader) as well as
% PDF document information (title, subject, author and keywords) that is
% viewable in Acrobat reader's Document_Properties menu. PDF document
% information is also used extensively to automate the cataloging of PDF
% documents. The above set of options ensures that hyperlinks will not be
% colored in the text and thus will not be visible in the printed page,
% but will be active on "mouse over". USING COLORS OR OTHER HIGHLIGHTING
% OF HYPERLINKS CAN RESULT IN DOCUMENT REJECTION BY THE IEEE, especially if
% these appear on the "printed" page. IF IN DOUBT, ASK THE RELEVANT
% SUBMISSION EDITOR. You may need to add the option hypertexnames=false if
% you used duplicate equation numbers, etc., but this should not be needed
% in normal IEEE work.
% The latest version of hyperref and its documentation can be obtained at:
% http://www.ctan.org/tex-archive/macros/latex/contrib/hyperref/

% *** Do not adjust lengths that control margins, column widths, etc. ***
% *** Do not use packages that alter fonts (such as pslatex).         ***
% There should be no need to do such things with IEEEtran.cls V1.6 and later.
% (Unless specifically asked to do so by the journal or conference you plan
% to submit to, of course. )

% correct bad hyphenation here
\hyphenation{op-tical net-works semi-conduc-tor}

\begin{document}
%
% paper title
% can use linebreaks \\ within to get better formatting as desired
\title{How to Write an EDIC Research Proposal}
%
%
% author names and IEEE memberships
% note positions of commas and nonbreaking spaces ( ~ ) LaTeX will not break
% a structure at a ~ so this keeps an author's name from being broken across
% two lines.
% use \thanks{} to gain access to the first footnote area
% a separate \thanks must be used for each paragraph as LaTeX2e's \thanks
% was not built to handle multiple paragraphs
%
%
%\IEEEcompsocitemizethanks is a special \thanks that produces the bulleted
% lists the Computer Society journals use for "first footnote" author
% affiliations. Use \IEEEcompsocthanksitem which works much like \item
% for each affiliation group. When not in compsoc mode,
% \IEEEcompsocitemizethanks becomes like \thanks and
% \IEEEcompsocthanksitem becomes a line break with idention. This
% facilitates dual compilation, although admittedly the differences in the
% desired content of \author between the different types of papers makes a
% one-size-fits-all approach a daunting prospect. For instance, compsoc 
% journal papers have the author affiliations above the "Manuscript
% received ..."  text while in non-compsoc journals this is reversed. Sigh.

\author{P.H.D.~Student% <-this % stops a space
\\ I\&C,  EPFL
\thanks{\normalsize Proposal submitted to committee: June 13th, 2009; Candidacy exam date: June 20th, 2009; Candidacy exam committee: Exam president, thesis director, co-examiner.}%
\IEEEcompsocitemizethanks{\IEEEcompsocthanksitem }
\thanks{\large This research plan has been approved:}%
\IEEEcompsocitemizethanks{\IEEEcompsocthanksitem \large}
\IEEEcompsocitemizethanks{\IEEEcompsocthanksitem \large}
\IEEEcompsocitemizethanks{\IEEEcompsocthanksitem \large Date:\hfill------------------------------------}
\IEEEcompsocitemizethanks{\IEEEcompsocthanksitem \large}
\IEEEcompsocitemizethanks{\IEEEcompsocthanksitem \large}
\IEEEcompsocitemizethanks{\IEEEcompsocthanksitem \large Doctoral candidate:\hfill------------------------------------}
\IEEEcompsocitemizethanks{\IEEEcompsocthanksitem \footnotesize \hfill                      (name and signature)\hspace{1.5cm}\hfill}
\IEEEcompsocitemizethanks{\IEEEcompsocthanksitem \large}
\IEEEcompsocitemizethanks{\IEEEcompsocthanksitem \large}
\IEEEcompsocitemizethanks{\IEEEcompsocthanksitem \large Thesis director:\hfill------------------------------------}
\IEEEcompsocitemizethanks{\IEEEcompsocthanksitem \footnotesize \hfill                      (name and signature)\hspace{1.5cm}\hfill}
\IEEEcompsocitemizethanks{\IEEEcompsocthanksitem \large}
\IEEEcompsocitemizethanks{\IEEEcompsocthanksitem \large}
\IEEEcompsocitemizethanks{\IEEEcompsocthanksitem \large Thesis co-director:\hfill------------------------------------}
\IEEEcompsocitemizethanks{\IEEEcompsocthanksitem \footnotesize (if applicable)\hfill  (name and signature)\hspace{1.5cm}\hfill}
\IEEEcompsocitemizethanks{\IEEEcompsocthanksitem \large}
\IEEEcompsocitemizethanks{\IEEEcompsocthanksitem \large}
\IEEEcompsocitemizethanks{\IEEEcompsocthanksitem \large Doct. prog. director:\hfill------------------------------------}
\IEEEcompsocitemizethanks{\IEEEcompsocthanksitem \footnotesize (R. Urbanke)  \hfill                    (signature)\hspace{1.5cm}\hfill}
\IEEEcompsocitemizethanks{\IEEEcompsocthanksitem \large}
\IEEEcompsocitemizethanks{\IEEEcompsocthanksitem \large}
\IEEEcompsocitemizethanks{\IEEEcompsocthanksitem \tiny EDIC-ru/05.05.2009}
}

% note the % following the last \IEEEmembership and also \thanks - 
% these prevent an unwanted space from occurring between the last author name
% and the end of the author line. i.e., if you had this:
% 
% \author{....lastname \thanks{...} \thanks{...} }
%                     ^------------^------------^----Do not want these spaces!
%
% a space would be appended to the last name and could cause every name on that
% line to be shifted left slightly. This is one of those "LaTeX things". For
% instance, "\textbf{A} \textbf{B}" will typeset as "A B" not "AB". To get
% "AB" then you have to do: "\textbf{A}\textbf{B}"
% \thanks is no different in this regard, so shield the last } of each \thanks
% that ends a line with a % and do not let a space in before the next \thanks.
% Spaces after \IEEEmembership other than the last one are OK (and needed) as
% you are supposed to have spaces between the names. For what it is worth,
% this is a minor point as most people would not even notice if the said evil
% space somehow managed to creep in.

% The paper headers
\markboth{EDIC Research Proposal}%
{Shell \MakeLowercase{\textit{et al.}}: EDIC Research Proposal}
% The only time the second header will appear is for the odd numbered pages
% after the title page when using the twoside option.
% 
% *** Note that you probably will NOT want to include the author's ***
% *** name in the headers of peer review papers.                   ***
% You can use \ifCLASSOPTIONpeerreview for conditional compilation here if
% you desire.

% The publisher's ID mark at the bottom of the page is less important with
% Computer Society journal papers as those publications place the marks
% outside of the main text columns and, therefore, unlike regular IEEE
% journals, the available text space is not reduced by their presence.
% If you want to put a publisher's ID mark on the page you can do it like
% this:
%\IEEEpubid{0000--0000/00\$00.00~\copyright~2007 IEEE}
% or like this to get the Computer Society new two part style.
%\IEEEpubid{\makebox[\columnwidth]{\hfill 0000--0000/00/\$00.00~\copyright~2007 IEEE}%
%\hspace{\columnsep}\makebox[\columnwidth]{Published by the IEEE Computer Society\hfill}}
% Remember, if you use this you must call \IEEEpubidadjcol in the second
% column for its text to clear the IEEEpubid mark (Computer Society jorunal
% papers don't need this extra clearance.)

% use for special paper notices
%\IEEEspecialpapernotice{(Invited Paper)}

% for Computer Society papers, we must declare the abstract and index terms
% PRIOR to the title within the \IEEEcompsoctitleabstractindextext IEEEtran
% command as these need to go into the title area created by \maketitle.
\IEEEcompsoctitleabstractindextext{%
\begin{abstract}
%\boldmath
About 60 to 100 words describing what this 4--8 page write-up is about. Start
with an abstract that clearly identifies the problem, the contribution of
the paper, and the results (if any) or ongoing research. 
This is not a whole lot of words.  This abstract is
48 words.

\end{abstract}
% IEEEtran.cls defaults to using nonbold math in the Abstract.
% This preserves the distinction between vectors and scalars. However,
% if the journal you are submitting to favors bold math in the abstract,
% then you can use LaTeX's standard command \boldmath at the very start
% of the abstract to achieve this. Many IEEE journals frown on math
% in the abstract anyway. In particular, the Computer Society does
% not want either math or citations to appear in the abstract.

% Note that keywords are not normally used for peerreview papers.
\begin{IEEEkeywords}
thesis proposal, candidacy exam write-up, EDIC, EPFL
\end{IEEEkeywords}}

% make the title area
\maketitle

% To allow for easy dual compilation without having to reenter the
% abstract/keywords data, the \IEEEcompsoctitleabstractindextext text will
% not be used in maketitle, but will appear (i.e., to be "transported")
% here as \IEEEdisplaynotcompsoctitleabstractindextext when compsoc mode
% is not selected <OR> if conference mode is selected - because compsoc
% conference papers position the abstract like regular (non-compsoc)
% papers do!
\IEEEdisplaynotcompsoctitleabstractindextext
% \IEEEdisplaynotcompsoctitleabstractindextext has no effect when using
% compsoc under a non-conference mode.

% For peer review papers, you can put extra information on the cover
% page as needed:
% \ifCLASSOPTIONpeerreview
% \begin{center} \bfseries EDICS Category: 3-BBND \end{center}
% \fi
%
% For peerreview papers, this IEEEtran command inserts a page break and
% creates the second title. It will be ignored for other modes.
\IEEEpeerreviewmaketitle

\section{Introduction}

% Computer Society journal papers do something a tad strange with the very
% first section heading (almost always called "Introduction"). They place it
% ABOVE the main text! IEEEtran.cls currently does not do this for you.
% However, You can achieve this effect by making LaTeX jump through some
% hoops via something like:
%
%\ifCLASSOPTIONcompsoc
%  \noindent\raisebox{2\baselineskip}[0pt][0pt]%
%  {\parbox{\columnwidth}{\section{Introduction}\label{sec:introduction}%
%  \global\everypar=\everypar}}%
%  \vspace{-1\baselineskip}\vspace{-\parskip}\par
%\else
%  \section{Introduction}\label{sec:introduction}\par
%\fi
%
% Admittedly, this is a hack and may well be fragile, but seems to do the
% trick for me. Note the need to keep any \label that may be used right
% after \section in the above as the hack puts \section within a raised box.

% The very first letter is a 2 line initial drop letter followed
% by the rest of the first word in caps (small caps for compsoc).
% 
% form to use if the first word consists of a single letter:
% \IEEEPARstart{A}{demo} file is ....
% 
% form to use if you need the single drop letter followed by
% normal text (unknown if ever used by IEEE):
% \IEEEPARstart{A}{}demo file is ....
% 
% Some journals put the first two words in caps:
% \IEEEPARstart{T}{his demo} file is ....
% 
% Here we have the typical use of a "T" for an initial drop letter
% and "HIS" in caps to complete the first word.
\IEEEPARstart{T}{his} template is not trying to suggest a specific
style for the content of your candidacy exam/thesis proposal paper; it
only concerns the style of the writeup itself.  This template is in LaTeX.
We encourage you to use it for your write-up.

If you insist on using a different word processing program please
match the style as much as possible.  Here is the basic format: The
paper size is A4, the font is 10 point, 
single-spacing, ``times roman like.''
Two columns, each 8.75~cm wide, with 1.5~cm left/right margins.
Sections and sub-sections have bold numbered headings.  Figures have
captions underneath, with a bold label and number, and also a short
caption sentence.  Ditto for tables.  The text itself is a mix of standard
paragraphs, bullets, numbered lists, and equations where appropriate.
References are at the end, in a numbered format, in a slightly smaller
font.

The overall length of this write-up should be 4-8 pages.

\subsection{Purpose of this Write-Up}
% needed in second column of first page if using \IEEEpubid
%\IEEEpubidadjcol
This write-up serves two purposes. First, it forms the basis for
your candidacy exam. As such you should summarize the three papers selected
by your advisor and yourself, and analyze as well as discuss them critically.
Second, the write-up 
is also your thesis proposal.  Therefore, the last one or two pages should
be dedicated to your own preliminary work. A road-map of how you plan to
advance the state of the art in your chosen area should also be given.
For further details please consult the document ``PhD Candidacy Exam Overview.'' You can find the latest
version at {\tt http://phd.epfl.ch/page57746-en.html}.

\subsection{What to Put into the Introduction}
Describe briefly the context, the problem, shortcomings in prior
approaches, and your proposed approach and solution. Forecast results.

% An example of a floating figure using the graphicx package.
% Note that \label must occur AFTER (or within) \caption.
% For figures, \caption should occur after the \includegraphics.
% Note that IEEEtran v1.7 and later has special internal code that
% is designed to preserve the operation of \label within \caption
% even when the captionsoff option is in effect. However, because
% of issues like this, it may be the safest practice to put all your
% \label just after \caption rather than within \caption{}.
%
% Reminder: the "draftcls" or "draftclsnofoot", not "draft", class
% option should be used if it is desired that the figures are to be
% displayed while in draft mode.
%
%\begin{figure}[!t]
%\centering
%\includegraphics[width=2.5in]{myfigure}
% where an .eps filename suffix will be assumed under latex, 
% and a .pdf suffix will be assumed for pdflatex; or what has been declared
% via \DeclareGraphicsExtensions.
%\caption{Simulation Results}
%\label{fig_sim}
%\end{figure}

% Note that IEEE typically puts floats only at the top, even when this
% results in a large percentage of a column being occupied by floats.
% However, the Computer Society has been known to put floats at the bottom.

% An example of a double column floating figure using two subfigures.
% (The subfig.sty package must be loaded for this to work.)
% The subfigure \label commands are set within each subfloat command, the
% \label for the overall figure must come after \caption.
% \hfil must be used as a separator to get equal spacing.
% The subfigure.sty package works much the same way, except \subfigure is
% used instead of \subfloat.
%
%\begin{figure*}[!t]
%\centerline{\subfloat[Case I]\includegraphics[width=2.5in]{subfigcase1}%
%\label{fig_first_case}}
%\hfil
%\subfloat[Case II]{\includegraphics[width=2.5in]{subfigcase2}%
%\label{fig_second_case}}}
%\caption{Simulation results}
%\label{fig_sim}
%\end{figure*}
%
% Note that often IEEE papers with subfigures do not employ subfigure
% captions (using the optional argument to \subfloat), but instead will
% reference/describe all of them (a), (b), etc., within the main caption.

% An example of a floating table. Note that, for IEEE style tables, the 
% \caption command should come BEFORE the table. Table text will default to
% \footnotesize as IEEE normally uses this smaller font for tables.
% The \label must come after \caption as always.
%
%\begin{table}[!t]
%% increase table row spacing, adjust to taste
%\renewcommand{\arraystretch}{1.3}
% if using array.sty, it might be a good idea to tweak the value of
% \extrarowheight as needed to properly center the text within the cells
%\caption{An Example of a Table}
%\label{table_example}
%\centering
%% Some packages, such as MDW tools, offer better commands for making tables
%% than the plain LaTeX2e tabular which is used here.
%\begin{tabular}{|c||c|}
%\hline
%One & Two\\
%\hline
%Three & Four\\
%\hline
%\end{tabular}
%\end{table}

% Note that IEEE does not put floats in the very first column - or typically
% anywhere on the first page for that matter. Also, in-text middle ("here")
% positioning is not used. Most IEEE journals use top floats exclusively.
% However, Computer Society journals sometimes do use bottom floats - bear
% this in mind when choosing appropriate optional arguments for the
% figure/table environments.
% Note that, LaTeX2e, unlike IEEE journals, places footnotes above bottom
% floats. This can be corrected via the \fnbelowfloat command of the
% stfloats package.

\section{Survey of the Selected Papers: The Giants upon whose shoulders I will stand}
Background --- Describe the three papers in detail, the problem they
tackle, the solutions and results, and their shortcomings, and how they
relate to your work.  This part builds the basis for the oral candidacy exam.

\section{Research Proposal: How I Plan to Revolutionize the World}
Describe your own work and include a  summary and references. Write how you
propose to advance the state of the art given the background. What
is new technically? How does it improve over prior work?
Summarize, suggest an approximate timeline, and list references.

\vspace{1cm}
This write-up should be 4-8 pages. \\
The quick brown fox jumps over the lazy dog.\\
The quick brown fox jumps over the lazy dog.\\
The quick brown fox jumps over the lazy dog.\\
The quick brown fox jumps over the lazy dog.\\
The quick brown fox jumps over the lazy dog.\\
The quick brown fox jumps over the lazy dog.\\
The quick brown fox jumps over the lazy dog.\\
The quick brown fox jumps over the lazy dog.\\
The quick brown fox jumps over the lazy dog.\\
The quick brown fox jumps over the lazy dog.\\
The quick brown fox jumps over the lazy dog.\\
The quick brown fox jumps over the lazy dog.\\
The quick brown fox jumps over the lazy dog.\\
The quick brown fox jumps over the lazy dog.\\
The quick brown fox jumps over the lazy dog.\\
The quick brown fox jumps over the lazy dog.\\
The quick brown fox jumps over the lazy dog.\\
The quick brown fox jumps over the lazy dog.\\
The quick brown fox jumps over the lazy dog.\\
The quick brown fox jumps over the lazy dog.\\
The quick brown fox jumps over the lazy dog.\\
The quick brown fox jumps over the lazy dog.\\
The quick brown fox jumps over the lazy dog.\\
The quick brown fox jumps over the lazy dog.\\
The quick brown fox jumps over the lazy dog.\\
The quick brown fox jumps over the lazy dog.\\
The quick brown fox jumps over the lazy dog.\\
The quick brown fox jumps over the lazy dog.\\
The quick brown fox jumps over the lazy dog.\\
The quick brown fox jumps over the lazy dog.\\
The quick brown fox jumps over the lazy dog.\\
The quick brown fox jumps over the lazy dog.\\
The quick brown fox jumps over the lazy dog.\\
The quick brown fox jumps over the lazy dog.\\
The quick brown fox jumps over the lazy dog.\\
The quick brown fox jumps over the lazy dog.\\
The quick brown fox jumps over the lazy dog.\\
The quick brown fox jumps over the lazy dog.\\
The quick brown fox jumps over the lazy dog.\\
The quick brown fox jumps over the lazy dog.\\
The quick brown fox jumps over the lazy dog.\\
The quick brown fox jumps over the lazy dog.\\
The quick brown fox jumps over the lazy dog.\\
The quick brown fox jumps over the lazy dog.\\
The quick brown fox jumps over the lazy dog.\\
The quick brown fox jumps over the lazy dog.\\
The quick brown fox jumps over the lazy dog.\\
The quick brown fox jumps over the lazy dog.\\
The quick brown fox jumps over the lazy dog.\\
The quick brown fox jumps over the lazy dog.\\
The quick brown fox jumps over the lazy dog.\\
The quick brown fox jumps over the lazy dog.\\
The quick brown fox jumps over the lazy dog.\\
The quick brown fox jumps over the lazy dog.\\
The quick brown fox jumps over the lazy dog.\\
The quick brown fox jumps over the lazy dog.\\
The quick brown fox jumps over the lazy dog.\\
The quick brown fox jumps over the lazy dog.\\
The quick brown fox jumps over the lazy dog.\\
The quick brown fox jumps over the lazy dog.\\
The quick brown fox jumps over the lazy dog.\\
The quick brown fox jumps over the lazy dog.\\
The quick brown fox jumps over the lazy dog.\\
The quick brown fox jumps over the lazy dog.\\
The quick brown fox jumps over the lazy dog.\\
The quick brown fox jumps over the lazy dog.\\
The quick brown fox jumps over the lazy dog.\\
The quick brown fox jumps over the lazy dog.\\
The quick brown fox jumps over the lazy dog.\\
The quick brown fox jumps over the lazy dog.\\
The quick brown fox jumps over the lazy dog.\\
The quick brown fox jumps over the lazy dog.\\
The quick brown fox jumps over the lazy dog.\\
The quick brown fox jumps over the lazy dog.\\
The quick brown fox jumps over the lazy dog.\\
The quick brown fox jumps over the lazy dog.\\
The quick brown fox jumps over the lazy dog.\\
The quick brown fox jumps over the lazy dog.\\
The quick brown fox jumps over the lazy dog.\\
The quick brown fox jumps over the lazy dog.\\
The quick brown fox jumps over the lazy dog.\\
The quick brown fox jumps over the lazy dog.\\
The quick brown fox jumps over the lazy dog.\\
The quick brown fox jumps over the lazy dog.\\
The quick brown fox jumps over the lazy dog.\\
The quick brown fox jumps over the lazy dog.\\
The quick brown fox jumps over the lazy dog.\\
The quick brown fox jumps over the lazy dog.\\
The quick brown fox jumps over the lazy dog.\\
The quick brown fox jumps over the lazy dog.\\
The quick brown fox jumps over the lazy dog.\\
The quick brown fox jumps over the lazy dog.\\
The quick brown fox jumps over the lazy dog.\\
The quick brown fox jumps over the lazy dog.\\
The quick brown fox jumps over the lazy dog.\\
The quick brown fox jumps over the lazy dog.\\
The quick brown fox jumps over the lazy dog.\\
The quick brown fox jumps over the lazy dog.\\
The quick brown fox jumps over the lazy dog.\\
The quick brown fox jumps over the lazy dog.\\
The quick brown fox jumps over the lazy dog.\\
The quick brown fox jumps over the lazy dog.\\
The quick brown fox jumps over the lazy dog.\\
The quick brown fox jumps over the lazy dog.\\
The quick brown fox jumps over the lazy dog.\\
The quick brown fox jumps over the lazy dog.\\
The quick brown fox jumps over the lazy dog.\\
The quick brown fox jumps over the lazy dog.\\
The quick brown fox jumps over the lazy dog.\\
The quick brown fox jumps over the lazy dog.\\
The quick brown fox jumps over the lazy dog.\\
The quick brown fox jumps over the lazy dog.\\
The quick brown fox jumps over the lazy dog.\\
The quick brown fox jumps over the lazy dog.\\
The quick brown fox jumps over the lazy dog.\\
The quick brown fox jumps over the lazy dog.\\
The quick brown fox jumps over the lazy dog.\\
The quick brown fox jumps over the lazy dog.\\
The quick brown fox jumps over the lazy dog.\\
The quick brown fox jumps over the lazy dog.\\
The quick brown fox jumps over the lazy dog.\\
The quick brown fox jumps over the lazy dog.\\
The quick brown fox jumps over the lazy dog.\\
The quick brown fox jumps over the lazy dog.\\
The quick brown fox jumps over the lazy dog.\\
The quick brown fox jumps over the lazy dog.\\
The quick brown fox jumps over the lazy dog.\\
The quick brown fox jumps over the lazy dog.\\
The quick brown fox jumps over the lazy dog.\\
The quick brown fox jumps over the lazy dog.\\
The quick brown fox jumps over the lazy dog.\\
The quick brown fox jumps over the lazy dog.\\
The quick brown fox jumps over the lazy dog.\\
The quick brown fox jumps over the lazy dog.\\
The quick brown fox jumps over the lazy dog.\\
The quick brown fox jumps over the lazy dog.\\
The quick brown fox jumps over the lazy dog.\\
The quick brown fox jumps over the lazy dog.\\
The quick brown fox jumps over the lazy dog.\\
The quick brown fox jumps over the lazy dog.\\
The quick brown fox jumps over the lazy dog.\\
The quick brown fox jumps over the lazy dog.\\
The quick brown fox jumps over the lazy dog.\\
The quick brown fox jumps over the lazy dog.\\
The quick brown fox jumps over the lazy dog.\\
The quick brown fox jumps over the lazy dog.\\
The quick brown fox jumps over the lazy dog.\\
The quick brown fox jumps over the lazy dog.\\
The quick brown fox jumps over the lazy dog.\\
The quick brown fox jumps over the lazy dog.\\
The quick brown fox jumps over the lazy dog.\\
The quick brown fox jumps over the lazy dog.\\
The quick brown fox jumps over the lazy dog.\\
The quick brown fox jumps over the lazy dog.\\
The quick brown fox jumps over the lazy dog.\\
The quick brown fox jumps over the lazy dog.\\
The quick brown fox jumps over the lazy dog.\\
The quick brown fox jumps over the lazy dog.\\
The quick brown fox jumps over the lazy dog.\\
The quick brown fox jumps over the lazy dog.\\
The quick brown fox jumps over the lazy dog.\\
The quick brown fox jumps over the lazy dog.\\
The quick brown fox jumps over the lazy dog.\\
The quick brown fox jumps over the lazy dog.\\
The quick brown fox jumps over the lazy dog.\\
The quick brown fox jumps over the lazy dog.\\
The quick brown fox jumps over the lazy dog.\\
The quick brown fox jumps over the lazy dog.\\
The quick brown fox jumps over the lazy dog.\\
The quick brown fox jumps over the lazy dog.\\
The quick brown fox jumps over the lazy dog.\\
The quick brown fox jumps over the lazy dog.\\
The quick brown fox jumps over the lazy dog.\\
The quick brown fox jumps over the lazy dog.\\
The quick brown fox jumps over the lazy dog.\\
The quick brown fox jumps over the lazy dog.\\
The quick brown fox jumps over the lazy dog.\\
The quick brown fox jumps over the lazy dog.\\
The quick brown fox jumps over the lazy dog.\\
The quick brown fox jumps over the lazy dog.\\
The quick brown fox jumps over the lazy dog.\\
The quick brown fox jumps over the lazy dog.\\
The quick brown fox jumps over the lazy dog.\\
The quick brown fox jumps over the lazy dog.\\
The quick brown fox jumps over the lazy dog.\\
The quick brown fox jumps over the lazy dog.\\
The quick brown fox jumps over the lazy dog.\\
The quick brown fox jumps over the lazy dog.\\
The quick brown fox jumps over the lazy dog.\\
The quick brown fox jumps over the lazy dog.\\
The quick brown fox jumps over the lazy dog.\\
The quick brown fox jumps over the lazy dog.\\
The quick brown fox jumps over the lazy dog.\\
The quick brown fox jumps over the lazy dog.\\
The quick brown fox jumps over the lazy dog.\\
The quick brown fox jumps over the lazy dog.\\
The quick brown fox jumps over the lazy dog.\\
The quick brown fox jumps over the lazy dog.\\
The quick brown fox jumps over the lazy dog.\\
The quick brown fox jumps over the lazy dog.\\
The quick brown fox jumps over the lazy dog.\\
The quick brown fox jumps over the lazy dog.\\
The quick brown fox jumps over the lazy dog.\\
The quick brown fox jumps over the lazy dog.\\
The quick brown fox jumps over the lazy dog.\\
The quick brown fox jumps over the lazy dog.\\
The quick brown fox jumps over the lazy dog.\\
The quick brown fox jumps over the lazy dog.\\
The quick brown fox jumps over the lazy dog.\\
The quick brown fox jumps over the lazy dog.\\
The quick brown fox jumps over the lazy dog.\\
The quick brown fox jumps over the lazy dog.\\
The quick brown fox jumps over the lazy dog.\\
The quick brown fox jumps over the lazy dog.\\
The quick brown fox jumps over the lazy dog.\\
The quick brown fox jumps over the lazy dog.\\
The quick brown fox jumps over the lazy dog.\\
The quick brown fox jumps over the lazy dog.\\
The quick brown fox jumps over the lazy dog.\\
The quick brown fox jumps over the lazy dog.\\
The quick brown fox jumps over the lazy dog.\\
The quick brown fox jumps over the lazy dog.\\
The quick brown fox jumps over the lazy dog.\\
The quick brown fox jumps over the lazy dog.\\
The quick brown fox jumps over the lazy dog.\\
The quick brown fox jumps over the lazy dog.\\
The quick brown fox jumps over the lazy dog.\\
The quick brown fox jumps over the lazy dog.\\
The quick brown fox jumps over the lazy dog.\\
The quick brown fox jumps over the lazy dog.\\
The quick brown fox jumps over the lazy dog.\\
The quick brown fox jumps over the lazy dog.\\
The quick brown fox jumps over the lazy dog.\\
The quick brown fox jumps over the lazy dog.\\
The quick brown fox jumps over the lazy dog.\\
The quick brown fox jumps over the lazy dog.\\
The quick brown fox jumps over the lazy dog.\\
The quick brown fox jumps over the lazy dog.\\
The quick brown fox jumps over the lazy dog.\\
The quick brown fox jumps over the lazy dog.\\
The quick brown fox jumps over the lazy dog.\\
The quick brown fox jumps over the lazy dog.\\
The quick brown fox jumps over the lazy dog.\\
The quick brown fox jumps over the lazy dog.\\
The quick brown fox jumps over the lazy dog.\\
The quick brown fox jumps over the lazy dog.\\
The quick brown fox jumps over the lazy dog.\\
The quick brown fox jumps over the lazy dog.\\
The quick brown fox jumps over the lazy dog.\\
The quick brown fox jumps over the lazy dog.\\
The quick brown fox jumps over the lazy dog.\\
The quick brown fox jumps over the lazy dog.\\
The quick brown fox jumps over the lazy dog.\\
The quick brown fox jumps over the lazy dog.\\
The quick brown fox jumps over the lazy dog.\\
The quick brown fox jumps over the lazy dog.\\
The quick brown fox jumps over the lazy dog.\\
The quick brown fox jumps over the lazy dog.\\
The quick brown fox jumps over the lazy dog.\\
The quick brown fox jumps over the lazy dog.\\
The quick brown fox jumps over the lazy dog.\\
The quick brown fox jumps over the lazy dog.\\
The quick brown fox jumps over the lazy dog.\\
The quick brown fox jumps over the lazy dog.\\
The quick brown fox jumps over the lazy dog.\\
The quick brown fox jumps over the lazy dog.\\
The quick brown fox jumps over the lazy dog.\\
The quick brown fox jumps over the lazy dog.\\
The quick brown fox jumps over the lazy dog.\\
The quick brown fox jumps over the lazy dog.\\
The quick brown fox jumps over the lazy dog.\\
The quick brown fox jumps over the lazy dog.\\
The quick brown fox jumps over the lazy dog.\\
The quick brown fox jumps over the lazy dog.\\
The quick brown fox jumps over the lazy dog.\\
The quick brown fox jumps over the lazy dog.\\
The quick brown fox jumps over the lazy dog.\\
The quick brown fox jumps over the lazy dog.\\
The quick brown fox jumps over the lazy dog.\\
The quick brown fox jumps over the lazy dog.\\
The quick brown fox jumps over the lazy dog.\\
The quick brown fox jumps over the lazy dog.\\
The quick brown fox jumps over the lazy dog.\\
The quick brown fox jumps over the lazy dog.\\
The quick brown fox jumps over the lazy dog.\\
The quick brown fox jumps over the lazy dog.\\
The quick brown fox jumps over the lazy dog.\\
The quick brown fox jumps over the lazy dog.\\
The quick brown fox jumps over the lazy dog.\\
The quick brown fox jumps over the lazy dog.\\
The quick brown fox jumps over the lazy dog.\\
The quick brown fox jumps over the lazy dog.\\
The quick brown fox jumps over the lazy dog.\\
The quick brown fox jumps over the lazy dog.\\
The quick brown fox jumps over the lazy dog.\\
The quick brown fox jumps over the lazy dog.\\
The quick brown fox jumps over the lazy dog.\\
The quick brown fox jumps over the lazy dog.\\
The quick brown fox jumps over the lazy dog.\\
The quick brown fox jumps over the lazy dog.\\
The quick brown fox jumps over the lazy dog.\\
The quick brown fox jumps over the lazy dog.\\
The quick brown fox jumps over the lazy dog.\\
The quick brown fox jumps over the lazy dog.\\
The quick brown fox jumps over the lazy dog.\\
The quick brown fox jumps over the lazy dog.\\
The quick brown fox jumps over the lazy dog.\\
The quick brown fox jumps over the lazy dog.\\
The quick brown fox jumps over the lazy dog.\\
The quick brown fox jumps over the lazy dog.\\
The quick brown fox jumps over the lazy dog.\\
The quick brown fox jumps over the lazy dog.\\
The quick brown fox jumps over the lazy dog.\\
The quick brown fox jumps over the lazy dog.\\
The quick brown fox jumps over the lazy dog.\\
The quick brown fox jumps over the lazy dog.\\
The quick brown fox jumps over the lazy dog.\\
The quick brown fox jumps over the lazy dog.\\
The quick brown fox jumps over the lazy dog.\\
The quick brown fox jumps over the lazy dog.\\
The quick brown fox jumps over the lazy dog.\\
The quick brown fox jumps over the lazy dog.\\
The quick brown fox jumps over the lazy dog.\\
The quick brown fox jumps over the lazy dog.\\
The quick brown fox jumps over the lazy dog.\\
The quick brown fox jumps over the lazy dog.\\
The quick brown fox jumps over the lazy dog.\\
The quick brown fox jumps over the lazy dog.\\
The quick brown fox jumps over the lazy dog.\\
The quick brown fox jumps over the lazy dog.\\
The quick brown fox jumps over the lazy dog.\\
The quick brown fox jumps over the lazy dog.\\
The quick brown fox jumps over the lazy dog.\\
The quick brown fox jumps over the lazy dog.\\
The quick brown fox jumps over the lazy dog.\\
The quick brown fox jumps over the lazy dog.\\
The quick brown fox jumps over the lazy dog.\\
The quick brown fox jumps over the lazy dog.\\
The quick brown fox jumps over the lazy dog.\\
The quick brown fox jumps over the lazy dog.\\
The quick brown fox jumps over the lazy dog.\\
The quick brown fox jumps over the lazy dog.\\
The quick brown fox jumps over the lazy dog.\\
The quick brown fox jumps over the lazy dog.\\
The quick brown fox jumps over the lazy dog.\\
The quick brown fox jumps over the lazy dog.\\
The quick brown fox jumps over the lazy dog.\\
The quick brown fox jumps over the lazy dog.\\

% if have a single appendix:
%\appendix[Proof of the Zonklar Equations]
% or
%\appendix  % for no appendix heading
% do not use \section anymore after \appendix, only \section*
% is possibly needed

% use appendices with more than one appendix
% then use \section to start each appendix
% you must declare a \section before using any
% \subsection or using \label (\appendices by itself
% starts a section numbered zero.)
%

%\appendices
%\section{Proof of the First Zonklar Equation}
%Appendix one text goes here.

% you can choose not to have a title for an appendix
% if you want by leaving the argument blank
%\section{}

% use section* for acknowledgement
%\ifCLASSOPTIONcompsoc
%  % The Computer Society usually uses the plural form
%  \section*{Acknowledgments}
%\else
%  % regular IEEE prefers the singular form
%  \section*{Acknowledgment}
%\fi
%
%
%The authors would like to thank...
%
%
% Can use something like this to put references on a page
% by themselves when using endfloat and the captionsoff option.
%\ifCLASSOPTIONcaptionsoff
%  \newpage
%\fi

% trigger a \newpage just before the given reference
% number - used to balance the columns on the last page
% adjust value as needed - may need to be readjusted if
% the document is modified later
%\IEEEtriggeratref{8}
% The "triggered" command can be changed if desired:
%\IEEEtriggercmd{\enlargethispage{-5in}}

% references section

% can use a bibliography generated by BibTeX as a .bbl file
% BibTeX documentation can be easily obtained at:
% http://www.ctan.org/tex-archive/biblio/bibtex/contrib/doc/
% The IEEEtran BibTeX style support page is at:
% http://www.michaelshell.org/tex/ieeetran/bibtex/
%\bibliographystyle{IEEEtran}
% argument is your BibTeX string definitions and bibliography database(s)
%\bibliography{IEEEabrv,../bib/paper}
%
% <OR> manually copy in the resultant .bbl file
% set second argument of \begin to the number of references
% (used to reserve space for the reference number labels box)
\begin{thebibliography}{1}

\bibitem{mittelbach}
F.~Mittelbach, M.~Goossens, J.~Braams, D.~Carlisle, and C.~Rowley, 
\emph{The {\LaTeX}} Companion, 2nd~ed.\hskip 1em plus
0.5em minus 0.4em\relax Addison-Wesley Professional, 2004.

\end{thebibliography}

% biography section
% 
% If you have an EPS/PDF photo (graphicx package needed) extra braces are
% needed around the contents of the optional argument to biography to prevent
% the LaTeX parser from getting confused when it sees the complicated
% \includegraphics command within an optional argument. (You could create
% your own custom macro containing the \includegraphics command to make things
% simpler here.)
%\begin{biography}[{\includegraphics[width=1in,height=1.25in,clip,keepaspectratio]{mshell}}]{Michael Shell}
% or if you just want to reserve a space for a photo:

%\begin{IEEEbiography}{Michael Shell}
%Biography text here.
%\end{IEEEbiography}

% if you will not have a photo at all:
%\begin{IEEEbiographynophoto}{John Doe}
%Biography text here.
%\end{IEEEbiographynophoto}

% insert where needed to balance the two columns on the last page with
% biographies
%\newpage

%\begin{IEEEbiographynophoto}{Jane Doe}
%Biography text here.
%\end{IEEEbiographynophoto}

% You can push biographies down or up by placing
% a \vfill before or after them. The appropriate
% use of \vfill depends on what kind of text is
% on the last page and whether or not the columns
% are being equalized.

%\vfill

% Can be used to pull up biographies so that the bottom of the last one
% is flush with the other column.
%\enlargethispage{-5in}

% that's all folks
\end{document}

### 2012-08-27 doc/candidacy/report.tex
changed titles

@@
 urlcolor={black},
-pdftitle={How to write an EDIC thesis proposal},%<!CHANGE!
-pdfsubject={Typesetting},%<!CHANGE!
-pdfauthor={EDIC},%<!CHANGE!
-pdfkeywords={EDIC, candidacy exam, EPFL}}%<^!CHANGE!
+pdftitle={Dependent Object Types},%<!CHANGE!
+pdfsubject={Programming Languages},%<!CHANGE!
+pdfauthor={Nada Amin},%<!CHANGE!
+pdfkeywords={Nada Amin, candidacy exam, EPFL}}%<^!CHANGE!
 %\ifCLASSINFOpdf
@@
 % can use linebreaks \\ within to get better formatting as desired
-\title{How to Write an EDIC Research Proposal}
+\title{Dependent Object Types}
 %
@@
 
-\author{P.H.D.~Student% <-this % stops a space
-\\ I\&C,  EPFL
-\thanks{\normalsize Proposal submitted to committee: June 13th, 2009; Candidacy exam date: June 20th, 2009; Candidacy exam committee: Exam president, thesis director, co-examiner.}%
+\author{Nada Amin\\
+\ \ \ \ \ \ \ \ \ \ \ \ \ \ \ \ \ \ LAMP, I\&C, EPFL
+\thanks{\normalsize Proposal submitted to committee: September 5th, 2012; Candidacy exam date: September 12th, 2012; Candidacy exam committee: Rachid Guerraoui, Martin Odersky, Viktor Kuncak.}%
 \IEEEcompsocitemizethanks{\IEEEcompsocthanksitem }
@@
 \IEEEcompsocitemizethanks{\IEEEcompsocthanksitem \footnotesize \hfill                      (name and signature)\hspace{1.5cm}\hfill}
-\IEEEcompsocitemizethanks{\IEEEcompsocthanksitem \large}
-\IEEEcompsocitemizethanks{\IEEEcompsocthanksitem \large}
-\IEEEcompsocitemizethanks{\IEEEcompsocthanksitem \large Thesis co-director:\hfill------------------------------------}
-\IEEEcompsocitemizethanks{\IEEEcompsocthanksitem \footnotesize (if applicable)\hfill  (name and signature)\hspace{1.5cm}\hfill}
+%\IEEEcompsocitemizethanks{\IEEEcompsocthanksitem \large}
+%\IEEEcompsocitemizethanks{\IEEEcompsocthanksitem \large}
+%\IEEEcompsocitemizethanks{\IEEEcompsocthanksitem \large Thesis co-director:\hfill------------------------------------}
+%\IEEEcompsocitemizethanks{\IEEEcompsocthanksitem \footnotesize (if applicable)\hfill  (name and signature)\hspace{1.5cm}\hfill}
 \IEEEcompsocitemizethanks{\IEEEcompsocthanksitem \large}
@@
 \markboth{EDIC Research Proposal}%
-{Shell \MakeLowercase{\textit{et al.}}: EDIC Research Proposal}
+{Amin \MakeLowercase{\textit{et al.}}: EDIC Research Proposal}
 % The only time the second header will appear is for the odd numbered pages
@@
 %\boldmath
+
+% TODO
 About 60 to 100 words describing what this 4--8 page write-up is about. Start
@@
 \begin{IEEEkeywords}
-thesis proposal, candidacy exam write-up, EDIC, EPFL
+calculus, objects, dependent types, scala
 \end{IEEEkeywords}}
@@
 % and "HIS" in caps to complete the first word.
+
+% TODO
+
 \IEEEPARstart{T}{his} template is not trying to suggest a specific

### 2012-08-27 doc/candidacy/report.tex
todos

@@
 % TODO
-About 60 to 100 words describing what this 4--8 page write-up is about. Start
-with an abstract that clearly identifies the problem, the contribution of
-the paper, and the results (if any) or ongoing research. 
-This is not a whole lot of words.  This abstract is
-48 words.
+
+%About 60 to 100 words describing what this 4--8 page write-up is about. Start
+%with an abstract that clearly identifies the problem, the contribution of
+%the paper, and the results (if any) or ongoing research. 
+%This is not a whole lot of words.  This abstract is
+%48 words.
 
@@
 
-\IEEEPARstart{T}{his} template is not trying to suggest a specific
-style for the content of your candidacy exam/thesis proposal paper; it
-only concerns the style of the writeup itself.  This template is in LaTeX.
-We encourage you to use it for your write-up.
-
-If you insist on using a different word processing program please
-match the style as much as possible.  Here is the basic format: The
-paper size is A4, the font is 10 point, 
-single-spacing, ``times roman like.''
-Two columns, each 8.75~cm wide, with 1.5~cm left/right margins.
-Sections and sub-sections have bold numbered headings.  Figures have
-captions underneath, with a bold label and number, and also a short
-caption sentence.  Ditto for tables.  The text itself is a mix of standard
-paragraphs, bullets, numbered lists, and equations where appropriate.
-References are at the end, in a numbered format, in a slightly smaller
-font.
-
-The overall length of this write-up should be 4-8 pages.
-
-\subsection{Purpose of this Write-Up}
+%\IEEEPARstart{T}{his} template is not trying to suggest a specific
+%style for the content of your candidacy exam/thesis proposal paper; it
+%only concerns the style of the writeup itself.  This template is in LaTeX.
+%We encourage you to use it for your write-up.
+
+%If you insist on using a different word processing program please
+%match the style as much as possible.  Here is the basic format: The
+%paper size is A4, the font is 10 point, 
+%single-spacing, ``times roman like.''
+%Two columns, each 8.75~cm wide, with 1.5~cm left/right margins.
+%Sections and sub-sections have bold numbered headings.  Figures have
+%captions underneath, with a bold label and number, and also a short
+%caption sentence.  Ditto for tables.  The text itself is a mix of standard
+%paragraphs, bullets, numbered lists, and equations where appropriate.
+%References are at the end, in a numbered format, in a slightly smaller
+%font.
+
+%The overall length of this write-up should be 4-8 pages.
+
+%\subsection{Purpose of this Write-Up}
 % needed in second column of first page if using \IEEEpubid
 %\IEEEpubidadjcol
-This write-up serves two purposes. First, it forms the basis for
-your candidacy exam. As such you should summarize the three papers selected
-by your advisor and yourself, and analyze as well as discuss them critically.
-Second, the write-up 
-is also your thesis proposal.  Therefore, the last one or two pages should
-be dedicated to your own preliminary work. A road-map of how you plan to
-advance the state of the art in your chosen area should also be given.
-For further details please consult the document ``PhD Candidacy Exam Overview.'' You can find the latest
-version at {\tt http://phd.epfl.ch/page57746-en.html}.
-
-\subsection{What to Put into the Introduction}
-Describe briefly the context, the problem, shortcomings in prior
-approaches, and your proposed approach and solution. Forecast results.
+%This write-up serves two purposes. First, it forms the basis for
+%your candidacy exam. As such you should summarize the three papers selected
+%by your advisor and yourself, and analyze as well as discuss them critically.
+%Second, the write-up 
+%is also your thesis proposal.  Therefore, the last one or two pages should
+%be dedicated to your own preliminary work. A road-map of how you plan to
+%advance the state of the art in your chosen area should also be given.
+%For further details please consult the document ``PhD Candidacy Exam Overview.'' You can find the latest
+%version at {\tt http://phd.epfl.ch/page57746-en.html}.
+
+%\subsection{What to Put into the Introduction}
+%Describe briefly the context, the problem, shortcomings in prior
+%approaches, and your proposed approach and solution. Forecast results.
 
@@
 
-\section{Survey of the Selected Papers: The Giants upon whose shoulders I will stand}
-Background --- Describe the three papers in detail, the problem they
-tackle, the solutions and results, and their shortcomings, and how they
-relate to your work.  This part builds the basis for the oral candidacy exam.
-
-\section{Research Proposal: How I Plan to Revolutionize the World}
-Describe your own work and include a  summary and references. Write how you
-propose to advance the state of the art given the background. What
-is new technically? How does it improve over prior work?
-Summarize, suggest an approximate timeline, and list references.
-
-\vspace{1cm}
-This write-up should be 4-8 pages. \\
-The quick brown fox jumps over the lazy dog.\\
-The quick brown fox jumps over the lazy dog.\\
-The quick brown fox jumps over the lazy dog.\\
-The quick brown fox jumps over the lazy dog.\\
-The quick brown fox jumps over the lazy dog.\\
-The quick brown fox jumps over the lazy dog.\\
-The quick brown fox jumps over the lazy dog.\\
-The quick brown fox jumps over the lazy dog.\\
-The quick brown fox jumps over the lazy dog.\\
-The quick brown fox jumps over the lazy dog.\\
-The quick brown fox jumps over the lazy dog.\\
-The quick brown fox jumps over the lazy dog.\\
-The quick brown fox jumps over the lazy dog.\\
-The quick brown fox jumps over the lazy dog.\\
-The quick brown fox jumps over the lazy dog.\\
-The quick brown fox jumps over the lazy dog.\\
-The quick brown fox jumps over the lazy dog.\\
-The quick brown fox jumps over the lazy dog.\\
-The quick brown fox jumps over the lazy dog.\\
-The quick brown fox jumps over the lazy dog.\\
-The quick brown fox jumps over the lazy dog.\\
-The quick brown fox jumps over the lazy dog.\\
-The quick brown fox jumps over the lazy dog.\\
-The quick brown fox jumps over the lazy dog.\\
-The quick brown fox jumps over the lazy dog.\\
-The quick brown fox jumps over the lazy dog.\\
-The quick brown fox jumps over the lazy dog.\\
-The quick brown fox jumps over the lazy dog.\\
-The quick brown fox jumps over the lazy dog.\\
-The quick brown fox jumps over the lazy dog.\\
-The quick brown fox jumps over the lazy dog.\\
-The quick brown fox jumps over the lazy dog.\\
-The quick brown fox jumps over the lazy dog.\\
-The quick brown fox jumps over the lazy dog.\\
-The quick brown fox jumps over the lazy dog.\\
-The quick brown fox jumps over the lazy dog.\\
-The quick brown fox jumps over the lazy dog.\\
-The quick brown fox jumps over the lazy dog.\\
-The quick brown fox jumps over the lazy dog.\\
-The quick brown fox jumps over the lazy dog.\\
-The quick brown fox jumps over the lazy dog.\\
-The quick brown fox jumps over the lazy dog.\\
-The quick brown fox jumps over the lazy dog.\\
-The quick brown fox jumps over the lazy dog.\\
-The quick brown fox jumps over the lazy dog.\\
-The quick brown fox jumps over the lazy dog.\\
-The quick brown fox jumps over the lazy dog.\\
-The quick brown fox jumps over the lazy dog.\\
-The quick brown fox jumps over the lazy dog.\\
-The quick brown fox jumps over the lazy dog.\\
-The quick brown fox jumps over the lazy dog.\\
-The quick brown fox jumps over the lazy dog.\\
-The quick brown fox jumps over the lazy dog.\\
-The quick brown fox jumps over the lazy dog.\\
-The quick brown fox jumps over the lazy dog.\\
-The quick brown fox jumps over the lazy dog.\\
-The quick brown fox jumps over the lazy dog.\\
-The quick brown fox jumps over the lazy dog.\\
-The quick brown fox jumps over the lazy dog.\\
-The quick brown fox jumps over the lazy dog.\\
-The quick brown fox jumps over the lazy dog.\\
-The quick brown fox jumps over the lazy dog.\\
-The quick brown fox jumps over the lazy dog.\\
-The quick brown fox jumps over the lazy dog.\\
-The quick brown fox jumps over the lazy dog.\\
-The quick brown fox jumps over the lazy dog.\\
-The quick brown fox jumps over the lazy dog.\\
-The quick brown fox jumps over the lazy dog.\\
-The quick brown fox jumps over the lazy dog.\\
-The quick brown fox jumps over the lazy dog.\\
-The quick brown fox jumps over the lazy dog.\\
-The quick brown fox jumps over the lazy dog.\\
-The quick brown fox jumps over the lazy dog.\\
-The quick brown fox jumps over the lazy dog.\\
-The quick brown fox jumps over the lazy dog.\\
-The quick brown fox jumps over the lazy dog.\\
-The quick brown fox jumps over the lazy dog.\\
-The quick brown fox jumps over the lazy dog.\\
-The quick brown fox jumps over the lazy dog.\\
-The quick brown fox jumps over the lazy dog.\\
-The quick brown fox jumps over the lazy dog.\\
-The quick brown fox jumps over the lazy dog.\\
-The quick brown fox jumps over the lazy dog.\\
-The quick brown fox jumps over the lazy dog.\\
-The quick brown fox jumps over the lazy dog.\\
-The quick brown fox jumps over the lazy dog.\\
-The quick brown fox jumps over the lazy dog.\\
-The quick brown fox jumps over the lazy dog.\\
-The quick brown fox jumps over the lazy dog.\\
-The quick brown fox jumps over the lazy dog.\\
-The quick brown fox jumps over the lazy dog.\\
-The quick brown fox jumps over the lazy dog.\\
-The quick brown fox jumps over the lazy dog.\\
-The quick brown fox jumps over the lazy dog.\\
-The quick brown fox jumps over the lazy dog.\\
-The quick brown fox jumps over the lazy dog.\\
-The quick brown fox jumps over the lazy dog.\\
-The quick brown fox jumps over the lazy dog.\\
-The quick brown fox jumps over the lazy dog.\\
-The quick brown fox jumps over the lazy dog.\\
-The quick brown fox jumps over the lazy dog.\\
-The quick brown fox jumps over the lazy dog.\\
-The quick brown fox jumps over the lazy dog.\\
-The quick brown fox jumps over the lazy dog.\\
-The quick brown fox jumps over the lazy dog.\\
-The quick brown fox jumps over the lazy dog.\\
-The quick brown fox jumps over the lazy dog.\\
-The quick brown fox jumps over the lazy dog.\\
-The quick brown fox jumps over the lazy dog.\\
-The quick brown fox jumps over the lazy dog.\\
-The quick brown fox jumps over the lazy dog.\\
-The quick brown fox jumps over the lazy dog.\\
-The quick brown fox jumps over the lazy dog.\\
-The quick brown fox jumps over the lazy dog.\\
-The quick brown fox jumps over the lazy dog.\\
-The quick brown fox jumps over the lazy dog.\\
-The quick brown fox jumps over the lazy dog.\\
-The quick brown fox jumps over the lazy dog.\\
-The quick brown fox jumps over the lazy dog.\\
-The quick brown fox jumps over the lazy dog.\\
-The quick brown fox jumps over the lazy dog.\\
-The quick brown fox jumps over the lazy dog.\\
-The quick brown fox jumps over the lazy dog.\\
-The quick brown fox jumps over the lazy dog.\\
-The quick brown fox jumps over the lazy dog.\\
-The quick brown fox jumps over the lazy dog.\\
-The quick brown fox jumps over the lazy dog.\\
-The quick brown fox jumps over the lazy dog.\\
-The quick brown fox jumps over the lazy dog.\\
-The quick brown fox jumps over the lazy dog.\\
-The quick brown fox jumps over the lazy dog.\\
-The quick brown fox jumps over the lazy dog.\\
-The quick brown fox jumps over the lazy dog.\\
-The quick brown fox jumps over the lazy dog.\\
-The quick brown fox jumps over the lazy dog.\\
-The quick brown fox jumps over the lazy dog.\\
-The quick brown fox jumps over the lazy dog.\\
-The quick brown fox jumps over the lazy dog.\\
-The quick brown fox jumps over the lazy dog.\\
-The quick brown fox jumps over the lazy dog.\\
-The quick brown fox jumps over the lazy dog.\\
-The quick brown fox jumps over the lazy dog.\\
-The quick brown fox jumps over the lazy dog.\\
-The quick brown fox jumps over the lazy dog.\\
-The quick brown fox jumps over the lazy dog.\\
-The quick brown fox jumps over the lazy dog.\\
-The quick brown fox jumps over the lazy dog.\\
-The quick brown fox jumps over the lazy dog.\\
-The quick brown fox jumps over the lazy dog.\\
-The quick brown fox jumps over the lazy dog.\\
-The quick brown fox jumps over the lazy dog.\\
-The quick brown fox jumps over the lazy dog.\\
-The quick brown fox jumps over the lazy dog.\\
-The quick brown fox jumps over the lazy dog.\\
-The quick brown fox jumps over the lazy dog.\\
-The quick brown fox jumps over the lazy dog.\\
-The quick brown fox jumps over the lazy dog.\\
-The quick brown fox jumps over the lazy dog.\\
-The quick brown fox jumps over the lazy dog.\\
-The quick brown fox jumps over the lazy dog.\\
-The quick brown fox jumps over the lazy dog.\\
-The quick brown fox jumps over the lazy dog.\\
-The quick brown fox jumps over the lazy dog.\\
-The quick brown fox jumps over the lazy dog.\\
-The quick brown fox jumps over the lazy dog.\\
-The quick brown fox jumps over the lazy dog.\\
-The quick brown fox jumps over the lazy dog.\\
-The quick brown fox jumps over the lazy dog.\\
-The quick brown fox jumps over the lazy dog.\\
-The quick brown fox jumps over the lazy dog.\\
-The quick brown fox jumps over the lazy dog.\\
-The quick brown fox jumps over the lazy dog.\\
-The quick brown fox jumps over the lazy dog.\\
-The quick brown fox jumps over the lazy dog.\\
-The quick brown fox jumps over the lazy dog.\\
-The quick brown fox jumps over the lazy dog.\\
-The quick brown fox jumps over the lazy dog.\\
-The quick brown fox jumps over the lazy dog.\\
-The quick brown fox jumps over the lazy dog.\\
-The quick brown fox jumps over the lazy dog.\\
-The quick brown fox jumps over the lazy dog.\\
-The quick brown fox jumps over the lazy dog.\\
-The quick brown fox jumps over the lazy dog.\\
-The quick brown fox jumps over the lazy dog.\\
-The quick brown fox jumps over the lazy dog.\\
-The quick brown fox jumps over the lazy dog.\\
-The quick brown fox jumps over the lazy dog.\\
-The quick brown fox jumps over the lazy dog.\\
-The quick brown fox jumps over the lazy dog.\\
-The quick brown fox jumps over the lazy dog.\\
-The quick brown fox jumps over the lazy dog.\\
-The quick brown fox jumps over the lazy dog.\\
-The quick brown fox jumps over the lazy dog.\\
-The quick brown fox jumps over the lazy dog.\\
-The quick brown fox jumps over the lazy dog.\\
-The quick brown fox jumps over the lazy dog.\\
-The quick brown fox jumps over the lazy dog.\\
-The quick brown fox jumps over the lazy dog.\\
-The quick brown fox jumps over the lazy dog.\\
-The quick brown fox jumps over the lazy dog.\\
-The quick brown fox jumps over the lazy dog.\\
-The quick brown fox jumps over the lazy dog.\\
-The quick brown fox jumps over the lazy dog.\\
-The quick brown fox jumps over the lazy dog.\\
-The quick brown fox jumps over the lazy dog.\\
-The quick brown fox jumps over the lazy dog.\\
-The quick brown fox jumps over the lazy dog.\\
-The quick brown fox jumps over the lazy dog.\\
-The quick brown fox jumps over the lazy dog.\\
-The quick brown fox jumps over the lazy dog.\\
-The quick brown fox jumps over the lazy dog.\\
-The quick brown fox jumps over the lazy dog.\\
-The quick brown fox jumps over the lazy dog.\\
-The quick brown fox jumps over the lazy dog.\\
-The quick brown fox jumps over the lazy dog.\\
-The quick brown fox jumps over the lazy dog.\\
-The quick brown fox jumps over the lazy dog.\\
-The quick brown fox jumps over the lazy dog.\\
-The quick brown fox jumps over the lazy dog.\\
-The quick brown fox jumps over the lazy dog.\\
-The quick brown fox jumps over the lazy dog.\\
-The quick brown fox jumps over the lazy dog.\\
-The quick brown fox jumps over the lazy dog.\\
-The quick brown fox jumps over the lazy dog.\\
-The quick brown fox jumps over the lazy dog.\\
-The quick brown fox jumps over the lazy dog.\\
-The quick brown fox jumps over the lazy dog.\\
-The quick brown fox jumps over the lazy dog.\\
-The quick brown fox jumps over the lazy dog.\\
-The quick brown fox jumps over the lazy dog.\\
-The quick brown fox jumps over the lazy dog.\\
-The quick brown fox jumps over the lazy dog.\\
-The quick brown fox jumps over the lazy dog.\\
-The quick brown fox jumps over the lazy dog.\\
-The quick brown fox jumps over the lazy dog.\\
-The quick brown fox jumps over the lazy dog.\\
-The quick brown fox jumps over the lazy dog.\\
-The quick brown fox jumps over the lazy dog.\\
-The quick brown fox jumps over the lazy dog.\\
-The quick brown fox jumps over the lazy dog.\\
-The quick brown fox jumps over the lazy dog.\\
-The quick brown fox jumps over the lazy dog.\\
-The quick brown fox jumps over the lazy dog.\\
-The quick brown fox jumps over the lazy dog.\\
-The quick brown fox jumps over the lazy dog.\\
-The quick brown fox jumps over the lazy dog.\\
-The quick brown fox jumps over the lazy dog.\\
-The quick brown fox jumps over the lazy dog.\\
-The quick brown fox jumps over the lazy dog.\\
-The quick brown fox jumps over the lazy dog.\\
-The quick brown fox jumps over the lazy dog.\\
-The quick brown fox jumps over the lazy dog.\\
-The quick brown fox jumps over the lazy dog.\\
-The quick brown fox jumps over the lazy dog.\\
-The quick brown fox jumps over the lazy dog.\\
-The quick brown fox jumps over the lazy dog.\\
-The quick brown fox jumps over the lazy dog.\\
-The quick brown fox jumps over the lazy dog.\\
-The quick brown fox jumps over the lazy dog.\\
-The quick brown fox jumps over the lazy dog.\\
-The quick brown fox jumps over the lazy dog.\\
-The quick brown fox jumps over the lazy dog.\\
-The quick brown fox jumps over the lazy dog.\\
-The quick brown fox jumps over the lazy dog.\\
-The quick brown fox jumps over the lazy dog.\\
-The quick brown fox jumps over the lazy dog.\\
-The quick brown fox jumps over the lazy dog.\\
-The quick brown fox jumps over the lazy dog.\\
-The quick brown fox jumps over the lazy dog.\\
-The quick brown fox jumps over the lazy dog.\\
-The quick brown fox jumps over the lazy dog.\\
-The quick brown fox jumps over the lazy dog.\\
-The quick brown fox jumps over the lazy dog.\\
-The quick brown fox jumps over the lazy dog.\\
-The quick brown fox jumps over the lazy dog.\\
-The quick brown fox jumps over the lazy dog.\\
-The quick brown fox jumps over the lazy dog.\\
-The quick brown fox jumps over the lazy dog.\\
-The quick brown fox jumps over the lazy dog.\\
-The quick brown fox jumps over the lazy dog.\\
-The quick brown fox jumps over the lazy dog.\\
-The quick brown fox jumps over the lazy dog.\\
-The quick brown fox jumps over the lazy dog.\\
-The quick brown fox jumps over the lazy dog.\\
-The quick brown fox jumps over the lazy dog.\\
-The quick brown fox jumps over the lazy dog.\\
-The quick brown fox jumps over the lazy dog.\\
-The quick brown fox jumps over the lazy dog.\\
-The quick brown fox jumps over the lazy dog.\\
-The quick brown fox jumps over the lazy dog.\\
-The quick brown fox jumps over the lazy dog.\\
-The quick brown fox jumps over the lazy dog.\\
-The quick brown fox jumps over the lazy dog.\\
-The quick brown fox jumps over the lazy dog.\\
-The quick brown fox jumps over the lazy dog.\\
-The quick brown fox jumps over the lazy dog.\\
-The quick brown fox jumps over the lazy dog.\\
-The quick brown fox jumps over the lazy dog.\\
-The quick brown fox jumps over the lazy dog.\\
-The quick brown fox jumps over the lazy dog.\\
-The quick brown fox jumps over the lazy dog.\\
-The quick brown fox jumps over the lazy dog.\\
-The quick brown fox jumps over the lazy dog.\\
-The quick brown fox jumps over the lazy dog.\\
-The quick brown fox jumps over the lazy dog.\\
-The quick brown fox jumps over the lazy dog.\\
-The quick brown fox jumps over the lazy dog.\\
-The quick brown fox jumps over the lazy dog.\\
-The quick brown fox jumps over the lazy dog.\\
-The quick brown fox jumps over the lazy dog.\\
-The quick brown fox jumps over the lazy dog.\\
-The quick brown fox jumps over the lazy dog.\\
-The quick brown fox jumps over the lazy dog.\\
-The quick brown fox jumps over the lazy dog.\\
-The quick brown fox jumps over the lazy dog.\\
-The quick brown fox jumps over the lazy dog.\\
-The quick brown fox jumps over the lazy dog.\\
-The quick brown fox jumps over the lazy dog.\\
-The quick brown fox jumps over the lazy dog.\\
-The quick brown fox jumps over the lazy dog.\\
-The quick brown fox jumps over the lazy dog.\\
-The quick brown fox jumps over the lazy dog.\\
-The quick brown fox jumps over the lazy dog.\\
-The quick brown fox jumps over the lazy dog.\\
-The quick brown fox jumps over the lazy dog.\\
-The quick brown fox jumps over the lazy dog.\\
-The quick brown fox jumps over the lazy dog.\\
-The quick brown fox jumps over the lazy dog.\\
-The quick brown fox jumps over the lazy dog.\\
-The quick brown fox jumps over the lazy dog.\\
-The quick brown fox jumps over the lazy dog.\\
-The quick brown fox jumps over the lazy dog.\\
-The quick brown fox jumps over the lazy dog.\\
-The quick brown fox jumps over the lazy dog.\\
-The quick brown fox jumps over the lazy dog.\\
-The quick brown fox jumps over the lazy dog.\\
-The quick brown fox jumps over the lazy dog.\\
-The quick brown fox jumps over the lazy dog.\\
-The quick brown fox jumps over the lazy dog.\\
-The quick brown fox jumps over the lazy dog.\\
-The quick brown fox jumps over the lazy dog.\\
-The quick brown fox jumps over the lazy dog.\\
-The quick brown fox jumps over the lazy dog.\\
-The quick brown fox jumps over the lazy dog.\\
-The quick brown fox jumps over the lazy dog.\\
-The quick brown fox jumps over the lazy dog.\\
-The quick brown fox jumps over the lazy dog.\\
-The quick brown fox jumps over the lazy dog.\\
+\section{Survey of the Selected Papers}%: The Giants upon whose shoulders I will stand
+% TODO
+%Background --- Describe the three papers in detail, the problem they
+%tackle, the solutions and results, and their shortcomings, and how they
+%relate to your work.  This part builds the basis for the oral candidacy exam.
+
+\section{Research Proposal} %: How I Plan to Revolutionize the World
+% TODO
+%Describe your own work and include a  summary and references. Write how you
+%propose to advance the state of the art given the background. What
+%is new technically? How does it improve over prior work?
+%Summarize, suggest an approximate timeline, and list references.
 

### 2012-08-27 doc/candidacy/report.tex
bib

@@
 \section{Survey of the Selected Papers}%: The Giants upon whose shoulders I will stand
+
+I present three papers~\cite{nuObj,tribe,homl}, each of which models a
+calculus with path-dependent types.
+
 % TODO
@@
 % http://www.michaelshell.org/tex/ieeetran/bibtex/
-%\bibliographystyle{IEEEtran}
+\bibliographystyle{IEEEtran}
 % argument is your BibTeX string definitions and bibliography database(s)
-%\bibliography{IEEEabrv,../bib/paper}
+\bibliography{IEEEabrv,dot}
 %
@@
 % (used to reserve space for the reference number labels box)
-\begin{thebibliography}{1}
-
-\bibitem{mittelbach}
-F.~Mittelbach, M.~Goossens, J.~Braams, D.~Carlisle, and C.~Rowley, 
-\emph{The {\LaTeX}} Companion, 2nd~ed.\hskip 1em plus
-0.5em minus 0.4em\relax Addison-Wesley Professional, 2004.
-
-\end{thebibliography}
+%\begin{thebibliography}{1}
+
+%\bibitem{mittelbach}
+%F.~Mittelbach, M.~Goossens, J.~Braams, D.~Carlisle, and C.~Rowley, 
+%\emph{The {\LaTeX}} Companion, 2nd~ed.\hskip 1em plus
+%0.5em minus 0.4em\relax Addison-Wesley Professional, 2004.
+
+%\end{thebibliography}
 

### 2012-08-29 doc/candidacy/report.tex
intro to ml paper

@@
 
+\subsection{A Type-Theoretic Approach to Higher-Order Modules with Sharing}
+
+In~\cite{homl}, Harper and Lillibridge present a calculus for
+first-class modules, with complete control over the propagation of
+compile-time information.
+
+Their module system is inspired by Standard ML (SML). The SML module
+system has three constructs: {\it signatures}, {\it structures}, and
+{\it functors}. A {\it structure} is a module: it defines a collection
+of types, values and (sub)structures packaged as a program unit. A
+{\it signature} is an interface, the ``type'' of a structure: it
+describes the types, values and (sub)structures of a structure by
+giving their kinds, types and interfaces. A {\it functor} is a
+function mapping structures to structures. SML has a notion of ``type
+sharing'' which allows stating that types specified in separate
+modules are actually equal.
+
+Their module system addresses some shortcomings of the SML module
+system. Though the SML module system is higher-order, because SML
+provides functors and substructures, it still treats modules as
+second-class. For example, it isn't possible to return a structure
+from an if statement. In fact, allowing this might require obscuring
+the visibility of the resulting structure to ensure soundness. Another
+limitation of the SML module system is that ``type sharing'' is
+limited to equality between type names, not general type
+expressions. In effect, this means that the transparency / opaqueness
+of a signature cannot be fully fine-tuned. Their proposal introduces
+translucent signatures, which allows any type to be partially
+determined by a type expression.
+
+%Scala's abstract types have close
+%resemblances to abstract types of signatures in the module systems of
+%ML.
+
+\subsection{A Nominal Theory of Objects with Dependent Types~\cite{nuObj}}
+
+\subsection{Tribe: a simple virtual class calculus~\cite{tribe}}
+
 \section{Research Proposal} %: How I Plan to Revolutionize the World

### 2012-08-29 doc/candidacy/report.tex
first pass on ml paper

@@
 of a signature cannot be fully fine-tuned. Their proposal introduces
-translucent signatures, which allows any type to be partially
+translucent signatures, which allows any type to be partially or fully
 determined by a type expression.
 
-%Scala's abstract types have close
-%resemblances to abstract types of signatures in the module systems of
-%ML.
+Scala's abstract types have close resemblances to abstract types of
+signatures in these ML-like module systems. % TODO: develop
+
+Their calculus is grounded in type theory. It is based on Girard's
+$F_\omega$~\cite{tapl}. They add {\it translucent sums} to model
+modules, dependent functions to model functors, a notion of subtyping
+to model module implementation-interface matching. The crucial
+addition and novelty is that of a {\it translucent sum}. A translucent
+sum has the form of a possibly empty sequence of bindings. Similarly,
+a translucent sum type has the form of a possibly empty sequence of
+declarations. They differ from traditional records, because they can
+contain types and because later fields can depend on ealier
+ones. Translucent sums are ordinary values. Since they model modules,
+it means that modules are first-class values. There's a form of
+dependent type, since translucent sums are expressions that may
+contain types.
+
+In terms of theoretical results, Lillibridge shows in his thesis that
+a simplified kernel system based on this calculus is sound, even in
+the presence of side effects~\cite{lillibridge:thesis}. In the
+appendix of the paper, they show that subtyping, and hence type
+checking, is undecidable.
 

### 2012-08-29 doc/candidacy/report.tex
minor

@@
 
-Scala's abstract types have close resemblances to abstract types of
-signatures in these ML-like module systems. % TODO: develop
-
 Their calculus is grounded in type theory. It is based on Girard's
@@
 ones. Translucent sums are ordinary values. Since they model modules,
-it means that modules are first-class values. There's a form of
-dependent type, since translucent sums are expressions that may
-contain types.
+it means that modules are first-class values.
+
+Scala's abstract types have close resemblances to abstract types of
+signatures in these ML-like module systems. In particular, in their
+calculus, there's a form of dependent type, since translucent sums are
+expressions that may contain types. Compared to path-dependent types
+as found in Scala, these are more limited because recursive references
+are not allowed.
 
@@
 
-\subsection{A Nominal Theory of Objects with Dependent Types~\cite{nuObj}}
-
-\subsection{Tribe: a simple virtual class calculus~\cite{tribe}}
+\subsection{Tribe: a simple virtual class calculus}
+
+\subsection{A Nominal Theory of Objects with Dependent Types}
 

### 2012-08-29 doc/candidacy/report.tex
minor

@@
 ones. Translucent sums are ordinary values. Since they model modules,
-it means that modules are first-class values.
+it means that modules are first-class values, and their module system
+is higher-order ``for free''.
 
@@
 \subsection{Tribe: a simple virtual class calculus}
+
+In terms of theoretical results, there's a sketch of soundness in the
+paper, but it's unclear whether it is correct. Decidability and
+completeness of subtyping has not yet been established.
 

### 2012-08-29 doc/candidacy/report.tex
beg of tribe paper

@@
 
+In~\cite{tribe}, Clarke et al. present a calculus for virtual
+classes. Virtual classes allows {\it families} of classes to be
+inherited, rather than just single classes. Classes are lexically
+nested inside other classes. When a class is inherited, its nested
+inner classes are inherited along with its methods and fields. {\it
+  Further binding} enables the nested classes to be further extended.
+
+Though Scala doesn't have virtual classes, its abstract types enables
+it to emulate some (but not all) of the benefits of virtual classes as
+presented in the Tribe calculus.
+
 In terms of theoretical results, there's a sketch of soundness in the

### 2012-08-30 doc/candidacy/report.tex
path-dep types in tribe
rev

@@
 as found in Scala, these are more limited because recursive references
-are not allowed.
+and bounded quantifications are not allowed.
 
@@
 In~\cite{tribe}, Clarke et al. present a calculus for virtual
-classes. Virtual classes allows {\it families} of classes to be
+classes. Virtual classes allow {\it families} of classes to be
 inherited, rather than just single classes. Classes are lexically
@@
 inner classes are inherited along with its methods and fields. {\it
-  Further binding} enables the nested classes to be further extended.
+  Further binding} enables the nested classes to be further
+extended. Path-dependent types arise naturally to distinguish objects
+coming from different families. In Tribe, paths can additionally use
+an ``out'' field to move from an object to the object which surrounds
+it.
+
+%Tribe, like Scala, distinguishes object families by using
+%path-dependent types. An {\it object family type} such as {\code
+%  x.f.C} refers to class {\code C} which is nested within the object
+%at the end of path {\code x.f}. Paths have to be invariant, and so
+%consists of chains of immutable fields or variables.
 
@@
 it to emulate some (but not all) of the benefits of virtual classes as
-presented in the Tribe calculus.
+presented in the Tribe calculus. On the other hand, Tribe doesn't have
+virtual types. In a follow-up paper~\cite{tribalo}, an improved
+variant of Tribe with simpler types, cross-family inheritance and
+generics is presented and a formalisation of ownership types built
+upon it.
 
 In terms of theoretical results, there's a sketch of soundness in the
-paper, but it's unclear whether it is correct. Decidability and
-completeness of subtyping has not yet been established.
+paper, but it's unclear whether it is correct. In particular, the
+substitution lemma does not seem valid when null is the value to be
+substituted and the expression has a type which contains the variable
+to be substituted. In the follow-up paper~\cite{tribalo}, the
+soundness proof sketch seems more plausible but then is tied to the
+ownership system formalism. Decidability and completeness of subtyping
+has not yet been established.
 

### 2012-08-30 doc/candidacy/report.tex
a bit more about tribe

@@
 extended. Path-dependent types arise naturally to distinguish objects
-coming from different families. In Tribe, paths can additionally use
-an ``out'' field to move from an object to the object which surrounds
-it.
+coming from different families. In Tribe, path types can depend
+simultaneously on both classes and objects. Additionally, paths can
+use an ``out'' field to move from an object to the object which
+surrounds it. This ``out'' feature enables ubiquitous access to an
+object's family without the need to drag around family arguments.
 

### 2012-08-30 doc/candidacy/report.tex
typo

@@
 
-Though Scala doesn't have virtual classes, its abstract types enables
+Though Scala doesn't have virtual classes, its abstract types enable
 it to emulate some (but not all) of the benefits of virtual classes as

### 2012-08-30 doc/candidacy/report.tex
blurb about nuObj

@@
 
+In~\cite{nuObj}, Odersky et al. presents ${\nu}Obj$, a calculus for
+classes and objects which can have types as members. It proposes three
+kinds of type members: aliases, abstract types, and new types. The
+calculus can encode Java's inner classes, virtual types, family
+polymorphism, essential aspects of the ML-like module systems
+including sharing consraints and higher-order functors, and System
+$F_{<:}$~\cite{tapl}. The calculus can be used as a basis for unifying
+concepts that so far existed in parallel in advanced object systems
+and module systems. ${\nu}Obj$ makes the following connections: Object =
+Module, Object type = Signature, Class = Method = Functor.
+
+In terms of theoretical results, the paper confirms type soundness of
+the calculus and the undecidability of type checking.
+
 \section{Research Proposal} %: How I Plan to Revolutionize the World

### 2012-08-30 doc/candidacy/report.tex
intro

@@
 
-% TODO
+% TODO intro
+
+\IEEEPARstart{T}{he} aim of this research is to propose a new sound
+and tractable type-theoretic foundation of Scala and languages like
+it: the Dependent Object Types calculus (DOT). The properties we are
+interested in modeling are Scala's path-dependent types and abstract
+type members, as well as its mixture of nominal and structural typing
+through the use of refinement types. Compared to previous
+approaches~\cite{nuObj,FS}, we make no attempt to model inheritance or
+mixing composition. Furthermore, the calculus is more normative than
+descriptive, in that it does not precisely describe what's currently
+in Scala.
+
+Path-dependent types are a key concept in the DOT calculus, and a
+unifying concept of the three papers we choose to study. A {\it
+  dependent type} is a type which depends on a term. A {\it
+  path-dependent type} is a limited form of dependent type, in which a
+type depends on a path: a chain of immutable fields or
+variables. Path-dependent types arise when terms, usually objects, can
+have types as members.
 
@@
 
-I present three papers~\cite{nuObj,tribe,homl}, each of which models a
+We present three papers~\cite{nuObj,tribe,homl}, each of which models a
 calculus with path-dependent types.

### 2012-08-30 doc/candidacy/macros.tex
macros for displaying Scala code

--- 2012-08-10 article.tex
@@
-\documentclass[9pt]{sigplanconf}
-
-% The following \documentclass options may be useful:
-%
-% 10pt          To set in 10-point type instead of 9-point.
-% 11pt          To set in 11-point type instead of 9-point.
-% authoryear    To obtain author/year citation style instead of numeric.
-
-\usepackage{amsmath}
-\usepackage{amssymb}
-
-\usepackage{fleqn}
 \usepackage{listings}
-\usepackage{math}
-\usepackage{amsmath}
-\usepackage{latexsym}
-\usepackage{bcprules}
-\usepackage[scaled=0.848971]{luximono} % This is for 11 pt Default font
-\usepackage[T1]{fontenc}
-
-% Prooftree formatting
-\usepackage{prooftree}
-
-\usepackage{multicol}
-\usepackage{framed}
-
-\usepackage{stmaryrd}
-
-%\usepackage{float}
-%\floatstyle{boxed} 
-%\restylefloat{figure}
-
-% support for generating PDF files
-%\newif\ifpdf
-%    \ifx\pdfoutput\undefined
-%    \pdffalse
-%\else
-%    \pdftrue
-%    \pdfoutput=1
-%\fi
-
-%versions
-% Use dependent function types
-\newif\ifdep\depfalse
 
@@
 }
-
-%% set latex/pdflatex specific stuff
-%\ifpdf
-    \usepackage[pdftex,
-                hyperindex,
-                plainpages=false,
-                breaklinks,
-                colorlinks,
-                citecolor=black,
-                filecolor=black,
-                linkcolor=black,
-                pagecolor=black,
-                urlcolor=black]{hyperref}
-    \usepackage[pdftex]{graphicx}
-    \DeclareGraphicsExtensions{.jpg,.pdf}
-    \pdfcatalog {
-        /PageMode (/UseNone)
-    }
-    \usepackage{thumbpdf}
-    \usepackage[pdftex]{color}
-%\else
-%    \usepackage[ps2pdf]{hyperref}
-%    \usepackage{graphicx}
-%    \DeclareGraphicsExtensions{.eps,.jpg}
-%    \usepackage{color}
-%\fi
-
-%\setlength{\parindent}{0pt}
-%\setlength{\parskip}{5pt}
-
-% verbfilter stuff
-\newcommand{\prog}[1]{{\sl #1}}
-\newenvironment{program}[1][10.5]
-  {\fontsize{#1}{13.6}\tt\begin{tabbing}\hspace*{0.5\parindent}\=\+\kill}
-  {\end{tabbing}\noindent}
-\newcommand{\blockcomment}[1]{{\color{grayPoint3}#1}}
-\newcommand{\linecomment}{\color{grayPoint3}}
-\newcommand{\grey}{\color{grey}}
-
-%\newenvironment{program}{\ \ \ \ \begin{minipage}{\textwidth}\renewcommand{\baselinestretch}{1.0}\sl\begin{tabbing}}{\end{tabbing}\end{minipage}}
-\newcommand{\vem}{\bfseries}
-\newcommand{\quotedstring}[1]{{#1}}
-\newcommand{\typename}[1]{{#1}}
-\newcommand{\literal}[1]{{#1}}
-\newcommand{\mi}[1]{\mathit{#1}}
-
-% comments and notes
-\newcommand{\comment}[1]{}
-%\newcommand{\note}[1]{{\bf $\clubsuit$ #1 $\spadesuit$}}
-
-% figures
-\newcommand{\figurebox}[1]
-        {\fbox{\begin{minipage}{\textwidth} #1 \medskip\end{minipage}}}
-%        {\fbox{\begin{minipage}{\textwidth}\begin{center} #1 \end{center}\medskip\end{minipage}}}
-\newcommand{\boxfig}[3]
-        {\begin{figure*}\figurebox{#3\caption{\label{fig:#1}#2}}\end{figure*}}
-\newcommand{\figref}[1]
-        {Figure~\ref{fig:#1}}
-
-% typing rules (not used here)
-\newcommand{\ttag}[1]{\mbox{\textsc{\small(#1)}}}
-\newcommand{\infer}[3]{\mbox{#1 }\ba{c} #2 \\ \hline #3 \ea}
-\newcommand{\irule}[2]{{\renewcommand{\arraystretch}{1.2}\ba{c} #1 
-                        \\ \hline #2 \ea}}
-\newlength{\trulemargin}
-\newlength{\trulewidth}
-\newlength{\srulewidth}
-\setlength{\trulemargin}{0.80cm}
-\setlength{\trulewidth}{40.0mm}
-\setlength{\srulewidth}{3.0cm}
-\newenvironment{trules}{$\vspace{0.5em}\ba{p{\trulemargin}@{~}p{\trulewidth}@{~}p{\trulemargin}}}{\ea$}
-\newenvironment{srules}{$\vspace{0.5em}\ba{p{\trulemargin}@{~}p{\srulewidth}}}{\ea$}
-\newcommand{\laxiom}[2]{\ttag{#1} & $ #2 \hfill\ }
-\newcommand{\raxiom}[2]{\hfill #2 $& \hfill \ttag{#1}}
-\newcommand{\caxiom}[2]{\ttag{#1} & $\hfill #2 \hfill $& \ }
-\newcommand{\lrule}[3]{\laxiom{#1}{\irule{#2}{#3}}}
-\newcommand{\rrule}[3]{\raxiom{#1}{\irule{#2}{#3}}}
-\newcommand{\crule}[3]{\caxiom{#1}{\irule{#2}{#3}}}
-\newcommand{\lsrule}[3]{\lsaxiom{#1}{\irule{#2}{#3}}}
-\newcommand{\rsrule}[3]{\rsaxiom{#1}{\irule{#2}{#3}}}
-\newcommand{\nl}{\end{trules}\\[0.5em] \begin{trules}}
-\newcommand{\snl}{\end{srules}\\[0.5em] \begin{srules}}
-
-% commas and semicolons
-\newcommand{\comma}{,\,}
-\newcommand{\commadots}{\comma \ldots \comma}
-\newcommand{\semi}{;\mbox{;};}
-\newcommand{\semidots}{\semi \ldots \semi}
-
-% spacing
-\newcommand{\gap}{\quad\quad}
-\newcommand{\biggap}{\quad\quad\quad}
-\newcommand{\nextline}{\\ \\}
-\newcommand{\htabwidth}{0.5cm}
-\newcommand{\tabwidth}{1cm}
-\newcommand{\htab}{\hspace{\htabwidth}}
-\newcommand{\tab}{\hspace{\tabwidth}}
-\newcommand{\linesep}{\ \hrulefill \ \smallskip}
-
-% math stuff
-\newenvironment{myproof}{{\em Proof:}}{$\Box$}
-\newenvironment{proofsketch}{{\em Proof Sketch:}}{$\Box$}
-\newcommand{\Case}{{\em Case\ }}
-
-% make ; a delimiter in math mode
-% \mathcode`\;="8000 % Makes ; active in math mode
-% {\catcode`\;=\active \gdef;{\;}}
-% \mathchardef\semicolon="003B
-
-% reserved words
-\newcommand{\mathem}{\bf}
-
-% brackets
-\newcommand{\set}[1]{\{#1\}}
-\newcommand{\sbs}[1]{\lquote #1 \rquote}
-
-% arrays
-\newcommand{\ba}{\begin{array}}
-\newcommand{\ea}{\end{array}}
-\newcommand{\bda}{\[\ba}
-\newcommand{\eda}{\ea\]}
-\newcommand{\ei}{\end{array}}
-\newcommand{\bcases}{\left\{\begin{array}{ll}}
-\newcommand{\ecases}{\end{array}\right.}
-
-% \cal ids
-\renewcommand{\AA}{{\cal A}}
-\newcommand{\BB}{{\cal B}}
-\newcommand{\CC}{{\cal C}}
-\newcommand{\DD}{{\cal D}}
-\newcommand{\EE}{{\cal E}}
-\newcommand{\FF}{{\cal F}}
-\newcommand{\GG}{{\cal G}}
-\newcommand{\HH}{{\cal H}}
-\newcommand{\II}{{\cal I}}
-\newcommand{\JJ}{{\cal J}}
-\newcommand{\KK}{{\cal K}}
-\newcommand{\LL}{{\cal L}}
-\newcommand{\MM}{{\cal M}}
-\newcommand{\NN}{{\cal N}}
-\newcommand{\OO}{{\cal O}}
-\newcommand{\PP}{{\cal P}}
-\newcommand{\QQ}{{\cal Q}}
-\newcommand{\RR}{{\cal R}}
-\newcommand{\TT}{{\cal T}}
-\newcommand{\UU}{{\cal U}}
-\newcommand{\VV}{{\cal V}}
-\newcommand{\WW}{{\cal W}}
-\newcommand{\XX}{{\cal X}}
-\newcommand{\YY}{{\cal Y}}
-\newcommand{\ZZ}{{\cal Z}}
-
-% misc symbols
-\newcommand{\dhd}{\!\!\!\!\!\rightarrow}
-\newcommand{\Dhd}{\!\!\!\!\!\Rightarrow}
-\newcommand{\ts}{\,\vdash\,}
-\newcommand{\la}{\langle}
-\newcommand{\ra}{\rangle}
-\newcommand{\eg}{{\em e.g.}}
-
-% misc identifiers
-\newcommand{\dom}{\mbox{\sl dom}}
-\newcommand{\fn}{\mbox{\sl fn}}
-\newcommand{\bn}{\mbox{\sl bn}}
-\newcommand{\sig}{\mbox{\sl sig}}
-\newcommand{\IF}{\mbox{\mathem if}}
-\newcommand{\OTHERWISE}{\mbox{\mathem otherwise}}
-\newcommand{\expand}{\prec}
-\newcommand{\weakexpand}{\prec^W}
-\newcommand{\spcomma}{~,~}
-
-%\newcommand{\inst}{\mbox{\mathem inst}}
-\newcommand{\trans}[1]{\la\!\la#1\ra\!\ra}
-\newcommand{\remark}[1]{{\bf $\clubsuit$ #1 $\spadesuit$}}
-\newcommand{\todo}[1]{\remark{to do: #1}}
-%\newcommand{\J}{\justifies}
-%\newcommand{\U}{\using}
-
-% names
-\newcommand{\Scala}{\mbox{\textsc{Scala}}}
-\newcommand{\Java}{\mbox{\textsc{Java}}}
-
-%\renewcommand\textfraction{.05}
-%\renewcommand\floatpagefraction{.9}
-%\renewcommand\topfraction{.8}
-
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-%   Language abstraction commands     %
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-
-%% Relations
-% Subtype 
-\newcommand{\sub}{<:}
-% Type assignment
-\newcommand{\typ}{:}
-% reduction
-\newcommand{\reduces}{\;\rightarrow\;}
-% well-formedness
-\newcommand{\wf}{\;\mbox{\textbf{wf}}}
-\newcommand{\wfe}{\;\mbox{\textbf{wfe}}}
-\newcommand{\nswfe}{\mbox{\textbf{wfe}}}
-
-%% Operators
-% Type selection
-\newcommand{\tsel}{\#}
-% Function type
-\newcommand{\tfun}{\rightarrow}
-\newcommand{\dfun}[3]{(#1\!:\!#2) \Rightarrow #3}
-% Conjunction
-\newcommand{\tand}{\wedge}
-% Disjunction
-\newcommand{\tor}{\vee}
-% Singleton type suffix
-\newcommand{\sing}{.\textbf{type}}
-
-%% Syntax
-% Header for typing rules
-\newcommand{\judgement}[2]{{\bf #1} \hfill #2}
-% Refinement
-\newcommand{\refine}[2]{\left\{#1 \Rightarrow #2 \right\}}
-\newcommand{\mlrefine}[2]{\{#1 \Rightarrow #2 \}}
-% Field definitions
-\newcommand{\ldefs}[1]{\left\{#1\right\}}
-\newcommand{\mlldefs}[1]{\{#1\}}
-% Member sequences
-\newcommand{\seq}[1]{\overline{#1}}
-% Lambda
-\newcommand{\dabs}[3]{(#1\!:\!#2)\Rightarrow #3}
-\newcommand{\abs}[3]{\lambda #1\!:\!#2.#3}
-% Method Application
-\newcommand{\mapp}[3]{#1.#2(#3)}
-% Substitution
-\newcommand{\subst}[3]{[#1/#2]#3}
-% Object creation
-\newcommand{\new}[3]{\textbf{val }#1 = \textbf{new }#2 ;\; #3}
-\newcommand{\mlnew}[3]{\textbf{val }#1 = \textbf{new }#2 ;\;\\&#3}
-%\renewcommand{\new}[3]{#1 \leftarrow #2 \,\textbf{in}\, #3}
-% Field declaration
-\newcommand{\Ldecl}[3]{#1 : #2..#3}%{#1 \operatorname{>:} #2 \operatorname{<:} #3}
-\newcommand{\ldecl}[2]{#1 : #2}
-\newcommand{\mdecl}[3]{#1 : #2 \tfun #3}
-% Top and Bottom
-\newcommand{\Top}{\top}%{\textbf{Top}}
-\newcommand{\Bot}{\bot}%\textbf{Bot}}
-% Environment extension
-%\newcommand{\envplus}[1]{\uplus \{ #1 \}}
-\newcommand{\envplus}[1]{, #1}
-% Reduction
-\newcommand{\reduction}[4]{#1 \operatorname{|} #2 \reduces #3 \operatorname{|} #4}
-
-% Sugar
-\newcommand{\arrow}[2]{#1\rightarrow_s#2}
-\newcommand{\fun}[4]{\textbf{fun } (#1:#2)\;#3\;#4}
-\newcommand{\app}[2]{(\textbf{app }#1\;#2)}
-\newcommand{\mlapp}[2]{(\textbf{app }#1\;\\&#2)}
-\newcommand{\cast}[2]{(\textbf{cast }#1\;#2)}
-
-\newcommand{\lindent}{\hspace{-4mm}}
-
-% Logical relations
-\newcommand{\relv}[4]{\mathcal{V}_{#1;#2;#3}\llbracket#4\rrbracket}
-\newcommand{\rele}[4]{\mathcal{E}_{#1;#2;#3}\llbracket#4\rrbracket}
-\newcommand{\rels}[3]{\mathcal{\supseteq}_{#1}\llbracket#2;#3\rrbracket}
-\newcommand{\relg}[3]{\mathcal{\supseteq^!}_{#1;#2}\llbracket#3\rrbracket}
-\newcommand{\irred}[2]{\text{irred }(#1,#2)}
-\newcommand{\andl}{\;\wedge\;}
-\newcommand{\orl}{\vee}
-\newcommand{\impliesl}{\rightarrow}
-\newcommand{\reductionl}[5]{#1 \operatorname{|} #2 \;\rightarrow^{#5}\; #3 \operatorname{|} #4}
-\newcommand{\ds}{\,\vDash\,}
-
-\begin{document}
-
-\conferenceinfo{FOOL '12}{October 22, 2012, Tucson, AZ, USA.} 
-\copyrightyear{2012} 
-\copyrightdata{[to be supplied]}
-
-\title{Dependent Object Types}
-\subtitle{Towards a foundation for Scala's type system}
-
-\authorinfo{Nada Amin \and Adriaan Moors \and Martin Odersky}
-           {EPFL}
-           {first.last@epfl.ch}
-
-\maketitle
-
-\begin{abstract}
-We propose a new type-theoretic foundation of Scala and languages like
-it: the Dependent Object Types calculus (DOT). DOT models Scala's
-path-dependent types and abstract type members, as well as its mixture
-of nominal and structural typing through the use of refinement
-types. It makes no attempt to model inheritance or mixing
-composition. The calculus does not model what's currently in Scala: it
-is more normative than descriptive.
-
-We show that DOT and its patched-up variants are not syntactically
-sound, by exhibiting counterexamples to preservation. Nevertheless, we
-sketch a proof of type-safety of the calculus via step-indexed logical
-relations.
-\end{abstract}
-
-\category{D.3.3}{Language Constructs and Features}{Abstract data types, Classes and objects, polymorphism}
-\category{D.3.1}{Formal Definitions and Theory}{Syntax, Semantics}
-\category{F.3.1}{Specifying and Verifying and Reasoning about Programs}{}
-\category{F.3.3}{Studies of Program Constructs}{Object-oriented constructs, type structure}
-\category{F.3.2}{Semantics or Programming Languages}{Operational semantics}
-
-\terms
-Languages, Theory, Verification
-
-\keywords
-calculus, objects, dependent types, step-indexed logical relations
-
-\section{Introduction}
-
-This paper presents a proposal for a new type-theoretic foundation of
-Scala and languages like it. The properties we are interested in
-modeling are Scala's path-dependent types and abstract type members,
-as well as its mixture of nominal and structural typing through the
-use of refinement types. Compared to previous approaches~\cite{nuObj,FS},
-we make no attempt to model inheritance or mixin composition. Indeed
-we will argue that such concepts are better modeled in a different
-setting.
-
-The calculus does not precisely describe what's currently in Scala. It
-is more normative than descriptive. The main point of deviation
-concerns the difference between Scala's compound type formation using
-{\bf with} and classical type intersection, as it is modeled in the
-calculus. Scala, and the previous calculi attempting to model it
-conflates the concepts of compound types (which inherit the members of
-several parent types) and mixin composition (which build classes from
-other classes and traits). At first glance, this offers an economy of
-concepts. However, it is problematic because mixin composition and
-intersection types have quite different properties. In the case of
-several inherited members with the same name, mixin composition has to
-pick one which overrides the others. It uses for that the concept of
-linearization of a trait hierarchy. Typically, given two independent
-traits $T_1$ and $T_2$ with a common method $m$, the mixin composition
-\code@$T_1$ with $T_2$@ would pick the $m$ in $T_2$, whereas the member in
-$T_1$ would be available via a super-call. All this makes sense from
-an implementation standpoint. From a typing standpoint it is more
-awkward, because it breaks commutativity and with it several
-monotonicity properties.
-
-In the present calculus, we replace Scala's compound types by
-classical intersection types, which are commutative. We also
-complement this by classical union types. Intersections and unions
-form a lattice wrt subtyping. This addresses another problematic
-feature of Scala: In Scala's current type system, least upper bounds
-and greatest lower bounds do not always exist. Here is an example:
-Given two traits
-\begin{lstlisting}
-  trait A { type T <: A }
-  trait B { type T <: B }
-\end{lstlisting}
-The greatest lower bound of \code@A@ and \code@B@ is approximated by the
-infinite sequence
-\begin{lstlisting}
-  A with B { type T <: A with B { type T <: A with B {
-    type T < ...
-  }}}
-\end{lstlisting}
-The limit of this sequence does not exist as a type in Scala.
-
-This is problematic because glbs and lubs play a central role in
-Scala's type inference. The absence of universal glbs and lubs makes
-type inference more brittle and more unpredictable.
-
-%The primary contributions of this paper are:
-%\begin{itemize}
-%\item Development of DOT -- a core calculus which models
-%  path-dependent types. We formally specify a small-step operational
-%  semantics and a static typing assignment in
-%  section~\ref{dot-calculus} and show some examples in
-%  section~\ref{dot-examples}.
-%\item Counterexamples to the standard theorem of preservation --
-%  section~\ref{dot-preservation}.
-%\item A sketch of a type-safety proof based on logical relations --
-%  section~\ref{dot-type-safety}.
-%\end{itemize}
-
-We propose DOT as a core calculus for path-dependent types. We present
-the calculus formally in section \ref{dot-calculus} and through examples in
-section~\ref{dot-examples}. Though we show that the calculus does not
-satisfy the standard theorem of preservation in
-section~\ref{dot-preservation}, we contribute a proof sketch of type
-safety using logical relations in section~\ref{dot-type-safety}. In
-section~\ref{discussion}, we discuss choices and variants of the
-calculus, as well as related work, and conclude in
-section~\ref{conclusion}.
-
-\section{The DOT Calculus}\label{dot-calculus}
-
-\boxfig{dot-one}{The DOT Calculus : Syntax, Reduction, Type / Declaration Assignment}{
-{\bf Syntax}\medskip
-\begin{center}    
-$\ba{l@{\hspace{0.2mm}}|@{\hspace{0.2mm}}l}
-\ba[t]{l@{\hspace{10mm}}l}
-x, y, z    & \lindent{\mbox{Variable}} \\
-l          & \lindent{\mbox{Value label}}\\
-m          & \lindent{\mbox{Method label}}\\[0.2em]
-v ::=      & \lindent{\mbox{Value}} \\
-\gap x     & \mbox{variable} \\[0.2em]
-t ::=      & \lindent{\mbox{Term}} \\
-\gap v     & \mbox{value} \\
-\gap \new x c t & \mbox{new instance} \\
-\gap t.l  & \mbox{field selection} \\
-\gap \mapp t m t  & \mbox{method invocation} \\[0.2em]
-p ::= & \lindent \mbox{Path} \\
-\gap x & \mbox{variable} \\
-\gap p.l & \mbox{selection} \\
-c ::= T_c \ldefs{\seq{d}} & \lindent{\mbox{Constructor}} \\[0.2em]
-d ::= & \lindent{\mbox{Initialization}}\\
-\gap l = v & \mbox{field initialization}\\
-\gap m(x) = t & \mbox{method initialization}\\
-s      ::= \seq{x \mapsto c} & \lindent\mbox{Store}
-\ea
-&
-\ba[t]{l@{\hspace{10mm}}l}
-L ::=      & \lindent{\mbox{Type label}} \\
-\gap L_c   & \mbox{class label} \\
-\gap L_a   & \mbox{abstract type label} \\[0.2em]
-S,T,U,V,W ::= & \lindent\mbox{Type}\\
-\gap p.L & \mbox{type selection} \\
-\gap T \refine z {\seq D} & \mbox{refinement} \\
-\gap T \tand T & \mbox{intersection type} \\
-\gap T \tor T & \mbox{union type} \\
-\gap \Top  & \mbox{top type} \\
-\gap \Bot  & \mbox{bottom type} \\[0.2em]
-S_c, T_c ::= & \lindent \mbox{Concrete type} \\
-\multicolumn{2}{l}{\gap p.L_c ~|~ T_c \refine z {\seq D} ~|~ T_c \wedge T_c  ~|~ \Top} \\[0.2em]
-D ::= & \lindent\mbox{Declaration} \\
-\gap \Ldecl L S U & \mbox{type declaration} \\
-\gap \ldecl l T   & \mbox{value declaration} \\
-\gap \mdecl m S U & \mbox{method declaration}\\
-\ \\
-\Gamma ::= \seq{x \typ T} & \lindent\mbox{Environment}
-\ea
-\ea$
-\end{center}
-\medskip
-
-\linesep
-
-\begin{multicols}{2}[\judgement{Reduction}{\fbox{$\reduction t s {t'} {s'}$}}]
-
-\infrule[\textsc{msel}]
-{y \mapsto T_c \ldefs{\seq{l = v'}\;\seq{m(x)=t}} \in s}
-{\reduction {\mapp y {m_i} v} s {\subst v {x_i} {t_i}} s}
-
-\infrule[\textsc{sel}]
-{y \mapsto T_c \ldefs{\seq{l = v}\;\seq{m(x)=t}} \in s}
-{\reduction {y.l_i} s {v_i} s}
-
-\infax[\textsc{new}]
-{\reduction {\new x c t} s t {s \envplus{x \mapsto c}}}
-
-\infrule[\textsc{context}]
-{\reduction t s {t'} {s'}}
-{\reduction {e[t]} s {e[t']} s'}
-\end{multicols}
-
-\hfill {\bf where} evaluation context $\gap e ::= [\,] ~|~ \mapp e m t ~|~ \mapp v m e ~|~ e.l\hspace{2cm}$
-
-\linesep
-
-\begin{multicols}{2}[\judgement{Type Assignment}{\fbox{$\Gamma \ts t \typ T$}}]
-
-\infrule[\textsc{var}]
-{x \typ T \in \Gamma}
-{\Gamma \ts x \typ T}
-
-\infrule[\textsc{msel}]
-{\Gamma \ts t \ni m \typ {S \tfun T} \\
- \Gamma \ts t' \typ T' \spcomma T' \sub S}
-{\Gamma \ts \mapp t m {t'} \typ T}
-
-\infrule[\textsc{sel}]
-{\Gamma \ts t \ni l \typ T'}
-{\Gamma \ts t.l \typ T'}
-
-\infrule[\textsc{new}]
-{
-y \notin \fn(T') \\
-\Gamma \ts T_c \wfe \spcomma T_c \expand_y \seq{\Ldecl L S U}, \seq{D} \\
-\Gamma \envplus{y: T_c} \ts \seq{S <: U} \spcomma \seq{d} \typ \seq{D} \spcomma {t'} \typ {T'}}
-{\Gamma \ts \new y {T_c \ldefs{\seq{d}}} {t'} \typ T'}
-
-\end{multicols}
-
-\linesep
-
-\begin{multicols}{2}[\judgement{Declaration Assignment}{\fbox{$\Gamma \ts d \typ D$}}]
-
-\infrule[\textsc{vdecl}]
-{\Gamma \ts v \typ {V'} \spcomma {V'} \sub V}
-{\Gamma \ts l = v \typ \ldecl l V}
-
-\infrule[\textsc{mdecl}]
-{\Gamma \ts S \wfe\\
- \Gamma \envplus{x : S} \ts t \typ {T'} \spcomma {T'} \sub T}
-{\Gamma \ts m(x) = t \typ \mdecl m S T }
-
-\end{multicols}
-
-} % END dot-one
-
-The DOT calculus is a simple system of dependent
-object-types. \figref{dot-one} gives its syntax, reduction rules,
-and type assignment rules.
-
-\subsection{Notation} We use standard notational conventions for
-sets. The notation $\seq{X}$ denotes a set of elements $X$. Given a
-such a set $\seq X$ in a typing rule, $X_i$ denotes an arbitrary
-element of $X$. 
-%The $\uplus$ operator extends a set of bindings. It is required that the added binding does not
-%introduce a variable which is already bound in the base-set.
-We use an
-abbreviation for preconditions in typing judgements. Given an
-environment $\Gamma$ and some predicates $P$ and $Q$, the condition $\Gamma \ts P \spcomma Q$
-is a shorthand for the two conditions $\Gamma \ts P$ and $\Gamma \ts Q$.
-
-\subsection{Syntax}
-
-There are four alphabets: Variable names $x$, $y$, $z$ are freely
-alpha-renamable. They occur as parameters of methods, as
-binders for objects created by \verb@new@-expressions, and as self
-references in refinements. Value labels $l$ denote fields in objects,
-which are bound to values at run-time. Similarly, method labels $m$
-denote methods in objects. Type labels $L$ denote type members of
-objects. Type labels are further separated into labels for abstract
-types $L_a$ and labels for classes $L_c$. It is assumed that in each
-program every class label $L_c$ is declared at most once.
-
-We assume that the label alphabets $l$, $m$ and $L$ are finite. This is
-not a restriction in practice, because one can include in these 
-alphabets every label occurring in a given program.
-
-The terms $t$ in DOT consist of variables $x$, $y$, $z$, field
-selections $t.l$, method invocations $t.m(t)$ and object creation
-expressions $\new y c {t'}$ where $c$ is a constructor $T_c \ldefs{\seq{l
-    = v}\;\seq{m(x) = t}}$. The latter binds a variable $y$ to a new
-instance of type $T_c$ with fields $\seq l$ initialized to values
-$\seq v$ and methods $\seq m$ initialized to methods of one parameter
-$\seq{x}$ and body $\seq{t}$.  The scope of $y$ extends through the term
-${t'}$.
-
-Two subclasses of terms are values $v$, which consist of just
-variables, and paths $p$ which consist of just variables and field
-selections.
-
-The types in DOT are denoted by letters $S$, $T$, $U$, $V$, or $W$. They consist of the following:
-\begin{itemize}
-\item[-] Type selections $p.L$, which denote the type member $L$ of path $p$.
-\item[-] Refinement types $T \refine z {\seq D}$, which refine a type $T$ by a set of declarations $D$.
-         The variable $z$ refers to the ``self''-reference of the type. Declarations can refer to
-         other declarations in the same type by selecting from $z$.
-\item[-] Type intersections $T \tand T'$, which carry the declarations of members present in either $T$ or $T'$.
-\item[-] Type unions $T \tor T'$, which carry only the declarations of members present in both $T$ and $T'$.
-\item[-] A top type $\Top$, which corresponds to an empty object.
-\item[-] A bottom type $\Bot$, which represents a non-terminating computation.
-\end{itemize}
-A subset of types $T_c$ are called {\em concrete types}. These are type selections
-$p.L_c$ of class labels,
-the top type $\Top$, intersections of concrete types, and refinements $T_c \refine z {\seq D}$ of concrete types. Only concrete types are allowed in constructors $c$.
-
-There are only three forms of declarations in DOT, which are all part
-of refinement types.  A value declaration $\ldecl l T$ introduces a
-field with type $T$.  A method declaration $\mdecl m S U$ introduces a
-method with parameter of type $S$ and result of type $U$. A type
-declaration $\Ldecl L S U$ introduces a type member $L$ with a lower
-bound type $S$ and an upper bound type $U$. There are no type aliases,
-but a type alias can be simulated by a type declaration $\Ldecl L T T$
-where the lower bound and the upper bound are the same type $T$.
-
-\boxfig{dot-decls}{The DOT Calculus : Declaration Lattice}{
-
-\bda{lcl@{\gap}l}
-
-      \dom(\seq D \tand \seq {D'}) &~=~& \dom(\seq{D}) \cup \dom(\seq{D'}) \\
-      \dom(\seq D \tor \seq {D'}) &=& \dom(\seq{D}) \cap \dom(\seq{D'}) \\[0.5em]
-      (D \tand D')(L) &=&
-        \Ldecl L {(S \tor S')} {(U \tand U')} & \mbox{if~} (\Ldecl L S U) \in \seq{D} \;\mbox{and}\; (\Ldecl L {S'} {U'}) \in \seq{D'} \\
-        &=& D(L) & \mbox{if~} L \notin \dom(\seq{D'}) \\
-        &=& D'(L) & \mbox{if~} L \notin \dom(\seq{D}) \\
-      (D \tand D')(m) &=&
-        \mdecl m {(S \tor S')} {(U \tand U')} & \mbox{if~} (\mdecl m S U) \in \seq{D} \;\mbox{and}\; (\mdecl m {S'} {U'}) \in \seq{D'} \\
-        &=& D(m) & \mbox{if~} m \notin \dom(\seq{D'}) \\
-        &=& D'(m) & \mbox{if~} m \notin \dom(\seq{D}) \\
-      (D \tand D')(l) &=&
-        \ldecl l {T \tand T'} & \mbox{if~} (\ldecl l T) \in \seq{D} \;\mbox{and}\; (\ldecl l {T'}) \in \seq{D'} \\
-        &=& D(l) & \mbox{if~} l \notin \dom(\seq{D'}) \\
-        &=& D'(l) & \mbox{if~} l \notin \dom(\seq{D}) \\[0.5em]
-      (D \tor D')(L) &=&
-        \Ldecl L {(S \tand S')} {(U \tor U')} & \mbox{if~} (\Ldecl L S U) \in \seq{D} \;\mbox{and}\; (\Ldecl L {S'} {U'}) \in \seq{D'} \\
-      (D \tor D')(m) &=&
-        \mdecl m {(S \tand S')} {(U \tor U')} & \mbox{if~} (\mdecl m S U) \in \seq{D} \;\mbox{and}\; (\mdecl m {S'} {U'}) \in \seq{D'} \\
-      (D \tor D')(l) &=&
-        \ldecl l {T \tor T'} & \mbox{if~} (\ldecl l T) \in \seq{D} \;\mbox{and}\; (\ldecl l {T'}) \in \seq{D'} 
-    \eda
-
-Sets of declarations form a lattice with the given meet $\wedge$ and
-join $\vee$, the empty set of declarations as the top element, and the
-bottom element $\seq{D_\Bot}$, Here $\seq{D_\Bot}$ is the set of
-declarations that contains for every term label $l$ the declaration
-$\ldecl l \Bot$, for every type label $L$ the declaration $\Ldecl L
-\Top \Bot$ and for every method label $m$ the declaration $\mdecl m
-\Top \Bot$.
-}
-
-Every field or type label can be declared only once in a set of
-declarations $\seq D$. A set of declarations can hence be seen as a map from
-labels to their declarations.  Meets $\wedge$ and joins $\vee$ on sets of
-declarations are defined in ~\figref{dot-decls}.
-
-\subsection{Reduction rules}
-
-Reduction rules $\reduction t s {t'} {s'}$ in DOT rewrite pairs of
-terms $t$ and stores $s$, where stores map variables to constructors.
-There are three main reduction rules: Rule (\textsc{msel}) rewrites a
-method invocation $\mapp y {m_i} v$ by retrieving the corresponding method
-definition from the store, and performing a substitution of the
-argument for the parameter in the body. Rule (\textsc{sel}) rewrites a
-field selection $x.l$ by retrieving the corresponding value from the
-store. Rule (\textsc{new}) rewrites an object creation $\new x c t$ by
-placing the binding of variable $x$ to constructor $c$ in the store
-and continuing with term $t$.  These reduction rules can be applied
-anywhere in a term where the hole $[\,]$ of an evaluation context $e$
-can be situated.
-
-\subsection{Type assignment rules}
-
-The last part of \figref{dot-one} presents rules for type
-assignment.
-
-Rules (\textsc{sel}) and (\textsc{msel}) type field selections and
-method invocations by means of an auxiliary membership relation $\ni$,
-which determines whether a given term contains a given declaration as
-one of its members. The membership relation is defined in
-\figref{dot-two} and is further explained in section~\ref{mem}.
-
-The last rule, (\textsc{new}), assigns types to object creation
-expressions. It is the most complex of DOT's typing rules.  To
-type-check an object creation $\new y {T_c \ldefs {\seq{l =
-      v}\;\seq{m(x) = t}}} t'$, one verifies first that the type $T_c$
-is well-formed (see \figref{dot-wf} for a definition of
-well-formedness).  One then determines the set of all declarations
-that this type carries, using the expansion relation $\expand$ defined
-in \figref{dot-two}.  Every type declaration $\Ldecl L S U$ in
-this set must be realizable, i.e.\ its lower bound $S$ must be a
-subtype of its upper bound $U$.  Every field declaration $\ldecl l V$
-in this set must have a corresponding initializing value of $v$ of
-type $V$.  These checks are made in an environment which is extended
-by the binding $y: T_c$. In particular this allows field values that
-recurse on ``self'' by referring to the bound variable $x$. Similarly,
-every method declaration $\mdecl m T W$ must have a corresponding
-initializing method definition $m(x) = t$. The parameter type $T$ must
-be $\nswfe$ (well-formed and expanding; see \figref{dot-wf}), and
-the body $t$ must type check to $W$ in an environment extended by the
-bindings $y : T_c$ and $x : T$.
-
-Instead of adding a separate subsumption rule, subtyping is expressed
-by preconditions in rules (\textsc{msel}) and (\textsc{new}).
-
-\boxfig{dot-two}{The DOT Calculus : Membership and Expansion}{
-    \begin{multicols}{2}[\judgement{Membership}{\fbox{$\Gamma \ts t \ni D$}}]
-
-      \infrule[\textsc{path-$\ni$}]
-      {\Gamma \ts p \typ T \spcomma T \expand_z \seq D}
-      {\Gamma \ts p \ni \subst p z {D_i}}
-
-      \infrule[\textsc{term-$\ni$}]
-      {z \not\in \fn(D_i) \andalso \Gamma \ts t \typ T \spcomma T \expand_z \seq D}
-      {\Gamma \ts t \ni D_i}
-   \end{multicols}
-
-   \linesep
-
-    \begin{multicols}{2}[\judgement{Expansion}{\fbox{$\Gamma \ts T \expand_z \seq D$}}]
-
-      \infrule[\textsc{rfn-$\expand$}]
-      {\Gamma \ts T \expand_z {\seq D'}}
-      {\Gamma \ts T \refine z {\seq D} \expand_z \seq {D'} \tand \seq D}
-
-      \infrule[\textsc{$\tand$-$\expand$}]
-      {\Gamma \ts T_1 \expand_z {\seq D_1} \spcomma T_2 \expand_z {\seq D_2}}
-      {\Gamma \ts T_1 \tand T_2 \expand_z {\seq D_1 \tand \seq D_2}}
-
-      \infax[\textsc{$\Top$-$\expand$}]
-      {\Gamma \ts \Top \expand_z \{\}}
-
-      \infrule[\textsc{tsel-$\expand$}]
-      {\Gamma \ts p \ni \Ldecl L S U \spcomma U \expand_z \seq D}
-      {\Gamma \ts p.L \expand_z \seq D}
-
-      \infrule[\textsc{$\tor$-$\expand$}]
-      {\Gamma \ts T_1 \expand_z {\seq D_1} \spcomma T_2 \expand_z {\seq D_2}}
-      {\Gamma \ts T_1 \tor T_2 \expand_z {\seq D_1 \tor \seq D_2}}
-
-      \infax[\textsc{$\Bot$-$\expand$}]
-      {\Gamma \ts \Bot \expand_z \seq{D_\Bot}}
-    \end{multicols}
-
-} % END dot-two
-
-\subsection{Membership}\label{mem}
-
-\figref{dot-two} presents typing rules for membership and expansion.
-The membership judgement $\Gamma \ts t \ni D$ states that in
-environment $\Gamma$ a term $t$ has a declaration $D$ as a member. The
-membership rules rely on expansion. There are two rules, one for paths
-(\textsc{path-$\ni$}) and one for general terms
-(\textsc{term-$\ni$}). For general terms, the ``self''-reference of
-the type must not occur in the resulting declaration $D$, since, to
-guarantee syntactic validity, we can only substitute a path for the
-``self''-reference.
-
-\subsection{Expansion}
-
-The expansion relation $\expand$ is needed to typecheck the complete
-set of declarations carried by a concrete type that is used in a
-\textbf{new}-expression. Expansion is also used by the membership
-rules and in subtyping refinements on the right (see
-\figref{dot-sub}).
-
-Rule (\textsc{rfn-$\expand$}) states that a refinement type $T
-\expand_z {\seq D}$ expands to the conjunction of the expansion $D'$
-of $T$ and the newly added declarations $D$. Rule
-(\textsc{tsel-$\expand$}) states that a type selection $p.L$ carries
-the same declarations as the upper bound $U$ of $L$ in $T$.  Rules
-($\tand$-$\expand$) and ($\tor$-$\expand$) states that expansion
-distributes through meets and joins.  Rule (\textsc{$\top$-$\expand$})
-states that the top type $\top$ expands to the empty set. Rule
-(\textsc{$\bot$-$\expand$}) states that the bottom type $\bot$ expands
-to the bottom element $\seq{D_\Bot}$ of the lattice of sets of
-declarations (recall \figref{dot-decls}).
-
-\boxfig{dot-sub}{The DOT Calculus : Subtyping and Declaration Subsumption}{
-
-    \begin{multicols}{2}[\judgement{Subtyping}{\fbox{$\Gamma \ts S \sub T$}}]
-
-%      \infrule[\textsc{trans}]
-%      {\Gamma \ts S \sub T \spcomma T \sub U}
-%      {\Gamma \ts S \sub U}
-
-      \infrule[\textsc{refl}]
-      {\Gamma \ts T \wfe}
-      {\Gamma \ts T \sub T}
-
-      \infrule[\textsc{$\sub$-rfn}]
-      {\Gamma \ts {T \refine z {\seq D}} \wfe
-       \spcomma S \sub T \spcomma S \expand_z \seq{D'} \\
-       \Gamma \envplus{z: S} \ts \seq{D' \sub D}}
-      {\Gamma \ts S \sub T \refine z {\seq D}}
-
-      \infrule[\textsc{$\sub$-tsel}]
-      {\Gamma \ts p \ni \Ldecl L S U \spcomma S \sub U \spcomma S' \sub S }
-      {\Gamma \ts S' \sub p.L}
-
-      \infrule[\textsc{$\sub$-$\tand$}]
-      {\Gamma \ts T \sub T_1 \spcomma T \sub T_2}
-      {\Gamma \ts T \sub T_1 \tand T_2}
-
-      \infrule[\textsc{$\sub$-$\tor_1$}]
-      {\Gamma \ts T_2 \wfe \spcomma T \sub T_1}
-      {\Gamma \ts T \sub T_1 \tor T_2}
-
-      \infrule[\textsc{$\sub$-$\tor_2$}]
-      {\Gamma \ts T_1 \wfe \spcomma T \sub T_2}
-      {\Gamma \ts T \sub T_1 \tor T_2}
-
-      %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-
-      \infrule[\textsc{$\sub$-$\Top$}]
-      {\Gamma \ts T \wfe}
-      {\Gamma \ts T \sub \Top}
-
-      \infrule[\textsc{$\Bot$-$\sub$}]
-      {\Gamma \ts T \wfe}
-      {\Gamma \ts \Bot \sub T}
-
-      \infrule[\textsc{rfn-$\sub$}]
-      {\Gamma \ts {T \refine z {\seq D}} \wfe
-       \spcomma T \sub T'}
-      {\Gamma \ts T \refine z {\seq D} \sub T'}
-
-      \infrule[\textsc{tsel-$\sub$}]
-      {\Gamma \ts p \ni \Ldecl L S U \spcomma S \sub U \spcomma U \sub U'}
-      {\Gamma \ts p.L \sub U'}
-
-      \infrule[\textsc{$\tor$-$\sub$}]
-      {\Gamma \ts T_1 \sub T \spcomma T_2 \sub T}
-      {\Gamma \ts T_1 \tor T_2 \sub T}
-
-      \infrule[\textsc{$\tand_1$-$\sub$}]
-      {\Gamma \ts T_2 \wfe \spcomma T_1 \sub T}
-      {\Gamma \ts T_1 \tand T_2 \sub T}
-
-      \infrule[\textsc{$\tand_2$-$\sub$}]
-      {\Gamma \ts T_1 \wfe \spcomma T_2 \sub T}
-      {\Gamma \ts T_1 \tand T_2 \sub T}
-
-      \end{multicols}
-
-    \linesep
-
-    \begin{multicols}{2}[\judgement{Declaration subsumption}{\fbox{$\Gamma \ts D \sub D'$}}]
-
-    \infrule[\textsc{tdecl-$\sub$}]
-            {\Gamma \ts S' \sub S \spcomma T \sub T'}
-            {\Gamma \ts (\Ldecl L S T) \sub (\Ldecl L {S'} {T'})}
-
-    \infrule[\textsc{vdecl-$\sub$}]
-            {\Gamma \ts T \sub T'}
-            {\Gamma \ts (\ldecl l T) \sub (\ldecl l {T'})}
-
-    \infrule[\textsc{mdecl-$\sub$}]
-            {\Gamma \ts S' \sub S \spcomma T \sub T'}
-            {\Gamma \ts (\mdecl m S T) \sub (\mdecl m {S'} {T'})}
-
-    \end{multicols}
-} % END dot-sub
-
-\subsection{Subtyping}
-
-\figref{dot-sub} defines the subtyping judgement $\Gamma \ts S
-\sub T$ which states that in environment $\Gamma$ type $S$ is a
-subtype of type $T$. Subtyping is regular wrt $\nswfe$: if
-type $S$ is a subtype of type $T$, then $S$ and $T$ are well-formed
-and expanding. Though this regularity limits our calculus to
-$\nswfe$-types, this limitation allows us to show that subtyping is
-transitive, as discussed in section~\ref{dot-preservation}.
-
-\subsection{Declaration Subsumption}
-
-The declaration subsumption judgement $\Gamma \ts D \sub D'$ in
-\figref{dot-sub} states that in environment $\Gamma$ the declaration
-$D$ subsumes the declaration $D'$. There are three rules, one for each
-kind (type, value, method) of declarations.  Rule
-(\textsc{tdecl-$\sub$}) states that a type declaration $\Ldecl L S U$
-subsumes another type declaration $\Ldecl L {S'} {U'}$ if $S'$ is a
-subtype of $S$ and $U$ is a subtype of $U'$. In other words, the set
-of types between $S$ and $U$ is contained in the set of types between
-$S'$ and $U'$.  Rule (\textsc{vdecl-$\sub$}) states that a value
-declaration $\ldecl l T$ subsumes another value declaration $\ldecl l
-T'$ if $T$ is a subtype of $T'$. Rule (\textsc{mdecl-$\sub$}) is
-similar to (\textsc{tdecl-$\sub$}), as the parameter type varies
-contravariantly and the return type covariantly.
-
-Declaration subsumption is extended to a binary relation between
-sequences of declarations: $\seq{D} <: \seq{D'}$ iff $\forall D'_i,
-\exists D_j. {D_j \sub D'_i}$.
-
-\boxfig{dot-wf}{The DOT Calculus : Well-Formedness}{
-
-       \begin{multicols}{2}[\judgement{Well-formed types}{\fbox{$\Gamma \ts T \wf$}}]
-
-      \infrule[\textsc{rfn-wf}]
-      {\Gamma \ts T \wf \\ 
-       \Gamma \envplus {z: T \refine z {\seq D}} \ts \seq {D \wf}}
-      {\Gamma \ts T \refine z {\seq D} \wf}
-
-      \infrule[\textsc{tsel-wf$_1$}]
-      {\Gamma \ts p \ni \Ldecl L S U \spcomma S \wf \spcomma U \wf}
-      {\Gamma \ts p.L \wf}
-
-      \infrule[\textsc{$\tand$-wf}]
-      {\Gamma \ts T \wf \spcomma T' \wf}
-      {\Gamma \ts T \tand T' \wf}
-
-      %%%%%%%%%%%%%%%%%%%%%%%%%%%
-
-      \infax[\textsc{$\Top$-wf}]
-      {\Gamma \ts \Top \wf}
-
-      \infax[\textsc{$\Bot$-wf}]
-      {\Gamma \ts \Bot \wf}
-
-      \infrule[\textsc{tsel-wf$_2$}]
-      {\Gamma \ts p \ni \Ldecl L \Bot U}
-      {\Gamma \ts p.L \wf}
-
-      \infrule[\textsc{$\tor$-wf}]
-      {\Gamma \ts T \wf \spcomma T' \wf}
-      {\Gamma \ts T \tor T' \wf}
-
-    \end{multicols}
-
-    \linesep
-
-    \begin{multicols}{2}[\judgement{Well-formed declarations}{\fbox{$\Gamma \ts D \wf$}}]
-      \infrule[\textsc{tdecl-wf}]
-      {\Gamma \ts S \wf \spcomma U \wf}
-      {\Gamma \ts \Ldecl L S U \wf}
-
-      \infrule[\textsc{vdecl-wf}]
-      {\Gamma \ts T \wf}
-      {\Gamma \ts \ldecl l T \wf}
-
-      \infrule[\textsc{mdecl-wf}]
-      {\Gamma \ts S \wf \spcomma U \wf}
-      {\Gamma \ts \mdecl m S U \wf}
-
-    \end{multicols}
-
-    \linesep
-
-    \begin{multicols}{2}[\judgement{Well-formed and expanding types}{\fbox{$\Gamma \ts T \wfe$}}]
-      \infrule[\textsc{wfe}]
-      {\Gamma \ts T \wf \spcomma T \expand_z \seq{D}}
-      {\Gamma \ts T \wfe}
-    \end{multicols}
-} % END dot-wf
-
-\subsection{Well-formedness}
-
-The well-formedness judgement $\Gamma \ts T \wf$ in \figref{dot-wf}
-states that in environment $\Gamma$ the type $T$ is
-well-formed.
-
-A refinement type $T \refine z {\seq D}$ is
-well-formed if the parent type $T$ is well-formed and every
-declaration in $\seq D$ is well-formed in an environment augmented by
-the binding of the self-reference $z$ to the refinement type itself
-(\textsc{rfn-wf}).
-
-A type selection $p.L$ is well-formed if $L$ is a member of $p$, and
-the lower bound of $L$ is also well-formed (\textsc{tsel-wf$_1$} and
-\textsc{tsel-wf$_2$}). The latter condition has the effect that the
-lower bound of a type $p.L$ may not refer directly or indirectly to a
-type containing $p.L$ itself --- if it would, the well-formedness
-judgement of $p.L$ would not have a finite proof. No such restriction
-exists for the upper bound of $L$ if the lower bound is $\bot$
-(\textsc{tsel-wf$_2$}). The upper bound may in fact refer back to the
-type. Hence, recursive class types and F-bounded abstract types are
-both expressible.
-
-The other forms of types in DOT are all well-formed if their
-constituent types are well-formed.
-
-Well-formedness extends straightforwardly to declarations with the
-judgement $\Gamma \ts D \wf$. All declarations are well-formed if
-their constituent types are well-formed.
-
-\section{Examples}\label{dot-examples}
-
-\subsection{Sugar: Functions}
-Like in Scala, we can encode functions as objects with a special
-method. We will freely use the following sugar in the remaining of
-this paper. Note that the variable $z$ must be fresh.
-\begin{align*}
-\arrow S T &\iff {\Top \refine z {\mdecl {\mi{apply}} S T}}\\
-\fun x S T t &\iff \new z {{\arrow S T}\ldefs{\mi{apply}(x) = t}} z\\
-\app f x &\iff \mapp f {\mi{apply}} x\\
-\cast T t&\iff \app {\fun x T T x} t
-\end{align*}
-
-In the remaining examples, for brevity, we will also use $\abs x S t$
-for $\fun x S \_ t$ where the return type can be somewhat inferred
-from the context.
-
-\subsection{Basics: Booleans, Error, \ldots}
-This program defines a $\mi{root}$ object with basic types
-($\mi{Unit}$, $\mi{Boolean}$, $\mi{Nat}$) and methods ($\mi{true}$,
-$\mi{false}$, $\mi{error}$, $\mi{zero}$). For brevity, we've omitted
-the code related to natural numbers. During the object creation, the
-method labels such as $\mi{false}$ are all initialized.
-\begin{align*}
-&\mlnew {\mi{root}} {\Top \mlrefine r {\\\displaybreak[3]
-&\ \Ldecl {\mi{Unit}} \Bot \Top\\\displaybreak[3]
-&\ \mdecl {\mi{unit}} \Top {r.\mi{Unit}}\\\displaybreak[3]
-&\ \Ldecl {\mi{Boolean}} \Bot {\Top \mlrefine z {\\\displaybreak[3]
-&\ \gap \ldecl {\mi{ifNat}} \arrow {(\arrow {r.\mi{Unit}} {r.\mi{Nat}})} {\arrow {(\arrow {r.\mi{Unit}} {r.\mi{Nat}})} {r.\mi{Nat}}}\\\displaybreak[3]
-&\gap}}\\\displaybreak[3]
-&\ \mdecl {\mi{false}} {r.\mi{Unit}} {r.\mi{Boolean}}\\\displaybreak[3]
-&\ \mdecl {\mi{true}} {r.\mi{Unit}} {r.\mi{Boolean}}\\\displaybreak[3]
-&\ \mdecl {\mi{error}} {r.\mi{Unit}} \Bot\\\displaybreak[3]
-& \ \ldots\\\displaybreak[3]
-&}\mlldefs{\\\displaybreak[3]
-&\ \mi{unit}(x) = {\new u {\mi{root}.\mi{Unit}} u}\\\displaybreak[3]
-%
-&\ \mi{false}(u) =\\\displaybreak[3]
-&\ \gap \mlnew {\mi{ff}} {\mi{root}.\mi{Boolean} \mlldefs{\\\displaybreak[3]
-&\ \gap\gap {\mi{ifNat}} = \abs t {\arrow {\mi{root}.\mi{Unit}} {\mi{root}.\mi{Nat}}} {\\&\gap\gap\gap\abs e {\arrow {\mi{root}.\mi{Unit}} {\mi{root}.\mi{Nat}}}\\&\gap\gap\gap\app e {\mi{root}.\mi{unit}}}\\\displaybreak[3]
-&\ \gap}}{
-\ \gap \mi{ff}}\\\displaybreak[3]
-%
-&\ \mi{true}(u) = \ldots\\\displaybreak[3]
-%
-&\ \mi{error}(u) = {\app {\mi{root}.\mi{error}} u}\\\displaybreak[3]
-&\ \ldots\\\displaybreak[3]
-&}}{
-\ldots
-}
-\end{align*}
-
-\subsection{Polymorphic Lists}
-Polymorphic lists can be expressed using an abstract type member for
-the element type ($\mi{Elem}$). We can instantiate a refinement of the
-list package to manipulate lists with a particular element type. We
-can choose to make the lists invariant or covariant in the element
-type by fixing the lower bound of the $\mi{Elem}$ re-declaration to
-the upper bound or to $\Bot$, respectively.
-\begin{align*}
-&\mlnew {\mi{genLists}} {\Top \mlrefine g {\Ldecl {\mi{ListPackage}} \Bot {\Top \mlrefine p {\\\displaybreak[3]
-&\ \Ldecl {\mi{Elem}} \Bot \Top\\\displaybreak[3]
-&\ \Ldecl {\mi{ListOfElement}} {g.\mi{List} \refine z {\Ldecl {\mi{Elem}} {p.\mi{Elem}} {p.\mi{Elem}}}} {\\&\gap\gap\gap\gap g.\mi{List} \refine z {\Ldecl {\mi{Elem}} {p.\mi{Elem}} {p.\mi{Elem}}}}\\\displaybreak[3]
-&\ \Ldecl {\mi{List}} \Bot {\Top \mlrefine z {\\\displaybreak[3]
-&\ \gap \mdecl {\mi{isEmpty}} {\mi{root}.\mi{Unit}} {\mi{root}.\mi{Boolean}}\\\displaybreak[3]
-&\ \gap \mdecl {\mi{head}} {\mi{root}.\mi{Unit}} {g.\mi{Elem}}\\\displaybreak[3]
-&\ \gap \mdecl {\mi{tail}} {\mi{root}.\mi{Unit}} {g.\mi{ListOfElem}}\\\displaybreak[3]
-&\gap}}\\\displaybreak[3]
-&\ \mdecl {\mi{nil}} {\mi{root}.\mi{Unit}} {g.\mi{ListOfElem}}\\\displaybreak[3]
-&\ \mdecl {\mi{cons}} {g.\mi{Elem}} {g.\mi{ListOfElem}}\\\displaybreak[3]
-&}}}}{
-\mlnew {\mi{natLists}} {{\mi{genLists}.\mi{ListPackage}} \mlrefine p {\\\displaybreak[3]
-&\gap \Ldecl {\mi{Elem}} {\mi{root}.\mi{Nat}} {\mi{root}.\mi{Nat}} }\mlldefs{\\\displaybreak[3]
-&\ {\mi{nil}}(u) = \ldots\\\displaybreak[3]
-&\ {\mi{cons}}(u) = \ldots\\\displaybreak[3]
-&}}}{
-\ldots
-}
-\end{align*}
-
-\section{Counterexamples to Preservation}\label{dot-preservation}
-
-We sketch a proof that the DOT calculus is type-safe using logical
-relations in section~\ref{dot-type-safety}. However, we first tried to
-prove the calculus type-safe using the standard theorems of
-preservation and progress~\cite{soundness,tapl}. Unfortunately, for
-the calculus as presented, and any of its variants that we devised,
-preservation doesn't hold. In this section, we review some of the most
-salient counterexamples to preservation that we found. These
-counterexamples have been checked with PLT Redex~\cite{plt_redex}.
-
-\subsection{\texorpdfstring{\textsc{term-$\ni$}}{Term-Mem} Restriction}\label{term_mem}
-
-There are two membership ($t \ni D$) rules: one for when the term $t$
-is a path, and one for an arbitrary term $t$. For paths, we can
-substitute the self-references in the declarations, but we cannot do
-so for arbitrary terms as the resulting types wouldn't be well-formed
-syntactically. Hence, the $\textsc{term-$\ni$}$ has the restriction
-that self-occurrences are not allowed. Here is a counterexample related
-to this restriction.
-
-Let $X$ be a shorthand for the type:
-\begin{align*}
-\Top & \mlrefine z {\\
-&\ \Ldecl {L_a} \Top \Top\\
-&\ \ldecl l {z.L_a}\\
-}&
-\end{align*}
-
-Let $Y$ be a shorthand for the type:
-\begin{align*}
-\Top & \mlrefine z {\\
-&\ \ldecl l \Top\\
-}&
-\end{align*}
-
-Now, consider the term
-\begin{align*}
-&\mlnew u {X \ldefs{ l = u }}{
-\app {(\abs y {\Top \tfun Y} {\app y u})} {(\abs d \Top {\cast X u})}.l
-}
-\end{align*}
-
-The term type-checks because the term $t=\app {(\abs y {\Top \tfun Y}
-  {\app y u})} {(\abs d \Top {\cast X u})}$ has type
-$Y$, so we can apply $\textsc{term-$\ni$}$ for $l$. However, the term
-$t$ eventually steps to ${\cast X u}$ which has type $X$,
-so we cannot apply $\textsc{term-$\ni$}$ for $l$ because of the
-self-reference ($z.L_a$).
-
-\subsection{Expansion Lost}\label{narrowing_exp}
-
-Expansion is not preserved by narrowing. Here, we create two type
-selections that are mutually recursive in their upper bounds after
-narrowing: $z_0.C_2$ initially expands, but after narrowing, $z_0.C_2$
-expands to what $z_0.A_2$ expands to, which expands to what $z_0.A_1$
-expands to, which expands to what $z_0.A_2$ expands to, and thus we
-have an infinite expansion. Thus, the last new expression initially
-type-checks, but after narrowing, it doesn't because the precise
-expansion needed by $\textsc{new}$ cannot be inferred.
-
-\begin{align*}
-&\mlnew {x_0} {\Top \mlrefine z { \Ldecl {A_1} \Bot {\Top \mlrefine z {\\
-&\gap\Ldecl {A_2} \Bot \Top\\
-&\gap\Ldecl {A_3} \Bot \Top\\
-&\gap\Ldecl {C_2} \Bot {z.A_2}}}}\ldefs{}}{
-\mlnew {x_1} {\Top \refine z {\Ldecl {C_1} \Bot {\Top \refine z {\Ldecl {A_1} \Bot {x_0.A_1}}}}\ldefs{}}}{
-\mlnew {x_2} {{x_1.C_1} \refine z {\Ldecl {A_1} \Bot {{x_0.A_1} \refine z {\Ldecl {A_2} \Bot {z.A_3}}}}\ldefs{}}}{
-\mlnew {x_3} {{x_1.C_1} \refine z {\Ldecl {A_1} \Bot {{x_0.A_1} \refine z {\Ldecl {A_3} \Bot {z.A_2}}}}\ldefs{}}}{
-\mlapp {\abs x {x_1.C_1} {(\abs {z_0} {x.A_1 \tand x_3.A_1} {\\&\gap\gap\gap\new z {z_0.C_2} {\app {(\abs x \Top x)} z}})}} {\ x_2}}
-\end{align*}
-
-\subsection{Well-Formedness Lost}\label{narrowing_wf}
-
-Even well-formedness is not preserved by narrowing. The trick is that
-if the lower bound of a type selection is not $\Bot$, then the
-bounds needs to be checked for well-formedness. Here, we create two
-type selections that are mutually recursive in their bounds
-after narrowing. $y.A$ is initially well-formed, but after narrowing,
-it isn't because we run into an infinite derivation trying to prove
-the well-formedness of its bounds.
-
-\begin{align*}
-&\mlnew v {\Top \refine z {\Ldecl L \Bot {\Top \refine z {\Ldecl A \Bot \Top, \Ldecl B {z.A} {z.A}}}} \ldefs{}}{
-\mlapp {(\abs x {\Top \refine z {\Ldecl L \Bot {\Top \refine z {\Ldecl A \Bot \Top, \Ldecl B \Bot \Top}}}} {\\&\gap
-\mlnew z {\Top \mlrefine z {\\&\gap\gap\gap
-\mdecl l {x.L \tand {\Top \refine z {\Ldecl A {z.B} {z.B}, \Ldecl B \Bot \Top}}} \Top}\mlldefs{\\&\gap\gap\gap
-l(y) = \fun a {y.A} \Top a}}{
-\gap \cast {\top} z
-}})}{\ v}}
-\end{align*}
-
-\subsection{Path Equality}\label{patheq}
-
-For preservation, we need to be able to relate path-dependent types
-after reduction. Here is a motivating example:
-
-\begin{align*}
-&\mlnew b {\Top \mlrefine z {&&\Ldecl X \Top \Top&\\
-&&&\ \ldecl l {z.X}&}\ldefs{l = b}}{
-\mlnew a {\Top \mlrefine z {\ldecl i {\Top \mlrefine z {&&&\\
-&&&\Ldecl X \Bot \Top\\
-&&&\ldecl l {z.X}}}&}\ldefs{i = b}}}{
-\app {(\abs x \Top x)} {\app {(\abs x {a.i.X} x)} {a.i.l}}}
-\end{align*}
-
-$a.i.l$ reduces to $b.l$. $b.l$ has type $b.X$, so we need $b.X <:
-a.i.X$. This cannot be established with the current rules: it is not
-true in general, but true here because $a.i$ reduces to $b$. Hence,
-we need to acknowledge path equality for preservation to hold.
-
-In section~\ref{pres-patch}, we discuss our failure to patch the
-calculus for preservation to hold.
-
-\section{Type Safety via Logical Relations}\label{dot-type-safety}
-We sketch a proof of type-safety of the DOT calculus via step-indexed
-logical relations~\cite{ahmed04,ahmed06,step_indexed_obj}.
-
-\subsection{Type Safety}
-Type-safety states that a well-typed program doesn't get stuck. More
-formally: If $\emptyset \ts t \typ T$ and $\reductionl t \emptyset {t'} {s'} *$ then
-either $t'$ is a value or $\exists t'', s''. \reduction {t'} {s'}
-{t''} {s''}$.
-
-Our strategy is to define a logical relation $\Gamma \ds t : T$, such
-that $\Gamma \ts t : T$ implies $\Gamma \ds t : T$ implies
-type-safety.
-
-\subsection{Step-Indexed Logical Relations}
-
-In order to ensure that our logical relation is well-founded, we use a
-step index. For each step index $k$, we define the set of values and
-the set of terms that appear to belong to a given type, when taking at
-most $k$ steps. $\Gamma \ds t : T$ is then defined in terms of the
-step-indexed logical relation by requiring it to hold $\forall k$.
-
-\subsubsection{Set of Values}
-$\relv k \Gamma s T$ defines the set of values that appear to have
-type $T$ when taking at most $k$ steps. $\Gamma$ and $s$ must agree:
-$\dom(\Gamma) = \dom(s)$ (ordered) and $\forall (x : T) \in \Gamma, x
-\in \relv k \Gamma s T$. A variable $y$ belongs to $\relv 0 \Gamma s
-T$ simply by being in the store. In addition, it belongs to $\relv k
-\Gamma s T$ for $k > 0$, if it defines all type, method and value labels in
-the expansion of $T$ appropriately for $j < k$ steps.
-\begin{align*}
-&\relv k \Gamma s T = \{ y \ |\\
-&\gap y \in \dom(s) \andl (\\
-&(\Gamma \ts T \wfe \andl\\
-&\gap \forall j < k,\\
-&\gap y \mapsto T_c \ldefs{\seq{l = v}\;\seq{m(x)=t}} \in s,\\
-&\gap \Gamma \ts T \expand_y \seq{D},\\
-&\gap (\forall m_i : S \rightarrow U \in \seq{D},\\
-&\gap\gap t_i \in \rele j {\Gamma \envplus{x_i : S}} s U) \andl\\
-&\gap (\forall l_i : V \in \seq{D},\\
-&\gap\gap v_i \in \relv j \Gamma s V)) \orl\\
-& (T = T_1 \tand T_2 \andl y \in \relv k \Gamma s {T_1} \andl y \in \relv k \Gamma s {T_2}) \orl\\
-& (T = T_1 \tor T_2 \andl  (y \in \relv k \Gamma s {T_1} \orl y \in \relv k \Gamma s {T_2}))\\
-&\}
-\end{align*}
-
-This relation captures the observation that the only ways for a term
-to get stuck is to have a field selection on an uninitialized field or
-a method invocation on an uninitialized method. However, a {\it potential
-  pitfall} is that the value itself might occur in the types $S$, $U$,
-$V$.
-
-\subsubsection{Set of Terms}
-$\rele k \Gamma s T$ defines the set of terms that appear to have type
-$T$ when taking at most $k$ steps. $s$ must agree with a {\it prefix}
-of $\Gamma$, so $\Gamma$ can additionally contain variables not in
-$s$. This is needed for checking methods in $\mathcal{V}$ above, and
-for relating open terms. If $k > 0$, $\mathcal{E}$ extends $\Gamma$
-and $s$ so that they agree. It then states that if it can reduce $t$
-in the extended store to an irreducible term in $j < k$ steps, then
-this term must be in a corresponding $\mathcal{V}$ set with $\Gamma$
-now extended to agree with the store resulting from the reduction
-steps.
-
-$\irred {t} {s}$ is a shorthand for $\neg\exists t', s'. \reduction t s
-{t'} {s'}$. $\mathcal{\supseteq}$ is used initially for the possibly
-shorter store to agree with the environment, and can extend both in
-many different ways. $\mathcal{\supseteq}^!$ is used finally for the
-possibly shorter environment to agree with the store, and just extends
-the environment in one straightforward way: hence, it defines
-singleton sets.
-
-\begin{align*}
-&\rele k \Gamma s T = \{ t \ |\\
-&\gap \Gamma \ts T \wfe \andl\\
-&\gap k = 0 \orl (\forall j < k,\\
-&\gap\gap \forall (\Gamma'; s') \in \rels k \Gamma s ,\\
-&\gap\gap \reductionl t {s'} {t'} {s''} j \andl\\
-&\gap\gap \irred {t'} {s''} \impliesl\\
-&\gap\gap \forall \Gamma'' \in \relg k {s''} {\Gamma'} ,\\
-&\gap\gap {t'} \in \relv {k-j-1} {\Gamma''} {s''} T)\\
-&\}
-\end{align*}
-
-\subsubsection{Extending the environment and the store}
-$\rels k \Gamma s$ for $k > 0$ defines the set of completed
-environment and stores that agree on $k-1$ steps, and that extend
-$\Gamma$ and $s$. $s$ must agree with a {\it prefix} of $\Gamma$.
-Both $\Gamma$ and $s$ are ordered maps. For $s$, $s'$ extends $s$ if
-$s$ is a prefix of $s'$. For $\Gamma$, $\Gamma'$ extends $\Gamma$ if
-we get back $\Gamma$ by keeping only the elements of $\Gamma'$ that
-belong to $\Gamma$. Furthermore, a prefix of $\Gamma'$ agrees with
-$s$.
-\begin{align*}
-&\rels k \Gamma s = \{\\
-&\ (\seq{x : T}^m, \seq{x_{ij} : T_{ij}}^{m \leq i < n; 0 \leq j < i_n}; s, \seq{x_{ij} \mapsto c_{ij}}^{m \leq i < n; 0 \leq j < i_n}) |\\
-&\ s = \seq{x \mapsto c}^m \andl \Gamma = \seq{x : T}^n \andl \\
-&\ m\leq n \andl \forall i, m \leq i < n, \forall {i_n}, j, 0 \leq j < i_n,\\
-&\ \forall T_{ij}, c_{ij}, T_{i{(i_n-1)}} = T_i, \forall n' \leq n, i_n' \leq i_n,\\
-&\ c_{ij} \in \relv {k-1} {\seq{x : T}^m, \seq{x_{ij} : T_{ij}}^{m \leq i' < n; 0 \leq j < i_n'}} {s, \seq{x_{ij} \mapsto c_{ij}}^{m \leq i < n'; 0 \leq j < i_n'}} {T_{ij}}\\
-&\}
-\end{align*}
-
-\subsubsection{Completing the environment to agree with the store}
-$\relg k s \Gamma$ defines a singleton set of a completed environment
-that agrees with a store $s$ by simply copying the constructor type
-from the store for each missing variable.
-\begin{align*}
-&\relg k s \Gamma = \{ \Gamma, \seq{x_i : T_{c_i}}^{m \leq i < n} \ |\\
-&\gap \Gamma = \seq{x : T}^m \andl s = \seq{x \mapsto c}^n \andl\\
-&\}
-\end{align*}
-
-\subsubsection{Terms in the Logical Relation}
-
-$\Gamma \ds t : T$ is simply defined as $t \in \rele k \Gamma \emptyset T,
-\forall k$.
-
-\subsection{Statements and Proofs}
-
-\subsubsection{Fundamental Theorem}
-
-The fundamental theorem is the implication from $\Gamma \ts t : T$ to
-$\Gamma \ds t : T$. Type safety is a straightforward corollary of this
-theorem.
-
-\begin{myproof}
-The proof is on induction on the derivation of $\Gamma \ts t : T$. For
-each case, we need to show $t \in \rele k \Gamma \emptyset T, \forall
-k$. The non-trivial case is when $k > 0$ and for $(\Gamma'; s') \in
-\rels k \Gamma s$ and some $j < k$, $\reductionl {t} s {t'} {s'} j
-\andl \irred {t'} {s'}$. Then, we need to show ${t'} \in \relv {k-j-1}
-      {\Gamma''} {s'} T$ for ${\Gamma''} \in \relg \Gamma k {s'}
-      {\Gamma'}$.
-
-\ \\
-
-\Case \textsc{var}: $\Gamma \ts x : T$ knowing $(x : T) \in
-\Gamma$. $x \in \relv {k-1} {\Gamma'} s T$ follows from the definition of
-$\rels k \Gamma \emptyset$.
-
-\ \\
-
-\Case \textsc{sel}: $\Gamma \ts t_1.l_i : T$ knowing $\Gamma \ts t_1 :
-T_1$, $\Gamma \ts T_1 \expand_z \seq{D}$, $\ldecl {l_i} {V_i} \in
-\seq{D}$ and knowing either that $t_1 = p_1 \andl T = \subst p z
-    {V_i}$ or that $z \not\in \fn(V_i) \andl T = V_i$.
-
-By operational semantics and induction hypothesis, $\reductionl {t_1}
-s {t_1'} {s'} {j-1}$ and $\irred {t_1'} {s'}$ and ${t_1'} \in \relv
-{k-j+1-1} {\Gamma'} {s'} {T_1}$.
-
-By operational semantics and the above, $\reductionl {t_1'.l_i} {s'}
-{t'} {s'} 1$, and we can conclude $t' \in \relv {k-j-1} {\Gamma''} {s'}
-T$ from the clause for value labels of ${t_1'} \in \relv {k-j}
-{\Gamma''} {s'} {T_1}$.
-
-\ \\
-
-\Case \textsc{msel}: $\Gamma \ts t_1.m_i(t_2) : T$ knowing $\Gamma \ts
-t_1 : T_1$, $\Gamma \ts t_2 : T_2$, $\Gamma \ts T_1 \expand_z
-\seq{D}$, $\mdecl {m_i} {S_i} {U_i} \in \seq{D}$ and knowing either
-that $t_1 = p_1 \andl S = \subst p z {S_i} \andl T = \subst p z {U_i}$
-or that $z \not\in \fn(S_i) \andl z \not\in \fn(U_i) \andl S = S_i \andl
-T = U_i$, and knowing that $\Gamma \ts T_2 \sub S$.
-
-By operational semantics and induction hypotheses, $\reductionl {t_1}
-s {t_1'} {s_1} {j_1}$ and $\irred {t_1'} {s_1}$ and $\reductionl {t_2}
-{s} {t_2'} {s_2} {j_2}$ and $\irred {t_2'} {s_2}$ and ${t_1'} \in
-\relv {k-j_1-1} {\Gamma_1} {s_1} {T_1}$ and ${t_2'} \in \relv
-{k-j_2-1} {\Gamma_2} {s_2} {T_2}$.
-
-Because $t_2$ reduces to a value $t_2'$ starting in store $s$, it
-should also reduce to a value $v_2$ in the same number of steps
-starting in store $s_1$, since $s_1$ extends $s$. So let $\reductionl
-{t_2} {s_1} {v_2} {s_{12}} {j_2}$ with $v_2 \in \relv {k-j_2-1}
-{\Gamma_{12}} {s_{12}} {T_2}$.
-
-By the above and operational semantics, $\reductionl {t_1'.m_i(v_2)}
-{s_{12}} {\subst {v_2} {x_i} {t_i}} {s_{12}} 1$.
-
-By the substitution lemma, ${\subst {v_2} {x_i} {t_i}} \in \rele
-{k-\max(j_1,j_2)-1} {\Gamma_{12}} {s_{12}} {T}$. Supposing,
-$\reductionl {\subst {v_2} {x_i} {t_i}} {s_{12}} {t'} {s'} {j_3}$,
-with $j_1 + j_2 + j_3 + 1 = j$, this completes the case, by
-monotonicity of $\mathcal{V}$.
-
-\ \\
-
-\Case \textsc{new}: $\Gamma \ts \new y c {t_b} : T$ knowing ...
-
-By operational semantics, $\reductionl {\new y c {t_b}} s {t_b} {s_b} 1$
-where $s_b = s \envplus{y \mapsto c}$. So $\reductionl {t_b} {s_b} {t'}
-{s'} {j-1}$.
-
-By induction hypotheses, $y \in \relv k {\Gamma_b} {s_b} {T_c}$ and $t_b \in
-\rele k {\Gamma_b} {s_b} {T}$.
-
-Result follows by monotonicity of $\mathcal{V}$.
-
-\end{myproof}
-
-\subsubsection{Substitution Lemma}
-
-The substitution lemma states that if (1) $v \in \relv {k_2}
-{\Gamma_{12}} {s_{12}} {T_2}$ and (2) $t \in \rele {k_1} {\Gamma_1
-  \envplus{{x} : S}} {s_1} T$ and (3) $\Gamma \ts {T_2} \sub S$ with
-(4) $x \not\in \fn(T)$ and $\Gamma_1$ extends $\Gamma$ and
-$\Gamma_{12}$ extends $\Gamma_1$ and $s_{12}$ extends $s_1$ and
-$\Gamma_1$ agrees with $s_1$ and $\Gamma_{12}$ agrees with $s_{12}$
-and a prefix of $\Gamma_{12}$ agrees with $s_1$, then $\subst {v} {x}
-{t} \in \rele {\min(k_1,k_2)} {\Gamma_{12}} {s_{12}} T$.
-
-\begin{proofsketch}
-By (1) and (3), it should hold that (5) $v \in \relv {k_2}
-{\Gamma_{12}} {s_{12}} {S}$ by the subset semantics lemma.  Since (2)
-holds, it should also hold that ${t} \in \rele {\min(k_1,k_2)}
-{\Gamma_{12} \envplus{{x} : S}} {s_{12}} T$ by the extended
-monotonicity lemma. Then, we can instantiate $x$ in the complete store
-to map to what $v$ maps to. This should be fine by (5) and
-monotonicity. Thus, ${t} \in \rele {\min(k_1,k_2)} {\Gamma_{12}
-  \envplus{{x} : S}} {s_{12} \envplus{{x} \mapsto s_{12}(v)}}
-T$. Thanks to (4), we don't actually need $x$ to be held abstract in
-the environment, because it won't occur in $T$ or its expansion (a
-{\it potential pitfall} is whether its occurrences in $t_i$ could
-still cause a check to fail through narrowing issues), so we can use
-the type of $v$ in the environment instead of $S$ for $x$: ${t} \in
-\rele {\min(k_1,k_2)} {\Gamma_{12} \envplus{{x} : {\Gamma_{12}(v)}}}
-      {s_{12} \envplus{{x} \mapsto s_{12}(v)}} T$. This implies what
-      needs to be shown.
-\end{proofsketch}
-
-\subsubsection{Subset Semantics Lemma}
-
-The subset semantics lemma states that if $v \in \relv k \Gamma s S$
-and $\Gamma \ts S <: U$, then $v \in \relv k \Gamma s U$.
-
-\begin{proofsketch}
-Because $S$ is a subtype of $U$, it should hold that the expansion of
-$S$ subsumes the expansion of $U$, when the ``self'' occurrences are
-of type $S$. Therefore, for $v \in \relv k \Gamma s U$, we have fewer
-declarations to check than for $v \in \relv k \Gamma s S$.
-
-A {\it potential pitfall} is whether some types of the expansion of
-$U$ can become non-expanding when the ``self'' occurrences are
-actually $v$ instead of just abstractly of type $S$, causing a check
-to fail. Another worry is that such a non-expanding type results from
-narrowing of a parameter type.
-%However, we believe this is fine by
-%subtyping regularity wtr to $\nswfe$ and $v \in \relv k \Gamma s S$
-%which cheks bounds and $\nswfe$ of the type labels.
-\end{proofsketch}
-
-\subsubsection{Extended Monotonicity Lemma}
-
-The extended monotonicity lemma states that if $t \in \rele k {\Gamma
-  \envplus{x : S}} s T$ then $t \in \rele j {\Gamma' \envplus{x : S}}
-{s'} T$ for $j \leq k$, $\Gamma'$ extends $\Gamma$, $s'$
-extends $s$, and $\Gamma$ agrees with $s$ and a prefix of $\Gamma'$
-agrees with $s$.
-
-\begin{proofsketch}
-For the monotonicity with regards to the step index, this follows
-directly from the definitions of $\mathcal{E}$ and $\mathcal{V}$. For
-the environment and the store, this follows by design from the
-definition of $\rels k {\Gamma \envplus{x : S}} s$. To extend the
-environment and the store for $x : S$, we can append as much as we
-want to $\Gamma$ and $s$, to get $\Gamma'$ and $s'$, and then ignore
-the last element which is for $x : S$.
-\end{proofsketch}
-
-\section{Discussion}\label{discussion}
-
-\subsection{Why No Inheritance?}\label{why-no-inheritance}
-
-In the calculus we made the deliberate choice not to model any form of
-inheritance. This is, first and foremost, to keep the calculus simple.
-Secondly, there are many different approaches to inheritance and
-mixin composition, so that it looks advantageous not to tie the basic
-calculus to a specific one. Finally, it seems that the modelization of
-inheritance lends itself to a different approach than the basic
-calculus. For the latter, we need to prove type safety of the calculus.
-One might try to do this also for a calculus with inheritance, but our
-experience suggests that this complicates the proofs considerably.  An
-alternative approach that might work better is to model inheritance
-as a form of code-reuse. Starting with an enriched type system with
-inheritance, and a translation to the basic calculus, one needs to
-show type safety wrt the translation. This might be easier than
-to prove type safety wrt reduction.
-
-\subsection{Variants of the DOT Calculus}\label{dot-variants}
-
-\subsubsection{Why limit the calculus to $\nswfe$-types?}
-
-Currently, the proof of type-safety via logical relations
-fundamentally relies on types having an expansion. However, this was
-not our original motivation for limiting the calculus to $\nswfe$-types.
-
-Originally, subtyping was not regular wrt $\nswfe$. Roughly,
-all the $\nswfe$ preconditions in subtyping were dropped. In this
-broader calculus, subtyping transitivity doesn't hold, because of the
-rule (\textsc{$\sub$-rfn}) which requires expansion of the left type.
-
-The problem is deep, as attested by this elaborate counterexample
-that is not so easily patched, and directly leads to a counterexample
-to preservation.
-
-Consider an environment where $u$ is bound to:
-\begin{align*}
-\Top & \mlrefine u {\\
-&\ \Ldecl {\mathit{Bad}} {\Bot} {u.\mathit{Bad}}\\
-&\ \Ldecl {\mathit{Good}} {\Top \refine z {\Ldecl L \Bot \Top}} {\Top \refine z {\Ldecl L \Bot \Top}}\\
-&\ \Ldecl {\mathit{Lower}} {u.\mathit{Bad} \tand u.\mathit{Good}} {u.\mathit{Good}}\\
-&\ \Ldecl {\mathit{Upper}} {u.\mathit{Good}} {u.\mathit{Bad} \tor u.\mathit{Good}}\\
-&\ \Ldecl X {u.\mathit{Lower}} {u.\mathit{Upper}}\\
-}&
-\end{align*}
-
-Now, consider the types $S$, $T$, $U$ defined in terms of $u$:
-\begin{align*}
-S &= u.\mathit{Bad} \tand u.\mathit{Good}\\
-T &= u.\mathit{Lower}\\
-U &= u.X \refine z {\Ldecl L \Bot \Top}
-\end{align*}
-We have $S \sub T$ and $T \sub U$, but we cannot derive $S \sub U$ because
-$S$ doesn't expand.
-
-Note that $u$ is realizable, since each lower bound is a subtype of
-its upper bound. So it is straightforward to turn this counterexample
-to subtyping transitivity into a counterexample to preservation:
-\begin{align*}
-&\mlnew u \ldots {
-\ \mlapp{\abs x \Top x}{
-\ \ \mlapp{\abs f {S \tfun U} f}{
-\ \ \ \mlapp{\abs f {S \tfun T} f}{
-\ \ \ \ \mlapp{\abs f {S \tfun S} f}{
-\ \ \  \ \ \abs x S x}}}}}
-\end{align*}
-
-The idea is to start with a function from $S \tfun S$ and cast it
-successively to $S \tfun T$ then $S \tfun U$. To typecheck the
-expression initially, we need to check $S \sub T$ and $T \sub
-U$. After some reduction steps, the first few casts vanish, and the
-reduced expression casts directly from $S \tfun S$ to $S \tfun U$, so
-we need to check $S \sub U$.
-
-\subsubsection{Why not include the lambda-calculus instead of methods?}
-
-Originally, the DOT calculus included the lambda-calculus, and
-explicit methods were not needed since they could be represented by a
-value label with a function type. However, the expansion of the
-function type was defined to be the empty set of declarations (like
-for $\Top$), which caused a real breach of type-safety.
-
-A concrete object could be a subtype of a function type without a
-function ever being defined. Consider:
-\begin{align*}
-&\mlnew u {\Top \refine z {\Ldecl C {\Top \tfun \Top} {\Top \tfun \Top}} \ldefs{}} {
-\mlnew f {u.C \ldefs{}} {
-\ldots
-}}
-\end{align*}
-
-Now, $f$ was a subtype of $\Top \tfun \Top$, but $\app f {(\abs x \Top
-  x)}$ was stuck (and, rightfully, didn't typecheck). But we could use
-narrowing to create a counterexample to type safety: $\app {(\abs g
-  {\Top \tfun \Top} {\app g (\abs x \Top x)})} f$.
-
-Because of this complication, we decided to drop the lambda-calculus
-from DOT, and instead introduce methods with one parameter. Like in
-Scala, functions are then just sugar for objects with a special
-method.
-
-An alternative design would have been to change the expansion of the
-function type to have a declaration for a special marker value label.
-
-\subsubsection{Why not patch the DOT calculus for preservation to hold?}\label{pres-patch}
-
-We tried! However, the resulting calculi were not elegant, and
-furthermore, we still found issues with preservation. Below, we give a
-brief summary of one failed attempt to patch the calculus for
-preservation to hold.
-
-Because many of the counterexamples to preservation are related to
-narrowing, we tried to make widening an explicit operation and change
-rules with implicit relaxations (\textsc{msel} and \textsc{new}) to be
-strict. From a typing perspective, the change was straightforward, but
-reduction became more complicated and dependent on typing because the
-type information in widenings needed to be propagated correctly.
-
-We added path equality provisions in the subtyping rules, in the same
-spirit as the Tribe calculus~\cite{tribe}.
-
-Unfortunately, these two patches interacted badly, and we were left
-with a disturbing counterexample to type safety.
-
-\begin{align*}
-&\mlnew a {\Top \refine z {\Ldecl C \Bot {\Top \refine z {\Ldecl D \Bot {z.X}, \Ldecl X \Bot \Top}}}}{
-\mlnew b {{a.C} \refine z {\Ldecl X \Bot \Bot}}}{
-\mlnew c {a.C}}{
-\mlnew d {{\cast {a.C} b}.D}}{
-\app {(\abs x \Bot {x.\mathit{foo}})} d
-}
-\end{align*}
-
-Notice that $d$ has type $\Bot$ if the cast on $b$ is ignored. This
-example didn't typecheck initially because the path-equality
-provisions only applied when objects are in the store, so the
-application was not well-typed. But if we started preservation in a store
-which had $a$, $b$, $c$ and $d$ then the application type-checked,
-because, through one of the path-equality provision, we could find that
-the type of $d$ was a subtype of $\Bot$. Now, of course, when we got to
-$d.\mathit{foo}$, reduction failed.
-
-\subsection{Related Work}
-
-In addition to Scala's previous models~\cite{nuObj,FS}, several
-calculi present some form of path-dependent types.
-
-The {\it vc} calculus~\cite{vc} models virtual classes with
-path-dependent types. {\it vc} restricts paths to start with ``this'',
-though it provides a way (``out'') to refer to the enclosing object.
-
-The Tribe calculus~\cite{tribe} builds an ownership types
-system~\cite{tribalo} on top of a core calculus which models virtual
-classes. The soundness proof for the core calculus seems to be tied to
-the ownership types system.
-
-Some ML-style module systems~\cite{homl,mixinml} have a form of
-stratified path-dependent types. Because of the stratification,
-recursion is not allowed. In MixML~\cite{mixml} like in Scala, this
-restriction is lifted.
-
-\section{Conclusion}\label{conclusion}
-
-We have presented DOT, a calculus aimed as a new foundation of Scala
-and languages like it. DOT features path-dependent types, refinement
-types, and abstract type members.
-
-Proving the DOT calculus type-safe has been an interesting
-adventure. We have shown that DOT does not satisfy preservation (also
-known as subject-reduction). However, the standard theorems of
-preservation and progress are just one way to prove type safety. We
-have sketched a plausible proof of type safety using the powerful
-method of logical relations.
-
-%\appendix
-%\section{Appendix Title}
-
-\acks
-
-We thank Amal Ahmed for many discussions and insights about applying
-logical relations to DOT. We thank Donna Malayeri and Geoffrey
-Washburn for preliminary work on DOT. We thank Tiark Rompf for helpful
-comments.
-
-\bibliographystyle{abbrvnat}
-\bibliography{dot}
-
-\end{document}

### 2012-08-30 doc/candidacy/report.tex
minor
macros for displaying Scala code
cite scala overview

@@
 \documentclass[10pt,journal,a4paper]{IEEEtran}
+
+\input{macros}
+
 % If IEEEtran.cls has not been installed into the LaTeX system files,
@@
 \IEEEPARstart{T}{he} aim of this research is to propose a new sound
-and tractable type-theoretic foundation of Scala and languages like
+and tractable type-theoretic foundation of Scala~\cite{ScalaOverview} and languages like
 it: the Dependent Object Types calculus (DOT). The properties we are
@@
 soundness proof sketch seems more plausible but then is tied to the
-ownership system formalism. Decidability and completeness of subtyping
-has not yet been established.
+ownership system formalism. Decidability of subtyping, and hence
+typechecking, has not yet been established.
 

### 2012-08-30 doc/candidacy/report.tex
just copied the context-free syntax comment from nuObj. Requires editing

@@
 \input{macros}
+\input{nuobj}
 
@@
 
+%% TODO beg edit
+A {\em term} denotes an object or a class. It can be of the following
+five forms.
+\begin{itemize}
+\item A {\em simple name} $x$, which denotes an object.
+\item A {\em selection} $t.l$, which can denote either an object or a class.
+\item An {\em object creation} \mbox{$\newin{x \larrowc t}{u}$},
+which defines a fresh instance $x$ of class $t$.
+The scope of this object is the term $u$.
+\item 
+A {\em class template} $\cls{x: S \sep \tup d}$ where $\tup d$ is a
+sequence of {\em definitions} which associate term labels with values
+and type labels with types. This acts as a template to construct
+objects with the members defined by the definitions.  The name $x$ of
+type $S$ stands for ``self'', i.e.\ the object being constructed from
+the template.  Its scope is the definition sequence $\tup d$.  A term
+or type can refer via $x.l$ \comment{or \mbox{$x.\type \Tsel L$}} to some other
+member of that object. No textual sequence constraint applies to such
+references; in particular it is possible that a binding refers to
+itself or to bindings defined later in the same record.  This
+distinguishes our type system from earlier type systems for records
+\cite{cardelli-mitchell:records} or modules \cite{harper94type}.
+\item 
+A {\em mixin composition} $t \withsub{S} u$, which forms a combined
+class from the two classes to which $t$ and $u$ evaluate. Here, $S$
+is the type of ``self'' in the combined class. 
+\end{itemize}
+A {\em value} is a simple name or a class template.
+A {\em path} $p$ is a name $x$ followed
+by a possibly empty sequence of selections, e.g.\ $x . l_1 .; \ldots;
+. l_n$.
+
+The syntax of {\em types} in our system closely follows the syntax of
+ terms.  A type can be of the following five forms.
+\begin{itemize}
+\item A {\em singleton type} $p.\type$. This type represents the set of
+values which has as only element the object referenced by the path
+$p$. Singleton types are the only way a type can depend on a term in
+$\NTO$.
+\item 
+A {\em type selection} $T\Tsel L$, which represents the type component labelled $L$ of type $T$.
+\item
+A {\em record type} $\obj{x \sep \tup D}$ where $\tup D$ is a sequence
+of {\em declarations} which can be value bindings or type bindings.  A
+{\em value binding} $l: T$ associates a term label $l$ with its type
+$T$.  {\em Type bindings} come in three different forms:
+
+First, the binding $L = T$ defines $L$ to be an {\em alias} for $T$. 
+Second, the binding $L \expand T$ defines $L$ to be a {\em new type} which
+{\em expands} to type $T$. That is, $L$ is a subtype of $T$ which has
+exactly the members defined by $T$; furthermore, one can create
+objects of type $L$ from a class which defines all members of $T$.
+Third, the binding $L <: T$ defines $L$ to be an {\em abstract type}
+which is known to be a subtype of its bound $T$.
+
+We let the meta-variable $\preceq$ range over $=$ and $\expand$, and
+let $\tbind$ range over $=$, $\expand$, and $<:$.  The name $x$ stands
+for ``self''; its type is assumed to be the record type itself. We let
+the letter $R$ range over record types.
+\item
+A {\em compound type} $T \with U$. This type contains all members of
+types $T$ and $U$. The subtyping relation for compound types is the
+same as the one for intersection types \cite{barendregt-coppo:filter},
+but the formation rules are more restrictive.  Where $T$ and $U$ have
+a member with the same label, the compound type contains the member
+defined in $U$. That member definition must be more specific
+(see Section~\ref{sec:type-system}) than the corresponding member
+definition in $T$.
+\item
+A {\em class type} $\cls{x: S \sep \tup D}$, which contains as values
+classes that instantiate to objects of type $\obj{x \sep \tup D}$, or
+some subtype of it.  $x$ is again the name for ``self''.  It now comes
+with an explicit type $S$ which may be different from $\obj{x \sep
+\tup D}$. Definitions in $S$ which are missing from $\tup D$ play the
+role of abstract members. Such members can be referred to from other
+definitions in the class, but they are not defined in the class
+itself. Instead, these members must be defined in other classes which
+are composed with the class itself in a mixin composition.
+Definitions which are present in $\tup D$ but missing in $S$ play in
+some sense the role of non-virtual members -- they are not referred to
+via ``self'' from inside the class, so overriding them does not change
+existing behavior. Definitions present in both $S$ and $\tup D$ play
+the role of virtual members.
+\end{itemize}
+%% TODO end edit
+
 In terms of theoretical results, the paper confirms type soundness of

### 2012-08-30 doc/candidacy/report.tex
summary of context-free syntax of nuObj

@@
 
-%% TODO beg edit
+The terms in ${\nu}Obj$ denotes objects or classes. They consist of
+variables $x$ denoting objects, selections $t.l$, object creations
+\mbox{$\newin{x \larrowc t}{u}$} defining an object $x$ of class $t$
+whose scope is the term $u$, class templates $\cls{x: S \sep \tup d}$,
+and mixing compositions $t \withsub{S} u$ denoting a class of self
+type $S$ combined from two classes $t$ and $u$. A value is a variable
+or a class template. A path is a variable followed by a possibly empty
+sequence of selections.
+
+The types in ${\nu}Obj$ consists of singleton types $p.\type$, type
+selections $T\Tsel L$, record types $\obj{x \sep \tup D}$, compound
+types $T \with U$, and class types $\cls{x: S \sep \tup D}$.
+
+A class template $\cls{x: S \sep \tup d}$ is closely related to a
+translucent sum, except that it needs to be instantiated into an
+object while a translucent sum is a module. Like in a translucent sum,
+$\tup d$ binds term labels to values and type labels to types. The $x$
+of type $S$ stands for the self, i.e.\ the object being constructed
+from the template. Its scope is the bindings $\tup d$. Since the self
+is explicitly typed, the bindings in class templates can freely refer
+to each other (and even to themselves) via $x$, while recall that, in
+translucent sums, only later fields can depend on earlier ones.
+
+Contrasting concrete type bindings $\tup d$ used in class template
+terms with type declarations $\tup D$ used in record and class types,
+the possible type binders in type declarations are type aliases $=$,
+new types $\expand$ and abstract types $<:$. Only the first two are
+also concrete type binders available in type bindings of class
+templates.
+
+\comment{
 A {\em term} denotes an object or a class. It can be of the following
@@
 \end{itemize}
-%% TODO end edit
+}
 

### 2012-08-30 doc/candidacy/report.tex
a bit more

@@
 from the template. Its scope is the bindings $\tup d$. Since the self
-is explicitly typed, the bindings in class templates can freely refer
-to each other (and even to themselves) via $x$, while recall that, in
-translucent sums, only later fields can depend on earlier ones.
+is explicitly typed, the bindings in class templates can contractively
+refer to each other (and even to themselves) via $x$, while recall
+that, in translucent sums, only later fields can depend on earlier
+ones.
 
@@
 also concrete type binders available in type bindings of class
-templates.
+templates. A consequence is that classes with abstract type members
+cannot be instantiated, because each type declaration of the self type
+needs to have a corresponding type binding when an object is
+instantiated.
+
+Mixing composition is not commutative. When expanding a type to a
+record type, a compound type $R_1 \with R_2$ expands to a record type
+containing the concatenation of the declarations in $R_1$ and $R_2$,
+but if some label is defined in both $R_1$ and $R_2$, the definition
+in $R_2$ overrides (and so must be more specific than) the definition
+in $R_1$. Though this break of commutativity is justified from an
+implementation standpoint, it is awkward from a typing standpoint: in
+particular, it means that least upper bounds and greatest lower bounds
+do not always exist.
 
@@
 
-In terms of theoretical results, the paper confirms type soundness of
-the calculus and the undecidability of type checking.
+In terms of theoretical results, the paper confirms the type soundness
+of the calculus and the undecidability of its type checking.
 

### 2012-08-30 doc/candidacy/report.tex
remove commented out copy of nuObj paper

@@
 
-\comment{
-A {\em term} denotes an object or a class. It can be of the following
-five forms.
-\begin{itemize}
-\item A {\em simple name} $x$, which denotes an object.
-\item A {\em selection} $t.l$, which can denote either an object or a class.
-\item An {\em object creation} \mbox{$\newin{x \larrowc t}{u}$},
-which defines a fresh instance $x$ of class $t$.
-The scope of this object is the term $u$.
-\item 
-A {\em class template} $\cls{x: S \sep \tup d}$ where $\tup d$ is a
-sequence of {\em definitions} which associate term labels with values
-and type labels with types. This acts as a template to construct
-objects with the members defined by the definitions.  The name $x$ of
-type $S$ stands for ``self'', i.e.\ the object being constructed from
-the template.  Its scope is the definition sequence $\tup d$.  A term
-or type can refer via $x.l$ \comment{or \mbox{$x.\type \Tsel L$}} to some other
-member of that object. No textual sequence constraint applies to such
-references; in particular it is possible that a binding refers to
-itself or to bindings defined later in the same record.  This
-distinguishes our type system from earlier type systems for records
-\cite{cardelli-mitchell:records} or modules \cite{harper94type}.
-\item 
-A {\em mixin composition} $t \withsub{S} u$, which forms a combined
-class from the two classes to which $t$ and $u$ evaluate. Here, $S$
-is the type of ``self'' in the combined class. 
-\end{itemize}
-A {\em value} is a simple name or a class template.
-A {\em path} $p$ is a name $x$ followed
-by a possibly empty sequence of selections, e.g.\ $x . l_1 .; \ldots;
-. l_n$.
-
-The syntax of {\em types} in our system closely follows the syntax of
- terms.  A type can be of the following five forms.
-\begin{itemize}
-\item A {\em singleton type} $p.\type$. This type represents the set of
-values which has as only element the object referenced by the path
-$p$. Singleton types are the only way a type can depend on a term in
-$\NTO$.
-\item 
-A {\em type selection} $T\Tsel L$, which represents the type component labelled $L$ of type $T$.
-\item
-A {\em record type} $\obj{x \sep \tup D}$ where $\tup D$ is a sequence
-of {\em declarations} which can be value bindings or type bindings.  A
-{\em value binding} $l: T$ associates a term label $l$ with its type
-$T$.  {\em Type bindings} come in three different forms:
-
-First, the binding $L = T$ defines $L$ to be an {\em alias} for $T$. 
-Second, the binding $L \expand T$ defines $L$ to be a {\em new type} which
-{\em expands} to type $T$. That is, $L$ is a subtype of $T$ which has
-exactly the members defined by $T$; furthermore, one can create
-objects of type $L$ from a class which defines all members of $T$.
-Third, the binding $L <: T$ defines $L$ to be an {\em abstract type}
-which is known to be a subtype of its bound $T$.
-
-We let the meta-variable $\preceq$ range over $=$ and $\expand$, and
-let $\tbind$ range over $=$, $\expand$, and $<:$.  The name $x$ stands
-for ``self''; its type is assumed to be the record type itself. We let
-the letter $R$ range over record types.
-\item
-A {\em compound type} $T \with U$. This type contains all members of
-types $T$ and $U$. The subtyping relation for compound types is the
-same as the one for intersection types \cite{barendregt-coppo:filter},
-but the formation rules are more restrictive.  Where $T$ and $U$ have
-a member with the same label, the compound type contains the member
-defined in $U$. That member definition must be more specific
-(see Section~\ref{sec:type-system}) than the corresponding member
-definition in $T$.
-\item
-A {\em class type} $\cls{x: S \sep \tup D}$, which contains as values
-classes that instantiate to objects of type $\obj{x \sep \tup D}$, or
-some subtype of it.  $x$ is again the name for ``self''.  It now comes
-with an explicit type $S$ which may be different from $\obj{x \sep
-\tup D}$. Definitions in $S$ which are missing from $\tup D$ play the
-role of abstract members. Such members can be referred to from other
-definitions in the class, but they are not defined in the class
-itself. Instead, these members must be defined in other classes which
-are composed with the class itself in a mixin composition.
-Definitions which are present in $\tup D$ but missing in $S$ play in
-some sense the role of non-virtual members -- they are not referred to
-via ``self'' from inside the class, so overriding them does not change
-existing behavior. Definitions present in both $S$ and $\tup D$ play
-the role of virtual members.
-\end{itemize}
-}
-
 In terms of theoretical results, the paper confirms the type soundness

### 2012-08-30 doc/candidacy/report.tex
more cautious formulation

@@
 implementation standpoint, it is awkward from a typing standpoint: in
-particular, it means that least upper bounds and greatest lower bounds
-do not always exist.
+particular, least upper bounds and greatest lower bounds
+do not always exist in Scala.
 

### 2012-08-30 doc/candidacy/report.tex
spellcheck

@@
 declarations. They differ from traditional records, because they can
-contain types and because later fields can depend on ealier
+contain types and because later fields can depend on earlier
 ones. Translucent sums are ordinary values. Since they model modules,
@@
 variant of Tribe with simpler types, cross-family inheritance and
-generics is presented and a formalisation of ownership types built
+generics is presented and a formalization of ownership types built
 upon it.
@@
 polymorphism, essential aspects of the ML-like module systems
-including sharing consraints and higher-order functors, and System
+including sharing constraints and higher-order functors, and System
 $F_{<:}$~\cite{tapl}. The calculus can be used as a basis for unifying
@@
 
+Let's now illustrate with a few examples the correspondences outlined
+earlier. %TODO
+
 In terms of theoretical results, the paper confirms the type soundness

### 2012-08-30 doc/candidacy/report.tex
some corrections

@@
 approaches~\cite{nuObj,FS}, we make no attempt to model inheritance or
-mixing composition. Furthermore, the calculus is more normative than
+mixin composition. Furthermore, the calculus is more normative than
 descriptive, in that it does not precisely describe what's currently
@@
 whose scope is the term $u$, class templates $\cls{x: S \sep \tup d}$,
-and mixing compositions $t \withsub{S} u$ denoting a class of self
+and mixin compositions $t \withsub{S} u$ denoting a class of self
 type $S$ combined from two classes $t$ and $u$. A value is a variable
@@
 also concrete type binders available in type bindings of class
-templates. A consequence is that classes with abstract type members
-cannot be instantiated, because each type declaration of the self type
-needs to have a corresponding type binding when an object is
-instantiated.
-
-Mixing composition is not commutative. When expanding a type to a
+templates.
+
+A class type $\cls{x: S \sep \tup D}$ contains as values classes that
+instantiate to objects of (sub)type $\obj{x \sep \tup D}$. The
+explicit self type $S$ may be different from $\obj{x \sep \tup
+  D}$. Declarations in $S$ which are not in $\tup D$ play the role of
+abstract members -- they must be defined by mixin composition in order
+for the class to be instantiated.
+
+Mixin composition is not commutative. When expanding a type to a
 record type, a compound type $R_1 \with R_2$ expands to a record type
@@
 Let's now illustrate with a few examples the correspondences outlined
-earlier. %TODO
+earlier.
+
+First, we show how to encode monomorphic functions into the
+calculus. The paper goes further and generalizes the encoding to
+system $F_{<:}$
 

### 2012-08-30 doc/candidacy/report.tex
ex

@@
 calculus. The paper goes further and generalizes the encoding to
-system $F_{<:}$
+system $F_{<:}$. A $\lambda$-abstraction $\lambda ( x : T )\; t$ is
+represented as a class with an abstract member \code!arg! for the
+function argument:
+\begin{lstlisting}
+[x: {arg: T}| fun = [res = $t'$]]
+\end{lstlisting}
+$t'$ corresponds to term $t$ in which all occurrences of $x$ are
+replaced by \code!$x$.arg!. We cannot directly access \code!arg! on
+the right-hand side of \code!fun!: this wouldn't be ``contractive'',
+because \code!arg! might not be initialised before \code!fun!. Hence,
+we pack $t'$ in another class to ensure contractiveness. Now, an
+application $g(e)$ gets decomposed into three subsequent steps:
+\begin{lstlisting}
+$\nu\,$g$_{app}$  <- g & [arg = e];
+$\nu\,$g$_{eval}$ <- g$_{app}$.fun;
+g$_{eval}$.res
+\end{lstlisting}
+First we instantiate function \code!g! with a concrete argument
+yielding a thunk \code!g$_{app}$!. Then we evaluate this thunk by
+creating an instance \code!g$_{eval}$! of it. Finally we extract the
+result by querying field \code!res! of \code!g$_{eval}$!.
 

### 2012-08-30 doc/candidacy/macros.tex
ex
basic example for tribe

@@
   {<:}{$<:\;$}{1}
+  {<-}{$\leftarrow\;$}{2}
 }
@@
 \lstset{
-    xleftmargin=2em,%
+    xleftmargin=1em,%
     framesep=5pt,%
@@
     tabsize=2,%
-    basicstyle=\small\tt,% \small\tt
+    basicstyle=\footnotesize\tt,% \small\tt
     commentstyle=\textit,%

### 2012-08-30 doc/candidacy/report.tex
added TODOs for examples
basic example for tribe

@@
 
+% TODO: example of translucent module and its utility. woudl be nice to re-use this example in nuObj to show how it can be encoded
+
 In terms of theoretical results, Lillibridge shows in his thesis that
@@
 upon it.
+
+Let's now re-use an example from the paper, and discuss how we would
+emulate it in Scala.
+\begin{lstlisting}
+class Graph {
+  class Node {
+    Edge connect(Node other) {
+      return new Edge(this, other);
+    }
+  }
+  class Edge {
+    Node from, to;
+    Edge(Node f, Node t) { from = f; to = t; }
+  }
+}
+
+class ColouredGraph extends Graph { // subclassing
+  class Node {                      // further binding
+    Colour nodeColour;
+  }
+}
+
+class Library {
+  int distance(Graph.Node n1, n1.out.Node n2) { ... }
+
+  e.out.Edge copyEdge(Graph.Edge e) {
+    e.out.Node from = e.from;
+    e.out.Node to = e.to;
+    new e.out.Edge(from, to);
+  }
+}
+\end{lstlisting}
+
+The ``tests'' are:
+\begin{lstlisting}
+final ColouredGraph cg1, cg2;
+cg1.Node cn1, cn3;
+cg2.Node cn2;
+cn1.connect(cn3); // Type Correct
+cn2.connect(cn3); // Type Error!!!
+\end{lstlisting}
+
+Now, we can emulate this Tribe example in Scala at the expense of some
+synctactic overhead, since further binding and ``out'' selection are
+not native.
+\lstinputlisting{tribe.scala}
 
@@
 
+% TODO: example of module encoding... perhaps in Scala rather than nuObj?
+
 In terms of theoretical results, the paper confirms the type soundness

### 2012-08-30 doc/candidacy/tribe.scala

@@ -1,0 +1,51 @@
+abstract class AbstractGraph { graph =>
+  type Node <: BasicNode
+  type Edge <: BasicEdge
+
+  def newNode(): Node
+  def newEdge(from: Node, to: Node): Edge
+
+  class BasicNode { this: Node =>
+    def connect(other: Node) = newEdge(this, other)
+    val out : graph.type = graph
+  }
+
+  class BasicEdge(val from: Node, val to: Node) {
+    val out : graph.type = graph
+  }
+}
+
+class BasicGraph extends AbstractGraph {
+  type Node = BasicNode
+  type Edge = BasicEdge
+
+  def newNode(): Node =
+    new BasicNode()
+  def newEdge(from: Node, to: Node): Edge =
+    new BasicEdge(from, to)
+}
+
+class ColouredGraph extends AbstractGraph {
+  type Node = ColouredNode
+  type Edge = BasicEdge
+
+  def newNode(): Node =
+    new ColouredNode("blue")
+  def newEdge(from: Node, to: Node): Edge =
+    new BasicEdge(from, to)
+
+  class ColouredNode(var colour: String) extends BasicNode
+}
+
+object Test extends App {
+  val cg1, cg2 = new ColouredGraph()
+  val cn1, cn3 = cg1.newNode()
+  val cn2 = cg2.newNode()
+  cn1.connect(cn3)
+  // cn2.connect(cn3)
+  // error: type mismatch;
+  // found   : Test.cg1.Node
+  //  (which expands to)  Test.cg1.ColouredNode
+  // required: Test.cg2.Node
+  //  (which expands to)  Test.cg2.ColouredNode
+}

### 2012-08-30 doc/candidacy/tribe.scala
implement lib for tribe ex. give up on emulating out.

@@ -1,4 +1,4 @@
-abstract class AbstractGraph { graph =>
+abstract class Graph {
   type Node <: BasicNode
   type Edge <: BasicEdge
 
@@ -7,15 +7,12 @@ abstract class AbstractGraph { graph =>
 
   class BasicNode { this: Node =>
     def connect(other: Node) = newEdge(this, other)
-    val out : graph.type = graph
   }
 
-  class BasicEdge(val from: Node, val to: Node) {
-    val out : graph.type = graph
-  }
+  class BasicEdge(val from: Node, val to: Node)
 }
 
-class BasicGraph extends AbstractGraph {
+class BasicGraph extends Graph {
   type Node = BasicNode
   type Edge = BasicEdge
 
@@ -25,7 +22,7 @@ class BasicGraph extends AbstractGraph {
     new BasicEdge(from, to)
 }
 
-class ColouredGraph extends AbstractGraph {
+class ColouredGraph extends Graph {
   type Node = ColouredNode
   type Edge = BasicEdge
 
@@ -37,11 +34,17 @@ class ColouredGraph extends AbstractGraph {
   class ColouredNode(var colour: String) extends BasicNode
 }
 
+abstract class Library {
+  def distance(g: Graph)(n1: g.Node, n2: g.Node): Int
+  def copyEdge(g: Graph)(e: g.Edge): g.Edge =
+    g.newEdge(e.from, e.to)
+}
+
 object Test extends App {
   val cg1, cg2 = new ColouredGraph()
   val cn1, cn3 = cg1.newNode()
   val cn2 = cg2.newNode()
-  cn1.connect(cn3)
+  val e = cn1.connect(cn3)
   // cn2.connect(cn3)
   // error: type mismatch;
   // found   : Test.cg1.Node

### 2012-08-30 doc/candidacy/tribe.scala
out works now

@@ -1,42 +1,57 @@
-abstract class Graph {
-  type Node <: BasicNode
-  type Edge <: BasicEdge
+abstract class Graph { graph =>
+  type Node <: AbstractNode
+  type Edge <: AbstractEdge
 
   def newNode(): Node
   def newEdge(from: Node, to: Node): Edge
 
-  class BasicNode { this: Node =>
-    def connect(other: Node) = newEdge(this, other)
+  class AbstractEntity {
+    val out: graph.type = graph
   }
 
-  class BasicEdge(val from: Node, val to: Node)
+  abstract class AbstractNode extends AbstractEntity { this: Node =>
+    def connect(other: out.Node): out.Edge = newEdge(this, other)
+  }
+
+  abstract class AbstractEdge extends AbstractEntity { this: Edge =>
+    val from: out.Node
+    val to: out.Node
+  }
 }
 
 class BasicGraph extends Graph {
-  type Node = BasicNode
-  type Edge = BasicEdge
+  class Node extends AbstractNode
+  class Edge(val from: Node, val to: Node) extends AbstractEdge
 
   def newNode(): Node =
-    new BasicNode()
+    new Node()
   def newEdge(from: Node, to: Node): Edge =
-    new BasicEdge(from, to)
+    new Edge(from, to)
 }
 
 class ColouredGraph extends Graph {
-  type Node = ColouredNode
-  type Edge = BasicEdge
+  class Node(var colour: String) extends AbstractNode
+  class Edge(val from: Node, val to: Node) extends AbstractEdge
 
   def newNode(): Node =
-    new ColouredNode("blue")
+    new Node("blue")
   def newEdge(from: Node, to: Node): Edge =
-    new BasicEdge(from, to)
-
-  class ColouredNode(var colour: String) extends BasicNode
+    new Edge(from, to)
 }
 
-abstract class Library {
-  def distance(g: Graph)(n1: g.Node, n2: g.Node): Int
-  def copyEdge(g: Graph)(e: g.Edge): g.Edge =
+object Library {
+  def distance(n1: Graph#Node)(n2: n1.out.Node): Int = {
+    val e: n1.out.Edge = n1.connect(n2)
+    0 // ...
+  }
+  def copyEdge(e: Graph#Edge): e.out.Edge =
+    e.out.newEdge(e.from, e.to)
+
[…]

### 2012-08-30 doc/candidacy/macros.tex
compact code

@@
     tabsize=2,%
-    basicstyle=\footnotesize\tt,% \small\tt
+    basicstyle=\scriptsize\tt,% \small\tt
     commentstyle=\textit,%

### 2012-08-30 doc/candidacy/report.tex
implement lib for tribe ex. give up on emulating out.
out works now
compact code

@@
 synctactic overhead, since further binding and ``out'' selection are
-not native.
+not native. Emulating ``out'' faithfully involves a bit of hackery in
+making Scala recognize that, for example, \code{e.from.type} is a
+subtype of \code{e.out.Node}. As noted in the paper, using ``out'' in
+library code obscures the symmetry between the parameters (for
+example, in the \code{distance} method below) but allows the family
+object to be recovered without being explicitly passed around
+(contrast with the method \code{distance2}).
 \lstinputlisting{tribe.scala}

### 2012-08-30 doc/candidacy/tribe.scala

@@ -1,27 +1,28 @@
 abstract class Graph { graph =>
-  type Node <: AbstractNode
-  type Edge <: AbstractEdge
+  type Node <: ANode
+  type Edge <: AEdge
 
   def newNode(): Node
   def newEdge(from: Node, to: Node): Edge
 
-  class AbstractEntity {
+  class WithOut {
     val out: graph.type = graph
   }
 
-  abstract class AbstractNode extends AbstractEntity { this: Node =>
-    def connect(other: out.Node): out.Edge = newEdge(this, other)
+  abstract class ANode extends WithOut { this: Node =>
+    def connect(other: out.Node): out.Edge =
+      newEdge(this, other)
   }
 
-  abstract class AbstractEdge extends AbstractEntity { this: Edge =>
+  abstract class AEdge extends WithOut { this: Edge =>
     val from: out.Node
     val to: out.Node
   }
 }
 
 class BasicGraph extends Graph {
-  class Node extends AbstractNode
-  class Edge(val from: Node, val to: Node) extends AbstractEdge
+  class Node extends ANode
+  class Edge(val from: Node, val to: Node) extends AEdge
 
   def newNode(): Node =
     new Node()
@@ -30,8 +31,8 @@ class BasicGraph extends Graph {
 }
 
 class ColouredGraph extends Graph {
-  class Node(var colour: String) extends AbstractNode
-  class Edge(val from: Node, val to: Node) extends AbstractEdge
+  class Node(var colour: String) extends ANode
+  class Edge(val from: Node, val to: Node) extends AEdge
 
   def newNode(): Node =
     new Node("blue")

### 2012-08-31 doc/candidacy/tribe.scala
simpler (too?) tribe emulation

@@ -1,43 +1,23 @@
-abstract class Graph { graph =>
-  type Node <: ANode
-  type Edge <: AEdge
-
-  def newNode(): Node
-  def newEdge(from: Node, to: Node): Edge
-
+class Graph { graph =>
   class WithOut {
     val out: graph.type = graph
   }
 
-  abstract class ANode extends WithOut { this: Node =>
+  class Node extends WithOut {
     def connect(other: out.Node): out.Edge =
-      newEdge(this, other)
+      new out.Edge(this, other)
   }
 
-  abstract class AEdge extends WithOut { this: Edge =>
-    val from: out.Node
-    val to: out.Node
+  class Edge(f: Node, t: Node) extends WithOut {
+    val from: out.Node = f
+    val to: out.Node = t
   }
-}
-
-class BasicGraph extends Graph {
-  class Node extends ANode
-  class Edge(val from: Node, val to: Node) extends AEdge
-
-  def newNode(): Node =
-    new Node()
-  def newEdge(from: Node, to: Node): Edge =
-    new Edge(from, to)
 }
 
 class ColouredGraph extends Graph {
-  class Node(var colour: String) extends ANode
-  class Edge(val from: Node, val to: Node) extends AEdge
-
-  def newNode(): Node =
-    new Node("blue")
-  def newEdge(from: Node, to: Node): Edge =
-    new Edge(from, to)
+  class Node extends super.Node {
+    var colour: String = "blue"
+  }
 }
 
 object Library {
@@ -46,27 +26,24 @@ object Library {
     0 // ...
   }
   def copyEdge(e: Graph#Edge): e.out.Edge =
-    e.out.newEdge(e.from, e.to)
+    new e.out.Edge(e.from, e.to)
 
   def distance2(g: Graph)(n1: g.Node, n2: g.Node): Int = {
     val e: g.Edge = n1.connect(n2)
     0 // ...
   }
   def copyEdge2(g: Graph)(e: g.Edge): g.Edge =
-    g.newEdge(e.from, e.to)
+    new g.Edge(e.from, e.to)
 }
 
 object Test extends App {
   val cg1, cg2 = new ColouredGraph()
-  val cn1, cn3 = cg1.newNode()
-  val cn2 = cg2.newNode()
+  val cn1, cn3 = new cg1.Node()
+  val cn2 = new cg2.Node()
   val e = cn1.connect(cn3)
+  // this is a type mismatch
[…]

### 2012-08-31 doc/candidacy/report.tex
typo

@@
 Now, we can emulate this Tribe example in Scala at the expense of some
-synctactic overhead, since further binding and ``out'' selection are
+syntactic overhead, since further binding and ``out'' selection are
 not native. Emulating ``out'' faithfully involves a bit of hackery in

### 2012-08-31 doc/candidacy/tribe2.scala
the canonical encoding of virtual classes using virtual types as explained http://lambda-the-ultimate.org/node/1960#comment-23826

@@ -1,0 +1,64 @@
+trait GraphBase { graph =>
+  type Node <: NodeBase
+  type Edge <: EdgeBase
+  def newEdge(from: Node, to: Node): Edge
+
+  trait WithOut {
+    val out: graph.type = graph
+  }
+  trait NodeBase extends WithOut { this: Node =>
+    def connect(other: out.Node): out.Edge =
+      newEdge(this, other)
+  }
+  trait EdgeBase extends WithOut { this: Edge =>
+    val from: out.Node
+    val to: out.Node
+  }
+}
+
+class Graph extends GraphBase {
+  class Node extends NodeBase
+  class Edge(val from: Node, val to: Node) extends EdgeBase
+  def newEdge(from: Node, to: Node) = new Edge(from, to)
+}
+
+class ColouredGraph extends GraphBase {
+  class Node(val colour: String) extends NodeBase
+  class Edge(val from: Node, val to: Node) extends EdgeBase
+  def newEdge(from: Node, to: Node) = new Edge(from, to)
+}
+
+object Library {
+  def distance(n1: GraphBase#Node)(n2: n1.out.Node): Int = {
+    val e: n1.out.Edge = n1.connect(n2)
+    0 // ...
+  }
+  def copyEdge(e: GraphBase#Edge): e.out.Edge =
+    e.out.newEdge(e.from, e.to)
+
+  def distance2(g: GraphBase)(n1: g.Node, n2: g.Node): Int = {
+    val e: g.Edge = n1.connect(n2)
+    0 // ...
+  }
+  def copyEdge2(g: GraphBase)(e: g.Edge): g.Edge =
+    g.newEdge(e.from, e.to)
+}
+
+object Test extends App {
+  val cg1, cg2 = new ColouredGraph()
+  val cn1 = new cg1.Node("blue")
+  val cn2 = new cg2.Node("green")
+  val cn3 = new cg1.Node("red")
+  val e = cn1.connect(cn3)
+  // this is a type mismatch
+  // cn2.connect(cn3)
+
+  val d1 = Library.distance(cn1)(cn3)
+  val d2 = Library.distance2(cg1)(cn1, cn3)
+  val e1 = Library.copyEdge(e)
+  val e2 = Library.copyEdge2(cg1)(e)
+  // these are all type mismatches
+  // Library.distance(cn1)(cn2)
+  // Library.distance2(cg2)(cn1, cn3)
+  // Library.copyEdge2(cg2)(e)
+}

### 2012-08-31 doc/candidacy/tribe.scala
playing with the diffs

@@ -12,12 +12,16 @@ class Graph { graph =>
     val from: out.Node = f
     val to: out.Node = t
   }
+
+  def newNode() = new Node
 }
 
 class ColouredGraph extends Graph {
   class Node extends super.Node {
     var colour: String = "blue"
   }
+
+  override def newNode() = new Node
 }
 
 object Library {

### 2012-08-31 doc/candidacy/tribe2.scala

@@ -1,6 +1,7 @@
 trait GraphBase { graph =>
   type Node <: NodeBase
   type Edge <: EdgeBase
+  def newNode(): Node
   def newEdge(from: Node, to: Node): Edge
 
   trait WithOut {
@@ -19,12 +20,14 @@ trait GraphBase { graph =>
 class Graph extends GraphBase {
   class Node extends NodeBase
   class Edge(val from: Node, val to: Node) extends EdgeBase
+  def newNode() = new Node
   def newEdge(from: Node, to: Node) = new Edge(from, to)
 }
 
 class ColouredGraph extends GraphBase {
   class Node(val colour: String) extends NodeBase
   class Edge(val from: Node, val to: Node) extends EdgeBase
+  def newNode() = new Node("blue")
   def newEdge(from: Node, to: Node) = new Edge(from, to)
 }
 

### 2012-08-31 doc/candidacy/tribe.scala
use the canonical more complicated encoding

@@ -1,49 +1,57 @@
-class Graph { graph =>
-  class WithOut {
+trait GraphBase { graph =>
+  type Node <: NodeBase
+  type Edge <: EdgeBase
+  def newNode(): Node
+  def newEdge(from: Node, to: Node): Edge
+
+  trait WithOut {
     val out: graph.type = graph
   }
-
-  class Node extends WithOut {
+  trait NodeBase extends WithOut { this: Node =>
     def connect(other: out.Node): out.Edge =
-      new out.Edge(this, other)
+      newEdge(this, other)
   }
-
-  class Edge(f: Node, t: Node) extends WithOut {
-    val from: out.Node = f
-    val to: out.Node = t
+  trait EdgeBase extends WithOut { this: Edge =>
+    val from: out.Node
+    val to: out.Node
   }
-
-  def newNode() = new Node
 }
 
-class ColouredGraph extends Graph {
-  class Node extends super.Node {
-    var colour: String = "blue"
-  }
+class Graph extends GraphBase {
+  class Node extends NodeBase
+  class Edge(val from: Node, val to: Node) extends EdgeBase
+  def newNode() = new Node
+  def newEdge(from: Node, to: Node) = new Edge(from, to)
+}
 
-  override def newNode() = new Node
+class ColouredGraph extends GraphBase {
+  class Node(val colour: String) extends NodeBase
+  class Edge(val from: Node, val to: Node) extends EdgeBase
+  def newNode() = new Node("blue")
+  def newEdge(from: Node, to: Node) = new Edge(from, to)
 }
 
 object Library {
-  def distance(n1: Graph#Node)(n2: n1.out.Node): Int = {
+  def distance(n1: GraphBase#Node)(n2: n1.out.Node): Int = {
     val e: n1.out.Edge = n1.connect(n2)
     0 // ...
   }
-  def copyEdge(e: Graph#Edge): e.out.Edge =
-    new e.out.Edge(e.from, e.to)
+  def copyEdge(e: GraphBase#Edge): e.out.Edge =
+    e.out.newEdge(e.from, e.to)
 
-  def distance2(g: Graph)(n1: g.Node, n2: g.Node): Int = {
+  def distance2(g: GraphBase)(n1: g.Node, n2: g.Node): Int = {
     val e: g.Edge = n1.connect(n2)
     0 // ...
   }
[…]

### 2012-08-31 doc/candidacy/tribe2.scala

@@ -1,67 +1,0 @@
-trait GraphBase { graph =>
-  type Node <: NodeBase
-  type Edge <: EdgeBase
-  def newNode(): Node
-  def newEdge(from: Node, to: Node): Edge
-
-  trait WithOut {
-    val out: graph.type = graph
-  }
-  trait NodeBase extends WithOut { this: Node =>
-    def connect(other: out.Node): out.Edge =
-      newEdge(this, other)
-  }
-  trait EdgeBase extends WithOut { this: Edge =>
-    val from: out.Node
-    val to: out.Node
-  }
-}
-
-class Graph extends GraphBase {
-  class Node extends NodeBase
-  class Edge(val from: Node, val to: Node) extends EdgeBase
-  def newNode() = new Node
-  def newEdge(from: Node, to: Node) = new Edge(from, to)
-}
-
-class ColouredGraph extends GraphBase {
-  class Node(val colour: String) extends NodeBase
-  class Edge(val from: Node, val to: Node) extends EdgeBase
-  def newNode() = new Node("blue")
-  def newEdge(from: Node, to: Node) = new Edge(from, to)
-}
-
-object Library {
-  def distance(n1: GraphBase#Node)(n2: n1.out.Node): Int = {
-    val e: n1.out.Edge = n1.connect(n2)
-    0 // ...
-  }
-  def copyEdge(e: GraphBase#Edge): e.out.Edge =
-    e.out.newEdge(e.from, e.to)
-
-  def distance2(g: GraphBase)(n1: g.Node, n2: g.Node): Int = {
-    val e: g.Edge = n1.connect(n2)
-    0 // ...
-  }
-  def copyEdge2(g: GraphBase)(e: g.Edge): g.Edge =
-    g.newEdge(e.from, e.to)
-}
-
-object Test extends App {
-  val cg1, cg2 = new ColouredGraph()
-  val cn1 = new cg1.Node("blue")
-  val cn2 = new cg2.Node("green")
-  val cn3 = new cg1.Node("red")
-  val e = cn1.connect(cn3)
-  // this is a type mismatch
-  // cn2.connect(cn3)
-
-  val d1 = Library.distance(cn1)(cn3)
-  val d2 = Library.distance2(cg1)(cn1, cn3)
-  val e1 = Library.copyEdge(e)
-  val e2 = Library.copyEdge2(cg1)(e)
-  // these are all type mismatches
[…]

### 2012-08-31 doc/candidacy/report.tex
context
a bit more context

--- 2012-08-30 doc/candidacy/report.tex
@@
 Now, we can emulate this Tribe example in Scala at the expense of some
-synctactic overhead, since further binding and ``out'' selection are
-not native.
-\lstinputlisting{tribe.scala}
+syntactic overhead, since further binding and ``out'' selection are
+not native. We use virtual types to ensure that the inner classes are
+indeed ``virtual'', i.e. dynamically instead of statically
+dispatched. In order to conform to the ``virtual class'' pattern, the
+leaf classes \code{Graph} and \code{ColourGraph} are no longer in a
+subclassing relationship with one another, as this would only bring
+confusion when ascribing a coloured graph to a graph. Emulating
+``out'' faithfully involves a bit of hackery in making Scala recognize
+that, for example, \code{e.from.type} is a subtype of
+\code{e.out.Node}. As noted in the paper, using ``out'' in library
+code obscures the symmetry between the parameters (for example, in the
+\code{distance} method below) but allows the family object to be
+recovered without being explicitly passed around (contrast with the
+method \code{distance2}).  \lstinputlisting{tribe.scala}
 

### 2012-08-31 doc/candidacy/report.tex
intro blurb

@@
 
-We present three papers~\cite{nuObj,tribe,homl}, each of which models a
-calculus with path-dependent types.
+We present three papers~\cite{homl,tribe,nuObj}, each of which models
+a calculus with path-dependent types.
+\begin{itemize}
+\item The first paper, {\it A Type-Theoretic Approach to Higher-Order
+  Modules with Sharing} by Robert Harper and Mark
+  Lillibridge~\cite{homl}, presents a calculus for modules in the ML
+  tradition. Path-dependent types arise because modules, which may
+  contain type components, are first-class values in their calculus.
+\item The second paper, {\it Tribe: a simple virtual class calculus}
+  by Dave Clarke, Sophia Drossopoulou, James Noble and Tobias
+  Wrigstad~\cite{tribe}, presents a core calculus for virtual
+  classes. Path-dependent types arise because, in Tribe, paths are
+  types, and, almost vice versa: each type is a generalized form of
+  path.
+\item The third paper, {\it A Nomial Theory of Objects with Dependent
+  Types} by Martin Odersky, Vincent Cremet, Christine R{\"o}ckl and
+  Matthias Zenger~\cite{nuObj}, presents a calculus which unifies
+  concepts from advanced object-oriented systems and module systems
+  and, with~\cite{FS}, serves as a type-theoretic foundation for many
+  features of Scala's type system. Path-dependent types arise because
+  any path $p$ has a singleton type $p.\type$.
+\end{itemize}
 
@@
 \section{Research Proposal} %: How I Plan to Revolutionize the World
+
 % TODO

### 2012-08-31 doc/candidacy/report.tex
def contractiveness

@@
 from the template. Its scope is the bindings $\tup d$. Since the self
-is explicitly typed, the bindings in class templates can contractively
-refer to each other (and even to themselves) via $x$, while recall
-that, in translucent sums, only later fields can depend on earlier
-ones.
+is explicitly typed, the bindings in class templates can
+``contractively'' refer to each other (and even to themselves) via
+$x$, while recall that, in translucent sums, only later fields can
+refer to earlier ones but the references don't have any
+``contractive'' restrictions. Roughly, ``contractiveness'' in
+${\nu}Obj$ guarantees that a field cannot be referenced before it is
+initalized.
 

### 2012-08-31 doc/candidacy/report.tex
reorg with examples at the end of each section

@@
 
-% TODO: example of translucent module and its utility. woudl be nice to re-use this example in nuObj to show how it can be encoded
-
 In terms of theoretical results, Lillibridge shows in his thesis that
@@
 checking, is undecidable.
+
+\subsubsection*{SML Example and Scala Encoding}
+TODO
 
@@
 
+In terms of theoretical results, there's a sketch of soundness in the
+paper, but it's unclear whether it is correct. In particular, the
+substitution lemma does not seem valid when null is the value to be
+substituted and the expression has a type which contains the variable
+to be substituted. In the follow-up paper~\cite{tribalo}, the
+soundness proof sketch seems more plausible but then is tied to the
+ownership system formalism. Decidability of subtyping, and hence
+typechecking, has not yet been established.
+
+\subsubsection*{Tribe Example and Scala Encoding}
 Let's now re-use an example from the paper, and discuss how we would
@@
 recovered without being explicitly passed around (contrast with the
-method \code{distance2}).  \lstinputlisting{tribe.scala}
-
-In terms of theoretical results, there's a sketch of soundness in the
-paper, but it's unclear whether it is correct. In particular, the
-substitution lemma does not seem valid when null is the value to be
-substituted and the expression has a type which contains the variable
-to be substituted. In the follow-up paper~\cite{tribalo}, the
-soundness proof sketch seems more plausible but then is tied to the
-ownership system formalism. Decidability of subtyping, and hence
-typechecking, has not yet been established.
+method \code{distance2}).
+\lstinputlisting{tribe.scala}
 
@@
 
-Let's now illustrate with a few examples the correspondences outlined
-earlier.
-
-First, we show how to encode monomorphic functions into the
+In terms of theoretical results, the paper confirms the type soundness
+of the calculus and the undecidability of its type checking.
+
+\subsubsection*{${\nu}Obj$ Example}
+
+We show how to encode monomorphic functions into the
 calculus. The paper goes further and generalizes the encoding to
-system $F_{<:}$. A $\lambda$-abstraction $\lambda ( x : T )\; t$ is
-represented as a class with an abstract member \code!arg! for the
-function argument:
+system $F_{<:}$.
+
+A $\lambda$-abstraction $\lambda ( x : T )\; t$ is represented as a
+class with an abstract member \code!arg! for the function argument:
 \begin{lstlisting}
@@
 
-% TODO: example of module encoding... perhaps in Scala rather than nuObj?
-
-In terms of theoretical results, the paper confirms the type soundness
-of the calculus and the undecidability of its type checking.
+This example illustrates the correspondence between classes, methods
+and functors claimed by ${\nu}Obj$.
 

### 2012-08-31 doc/candidacy/report.tex
ML example

@@
 
-\subsection{A Type-Theoretic Approach to Higher-Order Modules with Sharing}
+\subsection{A Type-Theoretic Approach to Higher-Order Modules with Sharing}\label{sec:homl}
 
@@
 
-\subsubsection*{SML Example and Scala Encoding}
-TODO
-
-\subsection{Tribe: a simple virtual class calculus}
+\subsubsection*{Pseudo-SML Example and Scala Encoding}
+Consider this example, in a ML-like language implementing the calculus
+of the paper. Since modules are first-class, we can \code{flip} a coin
+and return a module depending on the outcome:
+\begin{lstlisting}
+structure XInt = struct
+  type T = int
+  val v = 3
+  val f = negate
+end
+structure XBool = struct
+  type T = bool
+  val v = true
+  val f = not
+end
+structure X =
+  if flip() then XInt else XBool
+\end{lstlisting}
+But now, what should be the static type or signature of structure
+\code{X}? In ML, in order to make the \code{if} type check, the two
+branches must have equal types. In the calculus, the subsumption rule
+is used to coerce both types to their least upper bound, represented
+by the signature:
+\begin{lstlisting}
+signature XType = sig
+  type T
+  val v = T
+  val f = T -> T
+end
+\end{lstlisting}
+
+It is straightforward to encode this example in Scala, with the
+correspondence Object = Module, and Object Type = Signature. Indeed,
+the ${\nu}Obj$ calculus, which we study in section~\ref{sec:nuObj}
+embraces this correspondence, and futher claims: Class = Method =
+Functor, as we will see.
+
+\subsection{Tribe: a simple virtual class calculus}\label{sec:tribe}
 
@@
 
-\subsection{A Nominal Theory of Objects with Dependent Types}
+\subsection{A Nominal Theory of Objects with Dependent Types}\label{sec:nuObj}
 

### 2012-08-31 doc/candidacy/ml.scala

@@ -1,0 +1,24 @@
+object Test extends App {
+  object coin {
+    val r = new util.Random()
+    def flip() = r.nextBoolean()
+  }
+  type XType = {
+    type T
+    val v: T
+    val f: T => T
+  }
+  val xInt = new {
+    type T = Int
+    val v = 3
+    val f = - (_:T)
+  }
+  val xBool = new {
+    type T = Boolean
+    val v = true
+    val f = ! (_:T)
+  }
+  val x: XType = if (coin.flip()) xInt else xBool
+
+  println(x.f(x.v))
+}

### 2012-08-31 doc/candidacy/report.tex
show the scala ex... though not that insightful to.
little rev in ml section

@@
 from an if statement. In fact, allowing this might require obscuring
-the visibility of the resulting structure to ensure soundness. Another
-limitation of the SML module system is that ``type sharing'' is
-limited to equality between type names, not general type
-expressions. In effect, this means that the transparency / opaqueness
-of a signature cannot be fully fine-tuned. Their proposal introduces
-translucent signatures, which allows any type to be partially or fully
-determined by a type expression.
+the visibility of the resulting structure to ensure soundness, which
+we illustrate with an example below. Another limitation of the SML
+module system is that ``type sharing'' is limited to equality between
+type names, not general type expressions. In effect, this means that
+the transparency / opaqueness of a signature cannot be fully
+fine-tuned. Their proposal introduces translucent signatures, which
+allows any type to be partially or fully determined by a type
+expression.
 
@@
 
-\subsubsection*{Pseudo-SML Example and Scala Encoding}
+\subsubsection*{ML-like Example and Scala Encoding}
 Consider this example, in a ML-like language implementing the calculus
@@
 embraces this correspondence, and futher claims: Class = Method =
-Functor, as we will see.
+Functor, as we will see. Note that type members, and hence
+path-dependent types, are crucial for this correspondence.
+\lstinputlisting{ml.scala}
 

### 2012-08-31 doc/candidacy/report.tex
minor rev

@@
 calculus, there's a form of dependent type, since translucent sums are
-expressions that may contain types. Compared to path-dependent types
+terms that may contain types. Compared to path-dependent types
 as found in Scala, these are more limited because recursive references
@@
 A class template $\cls{x: S \sep \tup d}$ is closely related to a
-translucent sum, except that it needs to be instantiated into an
+translucent sum of section~\ref{sec:homl}, except that it needs to be instantiated into an
 object while a translucent sum is a module. Like in a translucent sum,

### 2012-08-31 doc/candidacy/report.tex
minor
abstract

@@
 
-% TODO
+Our aim is to propose a new type-theoretic foundation of Scala based
+on a core calculus for path-dependent types. Towards this end, we
+review three papers presenting calculi in which path-dependent types
+are essential: a calculus for first-class translucent ML-like modules,
+the Tribe calculus for virtual classes, and the ${\nu}Obj$ calculus
+for reasoning about classes and objects with type members.
 
@@
 In terms of theoretical results, the paper confirms the type soundness
-of the calculus and the undecidability of its type checking.
+of the calculus and the undecidability of its type checking by
+reduction to $F_{<:}$.
 

### 2012-08-31 doc/candidacy/report.tex
first draft of my work sec

@@
 
-% TODO
+In our own research, we have been exploring a new core calculus for
+path-dependent types, the Dependent Object Types (DOT) calculus, which
+could serve as a new type-theoretic foundation for Scala and languages
+like it. The three calculi presented demonstrate that path-dependent
+types arise in diverse settings, and thus, it is valuable to identify
+a core calculus which captures their essence. In DOT, we are not
+concerned with modeling orthogonal concepts such as inheritance and
+mixin composition. In contrast to the three previous calculi,
+type-checking in DOT is decidable, since all the judgments are
+syntax-directed (in the current design).
+
+DOT aims to bring more uniformity and simplicity to Scala. Scala, and
+${\nu}Obj$ which models it, conflate the concepts of compound types
+(which inherit the members of several parent types) and mixin
+composition (which build classes from other classes and traits). At
+first glance, this offers an economy of concepts. However, it is
+problematic because mixin composition and intersection types have
+quite different properties. In particular, we have seen in
+section~\ref{sec:nuObj} that mixin composition is not commutative,
+unlike classical intersection types.
+
+In DOT, we replace Scala's compound types by classical intersection
+types, which are commutative. We also complement this by classical
+union types. Intersections and unions form a lattice wrt
+subtyping. This addresses another problematic feature of Scala: In
+Scala's current type system, least upper bounds and greatest lower
+bounds do not always exist. Here is an example: Given two traits
+\begin{lstlisting}
+  trait A { type T <: A }
+  trait B { type T <: B }
+\end{lstlisting}
+The greatest lower bound of \code@A@ and \code@B@ is approximated by the
+infinite sequence
+\begin{lstlisting}
+  A with B { type T <: A with B { type T <: A with B {
+    type T < ...
+  }}}
+\end{lstlisting}
+The limit of this sequence does not exist as a type in Scala. This is
+problematic because glbs and lubs play a central role in Scala's type
+inference. The absence of universal glbs and lubs makes type inference
+more brittle and more unpredictable.
+
+Our immediate goal is to prove type soundness of the DOT calculus. We
+first tried to prove the calculus type-safe using the standard
+theorems of preservation and progress~\cite{soundness,
+  tapl}. Unfortunately, for the DOT calculus and any variants that we
+devise, preservation doesn't hold. Indeed, we found many
+counterexamples to preservation due to narrowing, i.e.\ a type becoming more
+precise after substitution. However, the standard theorems of
+preservation and progress are just one way to prove type soundness. We
+are exploring a promising avenue to prove type safey using the
+powerful method of step-indexed logical
+relations~\cite{ahmed04,ahmed06,step_indexed_obj}. In the process of
+proving the calculus sound, we are also refining it.
+
+In future work, we plan to investigate translations of Scala features
+into the core DOT calculus, and implement an experimental compiler
+front-end which uses the DOT calculus for type-checking.
+
+%\clearpage
+
 %Describe your own work and include a  summary and references. Write how you

### 2012-08-31 doc/candidacy/report.tex
typo

@@
   path.
-\item The third paper, {\it A Nomial Theory of Objects with Dependent
+\item The third paper, {\it A Nominal Theory of Objects with Dependent
   Types} by Martin Odersky, Vincent Cremet, Christine R{\"o}ckl and

### 2012-09-03 doc/candidacy/report.tex
feedback of eugene

@@
   features of Scala's type system. Path-dependent types arise because
-  any path $p$ has a singleton type $p.\type$.
+  any path $p$ has a singleton type $p.\type$, from which type members
+  can be selected.
 \end{itemize}
@@
 first-class modules, with complete control over the propagation of
-compile-time information.
+information between compile-time program units.
 
@@
 second-class. For example, it isn't possible to return a structure
-from an if statement. In fact, allowing this might require obscuring
+from an if expression. In fact, allowing this might require obscuring
 the visibility of the resulting structure to ensure soundness, which
@@
 leaf classes \code{Graph} and \code{ColourGraph} are no longer in a
-subclassing relationship with one another, as this would only bring
-confusion when ascribing a coloured graph to a graph. Emulating
+subclassing / further-binding with one another, as this would only bring
+confusion when ascribing a coloured graph to a graph, since the constructors are statically dispatched. Emulating
 ``out'' faithfully involves a bit of hackery in making Scala recognize
 that, for example, \code{e.from.type} is a subtype of
-\code{e.out.Node}. As noted in the paper, using ``out'' in library
+\code{e.out.Node}. As noted in the Tribe paper, using ``out'' in library
 code obscures the symmetry between the parameters (for example, in the
@@
 
-The terms in ${\nu}Obj$ denotes objects or classes. They consist of
+The terms in ${\nu}Obj$ denote objects or classes. They consist of
 variables $x$ denoting objects, selections $t.l$, object creations
@@
 
-The types in ${\nu}Obj$ consists of singleton types $p.\type$, type
+The types in ${\nu}Obj$ consist of singleton types $p.\type$, type
 selections $T\Tsel L$, record types $\obj{x \sep \tup D}$, compound
@@
 preservation and progress are just one way to prove type soundness. We
-are exploring a promising avenue to prove type safey using the
+are exploring a promising avenue to prove type safety using the
 powerful method of step-indexed logical

### 2012-09-03 doc/candidacy/report.tex
rev

@@
 second-class. For example, it isn't possible to return a structure
-from an if expression. In fact, allowing this might require obscuring
+from an \code{if} expression. In fact, allowing this might require obscuring
 the visibility of the resulting structure to ensure soundness, which
 we illustrate with an example below. Another limitation of the SML
-module system is that ``type sharing'' is limited to equality between
+module system is that ``type sharing'' is restricted to equality between
 type names, not general type expressions. In effect, this means that
@@
 But now, what should be the static type or signature of structure
-\code{X}? In ML, in order to make the \code{if} type check, the two
+\code{X}? In ML, in order to make the \code{if} expression typecheck, the two
 branches must have equal types. In the calculus, the subsumption rule
@@
 simultaneously on both classes and objects. Additionally, paths can
-use an ``out'' field to move from an object to the object which
-surrounds it. This ``out'' feature enables ubiquitous access to an
+use an \code{out} field to move from an object to the object which
+surrounds it. This \code{out} feature enables ubiquitous access to an
 object's family without the need to drag around family arguments.
@@
 Now, we can emulate this Tribe example in Scala at the expense of some
-syntactic overhead, since further binding and ``out'' selection are
+syntactic overhead, since further binding and \code{out} selection are
 not native. We use virtual types to ensure that the inner classes are
@@
 confusion when ascribing a coloured graph to a graph, since the constructors are statically dispatched. Emulating
-``out'' faithfully involves a bit of hackery in making Scala recognize
+\code{out} faithfully involves a bit of hackery in making Scala recognize
 that, for example, \code{e.from.type} is a subtype of
-\code{e.out.Node}. As noted in the Tribe paper, using ``out'' in library
+\code{e.out.Node}. As noted in the Tribe paper, using \code{out} in library
 code obscures the symmetry between the parameters (for example, in the
@@
   tapl}. Unfortunately, for the DOT calculus and any variants that we
-devise, preservation doesn't hold. Indeed, we found many
+devised, preservation doesn't hold. Indeed, we found many
 counterexamples to preservation due to narrowing, i.e.\ a type becoming more

### 2012-09-03 doc/candidacy/slides.tex
basic template for slides

\documentclass{beamer}
\usepackage{minted}
\usemintedstyle{eclipse}
\begin{document}

\begin{frame}[fragile]{Hello}
\begin{minted}{scala}
object HelloWorld extends App {
  println("Hello, World")
}
\end{minted}
\end{frame}

\end{document}

### 2012-09-03 doc/candidacy/report.tex
a bit more intro

@@
 variables. Path-dependent types arise when terms, usually objects, can
-have types as members.
+have types as members. Abstract type members, together with explicit
+selftypes and modular mixin composition have been identified as key
+abstractions for the construction of reusable components~\cite{SCA}.
+
+\subsubsection*{Scala Example} In the code below,
+\code{type T} is an abstract type member of \code{AbsCell}. In a
+library method taking an abstract cell \code{c}, the expression
+\code{c.set(c.init)} is well-typed because \code{c.init} has type
+\code{c.T} and \code{c.set} has the type of a function that expects a
+parameter of type \code{c.T}. Note that \code{c.T} is a path-dependent
+type. \lstinputlisting{intro.scala}
 
@@
 
-\section{Survey of the Selected Papers}%: The Giants upon whose shoulders I will stand
+In section~\ref{survey}, we survey selected previous work related to
+path-dependent types. Whenever possible, we contrast the formalisms
+with concrete examples in Scala. In section~\ref{proposal}, we discuss
+in more details our proposal for a new type-theoretic foundation of
+Scala based on path-dependent types.
+
+\section{Survey of the Selected Papers}\label{survey}%: The Giants upon whose shoulders I will stand
 
@@
 
-\section{Research Proposal} %: How I Plan to Revolutionize the World
+\section{Research Proposal}\label{proposal}%: How I Plan to Revolutionize the World
 

### 2012-09-03 doc/candidacy/slides.tex
a bit more intro

@@
 \documentclass{beamer}
+
 \usepackage{minted}
 \usemintedstyle{eclipse}
+
 \begin{document}
+
+\begin{frame}
+\frametitle{DOT: Dependent Object Types}
+
+The DOT calculus proposes a new \emph{type-theoretic foundation} for Scala
+and languages like it. It models
+\begin{itemize}
+\item \emph{path-dependent types}
+\item abstract type members
+\item mixture of nominal and structural typing via refinement types
+\end{itemize}
+
+It does not model
+\begin{itemize}
+\item inheritance and mixin composition
+\item what's currently in Scala
+\end{itemize}
+
+\end{frame}
+
+\begin{frame}[fragile]{Path-dependent types}
+
+\begin{description}[path-dependent type]
+\item[path-dependent type] limited form of \emph{dependent type}, in which a type depends on a \emph{path}
+\item[dependent type] a type which depends on a term
+\item[path] a chain of immutable fields or variables
+\end{description}
+
+\inputminted[frame=lines,fontsize=\footnotesize]{scala}{intro.scala}
+
+\end{frame}
 

### 2012-09-03 doc/candidacy/intro.scala

@@ -1,0 +1,13 @@
+abstract class AbsCell {
+  type T
+  val init: T
+  private var value: T = init
+  def get = value
+  def set(x: T) { value = x }
+}
+object Library {
+  def reset(c: AbsCell) { c.set(c.init) }
+  def update(c: AbsCell)(oldval: c.T, newval: c.T) =
+    if (oldval == c.get) { c.set(newval); true }
+    else false
+}

### 2012-09-03 doc/candidacy/intro_test.scala

@@ -1,0 +1,6 @@
+object Test extends App {
+  val c = new AbsCell { type T = Int; val init = 0 }
+  println(Library.update(c)(0, 1))
+  println(Library.update(c)(0, 1))
+  println(Library.update(c)(1, 0))
+}

### 2012-09-03 doc/candidacy/slides.tex
selected papers

@@
 \inputminted[frame=lines,fontsize=\footnotesize]{scala}{intro.scala}
+\end{frame}
 
+\begin{frame}{Selected Papers}
+\begin{enumerate}
+\item {\it A Type-Theoretic Approach to Higher-Order
+  Modules with Sharing}\\{\footnotesize by Robert Harper and Mark
+  Lillibridge}\begin{itemize}
+\item calculus for modules in the ML tradition
+\item modules are first-class values
+\end{itemize}
+\item {\it Tribe: a simple virtual class calculus}\\
+  {\footnotesize by Dave Clarke, Sophia Drossopoulou, James Noble and Tobias
+  Wrigstad}\begin{itemize}
+\item calculus for virtual classes
+\end{itemize}
+\item {\it A Nominal Theory of Objects with Dependent
+  Types}\\{\footnotesize by Martin Odersky, Vincent Cremet, Christine R{\"o}ckl and
+  Matthias Zenger}\begin{itemize}
+\item calculus unifying advanced object-oriented and module systems
+\item foundation for Scala
+\end{itemize}
+\end{enumerate}
 \end{frame}

### 2012-09-03 doc/candidacy/report.tex
ML rev

@@
 type depends on a path: a chain of immutable fields or
-variables. Path-dependent types arise when terms, usually objects, can
+stable values. Path-dependent types arise when terms, usually objects, can
 have types as members. Abstract type members, together with explicit
@@
 calculus, there's a form of dependent type, since translucent sums are
-terms that may contain types. Compared to path-dependent types
-as found in Scala, these are more limited because recursive references
-and bounded quantifications are not allowed.
+terms that may contain types. More concretely, their so-called VALUE
+rules enable a translucent sum to be given a more precise type by
+referring to the names of its type members with a path selection --
+this ability based on path-dependent types is not only essential for
+breaking the dependencies between (sub)fields but also critical for
+the propagation of typing information. Compared to Scala, these
+path-dependent types are more limited because recursive references and
+bounded quantifications are not allowed.
 

### 2012-09-03 doc/candidacy/slides.tex
section titles
sym
ML rev

--- 2012-09-03 doc/candidacy/slides.tex
@@
 
+\title{Dependent Object Types}
+\subtitle{EDIC Candidacy Exam}
+\author{Nada Amin}
+\institute{LAMP, I\&C, EPFL}
+\date{September 12, 2012}
+
 \begin{document}
+
+\frame{\titlepage}
+
+\section{Introduction}
 
@@
 \item[dependent type] a type which depends on a term
-\item[path] a chain of immutable fields or variables
+\item[path] a chain of immutable fields or stable values
 \end{description}
@@
 \inputminted[frame=lines,fontsize=\footnotesize]{scala}{intro.scala}
+\end{frame}
 
+\begin{frame}{Selected Papers}
+\begin{enumerate}
+\item {\it A Type-Theoretic Approach to Higher-Order
+  Modules with Sharing}\\{\footnotesize\textcolor{gray}{by Robert Harper and Mark
+  Lillibridge}}\begin{itemize}
+\item calculus for modules in the ML tradition
+\item modules are first-class values
+\end{itemize}
+\item {\it Tribe: a simple virtual class calculus}\\
+  {\footnotesize\textcolor{gray}{by Dave Clarke, Sophia Drossopoulou, James Noble and Tobias Wrigstad}}\begin{itemize}
+\item calculus for virtual classes
+\item paths are types, and types are ``generalized'' paths
+\end{itemize}
+\item {\it A Nominal Theory of Objects with Dependent
+  Types}\\{\footnotesize\textcolor{gray}{by Martin Odersky, Vincent Cremet, Christine R{\"o}ckl and
+  Matthias Zenger}}\begin{itemize}
+\item calculus unifying advanced object-oriented and module systems
+\item foundation for Scala
+\end{itemize}
+\end{enumerate}
 \end{frame}
+
+\section{Survey of Selected Papers}
+
+\subsection{``Translucent'' ML}
+
+\subsection{Tribe Calculus}
+
+\subsection{${\nu}Obj$}
+
+\section{Proposal}
 

### 2012-09-03 doc/candidacy/report.tex
syntax-directed does not imply decidable
tribe rev

@@
 have types as members. Abstract type members, together with explicit
-selftypes and modular mixin composition have been identified as key
+selftypes and modular mixin composition, have been identified as key
 abstractions for the construction of reusable components~\cite{SCA}.
@@
 extended. Path-dependent types arise naturally to distinguish objects
-coming from different families. In Tribe, path types can depend
-simultaneously on both classes and objects. Additionally, paths can
-use an \code{out} field to move from an object to the object which
-surrounds it. This \code{out} feature enables ubiquitous access to an
-object's family without the need to drag around family arguments.
+coming from different families. Yet, thanks to {\it family
+  polymorphism}, code written for one family also works for extension
+of that family. In Tribe, path types can depend simultaneously on both
+classes and objects. Additionally, paths can use an \code{out} field
+to move from an object to the object which surrounds it. This
+\code{out} feature enables ubiquitous access to an object's family
+without the need to drag around family arguments.
 
@@
 concerned with modeling orthogonal concepts such as inheritance and
-mixin composition. In contrast to the three previous calculi,
-type-checking in DOT is decidable, since all the judgments are
-syntax-directed (in the current design).
+mixin composition.
 

### 2012-09-04 doc/candidacy/report.tex
typos

@@
 the ${\nu}Obj$ calculus, which we study in section~\ref{sec:nuObj}
-embraces this correspondence, and futher claims: Class = Method =
+embraces this correspondence, and further claims: Class = Method =
 Functor, as we will see. Note that type members, and hence
@@
 nested inside other classes. When a class is inherited, its nested
-inner classes are inherited along with its methods and fields. {\it
+inner classes are inherited along with their methods and fields. {\it
   Further binding} enables the nested classes to be further
@@
 dispatched. In order to conform to the ``virtual class'' pattern, the
-leaf classes \code{Graph} and \code{ColourGraph} are no longer in a
+leaf classes \code{Graph} and \code{ColouredGraph} are no longer in a
 subclassing / further-binding with one another, as this would only bring
@@
 ${\nu}Obj$ guarantees that a field cannot be referenced before it is
-initalized.
+initialized.
 
@@
 the right-hand side of \code!fun!: this wouldn't be ``contractive'',
-because \code!arg! might not be initialised before \code!fun!. Hence,
+because \code!arg! might not be initialized before \code!fun!. Hence,
 we pack $t'$ in another class to ensure contractiveness. Now, an

### 2012-09-04 doc/candidacy/report.tex
a bit more about tribe paths

@@
 coming from different families. Yet, thanks to {\it family
-  polymorphism}, code written for one family also works for extension
+  polymorphism}, code written for one family also works for extensions
 of that family. In Tribe, path types can depend simultaneously on both
@@
 %consists of chains of immutable fields or variables.
+
+In Tribe, types are a form of generalized paths: a final variable,
+\code{this} or a class name then followed by a possibly empty sequence
+of field or class selections. Given the code below, we can read the
+type \code{kitt.Passenger.name} as the name of one of Kitt's
+passengers (including the driver). Subtyping is natural:
+\code{kitt.driver <: Car.driver <: Car.Passenger <: Car.Traveller <: Vehicle.Traveller}. Crucially, \code{kitt.driver} is not a subtype
+of \code{karr.driver} and similarly for \code{kitt.Passenger} and
+\code{karr.Passenger}.
+\begin{lstlisting}
+class Vehicle { class Traveller { ... } }
+class Car extends Vehicle {
+  class Passenger extends Traveller {
+    class String { ... }
+    final String name;
+  }
+  final Passenger driver;
+}
+final Car kitt;
+final Car karr;
+\end{lstlisting}
 
@@
 paper, but it's unclear whether it is correct. In particular, the
-substitution lemma does not seem valid when null is the value to be
+substitution lemma does not seem valid when \code{null} is the value to be
 substituted and the expression has a type which contains the variable

### 2012-09-05 doc/candidacy/slides.tex
ML slides outline
tribe rev
notes to self
ml slides

@@
 
+\begin{frame}{SML modules}
+\begin{description}[structure]
+\item[structure] a module; a collection of types, values and
+  (sub)structures packaged as a program unit.
+\item[signature] an interface; the type of a structure; description of
+  the types, values and (sub)structures of a structure given by their
+  kinds, types and interfaces.
+\item[functor] a function mapping structures to structures
+\item[type-sharing] enables stating that types specified in separate
+  modules are actually equal
+\end{description}
+% example?
+\end{frame}
+
+\begin{frame}{Limitations of SML modules}
+\begin{itemize}
+\item modules are not first-class; e.g. not possible to return a
+  structure from an {\tt if} expression.
+\item type sharing is restricted to equality between type names, not
+  general type expressions; in effect, transparency / opaqueness of a
+  signature cannot be fully fined-tuned
+\end{itemize}
+\end{frame}
+
+\begin{frame}{Translucent Sums}
+\begin{itemize}
+\item values representing modules
+\item translucent sum: sequence of bindings; translucent sum type:
+  sequence of declarations
+\item differs from traditional records, because later fields can
+  depend on earlier ones
+\item allows any type to be partially or fully deermined by a type
+\end{itemize}
+\end{frame}
+
+\begin{frame}[fragile]{Example}
+\begin{minted}{sml}
+structure XInt = struct
+  type T = int
+  val v = 3
+  val f = negate
+end
+structure XBool = struct
+  type T = bool
+  val v = true
+  val f = not
+end
+structure X =
+  if flip() then XInt else XBool
+signature XType = sig
+  type T
+  val v = T
+  val f = T -> T
+end
+\end{minted}
+\end{frame}
+
+% Maintaing Soundness?
+
+\begin{frame}{Path-Dependent Types}
+\begin{itemize}
+\item Scala's abstract types have close resemblances to abstract types
+  of signatures in these ML-like module systems. In particular, in
+  their calculus, there's a form of dependent type, since translucent
+  sums are terms that may contain types.
+\item More concretely, their so-called VALUE rules enable a
+  translucent sum to be given a more precise type by referring to the
+  names of its type members with a path selection -- this ability
+  based on path-dependent types is not only essential for breaking the
+  dependencies between (sub)fields but also critical for the
+  propagation of typing information.
+\item Compared to Scala, these path-dependent types are more limited
+  because recursive references and bounded quantifications are not
+  allowed.
+\end{itemize}
+\end{frame}
+
 \subsection{Tribe Calculus}
+% Virtual Classes
+% Example
+% Tribe Types (generalized paths & subtyping relation)
+% Follow-up
+  % Ownership System?
+  % Soundness?
+  % Cross-Family Inheritance
 
@@
 
+\section{Backup Slides}
+
+% System F omega
+% System F sub
 \end{document}

### 2012-09-05 doc/candidacy/slides.tex
starting tribe section

@@
 
-\begin{frame}{Translucent Sums}
+\begin{frame}{ML: Translucent Sums}
 \begin{itemize}
@@
 
-\begin{frame}[fragile]{Example}
+\begin{frame}[fragile]{ML: Example}
 \begin{minted}{sml}
@@
 
-\begin{frame}{Path-Dependent Types}
+\begin{frame}{ML: Path-Dependent Types}
 \begin{itemize}
@@
 \subsection{Tribe Calculus}
+
+\begin{frame}{Tribe Calculus: Virtual Classes}
+\begin{itemize}
+\item {\it familes} of classes inherited together
+\item classes are lexically nested inside other classes: when a class
+  is inherited, its nested inner classes are inherited along with
+  their methods and fields
+\item two form of inheritance: subclassing and further binding
+\item {\it family polymorphism}: code written for one family
+  also works for extensions of that family
+\end{itemize}
+\end{frame}
+
+\begin{frame}[fragile]{Tribe Types}
+\begin{itemize}
+\item In Tribe, types are a form of generalized paths: a final
+  variable, {\tt this} or a class name then followed by a possibly
+  empty sequence of field, {\tt out} or class selections.
+\item {\tt kitt.Passenger.name}: the name of one of Kitt's passengers
+\item {\tt kitt.driver <: Car.driver <: Car.Passenger <: Car.Traveller <:
+  Vehicle.Traveller}.
+\item {\tt kitt.driver $\not<:$ karr.driver} and {\tt kitt.Passenger $\not<:$ karr.Passenger}.
+\end{itemize}
+\begin{minted}[frame=lines,fontsize=\footnotesize]{java}
+class Vehicle { class Traveller { ... } }
+class Car extends Vehicle {
+  class Passenger extends Traveller {
+    class String { ... }
+    final String name;
+  }
+  final Passenger driver;
+}
+final Car kitt;
+final Car karr;
+\end{minted}
+\end{frame}
+
 % Virtual Classes

### 2012-09-05 doc/candidacy/slides.tex
a bit more tribalism

@@
   also works for extensions of that family
+\item in Tribe:\begin{itemize}
+\item path types can depend simultaneously on both classes and objects
+\item paths can use an {\tt out} field to move from an object to the
+  object which surrounds it; enables ubiquitous access to an object's
+  family without the need to drag around family arguments.
 \end{itemize}
+\end{itemize}
+\end{frame}
+
+\begin{frame}[fragile]{Tribe: Example}
+\begin{minted}{java}
+class Graph {
+  class Node {
+    Edge connect(Node other) {
+      return new Edge(this, other);
+    }
+  }
+  class Edge {
+    Node from, to;
+    Edge(Node f, Node t) { from = f; to = t; }
+  }
+}
+
+class ColouredGraph extends Graph { // subclassing
+  class Node {                      // further binding
+    Colour nodeColour;
+  }
+}
+\end{minted}
+\end{frame}
+
+\begin{frame}[fragile]{Tribe: Example (continued)}
+\begin{minted}{java}
+final ColouredGraph cg1, cg2;
+cg1.Node cn1, cn3;
+cg2.Node cn2;
+cn1.connect(cn3); // Type Correct
+cn2.connect(cn3); // Type Error!!!
+
+class Library {
+  int distance(Graph.Node n1, n1.out.Node n2) { ... }
+
+  e.out.Edge copyEdge(Graph.Edge e) {
+    e.out.Node from = e.from;
+    e.out.Node to = e.to;
+    new e.out.Edge(from, to);
+  }
+}
+\end{minted}
 \end{frame}

### 2012-09-05 doc/candidacy/slides.tex
tribe first pass

@@
 
+\begin{frame}{Tribe: Discussion}
+\begin{itemize}
+\item no virtual classes in Scala; its virtual types can emulate some (but not all) of the benefits
+\item because of {\tt out} field, Tribe has limited cross-family inheritance
+\item soundness? substitution lemma seems wrong when {\tt null} is
+  substituted for a variable in an expression whose type contains
+  that variable
+\item aims to have decidable type-checking, but still not proved so
+\end{itemize}
+\end{frame}
+
 % Virtual Classes

### 2012-09-06 doc/candidacy/slides.tex
rev

@@
 
+\begin{frame}[fragile]{SML modules: Example}
+TODO
+\end{frame}
+
 \begin{frame}{Limitations of SML modules}
@@
 \begin{itemize}
-\item values representing modules
-\item translucent sum: sequence of bindings; translucent sum type:
-  sequence of declarations
-\item differs from traditional records, because later fields can
-  depend on earlier ones
-\item allows any type to be partially or fully deermined by a type
+\item translucent sums are values representing modules
+\item translucent sum: sequence of bindings
+\item translucent sum type: sequence of declarations
+\item unlike traditional records, later fields can depend on earlier ones
+\item any type can be partially or fully determined by a type
+\item dependent typing since translucent sums are terms that may contain type components
 \end{itemize}
@@
 \begin{itemize}
-\item Scala's abstract types have close resemblances to abstract types
-  of signatures in these ML-like module systems. In particular, in
-  their calculus, there's a form of dependent type, since translucent
-  sums are terms that may contain types.
-\item More concretely, their so-called VALUE rules enable a
-  translucent sum to be given a more precise type by referring to the
-  names of its type members with a path selection -- this ability
-  based on path-dependent types is not only essential for breaking the
-  dependencies between (sub)fields but also critical for the
-  propagation of typing information.
-\item Compared to Scala, these path-dependent types are more limited
-  because recursive references and bounded quantifications are not
-  allowed.
+\item Scala's abstract types  $\approx$ ML's abstract types of signatures; limitations:
+\begin{itemize}
+\item recursive references
+\item bounded quantifications
+\end{itemize}
+\item translucent sum can be given a more precise type by referring to the name of its member path with a path selection; essential for
+\begin{itemize}
+\item breaking the dependencies between (sub)fields
+\item propagation of typing information
+\end{itemize}
 \end{itemize}
@@
 
-\begin{frame}[fragile]{Hello}
-\begin{minted}{scala}
-object HelloWorld extends App {
-  println("Hello, World")
-}
-\end{minted}
-\end{frame}
-
 \section{Backup Slides}

### 2012-09-06 doc/candidacy/slides.tex
wip vobj

@@
 \documentclass{beamer}
+
+\input{nuobj}
 
@@
 
+\begin{frame}{${\nu}Obj$}
+A calculus for classes and objects which can have types as members. It can encode:
+\begin{itemize}
+\item Java's inner classes
+\item virtual types
+\item family polymorphism
+\item essential aspects of ML-like module systems, including
+\begin{itemize}
+\item sharing constraints
+\item higher-order functors
+\end{itemize}
+\end{itemize}
+A basis for unifying concepts  in advaced object and module systems. Connections:
+\begin{itemize}
+\item Object = Module
+\item Object type = Signature
+\item Class = Method = Functor
+\end{itemize}
+\end{frame}
+
+\begin{frame}{${\nu}Obj$: Syntax}
+The terms in ${\nu}Obj$ denote objects or classes. They consist of
+variables $x$ denoting objects, selections $t.l$, object creations
+\mbox{$\newin{x \larrowc t}{u}$} defining an object $x$ of class $t$
+whose scope is the term $u$, class templates $\cls{x: S \sep \tup d}$,
+and mixin compositions $t \withsub{S} u$ denoting a class of self
+type $S$ combined from two classes $t$ and $u$. A value is a variable
+or a class template. A path is a variable followed by a possibly empty
+sequence of selections.
+
+The types in ${\nu}Obj$ consist of singleton types $p.\type$, type
+selections $T\Tsel L$, record types $\obj{x \sep \tup D}$, compound
+types $T \with U$, and class types $\cls{x: S \sep \tup D}$.
+\begin{itemize}
+\item TODO
+\end{itemize}
+\end{frame}
+
 \section{Proposal}

### 2012-09-06 doc/candidacy/slides.tex
wip vobj

--- 2012-09-06 doc/candidacy/slides.tex
@@
 
+\begin{frame}{${\nu}Obj$}
+A calculus for classes and objects which can have types as members. It can encode:
+\begin{itemize}
+\item Java's inner classes
+\item virtual types
+\item family polymorphism
+\item essential aspects of ML-like module systems, including
+\begin{itemize}
+\item sharing constraints
+\item higher-order functors
+\end{itemize}
+\end{itemize}
+A basis for unifying concepts  in advaced object and module systems. Connections:
+\begin{itemize}
+\item Object = Module
+\item Object type = Signature
+\item Class = Method = Functor
+\end{itemize}
+\end{frame}
+
+\begin{frame}{${\nu}Obj$: Syntax}
+\begin{itemize}
+\item terms in ${\nu}Obj$ denote objects or classes; consist of:
+\begin{itemize}
+\item variables $x$ denoting objects
+\item selections $t.l$
+\item object creations
+\item \mbox{$\newin{x \larrowc t}{u}$} defining an object $x$ of class $t$
+whose scope is the term $u$
+\item class templates $\cls{x: S \sep \tup d}$,
+\item mixin compositions $t \withsub{S} u$ denoting a class of self
+type $S$ combined from two classes $t$ and $u$
+\end{itemize}
+\item a value is a variable or class template
+\item a path is a variable followed by a possibly empty sequence of selections
+\item types in ${\nu}Obj$ consist of:
+\begin{itemize}
+\item singleton types $p.\type$
+\item type selections $T\Tsel L$
+\item record types $\obj{x \sep \tup D}$
+\item compound types $T \with U$
+\item class types $\cls{x: S \sep \tup D}$.
+\end{itemize}
+\end{itemize}
+\end{frame}
+
 \section{Proposal}

### 2012-09-06 doc/candidacy/slides.tex
vobj syntax
vobj: details

@@
 \item terms in ${\nu}Obj$ denote objects or classes; consist of:
-\begin{itemize}
-\item variables $x$ denoting objects
-\item selections $t.l$
-\item object creations
-\item \mbox{$\newin{x \larrowc t}{u}$} defining an object $x$ of class $t$
-whose scope is the term $u$
-\item class templates $\cls{x: S \sep \tup d}$,
-\item mixin compositions $t \withsub{S} u$ denoting a class of self
-type $S$ combined from two classes $t$ and $u$
-\end{itemize}
+\begin{description}[mixin compositions]
+\item[variables] $x$
+\item[selections] $t.l$
+\item[object creations] \mbox{$\newin{x \larrowc t}{u}$}
+\item[class templates] $\cls{x: S \sep \tup d}$
+\item[mixin compositions] $t \withsub{S} u$
+\end{description}
 \item a value is a variable or class template
@@
 \item types in ${\nu}Obj$ consist of:
+\begin{description}[mixin compositions]
+\item[singleton types] $p.\type$
+\item[type selections] $T\Tsel L$
+\item[record types] $\obj{x \sep \tup D}$
+\item[compound types] $T \with U$
+\item[class types] $\cls{x: S \sep \tup D}$
+\end{description}
+\end{itemize}
+\end{frame}
+
+\begin{frame}{${\nu}Obj$: Details}
 \begin{itemize}
-\item singleton types $p.\type$
-\item type selections $T\Tsel L$
-\item record types $\obj{x \sep \tup D}$
-\item compound types $T \with U$
-\item class types $\cls{x: S \sep \tup D}$.
+\item type bindings
+\begin{description}[abstract types]
+\item[type aliases] $=$
+\item[new types] $\expand$
+\item[abstract types] $<:$
+\end{description}
+\item class template $\cls{x: S \sep \tup d}$ related to translucent sum; some differences:
+\begin{itemize}
+\item needs to be instantiated
+\item ``contractive'' restriction
+\end{itemize}
+\item class type $\cls{x: S \sep \tup D}$
+\begin{itemize}
+\item contains as values classes that instantiate to objects of (sub)type $\obj{x \sep \tup D}$
+\item explicit self type $S$ may be different from $\obj{x \sep \tup D}$
+\item declarations in $S$ which are not in $\tup D$ play the role of abstract members, defined by mixin composition during instantiation
 \end{itemize}

### 2012-09-06 doc/candidacy/slides.tex
vobj: example

@@
 
+\input{macros}
 \input{nuobj}
@@
 
+\begin{frame}[fragile]{${\nu}Obj$: Example (encoding of monomorphic functions)}
+A $\lambda$-abstraction $\lambda ( x : T )\; t$ is represented as a
+class with an abstract member \code!arg! for the function argument:
+\begin{lstlisting}
+[x: {arg: T}| fun = [res = $t'$]]
+\end{lstlisting}
+$t'$ corresponds to term $t$ in which all occurrences of $x$ are
+replaced by \code!$x$.arg!. We cannot directly access \code!arg! on
+the right-hand side of \code!fun!: this wouldn't be ``contractive'',
+because \code!arg! might not be initialized before \code!fun!. Hence,
+we pack $t'$ in another class to ensure contractiveness. Now, an
+application $g(e)$ gets decomposed into three subsequent steps:
+\begin{lstlisting}
+$\nu\,$g$_{app}$  <- g & [arg = e];
+$\nu\,$g$_{eval}$ <- g$_{app}$.fun;
+g$_{eval}$.res
+\end{lstlisting}
+First we instantiate function \code!g! with a concrete argument
+yielding a thunk \code!g$_{app}$!. Then we evaluate this thunk by
+creating an instance \code!g$_{eval}$! of it. Finally we extract the
+result by querying field \code!res! of \code!g$_{eval}$!.
+\end{frame}
+
 \section{Proposal}

### 2012-09-06 doc/candidacy/slides.tex
vobj: discussion

@@
 
+\begin{frame}[fragile]{${\nu}Obj$: Discussion}
+\begin{itemize}
+\item conflate the concepts of compound types (which inherit the
+  members of several parent types) and mixin composition (which build
+  classes from other classes and traits)
+\item mixin composition is not commutative, unlike classical
+  intersection types
+\item in Scala, least upper bounds and greatest lower bounds do not always exist, e.g:
+\end{itemize}
+\begin{minted}{scala}
+  trait A { type T <: A }
+  trait B { type T <: B }
+  // glb is approximated by an infinite sequence
+  A with B { type T <: A with B { type T <: A with B {
+    type T < ...
+  }}}
+\end{minted}
+\end{frame}
+
 \section{Proposal}

### 2012-09-06 doc/candidacy/slides.tex
minor
red
punctuation

@@
 \item modules are not first-class; e.g. not possible to return a
-  structure from an {\tt if} expression.
+  structure from an {\tt if} expression
 \item type sharing is restricted to equality between type names, not
@@
 
-\begin{frame}{ML: Translucent Sums}
+\begin{frame}{ML-Like: Translucent Sums}
 \begin{itemize}
@@
 \item unlike traditional records, later fields can depend on earlier ones
-\item any type can be partially or fully determined by a type
+\item any type can be partially or fully determined by a type expression
 \item dependent typing since translucent sums are terms that may contain type components
@@
 
-\begin{frame}[fragile]{ML: Example}
+\begin{frame}[fragile]{ML-Like: Example}
 \begin{minted}{sml}
@@
 
-\begin{frame}{ML: Path-Dependent Types}
+\begin{frame}{ML-Like: Path-Dependent Types}
 \begin{itemize}
@@
   object which surrounds it; enables ubiquitous access to an object's
-  family without the need to drag around family arguments.
+  family without the need to drag around family arguments
 \end{itemize}
@@
 \begin{itemize}
-\item In Tribe, types are a form of generalized paths: a final
+\item in Tribe, types are a form of generalized paths: a final
   variable, {\tt this} or a class name then followed by a possibly
-  empty sequence of field, {\tt out} or class selections.
+  empty sequence of field, {\tt out} or class selections
 \item {\tt kitt.Passenger.name}: the name of one of Kitt's passengers
 \item {\tt kitt.driver <: Car.driver <: Car.Passenger <: Car.Traveller <:
-  Vehicle.Traveller}.
-\item {\tt kitt.driver $\not<:$ karr.driver} and {\tt kitt.Passenger $\not<:$ karr.Passenger}.
+  Vehicle.Traveller}
+\item {\tt kitt.driver $\not<:$ karr.driver} and {\tt kitt.Passenger $\not<:$ karr.Passenger}
 \end{itemize}

### 2012-09-06 doc/candidacy/slides.tex
rev

@@
 
+\begin{frame}{ML-Like: Calculus}
+grounded in type theory; based on Girard's $F_\omega$, adding:
+\begin{itemize}
+\item {\it translucent sums} to model modules
+\item dependent functions to model functors
+\item a notion of subtyping to model module implementation-interface matching
+\end{itemize}
+\end{frame}
+
 \begin{frame}{ML-Like: Translucent Sums}
@@
 \end{itemize}
-\item translucent sum can be given a more precise type by referring to the name of its member path with a path selection; essential for
+\item translucent sum can be given a more precise type by referring to the name of its type member with a path selection; essential for
 \begin{itemize}

### 2012-09-06 doc/candidacy/slides.tex
rev

@@
 \end{itemize}
-\item translucent sum can be given a more precise type by referring to the name of its type member with a path selection; essential for
+\item translucent sum can be given a more precise type by referring to the name of its type member with a path selection; essential for %example?
 \begin{itemize}
 \item breaking the dependencies between (sub)fields
-\item propagation of typing information
+\item propagating typing information
 \end{itemize}
@@
 \begin{itemize}
-\item {\it familes} of classes inherited together
+\item {\it families} of classes inherited together
 \item classes are lexically nested inside other classes: when a class
@@
 \item no virtual classes in Scala; its virtual types can emulate some (but not all) of the benefits
-\item because of {\tt out} field, Tribe has limited cross-family inheritance
+\item because of {\tt out} field, Tribe has limited cross-family inheritance %example?
 \item soundness? substitution lemma seems wrong when {\tt null} is
   substituted for a variable in an expression whose type contains
-  that variable
+  that variable %details?
 \item aims to have decidable type-checking, but still not proved so

### 2012-09-06 doc/candidacy/slides.tex
syntax

@@
   A with B { type T <: A with B { type T <: A with B {
-    type T < ...
+    type T <: ...
   }}}

### 2012-09-06 doc/candidacy/slides.tex
intro sml ex

@@
 \begin{frame}[fragile]{SML modules: Example}
-TODO
+\inputminted[fontsize=\footnotesize]{sml}{intro.sml}
 \end{frame}
@@
   type T
-  val v = T
-  val f = T -> T
+  val v: T
+  val f: T -> T
 end

### 2012-09-06 doc/candidacy/slides.tex
rev
notes
slide numbers, thanks xeno-by

@@
 \usemintedstyle{eclipse}
+
+\useoutertheme{infolines}
+\setbeamertemplate{headline}{} % removes the headline that infolines inserts
+% \setbeamertemplate{footline}{
+%   \hfill%
+%   \usebeamercolor[fg]{page number in head/foot}%
+%   \usebeamerfont{page number in head/foot}%
+%   \insertpagenumber\,/\,\insertpresentationendpage\kern1em\vskip2pt%
+% }
+\setbeamertemplate{footline}{
+  \hfill%
+  \usebeamercolor[fg]{page number in head/foot}%
+  \usebeamerfont{page number in head/foot}%
+  \insertpagenumber\kern1em\vskip2pt%
+}
+\setbeamertemplate{navigation symbols}{}
 
@@
   int distance(Graph.Node n1, n1.out.Node n2)  { ... }
+  int distance2(Graph g, g.Node n1, g.Node n2) { ... }
 
@@
 \end{itemize}
-A basis for unifying concepts  in advaced object and module systems. Connections:
+A basis for unifying concepts  in advanced object and module systems. Connections:
 \begin{itemize}
@@
 \item[type selections] $T\Tsel L$
-\item[record types] $\obj{x \sep \tup D}$
+\item[record types] $\obj{x \sep \tup D}$ % explain difference between record and class types?
 \item[compound types] $T \with U$

### 2012-09-06 doc/candidacy/slides.tex
notes

@@
   general type expressions; in effect, transparency / opaqueness of a
-  signature cannot be fully fined-tuned
+  signature cannot be fully fined-tuned % example?
 \end{itemize}
@@
 \begin{frame}{ML-Like: Calculus}
-grounded in type theory; based on Girard's $F_\omega$, adding:
+grounded in type theory; based on Girard's $F_\omega$, adding: % quick summary of F_\omega
 \begin{itemize}
@@
 \end{itemize}
-\item translucent sum can be given a more precise type by referring to the name of its type member with a path selection; essential for %example?
+\item translucent sum can be given a more precise type by referring to the name of its type member with a path selection; essential for %example with justification for two items below?
 \begin{itemize}
@@
 
+%maybe one slide on implementation?
+
 \begin{frame}{Tribe: Discussion}
@@
 \item type bindings
-\begin{description}[abstract types]
+\begin{description}[abstract types] %examples?
 \item[type aliases] $=$

### 2012-09-07 doc/candidacy/slides.tex
proposal slide

@@
 
+% can you make this slide less cluttered?
 \begin{frame}[fragile]{${\nu}Obj$: Example (encoding of monomorphic functions)}
@@
   intersection types
-\item in Scala, least upper bounds and greatest lower bounds do not always exist, e.g:
+\item in Scala, least upper bounds and greatest lower bounds do not always exist, e.g: %can you translate this to nuObj?
 \end{itemize}
@@
 \section{Proposal}
+\begin{frame}{DOT: Dependent Object Types}
+\begin{itemize}
+\item core calculus for modeling {\it path-dependent types}
+\item we've seen path-dependent types arise in three settings:
+\begin{itemize}
+\item ML-like module systems
+\item virtual classes
+\item ${\nu}Obj$ / Scala
+\end{itemize}
+\item DOT aims to bring more uniformity and simplicity to Scala
+\begin{itemize}
+\item replace Scala's compound types with classical intersection types
+\item complement calculus with classical union types
+\item intersections and unions form a lattice wrt subtyping
+\end{itemize}
+\end{itemize}
+\end{frame}
 

### 2012-09-07 doc/candidacy/slides.tex
cleanup dead-code
rev

@@
 \useoutertheme{infolines}
-\setbeamertemplate{headline}{} % removes the headline that infolines inserts
-% \setbeamertemplate{footline}{
-%   \hfill%
-%   \usebeamercolor[fg]{page number in head/foot}%
-%   \usebeamerfont{page number in head/foot}%
-%   \insertpagenumber\,/\,\insertpresentationendpage\kern1em\vskip2pt%
-% }
+\setbeamertemplate{headline}{}
 \setbeamertemplate{footline}{
-  \hfill%
-  \usebeamercolor[fg]{page number in head/foot}%
-  \usebeamerfont{page number in head/foot}%
-  \insertpagenumber\kern1em\vskip2pt%
+  \hfill
+  \usebeamercolor[fg]{page number in head/foot}
+  \usebeamerfont{page number in head/foot}
+  \insertpagenumber\kern1em\vskip2pt
 }
@@
 \begin{frame}{SML modules}
-\begin{description}[structure]
-\item[structure] a module; a collection of types, values and
+\begin{description}[type-sharing]
+\item[structure] a module\\a collection of types, values and
   (sub)structures packaged as a program unit.
-\item[signature] an interface; the type of a structure; description of
+\item[signature] an interface; the type of a structure\\description of
   the types, values and (sub)structures of a structure given by their
@@
 \item unlike traditional records, later fields can depend on earlier ones
+\item fields can be types or terms
 \item any type can be partially or fully determined by a type expression
@@
 \item translucent sum can be given a more precise type by referring to the name of its type member with a path selection; essential for %example with justification for two items below?
+not clear without introducing internal names?
 \begin{itemize}
@@
 \item[class templates] $\cls{x: S \sep \tup d}$
-\item[mixin compositions] $t \withsub{S} u$
+\item[mixin compositions] $t_1 \withsub{S} t_2$\\(if omitted, $S = {S_1 \with (S_2 \with
+  \obj{x \sep D_1 \uplus D_2})}$)
 \end{description}

### 2012-09-07 doc/candidacy/slides.tex
ex of limitation, with caveat

@@
 \end{description}
-% example?
 \end{frame}
@@
 
-\begin{frame}{Limitations of SML modules}
+\begin{frame}[fragile]{Limitations of SML modules}
 \begin{itemize}
@@
 \item type sharing is restricted to equality between type names, not
-  general type expressions; in effect, transparency / opaqueness of a
-  signature cannot be fully fined-tuned % example?
+  general type expressions (limitation of SML '90; lifted in SML '97)
+\begin{minted}{sml}
+signature X' = sig
+  type T
+  type T2T = T -> T
+  val v: T
+  val f: T2T
+end
+\end{minted}
+\item in effect, transparency / opaqueness of a signature cannot be
+  fully fined-tuned
 \end{itemize}

### 2012-09-07 doc/candidacy/slides.tex
confs

@@
   Modules with Sharing}\\{\footnotesize\textcolor{gray}{by Robert Harper and Mark
-  Lillibridge}}\begin{itemize}
+  Lillibridge\\POPL '94}}\begin{itemize}
 \item calculus for modules in the ML tradition
@@
 \item {\it Tribe: a simple virtual class calculus}\\
-  {\footnotesize\textcolor{gray}{by Dave Clarke, Sophia Drossopoulou, James Noble and Tobias Wrigstad}}\begin{itemize}
+  {\footnotesize\textcolor{gray}{by Dave Clarke, Sophia Drossopoulou, James Noble and Tobias Wrigstad\\AOSD '07}}\begin{itemize}
 \item calculus for virtual classes
@@
   Types}\\{\footnotesize\textcolor{gray}{by Martin Odersky, Vincent Cremet, Christine R{\"o}ckl and
-  Matthias Zenger}}\begin{itemize}
+  Matthias Zenger\\ECOOP '03}}\begin{itemize}
 \item calculus unifying advanced object-oriented and module systems

### 2012-09-07 doc/candidacy/slides.tex
formatting
system F_omega backup slide

@@
 
-\inputminted[frame=lines,fontsize=\footnotesize]{scala}{intro.scala}
+\inputminted[fontsize=\footnotesize]{scala}{intro.scala}
 \end{frame}
@@
 \begin{frame}{ML-Like: Calculus}
-grounded in type theory; based on Girard's $F_\omega$, adding: % quick summary of F_\omega
+grounded in type theory; based on Girard's $F_\omega$, adding:
 \begin{itemize}
@@
 \end{itemize}
-\begin{minted}[frame=lines,fontsize=\footnotesize]{java}
+\begin{minted}[fontsize=\footnotesize]{java}
 class Vehicle { class Traveller { ... } }
@@
 
-% System F omega
+\begin{frame}[fragile]{System $F_\omega$: Higher-order polymorphic lambda-calculus}
+\begin{description}[1234567890]
+\item[terms $t$]
+\begin{description}[123456789012345678901234567890]
+\item[typed lambda-calculus terms] $x$ -- $\lambda x:T.t$ -- $t t$
+\item[type abstraction] $\lambda X::K.t$
+\item[type application] $t [T]$
+\end{description}
+\item[types $T$]
+\begin{description}[123456789012345678901234567890]
+\item[type variable] $X$
+\item[type of functions] $T \rightarrow T$
+\item[universal type] $\forall X::K.T$
+\item[operator abstraction] $\lambda X::K.T$
+\item[operator application] $T T$
+\end{description}
+\item[kinds $K$]
+\begin{description}[123456789012345678901234567890]
+\item[kind of proper types] $*$
+\item[kind of operators] $K \Rightarrow K$
+\end{description}
+\end{description}
+\end{frame}
+
 % System F sub

### 2012-09-09 doc/fool/article.tex
simple changes

--- 2012-08-11 article.tex
@@
 of nominal and structural typing through the use of refinement
-types. It makes no attempt to model inheritance or mixing
+types. It makes no attempt to model inheritance or mixin
 composition. The calculus does not model what's currently in Scala: it
@@
 {\bf with} and classical type intersection, as it is modeled in the
-calculus. Scala, and the previous calculi attempting to model it
+calculus. Scala, and the previous calculi attempting to model it,
 conflates the concepts of compound types (which inherit the members of
@@
 and greatest lower bounds do not always exist. Here is an example:
-Given two traits
+given two traits
 \begin{lstlisting}
@@
 
-The DOT calculus is a simple system of dependent
+The DOT calculus is a small system of dependent
 object-types. \figref{dot-one} gives its syntax, reduction rules,
@@
 \subsection{Notation} We use standard notational conventions for
-sets. The notation $\seq{X}$ denotes a set of elements $X$. Given a
+sets. The notation $\seq{X}$ denotes a set of elements $X$. Given
 such a set $\seq X$ in a typing rule, $X_i$ denotes an arbitrary
@@
 
-Two subclasses of terms are values $v$, which consist of just
+Two sub-sorts of terms are values $v$, which consist of just
 variables, and paths $p$ which consist of just variables and field
@@
       \infrule[\textsc{rfn-$\expand$}]
-      {\Gamma \ts T \expand_z {\seq D'}}
+      {\Gamma \ts T \expand_z {\seq {D'}}}
       {\Gamma \ts T \refine z {\seq D} \expand_z \seq {D'} \tand \seq D}
@@
 Rule (\textsc{rfn-$\expand$}) states that a refinement type $T
-\expand_z {\seq D}$ expands to the conjunction of the expansion $D'$
-of $T$ and the newly added declarations $D$. Rule
+\expand_z {\seq D}$ expands to the conjunction of the expansion $\seq {D'}$
+of $T$ and the newly added declarations $\seq D$. Rule
 (\textsc{tsel-$\expand$}) states that a type selection $p.L$ carries
@@
 \app f x &\iff \mapp f {\mi{apply}} x\\
-\cast T t&\iff \app {\fun x T T x} t
+\cast T t&\iff \app {(\fun x T T x)} t
 \end{align*}
@@
 & (T = T_1 \tand T_2 \andl y \in \relv k \Gamma s {T_1} \andl y \in \relv k \Gamma s {T_2}) \orl\\
-& (T = T_1 \tor T_2 \andl  (y \in \relv k \Gamma s {T_1} \orl y \in \relv k \Gamma s {T_2})) \orl\\
-& (T = p.L \andl \Gamma \ts p \ni L \typ {S'..U'} \andl \forall j < k, y \in \relv j \Gamma s {U'})\\
+& (T = T_1 \tor T_2 \andl  (y \in \relv k \Gamma s {T_1} \orl y \in \relv k \Gamma s {T_2}))\\
 &)\}

### 2012-09-11 doc/candidacy/slides.tex
system Fsub
ml section rev

@@
 
-\begin{frame}{SML modules}
+\begin{frame}[fragile]{SML modules: Structures and Signatures}
+\begin{columns}
+\begin{column}[t]{5cm}
+\begin{minted}{sml}
+structure A = struct
+  type T = int
+  val v = 0
+  val f = fn x => x + 1
+end
+
+signature X = sig
+  type T
+  val v: T
+  val f: T -> T
+end
+
+structure B :> X = A
+\end{minted}
+\end{column}
+\begin{column}[t]{5cm}
+\begin{minted}{sml}
+(* A.T transparent *)
+val vA: int = A.v
+
+(* B.T opaque *)
+val vB: B.T = B.v
+\end{minted}
+\end{column}
+\end{columns}
+\end{frame}
+
+\begin{frame}[fragile]{SML modules: Functions and Type Sharing}
+\begin{minted}{sml}
+functor mkDoubler(arg: X) :> X where type T = arg.T = struct
+  type T = arg.T
+  val v = arg.v
+  val f = arg.f o arg.f
+end
+
+(* A2.T = A.T = int *)
+structure A2 = mkDoubler(A);
+val a: int = A2.f(A.v);
+
+(* B2.T = B.T *)
+structure B2 = mkDoubler(B);
+val b: B.T = B2.f(B.v);
+\end{minted}
+\end{frame}
+
+\begin{frame}{SML modules: Recap}
 \begin{description}[type-sharing]
@@
 \end{description}
-\end{frame}
-
-\begin{frame}[fragile]{SML modules: Example}
-\inputminted[fontsize=\footnotesize]{sml}{intro.sml}
 \end{frame}
@@
 
-\begin{frame}{ML-Like: Calculus}
+\begin{frame}{``Translucent'' ML: Calculus}
 grounded in type theory; based on Girard's $F_\omega$, adding:
@@
 
-\begin{frame}{ML-Like: Translucent Sums}
+\begin{frame}{``Translucent'' ML: Translucent Sums}
 \begin{itemize}
@@
 
-\begin{frame}[fragile]{ML-Like: Example}
+\begin{frame}[fragile]{``Translucent'' ML: Example}
 \begin{minted}{sml}
@@
 
-% Maintaing Soundness?
-
-\begin{frame}{ML-Like: Path-Dependent Types}
+\begin{frame}{``Translucent'' ML: Path-Dependent Types}
 \begin{itemize}
@@
 \begin{description}[123456789012345678901234567890]
-\item[typed lambda-calculus terms] $x$ -- $\lambda x:T.t$ -- $t t$
+\item[typed lambda-calculus terms] $x$ -- $\lambda x:T.t$ -- $t\ t$
 \item[type abstraction] $\lambda X::K.t$
@@
 \item[operator abstraction] $\lambda X::K.T$
-\item[operator application] $T T$
+\item[operator application] $T\ T$
 \end{description}
@@
 
-% System F sub
+\begin{frame}[fragile]{System $F_{<:}$: Bounded Quantification}
+\begin{description}[1234567890]
+\item[terms $t$]
+\begin{description}[123456789012345678901234567890]
+\item[typed lambda-calculus terms] $x$ -- ${\lambda}x:T.t$ -- $t\ t$
+\item[type abstraction] $\lambda X<:T.t$
+\item[type application] $t [T]$
+\end{description}
+\item[types $T$]
+\begin{description}[123456789012345678901234567890]
+\item[type variable] $X$
+\item[maximum type] $\top$
+\item[type of functions] $T \rightarrow T$
+\item[universal type] $\forall X<:T.T$
+\end{description}
+\end{description}
+\end{frame}
+
 \end{document}

### 2012-09-11 doc/candidacy/slides.tex
tribe rev
nuobj rev

@@
 
-\begin{frame}{Tribe Calculus: Virtual Classes}
-\begin{itemize}
-\item {\it families} of classes inherited together
-\item classes are lexically nested inside other classes: when a class
-  is inherited, its nested inner classes are inherited along with
-  their methods and fields
-\item two form of inheritance: subclassing and further binding
-\item {\it family polymorphism}: code written for one family
-  also works for extensions of that family
-\item in Tribe:\begin{itemize}
-\item path types can depend simultaneously on both classes and objects
-\item paths can use an {\tt out} field to move from an object to the
-  object which surrounds it; enables ubiquitous access to an object's
-  family without the need to drag around family arguments
-\end{itemize}
-\end{itemize}
-\end{frame}
-
-\begin{frame}[fragile]{Tribe: Example}
+\begin{frame}[fragile]{Tribe: Example (Virtual Classes)}
 \begin{minted}{java}
@@
 
-\begin{frame}[fragile]{Tribe: Example (continued)}
+\begin{frame}[fragile]{Tribe: Example (Object Family and Family Polymorphism)}
 \begin{minted}{java}
@@
 
-%maybe one slide on implementation?
+\begin{frame}{Tribe Calculus: Recap}
+\begin{itemize}
+\item {\it families} of classes inherited together
+\item classes are lexically nested inside other classes: when a class
+  is inherited, its nested inner classes are inherited along with
+  their methods and fields
+\item two form of inheritance: subclassing and further binding
+\item {\it family polymorphism}: code written for one family
+  also works for extensions of that family
+\item in Tribe:\begin{itemize}
+\item path types can depend simultaneously on both classes and objects
+\item paths can use an {\tt out} field to move from an object to the
+  object which surrounds it; enables ubiquitous access to an object's
+  family without the need to drag around family arguments
+\end{itemize}
+\end{itemize}
+\end{frame}
 
-\begin{frame}{Tribe: Discussion}
+\begin{frame}[fragile]{Tribe: Discussion}
 \begin{itemize}
@@
 \item because of {\tt out} field, Tribe has limited cross-family inheritance %example?
+\begin{minted}[fontsize=\footnotesize]{java}
+class A {
+  class D { ... }
+  class B {
+    class C {
+      this.out.out.D foo;
+    }
+  }
+}
+class E {
+  class F extends A.B.C { ... }
+}
+\end{minted}
 \item soundness? substitution lemma seems wrong when {\tt null} is
@@
 \item[record types] $\obj{x \sep \tup D}$ % explain difference between record and class types?
+\item[class types] $\cls{x: S \sep \tup D}$
 \item[compound types] $T \with U$
-\item[class types] $\cls{x: S \sep \tup D}$
 \end{description}
@@
 
-% can you make this slide less cluttered?
 \begin{frame}[fragile]{${\nu}Obj$: Example (encoding of monomorphic functions)}
-A $\lambda$-abstraction $\lambda ( x : T )\; t$ is represented as a
-class with an abstract member \code!arg! for the function argument:
-\begin{lstlisting}
+\begin{itemize}
+\item Encoding for a $\lambda$-abstraction $\lambda ( x : T )\; t$:
+\begin{lstlisting}[basicstyle=\large]
 [x: {arg: T}| fun = [res = $t'$]]
 \end{lstlisting}
-$t'$ corresponds to term $t$ in which all occurrences of $x$ are
-replaced by \code!$x$.arg!. We cannot directly access \code!arg! on
-the right-hand side of \code!fun!: this wouldn't be ``contractive'',
-because \code!arg! might not be initialized before \code!fun!. Hence,
-we pack $t'$ in another class to ensure contractiveness. Now, an
-application $g(e)$ gets decomposed into three subsequent steps:
-\begin{lstlisting}
+($t'$ = $t$ with \code{$x$.arg} substituted for $x$)
+
+\item Encoding for an application $g(e)$:
+\begin{lstlisting}[basicstyle=\large]
 $\nu\,$g$_{app}$  <- g & [arg = e];
@@
 \end{lstlisting}
-First we instantiate function \code!g! with a concrete argument
-yielding a thunk \code!g$_{app}$!. Then we evaluate this thunk by
-creating an instance \code!g$_{eval}$! of it. Finally we extract the
-result by querying field \code!res! of \code!g$_{eval}$!.
+\end{itemize}
 \end{frame}
@@
   trait B { type T <: B }
-  // glb is approximated by an infinite sequence
+  // glb is an infinite sequence
   A with B { type T <: A with B { type T <: A with B {

### 2012-09-11 doc/candidacy/slides.tex
formatting

@@
 \begin{frame}[fragile]{``Translucent'' ML: Example}
+\begin{columns}
+\begin{column}[t]{5cm}
 \begin{minted}{sml}
@@
 end
+
 structure X =
   if flip() then XInt else XBool
+\end{minted}
+\end{column}
+\begin{column}[t]{5cm}
+\begin{minted}{sml}
 signature XType = sig
@@
 \end{minted}
+\end{column}
+\end{columns}
 \end{frame}

### 2012-09-12 doc/candidacy/slides.tex
bump the slide numbers a bit up for projector to show them hopefully
typo
more refined intro

@@
   \usebeamerfont{page number in head/foot}
-  \insertpagenumber\kern1em\vskip2pt
+  \insertpagenumber\kern1em\vskip10pt
 }
@@
 
-\inputminted[fontsize=\footnotesize]{scala}{intro.scala}
+\begin{columns}
+\begin{column}[t]{5cm}
+\begin{minted}[fontsize=\footnotesize]{scala}
+abstract class AbsCell {
+  type T
+  val init: T
+  private var value: T = init
+  def get = value
+  def set(x: T) { value = x }
+}
+
+object Library {
+  def update(c: AbsCell)(oldval: c.T, newval: c.T) =
+    if (oldval == c.get) { c.set(newval); true }
+    else false
+}
+\end{minted}
+\end{column}
+\begin{column}[t]{5cm}
+\begin{minted}[fontsize=\footnotesize]{scala}
+val c = new AbsCell {
+  type T = Int
+  val init = 1
+}
+c.set(2)
+val a: AbsCell = c
+// a.T is opaque
+a.set(a.init)
+\end{minted}
+\end{column}
+\end{columns}
 \end{frame}
@@
 
-\begin{frame}[fragile]{SML modules: Functions and Type Sharing}
+\begin{frame}[fragile]{SML modules: Functors and Type Sharing}
 \begin{minted}{sml}

### 2012-09-12 doc/candidacy/slides.tex
proposal

@@
 \input{nuobj}
+\input{dot_macros}
 
@@
 
+\begin{frame}{DOT: Status}
+\begin{itemize}
+\item basic calculus
+\item type safety
+\begin{itemize}
+\item showed by counterexamples that subject reduction doesn't hold
+\item sketch of plausible proof via logical relations
+\end{itemize}
+\item submission about this work-in-progress accepted to Foundations
+  of Object-Oriented Languages (FOOL '12)
+\item future work
+\begin{itemize}
+\item confirm proof of type safety and/or refine calculus
+\item investigate translations of Scala features into DOT
+\item implement a compiler front-end which uses DOT for typechecking
+\end{itemize}
+\end{itemize}
+\end{frame}
+
 \section{Backup Slides}
+
+\begin{frame}[fragile]{DOT: Syntax}
+\begin{itemize}
+\item terms
+\begin{description}[method invocations]
+\item[variables] $x$, $y$, $z$
+\item[selections] $t.l$
+\item[method invocations] $t.m(t)$
+\item[object creations] $\new y c {t'}$\\
+$c$ is a constructor $T_c \ldefs{\seq{l = v}\;\seq{m(x) = t}}$
+\end{description}
+\item types
+\begin{description}[type intersections]
+\item[type selections] $p.L$
+\item[refinement types] $T \refine z {\seq D}$
+\item[type intersections] $T \tand T'$
+\item[type unions] $T \tor T'$
+\item[a top type] $\Top$
+\item[a bottom type] $\Bot$
+\end{description}
+\end{itemize}
+\end{frame}
+
+\begin{frame}[fragile]
+\frametitle{DOT: Preservation Counterex. (Well-Formedness Lost)}
+\center{// $y.A$ is not well-formed after $v$ is substituted for $x$.}
+\begin{align*}
+&\mlnew v {\Top \refine z {\Ldecl L \Bot {\Top \refine z {\Ldecl A \Bot \Top, \Ldecl B {z.A} {z.A}}}} \ldefs{}}{
+\mlapp {\dotabs x {\Top \refine z {\Ldecl L \Bot {\Top \refine z {\Ldecl A \Bot \Top, \Ldecl B \Bot \Top}}}} {\\&\gap
+\mlnew z {\Top \refine z {\ldecl l {\Bot \tfun \Top}}\mlldefs{\\&\gap\gap l = \dotabs y {x.L \tand {\Top \refine z {\Ldecl A {z.B} {z.B}, \Ldecl B \Bot \Top}}} {\\&\gap\gap\gap\dotabs a {y.A} {\app {(\dotabs x \Top x)} a}}}}{
+\ \app {(\dotabs x \Top x)} z
+}}}{\ v}}
+\end{align*}
+\end{frame}
 

### 2012-09-13 doc/fool/article.tex
put the redex generated example in
comment out wip ex section
minor clarifications

@@
 {\Gamma \ts v \typ {V'} \spcomma {V'} \sub V}
-{\Gamma \ts l = v \typ \ldecl l V}
+{\Gamma \ts (l = v) \typ (\ldecl l V)}
 
@@
  \Gamma \envplus{x : S} \ts t \typ {T'} \spcomma {T'} \sub T}
-{\Gamma \ts m(x) = t \typ \mdecl m S T }
+{\Gamma \ts (m(x) = t) \typ (\mdecl m S T) }
 
@@
 
-Two sub-sorts of terms are values $v$, which consist of just
-variables, and paths $p$ which consist of just variables and field
+Two sub-sorts of terms are values and paths. Values $v$ consist of
+just variables. Paths $p$ consist of just variables and field
 selections.
@@
 $\nswfe$-types, this limitation allows us to show that subtyping is
-transitive, as discussed in section~\ref{dot-preservation}.
+transitive, as discussed in section~\ref{dot-variants-why-wfe}.
 
@@
 
-\subsubsection{Why limit the calculus to $\nswfe$-types?}
+\subsubsection{Why limit the calculus to $\nswfe$-types?}\label{dot-variants-why-wfe}
 
@@
 
+%\section{Examples}
+%TODO: WIP
+%
+%\includegraphics[scale=0.7]{../../src/redex/ex_glb}
+
 %\appendix

### 2012-09-13 doc/fool/article.tex
extra and, but nothing missing
fixing and reducing claim of list example -- remove it?

@@
 
-\subsection{Polymorphic Lists}
+\subsection{Lists}
 Polymorphic lists can be expressed using an abstract type member for
 the element type ($\mi{Elem}$). We can instantiate a refinement of the
-list package to manipulate lists with a particular element type. We
-can choose to make the lists invariant or covariant in the element
-type by fixing the lower bound of the $\mi{Elem}$ re-declaration to
-the upper bound or to $\Bot$, respectively.
+list package to manipulate lists with a particular element type.
 \begin{align*}
@@
 &\ \Ldecl {\mi{Elem}} \Bot \Top\\\displaybreak[3]
-&\ \Ldecl {\mi{ListOfElement}} {g.\mi{List} \refine z {\Ldecl {\mi{Elem}} {p.\mi{Elem}} {p.\mi{Elem}}}} {\\&\gap\gap\gap\gap g.\mi{List} \refine z {\Ldecl {\mi{Elem}} {p.\mi{Elem}} {p.\mi{Elem}}}}\\\displaybreak[3]
 &\ \Ldecl {\mi{List}} \Bot {\Top \mlrefine z {\\\displaybreak[3]
 &\ \gap \mdecl {\mi{isEmpty}} {\mi{root}.\mi{Unit}} {\mi{root}.\mi{Boolean}}\\\displaybreak[3]
-&\ \gap \mdecl {\mi{head}} {\mi{root}.\mi{Unit}} {g.\mi{Elem}}\\\displaybreak[3]
-&\ \gap \mdecl {\mi{tail}} {\mi{root}.\mi{Unit}} {g.\mi{ListOfElem}}\\\displaybreak[3]
+&\ \gap \mdecl {\mi{head}} {\mi{root}.\mi{Unit}} {p.\mi{Elem}}\\\displaybreak[3]
+&\ \gap \mdecl {\mi{tail}} {\mi{root}.\mi{Unit}} {p.\mi{List}}\\\displaybreak[3]
 &\gap}}\\\displaybreak[3]
-&\ \mdecl {\mi{nil}} {\mi{root}.\mi{Unit}} {g.\mi{ListOfElem}}\\\displaybreak[3]
-&\ \mdecl {\mi{cons}} {g.\mi{Elem}} {g.\mi{ListOfElem}}\\\displaybreak[3]
+&\ \mdecl {\mi{nil}} {\mi{root}.\mi{Unit}} {p.\mi{List}}\\\displaybreak[3]
+&\ \mdecl {\mi{cons}} {p.\mi{Elem}} {\arrow {p.\mi{List}} {p.\mi{List}}}\\\displaybreak[3]
 &}}}}{
@@
 &\ {\mi{nil}}(u) = \ldots\\\displaybreak[3]
-&\ {\mi{cons}}(u) = \ldots\\\displaybreak[3]
+&\ {\mi{cons}}(hd) = \ldots\\\displaybreak[3]
 &}}}{
@@
 &\relg k s \Gamma = \{ \Gamma, \seq{x_i : T_{c_i}}^{m \leq i < n} \ |\\
-&\gap \Gamma = \seq{x : T}^m \andl s = \seq{x \mapsto c}^n \andl\\
+&\gap \Gamma = \seq{x : T}^m \andl s = \seq{x \mapsto c}^n\\
 &\}

### 2012-09-13 doc/fool/article.tex
separate out logical relation proof

@@
 
-\section{Type Safety via Logical Relations}\label{dot-type-safety}
-We sketch a proof of type-safety of the DOT calculus via step-indexed
-logical relations~\cite{ahmed04,ahmed06,step_indexed_obj}.
-
-\subsection{Type Safety}
-Type-safety states that a well-typed program doesn't get stuck. More
-formally: If $\emptyset \ts t \typ T$ and $\reductionl t \emptyset {t'} {s'} *$ then
-either $t'$ is a value or $\exists t'', s''. \reduction {t'} {s'}
-{t''} {s''}$.
-
-Our strategy is to define a logical relation $\Gamma \ds t : T$, such
-that $\Gamma \ts t : T$ implies $\Gamma \ds t : T$ implies
-type-safety.
-
-\subsection{Step-Indexed Logical Relations}
-
-In order to ensure that our logical relation is well-founded, we use a
-step index. For each step index $k$, we define the set of values and
-the set of terms that appear to belong to a given type, when taking at
-most $k$ steps. $\Gamma \ds t : T$ is then defined in terms of the
-step-indexed logical relation by requiring it to hold $\forall k$.
-
-\subsubsection{Set of Values}
-$\relv k \Gamma s T$ defines the set of values that appear to have
-type $T$ when taking at most $k$ steps. $\Gamma$ and $s$ must agree:
-$\dom(\Gamma) = \dom(s)$ (ordered) and $\forall (x : T) \in \Gamma, x
-\in \relv k \Gamma s T$. A variable $y$ belongs to $\relv 0 \Gamma s
-T$ simply by being in the store. In addition, it belongs to $\relv k
-\Gamma s T$ for $k > 0$, if it defines all type, method and value labels in
-the expansion of $T$ appropriately for $j < k$ steps.
-\begin{align*}
-&\relv k \Gamma s T = \{ y \ | y \in \dom(s) \andl (\\
-& (\Gamma \ts T \wfe \andl\\
-&\gap \forall j < k,\\
-&\gap y \mapsto T_c \ldefs{\seq{l = v}\;\seq{m(x)=t}} \in s,\\
-&\gap \Gamma \ts T \expand_y \seq{D},\\
-&\gap (\forall L_i : S \rightarrow U \in \seq{D},\\
-&\gap\gap \Gamma \ts y \ni L_i \typ {S'..U'}) \andl\\
-&\gap (\forall m_i : S \rightarrow U \in \seq{D},\\
-&\gap\gap t_i \in \rele j {\Gamma \envplus{x_i : S}} s U) \andl\\
-&\gap (\forall l_i : V \in \seq{D},\\
-&\gap\gap v_i \in \relv j \Gamma s V)) \orl\\
-& (T = T_1 \tand T_2 \andl y \in \relv k \Gamma s {T_1} \andl y \in \relv k \Gamma s {T_2}) \orl\\
-& (T = T_1 \tor T_2 \andl  (y \in \relv k \Gamma s {T_1} \orl y \in \relv k \Gamma s {T_2}))\\
-&)\}
-\end{align*}
-
-This relation captures the observation that the only ways for a term
-to get stuck is to have a field selection on an uninitialized field or
-a method invocation on an uninitialized method. However, a {\it potential
-  pitfall} is that the value itself might occur in the types $S$, $U$,
-$V$, because we substitute it for the ``self'' occurrences in the
-expansion, so the relation makes sure that the required type labels exist.
-
-\subsubsection{Set of Terms}
-$\rele k \Gamma s T$ defines the set of terms that appear to have type
-$T$ when taking at most $k$ steps. $s$ must agree with a {\it prefix}
-of $\Gamma$, so $\Gamma$ can additionally contain variables not in
-$s$. This is needed for checking methods in $\mathcal{V}$ above, and
-for relating open terms. If $k > 0$, $\mathcal{E}$ extends $\Gamma$
-and $s$ so that they agree. It then states that if it can reduce $t$
-in the extended store to an irreducible term in $j < k$ steps, then
-this term must be in a corresponding $\mathcal{V}$ set with $\Gamma$
-now extended to agree with the store resulting from the reduction
-steps.
-
-$\irred {t} {s}$ is a shorthand for $\neg\exists t', s'. \reduction t s
-{t'} {s'}$. $\mathcal{\supseteq}$ is used initially for the possibly
-shorter store to agree with the environment, and can extend both in
-many different ways. $\mathcal{\supseteq}^!$ is used finally for the
-possibly shorter environment to agree with the store, and just extends
-the environment in one straightforward way: hence, it defines
-singleton sets.
-
-\begin{align*}
-&\rele k \Gamma s T = \{ t \ |\\
-&\gap k = 0 \orl (\forall j < k,\\
-&\gap\gap \forall (\Gamma'; s') \in \rels k \Gamma s ,\\
-&\gap\gap \reductionl t {s'} {t'} {s''} j \andl\\
-&\gap\gap \irred {t'} {s''} \impliesl\\
-&\gap\gap \forall \Gamma'' \in \relg k {s''} {\Gamma'} ,\\
-&\gap\gap {t'} \in \relv {k-j-1} {\Gamma''} {s''} T)\\
-&\}
-\end{align*}
-
-\subsubsection{Extending the environment and the store}
-$\rels k \Gamma s$ for $k > 0$ defines the set of completed
-environment and stores that agree on $k-1$ steps, and that extend
-$\Gamma$ and $s$. $s$ must agree with a {\it prefix} of $\Gamma$.
-Both $\Gamma$ and $s$ are ordered maps. For $s$, $s'$ extends $s$ if
-$s$ is a prefix of $s'$. For $\Gamma$, $\Gamma'$ extends $\Gamma$ if
-we get back $\Gamma$ by keeping only the elements of $\Gamma'$ that
-belong to $\Gamma$. Furthermore, a prefix of $\Gamma'$ agrees with
-$s$.
-\begin{align*}
-&\rels k \Gamma s = \{\\
-&\ (\seq{x : T}^m, \seq{x_{ij} : T_{ij}}^{m \leq i < n; 0 \leq j < i_n}; s, \seq{x_{ij} \mapsto c_{ij}}^{m \leq i < n; 0 \leq j < i_n}) |\\
-&\ s = \seq{x \mapsto c}^m \andl \Gamma = \seq{x : T}^n \andl \\
-&\ m\leq n \andl \forall i, m \leq i < n, \forall {i_n}, j, 0 \leq j < i_n,\\
-&\ \forall T_{ij}, c_{ij}, T_{i{(i_n-1)}} = T_i, \forall n' \leq n, i_n' \leq i_n,\\
-&\ c_{ij} \in \relv {k-1} {\seq{x : T}^m, \seq{x_{ij} : T_{ij}}^{m \leq i' < n; 0 \leq j < i_n'}} {s, \seq{x_{ij} \mapsto c_{ij}}^{m \leq i < n'; 0 \leq j < i_n'}} {T_{ij}}\\
-&\}
-\end{align*}
-
-\subsubsection{Completing the environment to agree with the store}
-$\relg k s \Gamma$ defines a singleton set of a completed environment
-that agrees with a store $s$ by simply copying the constructor type
-from the store for each missing variable.
-\begin{align*}
-&\relg k s \Gamma = \{ \Gamma, \seq{x_i : T_{c_i}}^{m \leq i < n} \ |\\
-&\gap \Gamma = \seq{x : T}^m \andl s = \seq{x \mapsto c}^n\\
-&\}
-\end{align*}
-
-\subsubsection{Terms in the Logical Relation}
-
-$\Gamma \ds t : T$ is simply defined as $t \in \rele k \Gamma \emptyset T,
-\forall k$.
-
-\subsection{Statements and Proofs}
-
-\subsubsection{Fundamental Theorem}
-
-The fundamental theorem is the implication from $\Gamma \ts t : T$ to
-$\Gamma \ds t : T$. Type safety is a straightforward corollary of this
-theorem.
-
-\begin{myproof}
-The proof is on induction on the derivation of $\Gamma \ts t : T$. For
-each case, we need to show $t \in \rele k \Gamma \emptyset T, \forall
-k$. The non-trivial case is when $k > 0$ and for $(\Gamma'; s') \in
-\rels k \Gamma s$ and some $j < k$, $\reductionl {t} s {t'} {s'} j
-\andl \irred {t'} {s'}$. Then, we need to show ${t'} \in \relv {k-j-1}
-      {\Gamma''} {s'} T$ for ${\Gamma''} \in \relg \Gamma k {s'}
-      {\Gamma'}$.
-
-\ \\
-
-\Case \textsc{var}: $\Gamma \ts x : T$ knowing $(x : T) \in
-\Gamma$. $x \in \relv {k-1} {\Gamma'} s T$ follows from the definition of
-$\rels k \Gamma \emptyset$.
-
-\ \\
-
-\Case \textsc{sel}: $\Gamma \ts t_1.l_i : T$ knowing $\Gamma \ts t_1 :
-T_1$, $\Gamma \ts T_1 \expand_z \seq{D}$, $\ldecl {l_i} {V_i} \in
-\seq{D}$ and knowing either that $t_1 = p_1 \andl T = \subst p z
-    {V_i}$ or that $z \not\in \fn(V_i) \andl T = V_i$.
-
-By operational semantics and induction hypothesis, $\reductionl {t_1}
-s {t_1'} {s'} {j-1}$ and $\irred {t_1'} {s'}$ and ${t_1'} \in \relv
-{k-j+1-1} {\Gamma'} {s'} {T_1}$.
-
-By operational semantics and the above, $\reductionl {t_1'.l_i} {s'}
-{t'} {s'} 1$, and we can conclude $t' \in \relv {k-j-1} {\Gamma''} {s'}
-T$ from the clause for value labels of ${t_1'} \in \relv {k-j}
-{\Gamma''} {s'} {T_1}$.
-
-\ \\
-
-\Case \textsc{msel}: $\Gamma \ts t_1.m_i(t_2) : T$ knowing $\Gamma \ts
-t_1 : T_1$, $\Gamma \ts t_2 : T_2$, $\Gamma \ts T_1 \expand_z
-\seq{D}$, $\mdecl {m_i} {S_i} {U_i} \in \seq{D}$ and knowing either
-that $t_1 = p_1 \andl S = \subst p z {S_i} \andl T = \subst p z {U_i}$
-or that $z \not\in \fn(S_i) \andl z \not\in \fn(U_i) \andl S = S_i \andl
-T = U_i$, and knowing that $\Gamma \ts T_2 \sub S$.
-
-By operational semantics and induction hypotheses, $\reductionl {t_1}
-s {t_1'} {s_1} {j_1}$ and $\irred {t_1'} {s_1}$ and $\reductionl {t_2}
-{s} {t_2'} {s_2} {j_2}$ and $\irred {t_2'} {s_2}$ and ${t_1'} \in
-\relv {k-j_1-1} {\Gamma_1} {s_1} {T_1}$ and ${t_2'} \in \relv
-{k-j_2-1} {\Gamma_2} {s_2} {T_2}$.
-
-Because $t_2$ reduces to a value $t_2'$ starting in store $s$, it
-should also reduce to a value $v_2$ in the same number of steps
-starting in store $s_1$, since $s_1$ extends $s$. So let $\reductionl
-{t_2} {s_1} {v_2} {s_{12}} {j_2}$ with $v_2 \in \relv {k-j_2-1}
-{\Gamma_{12}} {s_{12}} {T_2}$.
-
-By the above and operational semantics, $\reductionl {t_1'.m_i(v_2)}
-{s_{12}} {\subst {v_2} {x_i} {t_i}} {s_{12}} 1$.
-
-By the substitution lemma, ${\subst {v_2} {x_i} {t_i}} \in \rele
-{k-\max(j_1,j_2)-1} {\Gamma_{12}} {s_{12}} {T}$. Supposing,
-$\reductionl {\subst {v_2} {x_i} {t_i}} {s_{12}} {t'} {s'} {j_3}$,
-with $j_1 + j_2 + j_3 + 1 = j$, this completes the case, by
-monotonicity of $\mathcal{V}$.
-
-\ \\
-
-\Case \textsc{new}: $\Gamma \ts \new y c {t_b} : T$ knowing ...
-
-By operational semantics, $\reductionl {\new y c {t_b}} s {t_b} {s_b} 1$
-where $s_b = s \envplus{y \mapsto c}$. So $\reductionl {t_b} {s_b} {t'}
-{s'} {j-1}$.
-
-By induction hypotheses, $y \in \relv k {\Gamma_b} {s_b} {T_c}$ and $t_b \in
-\rele k {\Gamma_b} {s_b} {T}$.
-
-Result follows by monotonicity of $\mathcal{V}$.
-
-\end{myproof}
-
-\subsubsection{Substitution Lemma}
-
-The substitution lemma states that if (1) $v \in \relv {k_2}
-{\Gamma_{12}} {s_{12}} {T_2}$ and (2) $t \in \rele {k_1} {\Gamma_1
-  \envplus{{x} : S}} {s_1} T$ and (3) $\Gamma \ts {T_2} \sub S$ with
-(4) $x \not\in \fn(T)$ and $\Gamma_1$ extends $\Gamma$ and
-$\Gamma_{12}$ extends $\Gamma_1$ and $s_{12}$ extends $s_1$ and
-$\Gamma_1$ agrees with $s_1$ and $\Gamma_{12}$ agrees with $s_{12}$
-and a prefix of $\Gamma_{12}$ agrees with $s_1$, then $\subst {v} {x}
-{t} \in \rele {\min(k_1,k_2)} {\Gamma_{12}} {s_{12}} T$.
-
-\begin{proofsketch}
-By (1) and (3), it should hold that (5) $v \in \relv {k_2}
-{\Gamma_{12}} {s_{12}} {S}$ by the subset semantics lemma.  Since (2)
-holds, it should also hold that ${t} \in \rele {\min(k_1,k_2)}
-{\Gamma_{12} \envplus{{x} : S}} {s_{12}} T$ by the extended
-monotonicity lemma. Then, we can instantiate $x$ in the complete store
-to map to what $v$ maps to. This should be fine by (5) and
-monotonicity. Thus, ${t} \in \rele {\min(k_1,k_2)} {\Gamma_{12}
-  \envplus{{x} : S}} {s_{12} \envplus{{x} \mapsto s_{12}(v)}}
-T$. Thanks to (4), we don't actually need $x$ to be held abstract in
-the environment, because it won't occur in $T$ or its expansion (a
-{\it potential pitfall} is whether its occurrences in $t_i$ could
-still cause a check to fail through narrowing issues), so we can use
-the type of $v$ in the environment instead of $S$ for $x$: ${t} \in
-\rele {\min(k_1,k_2)} {\Gamma_{12} \envplus{{x} : {\Gamma_{12}(v)}}}
-      {s_{12} \envplus{{x} \mapsto s_{12}(v)}} T$. This implies what
-      needs to be shown.
-\end{proofsketch}
-
-\subsubsection{Subset Semantics Lemma}
-
-The subset semantics lemma states that if $v \in \relv k \Gamma s S$
-and $\Gamma \ts S <: U$, then $v \in \relv k \Gamma s U$.
-
-\begin{proofsketch}
-Because $S$ is a subtype of $U$, it should hold that the expansion of
-$S$ subsumes the expansion of $U$, when the ``self'' occurrences are
-of type $S$. Therefore, for $v \in \relv k \Gamma s U$, we have fewer
-declarations to check than for $v \in \relv k \Gamma s S$.
-
-A {\it potential pitfall} is whether some types of the expansion of
-$U$ can become non-expanding when the ``self'' occurrences are
-actually $v$ instead of just abstractly of type $S$, causing a check
-to fail. Another worry is that such a non-expanding type results from
-narrowing of a parameter type.
-%However, we believe this is fine by
-%subtyping regularity wtr to $\nswfe$ and $v \in \relv k \Gamma s S$
-%which cheks bounds and $\nswfe$ of the type labels.
-\end{proofsketch}
-
-\subsubsection{Extended Monotonicity Lemma}
-
-The extended monotonicity lemma states that if $t \in \rele k {\Gamma
-  \envplus{x : S}} s T$ then $t \in \rele j {\Gamma' \envplus{x : S}}
-{s'} T$ for $j \leq k$, $\Gamma'$ extends $\Gamma$, $s'$
-extends $s$, and $\Gamma$ agrees with $s$ and a prefix of $\Gamma'$
-agrees with $s$.
-
-\begin{proofsketch}
-For the monotonicity with regards to the step index, this follows
-directly from the definitions of $\mathcal{E}$ and $\mathcal{V}$. For
-the environment and the store, this follows by design from the
-definition of $\rels k {\Gamma \envplus{x : S}} s$. To extend the
-environment and the store for $x : S$, we can append as much as we
-want to $\Gamma$ and $s$, to get $\Gamma'$ and $s'$, and then ignore
-the last element which is for $x : S$.
-\end{proofsketch}
-
-\section{Discussion}\label{discussion}
-
-\subsection{Why No Inheritance?}\label{why-no-inheritance}
-
-In the calculus we made the deliberate choice not to model any form of
-inheritance. This is, first and foremost, to keep the calculus simple.
-Secondly, there are many different approaches to inheritance and
-mixin composition, so that it looks advantageous not to tie the basic
-calculus to a specific one. Finally, it seems that the modelization of
-inheritance lends itself to a different approach than the basic
-calculus. For the latter, we need to prove type safety of the calculus.
-One might try to do this also for a calculus with inheritance, but our
-experience suggests that this complicates the proofs considerably.  An
-alternative approach that might work better is to model inheritance
-as a form of code-reuse. Starting with an enriched type system with
-inheritance, and a translation to the basic calculus, one needs to
-show type safety wrt the translation. This might be easier than
-to prove type safety wrt reduction.
-
 \subsection{Variants of the DOT Calculus}\label{dot-variants}

### 2012-09-13 doc/fool/proof.tex
separate out logical relation proof

--- 2012-08-11 article.tex
@@
-\documentclass[9pt]{sigplanconf}
-
-% The following \documentclass options may be useful:
-%
-% 10pt          To set in 10-point type instead of 9-point.
-% 11pt          To set in 11-point type instead of 9-point.
-% authoryear    To obtain author/year citation style instead of numeric.
-
-\usepackage{amsmath}
-\usepackage{amssymb}
-
-\usepackage{fleqn}
-\usepackage{listings}
-\usepackage{math}
-\usepackage{amsmath}
-\usepackage{latexsym}
-\usepackage{bcprules}
-\usepackage[scaled=0.848971]{luximono} % This is for 11 pt Default font
-\usepackage[T1]{fontenc}
-
-% Prooftree formatting
-\usepackage{prooftree}
-
-\usepackage{multicol}
-\usepackage{framed}
-
-\usepackage{stmaryrd}
-
-%\usepackage{float}
-%\floatstyle{boxed} 
-%\restylefloat{figure}
-
-% support for generating PDF files
-%\newif\ifpdf
-%    \ifx\pdfoutput\undefined
-%    \pdffalse
-%\else
-%    \pdftrue
-%    \pdfoutput=1
-%\fi
-
-%versions
-% Use dependent function types
-\newif\ifdep\depfalse
-
-\lstset{
-  literate=
-  {=>}{$\Rightarrow\;$}{2}
-  {<:}{$<:\;$}{1}
-}
-
-\lstdefinelanguage{scala}{% 
-       morekeywords={% 
-                try, catch, throw, private, public, protected, import, package, implicit, final, package, trait, type, class, val, def, var, if, for, this, else, extends, with, while, new, abstract, object, case, match, sealed,override},
-         sensitive=t, % 
-   morecomment=[s]{/*}{*/},morecomment=[l]{\//},% 
-   mathescape,
-%   escapeinside={/*\%}{*/},%
-   rangeprefix= /*< ,rangesuffix= >*/,%
-   morestring=[d]{"}% 
- }
-
-\lstset{breaklines=true,language=scala}
-
-\def\code{\lstinline}  % shorter version so you can write \code|String[Foo]|
-                       % -- \def must be in same file as uses for this to
-                       % work...
-\newcommand{\lstref}[1]{Listing~\ref{#1}}
-\newcommand{\Lstref}[1]{Listing~\ref{#1}} % only capitalise at beginning of sentence?
-\newcommand{\secref}[1]{Section~\ref{#1}}
-\newcommand{\Secref}[1]{Section~\ref{#1}} % only capitalise at beginning of sentence?
-
-% \lstset{basicstyle=\footnotesize\ttfamily, breaklines=true, language=scala, tabsize=2, columns=fixed, mathescape=false,includerangemarker=false}
-% thank you, Burak 
-% (lstset tweaking stolen from
-% http://lampsvn.epfl.ch/svn-repos/scala/scala/branches/typestate/docs/tstate-report/datasway.tex)
-\lstset{
-    xleftmargin=2em,%
-    framesep=5pt,%
-    frame=none,%
-    captionpos=b,%
-    fontadjust=true,%
-    columns=[c]fixed,%
-    keepspaces=false,%
-    basewidth={0.56em, 0.52em},%
-    tabsize=2,%
-    basicstyle=\small\tt,% \small\tt
-    commentstyle=\textit,%
-    keywordstyle=\bfseries,%
-    escapechar=\%,%
-}
-
-%% set latex/pdflatex specific stuff
-%\ifpdf
-    \usepackage[pdftex,
-                hyperindex,
-                plainpages=false,
-                breaklinks,
-                colorlinks,
-                citecolor=black,
-                filecolor=black,
-                linkcolor=black,
-                pagecolor=black,
-                urlcolor=black]{hyperref}
-    \usepackage[pdftex]{graphicx}
-    \DeclareGraphicsExtensions{.jpg,.pdf}
-    \pdfcatalog {
-        /PageMode (/UseNone)
-    }
-    \usepackage{thumbpdf}
-    \usepackage[pdftex]{color}
-%\else
-%    \usepackage[ps2pdf]{hyperref}
-%    \usepackage{graphicx}
-%    \DeclareGraphicsExtensions{.eps,.jpg}
-%    \usepackage{color}
-%\fi
-
-%\setlength{\parindent}{0pt}
-%\setlength{\parskip}{5pt}
-
-% verbfilter stuff
-\newcommand{\prog}[1]{{\sl #1}}
-\newenvironment{program}[1][10.5]
-  {\fontsize{#1}{13.6}\tt\begin{tabbing}\hspace*{0.5\parindent}\=\+\kill}
-  {\end{tabbing}\noindent}
-\newcommand{\blockcomment}[1]{{\color{grayPoint3}#1}}
-\newcommand{\linecomment}{\color{grayPoint3}}
-\newcommand{\grey}{\color{grey}}
-
-%\newenvironment{program}{\ \ \ \ \begin{minipage}{\textwidth}\renewcommand{\baselinestretch}{1.0}\sl\begin{tabbing}}{\end{tabbing}\end{minipage}}
-\newcommand{\vem}{\bfseries}
-\newcommand{\quotedstring}[1]{{#1}}
-\newcommand{\typename}[1]{{#1}}
-\newcommand{\literal}[1]{{#1}}
-\newcommand{\mi}[1]{\mathit{#1}}
-
-% comments and notes
-\newcommand{\comment}[1]{}
-%\newcommand{\note}[1]{{\bf $\clubsuit$ #1 $\spadesuit$}}
-
-% figures
-\newcommand{\figurebox}[1]
-        {\fbox{\begin{minipage}{\textwidth} #1 \medskip\end{minipage}}}
-%        {\fbox{\begin{minipage}{\textwidth}\begin{center} #1 \end{center}\medskip\end{minipage}}}
-\newcommand{\boxfig}[3]
-        {\begin{figure*}\figurebox{#3\caption{\label{fig:#1}#2}}\end{figure*}}
-\newcommand{\figref}[1]
-        {Figure~\ref{fig:#1}}
-
-% typing rules (not used here)
-\newcommand{\ttag}[1]{\mbox{\textsc{\small(#1)}}}
-\newcommand{\infer}[3]{\mbox{#1 }\ba{c} #2 \\ \hline #3 \ea}
-\newcommand{\irule}[2]{{\renewcommand{\arraystretch}{1.2}\ba{c} #1 
-                        \\ \hline #2 \ea}}
-\newlength{\trulemargin}
-\newlength{\trulewidth}
-\newlength{\srulewidth}
-\setlength{\trulemargin}{0.80cm}
-\setlength{\trulewidth}{40.0mm}
-\setlength{\srulewidth}{3.0cm}
-\newenvironment{trules}{$\vspace{0.5em}\ba{p{\trulemargin}@{~}p{\trulewidth}@{~}p{\trulemargin}}}{\ea$}
-\newenvironment{srules}{$\vspace{0.5em}\ba{p{\trulemargin}@{~}p{\srulewidth}}}{\ea$}
-\newcommand{\laxiom}[2]{\ttag{#1} & $ #2 \hfill\ }
-\newcommand{\raxiom}[2]{\hfill #2 $& \hfill \ttag{#1}}
-\newcommand{\caxiom}[2]{\ttag{#1} & $\hfill #2 \hfill $& \ }
-\newcommand{\lrule}[3]{\laxiom{#1}{\irule{#2}{#3}}}
-\newcommand{\rrule}[3]{\raxiom{#1}{\irule{#2}{#3}}}
-\newcommand{\crule}[3]{\caxiom{#1}{\irule{#2}{#3}}}
-\newcommand{\lsrule}[3]{\lsaxiom{#1}{\irule{#2}{#3}}}
-\newcommand{\rsrule}[3]{\rsaxiom{#1}{\irule{#2}{#3}}}
-\newcommand{\nl}{\end{trules}\\[0.5em] \begin{trules}}
-\newcommand{\snl}{\end{srules}\\[0.5em] \begin{srules}}
-
-% commas and semicolons
-\newcommand{\comma}{,\,}
-\newcommand{\commadots}{\comma \ldots \comma}
-\newcommand{\semi}{;\mbox{;};}
-\newcommand{\semidots}{\semi \ldots \semi}
-
-% spacing
-\newcommand{\gap}{\quad\quad}
-\newcommand{\biggap}{\quad\quad\quad}
-\newcommand{\nextline}{\\ \\}
-\newcommand{\htabwidth}{0.5cm}
-\newcommand{\tabwidth}{1cm}
-\newcommand{\htab}{\hspace{\htabwidth}}
-\newcommand{\tab}{\hspace{\tabwidth}}
-\newcommand{\linesep}{\ \hrulefill \ \smallskip}
-
-% math stuff
-\newenvironment{myproof}{{\em Proof:}}{$\Box$}
-\newenvironment{proofsketch}{{\em Proof Sketch:}}{$\Box$}
-\newcommand{\Case}{{\em Case\ }}
-
-% make ; a delimiter in math mode
-% \mathcode`\;="8000 % Makes ; active in math mode
-% {\catcode`\;=\active \gdef;{\;}}
-% \mathchardef\semicolon="003B
-
-% reserved words
-\newcommand{\mathem}{\bf}
-
-% brackets
-\newcommand{\set}[1]{\{#1\}}
-\newcommand{\sbs}[1]{\lquote #1 \rquote}
-
-% arrays
-\newcommand{\ba}{\begin{array}}
-\newcommand{\ea}{\end{array}}
-\newcommand{\bda}{\[\ba}
-\newcommand{\eda}{\ea\]}
-\newcommand{\ei}{\end{array}}
-\newcommand{\bcases}{\left\{\begin{array}{ll}}
-\newcommand{\ecases}{\end{array}\right.}
-
-% \cal ids
-\renewcommand{\AA}{{\cal A}}
-\newcommand{\BB}{{\cal B}}
-\newcommand{\CC}{{\cal C}}
-\newcommand{\DD}{{\cal D}}
-\newcommand{\EE}{{\cal E}}
-\newcommand{\FF}{{\cal F}}
-\newcommand{\GG}{{\cal G}}
-\newcommand{\HH}{{\cal H}}
-\newcommand{\II}{{\cal I}}
-\newcommand{\JJ}{{\cal J}}
-\newcommand{\KK}{{\cal K}}
-\newcommand{\LL}{{\cal L}}
-\newcommand{\MM}{{\cal M}}
-\newcommand{\NN}{{\cal N}}
-\newcommand{\OO}{{\cal O}}
-\newcommand{\PP}{{\cal P}}
-\newcommand{\QQ}{{\cal Q}}
-\newcommand{\RR}{{\cal R}}
-\newcommand{\TT}{{\cal T}}
-\newcommand{\UU}{{\cal U}}
-\newcommand{\VV}{{\cal V}}
-\newcommand{\WW}{{\cal W}}
-\newcommand{\XX}{{\cal X}}
-\newcommand{\YY}{{\cal Y}}
-\newcommand{\ZZ}{{\cal Z}}
-
-% misc symbols
-\newcommand{\dhd}{\!\!\!\!\!\rightarrow}
-\newcommand{\Dhd}{\!\!\!\!\!\Rightarrow}
-\newcommand{\ts}{\,\vdash\,}
-\newcommand{\la}{\langle}
-\newcommand{\ra}{\rangle}
-\newcommand{\eg}{{\em e.g.}}
-
-% misc identifiers
-\newcommand{\dom}{\mbox{\sl dom}}
-\newcommand{\fn}{\mbox{\sl fn}}
-\newcommand{\bn}{\mbox{\sl bn}}
-\newcommand{\sig}{\mbox{\sl sig}}
-\newcommand{\IF}{\mbox{\mathem if}}
-\newcommand{\OTHERWISE}{\mbox{\mathem otherwise}}
-\newcommand{\expand}{\prec}
-\newcommand{\weakexpand}{\prec^W}
-\newcommand{\spcomma}{~,~}
-
-%\newcommand{\inst}{\mbox{\mathem inst}}
-\newcommand{\trans}[1]{\la\!\la#1\ra\!\ra}
-\newcommand{\remark}[1]{{\bf $\clubsuit$ #1 $\spadesuit$}}
-\newcommand{\todo}[1]{\remark{to do: #1}}
-%\newcommand{\J}{\justifies}
-%\newcommand{\U}{\using}
-
-% names
-\newcommand{\Scala}{\mbox{\textsc{Scala}}}
-\newcommand{\Java}{\mbox{\textsc{Java}}}
-
-%\renewcommand\textfraction{.05}
-%\renewcommand\floatpagefraction{.9}
-%\renewcommand\topfraction{.8}
-
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-%   Language abstraction commands     %
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-
-%% Relations
-% Subtype 
-\newcommand{\sub}{<:}
-% Type assignment
-\newcommand{\typ}{:}
-% reduction
-\newcommand{\reduces}{\;\rightarrow\;}
-% well-formedness
-\newcommand{\wf}{\;\mbox{\textbf{wf}}}
-\newcommand{\wfe}{\;\mbox{\textbf{wfe}}}
-\newcommand{\nswfe}{\mbox{\textbf{wfe}}}
-
-%% Operators
-% Type selection
-\newcommand{\tsel}{\#}
-% Function type
-\newcommand{\tfun}{\rightarrow}
-\newcommand{\dfun}[3]{(#1\!:\!#2) \Rightarrow #3}
-% Conjunction
-\newcommand{\tand}{\wedge}
-% Disjunction
-\newcommand{\tor}{\vee}
-% Singleton type suffix
-\newcommand{\sing}{.\textbf{type}}
-
-%% Syntax
-% Header for typing rules
-\newcommand{\judgement}[2]{{\bf #1} \hfill #2}
-% Refinement
-\newcommand{\refine}[2]{\left\{#1 \Rightarrow #2 \right\}}
-\newcommand{\mlrefine}[2]{\{#1 \Rightarrow #2 \}}
-% Field definitions
-\newcommand{\ldefs}[1]{\left\{#1\right\}}
-\newcommand{\mlldefs}[1]{\{#1\}}
-% Member sequences
-\newcommand{\seq}[1]{\overline{#1}}
-% Lambda
-\newcommand{\dabs}[3]{(#1\!:\!#2)\Rightarrow #3}
-\newcommand{\abs}[3]{\lambda #1\!:\!#2.#3}
-% Method Application
-\newcommand{\mapp}[3]{#1.#2(#3)}
-% Substitution
-\newcommand{\subst}[3]{[#1/#2]#3}
-% Object creation
-\newcommand{\new}[3]{\textbf{val }#1 = \textbf{new }#2 ;\; #3}
-\newcommand{\mlnew}[3]{\textbf{val }#1 = \textbf{new }#2 ;\;\\&#3}
-%\renewcommand{\new}[3]{#1 \leftarrow #2 \,\textbf{in}\, #3}
-% Field declaration
-\newcommand{\Ldecl}[3]{#1 : #2..#3}%{#1 \operatorname{>:} #2 \operatorname{<:} #3}
-\newcommand{\ldecl}[2]{#1 : #2}
-\newcommand{\mdecl}[3]{#1 : #2 \tfun #3}
-% Top and Bottom
-\newcommand{\Top}{\top}%{\textbf{Top}}
-\newcommand{\Bot}{\bot}%\textbf{Bot}}
-% Environment extension
-%\newcommand{\envplus}[1]{\uplus \{ #1 \}}
-\newcommand{\envplus}[1]{, #1}
-% Reduction
-\newcommand{\reduction}[4]{#1 \operatorname{|} #2 \reduces #3 \operatorname{|} #4}
-
-% Sugar
-\newcommand{\arrow}[2]{#1\rightarrow_s#2}
-\newcommand{\fun}[4]{\textbf{fun } (#1:#2)\;#3\;#4}
-\newcommand{\app}[2]{(\textbf{app }#1\;#2)}
-\newcommand{\mlapp}[2]{(\textbf{app }#1\;\\&#2)}
-\newcommand{\cast}[2]{(\textbf{cast }#1\;#2)}
-
-\newcommand{\lindent}{\hspace{-4mm}}
-
-% Logical relations
-\newcommand{\relv}[4]{\mathcal{V}_{#1;#2;#3}\llbracket#4\rrbracket}
-\newcommand{\rele}[4]{\mathcal{E}_{#1;#2;#3}\llbracket#4\rrbracket}
-\newcommand{\rels}[3]{\mathcal{\supseteq}_{#1}\llbracket#2;#3\rrbracket}
-\newcommand{\relg}[3]{\mathcal{\supseteq^!}_{#1;#2}\llbracket#3\rrbracket}
-\newcommand{\irred}[2]{\text{irred }(#1,#2)}
-\newcommand{\andl}{\;\wedge\;}
-\newcommand{\orl}{\vee}
-\newcommand{\impliesl}{\rightarrow}
-\newcommand{\reductionl}[5]{#1 \operatorname{|} #2 \;\rightarrow^{#5}\; #3 \operatorname{|} #4}
-\newcommand{\ds}{\,\vDash\,}
-
-\begin{document}
-
-\conferenceinfo{FOOL '12}{October 22, 2012, Tucson, AZ, USA.} 
-\copyrightyear{2012} 
-\copyrightdata{[to be supplied]}
-
-\title{Dependent Object Types}
-\subtitle{Towards a foundation for Scala's type system}
-
-\authorinfo{Nada Amin \and Adriaan Moors \and Martin Odersky}
-           {EPFL}
-           {first.last@epfl.ch}
-
-\maketitle
-
-\begin{abstract}
-We propose a new type-theoretic foundation of Scala and languages like
-it: the Dependent Object Types calculus (DOT). DOT models Scala's
-path-dependent types and abstract type members, as well as its mixture
-of nominal and structural typing through the use of refinement
-types. It makes no attempt to model inheritance or mixing
-composition. The calculus does not model what's currently in Scala: it
-is more normative than descriptive.
-
-We show that DOT and its patched-up variants are not syntactically
-sound, by exhibiting counterexamples to preservation. Nevertheless, we
-sketch a proof of type-safety of the calculus via step-indexed logical
-relations.
-\end{abstract}
-
-\category{D.3.3}{Language Constructs and Features}{Abstract data types, Classes and objects, polymorphism}
-\category{D.3.1}{Formal Definitions and Theory}{Syntax, Semantics}
-\category{F.3.1}{Specifying and Verifying and Reasoning about Programs}{}
-\category{F.3.3}{Studies of Program Constructs}{Object-oriented constructs, type structure}
-\category{F.3.2}{Semantics or Programming Languages}{Operational semantics}
-
-\terms
-Languages, Theory, Verification
-
-\keywords
-calculus, objects, dependent types, step-indexed logical relations
-
-\section{Introduction}
-
-This paper presents a proposal for a new type-theoretic foundation of
-Scala and languages like it. The properties we are interested in
-modeling are Scala's path-dependent types and abstract type members,
-as well as its mixture of nominal and structural typing through the
-use of refinement types. Compared to previous approaches~\cite{nuObj,FS},
-we make no attempt to model inheritance or mixin composition. Indeed
-we will argue that such concepts are better modeled in a different
-setting.
-
-The calculus does not precisely describe what's currently in Scala. It
-is more normative than descriptive. The main point of deviation
-concerns the difference between Scala's compound type formation using
-{\bf with} and classical type intersection, as it is modeled in the
-calculus. Scala, and the previous calculi attempting to model it
-conflates the concepts of compound types (which inherit the members of
-several parent types) and mixin composition (which build classes from
-other classes and traits). At first glance, this offers an economy of
-concepts. However, it is problematic because mixin composition and
-intersection types have quite different properties. In the case of
-several inherited members with the same name, mixin composition has to
-pick one which overrides the others. It uses for that the concept of
-linearization of a trait hierarchy. Typically, given two independent
-traits $T_1$ and $T_2$ with a common method $m$, the mixin composition
-\code@$T_1$ with $T_2$@ would pick the $m$ in $T_2$, whereas the member in
-$T_1$ would be available via a super-call. All this makes sense from
-an implementation standpoint. From a typing standpoint it is more
-awkward, because it breaks commutativity and with it several
-monotonicity properties.
-
-In the present calculus, we replace Scala's compound types by
-classical intersection types, which are commutative. We also
-complement this by classical union types. Intersections and unions
-form a lattice wrt subtyping. This addresses another problematic
-feature of Scala: In Scala's current type system, least upper bounds
-and greatest lower bounds do not always exist. Here is an example:
-Given two traits
-\begin{lstlisting}
-  trait A { type T <: A }
-  trait B { type T <: B }
-\end{lstlisting}
-The greatest lower bound of \code@A@ and \code@B@ is approximated by the
-infinite sequence
-\begin{lstlisting}
-  A with B { type T <: A with B { type T <: A with B {
-    type T < ...
-  }}}
-\end{lstlisting}
-The limit of this sequence does not exist as a type in Scala.
-
-This is problematic because glbs and lubs play a central role in
-Scala's type inference. The absence of universal glbs and lubs makes
-type inference more brittle and more unpredictable.
-
-%The primary contributions of this paper are:
-%\begin{itemize}
-%\item Development of DOT -- a core calculus which models
-%  path-dependent types. We formally specify a small-step operational
-%  semantics and a static typing assignment in
-%  section~\ref{dot-calculus} and show some examples in
-%  section~\ref{dot-examples}.
-%\item Counterexamples to the standard theorem of preservation --
-%  section~\ref{dot-preservation}.
-%\item A sketch of a type-safety proof based on logical relations --
-%  section~\ref{dot-type-safety}.
-%\end{itemize}
-
-We propose DOT as a core calculus for path-dependent types. We present
-the calculus formally in section \ref{dot-calculus} and through examples in
-section~\ref{dot-examples}. Though we show that the calculus does not
-satisfy the standard theorem of preservation in
-section~\ref{dot-preservation}, we contribute a proof sketch of type
-safety using logical relations in section~\ref{dot-type-safety}. In
-section~\ref{discussion}, we discuss choices and variants of the
-calculus, as well as related work, and conclude in
-section~\ref{conclusion}.
-
-\section{The DOT Calculus}\label{dot-calculus}
-
-\boxfig{dot-one}{The DOT Calculus : Syntax, Reduction, Type / Declaration Assignment}{
-{\bf Syntax}\medskip
-\begin{center}    
-$\ba{l@{\hspace{0.2mm}}|@{\hspace{0.2mm}}l}
-\ba[t]{l@{\hspace{10mm}}l}
-x, y, z    & \lindent{\mbox{Variable}} \\
-l          & \lindent{\mbox{Value label}}\\
-m          & \lindent{\mbox{Method label}}\\[0.2em]
-v ::=      & \lindent{\mbox{Value}} \\
-\gap x     & \mbox{variable} \\[0.2em]
-t ::=      & \lindent{\mbox{Term}} \\
-\gap v     & \mbox{value} \\
-\gap \new x c t & \mbox{new instance} \\
-\gap t.l  & \mbox{field selection} \\
-\gap \mapp t m t  & \mbox{method invocation} \\[0.2em]
-p ::= & \lindent \mbox{Path} \\
-\gap x & \mbox{variable} \\
-\gap p.l & \mbox{selection} \\
-c ::= T_c \ldefs{\seq{d}} & \lindent{\mbox{Constructor}} \\[0.2em]
-d ::= & \lindent{\mbox{Initialization}}\\
-\gap l = v & \mbox{field initialization}\\
-\gap m(x) = t & \mbox{method initialization}\\
-s      ::= \seq{x \mapsto c} & \lindent\mbox{Store}
-\ea
-&
-\ba[t]{l@{\hspace{10mm}}l}
-L ::=      & \lindent{\mbox{Type label}} \\
-\gap L_c   & \mbox{class label} \\
-\gap L_a   & \mbox{abstract type label} \\[0.2em]
-S,T,U,V,W ::= & \lindent\mbox{Type}\\
-\gap p.L & \mbox{type selection} \\
-\gap T \refine z {\seq D} & \mbox{refinement} \\
-\gap T \tand T & \mbox{intersection type} \\
-\gap T \tor T & \mbox{union type} \\
-\gap \Top  & \mbox{top type} \\
-\gap \Bot  & \mbox{bottom type} \\[0.2em]
-S_c, T_c ::= & \lindent \mbox{Concrete type} \\
-\multicolumn{2}{l}{\gap p.L_c ~|~ T_c \refine z {\seq D} ~|~ T_c \wedge T_c  ~|~ \Top} \\[0.2em]
-D ::= & \lindent\mbox{Declaration} \\
-\gap \Ldecl L S U & \mbox{type declaration} \\
-\gap \ldecl l T   & \mbox{value declaration} \\
-\gap \mdecl m S U & \mbox{method declaration}\\
-\ \\
-\Gamma ::= \seq{x \typ T} & \lindent\mbox{Environment}
-\ea
-\ea$
-\end{center}
-\medskip
-
-\linesep
-
-\begin{multicols}{2}[\judgement{Reduction}{\fbox{$\reduction t s {t'} {s'}$}}]
-
-\infrule[\textsc{msel}]
-{y \mapsto T_c \ldefs{\seq{l = v'}\;\seq{m(x)=t}} \in s}
-{\reduction {\mapp y {m_i} v} s {\subst v {x_i} {t_i}} s}
-
-\infrule[\textsc{sel}]
-{y \mapsto T_c \ldefs{\seq{l = v}\;\seq{m(x)=t}} \in s}
-{\reduction {y.l_i} s {v_i} s}
-
-\infax[\textsc{new}]
-{\reduction {\new x c t} s t {s \envplus{x \mapsto c}}}
-
-\infrule[\textsc{context}]
-{\reduction t s {t'} {s'}}
-{\reduction {e[t]} s {e[t']} s'}
-\end{multicols}
-
-\hfill {\bf where} evaluation context $\gap e ::= [\,] ~|~ \mapp e m t ~|~ \mapp v m e ~|~ e.l\hspace{2cm}$
-
-\linesep
-
-\begin{multicols}{2}[\judgement{Type Assignment}{\fbox{$\Gamma \ts t \typ T$}}]
-
-\infrule[\textsc{var}]
-{x \typ T \in \Gamma}
-{\Gamma \ts x \typ T}
-
-\infrule[\textsc{msel}]
-{\Gamma \ts t \ni m \typ {S \tfun T} \\
- \Gamma \ts t' \typ T' \spcomma T' \sub S}
-{\Gamma \ts \mapp t m {t'} \typ T}
-
-\infrule[\textsc{sel}]
-{\Gamma \ts t \ni l \typ T'}
-{\Gamma \ts t.l \typ T'}
-
-\infrule[\textsc{new}]
-{
-y \notin \fn(T') \\
-\Gamma \ts T_c \wfe \spcomma T_c \expand_y \seq{\Ldecl L S U}, \seq{D} \\
-\Gamma \envplus{y: T_c} \ts \seq{S <: U} \spcomma \seq{d} \typ \seq{D} \spcomma {t'} \typ {T'}}
-{\Gamma \ts \new y {T_c \ldefs{\seq{d}}} {t'} \typ T'}
-
-\end{multicols}
-
-\linesep
-
-\begin{multicols}{2}[\judgement{Declaration Assignment}{\fbox{$\Gamma \ts d \typ D$}}]
-
-\infrule[\textsc{vdecl}]
-{\Gamma \ts v \typ {V'} \spcomma {V'} \sub V}
-{\Gamma \ts l = v \typ \ldecl l V}
-
-\infrule[\textsc{mdecl}]
-{\Gamma \ts S \wfe\\
- \Gamma \envplus{x : S} \ts t \typ {T'} \spcomma {T'} \sub T}
-{\Gamma \ts m(x) = t \typ \mdecl m S T }
-
-\end{multicols}
-
-} % END dot-one
-
-The DOT calculus is a simple system of dependent
-object-types. \figref{dot-one} gives its syntax, reduction rules,
-and type assignment rules.
-
-\subsection{Notation} We use standard notational conventions for
-sets. The notation $\seq{X}$ denotes a set of elements $X$. Given a
-such a set $\seq X$ in a typing rule, $X_i$ denotes an arbitrary
-element of $X$. 
-%The $\uplus$ operator extends a set of bindings. It is required that the added binding does not
-%introduce a variable which is already bound in the base-set.
-We use an
-abbreviation for preconditions in typing judgements. Given an
-environment $\Gamma$ and some predicates $P$ and $Q$, the condition $\Gamma \ts P \spcomma Q$
-is a shorthand for the two conditions $\Gamma \ts P$ and $\Gamma \ts Q$.
-
-\subsection{Syntax}
-
-There are four alphabets: Variable names $x$, $y$, $z$ are freely
-alpha-renamable. They occur as parameters of methods, as
-binders for objects created by \verb@new@-expressions, and as self
-references in refinements. Value labels $l$ denote fields in objects,
-which are bound to values at run-time. Similarly, method labels $m$
-denote methods in objects. Type labels $L$ denote type members of
-objects. Type labels are further separated into labels for abstract
-types $L_a$ and labels for classes $L_c$. It is assumed that in each
-program every class label $L_c$ is declared at most once.
-
-We assume that the label alphabets $l$, $m$ and $L$ are finite. This is
-not a restriction in practice, because one can include in these 
-alphabets every label occurring in a given program.
-
-The terms $t$ in DOT consist of variables $x$, $y$, $z$, field
-selections $t.l$, method invocations $t.m(t)$ and object creation
-expressions $\new y c {t'}$ where $c$ is a constructor $T_c \ldefs{\seq{l
-    = v}\;\seq{m(x) = t}}$. The latter binds a variable $y$ to a new
-instance of type $T_c$ with fields $\seq l$ initialized to values
-$\seq v$ and methods $\seq m$ initialized to methods of one parameter
-$\seq{x}$ and body $\seq{t}$.  The scope of $y$ extends through the term
-${t'}$.
-
-Two subclasses of terms are values $v$, which consist of just
-variables, and paths $p$ which consist of just variables and field
-selections.
-
-The types in DOT are denoted by letters $S$, $T$, $U$, $V$, or $W$. They consist of the following:
-\begin{itemize}
-\item[-] Type selections $p.L$, which denote the type member $L$ of path $p$.
-\item[-] Refinement types $T \refine z {\seq D}$, which refine a type $T$ by a set of declarations $D$.
-         The variable $z$ refers to the ``self''-reference of the type. Declarations can refer to
-         other declarations in the same type by selecting from $z$.
-\item[-] Type intersections $T \tand T'$, which carry the declarations of members present in either $T$ or $T'$.
-\item[-] Type unions $T \tor T'$, which carry only the declarations of members present in both $T$ and $T'$.
-\item[-] A top type $\Top$, which corresponds to an empty object.
-\item[-] A bottom type $\Bot$, which represents a non-terminating computation.
-\end{itemize}
-A subset of types $T_c$ are called {\em concrete types}. These are type selections
-$p.L_c$ of class labels,
-the top type $\Top$, intersections of concrete types, and refinements $T_c \refine z {\seq D}$ of concrete types. Only concrete types are allowed in constructors $c$.
-
-There are only three forms of declarations in DOT, which are all part
-of refinement types.  A value declaration $\ldecl l T$ introduces a
-field with type $T$.  A method declaration $\mdecl m S U$ introduces a
-method with parameter of type $S$ and result of type $U$. A type
-declaration $\Ldecl L S U$ introduces a type member $L$ with a lower
-bound type $S$ and an upper bound type $U$. There are no type aliases,
-but a type alias can be simulated by a type declaration $\Ldecl L T T$
-where the lower bound and the upper bound are the same type $T$.
-
-\boxfig{dot-decls}{The DOT Calculus : Declaration Lattice}{
-
-\bda{lcl@{\gap}l}
-
-      \dom(\seq D \tand \seq {D'}) &~=~& \dom(\seq{D}) \cup \dom(\seq{D'}) \\
-      \dom(\seq D \tor \seq {D'}) &=& \dom(\seq{D}) \cap \dom(\seq{D'}) \\[0.5em]
-      (D \tand D')(L) &=&
-        \Ldecl L {(S \tor S')} {(U \tand U')} & \mbox{if~} (\Ldecl L S U) \in \seq{D} \;\mbox{and}\; (\Ldecl L {S'} {U'}) \in \seq{D'} \\
-        &=& D(L) & \mbox{if~} L \notin \dom(\seq{D'}) \\
-        &=& D'(L) & \mbox{if~} L \notin \dom(\seq{D}) \\
-      (D \tand D')(m) &=&
-        \mdecl m {(S \tor S')} {(U \tand U')} & \mbox{if~} (\mdecl m S U) \in \seq{D} \;\mbox{and}\; (\mdecl m {S'} {U'}) \in \seq{D'} \\
-        &=& D(m) & \mbox{if~} m \notin \dom(\seq{D'}) \\
-        &=& D'(m) & \mbox{if~} m \notin \dom(\seq{D}) \\
-      (D \tand D')(l) &=&
-        \ldecl l {T \tand T'} & \mbox{if~} (\ldecl l T) \in \seq{D} \;\mbox{and}\; (\ldecl l {T'}) \in \seq{D'} \\
-        &=& D(l) & \mbox{if~} l \notin \dom(\seq{D'}) \\
-        &=& D'(l) & \mbox{if~} l \notin \dom(\seq{D}) \\[0.5em]
-      (D \tor D')(L) &=&
-        \Ldecl L {(S \tand S')} {(U \tor U')} & \mbox{if~} (\Ldecl L S U) \in \seq{D} \;\mbox{and}\; (\Ldecl L {S'} {U'}) \in \seq{D'} \\
-      (D \tor D')(m) &=&
-        \mdecl m {(S \tand S')} {(U \tor U')} & \mbox{if~} (\mdecl m S U) \in \seq{D} \;\mbox{and}\; (\mdecl m {S'} {U'}) \in \seq{D'} \\
-      (D \tor D')(l) &=&
-        \ldecl l {T \tor T'} & \mbox{if~} (\ldecl l T) \in \seq{D} \;\mbox{and}\; (\ldecl l {T'}) \in \seq{D'} 
-    \eda
-
-Sets of declarations form a lattice with the given meet $\wedge$ and
-join $\vee$, the empty set of declarations as the top element, and the
-bottom element $\seq{D_\Bot}$, Here $\seq{D_\Bot}$ is the set of
-declarations that contains for every term label $l$ the declaration
-$\ldecl l \Bot$, for every type label $L$ the declaration $\Ldecl L
-\Top \Bot$ and for every method label $m$ the declaration $\mdecl m
-\Top \Bot$.
-}
-
-Every field or type label can be declared only once in a set of
-declarations $\seq D$. A set of declarations can hence be seen as a map from
-labels to their declarations.  Meets $\wedge$ and joins $\vee$ on sets of
-declarations are defined in ~\figref{dot-decls}.
-
-\subsection{Reduction rules}
-
-Reduction rules $\reduction t s {t'} {s'}$ in DOT rewrite pairs of
-terms $t$ and stores $s$, where stores map variables to constructors.
-There are three main reduction rules: Rule (\textsc{msel}) rewrites a
-method invocation $\mapp y {m_i} v$ by retrieving the corresponding method
-definition from the store, and performing a substitution of the
-argument for the parameter in the body. Rule (\textsc{sel}) rewrites a
-field selection $x.l$ by retrieving the corresponding value from the
-store. Rule (\textsc{new}) rewrites an object creation $\new x c t$ by
-placing the binding of variable $x$ to constructor $c$ in the store
-and continuing with term $t$.  These reduction rules can be applied
-anywhere in a term where the hole $[\,]$ of an evaluation context $e$
-can be situated.
-
-\subsection{Type assignment rules}
-
-The last part of \figref{dot-one} presents rules for type
-assignment.
-
-Rules (\textsc{sel}) and (\textsc{msel}) type field selections and
-method invocations by means of an auxiliary membership relation $\ni$,
-which determines whether a given term contains a given declaration as
-one of its members. The membership relation is defined in
-\figref{dot-two} and is further explained in section~\ref{mem}.
-
-The last rule, (\textsc{new}), assigns types to object creation
-expressions. It is the most complex of DOT's typing rules.  To
-type-check an object creation $\new y {T_c \ldefs {\seq{l =
-      v}\;\seq{m(x) = t}}} t'$, one verifies first that the type $T_c$
-is well-formed (see \figref{dot-wf} for a definition of
-well-formedness).  One then determines the set of all declarations
-that this type carries, using the expansion relation $\expand$ defined
-in \figref{dot-two}.  Every type declaration $\Ldecl L S U$ in
-this set must be realizable, i.e.\ its lower bound $S$ must be a
-subtype of its upper bound $U$.  Every field declaration $\ldecl l V$
-in this set must have a corresponding initializing value of $v$ of
-type $V$.  These checks are made in an environment which is extended
-by the binding $y: T_c$. In particular this allows field values that
-recurse on ``self'' by referring to the bound variable $x$. Similarly,
-every method declaration $\mdecl m T W$ must have a corresponding
-initializing method definition $m(x) = t$. The parameter type $T$ must
-be $\nswfe$ (well-formed and expanding; see \figref{dot-wf}), and
-the body $t$ must type check to $W$ in an environment extended by the
-bindings $y : T_c$ and $x : T$.
-
-Instead of adding a separate subsumption rule, subtyping is expressed
-by preconditions in rules (\textsc{msel}) and (\textsc{new}).
-
-\boxfig{dot-two}{The DOT Calculus : Membership and Expansion}{
-    \begin{multicols}{2}[\judgement{Membership}{\fbox{$\Gamma \ts t \ni D$}}]
-
-      \infrule[\textsc{path-$\ni$}]
-      {\Gamma \ts p \typ T \spcomma T \expand_z \seq D}
-      {\Gamma \ts p \ni \subst p z {D_i}}
-
-      \infrule[\textsc{term-$\ni$}]
-      {z \not\in \fn(D_i) \andalso \Gamma \ts t \typ T \spcomma T \expand_z \seq D}
-      {\Gamma \ts t \ni D_i}
-   \end{multicols}
-
-   \linesep
-
-    \begin{multicols}{2}[\judgement{Expansion}{\fbox{$\Gamma \ts T \expand_z \seq D$}}]
-
-      \infrule[\textsc{rfn-$\expand$}]
-      {\Gamma \ts T \expand_z {\seq D'}}
-      {\Gamma \ts T \refine z {\seq D} \expand_z \seq {D'} \tand \seq D}
-
-      \infrule[\textsc{$\tand$-$\expand$}]
-      {\Gamma \ts T_1 \expand_z {\seq D_1} \spcomma T_2 \expand_z {\seq D_2}}
-      {\Gamma \ts T_1 \tand T_2 \expand_z {\seq D_1 \tand \seq D_2}}
-
-      \infax[\textsc{$\Top$-$\expand$}]
-      {\Gamma \ts \Top \expand_z \{\}}
-
-      \infrule[\textsc{tsel-$\expand$}]
-      {\Gamma \ts p \ni \Ldecl L S U \spcomma U \expand_z \seq D}
-      {\Gamma \ts p.L \expand_z \seq D}
-
-      \infrule[\textsc{$\tor$-$\expand$}]
-      {\Gamma \ts T_1 \expand_z {\seq D_1} \spcomma T_2 \expand_z {\seq D_2}}
-      {\Gamma \ts T_1 \tor T_2 \expand_z {\seq D_1 \tor \seq D_2}}
-
-      \infax[\textsc{$\Bot$-$\expand$}]
-      {\Gamma \ts \Bot \expand_z \seq{D_\Bot}}
-    \end{multicols}
-
-} % END dot-two
-
-\subsection{Membership}\label{mem}
-
-\figref{dot-two} presents typing rules for membership and expansion.
-The membership judgement $\Gamma \ts t \ni D$ states that in
-environment $\Gamma$ a term $t$ has a declaration $D$ as a member. The
-membership rules rely on expansion. There are two rules, one for paths
-(\textsc{path-$\ni$}) and one for general terms
-(\textsc{term-$\ni$}). For general terms, the ``self''-reference of
-the type must not occur in the resulting declaration $D$, since, to
-guarantee syntactic validity, we can only substitute a path for the
-``self''-reference.
-
-\subsection{Expansion}
-
-The expansion relation $\expand$ is needed to typecheck the complete
-set of declarations carried by a concrete type that is used in a
-\textbf{new}-expression. Expansion is also used by the membership
-rules and in subtyping refinements on the right (see
-\figref{dot-sub}).
-
-Rule (\textsc{rfn-$\expand$}) states that a refinement type $T
-\expand_z {\seq D}$ expands to the conjunction of the expansion $D'$
-of $T$ and the newly added declarations $D$. Rule
-(\textsc{tsel-$\expand$}) states that a type selection $p.L$ carries
-the same declarations as the upper bound $U$ of $L$ in $T$.  Rules
-($\tand$-$\expand$) and ($\tor$-$\expand$) states that expansion
-distributes through meets and joins.  Rule (\textsc{$\top$-$\expand$})
-states that the top type $\top$ expands to the empty set. Rule
-(\textsc{$\bot$-$\expand$}) states that the bottom type $\bot$ expands
-to the bottom element $\seq{D_\Bot}$ of the lattice of sets of
-declarations (recall \figref{dot-decls}).
-
-\boxfig{dot-sub}{The DOT Calculus : Subtyping and Declaration Subsumption}{
-
-    \begin{multicols}{2}[\judgement{Subtyping}{\fbox{$\Gamma \ts S \sub T$}}]
-
-%      \infrule[\textsc{trans}]
-%      {\Gamma \ts S \sub T \spcomma T \sub U}
-%      {\Gamma \ts S \sub U}
-
-      \infrule[\textsc{refl}]
-      {\Gamma \ts T \wfe}
-      {\Gamma \ts T \sub T}
-
-      \infrule[\textsc{$\sub$-rfn}]
-      {\Gamma \ts {T \refine z {\seq D}} \wfe
-       \spcomma S \sub T \spcomma S \expand_z \seq{D'} \\
-       \Gamma \envplus{z: S} \ts \seq{D' \sub D}}
-      {\Gamma \ts S \sub T \refine z {\seq D}}
-
-      \infrule[\textsc{$\sub$-tsel}]
-      {\Gamma \ts p \ni \Ldecl L S U \spcomma S \sub U \spcomma S' \sub S }
-      {\Gamma \ts S' \sub p.L}
-
-      \infrule[\textsc{$\sub$-$\tand$}]
-      {\Gamma \ts T \sub T_1 \spcomma T \sub T_2}
-      {\Gamma \ts T \sub T_1 \tand T_2}
-
-      \infrule[\textsc{$\sub$-$\tor_1$}]
-      {\Gamma \ts T_2 \wfe \spcomma T \sub T_1}
-      {\Gamma \ts T \sub T_1 \tor T_2}
-
-      \infrule[\textsc{$\sub$-$\tor_2$}]
-      {\Gamma \ts T_1 \wfe \spcomma T \sub T_2}
-      {\Gamma \ts T \sub T_1 \tor T_2}
-
-      %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-
-      \infrule[\textsc{$\sub$-$\Top$}]
-      {\Gamma \ts T \wfe}
-      {\Gamma \ts T \sub \Top}
-
-      \infrule[\textsc{$\Bot$-$\sub$}]
-      {\Gamma \ts T \wfe}
-      {\Gamma \ts \Bot \sub T}
-
-      \infrule[\textsc{rfn-$\sub$}]
-      {\Gamma \ts {T \refine z {\seq D}} \wfe
-       \spcomma T \sub T'}
-      {\Gamma \ts T \refine z {\seq D} \sub T'}
-
-      \infrule[\textsc{tsel-$\sub$}]
-      {\Gamma \ts p \ni \Ldecl L S U \spcomma S \sub U \spcomma U \sub U'}
-      {\Gamma \ts p.L \sub U'}
-
-      \infrule[\textsc{$\tor$-$\sub$}]
-      {\Gamma \ts T_1 \sub T \spcomma T_2 \sub T}
-      {\Gamma \ts T_1 \tor T_2 \sub T}
-
-      \infrule[\textsc{$\tand_1$-$\sub$}]
-      {\Gamma \ts T_2 \wfe \spcomma T_1 \sub T}
-      {\Gamma \ts T_1 \tand T_2 \sub T}
-
-      \infrule[\textsc{$\tand_2$-$\sub$}]
-      {\Gamma \ts T_1 \wfe \spcomma T_2 \sub T}
-      {\Gamma \ts T_1 \tand T_2 \sub T}
-
-      \end{multicols}
-
-    \linesep
-
-    \begin{multicols}{2}[\judgement{Declaration subsumption}{\fbox{$\Gamma \ts D \sub D'$}}]
-
-    \infrule[\textsc{tdecl-$\sub$}]
-            {\Gamma \ts S' \sub S \spcomma T \sub T'}
-            {\Gamma \ts (\Ldecl L S T) \sub (\Ldecl L {S'} {T'})}
-
-    \infrule[\textsc{vdecl-$\sub$}]
-            {\Gamma \ts T \sub T'}
-            {\Gamma \ts (\ldecl l T) \sub (\ldecl l {T'})}
-
-    \infrule[\textsc{mdecl-$\sub$}]
-            {\Gamma \ts S' \sub S \spcomma T \sub T'}
-            {\Gamma \ts (\mdecl m S T) \sub (\mdecl m {S'} {T'})}
-
-    \end{multicols}
-} % END dot-sub
-
-\subsection{Subtyping}
-
-\figref{dot-sub} defines the subtyping judgement $\Gamma \ts S
-\sub T$ which states that in environment $\Gamma$ type $S$ is a
-subtype of type $T$. Subtyping is regular wrt $\nswfe$: if
-type $S$ is a subtype of type $T$, then $S$ and $T$ are well-formed
-and expanding. Though this regularity limits our calculus to
-$\nswfe$-types, this limitation allows us to show that subtyping is
-transitive, as discussed in section~\ref{dot-preservation}.
-
-\subsection{Declaration Subsumption}
-
-The declaration subsumption judgement $\Gamma \ts D \sub D'$ in
-\figref{dot-sub} states that in environment $\Gamma$ the declaration
-$D$ subsumes the declaration $D'$. There are three rules, one for each
-kind (type, value, method) of declarations.  Rule
-(\textsc{tdecl-$\sub$}) states that a type declaration $\Ldecl L S U$
-subsumes another type declaration $\Ldecl L {S'} {U'}$ if $S'$ is a
-subtype of $S$ and $U$ is a subtype of $U'$. In other words, the set
-of types between $S$ and $U$ is contained in the set of types between
-$S'$ and $U'$.  Rule (\textsc{vdecl-$\sub$}) states that a value
-declaration $\ldecl l T$ subsumes another value declaration $\ldecl l
-T'$ if $T$ is a subtype of $T'$. Rule (\textsc{mdecl-$\sub$}) is
-similar to (\textsc{tdecl-$\sub$}), as the parameter type varies
-contravariantly and the return type covariantly.
-
-Declaration subsumption is extended to a binary relation between
-sequences of declarations: $\seq{D} <: \seq{D'}$ iff $\forall D'_i,
-\exists D_j. {D_j \sub D'_i}$.
-
-\boxfig{dot-wf}{The DOT Calculus : Well-Formedness}{
-
-       \begin{multicols}{2}[\judgement{Well-formed types}{\fbox{$\Gamma \ts T \wf$}}]
-
-      \infrule[\textsc{rfn-wf}]
-      {\Gamma \ts T \wf \\ 
-       \Gamma \envplus {z: T \refine z {\seq D}} \ts \seq {D \wf}}
-      {\Gamma \ts T \refine z {\seq D} \wf}
-
-      \infrule[\textsc{tsel-wf$_1$}]
-      {\Gamma \ts p \ni \Ldecl L S U \spcomma S \wf \spcomma U \wf}
-      {\Gamma \ts p.L \wf}
-
-      \infrule[\textsc{$\tand$-wf}]
-      {\Gamma \ts T \wf \spcomma T' \wf}
-      {\Gamma \ts T \tand T' \wf}
-
-      %%%%%%%%%%%%%%%%%%%%%%%%%%%
-
-      \infax[\textsc{$\Top$-wf}]
-      {\Gamma \ts \Top \wf}
-
-      \infax[\textsc{$\Bot$-wf}]
-      {\Gamma \ts \Bot \wf}
-
-      \infrule[\textsc{tsel-wf$_2$}]
-      {\Gamma \ts p \ni \Ldecl L \Bot U}
-      {\Gamma \ts p.L \wf}
-
-      \infrule[\textsc{$\tor$-wf}]
-      {\Gamma \ts T \wf \spcomma T' \wf}
-      {\Gamma \ts T \tor T' \wf}
-
-    \end{multicols}
-
-    \linesep
-
-    \begin{multicols}{2}[\judgement{Well-formed declarations}{\fbox{$\Gamma \ts D \wf$}}]
-      \infrule[\textsc{tdecl-wf}]
-      {\Gamma \ts S \wf \spcomma U \wf}
-      {\Gamma \ts \Ldecl L S U \wf}
-
-      \infrule[\textsc{vdecl-wf}]
-      {\Gamma \ts T \wf}
-      {\Gamma \ts \ldecl l T \wf}
-
-      \infrule[\textsc{mdecl-wf}]
-      {\Gamma \ts S \wf \spcomma U \wf}
-      {\Gamma \ts \mdecl m S U \wf}
-
-    \end{multicols}
-
-    \linesep
-
-    \begin{multicols}{2}[\judgement{Well-formed and expanding types}{\fbox{$\Gamma \ts T \wfe$}}]
-      \infrule[\textsc{wfe}]
-      {\Gamma \ts T \wf \spcomma T \expand_z \seq{D}}
-      {\Gamma \ts T \wfe}
-    \end{multicols}
-} % END dot-wf
-
-\subsection{Well-formedness}
-
-The well-formedness judgement $\Gamma \ts T \wf$ in \figref{dot-wf}
-states that in environment $\Gamma$ the type $T$ is
-well-formed.
-
-A refinement type $T \refine z {\seq D}$ is
-well-formed if the parent type $T$ is well-formed and every
-declaration in $\seq D$ is well-formed in an environment augmented by
-the binding of the self-reference $z$ to the refinement type itself
-(\textsc{rfn-wf}).
-
-A type selection $p.L$ is well-formed if $L$ is a member of $p$, and
-the lower bound of $L$ is also well-formed (\textsc{tsel-wf$_1$} and
-\textsc{tsel-wf$_2$}). The latter condition has the effect that the
-lower bound of a type $p.L$ may not refer directly or indirectly to a
-type containing $p.L$ itself --- if it would, the well-formedness
-judgement of $p.L$ would not have a finite proof. No such restriction
-exists for the upper bound of $L$ if the lower bound is $\bot$
-(\textsc{tsel-wf$_2$}). The upper bound may in fact refer back to the
-type. Hence, recursive class types and F-bounded abstract types are
-both expressible.
-
-The other forms of types in DOT are all well-formed if their
-constituent types are well-formed.
-
-Well-formedness extends straightforwardly to declarations with the
-judgement $\Gamma \ts D \wf$. All declarations are well-formed if
-their constituent types are well-formed.
-
-\section{Examples}\label{dot-examples}
-
-\subsection{Sugar: Functions}
-Like in Scala, we can encode functions as objects with a special
-method. We will freely use the following sugar in the remaining of
-this paper. Note that the variable $z$ must be fresh.
-\begin{align*}
-\arrow S T &\iff {\Top \refine z {\mdecl {\mi{apply}} S T}}\\
-\fun x S T t &\iff \new z {{\arrow S T}\ldefs{\mi{apply}(x) = t}} z\\
-\app f x &\iff \mapp f {\mi{apply}} x\\
-\cast T t&\iff \app {\fun x T T x} t
-\end{align*}
-
-In the remaining examples, for brevity, we will also use $\abs x S t$
-for $\fun x S \_ t$ where the return type can be somewhat inferred
-from the context.
-
-\subsection{Basics: Booleans, Error, \ldots}
-This program defines a $\mi{root}$ object with basic types
-($\mi{Unit}$, $\mi{Boolean}$, $\mi{Nat}$) and methods ($\mi{true}$,
-$\mi{false}$, $\mi{error}$, $\mi{zero}$). For brevity, we've omitted
-the code related to natural numbers. During the object creation, the
-method labels such as $\mi{false}$ are all initialized.
-\begin{align*}
-&\mlnew {\mi{root}} {\Top \mlrefine r {\\\displaybreak[3]
-&\ \Ldecl {\mi{Unit}} \Bot \Top\\\displaybreak[3]
-&\ \mdecl {\mi{unit}} \Top {r.\mi{Unit}}\\\displaybreak[3]
-&\ \Ldecl {\mi{Boolean}} \Bot {\Top \mlrefine z {\\\displaybreak[3]
-&\ \gap \ldecl {\mi{ifNat}} \arrow {(\arrow {r.\mi{Unit}} {r.\mi{Nat}})} {\arrow {(\arrow {r.\mi{Unit}} {r.\mi{Nat}})} {r.\mi{Nat}}}\\\displaybreak[3]
-&\gap}}\\\displaybreak[3]
-&\ \mdecl {\mi{false}} {r.\mi{Unit}} {r.\mi{Boolean}}\\\displaybreak[3]
-&\ \mdecl {\mi{true}} {r.\mi{Unit}} {r.\mi{Boolean}}\\\displaybreak[3]
-&\ \mdecl {\mi{error}} {r.\mi{Unit}} \Bot\\\displaybreak[3]
-& \ \ldots\\\displaybreak[3]
-&}\mlldefs{\\\displaybreak[3]
-&\ \mi{unit}(x) = {\new u {\mi{root}.\mi{Unit}} u}\\\displaybreak[3]
-%
-&\ \mi{false}(u) =\\\displaybreak[3]
-&\ \gap \mlnew {\mi{ff}} {\mi{root}.\mi{Boolean} \mlldefs{\\\displaybreak[3]
-&\ \gap\gap {\mi{ifNat}} = \abs t {\arrow {\mi{root}.\mi{Unit}} {\mi{root}.\mi{Nat}}} {\\&\gap\gap\gap\abs e {\arrow {\mi{root}.\mi{Unit}} {\mi{root}.\mi{Nat}}}\\&\gap\gap\gap\app e {\mi{root}.\mi{unit}}}\\\displaybreak[3]
-&\ \gap}}{
-\ \gap \mi{ff}}\\\displaybreak[3]
-%
-&\ \mi{true}(u) = \ldots\\\displaybreak[3]
-%
-&\ \mi{error}(u) = {\app {\mi{root}.\mi{error}} u}\\\displaybreak[3]
-&\ \ldots\\\displaybreak[3]
-&}}{
-\ldots
-}
-\end{align*}
-
-\subsection{Polymorphic Lists}
-Polymorphic lists can be expressed using an abstract type member for
-the element type ($\mi{Elem}$). We can instantiate a refinement of the
-list package to manipulate lists with a particular element type. We
-can choose to make the lists invariant or covariant in the element
-type by fixing the lower bound of the $\mi{Elem}$ re-declaration to
-the upper bound or to $\Bot$, respectively.
-\begin{align*}
-&\mlnew {\mi{genLists}} {\Top \mlrefine g {\Ldecl {\mi{ListPackage}} \Bot {\Top \mlrefine p {\\\displaybreak[3]
-&\ \Ldecl {\mi{Elem}} \Bot \Top\\\displaybreak[3]
-&\ \Ldecl {\mi{ListOfElement}} {g.\mi{List} \refine z {\Ldecl {\mi{Elem}} {p.\mi{Elem}} {p.\mi{Elem}}}} {\\&\gap\gap\gap\gap g.\mi{List} \refine z {\Ldecl {\mi{Elem}} {p.\mi{Elem}} {p.\mi{Elem}}}}\\\displaybreak[3]
-&\ \Ldecl {\mi{List}} \Bot {\Top \mlrefine z {\\\displaybreak[3]
-&\ \gap \mdecl {\mi{isEmpty}} {\mi{root}.\mi{Unit}} {\mi{root}.\mi{Boolean}}\\\displaybreak[3]
-&\ \gap \mdecl {\mi{head}} {\mi{root}.\mi{Unit}} {g.\mi{Elem}}\\\displaybreak[3]
-&\ \gap \mdecl {\mi{tail}} {\mi{root}.\mi{Unit}} {g.\mi{ListOfElem}}\\\displaybreak[3]
-&\gap}}\\\displaybreak[3]
-&\ \mdecl {\mi{nil}} {\mi{root}.\mi{Unit}} {g.\mi{ListOfElem}}\\\displaybreak[3]
-&\ \mdecl {\mi{cons}} {g.\mi{Elem}} {g.\mi{ListOfElem}}\\\displaybreak[3]
-&}}}}{
-\mlnew {\mi{natLists}} {{\mi{genLists}.\mi{ListPackage}} \mlrefine p {\\\displaybreak[3]
-&\gap \Ldecl {\mi{Elem}} {\mi{root}.\mi{Nat}} {\mi{root}.\mi{Nat}} }\mlldefs{\\\displaybreak[3]
-&\ {\mi{nil}}(u) = \ldots\\\displaybreak[3]
-&\ {\mi{cons}}(u) = \ldots\\\displaybreak[3]
-&}}}{
-\ldots
-}
-\end{align*}
-
-\section{Counterexamples to Preservation}\label{dot-preservation}
-
-We sketch a proof that the DOT calculus is type-safe using logical
-relations in section~\ref{dot-type-safety}. However, we first tried to
-prove the calculus type-safe using the standard theorems of
-preservation and progress~\cite{soundness,tapl}. Unfortunately, for
-the calculus as presented, and any of its variants that we devised,
-preservation doesn't hold. In this section, we review some of the most
-salient counterexamples to preservation that we found. These
-counterexamples have been checked with PLT Redex~\cite{plt_redex}.
-
-\subsection{\texorpdfstring{\textsc{term-$\ni$}}{Term-Mem} Restriction}\label{term_mem}
-
-There are two membership ($t \ni D$) rules: one for when the term $t$
-is a path, and one for an arbitrary term $t$. For paths, we can
-substitute the self-references in the declarations, but we cannot do
-so for arbitrary terms as the resulting types wouldn't be well-formed
-syntactically. Hence, the $\textsc{term-$\ni$}$ has the restriction
-that self-occurrences are not allowed. Here is a counterexample related
-to this restriction.
-
-Let $X$ be a shorthand for the type:
-\begin{align*}
-\Top & \mlrefine z {\\
-&\ \Ldecl {L_a} \Top \Top\\
-&\ \ldecl l {z.L_a}\\
-}&
-\end{align*}
-
-Let $Y$ be a shorthand for the type:
-\begin{align*}
-\Top & \mlrefine z {\\
-&\ \ldecl l \Top\\
-}&
-\end{align*}
-
-Now, consider the term
-\begin{align*}
-&\mlnew u {X \ldefs{ l = u }}{
-\app {(\abs y {\Top \tfun Y} {\app y u})} {(\abs d \Top {\cast X u})}.l
-}
-\end{align*}
-
-The term type-checks because the term $t=\app {(\abs y {\Top \tfun Y}
-  {\app y u})} {(\abs d \Top {\cast X u})}$ has type
-$Y$, so we can apply $\textsc{term-$\ni$}$ for $l$. However, the term
-$t$ eventually steps to ${\cast X u}$ which has type $X$,
-so we cannot apply $\textsc{term-$\ni$}$ for $l$ because of the
-self-reference ($z.L_a$).
-
-\subsection{Expansion Lost}\label{narrowing_exp}
-
-Expansion is not preserved by narrowing. Here, we create two type
-selections that are mutually recursive in their upper bounds after
-narrowing: $z_0.C_2$ initially expands, but after narrowing, $z_0.C_2$
-expands to what $z_0.A_2$ expands to, which expands to what $z_0.A_1$
-expands to, which expands to what $z_0.A_2$ expands to, and thus we
-have an infinite expansion. Thus, the last new expression initially
-type-checks, but after narrowing, it doesn't because the precise
-expansion needed by $\textsc{new}$ cannot be inferred.
-
-\begin{align*}
-&\mlnew {x_0} {\Top \mlrefine z { \Ldecl {A_1} \Bot {\Top \mlrefine z {\\
-&\gap\Ldecl {A_2} \Bot \Top\\
-&\gap\Ldecl {A_3} \Bot \Top\\
-&\gap\Ldecl {C_2} \Bot {z.A_2}}}}\ldefs{}}{
-\mlnew {x_1} {\Top \refine z {\Ldecl {C_1} \Bot {\Top \refine z {\Ldecl {A_1} \Bot {x_0.A_1}}}}\ldefs{}}}{
-\mlnew {x_2} {{x_1.C_1} \refine z {\Ldecl {A_1} \Bot {{x_0.A_1} \refine z {\Ldecl {A_2} \Bot {z.A_3}}}}\ldefs{}}}{
-\mlnew {x_3} {{x_1.C_1} \refine z {\Ldecl {A_1} \Bot {{x_0.A_1} \refine z {\Ldecl {A_3} \Bot {z.A_2}}}}\ldefs{}}}{
-\mlapp {\abs x {x_1.C_1} {(\abs {z_0} {x.A_1 \tand x_3.A_1} {\\&\gap\gap\gap\new z {z_0.C_2} {\app {(\abs x \Top x)} z}})}} {\ x_2}}
-\end{align*}
-
-\subsection{Well-Formedness Lost}\label{narrowing_wf}
-
-Even well-formedness is not preserved by narrowing. The trick is that
-if the lower bound of a type selection is not $\Bot$, then the
-bounds needs to be checked for well-formedness. Here, we create two
-type selections that are mutually recursive in their bounds
-after narrowing. $y.A$ is initially well-formed, but after narrowing,
-it isn't because we run into an infinite derivation trying to prove
-the well-formedness of its bounds.
-
-\begin{align*}
-&\mlnew v {\Top \refine z {\Ldecl L \Bot {\Top \refine z {\Ldecl A \Bot \Top, \Ldecl B {z.A} {z.A}}}} \ldefs{}}{
-\mlapp {(\abs x {\Top \refine z {\Ldecl L \Bot {\Top \refine z {\Ldecl A \Bot \Top, \Ldecl B \Bot \Top}}}} {\\&\gap
-\mlnew z {\Top \mlrefine z {\\&\gap\gap\gap
-\mdecl l {x.L \tand {\Top \refine z {\Ldecl A {z.B} {z.B}, \Ldecl B \Bot \Top}}} \Top}\mlldefs{\\&\gap\gap\gap
-l(y) = \fun a {y.A} \Top a}}{
-\gap \cast {\top} z
-}})}{\ v}}
-\end{align*}
-
-\subsection{Path Equality}\label{patheq}
-
-For preservation, we need to be able to relate path-dependent types
-after reduction. Here is a motivating example:
-
-\begin{align*}
-&\mlnew b {\Top \mlrefine z {&&\Ldecl X \Top \Top&\\
-&&&\ \ldecl l {z.X}&}\ldefs{l = b}}{
-\mlnew a {\Top \mlrefine z {\ldecl i {\Top \mlrefine z {&&&\\
-&&&\Ldecl X \Bot \Top\\
-&&&\ldecl l {z.X}}}&}\ldefs{i = b}}}{
-\app {(\abs x \Top x)} {\app {(\abs x {a.i.X} x)} {a.i.l}}}
-\end{align*}
-
-$a.i.l$ reduces to $b.l$. $b.l$ has type $b.X$, so we need $b.X <:
-a.i.X$. This cannot be established with the current rules: it is not
-true in general, but true here because $a.i$ reduces to $b$. Hence,
-we need to acknowledge path equality for preservation to hold.
-
-In section~\ref{pres-patch}, we discuss our failure to patch the
-calculus for preservation to hold.
-
 \section{Type Safety via Logical Relations}\label{dot-type-safety}
@@
 to prove type safety wrt reduction.
-
-\subsection{Variants of the DOT Calculus}\label{dot-variants}
-
-\subsubsection{Why limit the calculus to $\nswfe$-types?}
-
-Currently, the proof of type-safety via logical relations
-fundamentally relies on types having an expansion. However, this was
-not our original motivation for limiting the calculus to $\nswfe$-types.
-
-Originally, subtyping was not regular wrt $\nswfe$. Roughly,
-all the $\nswfe$ preconditions in subtyping were dropped. In this
-broader calculus, subtyping transitivity doesn't hold, because of the
-rule (\textsc{$\sub$-rfn}) which requires expansion of the left type.
-
-The problem is deep, as attested by this elaborate counterexample
-that is not so easily patched, and directly leads to a counterexample
-to preservation.
-
-Consider an environment where $u$ is bound to:
-\begin{align*}
-\Top & \mlrefine u {\\
-&\ \Ldecl {\mathit{Bad}} {\Bot} {u.\mathit{Bad}}\\
-&\ \Ldecl {\mathit{Good}} {\Top \refine z {\Ldecl L \Bot \Top}} {\Top \refine z {\Ldecl L \Bot \Top}}\\
-&\ \Ldecl {\mathit{Lower}} {u.\mathit{Bad} \tand u.\mathit{Good}} {u.\mathit{Good}}\\
-&\ \Ldecl {\mathit{Upper}} {u.\mathit{Good}} {u.\mathit{Bad} \tor u.\mathit{Good}}\\
-&\ \Ldecl X {u.\mathit{Lower}} {u.\mathit{Upper}}\\
-}&
-\end{align*}
-
-Now, consider the types $S$, $T$, $U$ defined in terms of $u$:
-\begin{align*}
-S &= u.\mathit{Bad} \tand u.\mathit{Good}\\
-T &= u.\mathit{Lower}\\
-U &= u.X \refine z {\Ldecl L \Bot \Top}
-\end{align*}
-We have $S \sub T$ and $T \sub U$, but we cannot derive $S \sub U$ because
-$S$ doesn't expand.
-
-Note that $u$ is realizable, since each lower bound is a subtype of
-its upper bound. So it is straightforward to turn this counterexample
-to subtyping transitivity into a counterexample to preservation:
-\begin{align*}
-&\mlnew u \ldots {
-\ \mlapp{\abs x \Top x}{
-\ \ \mlapp{\abs f {S \tfun U} f}{
-\ \ \ \mlapp{\abs f {S \tfun T} f}{
-\ \ \ \ \mlapp{\abs f {S \tfun S} f}{
-\ \ \  \ \ \abs x S x}}}}}
-\end{align*}
-
-The idea is to start with a function from $S \tfun S$ and cast it
-successively to $S \tfun T$ then $S \tfun U$. To typecheck the
-expression initially, we need to check $S \sub T$ and $T \sub
-U$. After some reduction steps, the first few casts vanish, and the
-reduced expression casts directly from $S \tfun S$ to $S \tfun U$, so
-we need to check $S \sub U$.
-
-\subsubsection{Why not include the lambda-calculus instead of methods?}
-
-Originally, the DOT calculus included the lambda-calculus, and
-explicit methods were not needed since they could be represented by a
-value label with a function type. However, the expansion of the
-function type was defined to be the empty set of declarations (like
-for $\Top$), which caused a real breach of type-safety.
-
-A concrete object could be a subtype of a function type without a
-function ever being defined. Consider:
-\begin{align*}
-&\mlnew u {\Top \refine z {\Ldecl C {\Top \tfun \Top} {\Top \tfun \Top}} \ldefs{}} {
-\mlnew f {u.C \ldefs{}} {
-\ldots
-}}
-\end{align*}
-
-Now, $f$ was a subtype of $\Top \tfun \Top$, but $\app f {(\abs x \Top
-  x)}$ was stuck (and, rightfully, didn't typecheck). But we could use
-narrowing to create a counterexample to type safety: $\app {(\abs g
-  {\Top \tfun \Top} {\app g (\abs x \Top x)})} f$.
-
-Because of this complication, we decided to drop the lambda-calculus
-from DOT, and instead introduce methods with one parameter. Like in
-Scala, functions are then just sugar for objects with a special
-method.
-
-An alternative design would have been to change the expansion of the
-function type to have a declaration for a special marker value label.
-
-\subsubsection{Why not patch the DOT calculus for preservation to hold?}\label{pres-patch}
-
-We tried! However, the resulting calculi were not elegant, and
-furthermore, we still found issues with preservation. Below, we give a
-brief summary of one failed attempt to patch the calculus for
-preservation to hold.
-
-Because many of the counterexamples to preservation are related to
-narrowing, we tried to make widening an explicit operation and change
-rules with implicit relaxations (\textsc{msel} and \textsc{new}) to be
-strict. From a typing perspective, the change was straightforward, but
-reduction became more complicated and dependent on typing because the
-type information in widenings needed to be propagated correctly.
-
-We added path equality provisions in the subtyping rules, in the same
-spirit as the Tribe calculus~\cite{tribe}.
-
-Unfortunately, these two patches interacted badly, and we were left
-with a disturbing counterexample to type safety.
-
-\begin{align*}
-&\mlnew a {\Top \refine z {\Ldecl C \Bot {\Top \refine z {\Ldecl D \Bot {z.X}, \Ldecl X \Bot \Top}}}}{
-\mlnew b {{a.C} \refine z {\Ldecl X \Bot \Bot}}}{
-\mlnew c {a.C}}{
-\mlnew d {{\cast {a.C} b}.D}}{
-\app {(\abs x \Bot {x.\mathit{foo}})} d
-}
-\end{align*}
-
-Notice that $d$ has type $\Bot$ if the cast on $b$ is ignored. This
-example didn't typecheck initially because the path-equality
-provisions only applied when objects are in the store, so the
-application was not well-typed. But if we started preservation in a store
-which had $a$, $b$, $c$ and $d$ then the application type-checked,
-because, through one of the path-equality provision, we could find that
-the type of $d$ was a subtype of $\Bot$. Now, of course, when we got to
-$d.\mathit{foo}$, reduction failed.
-
-\subsection{Related Work}
-
-In addition to Scala's previous models~\cite{nuObj,FS}, several
-calculi present some form of path-dependent types.
-
-The {\it vc} calculus~\cite{vc} models virtual classes with
-path-dependent types. {\it vc} restricts paths to start with ``this'',
-though it provides a way (``out'') to refer to the enclosing object.
-
-The Tribe calculus~\cite{tribe} builds an ownership types
-system~\cite{tribalo} on top of a core calculus which models virtual
-classes. The soundness proof for the core calculus seems to be tied to
-the ownership types system.
-
-Some ML-style module systems~\cite{homl,mixinml} have a form of
-stratified path-dependent types. Because of the stratification,
-recursion is not allowed. In MixML~\cite{mixml} like in Scala, this
-restriction is lifted.
-
-\section{Conclusion}\label{conclusion}
-
-We have presented DOT, a calculus aimed as a new foundation of Scala
-and languages like it. DOT features path-dependent types, refinement
-types, and abstract type members.
-
-Proving the DOT calculus type-safe has been an interesting
-adventure. We have shown that DOT does not satisfy preservation (also
-known as subject-reduction). However, the standard theorems of
-preservation and progress are just one way to prove type safety. We
-have sketched a plausible proof of type safety using the powerful
-method of logical relations.
-
-%\appendix
-%\section{Appendix Title}
-
-\acks
-
-We thank Amal Ahmed for many discussions and insights about applying
-logical relations to DOT. We thank Donna Malayeri and Geoffrey
-Washburn for preliminary work on DOT. We thank Tiark Rompf for helpful
-comments.
-
-\bibliographystyle{abbrvnat}
-\bibliography{dot}
-
-\end{document}

### 2012-09-13 src/redex/encoding.scala
starting to work on larger examples/encodings

@@ -1,0 +1,5 @@
+class Pet
+class Dog extends Pet
+class Cat extends Pet
+class Poodle extends Dog
+

### 2012-09-13 src/redex/encoding.scala
pets: class hierarchy

@@ -1,5 +1,7 @@
-class Pet
-class Dog extends Pet
-class Cat extends Pet
-class Poodle extends Dog
+object pets {
+  class Pet
+  class Dog extends Pet
+  class Cat extends Pet
+  class Poodle extends Dog
+}
 

### 2012-09-13 src/redex/encoding.scala
scoped polymorphic encoding

@@ -1,7 +1,42 @@
 object pets {
-  class Pet
-  class Dog extends Pet
-  class Cat extends Pet
-  class Poodle extends Dog
+  trait Pet
+  trait Dog extends Pet
+  trait Cat extends Pet
+  trait Poodle extends Dog
+  trait Dalmatian extends Dog
 }
 
+object choices {
+  trait Alt {
+    type C
+    type A <: C
+    type B <: C
+    val choose : A => B => C
+  }
+  def pickFirst[Cp,Ap<:Cp,Bp<:Cp] = new Alt {
+    type C = Cp
+    type A = Ap
+    type B = Bp
+    val choose = (a: A) => (b: B) => a
+  }
+  def pickLast[Cp,Ap<:Cp,Bp<:Cp] = new Alt {
+    type C = Cp
+    type A = Ap
+    type B = Bp
+    val choose = (a: A) => (b: B) => b
+  }
+}
+
+object Main extends App {
+  import pets._
+  import choices._
+
+  val potty = new Poodle { override def toString = "potty" }
+  val dotty = new Dalmatian { override def toString = "dotty" }
+
+  val picker = pickLast[Dog,Poodle,Dalmatian]
+  val p: picker.A = potty
+  val r: picker.B = picker.choose(potty)(dotty)
+  println(r)
+  //type error: println(picker.choose(dotty)(potty))
+}

### 2012-09-14 doc/fool/article.tex
new abs and rev intro

--- 2012-09-13 doc/fool/article.tex
@@
 it: the Dependent Object Types calculus (DOT). DOT models Scala's
-path-dependent types and abstract type members, as well as its mixture
-of nominal and structural typing through the use of refinement
-types. It makes no attempt to model inheritance or mixin
-composition. The calculus does not model what's currently in Scala: it
-is more normative than descriptive.
-
-We show that DOT and its patched-up variants are not syntactically
-sound, by exhibiting counterexamples to preservation. Nevertheless, we
-sketch a proof of type-safety of the calculus via step-indexed logical
-relations.
+path-dependent types, abstract type members and its mixture of nominal
+and structural typing through the use of refinement types.
+
+DOT does not precisely describe what's currently in Scala: it is more
+normative than descriptive, since our goal is to explore
+simplifications and unifications of concepts in the type system. In
+particular, DOT unifies Scala's constructs for type members, and
+provides classical intersection and union types, which simplify
+Scala's compound types and glb/lub computations.
+
+In this paper, we present and motivate the DOT calculus, our core
+calculus for path-dependent types, both formally and informally. We
+also discuss our work-in-progress to prove type-safety of the calculus.
 \end{abstract}
@@
 and greatest lower bounds do not always exist. Here is an example:
-given two traits
+given two traits \code{A} and \code{B}, where each declares an
+abstract upper-bounded type member \code{T},
 \begin{lstlisting}
@@
 \end{lstlisting}
-The greatest lower bound of \code@A@ and \code@B@ is approximated by the
+the greatest lower bound of \code{A} and \code{B} is approximated by the
 infinite sequence
@@
 This is problematic because glbs and lubs play a central role in
-Scala's type inference. The absence of universal glbs and lubs makes
-type inference more brittle and more unpredictable.
+Scala's type inference. For example, in to infer the type \code{if} expression such as
+\begin{lstlisting}
+  if (cond) ((a: A) => c: C) else ((b: B) => d: D)
+\end{lstlisting}
+type inference tries to compute the glb of \code{A} and \code{B} and
+the lub of \code{C} and \code{D} The absence of universal glbs and
+lubs makes type inference more brittle and more unpredictable.
 
@@
 
-We propose DOT as a core calculus for path-dependent types. We present
-the calculus formally in section \ref{dot-calculus} and through examples in
-section~\ref{dot-examples}. Though we show that the calculus does not
-satisfy the standard theorem of preservation in
+We propose DOT as a core calculus for path-dependent types. We first
+motivate the calculus informally in section~\ref{dot-motiv}. We
+present the calculus formally in section \ref{dot-calculus} and
+through examples in section~\ref{dot-examples}. Though we show that
+the calculus does not satisfy the standard theorem of preservation in
 section~\ref{dot-preservation}, we contribute a proof sketch of type
@@
 section~\ref{conclusion}.
+
+\section{Motivation}\label{dot-motiv}
+%TODO
 

### 2012-09-15 doc/fool/article.tex
simplif abstract
intro rev

@@
 We propose a new type-theoretic foundation of Scala and languages like
-it: the Dependent Object Types calculus (DOT). DOT models Scala's
+it: the Dependent Object Types (DOT) calculus. DOT models Scala's
 path-dependent types, abstract type members and its mixture of nominal
@@
 
-In this paper, we present and motivate the DOT calculus, our core
-calculus for path-dependent types, both formally and informally. We
-also discuss our work-in-progress to prove type-safety of the calculus.
+In this paper, we present the DOT calculus, both formally and
+informally. We also discuss our work-in-progress to prove type-safety
+of the calculus.
 \end{abstract}
@@
 \section{Introduction}
-
-This paper presents a proposal for a new type-theoretic foundation of
-Scala and languages like it. The properties we are interested in
-modeling are Scala's path-dependent types and abstract type members,
-as well as its mixture of nominal and structural typing through the
-use of refinement types. Compared to previous approaches~\cite{nuObj,FS},
-we make no attempt to model inheritance or mixin composition. Indeed
-we will argue that such concepts are better modeled in a different
-setting.
-
-The calculus does not precisely describe what's currently in Scala. It
+A scalable programming language is one in which the same concepts can
+describe small as well as large parts. Towards this goal, Scala
+unifies concepts from object and module systems. An essential
+ingredient of this unification is objects with type members. Given a
+stable path to an object, its type members can be accessed as types,
+called path-dependent types.
+
+This paper presents Dependent Object Types (DOT), a small object
+calculus with path-dependent types. In addition to path-dependent
+types, types in DOT are built from refinements, intersections and
+unions. A refinement extends a type by (re-)declaring members, which
+can be types, values or methods.
+
+We propose DOT as a new type-theoretic foundation of Scala and
+languages like it. The properties we are interested in modeling are
+Scala's path-dependent types and abstract type members, as well as its
+mixture of nominal and structural typing. Compared to previous
+approaches~\cite{nuObj,FS}, we make no attempt to model inheritance or
+mixin composition. Indeed we will argue that such concepts are better
+modeled in a different setting.
+
+The DOT calculus does not precisely describe what's currently in Scala. It
 is more normative than descriptive. The main point of deviation
@@
 This is problematic because glbs and lubs play a central role in
-Scala's type inference. For example, in to infer the type \code{if} expression such as
+Scala's type inference. For example, in order to infer the type of an
+\code{if} expression such as
 \begin{lstlisting}
@@
 type inference tries to compute the glb of \code{A} and \code{B} and
-the lub of \code{C} and \code{D} The absence of universal glbs and
+the lub of \code{C} and \code{D}. The absence of universal glbs and
 lubs makes type inference more brittle and more unpredictable.
 
-%The primary contributions of this paper are:
-%\begin{itemize}
-%\item Development of DOT -- a core calculus which models
-%  path-dependent types. We formally specify a small-step operational
-%  semantics and a static typing assignment in
-%  section~\ref{dot-calculus} and show some examples in
-%  section~\ref{dot-examples}.
-%\item Counterexamples to the standard theorem of preservation --
-%  section~\ref{dot-preservation}.
-%\item A sketch of a type-safety proof based on logical relations --
-%  section~\ref{dot-type-safety}.
-%\end{itemize}
-
-We propose DOT as a core calculus for path-dependent types. We first
-motivate the calculus informally in section~\ref{dot-motiv}. We
-present the calculus formally in section \ref{dot-calculus} and
-through examples in section~\ref{dot-examples}. Though we show that
-the calculus does not satisfy the standard theorem of preservation in
+Compared to Scala, DOT also simplifies type members. In DOT, a type
+member is declared by its lower and upper bounds. Furthermore, unlike
+Scala, there is no notion of uninstiated type member. Here is an
+example inspired by~\cite{vt}:
+\begin{lstlisting}
+trait Food
+trait Animal {
+  type Meal <: Food
+  def eat(meal: Meal) {}
+}
+\end{lstlisting}
+In Scala, \code{Meal} is an uninstiated type member of \code{Animal}:
+it is not possible to instantiate an \code{Animal} without providing a
+concrete binding for \code{Meal}. In DOT, \code{Meal} is a type member
+with a lower bound of $\Bot$ and an upper bound of \code{Food}. It is
+possible to instantiate an \code{Animal} without further specifying
+\code{Meal}, though it would be not possible to feed it. Now,
+we define \code{Cow} by refining \code{Animal}.
+\begin{lstlisting}
+trait Grass extends Food
+trait Cow extends Animal {
+  type Meal = Grass
+}
+\end{lstlisting}
+In Scala, the type alias \code{type Meal = Grass} defines a concrete
+binding for \code{Meal}. Now, we can instantiate \code{Cow}s and feed
+them \code{Grass}. In DOT, such a type alias is declared by giving it
+identical lower and upper bounds.
+
+DOT has a notion of concrete vs abstract type members, but it used to
+distinguish which types members are instantiable as opposed to
+instantiated. In the example above, all types introduced by
+\code{trait} would be concrete type members in DOT with lower bounds
+of $\Bot$ and upper bounds describing the \code{extends} clause:
+$\Top$ for \code{Food}, a refinement of $\Top$ with type member
+\code{Meal} and method member \code{eat} for \code{Animal},
+\code{Food} for \code{Grass} and a refinement of \code{Animal} with
+type member \code{Meal} for \code{Cow}. Concrete type members
+typically have a lower bound of $\Bot$ so that they are purely
+nominal: e.g. one needs to directly or indirectly instantiate a
+\code{Cow} to have an object of type \code{Cow}.
+
+We propose DOT as a core calculus for path-dependent types. We present
+the calculus formally in section \ref{dot-calculus} and through
+examples in section~\ref{dot-examples}. Though we show that the
+calculus does not satisfy the standard theorem of preservation in
 section~\ref{dot-preservation}, we contribute a proof sketch of type
@@
 section~\ref{conclusion}.
-
-\section{Motivation}\label{dot-motiv}
-%TODO
 
@@
 
+\subsection{Scala to DOT}\label{dot-examples-scala-sugar}
+// TODO
+
 \subsection{Sugar: Functions}

### 2012-09-16 src/redex/encoding.scala
suggestions from viktor

@@ -23,7 +23,7 @@ object choices {
     type C = Cp
     type A = Ap
     type B = Bp
-    val choose = (a: A) => (b: B) => b
+    val choose: A => B => B = a => b => b
   }
 }
 

### 2012-09-16 doc/fool/article.tex
exs

@@
 member is declared by its lower and upper bounds. Furthermore, unlike
-Scala, there is no notion of uninstiated type member. Here is an
+Scala, there is no notion of uninstantiated type member. Here is an
 example inspired by~\cite{vt}:
@@
 \end{lstlisting}
-In Scala, \code{Meal} is an uninstiated type member of \code{Animal}:
+In Scala, \code{Meal} is an uninstantiated type member of \code{Animal}:
 it is not possible to instantiate an \code{Animal} without providing a
@@
 
-\subsection{Scala to DOT}\label{dot-examples-scala-sugar}
-// TODO
-
-\subsection{Sugar: Functions}
+\subsection{Functions as Sugar}
 Like in Scala, we can encode functions as objects with a special
-method. We will freely use the following sugar in the remaining of
-this paper. Note that the variable $z$ must be fresh.
+method. Note that the variable $z$ must be fresh.
 \begin{align*}
@@
 
-In the remaining examples, for brevity, we will also use $\abs x S t$
-for $\fun x S \_ t$ where the return type can be somewhat inferred
-from the context.
-
-\subsection{Basics: Booleans, Error, \ldots}
-This program defines a $\mi{root}$ object with basic types
-($\mi{Unit}$, $\mi{Boolean}$, $\mi{Nat}$) and methods ($\mi{true}$,
-$\mi{false}$, $\mi{error}$, $\mi{zero}$). For brevity, we've omitted
-the code related to natural numbers. During the object creation, the
-method labels such as $\mi{false}$ are all initialized.
-\begin{align*}
-&\mlnew {\mi{root}} {\Top \mlrefine r {\\\displaybreak[3]
-&\ \Ldecl {\mi{Unit}} \Bot \Top\\\displaybreak[3]
-&\ \mdecl {\mi{unit}} \Top {r.\mi{Unit}}\\\displaybreak[3]
-&\ \Ldecl {\mi{Boolean}} \Bot {\Top \mlrefine z {\\\displaybreak[3]
-&\ \gap \ldecl {\mi{ifNat}} \arrow {(\arrow {r.\mi{Unit}} {r.\mi{Nat}})} {\arrow {(\arrow {r.\mi{Unit}} {r.\mi{Nat}})} {r.\mi{Nat}}}\\\displaybreak[3]
-&\gap}}\\\displaybreak[3]
-&\ \mdecl {\mi{false}} {r.\mi{Unit}} {r.\mi{Boolean}}\\\displaybreak[3]
-&\ \mdecl {\mi{true}} {r.\mi{Unit}} {r.\mi{Boolean}}\\\displaybreak[3]
-&\ \mdecl {\mi{error}} {r.\mi{Unit}} \Bot\\\displaybreak[3]
-& \ \ldots\\\displaybreak[3]
-&}\mlldefs{\\\displaybreak[3]
-&\ \mi{unit}(x) = {\new u {\mi{root}.\mi{Unit}} u}\\\displaybreak[3]
-%
-&\ \mi{false}(u) =\\\displaybreak[3]
-&\ \gap \mlnew {\mi{ff}} {\mi{root}.\mi{Boolean} \mlldefs{\\\displaybreak[3]
-&\ \gap\gap {\mi{ifNat}} = \abs t {\arrow {\mi{root}.\mi{Unit}} {\mi{root}.\mi{Nat}}} {\\&\gap\gap\gap\abs e {\arrow {\mi{root}.\mi{Unit}} {\mi{root}.\mi{Nat}}}\\&\gap\gap\gap\app e {\mi{root}.\mi{unit}}}\\\displaybreak[3]
-&\ \gap}}{
-\ \gap \mi{ff}}\\\displaybreak[3]
-%
-&\ \mi{true}(u) = \ldots\\\displaybreak[3]
-%
-&\ \mi{error}(u) = {\app {\mi{root}.\mi{error}} u}\\\displaybreak[3]
-&\ \ldots\\\displaybreak[3]
-&}}{
-\ldots
+We will freely use the following sugar in the remaining of this
+paper. We will also sometimes use $\abs x S t$ for $\fun x S \_ t$,
+omitting the return type for convenience and brevity.
+
+\subsection{Class Hierarchies}
+A class hierarchy such as
+\begin{lstlisting}
+object pets {
+  trait Pet
+  trait Cat extends Pet
+  trait Dog extends Pet
+  trait Poodle extends Dog
+  trait Dalmatian extends Dog
 }
-\end{align*}
-
-\subsection{Lists}
-Polymorphic lists can be expressed using an abstract type member for
-the element type ($\mi{Elem}$). We can instantiate a refinement of the
-list package to manipulate lists with a particular element type.
-\begin{align*}
-&\mlnew {\mi{genLists}} {\Top \mlrefine g {\Ldecl {\mi{ListPackage}} \Bot {\Top \mlrefine p {\\\displaybreak[3]
-&\ \Ldecl {\mi{Elem}} \Bot \Top\\\displaybreak[3]
-&\ \Ldecl {\mi{List}} \Bot {\Top \mlrefine z {\\\displaybreak[3]
-&\ \gap \mdecl {\mi{isEmpty}} {\mi{root}.\mi{Unit}} {\mi{root}.\mi{Boolean}}\\\displaybreak[3]
-&\ \gap \mdecl {\mi{head}} {\mi{root}.\mi{Unit}} {p.\mi{Elem}}\\\displaybreak[3]
-&\ \gap \mdecl {\mi{tail}} {\mi{root}.\mi{Unit}} {p.\mi{List}}\\\displaybreak[3]
-&\gap}}\\\displaybreak[3]
-&\ \mdecl {\mi{nil}} {\mi{root}.\mi{Unit}} {p.\mi{List}}\\\displaybreak[3]
-&\ \mdecl {\mi{cons}} {p.\mi{Elem}} {\arrow {p.\mi{List}} {p.\mi{List}}}\\\displaybreak[3]
-&}}}}{
-\mlnew {\mi{natLists}} {{\mi{genLists}.\mi{ListPackage}} \mlrefine p {\\\displaybreak[3]
-&\gap \Ldecl {\mi{Elem}} {\mi{root}.\mi{Nat}} {\mi{root}.\mi{Nat}} }\mlldefs{\\\displaybreak[3]
-&\ {\mi{nil}}(u) = \ldots\\\displaybreak[3]
-&\ {\mi{cons}}(hd) = \ldots\\\displaybreak[3]
-&}}}{
-\ldots
+\end{lstlisting}
+can be easily represented by concrete type members, setting the upper
+bounds appropriately:
+\begin{align*}
+\mlnew {\mi{pets}} {&\Top \mlrefine z {\\
+&\ \Ldecl {\mi{Pet_c}} \Bot \Top\\
+&\ \Ldecl {\mi{Cat_c}} \Bot {z.{\mi{Pet_c}}}\\
+&\ \Ldecl {\mi{Dog_c}} \Bot {z.{\mi{Pet_c}}}\\
+&\ \Ldecl {\mi{Poodle_c}} \Bot {z.{\mi{Dog_c}}}\\
+&\ \Ldecl {\mi{Dalmatian_c}} \Bot {z.{\mi{Dog_c}}}\\
+} \ldefs{}}{}
+\end{align*}
+
+The lower of bounds of $\Bot$ ensures that these concrete types are
+nominal as the \textsc{$\sub$-tsel} rule cannot be meaningfully
+applied.
+
+\subsection{Abstract Type Members}
+The \code{choices.Alt} trait takes three abstract type members:
+\code{C}, \code{A}, \code{B}. \code{A} and \code{B} are upper bounded
+by \code{C}. The intention is that the \code{choose} function takes an
+\code{A} and a \code{B} and returns one or the other.
+\begin{lstlisting}
+object choices {
+  trait Alt {
+    type C
+    type A <: C
+    type B <: C
+    val choose : A => B => C
   }
+}
+\end{lstlisting}
+In DOT, we can state more precisely the return type of $\mi{choose}$,
+thanks to union types:
+\begin{align*}
+&\new {\mi{choices}} {\Top \mlrefine z {\\
+&\gap\Ldecl {\mi{Alt_c}} {\Bot} {\Top \mlrefine a {\\
+&\gap\gap\Ldecl C \Bot \Top\\
+&\gap\gap\Ldecl A \Bot {a.C}\\
+&\gap\gap\Ldecl B \Bot {a.C}\\
+&\gap\gap\mdecl {\mi{choose}} {a.A} {\arrow {a.B} {a.A \tor a.B}}\\
+&\gap}}\\
+&}\ldefs{}}
+\end{align*}
+
+Using lower bounds of $\Bot$ for the abstract type members ensures
+they can vary covariantly. On the other hand, $\mi{choose}$ cannot be
+called meaningfully, unless $A$ and $B$ have realizable lower bounds.
+
+As expected, 
+\begin{align*}
+&{\mi{choices.Alt_c} \refine a {\Ldecl C \Bot {\mi{pets.Dog_c}}}}\\
+\sub\gap&{\mi{choices.Alt_c} \refine a {\Ldecl C \Bot {\mi{pets.Pet_c}}}}
+\end{align*}
+but
+\begin{align*}
+&{\mi{choices.Alt_c} \refine a {\Ldecl C {\mi{pets.Dog_c}} {\mi{pets.Dog_c}}}}\\
+\not\sub\gap&{\mi{choices.Alt_c} \refine a {\Ldecl C {\mi{pets.Pet_c}} {\mi{pets.Pet_c}}}}
+\end{align*}
+
+\subsection{Polymorphic Operators as Sugar}
+In Scala, we can implement a polymorphic operator \code{pickLast}
+which takes concrete types for \code{C}, \code{A} and \code{B} and
+implements a \code{choices.Alt} instance where the \code{choose}
+function picks the \code{B} element -- note the precision of the
+\code{choose} function which has been refined to return an element of
+exactly type \code{B}.
+\begin{lstlisting}
+def pickLast[Cp,Ap <: Cp,Bp <: Cp] = new Alt {
+  type C = Cp
+  type A = Ap
+  type B = Bp
+  val choose: A => B => B = a => b => b
+}
+
+val potty = new Poodle {}
+val dotty = new Dalmatian {}
+val picker = pickLast[Dog,Poodle,Dalmatian]
+val p: picker.A = potty
+val r: picker.B = picker.choose(potty)(dotty)
+\end{lstlisting}
+
+In DOT, we can implement such a polymorphic operator as sugar. Here,
+it is not convenient to just return a complex term like we did for
+functions as sugar because then invoking $\mi{choose}$ falls under the
+$\textsc{term-$\ni$}$ restriction, which doesn't allow the ``self''
+type to occur in the result of the method invocation. So the
+translation involves explicitly binding an object to the result of the
+polymorphic operator. We translate
+\begin{align*}
+\textbf{val } x^a = \textbf{pickLast}(T^C, T^A, T^B); e^a
+\end{align*}
+to
+\begin{align*}
+&\mlnew {x^a} {\mi{choices.Alt_c} \mlrefine {x^a} {\\
+&\gap\Ldecl C {T^C} {T^C}\\
+&\gap\Ldecl A {T^A} {T^A}\\
+&\gap\Ldecl B {T^B} {T^B}\\
+&\gap\mdecl {\mi{choose}} {x^a.A} {\arrow {x^a.B} {x^a.B}}\\
+&}\ldefs{\mi{choose}(a)=\fun b {x^a.B} {x^a.B} b}}
+{e^a}
+\end{align*}
+
+Now, given
+\begin{align*}
+&\mlnew p {\mi{pets.Poodle_c}}
+{\mlnew d {\mi{pets.Dalmatian_c}}
+{\textbf{val } a = \textbf{pickLast}({\mi{pets.Dog_c}}, {\mi{pets.Poodle_c}}, {\mi{pets.Dalmatian_c}});}}
+\end{align*}
+
+The type of $a$ is a subtype of ${\mi{choices.Alt_c}}$:
+\begin{align*}
+&\cast \Top {\\
+&\cast {\mi{choices.Alt_c}} {\\
+&\cast {{\mi{choices.Alt_c}} \refine a {\Ldecl C \Bot {\mi{pets.Dog_c}}}} {\\
+&a}}}
+\end{align*}
+
+The type of $p$ is a subtype of $a.A$:
+\begin{align*}
+\cast \Top {\cast {a.A} {p}}
+\end{align*}
+
+$a$ chooses a $\mi{pets.Dalmatian_c}$:
+\begin{align*}
+\cast \Top {\cast {\mi{pets.Dalmatian_c}} {\app {a.choose(p)} d}}
+\end{align*}
+
+$a$ enforces that it's first argument be a $\mi{pets.Poodle_c}$ and it's second a $\mi{pets.Dalmatian_c}$. The following does not typecheck for this reason:
+\begin{align*}
+\cast \Top {\app {a.choose(d)} p}
 \end{align*}

### 2012-09-17 doc/fool/article.tex
new abstract

@@
 path-dependent types, abstract type members and its mixture of nominal
-and structural typing through the use of refinement types.
-
-DOT does not precisely describe what's currently in Scala: it is more
-normative than descriptive, since our goal is to explore
-simplifications and unifications of concepts in the type system. In
-particular, DOT unifies Scala's constructs for type members, and
-provides classical intersection and union types, which simplify
-Scala's compound types and glb/lub computations.
+and structural typing through the use of refinement types. The core
+formalism makes no attempt to model inheritance and mixin
+composition. DOT normalizes Scala's type system by unifying the
+constructs for type members and by providing classical intersection
+and union types which simplify Scala's compound types and glb/lub
+computations.
 

### 2012-09-17 doc/fool/article.tex
rev

@@
 Scala's path-dependent types and abstract type members, as well as its
-mixture of nominal and structural typing. Compared to previous
-approaches~\cite{nuObj,FS}, we make no attempt to model inheritance or
-mixin composition. Indeed we will argue that such concepts are better
-modeled in a different setting.
+mixture of nominal and structural typing through the use of refinement
+types. Compared to previous approaches~\cite{nuObj,FS}, we make no
+attempt to model inheritance or mixin composition. Indeed we will
+argue that such concepts are better modeled in a different setting.
 
@@
 possible to instantiate an \code{Animal} without further specifying
-\code{Meal}, though it would be not possible to feed it. Now,
+\code{Meal}, though it would not be possible to feed it. Now,
 we define \code{Cow} by refining \code{Animal}.

### 2012-09-17 doc/fool/article.tex
Revert "separate out logical relation proof"
This reverts commit 7b8b484f17be31b1ef46de48aefef7e805b970a3.

@@
 
+\section{Type Safety via Logical Relations}\label{dot-type-safety}
+We sketch a proof of type-safety of the DOT calculus via step-indexed
+logical relations~\cite{ahmed04,ahmed06,step_indexed_obj}.
+
+\subsection{Type Safety}
+Type-safety states that a well-typed program doesn't get stuck. More
+formally: If $\emptyset \ts t \typ T$ and $\reductionl t \emptyset {t'} {s'} *$ then
+either $t'$ is a value or $\exists t'', s''. \reduction {t'} {s'}
+{t''} {s''}$.
+
+Our strategy is to define a logical relation $\Gamma \ds t : T$, such
+that $\Gamma \ts t : T$ implies $\Gamma \ds t : T$ implies
+type-safety.
+
+\subsection{Step-Indexed Logical Relations}
+
+In order to ensure that our logical relation is well-founded, we use a
+step index. For each step index $k$, we define the set of values and
+the set of terms that appear to belong to a given type, when taking at
+most $k$ steps. $\Gamma \ds t : T$ is then defined in terms of the
+step-indexed logical relation by requiring it to hold $\forall k$.
+
+\subsubsection{Set of Values}
+$\relv k \Gamma s T$ defines the set of values that appear to have
+type $T$ when taking at most $k$ steps. $\Gamma$ and $s$ must agree:
+$\dom(\Gamma) = \dom(s)$ (ordered) and $\forall (x : T) \in \Gamma, x
+\in \relv k \Gamma s T$. A variable $y$ belongs to $\relv 0 \Gamma s
+T$ simply by being in the store. In addition, it belongs to $\relv k
+\Gamma s T$ for $k > 0$, if it defines all type, method and value labels in
+the expansion of $T$ appropriately for $j < k$ steps.
+\begin{align*}
+&\relv k \Gamma s T = \{ y \ | y \in \dom(s) \andl (\\
+& (\Gamma \ts T \wfe \andl\\
+&\gap \forall j < k,\\
+&\gap y \mapsto T_c \ldefs{\seq{l = v}\;\seq{m(x)=t}} \in s,\\
+&\gap \Gamma \ts T \expand_y \seq{D},\\
+&\gap (\forall L_i : S \rightarrow U \in \seq{D},\\
+&\gap\gap \Gamma \ts y \ni L_i \typ {S'..U'}) \andl\\
+&\gap (\forall m_i : S \rightarrow U \in \seq{D},\\
+&\gap\gap t_i \in \rele j {\Gamma \envplus{x_i : S}} s U) \andl\\
+&\gap (\forall l_i : V \in \seq{D},\\
+&\gap\gap v_i \in \relv j \Gamma s V)) \orl\\
+& (T = T_1 \tand T_2 \andl y \in \relv k \Gamma s {T_1} \andl y \in \relv k \Gamma s {T_2}) \orl\\
+& (T = T_1 \tor T_2 \andl  (y \in \relv k \Gamma s {T_1} \orl y \in \relv k \Gamma s {T_2}))\\
+&)\}
+\end{align*}
+
+This relation captures the observation that the only ways for a term
+to get stuck is to have a field selection on an uninitialized field or
+a method invocation on an uninitialized method. However, a {\it potential
+  pitfall} is that the value itself might occur in the types $S$, $U$,
+$V$, because we substitute it for the ``self'' occurrences in the
+expansion, so the relation makes sure that the required type labels exist.
+
+\subsubsection{Set of Terms}
+$\rele k \Gamma s T$ defines the set of terms that appear to have type
+$T$ when taking at most $k$ steps. $s$ must agree with a {\it prefix}
+of $\Gamma$, so $\Gamma$ can additionally contain variables not in
+$s$. This is needed for checking methods in $\mathcal{V}$ above, and
+for relating open terms. If $k > 0$, $\mathcal{E}$ extends $\Gamma$
+and $s$ so that they agree. It then states that if it can reduce $t$
+in the extended store to an irreducible term in $j < k$ steps, then
+this term must be in a corresponding $\mathcal{V}$ set with $\Gamma$
+now extended to agree with the store resulting from the reduction
+steps.
+
+$\irred {t} {s}$ is a shorthand for $\neg\exists t', s'. \reduction t s
+{t'} {s'}$. $\mathcal{\supseteq}$ is used initially for the possibly
+shorter store to agree with the environment, and can extend both in
+many different ways. $\mathcal{\supseteq}^!$ is used finally for the
+possibly shorter environment to agree with the store, and just extends
+the environment in one straightforward way: hence, it defines
+singleton sets.
+
+\begin{align*}
+&\rele k \Gamma s T = \{ t \ |\\
+&\gap k = 0 \orl (\forall j < k,\\
+&\gap\gap \forall (\Gamma'; s') \in \rels k \Gamma s ,\\
+&\gap\gap \reductionl t {s'} {t'} {s''} j \andl\\
+&\gap\gap \irred {t'} {s''} \impliesl\\
+&\gap\gap \forall \Gamma'' \in \relg k {s''} {\Gamma'} ,\\
+&\gap\gap {t'} \in \relv {k-j-1} {\Gamma''} {s''} T)\\
+&\}
+\end{align*}
+
+\subsubsection{Extending the environment and the store}
+$\rels k \Gamma s$ for $k > 0$ defines the set of completed
+environment and stores that agree on $k-1$ steps, and that extend
+$\Gamma$ and $s$. $s$ must agree with a {\it prefix} of $\Gamma$.
+Both $\Gamma$ and $s$ are ordered maps. For $s$, $s'$ extends $s$ if
+$s$ is a prefix of $s'$. For $\Gamma$, $\Gamma'$ extends $\Gamma$ if
+we get back $\Gamma$ by keeping only the elements of $\Gamma'$ that
+belong to $\Gamma$. Furthermore, a prefix of $\Gamma'$ agrees with
+$s$.
+\begin{align*}
+&\rels k \Gamma s = \{\\
+&\ (\seq{x : T}^m, \seq{x_{ij} : T_{ij}}^{m \leq i < n; 0 \leq j < i_n}; s, \seq{x_{ij} \mapsto c_{ij}}^{m \leq i < n; 0 \leq j < i_n}) |\\
+&\ s = \seq{x \mapsto c}^m \andl \Gamma = \seq{x : T}^n \andl \\
+&\ m\leq n \andl \forall i, m \leq i < n, \forall {i_n}, j, 0 \leq j < i_n,\\
+&\ \forall T_{ij}, c_{ij}, T_{i{(i_n-1)}} = T_i, \forall n' \leq n, i_n' \leq i_n,\\
+&\ c_{ij} \in \relv {k-1} {\seq{x : T}^m, \seq{x_{ij} : T_{ij}}^{m \leq i' < n; 0 \leq j < i_n'}} {s, \seq{x_{ij} \mapsto c_{ij}}^{m \leq i < n'; 0 \leq j < i_n'}} {T_{ij}}\\
+&\}
+\end{align*}
+
+\subsubsection{Completing the environment to agree with the store}
+$\relg k s \Gamma$ defines a singleton set of a completed environment
+that agrees with a store $s$ by simply copying the constructor type
+from the store for each missing variable.
+\begin{align*}
+&\relg k s \Gamma = \{ \Gamma, \seq{x_i : T_{c_i}}^{m \leq i < n} \ |\\
+&\gap \Gamma = \seq{x : T}^m \andl s = \seq{x \mapsto c}^n\\
+&\}
+\end{align*}
+
+\subsubsection{Terms in the Logical Relation}
+
+$\Gamma \ds t : T$ is simply defined as $t \in \rele k \Gamma \emptyset T,
+\forall k$.
+
+\subsection{Statements and Proofs}
+
+\subsubsection{Fundamental Theorem}
+
+The fundamental theorem is the implication from $\Gamma \ts t : T$ to
+$\Gamma \ds t : T$. Type safety is a straightforward corollary of this
+theorem.
+
+\begin{myproof}
+The proof is on induction on the derivation of $\Gamma \ts t : T$. For
+each case, we need to show $t \in \rele k \Gamma \emptyset T, \forall
+k$. The non-trivial case is when $k > 0$ and for $(\Gamma'; s') \in
+\rels k \Gamma s$ and some $j < k$, $\reductionl {t} s {t'} {s'} j
+\andl \irred {t'} {s'}$. Then, we need to show ${t'} \in \relv {k-j-1}
+      {\Gamma''} {s'} T$ for ${\Gamma''} \in \relg \Gamma k {s'}
+      {\Gamma'}$.
+
+\ \\
+
+\Case \textsc{var}: $\Gamma \ts x : T$ knowing $(x : T) \in
+\Gamma$. $x \in \relv {k-1} {\Gamma'} s T$ follows from the definition of
+$\rels k \Gamma \emptyset$.
+
+\ \\
+
+\Case \textsc{sel}: $\Gamma \ts t_1.l_i : T$ knowing $\Gamma \ts t_1 :
+T_1$, $\Gamma \ts T_1 \expand_z \seq{D}$, $\ldecl {l_i} {V_i} \in
+\seq{D}$ and knowing either that $t_1 = p_1 \andl T = \subst p z
+    {V_i}$ or that $z \not\in \fn(V_i) \andl T = V_i$.
+
+By operational semantics and induction hypothesis, $\reductionl {t_1}
+s {t_1'} {s'} {j-1}$ and $\irred {t_1'} {s'}$ and ${t_1'} \in \relv
+{k-j+1-1} {\Gamma'} {s'} {T_1}$.
+
+By operational semantics and the above, $\reductionl {t_1'.l_i} {s'}
+{t'} {s'} 1$, and we can conclude $t' \in \relv {k-j-1} {\Gamma''} {s'}
+T$ from the clause for value labels of ${t_1'} \in \relv {k-j}
+{\Gamma''} {s'} {T_1}$.
+
+\ \\
+
+\Case \textsc{msel}: $\Gamma \ts t_1.m_i(t_2) : T$ knowing $\Gamma \ts
+t_1 : T_1$, $\Gamma \ts t_2 : T_2$, $\Gamma \ts T_1 \expand_z
+\seq{D}$, $\mdecl {m_i} {S_i} {U_i} \in \seq{D}$ and knowing either
+that $t_1 = p_1 \andl S = \subst p z {S_i} \andl T = \subst p z {U_i}$
+or that $z \not\in \fn(S_i) \andl z \not\in \fn(U_i) \andl S = S_i \andl
+T = U_i$, and knowing that $\Gamma \ts T_2 \sub S$.
+
+By operational semantics and induction hypotheses, $\reductionl {t_1}
+s {t_1'} {s_1} {j_1}$ and $\irred {t_1'} {s_1}$ and $\reductionl {t_2}
+{s} {t_2'} {s_2} {j_2}$ and $\irred {t_2'} {s_2}$ and ${t_1'} \in
+\relv {k-j_1-1} {\Gamma_1} {s_1} {T_1}$ and ${t_2'} \in \relv
+{k-j_2-1} {\Gamma_2} {s_2} {T_2}$.
+
+Because $t_2$ reduces to a value $t_2'$ starting in store $s$, it
+should also reduce to a value $v_2$ in the same number of steps
+starting in store $s_1$, since $s_1$ extends $s$. So let $\reductionl
+{t_2} {s_1} {v_2} {s_{12}} {j_2}$ with $v_2 \in \relv {k-j_2-1}
+{\Gamma_{12}} {s_{12}} {T_2}$.
+
+By the above and operational semantics, $\reductionl {t_1'.m_i(v_2)}
+{s_{12}} {\subst {v_2} {x_i} {t_i}} {s_{12}} 1$.
+
+By the substitution lemma, ${\subst {v_2} {x_i} {t_i}} \in \rele
+{k-\max(j_1,j_2)-1} {\Gamma_{12}} {s_{12}} {T}$. Supposing,
+$\reductionl {\subst {v_2} {x_i} {t_i}} {s_{12}} {t'} {s'} {j_3}$,
+with $j_1 + j_2 + j_3 + 1 = j$, this completes the case, by
+monotonicity of $\mathcal{V}$.
+
+\ \\
+
+\Case \textsc{new}: $\Gamma \ts \new y c {t_b} : T$ knowing ...
+
+By operational semantics, $\reductionl {\new y c {t_b}} s {t_b} {s_b} 1$
+where $s_b = s \envplus{y \mapsto c}$. So $\reductionl {t_b} {s_b} {t'}
+{s'} {j-1}$.
+
+By induction hypotheses, $y \in \relv k {\Gamma_b} {s_b} {T_c}$ and $t_b \in
+\rele k {\Gamma_b} {s_b} {T}$.
+
+Result follows by monotonicity of $\mathcal{V}$.
+
+\end{myproof}
+
+\subsubsection{Substitution Lemma}
+
+The substitution lemma states that if (1) $v \in \relv {k_2}
+{\Gamma_{12}} {s_{12}} {T_2}$ and (2) $t \in \rele {k_1} {\Gamma_1
+  \envplus{{x} : S}} {s_1} T$ and (3) $\Gamma \ts {T_2} \sub S$ with
+(4) $x \not\in \fn(T)$ and $\Gamma_1$ extends $\Gamma$ and
+$\Gamma_{12}$ extends $\Gamma_1$ and $s_{12}$ extends $s_1$ and
+$\Gamma_1$ agrees with $s_1$ and $\Gamma_{12}$ agrees with $s_{12}$
+and a prefix of $\Gamma_{12}$ agrees with $s_1$, then $\subst {v} {x}
+{t} \in \rele {\min(k_1,k_2)} {\Gamma_{12}} {s_{12}} T$.
+
+\begin{proofsketch}
+By (1) and (3), it should hold that (5) $v \in \relv {k_2}
+{\Gamma_{12}} {s_{12}} {S}$ by the subset semantics lemma.  Since (2)
+holds, it should also hold that ${t} \in \rele {\min(k_1,k_2)}
+{\Gamma_{12} \envplus{{x} : S}} {s_{12}} T$ by the extended
+monotonicity lemma. Then, we can instantiate $x$ in the complete store
+to map to what $v$ maps to. This should be fine by (5) and
+monotonicity. Thus, ${t} \in \rele {\min(k_1,k_2)} {\Gamma_{12}
+  \envplus{{x} : S}} {s_{12} \envplus{{x} \mapsto s_{12}(v)}}
+T$. Thanks to (4), we don't actually need $x$ to be held abstract in
+the environment, because it won't occur in $T$ or its expansion (a
+{\it potential pitfall} is whether its occurrences in $t_i$ could
+still cause a check to fail through narrowing issues), so we can use
+the type of $v$ in the environment instead of $S$ for $x$: ${t} \in
+\rele {\min(k_1,k_2)} {\Gamma_{12} \envplus{{x} : {\Gamma_{12}(v)}}}
+      {s_{12} \envplus{{x} \mapsto s_{12}(v)}} T$. This implies what
+      needs to be shown.
+\end{proofsketch}
+
+\subsubsection{Subset Semantics Lemma}
+
+The subset semantics lemma states that if $v \in \relv k \Gamma s S$
+and $\Gamma \ts S <: U$, then $v \in \relv k \Gamma s U$.
+
+\begin{proofsketch}
+Because $S$ is a subtype of $U$, it should hold that the expansion of
+$S$ subsumes the expansion of $U$, when the ``self'' occurrences are
+of type $S$. Therefore, for $v \in \relv k \Gamma s U$, we have fewer
+declarations to check than for $v \in \relv k \Gamma s S$.
+
+A {\it potential pitfall} is whether some types of the expansion of
+$U$ can become non-expanding when the ``self'' occurrences are
+actually $v$ instead of just abstractly of type $S$, causing a check
+to fail. Another worry is that such a non-expanding type results from
+narrowing of a parameter type.
+%However, we believe this is fine by
+%subtyping regularity wtr to $\nswfe$ and $v \in \relv k \Gamma s S$
+%which cheks bounds and $\nswfe$ of the type labels.
+\end{proofsketch}
+
+\subsubsection{Extended Monotonicity Lemma}
+
+The extended monotonicity lemma states that if $t \in \rele k {\Gamma
+  \envplus{x : S}} s T$ then $t \in \rele j {\Gamma' \envplus{x : S}}
+{s'} T$ for $j \leq k$, $\Gamma'$ extends $\Gamma$, $s'$
+extends $s$, and $\Gamma$ agrees with $s$ and a prefix of $\Gamma'$
+agrees with $s$.
+
+\begin{proofsketch}
+For the monotonicity with regards to the step index, this follows
+directly from the definitions of $\mathcal{E}$ and $\mathcal{V}$. For
+the environment and the store, this follows by design from the
+definition of $\rels k {\Gamma \envplus{x : S}} s$. To extend the
+environment and the store for $x : S$, we can append as much as we
+want to $\Gamma$ and $s$, to get $\Gamma'$ and $s'$, and then ignore
+the last element which is for $x : S$.
+\end{proofsketch}
+
+\section{Discussion}\label{discussion}
+
+\subsection{Why No Inheritance?}\label{why-no-inheritance}
+
+In the calculus we made the deliberate choice not to model any form of
+inheritance. This is, first and foremost, to keep the calculus simple.
+Secondly, there are many different approaches to inheritance and
+mixin composition, so that it looks advantageous not to tie the basic
+calculus to a specific one. Finally, it seems that the modelization of
+inheritance lends itself to a different approach than the basic
+calculus. For the latter, we need to prove type safety of the calculus.
+One might try to do this also for a calculus with inheritance, but our
+experience suggests that this complicates the proofs considerably.  An
+alternative approach that might work better is to model inheritance
+as a form of code-reuse. Starting with an enriched type system with
+inheritance, and a translation to the basic calculus, one needs to
+show type safety wrt the translation. This might be easier than
+to prove type safety wrt reduction.
+
 \subsection{Variants of the DOT Calculus}\label{dot-variants}

### 2012-09-17 doc/fool/article.tex
typos
wrong info

@@
 Compared to Scala, DOT also simplifies type members. In DOT, a type
-member is declared by its lower and upper bounds. Furthermore, unlike
-Scala, there is no notion of uninstantiated type member. Here is an
-example inspired by~\cite{vt}:
+member is declared by its lower and upper bounds. Here is an example
+inspired by~\cite{vt}:
 \begin{lstlisting}
@@
 \end{lstlisting}
-In Scala, \code{Meal} is an uninstantiated type member of \code{Animal}:
-it is not possible to instantiate an \code{Animal} without providing a
-concrete binding for \code{Meal}. In DOT, \code{Meal} is a type member
-with a lower bound of $\Bot$ and an upper bound of \code{Food}. It is
-possible to instantiate an \code{Animal} without further specifying
-\code{Meal}, though it would not be possible to feed it. Now,
-we define \code{Cow} by refining \code{Animal}.
+\code{Meal} is a type member with a lower bound of $\Bot$ and an upper
+bound of \code{Food}. It is possible to instantiate an \code{Animal}
+without further specifying \code{Meal}, though it would not be
+possible to feed it. Now, we define \code{Cow} by refining
+\code{Animal}.
 \begin{lstlisting}
@@
 In Scala, the type alias \code{type Meal = Grass} defines a concrete
-binding for \code{Meal}. Now, we can instantiate \code{Cow}s and feed
-them \code{Grass}. In DOT, such a type alias is declared by giving it
-identical lower and upper bounds.
-
-DOT has a notion of concrete vs abstract type members, but it used to
-distinguish which types members are instantiable as opposed to
-instantiated. In the example above, all types introduced by
-\code{trait} would be concrete type members in DOT with lower bounds
-of $\Bot$ and upper bounds describing the \code{extends} clause:
-$\Top$ for \code{Food}, a refinement of $\Top$ with type member
-\code{Meal} and method member \code{eat} for \code{Animal},
-\code{Food} for \code{Grass} and a refinement of \code{Animal} with
-type member \code{Meal} for \code{Cow}. Concrete type members
-typically have a lower bound of $\Bot$ so that they are purely
-nominal: e.g. one needs to directly or indirectly instantiate a
+binding for \code{Meal}. In DOT, such a type alias is declared by
+giving it identical lower and upper bounds. Now, we can instantiate
+\code{Cow}s and feed them \code{Grass}.
+
+DOT has a notion of concrete vs abstract type members, which is used
+to distinguish which types members are instantiable. In the example
+above, all types introduced by \code{trait} would be concrete type
+members in DOT with lower bounds of $\Bot$ and upper bounds describing
+the \code{extends} clause: $\Top$ for \code{Food}, a refinement of
+$\Top$ with type member \code{Meal} and method member \code{eat} for
+\code{Animal}, \code{Food} for \code{Grass} and a refinement of
+\code{Animal} with type member \code{Meal} for \code{Cow}. Concrete
+type members typically have a lower bound of $\Bot$ so that they are
+purely nominal: e.g. one needs to directly or indirectly instantiate a
 \code{Cow} to have an object of type \code{Cow}.
@@
 
-$a$ enforces that it's first argument be a $\mi{pets.Poodle_c}$ and it's second a $\mi{pets.Dalmatian_c}$. The following does not typecheck for this reason:
+$a$ enforces that its first argument be a $\mi{pets.Poodle_c}$ and its second a $\mi{pets.Dalmatian_c}$. The following does not typecheck for this reason:
 \begin{align*}

### 2012-09-17 doc/fool/article.tex
clarif

@@
 
-Using lower bounds of $\Bot$ for the abstract type members ensures
-they can vary covariantly. On the other hand, $\mi{choose}$ cannot be
-called meaningfully, unless $A$ and $B$ have realizable lower bounds.
-
-As expected, 
+Using lower bounds of $\Bot$ for the abstract type members means they
+can vary covariantly. However, we wouldn't want to pass in any
+${\mi{pets.Pet_c}}$ to a $\mi{choose}$ method that expects only a
+${\mi{pets.Dog_c}}$. Therefore,
 \begin{align*}
@@
 \end{align*}
+
+As expected, we cannot invoke $\mi{choose}$ meaningfully, unless $A$
+and $B$ have realizable lower bounds. For example, assuming we refine
+the types above so that $\Ldecl A C C$ and $\Ldecl B C C$, we cannot
+invoke $\mi{choose}$ when $C$ has a lower bound of $\Bot$ but only
+when it has a realizable lower bound such as $\mi{pets.Dog_c}$.
 

### 2012-09-17 doc/fool/article.tex
ack viktor
field vs value consistency

@@
 
-Every field or type label can be declared only once in a set of
+Every value, method or type label can be declared only once in a set of
 declarations $\seq D$. A set of declarations can hence be seen as a map from
@@
 An alternative design would have been to change the expansion of the
-function type to have a declaration for a special marker value label.
+function type to have a declaration for a special label that either
+prevents instantiation or requires an implementation for the
+function.
 
@@
 logical relations to DOT. We thank Donna Malayeri and Geoffrey
-Washburn for preliminary work on DOT. We thank Tiark Rompf for helpful
-comments.
+Washburn for preliminary work on DOT. We thank Tiark Rompf and Viktor
+Kuncak for helpful comments.
 

### 2012-09-17 src/redex/encoding.scala
F-bounded abstraction example, and bug fixes for it to work.

@@ -27,6 +27,30 @@ object choices {
   }
 }
 
+object metachoices {
+  trait MetaAlt extends choices.Alt {
+    type C = MetaAlt
+    type A = C
+    type B = C
+  }
+  val first = new MetaAlt {
+    val choose: C => C => C = a => b => a
+    override def toString = "<first>"
+  }
+  val last = new MetaAlt {
+    val choose: C => C => C = a => b => b
+    override def toString = "<last>"
+  }
+  val recfirst = new MetaAlt {
+    val choose: C => C => C = a => b => a.choose(a)(b)
+    override def toString = "<recfirst>"
+  }
+  val reclast = new MetaAlt {
+    val choose: C => C => C = a => b => b.choose(a)(b)
+    override def toString = "<reclast>"
+  }
+}
+
 object Main extends App {
   import pets._
   import choices._
@@ -38,5 +62,6 @@ object Main extends App {
   val p: picker.A = potty
   val r: picker.B = picker.choose(potty)(dotty)
   println(r)
+  println(metachoices.recfirst.choose(metachoices.first)(metachoices.reclast))
   //type error: println(picker.choose(dotty)(potty))
 }

### 2012-09-19 doc/fool/article.tex
expansion blurb
stub

@@
 
+The expansion judgement $\Gamma \ts T \expand_z {\seq {D}}$
+``flattens'' all the declarations of a type: it relates a type $T$ to
+a set of declarations that describe the type structurally. Expansion
+is precise and unique, though it doesn't always
+exist.
+
 The expansion relation $\expand$ is needed to typecheck the complete
@@
 \section{Examples}\label{dot-examples}
+
+\subsection{Glbs and Lubs}
+
+In DOT, the glb of two types is their intersection and their lub,
+their union. Recall the introductory example which was problematic in
+Scala:
+\begin{lstlisting}
+  trait A { type T <: A }
+  trait B { type T <: B }
+\end{lstlisting}
+
+We represent it as follows in DOT:
+\begin{align*}
+\mlnew p {&\Top \mlrefine p {\\
+&\ \Ldecl {A_c} \Bot {\Top \refine z {\Ldecl T \Bot {z.A_c}}}\\
+&\ \Ldecl {B_c} \Bot {\Top \refine z {\Ldecl T \Bot {z.B_c}}}\\
+&}\ldefs{}}{}
+\end{align*}
+
+The glb is of $p.A_c$ and $p.B_c$ is $p.A_c \tand p.B_c$ by
+definition.
+
+What is the expansion of $p.A_c \tand p.B_c$? It is the set ${\Ldecl T
+  \Bot {p.A_c \tand p.B_c}}$.
 

### 2012-09-21 doc/fool/article.tex
wip
a bit more of the expansion derivation

@@
 a set of declarations that describe the type structurally. Expansion
-is precise and unique, though it doesn't always
-exist.
+is precise and unique, though it doesn't always exist. See
+section~\ref{ex-exp} for examples.
 
@@
 
-\subsection{Glbs and Lubs}
+\subsection{Glbs and Lubs}\label{ex-exp}
 
@@
 
-We represent it as follows in DOT:
+We represent it as follows in DOT, wrapping the types in a namespace $p$:
 \begin{align*}
 \mlnew p {&\Top \mlrefine p {\\
-&\ \Ldecl {A_c} \Bot {\Top \refine z {\Ldecl T \Bot {z.A_c}}}\\
-&\ \Ldecl {B_c} \Bot {\Top \refine z {\Ldecl T \Bot {z.B_c}}}\\
+&\ \Ldecl {A_c} \Bot {\Top \refine z {\Ldecl T \Bot {p.A_c}}}\\
+&\ \Ldecl {B_c} \Bot {\Top \refine z {\Ldecl T \Bot {p.B_c}}}\\
 &}\ldefs{}}{}
@@
 
-What is the expansion of $p.A_c \tand p.B_c$? It is the set ${\Ldecl T
-  \Bot {p.A_c \tand p.B_c}}$.
+The expansion of $p.A_c \tand p.B_c$ is the set $\{\Ldecl T \Bot
+{p.A_c \tand p.B_c}\}$ by \textsc{$\tand$-$\expand$} rule:
+
+\
+
+{\tiny
+\prooftree
+     p.A_c \expand_z \{\Ldecl T \Bot {p.A_c}\}\quad
+     p.B_c \expand_z \{\Ldecl T \Bot {p.B_c}\}
+\justifies
+     p.A_c \tand p.B_c \expand_z \{\Ldecl T \Bot {p.A_c \tand p.B_c}\}
+\using
+     \textsc{$\tand$-$\expand$}
+\endprooftree
+}
+
+\
+
+In turn, $p.A_c \expand_z \{\Ldecl T \Bot {p.A_c}\}$ is derived as
+follows:
+
+\
+
+{\tiny
+\prooftree
+   \[
+   \ldots
+   \justifies
+   p \ni {\Ldecl {A_c} \Bot {\Top \refine z {\Ldecl T \Bot {p.A_c}}}}\quad
+   \using
+   \textsc{path-$\ni$}
+   \]
+   \[
+   \[
+   \justifies
+   \Top \expand_z \{\}
+   \using
+   \textsc{$\Top$-$\expand$}
+   \]
+   \justifies
+   {\Top \refine z {\Ldecl T \Bot {p.A_c}}} \expand_z \{\Ldecl T \Bot {p.A_c}\}
+   \using
+   \textsc{rfn-$\expand$}
+   \]
+\justifies
+   p.A_c \expand_z \{\Ldecl T \Bot {p.A_c}\}
+\using
+     \textsc{tsel-$\expand$}
+\endprooftree
+}
 

### 2012-09-21 doc/fool/article.tex
f-bounded quantification ex

@@
 &\gap}}\\
-&}\ldefs{}}
+&}\ldefs{}}{}
 \end{align*}
@@
 \end{align*}
+
+\subsection{F-bounded Quantification}
+
+F-bounded quantification describes an upper bound that itself contains
+the type being constrained: for example, \code{Int extends Ordered[Int]}.
+Here, we define \code{MetaAlt} to extend
+\code{choices.Alt} with \code{C} as an alias for \code{MetaAlt}.
+
+\begin{lstlisting}
+trait MetaAlt extends choices.Alt {
+  type C = MetaAlt
+  type A = C
+  type B = C
+}
+\end{lstlisting}
+
+Now, we can define some \code{MetaAlt} instances:
+\begin{lstlisting}
+val first = new MetaAlt {
+  val choose: C => C => C = a => b => a
+}
+val last = new MetaAlt {
+  val choose: C => C => C = a => b => b
+}
+val recfirst = new MetaAlt {
+  val choose: C => C => C = a => b => a.choose(a)(b)
+}
+val reclast = new MetaAlt {
+  val choose: C => C => C = a => b => b.choose(a)(b)
+}
+\end{lstlisting}
+
+The equivalent in DOT is straightforward. We wrap $\mi{MetaAlt}$ in a
+namespace, so that we can refer to it.
+\begin{align*}
+&\mlnew m {\Top \mlrefine m {\\
+&\gap\Ldecl {\mi{MetaAlt_c}} {\Bot} {{\mi{choices.Alt_c}} \mlrefine a {\\
+&\gap\gap\Ldecl C {\mi{m.MetaAlt_c}} {\mi{m.MetaAlt_c}}\\
+&\gap\gap\Ldecl A {\mi{a.C}} {\mi{a.C}}\\
+&\gap\gap\Ldecl B {\mi{a.C}} {\mi{a.C}}\\
+&\gap}}\\
+&}\ldefs{}}{}
+\end{align*}
+
+Now, we can create the equivalent of \code{first} ($f$), \code{last}
+($l$), \code{recfirst} ($\mi{rf}$) and \code{reclast} ($\mi{rl}$):
+\begin{align*}
+&\mlnew f {{\mi{m.MetaAlt_c}}\mlldefs{\\
+&\gap\gap\mi{choose}(a)=\fun b {\mi{m.MetaAlt_c}} {\mi{m.MetaAlt_c}} a}}{
+\mlnew l {{\mi{m.MetaAlt_c}}\mlldefs{\\
+&\gap\gap\mi{choose}(a)=\fun b {\mi{m.MetaAlt_c}} {\mi{m.MetaAlt_c}} b}}{
+\mlnew {\mi{rf}} {{\mi{m.MetaAlt_c}}\mlldefs{\\
+&\gap\gap\mi{choose}(a)=\fun b {\mi{m.MetaAlt_c}} {\mi{m.MetaAlt_c}} {\\
+&\gap\gap\gap \app {{\mi{a.choose}}(a)} b}}}{
+\mlnew {\mi{rf}} {{\mi{m.MetaAlt_c}}\mlldefs{\\
+&\gap\gap\mi{choose}(a)=\fun b {\mi{m.MetaAlt_c}} {\mi{m.MetaAlt_c}} {\\
+&\gap\gap\gap \app {{\mi{b.choose}}(a)} b}}}{
+}}}}
+\end{align*}
+
+Given these definitions, here is a valid expression, which evaluates to $f$:
+$\cast \Top \app {{\mi{rf.choose}}(f)} l$.
 

### 2012-09-22 doc/fool/article.tex
rev

@@
 constructs for type members and by providing classical intersection
-and union types which simplify Scala's compound types and glb/lub
-computations.
+and union types which simplify glb/lub computations.
 
@@
 
-\subsection{Glbs and Lubs}\label{ex-exp}
+\subsection{Glbs and Lubs}
 
@@
 definition.
+
+\subsection{Expansion}
 
@@
 
+Note that expansions do not always exist. %TODO
+
 \subsection{Functions as Sugar}

### 2012-09-22 doc/fool/article.tex
blurb about preservation not holding

@@
 
-Note that expansions do not always exist. %TODO
+Note that expansions do not always exist; see
+section~\ref{narrowing_exp} for illustration.
 
@@
 counterexamples have been checked with PLT Redex~\cite{plt_redex}.
+
+Most of these counterexamples are related to narrowing, the phenomenon
+that a type can become more precise after substitution: if a method
+takes a parameter $x$ of type $U$, then when it is invoked, any
+argument $v$ of type $S \sub U$ can be substituted -- this is the
+narrowing effect: it is as if the context was changed from $x : U$ to
+$x : S$. Sections~\ref{term_mem},~\ref{narrowing_exp}~and~\ref{narrowing_wf} each
+present a counterexample related to narrowing.
+
+The last counterexample, presented in section~\ref{patheq},
+illustrates the need to relate path-dependent types after
+reduction. This need for path-equality provisions in order for
+preservation to hold is well-known from other calculi such as
+Tribe~\cite{tribe} with path-dependent types and a small-step
+operational semantics.
+
+More generally, these counterexamples illustrate that preservation
+doesn't hold because a term that typechecks can step to a term that
+does not typecheck. However, these counterexamples to preservation are
+not counterexamples to type-safety: i.e. these programs don't get
+stuck -- they eventually step to a value.
 

### 2012-09-22 doc/fool/article.tex
a bit more about expansion counterex

@@
 \newcommand{\wf}{\;\mbox{\textbf{wf}}}
+\newcommand{\nswf}{\mbox{\textbf{wf}}}
 \newcommand{\wfe}{\;\mbox{\textbf{wfe}}}
@@
 
+First, let's illustrate why expansion does not always exist.
+
+Here is the simplest such example:
+\begin{align*}
+&\mlnew z {\Top \mlrefine z {\\
+&\gap \Ldecl {L} \Bot {z.L}\\
+&}\ldefs{}}{}
+\end{align*}
+The type $z.L$ is $\nswf$ but not $\nswfe$. Indeed, there is no finite
+derivation of an expansion for $z.L$, because by the
+$\textsc{tsel-$\expand$}$ rule, in order to expand $z.L$, we need to
+expand its upper bound, which is $z.L$! However, not that the object
+creation expression for $z$ would not typecheck because subtyping is
+regular wrt $\nswfe$, and so the type of a constructor in an object
+creation expression must have $\nswfe$ type members, since we check
+that the lower bound is a subtype of the upper bound for each type
+member.
+
+The example above relies on the $\textsc{tsel-wf$_2$}$, in order for
+the upper bound to refer to the type member being declared. Here is an
+example that does not rely on this rule, and that we will use below to
+create a counterexample to preservation. Let
+\begin{align*}
+T_1 = \Top &\mlrefine z {&\\
+&\gap\Ldecl A \Bot {z.B}&\\
+&\gap\Ldecl B \Bot \Top&}\\
+T_2 = \Top &\mlrefine z {&\\
+&\gap\Ldecl A \Bot \Top&\\
+&\gap\Ldecl B \Bot {z.A}&}
+\end{align*}
+
+Now, consider $T_1 \tand T_2$. The type is $\nswf$ and $\nswfe$, but
+its members $A$ and $B$ are not $\nswfe$, because the expansion of
+$T_1 \tand T_2$ with self-type $z$ has the set of declarations
+$\{\Ldecl A \Bot {z.B},\;\Ldecl B \Bot {z.A}\}$ -- thus, to expand
+$z.A$, we need to expand $z.B$, and to expand $z.B$, we need to expand
+$z.A$! There is no finite derivation of expansions for the type
+members $A$ and $B$.
+
 Expansion is not preserved by narrowing. Here, we create two type

### 2012-09-22 doc/fool/article.tex
formatting of ex
Removing proof of type safety from article.

@@
 \documentclass[9pt]{sigplanconf}
-
-% The following \documentclass options may be useful:
-%
-% 10pt          To set in 10-point type instead of 9-point.
-% 11pt          To set in 11-point type instead of 9-point.
-% authoryear    To obtain author/year citation style instead of numeric.
-
-\usepackage{amsmath}
-\usepackage{amssymb}
-
-\usepackage{fleqn}
-\usepackage{listings}
-\usepackage{math}
-\usepackage{amsmath}
-\usepackage{latexsym}
-\usepackage{bcprules}
-\usepackage[scaled=0.848971]{luximono} % This is for 11 pt Default font
-\usepackage[T1]{fontenc}
-
-% Prooftree formatting
-\usepackage{prooftree}
-
-\usepackage{multicol}
-\usepackage{framed}
-
-\usepackage{stmaryrd}
-
-%\usepackage{float}
-%\floatstyle{boxed} 
-%\restylefloat{figure}
-
-% support for generating PDF files
-%\newif\ifpdf
-%    \ifx\pdfoutput\undefined
-%    \pdffalse
-%\else
-%    \pdftrue
-%    \pdfoutput=1
-%\fi
-
-%versions
-% Use dependent function types
-\newif\ifdep\depfalse
-
-\lstset{
-  literate=
-  {=>}{$\Rightarrow\;$}{2}
-  {<:}{$<:\;$}{1}
-}
-
-\lstdefinelanguage{scala}{% 
-       morekeywords={% 
-                try, catch, throw, private, public, protected, import, package, implicit, final, package, trait, type, class, val, def, var, if, for, this, else, extends, with, while, new, abstract, object, case, match, sealed,override},
-         sensitive=t, % 
-   morecomment=[s]{/*}{*/},morecomment=[l]{\//},% 
-   mathescape,
-%   escapeinside={/*\%}{*/},%
-   rangeprefix= /*< ,rangesuffix= >*/,%
-   morestring=[d]{"}% 
- }
-
-\lstset{breaklines=true,language=scala}
-
-\def\code{\lstinline}  % shorter version so you can write \code|String[Foo]|
-                       % -- \def must be in same file as uses for this to
-                       % work...
-\newcommand{\lstref}[1]{Listing~\ref{#1}}
-\newcommand{\Lstref}[1]{Listing~\ref{#1}} % only capitalise at beginning of sentence?
-\newcommand{\secref}[1]{Section~\ref{#1}}
-\newcommand{\Secref}[1]{Section~\ref{#1}} % only capitalise at beginning of sentence?
-
-% \lstset{basicstyle=\footnotesize\ttfamily, breaklines=true, language=scala, tabsize=2, columns=fixed, mathescape=false,includerangemarker=false}
-% thank you, Burak 
-% (lstset tweaking stolen from
-% http://lampsvn.epfl.ch/svn-repos/scala/scala/branches/typestate/docs/tstate-report/datasway.tex)
-\lstset{
-    xleftmargin=2em,%
-    framesep=5pt,%
-    frame=none,%
-    captionpos=b,%
-    fontadjust=true,%
-    columns=[c]fixed,%
-    keepspaces=false,%
-    basewidth={0.56em, 0.52em},%
-    tabsize=2,%
-    basicstyle=\small\tt,% \small\tt
-    commentstyle=\textit,%
-    keywordstyle=\bfseries,%
-    escapechar=\%,%
-}
-
-%% set latex/pdflatex specific stuff
-%\ifpdf
-    \usepackage[pdftex,
-                hyperindex,
-                plainpages=false,
-                breaklinks,
-                colorlinks,
-                citecolor=black,
-                filecolor=black,
-                linkcolor=black,
-                pagecolor=black,
-                urlcolor=black]{hyperref}
-    \usepackage[pdftex]{graphicx}
-    \DeclareGraphicsExtensions{.jpg,.pdf}
-    \pdfcatalog {
-        /PageMode (/UseNone)
-    }
-    \usepackage{thumbpdf}
-    \usepackage[pdftex]{color}
-%\else
-%    \usepackage[ps2pdf]{hyperref}
-%    \usepackage{graphicx}
-%    \DeclareGraphicsExtensions{.eps,.jpg}
-%    \usepackage{color}
-%\fi
-
-%\setlength{\parindent}{0pt}
-%\setlength{\parskip}{5pt}
-
-% verbfilter stuff
-\newcommand{\prog}[1]{{\sl #1}}
-\newenvironment{program}[1][10.5]
-  {\fontsize{#1}{13.6}\tt\begin{tabbing}\hspace*{0.5\parindent}\=\+\kill}
-  {\end{tabbing}\noindent}
-\newcommand{\blockcomment}[1]{{\color{grayPoint3}#1}}
-\newcommand{\linecomment}{\color{grayPoint3}}
-\newcommand{\grey}{\color{grey}}
-
-%\newenvironment{program}{\ \ \ \ \begin{minipage}{\textwidth}\renewcommand{\baselinestretch}{1.0}\sl\begin{tabbing}}{\end{tabbing}\end{minipage}}
-\newcommand{\vem}{\bfseries}
-\newcommand{\quotedstring}[1]{{#1}}
-\newcommand{\typename}[1]{{#1}}
-\newcommand{\literal}[1]{{#1}}
-\newcommand{\mi}[1]{\mathit{#1}}
-
-% comments and notes
-\newcommand{\comment}[1]{}
-%\newcommand{\note}[1]{{\bf $\clubsuit$ #1 $\spadesuit$}}
-
-% figures
-\newcommand{\figurebox}[1]
-        {\fbox{\begin{minipage}{\textwidth} #1 \medskip\end{minipage}}}
-%        {\fbox{\begin{minipage}{\textwidth}\begin{center} #1 \end{center}\medskip\end{minipage}}}
-\newcommand{\boxfig}[3]
-        {\begin{figure*}\figurebox{#3\caption{\label{fig:#1}#2}}\end{figure*}}
-\newcommand{\figref}[1]
-        {Figure~\ref{fig:#1}}
-
-% typing rules (not used here)
-\newcommand{\ttag}[1]{\mbox{\textsc{\small(#1)}}}
-\newcommand{\infer}[3]{\mbox{#1 }\ba{c} #2 \\ \hline #3 \ea}
-\newcommand{\irule}[2]{{\renewcommand{\arraystretch}{1.2}\ba{c} #1 
-                        \\ \hline #2 \ea}}
-\newlength{\trulemargin}
-\newlength{\trulewidth}
-\newlength{\srulewidth}
-\setlength{\trulemargin}{0.80cm}
-\setlength{\trulewidth}{40.0mm}
-\setlength{\srulewidth}{3.0cm}
-\newenvironment{trules}{$\vspace{0.5em}\ba{p{\trulemargin}@{~}p{\trulewidth}@{~}p{\trulemargin}}}{\ea$}
-\newenvironment{srules}{$\vspace{0.5em}\ba{p{\trulemargin}@{~}p{\srulewidth}}}{\ea$}
-\newcommand{\laxiom}[2]{\ttag{#1} & $ #2 \hfill\ }
-\newcommand{\raxiom}[2]{\hfill #2 $& \hfill \ttag{#1}}
-\newcommand{\caxiom}[2]{\ttag{#1} & $\hfill #2 \hfill $& \ }
-\newcommand{\lrule}[3]{\laxiom{#1}{\irule{#2}{#3}}}
-\newcommand{\rrule}[3]{\raxiom{#1}{\irule{#2}{#3}}}
-\newcommand{\crule}[3]{\caxiom{#1}{\irule{#2}{#3}}}
-\newcommand{\lsrule}[3]{\lsaxiom{#1}{\irule{#2}{#3}}}
-\newcommand{\rsrule}[3]{\rsaxiom{#1}{\irule{#2}{#3}}}
-\newcommand{\nl}{\end{trules}\\[0.5em] \begin{trules}}
-\newcommand{\snl}{\end{srules}\\[0.5em] \begin{srules}}
-
-% commas and semicolons
-\newcommand{\comma}{,\,}
-\newcommand{\commadots}{\comma \ldots \comma}
-\newcommand{\semi}{;\mbox{;};}
-\newcommand{\semidots}{\semi \ldots \semi}
-
-% spacing
-\newcommand{\gap}{\quad\quad}
-\newcommand{\biggap}{\quad\quad\quad}
-\newcommand{\nextline}{\\ \\}
-\newcommand{\htabwidth}{0.5cm}
-\newcommand{\tabwidth}{1cm}
-\newcommand{\htab}{\hspace{\htabwidth}}
-\newcommand{\tab}{\hspace{\tabwidth}}
-\newcommand{\linesep}{\ \hrulefill \ \smallskip}
-
-% math stuff
-\newenvironment{myproof}{{\em Proof:}}{$\Box$}
-\newenvironment{proofsketch}{{\em Proof Sketch:}}{$\Box$}
-\newcommand{\Case}{{\em Case\ }}
-
-% make ; a delimiter in math mode
-% \mathcode`\;="8000 % Makes ; active in math mode
-% {\catcode`\;=\active \gdef;{\;}}
-% \mathchardef\semicolon="003B
-
-% reserved words
-\newcommand{\mathem}{\bf}
-
-% brackets
-\newcommand{\set}[1]{\{#1\}}
-\newcommand{\sbs}[1]{\lquote #1 \rquote}
-
-% arrays
-\newcommand{\ba}{\begin{array}}
-\newcommand{\ea}{\end{array}}
-\newcommand{\bda}{\[\ba}
-\newcommand{\eda}{\ea\]}
-\newcommand{\ei}{\end{array}}
-\newcommand{\bcases}{\left\{\begin{array}{ll}}
-\newcommand{\ecases}{\end{array}\right.}
-
-% \cal ids
-\renewcommand{\AA}{{\cal A}}
-\newcommand{\BB}{{\cal B}}
-\newcommand{\CC}{{\cal C}}
-\newcommand{\DD}{{\cal D}}
-\newcommand{\EE}{{\cal E}}
-\newcommand{\FF}{{\cal F}}
-\newcommand{\GG}{{\cal G}}
-\newcommand{\HH}{{\cal H}}
-\newcommand{\II}{{\cal I}}
-\newcommand{\JJ}{{\cal J}}
-\newcommand{\KK}{{\cal K}}
-\newcommand{\LL}{{\cal L}}
-\newcommand{\MM}{{\cal M}}
-\newcommand{\NN}{{\cal N}}
-\newcommand{\OO}{{\cal O}}
-\newcommand{\PP}{{\cal P}}
-\newcommand{\QQ}{{\cal Q}}
-\newcommand{\RR}{{\cal R}}
-\newcommand{\TT}{{\cal T}}
-\newcommand{\UU}{{\cal U}}
-\newcommand{\VV}{{\cal V}}
-\newcommand{\WW}{{\cal W}}
-\newcommand{\XX}{{\cal X}}
-\newcommand{\YY}{{\cal Y}}
-\newcommand{\ZZ}{{\cal Z}}
-
-% misc symbols
-\newcommand{\dhd}{\!\!\!\!\!\rightarrow}
-\newcommand{\Dhd}{\!\!\!\!\!\Rightarrow}
-\newcommand{\ts}{\,\vdash\,}
-\newcommand{\la}{\langle}
-\newcommand{\ra}{\rangle}
-\newcommand{\eg}{{\em e.g.}}
-
-% misc identifiers
-\newcommand{\dom}{\mbox{\sl dom}}
-\newcommand{\fn}{\mbox{\sl fn}}
-\newcommand{\bn}{\mbox{\sl bn}}
-\newcommand{\sig}{\mbox{\sl sig}}
-\newcommand{\IF}{\mbox{\mathem if}}
-\newcommand{\OTHERWISE}{\mbox{\mathem otherwise}}
-\newcommand{\expand}{\prec}
-\newcommand{\weakexpand}{\prec^W}
-\newcommand{\spcomma}{~,~}
-
-%\newcommand{\inst}{\mbox{\mathem inst}}
-\newcommand{\trans}[1]{\la\!\la#1\ra\!\ra}
-\newcommand{\remark}[1]{{\bf $\clubsuit$ #1 $\spadesuit$}}
-\newcommand{\todo}[1]{\remark{to do: #1}}
-%\newcommand{\J}{\justifies}
-%\newcommand{\U}{\using}
-
-% names
-\newcommand{\Scala}{\mbox{\textsc{Scala}}}
-\newcommand{\Java}{\mbox{\textsc{Java}}}
-
-%\renewcommand\textfraction{.05}
-%\renewcommand\floatpagefraction{.9}
-%\renewcommand\topfraction{.8}
-
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-%   Language abstraction commands     %
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-
-%% Relations
-% Subtype 
-\newcommand{\sub}{<:}
-% Type assignment
-\newcommand{\typ}{:}
-% reduction
-\newcommand{\reduces}{\;\rightarrow\;}
-% well-formedness
-\newcommand{\wf}{\;\mbox{\textbf{wf}}}
-\newcommand{\nswf}{\mbox{\textbf{wf}}}
-\newcommand{\wfe}{\;\mbox{\textbf{wfe}}}
-\newcommand{\nswfe}{\mbox{\textbf{wfe}}}
-
-%% Operators
-% Type selection
-\newcommand{\tsel}{\#}
-% Function type
-\newcommand{\tfun}{\rightarrow}
-\newcommand{\dfun}[3]{(#1\!:\!#2) \Rightarrow #3}
-% Conjunction
-\newcommand{\tand}{\wedge}
-% Disjunction
-\newcommand{\tor}{\vee}
-% Singleton type suffix
-\newcommand{\sing}{.\textbf{type}}
-
-%% Syntax
-% Header for typing rules
-\newcommand{\judgement}[2]{{\bf #1} \hfill #2}
-% Refinement
-\newcommand{\refine}[2]{\left\{#1 \Rightarrow #2 \right\}}
-\newcommand{\mlrefine}[2]{\{#1 \Rightarrow #2 \}}
-% Field definitions
-\newcommand{\ldefs}[1]{\left\{#1\right\}}
-\newcommand{\mlldefs}[1]{\{#1\}}
-% Member sequences
-\newcommand{\seq}[1]{\overline{#1}}
-% Lambda
-\newcommand{\dabs}[3]{(#1\!:\!#2)\Rightarrow #3}
-\newcommand{\abs}[3]{\lambda #1\!:\!#2.#3}
-% Method Application
-\newcommand{\mapp}[3]{#1.#2(#3)}
-% Substitution
-\newcommand{\subst}[3]{[#1/#2]#3}
-% Object creation
-\newcommand{\new}[3]{\textbf{val }#1 = \textbf{new }#2 ;\; #3}
-\newcommand{\mlnew}[3]{\textbf{val }#1 = \textbf{new }#2 ;\;\\&#3}
-%\renewcommand{\new}[3]{#1 \leftarrow #2 \,\textbf{in}\, #3}
-% Field declaration
-\newcommand{\Ldecl}[3]{#1 : #2..#3}%{#1 \operatorname{>:} #2 \operatorname{<:} #3}
-\newcommand{\ldecl}[2]{#1 : #2}
-\newcommand{\mdecl}[3]{#1 : #2 \tfun #3}
-% Top and Bottom
-\newcommand{\Top}{\top}%{\textbf{Top}}
-\newcommand{\Bot}{\bot}%\textbf{Bot}}
-% Environment extension
-%\newcommand{\envplus}[1]{\uplus \{ #1 \}}
-\newcommand{\envplus}[1]{, #1}
-% Reduction
-\newcommand{\reduction}[4]{#1 \operatorname{|} #2 \reduces #3 \operatorname{|} #4}
-
-% Sugar
-\newcommand{\arrow}[2]{#1\rightarrow_s#2}
-\newcommand{\fun}[4]{\textbf{fun } (#1:#2)\;#3\;#4}
-\newcommand{\app}[2]{(\textbf{app }#1\;#2)}
-\newcommand{\mlapp}[2]{(\textbf{app }#1\;\\&#2)}
-\newcommand{\cast}[2]{(\textbf{cast }#1\;#2)}
-
-\newcommand{\lindent}{\hspace{-4mm}}
-
-% Logical relations
-\newcommand{\relv}[4]{\mathcal{V}_{#1;#2;#3}\llbracket#4\rrbracket}
-\newcommand{\rele}[4]{\mathcal{E}_{#1;#2;#3}\llbracket#4\rrbracket}
-\newcommand{\rels}[3]{\mathcal{\supseteq}_{#1}\llbracket#2;#3\rrbracket}
-\newcommand{\relg}[3]{\mathcal{\supseteq^!}_{#1;#2}\llbracket#3\rrbracket}
-\newcommand{\irred}[2]{\text{irred }(#1,#2)}
-\newcommand{\andl}{\;\wedge\;}
-\newcommand{\orl}{\vee}
-\newcommand{\impliesl}{\rightarrow}
-\newcommand{\reductionl}[5]{#1 \operatorname{|} #2 \;\rightarrow^{#5}\; #3 \operatorname{|} #4}
-\newcommand{\ds}{\,\vDash\,}
-
+\input{dot_macros}
 \begin{document}
@@
 \keywords
-calculus, objects, dependent types, step-indexed logical relations
+calculus, objects, dependent types%, step-indexed logical relations
 
@@
 Given these definitions, here is a valid expression, which evaluates to $f$:
-$\cast \Top \app {{\mi{rf.choose}}(f)} l$.
+$\cast \Top {\app {{\mi{rf.choose}}(f)} l}$.
 
@@
 calculus for preservation to hold.
-
-\section{Type Safety via Logical Relations}\label{dot-type-safety}
-We sketch a proof of type-safety of the DOT calculus via step-indexed
-logical relations~\cite{ahmed04,ahmed06,step_indexed_obj}.
-
-\subsection{Type Safety}
-Type-safety states that a well-typed program doesn't get stuck. More
-formally: If $\emptyset \ts t \typ T$ and $\reductionl t \emptyset {t'} {s'} *$ then
-either $t'$ is a value or $\exists t'', s''. \reduction {t'} {s'}
-{t''} {s''}$.
-
-Our strategy is to define a logical relation $\Gamma \ds t : T$, such
-that $\Gamma \ts t : T$ implies $\Gamma \ds t : T$ implies
-type-safety.
-
-\subsection{Step-Indexed Logical Relations}
-
-In order to ensure that our logical relation is well-founded, we use a
-step index. For each step index $k$, we define the set of values and
-the set of terms that appear to belong to a given type, when taking at
-most $k$ steps. $\Gamma \ds t : T$ is then defined in terms of the
-step-indexed logical relation by requiring it to hold $\forall k$.
-
-\subsubsection{Set of Values}
-$\relv k \Gamma s T$ defines the set of values that appear to have
-type $T$ when taking at most $k$ steps. $\Gamma$ and $s$ must agree:
-$\dom(\Gamma) = \dom(s)$ (ordered) and $\forall (x : T) \in \Gamma, x
-\in \relv k \Gamma s T$. A variable $y$ belongs to $\relv 0 \Gamma s
-T$ simply by being in the store. In addition, it belongs to $\relv k
-\Gamma s T$ for $k > 0$, if it defines all type, method and value labels in
-the expansion of $T$ appropriately for $j < k$ steps.
-\begin{align*}
-&\relv k \Gamma s T = \{ y \ | y \in \dom(s) \andl (\\
-& (\Gamma \ts T \wfe \andl\\
-&\gap \forall j < k,\\
-&\gap y \mapsto T_c \ldefs{\seq{l = v}\;\seq{m(x)=t}} \in s,\\
-&\gap \Gamma \ts T \expand_y \seq{D},\\
-&\gap (\forall L_i : S \rightarrow U \in \seq{D},\\
-&\gap\gap \Gamma \ts y \ni L_i \typ {S'..U'}) \andl\\
-&\gap (\forall m_i : S \rightarrow U \in \seq{D},\\
-&\gap\gap t_i \in \rele j {\Gamma \envplus{x_i : S}} s U) \andl\\
-&\gap (\forall l_i : V \in \seq{D},\\
-&\gap\gap v_i \in \relv j \Gamma s V)) \orl\\
-& (T = T_1 \tand T_2 \andl y \in \relv k \Gamma s {T_1} \andl y \in \relv k \Gamma s {T_2}) \orl\\
-& (T = T_1 \tor T_2 \andl  (y \in \relv k \Gamma s {T_1} \orl y \in \relv k \Gamma s {T_2}))\\
-&)\}
-\end{align*}
-
-This relation captures the observation that the only ways for a term
-to get stuck is to have a field selection on an uninitialized field or
-a method invocation on an uninitialized method. However, a {\it potential
-  pitfall} is that the value itself might occur in the types $S$, $U$,
-$V$, because we substitute it for the ``self'' occurrences in the
-expansion, so the relation makes sure that the required type labels exist.
-
-\subsubsection{Set of Terms}
-$\rele k \Gamma s T$ defines the set of terms that appear to have type
-$T$ when taking at most $k$ steps. $s$ must agree with a {\it prefix}
-of $\Gamma$, so $\Gamma$ can additionally contain variables not in
-$s$. This is needed for checking methods in $\mathcal{V}$ above, and
-for relating open terms. If $k > 0$, $\mathcal{E}$ extends $\Gamma$
-and $s$ so that they agree. It then states that if it can reduce $t$
-in the extended store to an irreducible term in $j < k$ steps, then
-this term must be in a corresponding $\mathcal{V}$ set with $\Gamma$
-now extended to agree with the store resulting from the reduction
-steps.
-
-$\irred {t} {s}$ is a shorthand for $\neg\exists t', s'. \reduction t s
-{t'} {s'}$. $\mathcal{\supseteq}$ is used initially for the possibly
-shorter store to agree with the environment, and can extend both in
-many different ways. $\mathcal{\supseteq}^!$ is used finally for the
-possibly shorter environment to agree with the store, and just extends
-the environment in one straightforward way: hence, it defines
-singleton sets.
-
-\begin{align*}
-&\rele k \Gamma s T = \{ t \ |\\
-&\gap k = 0 \orl (\forall j < k,\\
-&\gap\gap \forall (\Gamma'; s') \in \rels k \Gamma s ,\\
-&\gap\gap \reductionl t {s'} {t'} {s''} j \andl\\
-&\gap\gap \irred {t'} {s''} \impliesl\\
-&\gap\gap \forall \Gamma'' \in \relg k {s''} {\Gamma'} ,\\
-&\gap\gap {t'} \in \relv {k-j-1} {\Gamma''} {s''} T)\\
-&\}
-\end{align*}
-
-\subsubsection{Extending the environment and the store}
-$\rels k \Gamma s$ for $k > 0$ defines the set of completed
-environment and stores that agree on $k-1$ steps, and that extend
-$\Gamma$ and $s$. $s$ must agree with a {\it prefix} of $\Gamma$.
-Both $\Gamma$ and $s$ are ordered maps. For $s$, $s'$ extends $s$ if
-$s$ is a prefix of $s'$. For $\Gamma$, $\Gamma'$ extends $\Gamma$ if
-we get back $\Gamma$ by keeping only the elements of $\Gamma'$ that
-belong to $\Gamma$. Furthermore, a prefix of $\Gamma'$ agrees with
-$s$.
-\begin{align*}
-&\rels k \Gamma s = \{\\
-&\ (\seq{x : T}^m, \seq{x_{ij} : T_{ij}}^{m \leq i < n; 0 \leq j < i_n}; s, \seq{x_{ij} \mapsto c_{ij}}^{m \leq i < n; 0 \leq j < i_n}) |\\
-&\ s = \seq{x \mapsto c}^m \andl \Gamma = \seq{x : T}^n \andl \\
-&\ m\leq n \andl \forall i, m \leq i < n, \forall {i_n}, j, 0 \leq j < i_n,\\
-&\ \forall T_{ij}, c_{ij}, T_{i{(i_n-1)}} = T_i, \forall n' \leq n, i_n' \leq i_n,\\
-&\ c_{ij} \in \relv {k-1} {\seq{x : T}^m, \seq{x_{ij} : T_{ij}}^{m \leq i' < n; 0 \leq j < i_n'}} {s, \seq{x_{ij} \mapsto c_{ij}}^{m \leq i < n'; 0 \leq j < i_n'}} {T_{ij}}\\
-&\}
-\end{align*}
-
-\subsubsection{Completing the environment to agree with the store}
-$\relg k s \Gamma$ defines a singleton set of a completed environment
-that agrees with a store $s$ by simply copying the constructor type
-from the store for each missing variable.
-\begin{align*}
-&\relg k s \Gamma = \{ \Gamma, \seq{x_i : T_{c_i}}^{m \leq i < n} \ |\\
-&\gap \Gamma = \seq{x : T}^m \andl s = \seq{x \mapsto c}^n\\
-&\}
-\end{align*}
-
-\subsubsection{Terms in the Logical Relation}
-
-$\Gamma \ds t : T$ is simply defined as $t \in \rele k \Gamma \emptyset T,
-\forall k$.
-
-\subsection{Statements and Proofs}
-
-\subsubsection{Fundamental Theorem}
-
-The fundamental theorem is the implication from $\Gamma \ts t : T$ to
-$\Gamma \ds t : T$. Type safety is a straightforward corollary of this
-theorem.
-
-\begin{myproof}
-The proof is on induction on the derivation of $\Gamma \ts t : T$. For
-each case, we need to show $t \in \rele k \Gamma \emptyset T, \forall
-k$. The non-trivial case is when $k > 0$ and for $(\Gamma'; s') \in
-\rels k \Gamma s$ and some $j < k$, $\reductionl {t} s {t'} {s'} j
-\andl \irred {t'} {s'}$. Then, we need to show ${t'} \in \relv {k-j-1}
-      {\Gamma''} {s'} T$ for ${\Gamma''} \in \relg \Gamma k {s'}
-      {\Gamma'}$.
-
-\ \\
-
-\Case \textsc{var}: $\Gamma \ts x : T$ knowing $(x : T) \in
-\Gamma$. $x \in \relv {k-1} {\Gamma'} s T$ follows from the definition of
-$\rels k \Gamma \emptyset$.
-
-\ \\
-
-\Case \textsc{sel}: $\Gamma \ts t_1.l_i : T$ knowing $\Gamma \ts t_1 :
-T_1$, $\Gamma \ts T_1 \expand_z \seq{D}$, $\ldecl {l_i} {V_i} \in
-\seq{D}$ and knowing either that $t_1 = p_1 \andl T = \subst p z
-    {V_i}$ or that $z \not\in \fn(V_i) \andl T = V_i$.
-
-By operational semantics and induction hypothesis, $\reductionl {t_1}
-s {t_1'} {s'} {j-1}$ and $\irred {t_1'} {s'}$ and ${t_1'} \in \relv
-{k-j+1-1} {\Gamma'} {s'} {T_1}$.
-
-By operational semantics and the above, $\reductionl {t_1'.l_i} {s'}
-{t'} {s'} 1$, and we can conclude $t' \in \relv {k-j-1} {\Gamma''} {s'}
-T$ from the clause for value labels of ${t_1'} \in \relv {k-j}
-{\Gamma''} {s'} {T_1}$.
-
-\ \\
-
-\Case \textsc{msel}: $\Gamma \ts t_1.m_i(t_2) : T$ knowing $\Gamma \ts
-t_1 : T_1$, $\Gamma \ts t_2 : T_2$, $\Gamma \ts T_1 \expand_z
-\seq{D}$, $\mdecl {m_i} {S_i} {U_i} \in \seq{D}$ and knowing either
-that $t_1 = p_1 \andl S = \subst p z {S_i} \andl T = \subst p z {U_i}$
-or that $z \not\in \fn(S_i) \andl z \not\in \fn(U_i) \andl S = S_i \andl
-T = U_i$, and knowing that $\Gamma \ts T_2 \sub S$.
-
-By operational semantics and induction hypotheses, $\reductionl {t_1}
-s {t_1'} {s_1} {j_1}$ and $\irred {t_1'} {s_1}$ and $\reductionl {t_2}
-{s} {t_2'} {s_2} {j_2}$ and $\irred {t_2'} {s_2}$ and ${t_1'} \in
-\relv {k-j_1-1} {\Gamma_1} {s_1} {T_1}$ and ${t_2'} \in \relv
-{k-j_2-1} {\Gamma_2} {s_2} {T_2}$.
-
-Because $t_2$ reduces to a value $t_2'$ starting in store $s$, it
-should also reduce to a value $v_2$ in the same number of steps
-starting in store $s_1$, since $s_1$ extends $s$. So let $\reductionl
-{t_2} {s_1} {v_2} {s_{12}} {j_2}$ with $v_2 \in \relv {k-j_2-1}
-{\Gamma_{12}} {s_{12}} {T_2}$.
-
-By the above and operational semantics, $\reductionl {t_1'.m_i(v_2)}
-{s_{12}} {\subst {v_2} {x_i} {t_i}} {s_{12}} 1$.
-
-By the substitution lemma, ${\subst {v_2} {x_i} {t_i}} \in \rele
-{k-\max(j_1,j_2)-1} {\Gamma_{12}} {s_{12}} {T}$. Supposing,
-$\reductionl {\subst {v_2} {x_i} {t_i}} {s_{12}} {t'} {s'} {j_3}$,
-with $j_1 + j_2 + j_3 + 1 = j$, this completes the case, by
-monotonicity of $\mathcal{V}$.
-
-\ \\
-
-\Case \textsc{new}: $\Gamma \ts \new y c {t_b} : T$ knowing ...
-
-By operational semantics, $\reductionl {\new y c {t_b}} s {t_b} {s_b} 1$
-where $s_b = s \envplus{y \mapsto c}$. So $\reductionl {t_b} {s_b} {t'}
-{s'} {j-1}$.
-
-By induction hypotheses, $y \in \relv k {\Gamma_b} {s_b} {T_c}$ and $t_b \in
-\rele k {\Gamma_b} {s_b} {T}$.
-
-Result follows by monotonicity of $\mathcal{V}$.
-
-\end{myproof}
-
-\subsubsection{Substitution Lemma}
-
-The substitution lemma states that if (1) $v \in \relv {k_2}
-{\Gamma_{12}} {s_{12}} {T_2}$ and (2) $t \in \rele {k_1} {\Gamma_1
-  \envplus{{x} : S}} {s_1} T$ and (3) $\Gamma \ts {T_2} \sub S$ with
-(4) $x \not\in \fn(T)$ and $\Gamma_1$ extends $\Gamma$ and
-$\Gamma_{12}$ extends $\Gamma_1$ and $s_{12}$ extends $s_1$ and
-$\Gamma_1$ agrees with $s_1$ and $\Gamma_{12}$ agrees with $s_{12}$
-and a prefix of $\Gamma_{12}$ agrees with $s_1$, then $\subst {v} {x}
-{t} \in \rele {\min(k_1,k_2)} {\Gamma_{12}} {s_{12}} T$.
-
-\begin{proofsketch}
-By (1) and (3), it should hold that (5) $v \in \relv {k_2}
-{\Gamma_{12}} {s_{12}} {S}$ by the subset semantics lemma.  Since (2)
-holds, it should also hold that ${t} \in \rele {\min(k_1,k_2)}
-{\Gamma_{12} \envplus{{x} : S}} {s_{12}} T$ by the extended
-monotonicity lemma. Then, we can instantiate $x$ in the complete store
-to map to what $v$ maps to. This should be fine by (5) and
-monotonicity. Thus, ${t} \in \rele {\min(k_1,k_2)} {\Gamma_{12}
-  \envplus{{x} : S}} {s_{12} \envplus{{x} \mapsto s_{12}(v)}}
-T$. Thanks to (4), we don't actually need $x$ to be held abstract in
-the environment, because it won't occur in $T$ or its expansion (a
-{\it potential pitfall} is whether its occurrences in $t_i$ could
-still cause a check to fail through narrowing issues), so we can use
-the type of $v$ in the environment instead of $S$ for $x$: ${t} \in
-\rele {\min(k_1,k_2)} {\Gamma_{12} \envplus{{x} : {\Gamma_{12}(v)}}}
-      {s_{12} \envplus{{x} \mapsto s_{12}(v)}} T$. This implies what
-      needs to be shown.
-\end{proofsketch}
-
-\subsubsection{Subset Semantics Lemma}
-
-The subset semantics lemma states that if $v \in \relv k \Gamma s S$
-and $\Gamma \ts S <: U$, then $v \in \relv k \Gamma s U$.
-
-\begin{proofsketch}
-Because $S$ is a subtype of $U$, it should hold that the expansion of
-$S$ subsumes the expansion of $U$, when the ``self'' occurrences are
-of type $S$. Therefore, for $v \in \relv k \Gamma s U$, we have fewer
-declarations to check than for $v \in \relv k \Gamma s S$.
-
-A {\it potential pitfall} is whether some types of the expansion of
-$U$ can become non-expanding when the ``self'' occurrences are
-actually $v$ instead of just abstractly of type $S$, causing a check
-to fail. Another worry is that such a non-expanding type results from
-narrowing of a parameter type.
-%However, we believe this is fine by
-%subtyping regularity wtr to $\nswfe$ and $v \in \relv k \Gamma s S$
-%which cheks bounds and $\nswfe$ of the type labels.
-\end{proofsketch}
-
-\subsubsection{Extended Monotonicity Lemma}
-
-The extended monotonicity lemma states that if $t \in \rele k {\Gamma
-  \envplus{x : S}} s T$ then $t \in \rele j {\Gamma' \envplus{x : S}}
-{s'} T$ for $j \leq k$, $\Gamma'$ extends $\Gamma$, $s'$
-extends $s$, and $\Gamma$ agrees with $s$ and a prefix of $\Gamma'$
-agrees with $s$.
-
-\begin{proofsketch}
-For the monotonicity with regards to the step index, this follows
-directly from the definitions of $\mathcal{E}$ and $\mathcal{V}$. For
-the environment and the store, this follows by design from the
-definition of $\rels k {\Gamma \envplus{x : S}} s$. To extend the
-environment and the store for $x : S$, we can append as much as we
-want to $\Gamma$ and $s$, to get $\Gamma'$ and $s'$, and then ignore
-the last element which is for $x : S$.
-\end{proofsketch}
 

### 2012-09-22 doc/fool/type_safety.tex
Removing proof of type safety from article.

--- 2012-09-13 doc/fool/proof.tex
@@
 \end{proofsketch}
-
-\section{Discussion}\label{discussion}
-
-\subsection{Why No Inheritance?}\label{why-no-inheritance}
-
-In the calculus we made the deliberate choice not to model any form of
-inheritance. This is, first and foremost, to keep the calculus simple.
-Secondly, there are many different approaches to inheritance and
-mixin composition, so that it looks advantageous not to tie the basic
-calculus to a specific one. Finally, it seems that the modelization of
-inheritance lends itself to a different approach than the basic
-calculus. For the latter, we need to prove type safety of the calculus.
-One might try to do this also for a calculus with inheritance, but our
-experience suggests that this complicates the proofs considerably.  An
-alternative approach that might work better is to model inheritance
-as a form of code-reuse. Starting with an enriched type system with
-inheritance, and a translation to the basic calculus, one needs to
-show type safety wrt the translation. This might be easier than
-to prove type safety wrt reduction.

### 2012-09-22 doc/fool/article.tex
type safety intuition

@@
 the calculus formally in section \ref{dot-calculus} and through
-examples in section~\ref{dot-examples}. Though we show that the
-calculus does not satisfy the standard theorem of preservation in
-section~\ref{dot-preservation}, we contribute a proof sketch of type
-safety using logical relations in section~\ref{dot-type-safety}. In
+examples in section~\ref{dot-examples}. In
+section~\ref{dot-preservation}, we show that the calculus does not
+satisfy the standard theorem of preservation, also known as subject
+reduction. However, we still believe that the calculus is type-safe,
+as discussed in section~\ref{dot-type-safety}. In
 section~\ref{discussion}, we discuss choices and variants of the
@@
 
-We sketch a proof that the DOT calculus is type-safe using logical
-relations in section~\ref{dot-type-safety}. However, we first tried to
-prove the calculus type-safe using the standard theorems of
-preservation and progress~\cite{soundness,tapl}. Unfortunately, for
-the calculus as presented, and any of its variants that we devised,
-preservation doesn't hold. In this section, we review some of the most
-salient counterexamples to preservation that we found. These
-counterexamples have been checked with PLT Redex~\cite{plt_redex}.
+We first tried to prove the calculus type-safe using the standard
+theorems of preservation and
+progress~\cite{soundness,tapl}. Unfortunately, for the calculus as
+presented, and any of its variants that we devised, preservation
+doesn't hold. In this section, we review some of the most salient
+counterexamples to preservation that we found. These counterexamples
+have been checked with PLT Redex~\cite{plt_redex}.
 
@@
 calculus for preservation to hold.
+
+\section{Type Safety via Logical Relations}\label{dot-type-safety}
+
+We believe that the DOT calculus is type-safe, and are developing a
+proof of type safety using step-indexed logical
+relations~\cite{ahmed04,ahmed06,step_indexed_obj}. In this section, we
+briefly summarize the main theorem of type safety, and the strategy
+based on step-indexed logical relations that we are using to prove
+it. All our development, including models in Coq and PLT Redex, is
+available from \url{http://dot.namin.net}.
+
+\subsection{Type Safety}
+Type safety states that a well-typed program doesn't get stuck. More
+formally: If $\emptyset \ts t \typ T$ and $\reductionl t \emptyset
+{t'} {s'} *$ then either $t'$ is a value or $\exists t'',
+s''. \reduction {t'} {s'} {t''} {s''}$. Note that this is stronger
+than the standard theorem of progress, which states that a well-type
+term can take \emph{a} step or is a value.
+
+\subsection{Strategy based on Logical Relations}
+Our strategy follows the standard technique of proving type safety
+using logical relations. We define a logical relation $\Gamma \ds t
+: T$, such that $\Gamma \ts t : T$ implies $\Gamma \ds t : T$ implies
+type safety.
+
+The main logical relation $\Gamma \ds t : T$ is based on a set of
+mutually recursive logical relations that are step-indexed in order to
+ensure their well-foundedness: $\rele k \Gamma s T$ defines the set of
+terms that appear to have type $T$ when taking at most $k$ steps, and
+$\relv k \Gamma s T$ defines the set of \emph{values} that appear to
+have type $T$ when taking at most $k$ steps. There are also two other
+logical relations, one for completing the store to match the context,
+and one for completing the context to match the store resulting from
+taking some number of steps.
+
+$\Gamma \ds t : T$ is defined as $t \in \rele k \Gamma \emptyset T,
+\forall k$. $t \in \rele k \Gamma s T$ roughly as follows: after
+reducing $t$ a number of steps $<k$, if the resulting term is
+irreducible, then it must be in $\relv {j'} {\Gamma'} {s'} T$ for
+appropriate $j', \Gamma',s'$. By definition, $\Gamma \ds t : T$
+implies that $t$ cannot be stuck.
+
+Then, to prove type safety, all that needs to be proved is the
+fundamental theorem -- or completeness -- of the logical relation:
+$\Gamma \ts t : T$ to $\Gamma \ds t : T$. Type safety is a
+straightforward corollary of this theorem, since $\Gamma \ds t : T$
+implies by definition that $t$ cannot be stuck. The proof of the
+fundamental theorem is on induction on the derivation of $\Gamma \ts t
+: T$. The logical relation approach enables us to have strong enough
+induction hypotheses to carry the proof through, without requiring us
+to strictly relate intermediate terms by types like preservation.
 

### 2012-09-22 doc/fool/article.tex
coq cite and minor rev
more elaborate details on why the patch failed. copied from report, mostly.

@@
 
-\subsection{Expansion}
+\subsection{Expansion}\label{ex-exp}
 
@@
 based on step-indexed logical relations that we are using to prove
-it. All our development, including models in Coq and PLT Redex, is
+it. All our development, including models in Coq~\cite{coq} and PLT Redex~\cite{plt_redex}, is
 available from \url{http://dot.namin.net}.
@@
 We tried! However, the resulting calculi were not elegant, and
-furthermore, we still found issues with preservation. Below, we give a
-brief summary of one failed attempt to patch the calculus for
-preservation to hold.
-
-Because many of the counterexamples to preservation are related to
-narrowing, we tried to make widening an explicit operation and change
-rules with implicit relaxations (\textsc{msel} and \textsc{new}) to be
+furthermore, we still found issues with preservation.
+
+Here is a summary of one attempt, which is further detailed
+below. Because many of the counterexamples to preservation are re-
+lated to narrowing, we tried to make widening an explicit operation
+and change rules with implicit relaxations (MSEL and NEW) to be
 strict. From a typing perspective, the change was straightforward, but
 reduction became more complicated and dependent on typing because the
-type information in widenings needed to be propagated correctly.
-
-We added path equality provisions in the subtyping rules, in the same
-spirit as the Tribe calculus~\cite{tribe}.
-
-Unfortunately, these two patches interacted badly, and we were left
-with a disturbing counterexample to type safety.
+type information in widenings needed to be propagated correctly.  We
+added path equality provisions in the subtyping rules, in the same
+spirit as the Tribe calculus~\cite{tribe}.  Unfortunately, these two
+patches interacted badly, and we were left with a disturbing
+counterexample to type safety.
+
+In any case, this attempt resulted in a patched calculus that was not
+as elegant as the original one, in addition to being unsound. We're
+open to ideas!
+
+In the DOT calculus as presented, the \textsc{app} and \textsc{new} typing
+rules have implicit relaxations. For instance, in \textsc{app}, the
+argument type may be a subtype of the declared parameter type. In
+order to deal with all the preservation counterexamples due to narrowing, we
+tried making widening an explicit operation and changing those rules to be
+strict by replacing those relaxed subtyping judgments with equality
+judgments. Two types $S$ and $T$ are judged to be equal if $S <: T$
+and $T <: S$.
+
+Syntactically, we add a widening term: $\wid t T$, and extend values
+with a case for widening: $\wid v T$. The typing rule for widening,
+\textsc{wid}, is then the only one admitting a subtyping relaxation:
+$\Gamma \ts (\wid t T) \typ T$ if $\Gamma \ts t \typ T'$ and $\Gamma \ts T'
+\sub T$.
+
+The reduction rules become more complicated because the type
+information in the widening needed to be propagated correctly. We will
+motivate this informally with examples.
+
+\begin{align*}
+&\mlnew v {\Top \refine z {\Ldecl {L_a} \Bot \Top, \ldecl l {\Top \refine z {\Ldecl {L_a} \Bot \Top}}}\\
+&\gap\gap\ldefs{l = \wid v {\Top \refine z {\Ldecl {L_a} \Bot \Top}}}}{
+\app {(\abs x \Top x)} {(\wid v {\Top \refine z {\ldecl l \Top}}).l}
+}
+\end{align*}
+
+The term $(\wid v {\Top \refine z {\ldecl l \Top}}).l$ first widens
+$v$ so that the label has type $\Top$ instead of $\Top \refine z
+{\Ldecl {L_a} \Bot \Top}$.
+
+How should reduction proceed? We cannot just strip the widening and
+then reduce, because then the strict function application would not
+accept the reduced term. In short, we need to do some type
+manipulations during reduction, by using the membership and expansion
+judgments. This is a bit unfortunate, because it means that reduction
+now needs to know about typing.
+
+Next, we look at path equality provisions. These are even more
+essential now in the presence of explicit widening. Consider this
+example:
+
+\begin{align*}
+&\mlnew b {\Top \refine z {\Ldecl X \Top \Top, \ldecl l {z.X}} \ldefs{l = \wid b {b.X}}}{
+\mlnew a {\Top \refine z {\ldecl i {\Top \refine z {\Ldecl X \Bot \Top, \ldecl l {z.X}}}}\\
+&\gap\gap\ldefs{i = \wid b {\Top \refine z {\Ldecl X \Bot \Top, \ldecl l {z.X}}}}}}{
+\wid {a.i.l} \Top
+}
+\end{align*}
+
+$a.i.l$ reduces to $\wid b {\Top \refine z {\Ldecl X \Bot \Top, \ldecl
+    l {z.X}}}$. Now, how can we continue? $b.l$ reduces to $b : b.X$
+which has bounds $\Top..\Top$, but $(\wid b {\Top \refine z {\Ldecl X
+    \Bot Top, \ldecl l {z.X}}}).l$ has bounds $\Bot..\Top$, so without
+some provision for path equality, we cannot widen $b.l$ to $(\wid b
+{\Top \refine z {\Ldecl X \Bot \Top, \ldecl l {z.X}}}).l$.
+
+We add the path equality provisions to the subtyping rules.
+
+Let's first ignore the extension of the calculus requiring explicit
+widenings introduced in~\ref{widening}. Then, we need to add one
+intuitive rule to the subtyping judgment: $\textsc{$\sub$-path}$. If
+$p$ (path-)reduces to $q$, and $T \sub q.L$, then $T \sub p.L$. Path
+reduction is a simplified form of reduction involving only
+paths. However, this means that the subtyping judgment, and
+indirectly, all the typing-related judgments, now need to carry the
+store in addition to the context so that path reductions can be
+calculated.
+
+Now, let's see how path equality provisions and explicit widening can
+fit together.
+
+First, path reduction is not isomorphic to reduction anymore, since we
+want to actually skip over widenings, as motivated by the last example
+in~\ref{widening}.
+
+In addition, we now also need a dual rule, $\textsc{path-$\sub$}$: if
+$p$ (path-dually)-reduces to $q$, and $q.L \sub T$ then $p.L \sub
+T$. This is because when we have a widening on an object on which a
+method is called, we have to upcast the argument to the parameter type
+expected by the original method. Here is a motivating example.
+
+Let $T_c$ be a shorthand for the type:
+\begin{align*}
+\Top &\mlrefine z {\\
+&\ \Ldecl A {\Top \refine z {\mdecl m \Bot \Top}} \Top\\
+&\ \Ldecl B \Top \Top\\
+&\ \mdecl m {z.A} \Top\\
+}&
+\end{align*}
+
+Let $T$ be a shorthand for the type:
+\begin{align*}
+\Top &\mlrefine z {\\
+&\ \Ldecl A {\Top \refine z {\mdecl m \Bot \Top}} \Top\\
+&\ \Ldecl B \Top \Top\\
+&\ \mdecl m {{z.A} \refine z {\Ldecl B \Top \Top}} \Top\\
+}&
+\end{align*}
+
+Now, consider the term:
+\begin{align*}
+&\mlnew v {T_c \ldefs{m(x)=\wid x \Top}}{
+(\wid v T).m(\wid v {((\wid v T).{A \refine z {\Ldecl B \Top \Top}})})
+}
+\end{align*}
+
+When we evaluate the method invocation, we need to cast $\wid v
+{((\wid v T).{A \refine z {\Ldecl B \Top \Top}})}$ to $v.A$, and for
+this, we need the newly introduced $\textsc{path-$\sub$}$ rule.
+
+Note that the path dual reduction can be a bit stricter with casts
+than the path reduction. In any case, introducing this
+$\textsc{path-$\sub$}$ rule into the subtyping judgment is
+problematic: it is now possible to say $p.L \sub T$, even though $T$
+can do more than what $p.L$ is defined to do. Here is an example,
+where we construct an object, with $T = \Bot$. (The convolution in the
+example is due to the requirement that concrete types be only
+mentioned once.)
 
@@
 \mlnew c {a.C}}{
-\mlnew d {{\cast {a.C} b}.D}}{
+\mlnew d {(\wid b {a.C}).D}}{
 \app {(\abs x \Bot {x.\mathit{foo}})} d
@@
 
-Notice that $d$ has type $\Bot$ if the cast on $b$ is ignored. This
-example didn't typecheck initially because the path-equality
-provisions only applied when objects are in the store, so the
-application was not well-typed. But if we started preservation in a store
-which had $a$, $b$, $c$ and $d$ then the application type-checked,
-because, through one of the path-equality provision, we could find that
-the type of $d$ was a subtype of $\Bot$. Now, of course, when we got to
-$d.\mathit{foo}$, reduction failed.
+Notice that $d$ has type $\Bot$ if you ignore the cast on $b$. This
+example doesn't typecheck initially because $\textsc{path-$\sub$}$
+only applies when objects are in the store, so the application is not
+well-typed. But if we start preservation in a store which has $a$,
+$b$, $c$ and $d$ then the application type-checks, because, through
+$\textsc{path-$\sub$}$, we can find that the type of $d$ is a subtype
+of $\Bot$. Now, of course, when we get to $d.\mathit{foo}$, reduction
+fails.
+
+So the preservation theorem as defined on a small-step semantics
+(where we start with an arbitrary well-formed environment) fails when
+we add the $\textsc{path-$\sub$}$ rule.
 
@@
 adventure. We have shown that DOT does not satisfy preservation (also
-known as subject-reduction). However, the standard theorems of
-preservation and progress are just one way to prove type safety. We
-have sketched a plausible proof of type safety using the powerful
-method of logical relations.
+known as subject-reduction), because a well-typed term can step to an
+intermediate term that is not well-typed. We're open to ideas on how
+to resolve this. In any case, the standard theorems of preservation
+and progress are just one way to prove type safety, which simply
+states that a well-typed term cannot be stuck. We are developing a
+proof of type safety using logical relations.
 

### 2012-09-22 doc/fool/article.tex
remove undef refs and conc

@@
 Let's first ignore the extension of the calculus requiring explicit
-widenings introduced in~\ref{widening}. Then, we need to add one
+widenings. Then, we need to add one
 intuitive rule to the subtyping judgment: $\textsc{$\sub$-path}$. If
@@
 First, path reduction is not isomorphic to reduction anymore, since we
-want to actually skip over widenings, as motivated by the last example
-in~\ref{widening}.
+want to actually skip over widenings, as motivated by the example above.
 
@@
 known as subject-reduction), because a well-typed term can step to an
-intermediate term that is not well-typed. We're open to ideas on how
-to resolve this. In any case, the standard theorems of preservation
-and progress are just one way to prove type safety, which simply
-states that a well-typed term cannot be stuck. We are developing a
-proof of type safety using logical relations.
+intermediate term that is not well-typed. In any case, the standard
+theorems of preservation and progress are just one way to prove type
+safety, which states that a well-typed term cannot be stuck -- a
+weaker statement that does not require typechecking intermediate
+steps. We are developing a proof of type safety using logical
+relations.
 

### 2012-09-22 doc/fool/article.tex
type safety to type-safety and typecheck to type-check

@@
 
-The expansion relation $\expand$ is needed to typecheck the complete
+The expansion relation $\expand$ is needed to type-check the complete
 set of declarations carried by a concrete type that is used in a
@@
 
-$a$ enforces that its first argument be a $\mi{pets.Poodle_c}$ and its second a $\mi{pets.Dalmatian_c}$. The following does not typecheck for this reason:
+$a$ enforces that its first argument be a $\mi{pets.Poodle_c}$ and its second a $\mi{pets.Dalmatian_c}$. The following does not type-check for this reason:
 \begin{align*}
@@
 More generally, these counterexamples illustrate that preservation
-doesn't hold because a term that typechecks can step to a term that
-does not typecheck. However, these counterexamples to preservation are
+doesn't hold because a term that type-checks can step to a term that
+does not type-check. However, these counterexamples to preservation are
 not counterexamples to type-safety: i.e. these programs don't get
@@
 expand its upper bound, which is $z.L$! However, not that the object
-creation expression for $z$ would not typecheck because subtyping is
+creation expression for $z$ would not type-check because subtyping is
 regular wrt $\nswfe$, and so the type of a constructor in an object
@@
 
-\section{Type Safety via Logical Relations}\label{dot-type-safety}
+\section{Type-Safety via Logical Relations}\label{dot-type-safety}
 
 We believe that the DOT calculus is type-safe, and are developing a
-proof of type safety using step-indexed logical
+proof of type-safety using step-indexed logical
 relations~\cite{ahmed04,ahmed06,step_indexed_obj}. In this section, we
-briefly summarize the main theorem of type safety, and the strategy
+briefly summarize the main theorem of type-safety, and the strategy
 based on step-indexed logical relations that we are using to prove
@@
 
-\subsection{Type Safety}
-Type safety states that a well-typed program doesn't get stuck. More
+\subsection{Type-Safety}
+Type-safety states that a well-typed program doesn't get stuck. More
 formally: If $\emptyset \ts t \typ T$ and $\reductionl t \emptyset
@@
 \subsection{Strategy based on Logical Relations}
-Our strategy follows the standard technique of proving type safety
+Our strategy follows the standard technique of proving type-safety
 using logical relations. We define a logical relation $\Gamma \ds t
 : T$, such that $\Gamma \ts t : T$ implies $\Gamma \ds t : T$ implies
-type safety.
+type-safety.
 
@@
 
-Then, to prove type safety, all that needs to be proved is the
+Then, to prove type-safety, all that needs to be proved is the
 fundamental theorem -- or completeness -- of the logical relation:
-$\Gamma \ts t : T$ to $\Gamma \ds t : T$. Type safety is a
+$\Gamma \ts t : T$ to $\Gamma \ds t : T$. Type-safety is a
 straightforward corollary of this theorem, since $\Gamma \ds t : T$
@@
 inheritance lends itself to a different approach than the basic
-calculus. For the latter, we need to prove type safety of the calculus.
+calculus. For the latter, we need to prove type-safety of the calculus.
 One might try to do this also for a calculus with inheritance, but our
@@
 inheritance, and a translation to the basic calculus, one needs to
-show type safety wrt the translation. This might be easier than
-to prove type safety wrt reduction.
+show type-safety wrt the translation. This might be easier than
+to prove type-safety wrt reduction.
 
@@
 The idea is to start with a function from $S \tfun S$ and cast it
-successively to $S \tfun T$ then $S \tfun U$. To typecheck the
+successively to $S \tfun T$ then $S \tfun U$. To type-check the
 expression initially, we need to check $S \sub T$ and $T \sub
@@
 Now, $f$ was a subtype of $\Top \tfun \Top$, but $\app f {(\abs x \Top
-  x)}$ was stuck (and, rightfully, didn't typecheck). But we could use
-narrowing to create a counterexample to type safety: $\app {(\abs g
+  x)}$ was stuck (and, rightfully, didn't type-check). But we could use
+narrowing to create a counterexample to type-safety: $\app {(\abs g
   {\Top \tfun \Top} {\app g (\abs x \Top x)})} f$.
@@
 patches interacted badly, and we were left with a disturbing
-counterexample to type safety.
+counterexample to type-safety.
 
@@
 Notice that $d$ has type $\Bot$ if you ignore the cast on $b$. This
-example doesn't typecheck initially because $\textsc{path-$\sub$}$
+example doesn't type-check initially because $\textsc{path-$\sub$}$
 only applies when objects are in the store, so the application is not
@@
 safety, which states that a well-typed term cannot be stuck -- a
-weaker statement that does not require typechecking intermediate
-steps. We are developing a proof of type safety using logical
+weaker statement that does not require type-checking intermediate
+steps. We are developing a proof of type-safety using logical
 relations.

### 2012-09-22 doc/fool/article.tex
minor rev

@@
 Here is a summary of one attempt, which is further detailed
-below. Because many of the counterexamples to preservation are re-
-lated to narrowing, we tried to make widening an explicit operation
-and change rules with implicit relaxations (MSEL and NEW) to be
+below. Because many of the counterexamples to preservation are related
+to narrowing, we tried to make widening an explicit operation and
+change rules with implicit relaxations (\textsc{msel} and \textsc{new}) to be
 strict. From a typing perspective, the change was straightforward, but

### 2012-09-30 doc/fool/article.tex
spell out glb and lub throughout

@@
 constructs for type members and by providing classical intersection
-and union types which simplify glb/lub computations.
+and union types which simplify greatest lower bound and least upper bound computations.
 
@@
 
-This is problematic because glbs and lubs play a central role in
+This is problematic because greatest lower bounds and least upper bounds play a central role in
 Scala's type inference. For example, in order to infer the type of an
@@
 \end{lstlisting}
-type inference tries to compute the glb of \code{A} and \code{B} and
-the lub of \code{C} and \code{D}. The absence of universal glbs and
-lubs makes type inference more brittle and more unpredictable.
+type inference tries to compute the greatest lower bound of \code{A} and \code{B} and
+the least upper bound of \code{C} and \code{D}. The absence of universal greatest lower bounds and
+least upper bounds makes type inference more brittle and more unpredictable.
 
@@
 
-\subsection{Glbs and Lubs}
-
-In DOT, the glb of two types is their intersection and their lub,
+\subsection{Greatest Lower Bounds and Least Upper Bounds}
+
+In DOT, the greatest lower bound of two types is their intersection and their least upper bound,
 their union. Recall the introductory example which was problematic in
@@
 
-The glb is of $p.A_c$ and $p.B_c$ is $p.A_c \tand p.B_c$ by
+The greatest lower bound of $p.A_c$ and $p.B_c$ is $p.A_c \tand p.B_c$ by
 definition.
@@
 
-%\section{Examples}
-%TODO: WIP
-%
-%\includegraphics[scale=0.7]{../../src/redex/ex_glb}
-
 %\appendix

### 2012-10-13 doc/fool/article.tex
minor
Template for slides.
Also refactored dot_macros, so that it can be partly used in the
slides.
f-bounded ex

@@
 \documentclass[9pt]{sigplanconf}
+\input{macros}
 \input{dot_macros}
@@
 reduction. However, we still believe that the calculus is type-safe,
-as discussed in section~\ref{dot-type-safety}. In
+as explained in section~\ref{dot-type-safety}. In
 section~\ref{discussion}, we discuss choices and variants of the
@@
 &\gap\gap\gap \app {{\mi{a.choose}}(a)} b}}}{
-\mlnew {\mi{rf}} {{\mi{m.MetaAlt_c}}\mlldefs{\\
+\mlnew {\mi{rl}} {{\mi{m.MetaAlt_c}}\mlldefs{\\
 &\gap\gap\mi{choose}(a)=\fun b {\mi{m.MetaAlt_c}} {\mi{m.MetaAlt_c}} {\\

### 2012-11-15 doc/fool/type_safety.tex
Math typos.

@@
 &\ m\leq n \andl \forall i, m \leq i < n, \forall {i_n}, j, 0 \leq j < i_n,\\
-&\ \forall T_{ij}, c_{ij}, T_{i{(i_n-1)}} = T_i, \forall n' \leq n, i_n' \leq i_n,\\
-&\ c_{ij} \in \relv {k-1} {\seq{x : T}^m, \seq{x_{ij} : T_{ij}}^{m \leq i' < n; 0 \leq j < i_n'}} {s, \seq{x_{ij} \mapsto c_{ij}}^{m \leq i < n'; 0 \leq j < i_n'}} {T_{ij}}\\
+&\ \forall T_{ij}, c_{ij}, T_{i{(i_n-1)}} = T_i, \forall n' \leq n, i_{n'} \leq i_n,\\
+&\ c_{ij} \in \relv {k-1} {\seq{x : T}^m, \seq{x_{ij} : T_{ij}}^{m \leq i < n'; 0 \leq j < i_{n'}}} {s, \seq{x_{ij} \mapsto c_{ij}}^{m \leq i < n'; 0 \leq j < i_{n'}}} {T_{ij}}\\
 &\}

### 2012-11-16 src/scala/old/src/scala/dot/Parser.scala
just make it compile

@@ -143,7 +143,7 @@ object TestParser extends Parsing with PrettyPrinting with Application with Eval
 
   def parse(in: String) = phrase(Parser.term)(new lexical.Scanner(in))
 
-  val source = Source.fromPath("../dot.txt")
+  val source = Source.fromFile("../dot.txt")
   val lines = source.getLines().mkString
   println("parsing: " + lines)
   println("******************")

### 2012-11-16 src/scala/old/src/scala/dot/Typer.scala

@@ -14,9 +14,9 @@ import scala.util.Equalities
  */
 trait Constraints extends StandardTyperMonad with  TyperSyntax {
   type State = Set[Constraint]
-  val initState = Set()
+  val initState = Set[Constraint]()
 
-  def addConstraint(c: Constraint): TyperMonad[()] =
+  def addConstraint(c: Constraint): TyperMonad[Unit] = ()
 
   implicit def constrainExpected(tp1: Expected[Type]) = new {def <:<(tp2: Expected[Type]): Constraint = new SubtypeConstraint(tp1, tp2) }
 
@@ -34,7 +34,8 @@ class Typer extends StandardTyperMonad with Constraints with TyperSyntax with No
   def ofT(tm: Term, pt: Expected[Type]): TyperMonad[Unit] = tm match {
     // Var
     case Var(x) => for(
-        _  <- pt := lookup(x)) yield ()
+        r  <- lookup(x);
+        _  <- pt := r) yield ()
     // Sel
     case Sel(t, l) => for(
         _  <- hasMem(t, l, pt)) yield ()
@@ -48,9 +49,9 @@ class Typer extends StandardTyperMonad with Constraints with TyperSyntax with No
         _   <- wf(Check(ta));
         tr  = Infer[Type]("tr");
         _   <- assume(x, ta)(ofT(b, tr));
-        tr  <- !tr;
-        _   <- fresh(x, tr);
-        _   <- pt := FunT(ta, tr.toMetaVar(MetaType));
+        tr <- !tr;
+        _   <- fresh(x, Check(tr));
+        _   <- pt := FunT(ta, Check(tr).toMetaVar(MetaType));
         _   <- wf(pt)) yield ()
     // New
     case New(tc, \\(x, (args, b))) => for( // parsing has already checked tc is a class type
@@ -73,7 +74,7 @@ class Typer extends StandardTyperMonad with Constraints with TyperSyntax with No
           <:<(Check(s), Check(u))
         case TypeDecl(l, tp) =>
           for(a <- exactlyOne(args.defs find {_.l === l});
-              _ <- ofT(a.rhs, tp)) yield ()
+              _ <- ofT(a.rhs, Check(tp))) yield ()
       }
 
   // for nom = Check(cb) check that x is fresh in nom

### 2012-11-16 src/scala/old/src/scala/dot/TyperMonad.scala

@@ -43,12 +43,12 @@ trait TyperMonad extends AbstractBindingSyntax {
 
     // chain the computations yielded by f(x) for all x in xs, so that, if f(x) is a failure for some x in xs,
     // the computation produced by forall fails; the computation succeeds iff, for all x in xs, f(x) succeeds
-    def forall[T](xs: Traversable[T])(f: T => TyperMonad[()]): TyperMonad[()] =
+    def forall[T](xs: Traversable[T])(f: T => TyperMonad[Unit]): TyperMonad[Unit] =
       (xs map f).foldLeft(result(())){(acc, fx) => acc >>= {_ => fx}}
 
 
     def exactlyOne[T](xs: Traversable[T], err: String="required exactly 1 element"): TyperMonad[T] =
-      if(xs.length == 1) result(xs.head)
+      if(xs.size == 1) result(xs.head)
       else fail(err)
   }
 

### 2012-11-16 src/scala/old/src/scala/dot/Parser.scala
path update

@@ -143,7 +143,7 @@ object TestParser extends Parsing with PrettyPrinting with Application with Eval
 
   def parse(in: String) = phrase(Parser.term)(new lexical.Scanner(in))
 
-  val source = Source.fromFile("../dot.txt")
+  val source = Source.fromFile("../../../../dot.txt")
   val lines = source.getLines().mkString
   println("parsing: " + lines)
   println("******************")

### 2012-11-16 src/scala/src/main/scala/dot/Monads.scala
Monads :-)

@@ -1,0 +1,99 @@
+package dot
+
+// based in part on:
+// Monadic Parser Combinator by Graham Hutton and Erik Meijer
+// technical report NOTTCS-TR-96-4, available at
+// http://www.cs.nott.ac.uk/~gmh/bib.html#monparsing
+
+trait MonadCompanion[M[_]] {
+  // returns the computation that yields the given value 
+  implicit def result[A](x: A): M[A]
+}
+
+trait Monad[A, M[_]] {
+  implicit val companion: MonadCompanion[M]  
+  import companion._
+  
+  // chains this computation with the next one, `f`
+  def >>=[B](f: A => M[B]): M[B] 
+  
+  
+  def map[B](f: A => B): M[B]        = >>= {x => result(f(x))}
+  def flatMap[B](f: A => M[B]): M[B] = >>=(f)
+}
+
+trait MonadZeroPlusCompanion[M[_]] extends MonadCompanion[M] {
+  // returns the trivial computation
+  implicit def zero[A]: M[A]
+}
+
+trait MonadZeroPlus[A, M[x] <: MonadZeroPlus[x, M]] extends Monad[A, M] {
+  implicit override val companion: MonadZeroPlusCompanion[M]
+  import companion._
+
+  // the choice operator for computations, chose either this computation or `o`
+  def ++(o: M[A]): M[A]
+}
+  
+trait StateMonadCompanion[M[_], State] extends MonadCompanion[M] {
+  def apply[A](fun: State => (A, State)): M[A]
+
+  implicit def result[A](x: A): M[A] = apply((x, _))
+  
+  def update(f: State => State): M[State] = apply{s: State => (s, f(s))}
+  def set(s: State): M[State] = update(x => s)
+  def fetch = update(x => x)
+}
+
+trait StateMonad[A, M[x] <: StateMonad[x, M, State], State] extends Monad[A, M] {
+  implicit override val companion: StateMonadCompanion[M, State]
+  import companion._
+  
+  val initState: State
+  val stateTrans: State => (A, State)
+
+  def >>=[B](f: A => M[B]) = apply { currState => 
+        val (a, nextState) = stateTrans(currState); f(a).stateTrans(nextState) }
+        
+  def run = stateTrans(initState)._1
+}
+
+
+trait ReaderMonadCompanion[M[_], State] extends MonadCompanion[M] {
[…]

### 2012-11-16 src/scala/src/test/scala/dot/TestMonads.scala

@@ -1,0 +1,30 @@
+package dot
+
+import org.scalatest._
+
+import scala.collection.mutable.HashMap
+
+object IntStateMonadCompanion extends StateMonadCompanion[IntStateMonad, HashMap[Int, Int]] {
+  def apply[A](fun: HashMap[Int, Int] => (A, HashMap[Int, Int])): IntStateMonad[A]  = new IntStateMonad(HashMap.empty, fun)
+}
+
+class IntStateMonad[A](override val initState: HashMap[Int, Int], override val stateTrans: HashMap[Int, Int] => (A, HashMap[Int, Int])) extends StateMonad[A, IntStateMonad, HashMap[Int, Int]] {
+  implicit override val companion = IntStateMonadCompanion
+}
+
+class TestMonads extends Suite {
+  import IntStateMonadCompanion._
+
+  type M[X] = IntStateMonad[X]
+
+  def fib(n: Int): M[Int] = n match {
+    case 0 => 0
+    case 1 => 1
+    case _ => for(r1 <- fib(n-1);
+                  r2 <- fib(n-2)) yield r1+r2
+  }
+
+  def testFib() = {
+    expect(144)(fib(12).run)
+  }
+}

### 2012-11-16 src/scala/src/test/scala/dot/TestMonads.scala
make the test run in eclipse

@@ -1,6 +1,8 @@
 package dot
 
 import org.scalatest._
+import org.junit.runner.RunWith
+import org.scalatest.junit.JUnitRunner
 
 import scala.collection.mutable.HashMap
 
@@ -12,6 +14,7 @@ class IntStateMonad[A](override val initState: HashMap[Int, Int], override val s
   implicit override val companion = IntStateMonadCompanion
 }
 
+@RunWith(classOf[JUnitRunner])
 class TestMonads extends Suite {
   import IntStateMonadCompanion._
 

### 2012-11-17 src/scala/src/main/scala/dot/BindingSyntax.scala
nominal binding syntax

@@ -1,0 +1,142 @@
+package dot
+
+trait AbstractBindingSyntax {
+  type ContainsBinders[T] 
+
+  trait Scoped[T] { self: \\[T] =>
+    def unabs: (Name, T)
+    def map[U: ContainsBinders](f: T => U): \\[U] = {
+      val (n, b) = unabs
+      \\(n, f(b)) // TODO: make a version where f gets to see the new binder?
+    }
+  }
+
+  type \\[T] <: Scoped[T] 
+  val \\ : ScopedCompanion
+  trait ScopedCompanion {
+    def apply[T: ContainsBinders](binder: Name, body: T): \\[T]
+    def unapply[T](scrutinee: \\[T]): Option[(Name, T)]
+  }
+  
+  type Name 
+  val Name : NameCompanion
+  trait NameCompanion {
+    def apply(s: String): Name
+  }
+
+//  implicit def listBinders[T: ContainsBinders]: ContainsBinders[List[T]]
+//  implicit def pairBinders[T: ContainsBinders, U: ContainsBinders]: ContainsBinders[(T, U)]
+
+  trait Substable[To, Res] {
+    type From[T] = T => Substable[To, Res] // for use as context-bound
+    def subst(from: Name, to: To): Res
+  }
+
+  // TODO: generalize similar to collect
+  implicit def listIsSubstable[T: Substable[To, Res]#From, To, Res](in: List[T]): Substable[To, List[Res]] = new Substable[To, List[Res]] {
+    def subst(from: Name, to: To): List[Res] = in map (_.subst(from, to))
+  }
+
+  implicit def pairIsSubstable[A: Substable[To, A2]#From, B: Substable[To, B2]#From, To, A2, B2](in: (A, B)): Substable[To, (A2, B2)] = new Substable[To, (A2, B2)] {
+    def subst(from: Name, to: To): (A2, B2) = (in._1 subst(from, to), in._2 subst(from, to))
+  }
+}
+
+trait NominalBindingSyntax extends AbstractBindingSyntax with Equalities {
+  type ContainsBinders[T] = T => Nominal[T]
+
+  trait Nominal[Self] {
+    def swap(a: Name, b: Name): Self
+    def fresh(a: Name): Boolean
+  }
+
+  def AtomicIsNominal[T](self: T): Nominal[T] = new Nominal[T] {
+    def swap(a: Name, b: Name) = self
+    def fresh(a: Name) = true
+  }
+
[…]

### 2012-11-17 src/scala/src/main/scala/dot/Equalities.scala

@@ -1,0 +1,17 @@
+package dot
+
+trait Equalities {
+  type ChecksEquality[T] = T => Equality[T]
+
+  trait Equality[-T] {
+    def ===(a: T): Boolean
+  }
+
+  implicit def OptionEq[T : ChecksEquality](self: Option[T]) = new Equality[Option[T]] {
+    def ===(other: Option[T]): Boolean = (for( s <- self; o <- other) yield s === o) getOrElse(true)
+  }
+
+  implicit def ListEq[T : ChecksEquality](self: List[T]) = new Equality[List[T]] {
+    def ===(other: List[T]): Boolean = (self, other).zipped.forall(_ === _) 
+  }
+}

### 2012-11-17 src/scala/src/test/scala/dot/TestBindingSyntax.scala

@@ -1,0 +1,45 @@
+package dot
+
+import org.scalatest._
+import org.junit.runner.RunWith
+import org.scalatest.junit.JUnitRunner
+
+trait Lambda extends AbstractBindingSyntax {
+  sealed abstract class Term
+  case class Var(n: Name) extends Term
+  case class App(fun: Term, arg: Term) extends Term
+  case class Lam(abs: \\[Term]) extends Term
+}
+
+trait NominalLambda extends Lambda with NominalBindingSyntax { self: Lambda =>
+  implicit val termHasBinders: ContainsBinders[Term] = (tm: Term) => new Nominal[Term] {
+    def swap(a: Name, b: Name): Term = {
+      tm match {
+	case Var(n) => Var(n swap(a, b))
+	case App(fun, arg) => App(fun swap(a, b), arg swap(a, b))
+	case Lam(abs) => Lam(abs swap(a, b))
+      }
+    }
+    def fresh(a: Name): Boolean = {
+      tm match {
+	case Var(n) => n fresh(a)
+	case App(fun, arg) => fun.fresh(a) && arg.fresh(a)
+	case Lam(abs) => abs fresh(a)
+      }
+    }
+  }
+}
+
+@RunWith(classOf[JUnitRunner])
+class TestBindingSyntax extends Suite {
+  def testAlphaEquivalence() = {
+    new NominalLambda {
+      val x = Name("x")
+      val y = Name("y")
+      val z = Name("z")
+      expect(true)(Lam(x\\Var(y)) == Lam(z\\Var(y)))
+      expect(false)(Lam(x\\Var(y)) == Lam(y\\Var(x)))
+      expect(true)(Lam(x\\Var(x)) == Lam(y\\Var(y)))
+    }
+  }
+}

### 2012-11-17 src/scala/src/main/scala/dot/BindingParsers.scala
binding parsers

@@ -1,0 +1,25 @@
+package dot
+
+import scala.util.parsing.combinator.Parsers
+import scala.collection.immutable.Map
+
+trait BindingParsers extends Parsers with AbstractBindingSyntax {
+  type BindingParser <: BindingParserCore
+  def BindingParser(env: Map[String, Name]): BindingParser
+  
+  class BindingParserCore(env: Map[String, Name]) {
+    type BindResult = (BindingParser, Name)
+    
+    def bind(nameParser: Parser[String]): Parser[BindResult] = nameParser ^^ {name =>
+      val n=Name(name)
+      (BindingParser(env.updated(name, n)), n)
+    }
+    
+    def bound(nameParser: Parser[String]): Parser[Name]  = nameParser ^?(env, {name => "Unbound variable: "+name})
+    
+    def under[T : ContainsBinders](binder: BindResult)(p: BindingParser => Parser[T]): Parser[\\[T]] = {
+      val (ctx, n) = binder
+      p(ctx) ^^ (\\(n, _))
+    }
+  }
+}

### 2012-11-17 src/scala/src/test/scala/dot/TestBindingParsers.scala

@@ -1,0 +1,62 @@
+package dot
+
+import org.scalatest._
+import org.junit.runner.RunWith
+import org.scalatest.junit.JUnitRunner
+
+import scala.util.parsing.combinator.syntactical.StdTokenParsers
+import scala.util.parsing.combinator.lexical.StdLexical
+import scala.collection.immutable._
+import util.parsing.combinator.{PackratParsers, ImplicitConversions}
+
+trait LambdaParsing extends StdTokenParsers with BindingParsers with PackratParsers with LambdaNominalSyntax with ImplicitConversions { theParser =>
+  type Tokens = StdLexical; val lexical = new StdLexical
+  lexical.delimiters ++= List("\\",".", "(", ")")
+
+  type P[T] = PackratParser[T]
+
+  def l[T](p: => Parser[T])(name: String): P[T] = Parser{ in =>
+    val r = p(in)
+    r
+  }
+  def BindingParser(env: Map[String, Name]): BindingParser = new BindingParser(env)
+  class BindingParser(env: Map[String, Name]) extends BindingParserCore(env) {
+    lazy val termSingle: P[Term] = (
+      l(bound(ident) ^^ {case x => Var(x)}) ("var") |
+      l("\\" ~> bind(ident) >> {x => "." ~> under(x)(_.term)} ^^ {case abs => Lam(abs)}) ("lam") |
+      l("(" ~> term <~ ")") ("paren")
+    )
+    lazy val term = chainl1(termSingle, l(success(App(_, _))) ("app"))
+  }
+  object Parser extends BindingParser(HashMap.empty)
+}
+
+@RunWith(classOf[JUnitRunner])
+class TestBindingParsers extends Suite with LambdaParsing {
+  def parse(in: String) = phrase(Parser.term)(new lexical.Scanner(in))
+
+  val x = Name("x")
+  val y = Name("y")
+  val z = Name("z")
+
+  def ok(expected: Term)(in: String) = parse(in) match {
+    case Success(actual, _) => expect(expected)(actual)
+    case _@r => fail("expected success, got " + r)
+  }
+
+  def bad(expectedMsg: String)(in: String) = parse(in) match {
+    case Failure(msg, _) => expect(expectedMsg)(msg)
[…]

### 2012-11-17 src/scala/src/test/scala/dot/TestBindingSyntax.scala

@@ -4,14 +4,14 @@ import org.scalatest._
 import org.junit.runner.RunWith
 import org.scalatest.junit.JUnitRunner
 
-trait Lambda extends AbstractBindingSyntax {
+trait LambdaSyntax extends AbstractBindingSyntax {
   sealed abstract class Term
   case class Var(n: Name) extends Term
   case class App(fun: Term, arg: Term) extends Term
   case class Lam(abs: \\[Term]) extends Term
 }
 
-trait NominalLambda extends Lambda with NominalBindingSyntax { self: Lambda =>
+trait LambdaNominalSyntax extends LambdaSyntax with NominalBindingSyntax { self: LambdaSyntax =>
   implicit val termHasBinders: ContainsBinders[Term] = (tm: Term) => new Nominal[Term] {
     def swap(a: Name, b: Name): Term = {
       tm match {
@@ -31,15 +31,13 @@ trait NominalLambda extends Lambda with NominalBindingSyntax { self: Lambda =>
 }
 
 @RunWith(classOf[JUnitRunner])
-class TestBindingSyntax extends Suite {
+class TestBindingSyntax extends Suite with LambdaNominalSyntax {
+  val x = Name("x")
+  val y = Name("y")
+  val z = Name("z")
   def testAlphaEquivalence() = {
-    new NominalLambda {
-      val x = Name("x")
-      val y = Name("y")
-      val z = Name("z")
-      expect(true)(Lam(x\\Var(y)) == Lam(z\\Var(y)))
-      expect(false)(Lam(x\\Var(y)) == Lam(y\\Var(x)))
-      expect(true)(Lam(x\\Var(x)) == Lam(y\\Var(y)))
-    }
+    expect(true)(Lam(x\\Var(y)) == Lam(z\\Var(y)))
+    expect(false)(Lam(x\\Var(y)) == Lam(y\\Var(x)))
+    expect(true)(Lam(x\\Var(x)) == Lam(y\\Var(y)))
   }
 }

### 2012-11-17 src/scala/src/test/scala/dot/TestBindingSyntax.scala
added test for substitution

@@ -30,8 +30,21 @@ trait LambdaNominalSyntax extends LambdaSyntax with NominalBindingSyntax { self:
   }
 }
 
+trait LambdaSubstitution extends LambdaNominalSyntax {
+  implicit def scopedIsTermSubstable[T](in: \\[T])(implicit wSubs: T => Substable[Term, T], wNom: T => Nominal[T]): Substable[Term, \\[T]] = scopedIsSubstable[T, Term, T](in)
+
+  implicit def termIsSubstable(in: Term): Substable[Term, Term] = new Substable[Term, Term] {
+    def subst(from: Name, to: Term): Term = in match {
+      case Var(`from`) => to
+      case Var(_) => in
+      case App(fun, arg) => App(fun subst(from, to), arg subst(from, to))
+      case Lam(abs) => Lam(abs subst(from, to))
+    }
+  }
+}
+
 @RunWith(classOf[JUnitRunner])
-class TestBindingSyntax extends Suite with LambdaNominalSyntax {
+class TestBindingSyntax extends Suite with LambdaNominalSyntax with LambdaSubstitution {
   val x = Name("x")
   val y = Name("y")
   val z = Name("z")
@@ -39,5 +52,9 @@ class TestBindingSyntax extends Suite with LambdaNominalSyntax {
     expect(true)(Lam(x\\Var(y)) == Lam(z\\Var(y)))
     expect(false)(Lam(x\\Var(y)) == Lam(y\\Var(x)))
     expect(true)(Lam(x\\Var(x)) == Lam(y\\Var(y)))
+  }
+  def testSubstitution() = {
+    expect(Lam(y\\Var(z)))(Lam(y\\Var(x)) subst(x, Var(z)))
+    expect(Lam(x\\Var(x)))(Lam(x\\Var(x)) subst(x, Var(z)))
   }
 }

### 2012-11-17 src/scala/src/main/scala/dot/TyperMonad.scala
typer monad

@@ -1,0 +1,247 @@
+package dot
+
+import collection.immutable.HashMap
+import collection.Traversable
+
+trait TyperMonad extends AbstractBindingSyntax {
+  // results of the monad
+  sealed abstract class Result[A] extends Monad[A, Result] {
+    object companion extends MonadCompanion[Result]{
+      implicit def result[A](x: A) = Success(x)
+    }
+    def filter(pred: A => Boolean): Result[A]     
+    def mapOutZero[M[_], B](f: A => M[B])(z: Result[B] => M[B]): M[B]
+  } 
+  
+//  def Success[A](r: A): Success[A] = Success(r, List())
+  
+  // TODO: generalise res to deal with multiple results (Stream / List)
+  case class Success[A](res: A) extends Result[A] {
+    def >>=[B](next: A => Result[B]) = next(res) //concat (more map ((mo: () => Result[A]) => () => mo() >>= next))
+    def filter(pred: A => Boolean): Result[A] = if(pred(res)) this else Failure("filtered out")
+    def mapOutZero[M[_], B](f: A => M[B])(z: Result[B] => M[B]): M[B] = f(res)
+  }
+  
+  case class Failure[A](msg: String) extends Result[A] {
+    def >>=[B](next: A => Result[B]) = Failure[B](msg)
+    def filter(pred: A => Boolean): Result[A] = this
+    def mapOutZero[M[_], B](f: A => M[B])(z: Result[B] => M[B]): M[B] = z(Failure[B](msg))
+  }
+    
+  object TyperMonad extends MonadCompanion[TyperMonad] {
+    def apply[A](f: From => To[A]) = new TyperMonad[A]{def apply(from: From) = f(from)}
+    
+    // inject x into the monad
+    implicit def result[A](x: A) = TyperMonad{_ mapEnvTo(_y => Success(x))}
+    def results[A](x: Iterable[A]): TyperMonad[A] = TyperMonad{ case From(st, env) => To(x.toStream map {x => (st, Success(x))})}
+    def fail[A](msg: String) = TyperMonad{_ mapEnvTo(_y => Failure[A](msg))}
+
+    def check(p: => Boolean) = result() filter(x => p)
+
+    // chain the computations yielded by f(x) for all x in xs, so that, if f(x) is a failure for some x in xs,
+    // the computation produced by forall fails; the computation succeeds iff, for all x in xs, f(x) succeeds
[…]

### 2012-11-17 src/scala/src/test/scala/dot/TestBindingParsers.scala

@@ -11,7 +11,7 @@ import util.parsing.combinator.{PackratParsers, ImplicitConversions}
 
 trait LambdaParsing extends StdTokenParsers with BindingParsers with PackratParsers with LambdaNominalSyntax with ImplicitConversions { theParser =>
   type Tokens = StdLexical; val lexical = new StdLexical
-  lexical.delimiters ++= List("\\",".", "(", ")")
+  lexical.delimiters ++= List("\\",".", "(", ")", ":", "->", "*")
 
   type P[T] = PackratParser[T]
 
@@ -21,12 +21,16 @@ trait LambdaParsing extends StdTokenParsers with BindingParsers with PackratPars
   }
   def BindingParser(env: Map[String, Name]): BindingParser = new BindingParser(env)
   class BindingParser(env: Map[String, Name]) extends BindingParserCore(env) {
-    lazy val termSingle: P[Term] = (
+    lazy val term1: P[Term] = (
       l(bound(ident) ^^ {case x => Var(x)}) ("var") |
-      l("\\" ~> bind(ident) >> {x => "." ~> under(x)(_.term)} ^^ {case abs => Lam(abs)}) ("lam") |
+      l("\\" ~> bind(ident) >> {x => ":" ~> ty ~("." ~> under(x)(_.term))} ^^ {case ty~abs => Lam(ty, abs)}) ("lam") |
       l("(" ~> term <~ ")") ("paren")
     )
-    lazy val term = chainl1(termSingle, l(success(App(_, _))) ("app"))
+    lazy val term = chainl1(term1, l(success(App(_, _))) ("app"))
+
+    lazy val ty: P[Type] = (
+      l(ty ~ ("->" ~> ty) ^^ {case ty1~ty2 => Fun(ty1, ty2)}) ("arrow type")) |
+      l(("*" | ident) ^^ {case base => T(base)}) ("base type")
   }
   object Parser extends BindingParser(HashMap.empty)
 }
@@ -38,6 +42,7 @@ class TestBindingParsers extends Suite with LambdaParsing {
   val x = Name("x")
   val y = Name("y")
   val z = Name("z")
+  val ty = T("*")
 
[…]

### 2012-11-17 src/scala/src/test/scala/dot/TestBindingSyntax.scala

@@ -8,7 +8,11 @@ trait LambdaSyntax extends AbstractBindingSyntax {
   sealed abstract class Term
   case class Var(n: Name) extends Term
   case class App(fun: Term, arg: Term) extends Term
-  case class Lam(abs: \\[Term]) extends Term
+  case class Lam(ty: Type, abs: \\[Term]) extends Term
+
+  abstract class Type
+  case class T(base: String) extends Type
+  case class Fun(ty1: Type, ty2: Type) extends Type
 }
 
 trait LambdaNominalSyntax extends LambdaSyntax with NominalBindingSyntax { self: LambdaSyntax =>
@@ -17,14 +21,14 @@ trait LambdaNominalSyntax extends LambdaSyntax with NominalBindingSyntax { self:
       tm match {
 	case Var(n) => Var(n swap(a, b))
 	case App(fun, arg) => App(fun swap(a, b), arg swap(a, b))
-	case Lam(abs) => Lam(abs swap(a, b))
+	case Lam(ty, abs) => Lam(ty, abs swap(a, b))
       }
     }
     def fresh(a: Name): Boolean = {
       tm match {
 	case Var(n) => n fresh(a)
 	case App(fun, arg) => fun.fresh(a) && arg.fresh(a)
-	case Lam(abs) => abs fresh(a)
+	case Lam(ty, abs) => abs fresh(a)
       }
     }
   }
@@ -38,7 +42,7 @@ trait LambdaSubstitution extends LambdaNominalSyntax {
       case Var(`from`) => to
       case Var(_) => in
       case App(fun, arg) => App(fun subst(from, to), arg subst(from, to))
-      case Lam(abs) => Lam(abs subst(from, to))
+      case Lam(ty, abs) => Lam(ty, abs subst(from, to))
     }
   }
 }
@@ -48,13 +52,15 @@ class TestBindingSyntax extends Suite with LambdaNominalSyntax with LambdaSubsti
   val x = Name("x")
   val y = Name("y")
   val z = Name("z")
+  val ty = T("*")
+
   def testAlphaEquivalence() = {
-    expect(true)(Lam(x\\Var(y)) == Lam(z\\Var(y)))
-    expect(false)(Lam(x\\Var(y)) == Lam(y\\Var(x)))
[…]

### 2012-11-17 src/scala/src/test/scala/dot/TestTyperMonad.scala

@@ -1,0 +1,71 @@
+package dot
+
+import org.scalatest._
+import org.junit.runner.RunWith
+import org.scalatest.junit.JUnitRunner
+
+trait LambdaTyper extends StandardTyperMonad with LambdaTyperSyntax with LambdaNominalSyntax {
+  import TyperMonad._
+
+  type State = AbsState
+  val initState = AbsState(0)
+  // state that is carried along
+  case class AbsState(currTag: Int) {
+    def next = AbsState(currTag+1)
+  }
+
+  def ofT(tm: Term, pt: Expected[Type]): TyperMonad[Unit] = {
+    println("type of " + tm + ":" + pt)
+    tm match {
+      case Var(x) => for(
+	r <- lookup(x);
+	_ <- pt := r) yield ()
+      case App(fun, arg) => for(
+	ty1 <- Infer[Type]("ty1").toMetaVar(MetaType);
+	ty2 <- Infer[Type]("ty2").toMetaVar(MetaType);
+	_ <- ofT(fun, Check(Fun(ty1, ty2)));
+	ty1 <- !ty1;
+	ty2 <- !ty2;
+	_ <- ofT(arg, Check(ty1));
+        _ <- pt := ty2) yield ()
+      case Lam(ty1, \\(x, b)) => for(
+	ety2 <- Infer[Type]("ty2");
+	_ <- assume(x, ty1)(ofT(b, ety2));
+	ty2 <- !ety2;
+	_ <- pt := Fun(ty1, ty2)) yield ()
+    }
+  }
+  def typecheck(tm: Term): Result[Type] = (for(
+    ein <- Infer[Type]("in");
+    _ <- ofT(tm, ein);
+    in <- !ein) yield in).run getOrElse Failure("")
+}
+
+trait LambdaTyperSyntax extends MetaVariablesNominal with LambdaSyntax {
+  implicit object MetaType extends MetaVarBuilder[Type, MetaType]("metaTp") {
+    def apply(n: String) = new MetaType(n)
+  }
+  class MetaType(override val name: String) extends Type with MetaVar[Type]
+  implicit def eqType(tp1: Type): Equality[Type] = new Equality[Type] {
+    def ===(tp2: Type) = (tp1, tp2) match {
+      case (x1: MetaType, _) => x1 === tp2
+      case (_, x2: MetaType) => x2 === tp1
[…]

### 2012-11-17 src/scala/src/main/scala/dot/TyperMonad.scala
put debug printing under a flag

@@ -4,6 +4,9 @@ import collection.immutable.HashMap
 import collection.Traversable
 
 trait TyperMonad extends AbstractBindingSyntax {
+  val debugMode: Boolean = true
+  def debug(msg: String) { if (debugMode) println(msg) }
+
   // results of the monad
   sealed abstract class Result[A] extends Monad[A, Result] {
     object companion extends MonadCompanion[Result]{
@@ -119,7 +122,7 @@ trait MetaVariables extends AbstractBindingSyntax with TyperMonad with Equalitie
   }
   class Infer[A : ChecksEquality](val name: String) extends Expected[A] {
     def :=(o: A): TyperMonad[Unit] = {
-      println("infer " + this + " is " + o)
+      debug("infer " + this + " is " + o)
       TyperMonad.result(res=Some(o))
     }
     def unary_! = res map {TyperMonad.result(_)} getOrElse TyperMonad.fail("could not infer "+name)
@@ -134,10 +137,10 @@ trait MetaVariables extends AbstractBindingSyntax with TyperMonad with Equalitie
 
     def :=(o: A): TyperMonad[Unit] =
       if(proto === o) {
-        println("check: "+ proto +" == "+ o)
+        debug("check: "+ proto +" == "+ o)
         TyperMonad.result(())
       } else {
-        println("check: "+ proto +" != "+ o)
+        debug("check: "+ proto +" != "+ o)
         TyperMonad.fail("type checking failed")
       }
 
@@ -175,7 +178,7 @@ trait MetaVariables extends AbstractBindingSyntax with TyperMonad with Equalitie
 
     def unify(o: T): Boolean = {
       assert(v.isEmpty) // or unify silently when o == v.get ?
-      println("unified "+this+" to "+o)
+      debug("unified "+this+" to "+o)
       v = unifies(o)
       !v.isEmpty
     }
@@ -240,8 +243,8 @@ trait StandardTyperMonad extends TyperMonad with MetaVariables {
 
   def lookup(n: Name): TyperMonad[Type] = getGamma >>= {
     _.get(n) match {
-    case Some(t) => println("lu"+t); TyperMonad.result(t)
-    case None => TyperMonad.fail("")
+    case Some(t) => debug("looked up "+t); TyperMonad.result(t)
+    case None => TyperMonad.fail("unbound " + n)
     }
   }
 }

### 2012-11-17 src/scala/src/test/scala/dot/TestTyperMonad.scala

@@ -6,6 +6,7 @@ import org.scalatest.junit.JUnitRunner
 
 trait LambdaTyper extends StandardTyperMonad with LambdaTyperSyntax with LambdaNominalSyntax {
   import TyperMonad._
+  override val debugMode = false
 
   type State = AbsState
   val initState = AbsState(0)
@@ -15,7 +16,7 @@ trait LambdaTyper extends StandardTyperMonad with LambdaTyperSyntax with LambdaN
   }
 
   def ofT(tm: Term, pt: Expected[Type]): TyperMonad[Unit] = {
-    println("type of " + tm + ":" + pt)
+    debug("type of " + tm + ":" + pt)
     tm match {
       case Var(x) => for(
 	r <- lookup(x);

### 2012-11-19 src/scala/src/main/scala/dot/DotSyntax.scala
dot syntax

@@ -1,0 +1,224 @@
+package dot
+
+trait DotSyntax extends AbstractBindingSyntax { syntax =>
+  sealed trait Level {
+    type Classifies <: Level
+  }
+  object Levels {
+    trait TypeBounds extends Level { type Classifies = Type }
+    trait Type extends Level { type Classifies = Term }
+    trait ClassType extends Type
+    trait AbstractType extends Type
+    trait Term extends Level { type Classifies = Nothing }
+    trait ArrowType extends Level { type Classifies = Method }
+    trait Method extends Level { type Classifies = Nothing }
+  }  
+
+  sealed abstract class Label[+T <: Level](val name: String)
+  sealed abstract class TypeLabel(name: String, val isConcrete: Boolean) extends Label[Levels.Type](name)
+  case class ConcreteTypeLabel(override val name: String) extends TypeLabel(name, true)
+  case class AbstractTypeLabel(override val name: String) extends TypeLabel(name, false)
+  object TypeLabel {
+    def unapply(tyl : TypeLabel): Option[(String, Boolean)] = Some((tyl.name, tyl.isConcrete))
+  }
+  case class ValueLabel(override val name: String) extends Label[Levels.Term](name)
+  case class MethodLabel(override val name: String) extends Label[Levels.Method](name)
+
+  sealed trait Entity {
+    type Level <: syntax.Level
+  }
+
+  trait Term extends Entity {
+    type Level = Levels.Term
+    def isPath = false
+  }
+
+  case class Method(body: \\[Term]) extends Entity {
+    type Level = Levels.Method
+  }
+
+  trait Type extends Entity {
+    type Level = Levels.Type
+    def isConcrete = false
+  }
+
+  sealed abstract class TypePair(val lo: Type, val hi: Type) extends Entity
+  case class TypeBounds(override val lo: Type, override val hi: Type) extends TypePair(lo, hi) {
+    type Level = Levels.TypeBounds
+  }
+  case class ArrowType(val paramType: Type, val bodyType: Type) extends TypePair(paramType, bodyType) {
+    type Level = Levels.ArrowType
+  }
+
+  object Members {
+    type Dcl = Decl[Level, Entity]
[…]

### 2012-11-19 src/scala/src/test/scala/dot/TestDotSyntax.scala

@@ -1,0 +1,32 @@
+package dot
+
+import org.scalatest._
+import org.junit.runner.RunWith
+import org.scalatest.junit.JUnitRunner
+
+@RunWith(classOf[JUnitRunner])
+class TestDotSyntax extends Suite with DotNominalSyntax {
+  import Terms._
+  import Types._
+  import Members._
+
+  val x = Name("x")
+  val y = Name("y")
+  val z = Name("z")
+
+  def testAlphaEquivalence() = {
+    expect(false)(Var(x) == Var(y))
+    expect(true)(New(Top, x\\(Defs(List()), Var(x))) == New(Top, z\\(Defs(List()), Var(z))))
+    expect(true)(Refine(Top, x\\(Decls(List()))) == Refine(Top, z\\Decls(List())))
+
+    expect(true)(
+      New(Top, x\\(Defs(List(MethodDef(MethodLabel("m"), Method(y\\Msel(Var(x), MethodLabel("m"), Var(y)))))), Var(z)))
+      ==
+      New(Top, y\\(Defs(List(MethodDef(MethodLabel("m"), Method(x\\Msel(Var(y), MethodLabel("m"), Var(x)))))), Var(z))))
+
+    expect(false)(
+      New(Top, x\\(Defs(List(MethodDef(MethodLabel("m"), Method(y\\Msel(Var(x), MethodLabel("m"), Var(y)))))), Var(z)))
+      ==
+      New(Top, z\\(Defs(List(MethodDef(MethodLabel("m"), Method(x\\Msel(Var(z), MethodLabel("m"), Var(x)))))), Var(z))))
+  }
+}

### 2012-11-19 src/scala/src/test/scala/dot/TestTyperMonad.scala
no need for extra state in the test app for the typer monad

@@ -8,12 +8,8 @@ trait LambdaTyper extends StandardTyperMonad with LambdaTyperSyntax with LambdaN
   import TyperMonad._
   override val debugMode = false
 
-  type State = AbsState
-  val initState = AbsState(0)
-  // state that is carried along
-  case class AbsState(currTag: Int) {
-    def next = AbsState(currTag+1)
-  }
+  type State = Option[Nothing]
+  val initState = None
 
   def ofT(tm: Term, pt: Expected[Type]): TyperMonad[Unit] = {
     debug("type of " + tm + ":" + pt)

### 2012-11-19 src/scala/src/main/scala/dot/Debug.scala
substitution

@@ -1,0 +1,6 @@
+package dot
+
+trait Debug {
+  val debugMode: Boolean = true
+  def debug(msg: String) { if (debugMode) println(msg) }
+}

### 2012-11-19 src/scala/src/main/scala/dot/DotSyntax.scala

@@ -222,3 +222,83 @@ trait DotNominalSyntax extends DotSyntax with NominalBindingSyntax { self: DotSy
     }
   }
 }
+
+trait DotSubstitution extends DotNominalSyntax {
+  import Terms._
+  import Types._
+  import Members._
+
+  implicit def scopedIsTermSubstable[T](in: \\[T])(implicit wSubs: T => Substable[Term, T], wNom: T => Nominal[T]): Substable[Term, \\[T]] = scopedIsSubstable[T, Term, T](in)
+  implicit def listIsTermSubstable[T: Substable[Term, T]#From](in: List[T]): Substable[Term, List[T]] = listIsSubstable[T, Term, T](in)
+
+  implicit def termIsSubstable(in: Term): Substable[Term, Term] = new Substable[Term, Term] {
+    def subst(from: Name, to: Term): Term = in match {
+      case Var(`from`) => to
+      case Var(_) => in
+      case Sel(obj, l) => Sel(obj subst(from, to), l)
+      case Msel(obj, m, arg) => Msel(obj subst(from, to), m, arg subst(from, to))
+      case New(tpe, args_scope) => New(tpe subst(from, to), args_scope subst(from, to))
+    }
+  }
+
+  implicit def typeIsSubstable(in: Type): Substable[Term, Type] = new Substable[Term, Type] {
+    def subst(from: Name, to: Term): Type = in match {
+      case Tsel(obj, l) => Tsel(obj subst(from, to), l)
+      case Refine(parent, decls) => Refine(parent subst(from, to), decls subst(from, to))
+      case Intersect(a, b) => Intersect(a subst(from, to), b subst(from, to))
+      case Union(a, b) => Union(a subst(from, to), b subst(from, to))
+      case Top => Top
+      case Bottom => Bottom
+    }
+  }
+
+  implicit def typeBoundsIsSubstable(in: TypeBounds): Substable[Term, TypeBounds] = new Substable[Term, TypeBounds] {
+    def subst(from: Name, to: Term): TypeBounds = in match {
+      case TypeBounds(lo, hi) => TypeBounds(lo subst(from, to), hi subst(from, to))
+    }
+  }
+
+  implicit def arrowTypeIsSubstable(in: ArrowType): Substable[Term, ArrowType] = new Substable[Term, ArrowType] {
+    def subst(from: Name, to: Term): ArrowType = in match {
[…]

### 2012-11-19 src/scala/src/main/scala/dot/TyperMonad.scala

@@ -3,10 +3,7 @@ package dot
 import collection.immutable.HashMap
 import collection.Traversable
 
-trait TyperMonad extends AbstractBindingSyntax {
-  val debugMode: Boolean = true
-  def debug(msg: String) { if (debugMode) println(msg) }
-
+trait TyperMonad extends AbstractBindingSyntax with Debug {
   // results of the monad
   sealed abstract class Result[A] extends Monad[A, Result] {
     object companion extends MonadCompanion[Result]{

### 2012-11-19 src/scala/src/test/scala/dot/TestDotSyntax.scala

@@ -5,7 +5,7 @@ import org.junit.runner.RunWith
 import org.scalatest.junit.JUnitRunner
 
 @RunWith(classOf[JUnitRunner])
-class TestDotSyntax extends Suite with DotNominalSyntax {
+class TestDotSyntax extends Suite with DotNominalSyntax with DotSubstitution {
   import Terms._
   import Types._
   import Members._
@@ -28,5 +28,14 @@ class TestDotSyntax extends Suite with DotNominalSyntax {
       New(Top, x\\(Defs(List(MethodDef(MethodLabel("m"), Method(y\\Msel(Var(x), MethodLabel("m"), Var(y)))))), Var(z)))
       ==
       New(Top, z\\(Defs(List(MethodDef(MethodLabel("m"), Method(x\\Msel(Var(z), MethodLabel("m"), Var(x)))))), Var(z))))
+  }
+
+  def testSubstitution() = {
+    expect(Var(x))(Var(y) subst(y, Var(x)))
+    expect {
+      New(Top, z\\(Defs(List(ValueDef(ValueLabel("l"), Var(x)))), Sel(Var(z), ValueLabel("l"))))
+    } {
+      New(Top, y\\(Defs(List(ValueDef(ValueLabel("l"), Var(z)))), Sel(Var(y), ValueLabel("l")))) subst(z, Var(x))
+    }
   }
 }

### 2012-11-19 src/scala/src/main/scala/dot/DotSyntax.scala
getting started with dot typer: check bare equalities

@@ -16,7 +16,7 @@ trait DotSyntax extends AbstractBindingSyntax { syntax =>
 
   sealed abstract class Label[+T <: Level](val name: String)
   sealed abstract class TypeLabel(name: String, val isConcrete: Boolean) extends Label[Levels.Type](name)
-  case class ConcreteTypeLabel(override val name: String) extends TypeLabel(name, true)
+  case class ClassLabel(override val name: String) extends TypeLabel(name, true)
   case class AbstractTypeLabel(override val name: String) extends TypeLabel(name, false)
   object TypeLabel {
     def unapply(tyl : TypeLabel): Option[(String, Boolean)] = Some((tyl.name, tyl.isConcrete))
@@ -32,6 +32,9 @@ trait DotSyntax extends AbstractBindingSyntax { syntax =>
     type Level = Levels.Term
     def isPath = false
   }
+  object Path {
+    def unapply(t: Term): Option[Term] = if(t.isPath) Some(t) else None
+  }
 
   case class Method(body: \\[Term]) extends Entity {
     type Level = Levels.Method
@@ -41,14 +44,14 @@ trait DotSyntax extends AbstractBindingSyntax { syntax =>
     type Level = Levels.Type
     def isConcrete = false
   }
+  object ClassType {
+    def unapply(tp: Type): Option[Type] = if(tp.isConcrete) Some(tp) else None
+  }
 
-  sealed abstract class TypePair(val lo: Type, val hi: Type) extends Entity
-  case class TypeBounds(override val lo: Type, override val hi: Type) extends TypePair(lo, hi) {
-    type Level = Levels.TypeBounds
-  }
-  case class ArrowType(val paramType: Type, val bodyType: Type) extends TypePair(paramType, bodyType) {
-    type Level = Levels.ArrowType
-  }
+  sealed abstract class TypePair[Level](val lo: Type, val hi: Type) extends Entity
+
+  case class TypeBounds(override val lo: Type, override val hi: Type) extends TypePair[Levels.TypeBounds](lo, hi)
+  case class ArrowType(val paramType: Type, val bodyType: Type) extends TypePair[Levels.ArrowType](paramType, bodyType)
 
   object Members {
     type Dcl = Decl[Level, Entity]
@@ -58,10 +61,10 @@ trait DotSyntax extends AbstractBindingSyntax { syntax =>
     case class MethodDecl(override val l: MethodLabel, override val cls: ArrowType) extends Decl(l, cls)
[…]

### 2012-11-19 src/scala/src/main/scala/dot/DotTyper.scala

@@ -1,0 +1,90 @@
+package dot
+
+trait Constraints extends StandardTyperMonad {
+  override type State = Set[Constraint]
+  override val initState = Set[Constraint]()
+
+  def addConstraint(c: Constraint): TyperMonad[Unit] = TyperMonad{from =>
+    from.mapStateTo({state => state + c}, {state => Success(())})
+  }
+
+  implicit def constraintExpected(tp1: Expected[Type]) = new {
+    def <:<(tp2: Expected[Type]): Constraint = new SubtypeConstraint(tp1, tp2)
+  }
+
+  trait Constraint
+  case class SubtypeConstraint(tv1: Expected[Type], tv2: Expected[Type]) extends Constraint
+}
+
+trait Typer extends StandardTyperMonad with Constraints with DotTyperSyntax with DotNominalSyntax with DotSubstitution {
+  import Terms._
+  import Types._
+  import Members._
+  import TyperMonad._
+}
+
+trait DotTyperSyntax extends MetaVariablesNominal with DotSyntax {
+  implicit def eqEntity(e1: Entity): Equality[Entity] = new Equality[Entity] {
+    def ===(e2: Entity) = (e1, e2) match {
+      case (a: TypeBounds, b: TypeBounds) => a === b
+      case (a: Type, b: Type) => a === b
+      case (a: Term, b: Term) => a === b
+      case (a: ArrowType, b: ArrowType) => a === b
+      case (a: Method, b: Method) => a === b
+      case _ => false
+    }
+  }
+
+  implicit def eqType(tp1: Type): Equality[Type] = new Equality[Type] {
+    import Types._
+    def ===(tp2: Type) = (tp1, tp2) match {
+      case (Tsel(o1, l1), Tsel(o2, l2)) => o1===o2 && l1===l2
+      case (Refine(parent1, decls1), Refine(parent2, decls2)) => parent1===parent2 && decls1===decls2
+      case (Intersect(a1, b1), Intersect(a2, b2)) => a1===a2 && b1===b2
+      case (Union(a1, b1), Union(a2, b2)) => a1===a2 && b1===b2
+      case (Top,Top) => true
+      case (Bottom,Bottom) => true
+      case _ => false
+    }
+  }
+
[…]

### 2012-11-19 src/scala/src/main/scala/dot/Equalities.scala

@@ -14,4 +14,8 @@ trait Equalities {
   implicit def ListEq[T : ChecksEquality](self: List[T]) = new Equality[List[T]] {
     def ===(other: List[T]): Boolean = (self, other).zipped.forall(_ === _) 
   }
+
+  implicit def PairEq[A : ChecksEquality, B: ChecksEquality](self: (A,B)) = new Equality[(A,B)] {
+    def ===(other: (A,B)): Boolean = self._1===other._1 && self._2===other._2
+  }
 }

### 2012-11-19 src/scala/src/main/scala/dot/DotSyntax.scala
baby step...

@@ -59,7 +59,8 @@ trait DotSyntax extends AbstractBindingSyntax { syntax =>
     case class TypeDecl(override val l: TypeLabel, override val cls: TypeBounds) extends Decl(l, cls)
     case class ValueDecl(override val l: ValueLabel, override val cls: Type) extends Decl(l, cls)
     case class MethodDecl(override val l: MethodLabel, override val cls: ArrowType) extends Decl(l, cls)
-    case class Decls(decls: List[Dcl])
+    abstract class Dcls
+    case class Decls(decls: List[Dcl]) extends Dcls
 
     type Defn = Def[Level, Entity]
     sealed abstract class Def[+L <: Level, +E <: Entity](val l: Label[L], val rhs: E)
@@ -87,7 +88,7 @@ trait DotSyntax extends AbstractBindingSyntax { syntax =>
       assert(o.isPath)
       override def isConcrete = label.isConcrete
     }
-    case class Refine(parent: Type, decls: \\[Members.Decls]) extends Type {
+    case class Refine(parent: Type, decls: \\[Members.Dcls]) extends Type {
       override def isConcrete = parent.isConcrete
     }
     case class Intersect(a: Type, b: Type) extends Type {
@@ -205,9 +206,9 @@ trait DotNominalSyntax extends DotSyntax with NominalBindingSyntax { self: DotSy
     }
   }
 
-  implicit def declsHasBinders: ContainsBinders[Members.Decls] = (ds: Members.Decls) => new Nominal[Members.Decls] {
+  implicit def dclsHasBinders: ContainsBinders[Members.Dcls] = (ds: Members.Dcls) => new Nominal[Members.Dcls] {
     import Members._
-    def swap(a: Name, b: Name): Decls = ds match {
+    def swap(a: Name, b: Name): Dcls = ds match {
       case Decls(ds) => Decls(ds swap(a, b))
     }
     def fresh(a: Name): Boolean = ds match {
@@ -293,8 +294,8 @@ trait DotSubstitution extends DotNominalSyntax {
     }
   }
 
-  implicit def declsIsSubstable(in: Decls): Substable[Term, Decls] = new Substable[Term, Decls] {
-    def subst(from: Name, to: Term): Decls = in match {
+  implicit def dclsIsSubstable(in: Dcls): Substable[Term, Dcls] = new Substable[Term, Dcls] {
+    def subst(from: Name, to: Term): Dcls = in match {
       case Decls(ds) => Decls(ds subst(from, to))
     }
   }

### 2012-11-19 src/scala/src/main/scala/dot/DotTyper.scala

@@ -16,14 +16,67 @@ trait Constraints extends StandardTyperMonad {
   case class SubtypeConstraint(tv1: Expected[Type], tv2: Expected[Type]) extends Constraint
 }
 
-trait Typer extends StandardTyperMonad with Constraints with DotTyperSyntax with DotNominalSyntax with DotSubstitution {
+trait DotTyper extends StandardTyperMonad with Constraints with DotTyperSyntax with DotNominalSyntax with DotSubstitution {
   import Terms._
   import Types._
   import Members._
   import TyperMonad._
+
+  override val debugMode = true
+
+  def typecheck(tm: Term): Result[Type] = (for(
+    ein <- Infer[Type]("in");
+    _ <- ofT(tm, ein);
+    in <- !ein) yield in).run getOrElse Failure("")
+
+  def ofT(tm: Term, pt: Expected[Type]): TyperMonad[Unit] = {
+    debug("type of " + tm + ":" + pt)
+    tm match {
+      case Var(x) => for(
+	r <- lookup(x);
+	_ <- pt := r) yield ()
+      //case Sel(o, l) => TODO
+      //case Msel(o, m, a) => TODO
+      case New(tc, \\(x, (args, b))) => for(
+	// TODO: complete stub
+	_ <- assume(x, tc){for(
+	  _ <- ofT(b, pt)) yield ()}) yield ()
+    }
+  }
 }
 
 trait DotTyperSyntax extends MetaVariablesNominal with DotSyntax {
+  implicit object MetaClassLabel extends MetaVarBuilder[ClassLabel, MetaClassLabel]("metaLc") {
+    def apply(n: String) = new MetaClassLabel(n)
+  }
+  class MetaClassLabel(override val name: String) extends ClassLabel(name) with MetaVar[ClassLabel]
+
+  implicit object MetaAbstractTypeLabel extends MetaVarBuilder[AbstractTypeLabel, MetaAbstractTypeLabel]("metaL") {
+    def apply(n: String) = new MetaAbstractTypeLabel(n)
+  }
+  class MetaAbstractTypeLabel(override val name: String) extends AbstractTypeLabel(name) with MetaVar[AbstractTypeLabel]
+
+  implicit object MetaValueLabel extends MetaVarBuilder[ValueLabel, MetaValueLabel]("metal") {
+    def apply(n: String) = new MetaValueLabel(n)
+  }
[…]

### 2012-11-19 src/scala/src/test/scala/dot/TestDotTyper.scala

@@ -1,0 +1,19 @@
+package dot
+
+import org.scalatest._
+import org.junit.runner.RunWith
+import org.scalatest.junit.JUnitRunner
+
+@RunWith(classOf[JUnitRunner])
+class TestDotTyper extends Suite with DotTyper {
+  import Terms._
+  import Types._
+  import Members._
+
+  val x = Name("x")
+  val nodefs = Defs(List())
+
+  def testTC1() = {
+    expect(Success(Top))(typecheck(New(Top,x\\(nodefs, Var(x)))))
+  }
+}

### 2012-11-19 src/scala/src/main/scala/dot/DotSyntax.scala
more baby steps

@@ -61,6 +61,7 @@ trait DotSyntax extends AbstractBindingSyntax { syntax =>
     case class MethodDecl(override val l: MethodLabel, override val cls: ArrowType) extends Decl(l, cls)
     abstract class Dcls
     case class Decls(decls: List[Dcl]) extends Dcls
+    case object BottomDecls extends Dcls
 
     type Defn = Def[Level, Entity]
     sealed abstract class Def[+L <: Level, +E <: Entity](val l: Label[L], val rhs: E)
@@ -88,7 +89,7 @@ trait DotSyntax extends AbstractBindingSyntax { syntax =>
       assert(o.isPath)
       override def isConcrete = label.isConcrete
     }
-    case class Refine(parent: Type, decls: \\[Members.Dcls]) extends Type {
+    case class Refine(parent: Type, decls: \\[Members.Decls]) extends Type {
       override def isConcrete = parent.isConcrete
     }
     case class Intersect(a: Type, b: Type) extends Type {
@@ -206,9 +207,9 @@ trait DotNominalSyntax extends DotSyntax with NominalBindingSyntax { self: DotSy
     }
   }
 
-  implicit def dclsHasBinders: ContainsBinders[Members.Dcls] = (ds: Members.Dcls) => new Nominal[Members.Dcls] {
+  implicit def declsHasBinders: ContainsBinders[Members.Decls] = (ds: Members.Decls) => new Nominal[Members.Decls] {
     import Members._
-    def swap(a: Name, b: Name): Dcls = ds match {
+    def swap(a: Name, b: Name): Decls = ds match {
       case Decls(ds) => Decls(ds swap(a, b))
     }
     def fresh(a: Name): Boolean = ds match {
@@ -294,8 +295,8 @@ trait DotSubstitution extends DotNominalSyntax {
     }
   }
 
-  implicit def dclsIsSubstable(in: Dcls): Substable[Term, Dcls] = new Substable[Term, Dcls] {
-    def subst(from: Name, to: Term): Dcls = in match {
+  implicit def declsIsSubstable(in: Decls): Substable[Term, Decls] = new Substable[Term, Decls] {
+    def subst(from: Name, to: Term): Decls = in match {
       case Decls(ds) => Decls(ds subst(from, to))
     }
   }

### 2012-11-19 src/scala/src/main/scala/dot/DotTyper.scala

@@ -24,10 +24,14 @@ trait DotTyper extends StandardTyperMonad with Constraints with DotTyperSyntax w
 
   override val debugMode = true
 
-  def typecheck(tm: Term): Result[Type] = (for(
-    ein <- Infer[Type]("in");
-    _ <- ofT(tm, ein);
-    in <- !ein) yield in).run getOrElse Failure("")
+  def typecheck(tm: Term): Result[Type] = {
+    val r = (for(
+      ein <- Infer[Type]("in");
+      _ <- ofT(tm, ein);
+      in <- !ein) yield in).findAll
+    assert(r.length==1)
+    r.head
+  }
 
   def ofT(tm: Term, pt: Expected[Type]): TyperMonad[Unit] = {
     debug("type of " + tm + ":" + pt)
@@ -40,42 +44,40 @@ trait DotTyper extends StandardTyperMonad with Constraints with DotTyperSyntax w
       case New(tc, \\(x, (args, b))) => for(
 	// TODO: complete stub
 	_ <- assume(x, tc){for(
+	  ds <- expand(x, tc);
+	  _ <- wfCtorMems(ds, args);
 	  _ <- ofT(b, pt)) yield ()}) yield ()
     }
+  }
+
+  def wfCtorMems(ds: Dcls, args: Defs): TyperMonad[Unit] = ds match {
+    case BottomDecls => fail("bottom expansion for constructor type")
+    case Decls(ds) => forall(ds) {
+      case TypeDecl(l, TypeBounds(s, u)) => sub(s, u)
+      // TODO...
+    }
+  }
+
+  def expand(x: Name, tp: Type): TyperMonad[Dcls] = tp match {
+    case Top => Decls(List())
+    case Bottom => BottomDecls
+    // TODO
+    case Refine(parent, \\(z, decls)) if parent===Top => decls swap(z, x)
+  }
+
+  def sub(tp1: Type, tp2: Type): TyperMonad[Unit] = {
+    if (tp2 === Top) ()
+    else if (tp1 === Bottom) ()
+    else fail(tp1 + " is not a subtype of " + tp2)
+    // TODO
   }
 }
 
 trait DotTyperSyntax extends MetaVariablesNominal with DotSyntax {
-  implicit object MetaClassLabel extends MetaVarBuilder[ClassLabel, MetaClassLabel]("metaLc") {
-    def apply(n: String) = new MetaClassLabel(n)
-  }
[…]

### 2012-11-19 src/scala/src/test/scala/dot/TestDotTyper.scala

@@ -11,9 +11,23 @@ class TestDotTyper extends Suite with DotTyper {
   import Members._
 
   val x = Name("x")
+  val y = Name("y")
+  val z = Name("z")
   val nodefs = Defs(List())
 
-  def testTC1() = {
-    expect(Success(Top))(typecheck(New(Top,x\\(nodefs, Var(x)))))
-  }
+  def testTC1() = expect(Success(Top))(typecheck(New(Top,x\\(nodefs, Var(x)))))
+
+  def testTC2() =
+    expect {
+      Success(Refine(Top, z\\Decls(List(TypeDecl(AbstractTypeLabel("L"), TypeBounds(Bottom, Top))))))
+    } {
+      typecheck(New(Refine(Top, z\\Decls(List(TypeDecl(AbstractTypeLabel("L"), TypeBounds(Bottom, Top))))), x\\(nodefs, Var(x))))
+    }
+
+  def testTC2BadBounds() =
+    expect {
+      Failure("Top is not a subtype of Bottom")
+    } {
+      typecheck(New(Refine(Top, z\\Decls(List(TypeDecl(AbstractTypeLabel("L"), TypeBounds(Top, Bottom))))), x\\(nodefs, Var(x))))
+    }
 }

### 2012-11-19 src/scala/src/main/scala/dot/DotTyper.scala
meet and join of decl lattice

@@ -58,6 +58,34 @@ trait DotTyper extends StandardTyperMonad with Constraints with DotTyperSyntax w
     }
   }
 
+  def mergeDecl[L <: Level, E <: Entity](funlo: (Type, Type) => Type, funhi: (Type, Type) => Type)(
+    d1: Decl[L,E], d2: Decl[L,E]): Dcl = (d1, d2) match {
+      case (TypeDecl(l1, TypeBounds(s1, u1)), TypeDecl(l2, TypeBounds(s2, u2))) if l1===l2 =>
+	TypeDecl(l1, TypeBounds(funlo(s1, s2), funhi(u1, u2)))
+      case (MethodDecl(l1, ArrowType(s1, u1)), MethodDecl(l2, ArrowType(s2, u2))) if l1===l2 =>
+	MethodDecl(l1, ArrowType(funlo(s1, s2), funhi(u1, u2)))
+      case (ValueDecl(l1, u1), ValueDecl(l2, u2)) if l1===l2 =>
+	ValueDecl(l1, funhi(u1, u2))
+    }
+  def intersectDecl = mergeDecl(Union, Intersect) _
+  def unionDecl = mergeDecl(Intersect, Union) _
+  //  intersection of declaration sets
+  def meet(ds1: Dcls, ds2: Dcls): Dcls = (ds1, ds2) match {
+    case (BottomDecls, _) => BottomDecls
+    case (_, BottomDecls) => BottomDecls
+    case (Decls(ds1), Decls(ds2)) => Decls(
+      ds1.flatMap{d1 => ds2.find{d2 => d1.l===d2.l}.map{d2 => intersectDecl(d1, d2)}} ++
+      ds1.filterNot{d1 => ds2.exists{d2 => d1.l===d2.l}} ++
+      ds2.filterNot{d2 => ds1.exists{d1 => d1.l===d2.l}})
+  }
+  // union of declaration sets
+  def join(ds1: Dcls, ds2: Dcls): Dcls = (ds1, ds2) match {
+    case (BottomDecls, other) => other
+    case (other, BottomDecls) => other
+    case (Decls(ds1), Decls(ds2)) => Decls(
+      ds1.flatMap{d1 => ds2.find{d2 => d1.l===d2.l}.map{d2 => unionDecl(d1, d2)}})
+  }
+
   def expand(x: Name, tp: Type): TyperMonad[Dcls] = tp match {
     case Top => Decls(List())
     case Bottom => BottomDecls

### 2012-11-19 src/scala/src/test/scala/dot/TestDotTyper.scala

@@ -30,4 +30,38 @@ class TestDotTyper extends Suite with DotTyper {
     } {
       typecheck(New(Refine(Top, z\\Decls(List(TypeDecl(AbstractTypeLabel("L"), TypeBounds(Top, Bottom))))), x\\(nodefs, Var(x))))
     }
+
+  val decls1 = Decls(List(
+    TypeDecl(AbstractTypeLabel("L1"), TypeBounds(Tsel(Var(x), AbstractTypeLabel("S1")), Tsel(Var(x), AbstractTypeLabel("U1")))),
+    TypeDecl(AbstractTypeLabel("L"), TypeBounds(Tsel(Var(x), AbstractTypeLabel("S1")), Tsel(Var(x), AbstractTypeLabel("U1")))),
+    TypeDecl(AbstractTypeLabel("L'"), TypeBounds(Tsel(Var(x), AbstractTypeLabel("S1'")), Tsel(Var(x), AbstractTypeLabel("U1'"))))))
+  val decls2 = Decls(List(
+    TypeDecl(AbstractTypeLabel("L2"), TypeBounds(Tsel(Var(x), AbstractTypeLabel("S2")), Tsel(Var(x), AbstractTypeLabel("U2")))),
+    TypeDecl(AbstractTypeLabel("L'"), TypeBounds(Tsel(Var(x), AbstractTypeLabel("S2'")), Tsel(Var(x), AbstractTypeLabel("U2'")))),
+    TypeDecl(AbstractTypeLabel("L"), TypeBounds(Tsel(Var(x), AbstractTypeLabel("S2")), Tsel(Var(x), AbstractTypeLabel("U2"))))))
+
+  def testMeet1() =
+    expect {
+      Decls(List(
+	TypeDecl(AbstractTypeLabel("L"), TypeBounds(Union(Tsel(Var(x), AbstractTypeLabel("S1")), Tsel(Var(x), AbstractTypeLabel("S2"))),
+						    Intersect(Tsel(Var(x), AbstractTypeLabel("U1")), Tsel(Var(x), AbstractTypeLabel("U2"))))),
+	TypeDecl(AbstractTypeLabel("L'"), TypeBounds(Union(Tsel(Var(x), AbstractTypeLabel("S1'")), Tsel(Var(x), AbstractTypeLabel("S2'"))),
[…]

### 2012-11-19 src/scala/src/main/scala/dot/DotTyper.scala
flesh out expansion

@@ -87,10 +87,20 @@ trait DotTyper extends StandardTyperMonad with Constraints with DotTyperSyntax w
   }
 
   def expand(x: Name, tp: Type): TyperMonad[Dcls] = tp match {
+    case Refine(parent, \\(z, ds)) =>
+      for (dsp <- expand(x, parent))
+      yield meet(dsp, ds swap(z, x))
+    case Intersect(a, b) =>
+      for (dsa <- expand(x, a);
+	   dsb <- expand(x, b))
+      yield meet(dsa, dsb)
+    case Union(a, b) =>
+      for (dsa <- expand(x, a);
+	   dsb <- expand(x, b))
+      yield join(dsa, dsb)
+    /*case Tsel(p, l) => // TODO */
     case Top => Decls(List())
     case Bottom => BottomDecls
-    // TODO
-    case Refine(parent, \\(z, decls)) if parent===Top => decls swap(z, x)
   }
 
   def sub(tp1: Type, tp2: Type): TyperMonad[Unit] = {

### 2012-11-19 src/scala/src/test/scala/dot/TestDotTyper.scala

@@ -63,5 +63,4 @@ class TestDotTyper extends Suite with DotTyper {
     } {
       join(decls1, decls2)
     }
-
 }

### 2012-11-19 src/scala/src/main/scala/dot/DotTyper.scala
pseudo freshness handling

@@ -1,6 +1,6 @@
 package dot
 
-trait Constraints extends StandardTyperMonad {
+trait Constraints extends StandardTyperMonad with NominalBindingSyntax {
   override type State = Set[Constraint]
   override val initState = Set[Constraint]()
 
@@ -8,15 +8,24 @@ trait Constraints extends StandardTyperMonad {
     from.mapStateTo({state => state + c}, {state => Success(())})
   }
 
-  implicit def constraintExpected(tp1: Expected[Type]) = new {
+  implicit def subtypingConstraintExpected(tp1: Expected[Type]) = new {
     def <:<(tp2: Expected[Type]): Constraint = new SubtypeConstraint(tp1, tp2)
+  }
+
+  implicit def freshnessConstraintExpected[T: ContainsBinders](tp: Expected[T]) = new {
+    def fresh(x: Name): Constraint = new FreshnessConstraint(x, tp)
   }
 
   trait Constraint
   case class SubtypeConstraint(tv1: Expected[Type], tv2: Expected[Type]) extends Constraint
+  case class FreshnessConstraint[T: ContainsBinders](x: Name, nom: Expected[T]) extends Constraint
 }
 
-trait DotTyper extends StandardTyperMonad with Constraints with DotTyperSyntax with DotNominalSyntax with DotSubstitution {
+trait DotTyper extends StandardTyperMonad with DotTyperSyntax with DotNominalSyntax with DotSubstitution {
+  // NOTE: don't use constraints for now (baby steps)
+  override type State = Option[Nothing]
+  override val initState = None
+
   import Terms._
   import Types._
   import Members._
@@ -46,7 +55,9 @@ trait DotTyper extends StandardTyperMonad with Constraints with DotTyperSyntax w
 	_ <- assume(x, tc){for(
 	  ds <- expand(x, tc);
 	  _ <- wfCtorMems(ds, args);
-	  _ <- ofT(b, pt)) yield ()}) yield ()
+	  _ <- ofT(b, pt);
+	  tp <- !pt;
+	  if tp fresh(x)) yield ()}) yield ()
     }
   }
 

### 2012-11-19 src/scala/src/main/scala/dot/DotTyper.scala
really give up on constraints for now... just know how to generate unambiguous friendly names

@@ -1,30 +1,11 @@
 package dot
 
-trait Constraints extends StandardTyperMonad with NominalBindingSyntax {
-  override type State = Set[Constraint]
-  override val initState = Set[Constraint]()
-
-  def addConstraint(c: Constraint): TyperMonad[Unit] = TyperMonad{from =>
-    from.mapStateTo({state => state + c}, {state => Success(())})
-  }
-
-  implicit def subtypingConstraintExpected(tp1: Expected[Type]) = new {
-    def <:<(tp2: Expected[Type]): Constraint = new SubtypeConstraint(tp1, tp2)
-  }
-
-  implicit def freshnessConstraintExpected[T: ContainsBinders](tp: Expected[T]) = new {
-    def fresh(x: Name): Constraint = new FreshnessConstraint(x, tp)
-  }
-
-  trait Constraint
-  case class SubtypeConstraint(tv1: Expected[Type], tv2: Expected[Type]) extends Constraint
-  case class FreshnessConstraint[T: ContainsBinders](x: Name, nom: Expected[T]) extends Constraint
-}
-
 trait DotTyper extends StandardTyperMonad with DotTyperSyntax with DotNominalSyntax with DotSubstitution {
-  // NOTE: don't use constraints for now (baby steps)
-  override type State = Option[Nothing]
-  override val initState = None
+  override type State = Int
+  override val initState = 0
+  def tag: TyperMonad[String] = TyperMonad{from =>
+    from.mapStateTo({state => state + 1}, {state => Success("$tag$" + state.toString)})
+  }
 
   import Terms._
   import Types._

### 2012-11-19 src/scala/src/main/scala/dot/DotTyper.scala
fresh name hack in dot typer

@@ -3,9 +3,10 @@ package dot
 trait DotTyper extends StandardTyperMonad with DotTyperSyntax with DotNominalSyntax with DotSubstitution {
   override type State = Int
   override val initState = 0
-  def tag: TyperMonad[String] = TyperMonad{from =>
-    from.mapStateTo({state => state + 1}, {state => Success("$tag$" + state.toString)})
+  def tag: TyperMonad[Int] = TyperMonad{from =>
+    from.mapStateTo({state => state + 1}, {state => Success(state)})
   }
+  def freshName(n: String): TyperMonad[Name] = for (s <- tag) yield (Name(n+"$tag$"+s))
 
   import Terms._
   import Types._

### 2012-11-19 src/scala/src/main/scala/dot/DotSyntax.scala
Attempt at mem. Compiles with warnings. Untested.

@@ -59,9 +59,20 @@ trait DotSyntax extends AbstractBindingSyntax { syntax =>
     case class TypeDecl(override val l: TypeLabel, override val cls: TypeBounds) extends Decl(l, cls)
     case class ValueDecl(override val l: ValueLabel, override val cls: Type) extends Decl(l, cls)
     case class MethodDecl(override val l: MethodLabel, override val cls: ArrowType) extends Decl(l, cls)
-    abstract class Dcls
-    case class Decls(decls: List[Dcl]) extends Dcls
-    case object BottomDecls extends Dcls
+    abstract class Dcls {
+      def findByLabel(l: Label[_]): Option[Dcl]
+    }
+    case class Decls(decls: List[Dcl]) extends Dcls {
+      override def findByLabel(l: Label[_]) = decls.find(d => d.l==l)
+    }
+    case object BottomDecls extends Dcls {
+      val s = Types.Top; val u = Types.Bottom
+      override def findByLabel(l: Label[_]) = Some(l match {
+	case (l: TypeLabel) => TypeDecl(l, TypeBounds(s, u))
+	case (l: MethodLabel) => MethodDecl(l, ArrowType(s, u))
+	case (l: ValueLabel) => ValueDecl(l, u)
+      })
+    }
 
     type Defn = Def[Level, Entity]
     sealed abstract class Def[+L <: Level, +E <: Entity](val l: Label[L], val rhs: E)
@@ -226,6 +237,23 @@ trait DotNominalSyntax extends DotSyntax with NominalBindingSyntax { self: DotSy
       case Defs(ds) => ds.fresh(a)
     }
   }
+
+  val entityHasBinders: ContainsBinders[Entity] = (e: Entity) => new Nominal[Entity] {
+    def swap(a: Name, b: Name): Entity = e match {
+      case (e: TypeBounds) => typeBoundsHasBinders(e) swap(a, b)
+      case (e: Type) => typeHasBinders(e) swap(a, b)
+      case (e: Term) => termHasBinders(e) swap(a, b)
+      case (e: ArrowType) => arrowTypeHasBinders(e) swap(a, b)
+      case (e: Method) => methodHasBinders(e) swap(a, b)
+    }
+    def fresh(a: Name): Boolean = e match {
+      case (e: TypeBounds) => typeBoundsHasBinders(e) fresh(a)
+      case (e: Type) => typeHasBinders(e) fresh(a)
[…]

### 2012-11-19 src/scala/src/main/scala/dot/DotTyper.scala

@@ -79,6 +79,17 @@ trait DotTyper extends StandardTyperMonad with DotTyperSyntax with DotNominalSyn
       ds1.flatMap{d1 => ds2.find{d2 => d1.l===d2.l}.map{d2 => unionDecl(d1, d2)}})
   }
 
+  def mem(tm: Term, d: Dcl): TyperMonad[Unit] = for(
+    etp <- Infer[Type]("memtp");
+    _ <- ofT(tm, etp);
+    tp <- !etp;
+    z <- freshName("z");
+    ds <- expand(z, tp);
+    Some(di) = ds.findByLabel(d.l);
+    if ((entityHasBinders(di.cls).fresh(z) && di.cls===d.cls) ||
+	(tm.isPath && entityIsSubstable(di.cls).subst(z, tm)===d.cls))) yield ()
+
+
   def expand(x: Name, tp: Type): TyperMonad[Dcls] = tp match {
     case Refine(parent, \\(z, ds)) =>
       for (dsp <- expand(x, parent))

### 2012-11-19 src/scala/src/main/scala/dot/DotTyper.scala
better err msg

@@ -85,7 +85,7 @@ trait DotTyper extends StandardTyperMonad with DotTyperSyntax with DotNominalSyn
     tp <- !etp;
     z <- freshName("z");
     ds <- expand(z, tp);
-    Some(di) = ds.findByLabel(d.l);
+    di <- exactlyOne(ds.findByLabel(d.l), "undeclared label " + d.l);
     if ((entityHasBinders(di.cls).fresh(z) && di.cls===d.cls) ||
 	(tm.isPath && entityIsSubstable(di.cls).subst(z, tm)===d.cls))) yield ()
 

### 2012-11-19 src/scala/src/main/scala/dot/DotTyper.scala
mem convenience methods

@@ -79,8 +79,27 @@ trait DotTyper extends StandardTyperMonad with DotTyperSyntax with DotNominalSyn
       ds1.flatMap{d1 => ds2.find{d2 => d1.l===d2.l}.map{d2 => unionDecl(d1, d2)}})
   }
 
+  def memValue(o: Term, l: ValueLabel): TyperMonad[Type] = for(
+    tp <- Infer[Type]("memValTp").toMetaVar(MetaType);
+    _ <- mem(o, ValueDecl(l, tp));
+    tp <- !tp) yield tp
+
+  def memMethod(o: Term, m: MethodLabel): TyperMonad[ArrowType] = for(
+    lo <- Infer[Type]("memMetTp").toMetaVar(MetaType);
+    hi <- Infer[Type]("memMetTp").toMetaVar(MetaType);
+    _ <- mem(o, MethodDecl(m, ArrowType(lo, hi)));
+    lo <- !lo;
+    hi <- !hi) yield ArrowType(lo, hi)
+
+  def memMethod(o: Term, l: TypeLabel): TyperMonad[TypeBounds] = for(
+    lo <- Infer[Type]("memTypTp").toMetaVar(MetaType);
+    hi <- Infer[Type]("memTypTp").toMetaVar(MetaType);
+    _ <- mem(o, TypeDecl(l, TypeBounds(lo, hi)));
+    lo <- !lo;
+    hi <- !hi) yield TypeBounds(lo, hi)
+
   def mem(tm: Term, d: Dcl): TyperMonad[Unit] = for(
-    etp <- Infer[Type]("memtp");
+    etp <- Infer[Type]("memObjTp");
     _ <- ofT(tm, etp);
     tp <- !etp;
     z <- freshName("z");

### 2012-11-19 src/scala/src/main/scala/dot/DotTyper.scala
sel + tests

@@ -30,7 +30,9 @@ trait DotTyper extends StandardTyperMonad with DotTyperSyntax with DotNominalSyn
       case Var(x) => for(
 	r <- lookup(x);
 	_ <- pt := r) yield ()
-      //case Sel(o, l) => TODO
+      case Sel(o, l) => for(
+	tp <- memValue(o, l);
+	_ <- pt := tp) yield ()
       //case Msel(o, m, a) => TODO
       case New(tc, \\(x, (args, b))) => for(
 	// TODO: complete stub
@@ -47,7 +49,12 @@ trait DotTyper extends StandardTyperMonad with DotTyperSyntax with DotNominalSyn
     case BottomDecls => fail("bottom expansion for constructor type")
     case Decls(ds) => forall(ds) {
       case TypeDecl(l, TypeBounds(s, u)) => sub(s, u)
-      // TODO...
+      case ValueDecl(l, u) => for(
+	ValueDef(_, v) <- exactlyOne(args.defs.find(d => d.l===l), "uninitialized value for label " + l);
+	tv <- Infer[Type]("valTp").toMetaVar(MetaType);
+	_ <- ofT(v, Check(tv));
+	tv <- !tv;
+	_ <- sub(tv, u)) yield ()
     }
   }
 

### 2012-11-19 src/scala/src/test/scala/dot/TestDotTyper.scala

@@ -6,6 +6,10 @@ import org.scalatest.junit.JUnitRunner
 
 @RunWith(classOf[JUnitRunner])
 class TestDotTyper extends Suite with DotTyper {
+  def bad[A](r: Result[A]) {
+    assert(r.isInstanceOf[Failure[A]])
+  }
+
   import Terms._
   import Types._
   import Members._
@@ -30,6 +34,26 @@ class TestDotTyper extends Suite with DotTyper {
     } {
       typecheck(New(Refine(Top, z\\Decls(List(TypeDecl(AbstractTypeLabel("L"), TypeBounds(Top, Bottom))))), x\\(nodefs, Var(x))))
     }
+
+  def testTC_Sel() =
+    expect(Success(Top))(typecheck(New(Refine(Top, z\\Decls(List(ValueDecl(ValueLabel("l"), Top)))), x\\(Defs(List(ValueDef(ValueLabel("l"), Var(x)))), Sel(Var(x), ValueLabel("l"))))))
+
+  def testTC_Sel_Bad() =
+    expect {
+      Failure("undeclared label ValueLabel(l')")
+    } {
+      typecheck(New(Refine(Top, z\\Decls(List(ValueDecl(ValueLabel("l"), Top)))), x\\(Defs(List(ValueDef(ValueLabel("l"), Var(x)))), Sel(Var(x), ValueLabel("l'")))))
+    }
+
+  def testTC_Sel_BadInit() =
+    expect {
+      Failure("uninitialized value for label ValueLabel(l)")
+    } {
+      typecheck(New(Refine(Top, z\\Decls(List(ValueDecl(ValueLabel("l"), Top)))), x\\(Defs(List(ValueDef(ValueLabel("l'"), Var(x)))), Sel(Var(x), ValueLabel("l")))))
+    }
+
+  def testTC_Sel_BadInitTC() =
+    bad(typecheck(New(Refine(Top, z\\Decls(List(ValueDecl(ValueLabel("l"), Bottom)))), x\\(Defs(List(ValueDef(ValueLabel("l"), Var(x)))), Sel(Var(x), ValueLabel("l"))))))
 
   val decls1 = Decls(List(
     TypeDecl(AbstractTypeLabel("L1"), TypeBounds(Tsel(Var(x), AbstractTypeLabel("S1")), Tsel(Var(x), AbstractTypeLabel("U1")))),

### 2012-11-19 src/scala/src/main/scala/dot/DotTyper.scala
Got started on Msel.
- Added some debug info.
- Fixed a bug in === for TypePairs!

@@ -33,7 +33,13 @@ trait DotTyper extends StandardTyperMonad with DotTyperSyntax with DotNominalSyn
       case Sel(o, l) => for(
 	tp <- memValue(o, l);
 	_ <- pt := tp) yield ()
-      //case Msel(o, m, a) => TODO
+      case Msel(o, m, a) => for(
+	ArrowType(s, u) <- memMethod(o, m);
+	_ <- pt := u;
+	ta <- Infer[Type]("argTp");
+	_ <- ofT(a, ta);
+	ta <- !ta;
+	_ <- sub(ta, s)) yield ()
       case New(tc, \\(x, (args, b))) => for(
 	// TODO: complete stub
 	_ <- assume(x, tc){for(
@@ -55,6 +61,12 @@ trait DotTyper extends StandardTyperMonad with DotTyperSyntax with DotNominalSyn
 	_ <- ofT(v, Check(tv));
 	tv <- !tv;
 	_ <- sub(tv, u)) yield ()
+      case MethodDecl(l, ArrowType(s, u)) => for(
+	MethodDef(_, Method(\\(x, b))) <- exactlyOne(args.defs.find(d => d.l===l), "uninitialized method for label " + l);
+	tb <- Infer[Type]("bodyTp").toMetaVar(MetaType);
+	_ <- assume(x, s)(ofT(b, Check(tb)));
+	tb <- !tb;
+	_ <- sub(tb, u)) yield ()
     }
   }
 
@@ -105,32 +117,54 @@ trait DotTyper extends StandardTyperMonad with DotTyperSyntax with DotNominalSyn
     lo <- !lo;
     hi <- !hi) yield TypeBounds(lo, hi)
 
-  def mem(tm: Term, d: Dcl): TyperMonad[Unit] = for(
-    etp <- Infer[Type]("memObjTp");
-    _ <- ofT(tm, etp);
-    tp <- !etp;
-    z <- freshName("z");
-    ds <- expand(z, tp);
-    di <- exactlyOne(ds.findByLabel(d.l), "undeclared label " + d.l);
-    if ((entityHasBinders(di.cls).fresh(z) && di.cls===d.cls) ||
-	(tm.isPath && entityIsSubstable(di.cls).subst(z, tm)===d.cls))) yield ()
+  def mem(tm: Term, d: Dcl): TyperMonad[Unit] = {
+    debug("mem? " + tm + " : " + d)
+    val r = for(
[…]

### 2012-11-19 src/scala/src/test/scala/dot/TestDotTyper.scala

@@ -55,6 +55,12 @@ class TestDotTyper extends Suite with DotTyper {
   def testTC_Sel_BadInitTC() =
     bad(typecheck(New(Refine(Top, z\\Decls(List(ValueDecl(ValueLabel("l"), Bottom)))), x\\(Defs(List(ValueDef(ValueLabel("l"), Var(x)))), Sel(Var(x), ValueLabel("l"))))))
 
+  def testTc_Msel() =
+    expect(Success(Top))(typecheck(New(Refine(Top, z\\Decls(List(MethodDecl(MethodLabel("m"), ArrowType(Top, Top))))),
+				       z\\(Defs(List(MethodDef(MethodLabel("m"),
+							       Method(x\\Var(x))))),
+					   Msel(Var(z), MethodLabel("m"), Var(z))))))
+
   val decls1 = Decls(List(
     TypeDecl(AbstractTypeLabel("L1"), TypeBounds(Tsel(Var(x), AbstractTypeLabel("S1")), Tsel(Var(x), AbstractTypeLabel("U1")))),
     TypeDecl(AbstractTypeLabel("L"), TypeBounds(Tsel(Var(x), AbstractTypeLabel("S1")), Tsel(Var(x), AbstractTypeLabel("U1")))),

### 2012-11-19 src/scala/src/test/scala/dot/TestDotTyper.scala
one more msel test case

@@ -55,10 +55,16 @@ class TestDotTyper extends Suite with DotTyper {
   def testTC_Sel_BadInitTC() =
     bad(typecheck(New(Refine(Top, z\\Decls(List(ValueDecl(ValueLabel("l"), Bottom)))), x\\(Defs(List(ValueDef(ValueLabel("l"), Var(x)))), Sel(Var(x), ValueLabel("l"))))))
 
-  def testTc_Msel() =
+  def testTc_Msel1() =
     expect(Success(Top))(typecheck(New(Refine(Top, z\\Decls(List(MethodDecl(MethodLabel("m"), ArrowType(Top, Top))))),
 				       z\\(Defs(List(MethodDef(MethodLabel("m"),
 							       Method(x\\Var(x))))),
+					   Msel(Var(z), MethodLabel("m"), Var(z))))))
+
+  def testTc_Msel2() =
+    expect(Success(Top))(typecheck(New(Refine(Top, z\\Decls(List(MethodDecl(MethodLabel("m"), ArrowType(Top, Top))))),
+				       z\\(Defs(List(MethodDef(MethodLabel("m"),
+							       Method(x\\Msel(Var(z), MethodLabel("m"), Var(x)))))),
 					   Msel(Var(z), MethodLabel("m"), Var(z))))))
 
   val decls1 = Decls(List(

### 2012-11-19 src/scala/src/main/scala/dot/DotSyntax.scala
wf and a few tweaks

@@ -55,7 +55,8 @@ trait DotSyntax extends AbstractBindingSyntax { syntax =>
 
   object Members {
     type Dcl = Decl[Level, Entity]
-    sealed abstract class Decl[+L <: Level, +E <: Entity](val l: Label[L], val cls: E)
+    sealed abstract class DeclOrDef[+L <: Level](val l: Label[L])
+    sealed abstract class Decl[+L <: Level, +E <: Entity](override val l: Label[L], val cls: E) extends DeclOrDef[L](l)
     case class TypeDecl(override val l: TypeLabel, override val cls: TypeBounds) extends Decl(l, cls)
     case class ValueDecl(override val l: ValueLabel, override val cls: Type) extends Decl(l, cls)
     case class MethodDecl(override val l: MethodLabel, override val cls: ArrowType) extends Decl(l, cls)
@@ -63,6 +64,7 @@ trait DotSyntax extends AbstractBindingSyntax { syntax =>
       def findByLabel(l: Label[_]): Option[Dcl]
     }
     case class Decls(decls: List[Dcl]) extends Dcls {
+      assert(uniqLabels(decls))
       override def findByLabel(l: Label[_]) = decls.find(d => d.l==l)
     }
     case object BottomDecls extends Dcls {
@@ -75,10 +77,17 @@ trait DotSyntax extends AbstractBindingSyntax { syntax =>
     }
 
     type Defn = Def[Level, Entity]
-    sealed abstract class Def[+L <: Level, +E <: Entity](val l: Label[L], val rhs: E)
+    sealed abstract class Def[+L <: Level, +E <: Entity](override val l: Label[L], val rhs: E) extends DeclOrDef[L](l)
     case class ValueDef(override val l: ValueLabel, override val rhs: Terms.Value) extends Def(l, rhs)
     case class MethodDef(override val l: MethodLabel, override val rhs: Method) extends Def(l, rhs)
-    case class Defs(defs: List[Defn])
+    case class Defs(defs: List[Defn]) {
+      assert(uniqLabels(defs))
+    }
+
+    def uniqLabels[D <: DeclOrDef[_]](ds: List[D]): Boolean = ds forall{ d1 =>
+      val n = ds.filter(d2 => d1.l==d2.l).size
+      n==1
+    }
   }
 
   object Terms {

### 2012-11-19 src/scala/src/main/scala/dot/DotTyper.scala

@@ -52,7 +52,10 @@ trait DotTyper extends StandardTyperMonad with DotTyperSyntax with DotNominalSyn
   }
 
   def wfCtorMems(ds: Dcls, args: Defs): TyperMonad[Unit] = ds match {
-    case BottomDecls => fail("bottom expansion for constructor type")
+    case BottomDecls =>
+      fail("bottom expansion for constructor type")
+    case Decls(ds) if !args.defs.forall{d1 => ds.exists{d2 => d1.l==d2.l}} =>
+      fail("initialized label is undeclared")
     case Decls(ds) => forall(ds) {
       case TypeDecl(l, TypeBounds(s, u)) => sub(s, u)
       case ValueDecl(l, u) => for(
@@ -110,7 +113,7 @@ trait DotTyper extends StandardTyperMonad with DotTyperSyntax with DotNominalSyn
     lo <- !lo;
     hi <- !hi) yield ArrowType(lo, hi)
 
-  def memMethod(o: Term, l: TypeLabel): TyperMonad[TypeBounds] = for(
+  def memType(o: Term, l: TypeLabel): TyperMonad[TypeBounds] = for(
     lo <- Infer[Type]("memTypTp").toMetaVar(MetaType);
     hi <- Infer[Type]("memTypTp").toMetaVar(MetaType);
     _ <- mem(o, TypeDecl(l, TypeBounds(lo, hi)));
@@ -161,7 +164,10 @@ trait DotTyper extends StandardTyperMonad with DotTyperSyntax with DotNominalSyn
 	for (dsa <- expand(x, a);
 	     dsb <- expand(x, b))
 	yield join(dsa, dsb)
-      /*case Tsel(p, l) => // TODO */
+      case Tsel(p, l) =>
+	for (TypeBounds(s, u) <- memType(p, l);
+	     dsu <- expand(x, u))
+	yield dsu
       case Top => Decls(List())
       case Bottom => BottomDecls
     }
@@ -172,6 +178,39 @@ trait DotTyper extends StandardTyperMonad with DotTyperSyntax with DotNominalSyn
     else if (tp1 === Bottom) ()
     else fail(tp1 + " is not a subtype of " + tp2)
     // TODO
+  }
+
+  def wf(tp: Type): TyperMonad[Unit] = tp match {
+    case Top => ()
+    case Bottom => ()
[…]

### 2012-11-19 src/scala/src/test/scala/dot/TestDotTyper.scala

@@ -49,7 +49,7 @@ class TestDotTyper extends Suite with DotTyper {
     expect {
       Failure("uninitialized value for label ValueLabel(l)")
     } {
-      typecheck(New(Refine(Top, z\\Decls(List(ValueDecl(ValueLabel("l"), Top)))), x\\(Defs(List(ValueDef(ValueLabel("l'"), Var(x)))), Sel(Var(x), ValueLabel("l")))))
+      typecheck(New(Refine(Top, z\\Decls(List(ValueDecl(ValueLabel("l"), Top)))), x\\(nodefs, Sel(Var(x), ValueLabel("l")))))
     }
 
   def testTC_Sel_BadInitTC() =

### 2012-11-19 src/scala/src/main/scala/dot/Debug.scala
Implemented the 'first' TyperMonad combinator.
Also, refactored debug infrastructure for more flexibility.

@@ -1,6 +1,15 @@
 package dot
 
 trait Debug {
-  val debugMode: Boolean = true
+  var debugMode: Boolean = false
   def debug(msg: String) { if (debugMode) println(msg) }
+  def inDebugMode[A](body: => A): A = {
+    val oldDebugMode = debugMode
+    debugMode = true
+    try {
+      body
+    } finally {
+      debugMode = oldDebugMode
+    }
+  }
 }

### 2012-11-19 src/scala/src/main/scala/dot/DotTyper.scala

@@ -13,8 +13,6 @@ trait DotTyper extends StandardTyperMonad with DotTyperSyntax with DotNominalSyn
   import Members._
   import TyperMonad._
 
-  override val debugMode = true
-
   def typecheck(tm: Term): Result[Type] = {
     val r = (for(
       ein <- Infer[Type]("in");
@@ -135,13 +133,13 @@ trait DotTyper extends StandardTyperMonad with DotTyperSyntax with DotNominalSyn
     if (tm.isPath) {
       debug("mem-path");
       for ((z, di) <- r;
-	   _ <- entityIsSubstable(di.cls).subst(z, tm)===d.cls) yield ()
+	   _ <- check(entityIsSubstable(di.cls).subst(z, tm)===d.cls)) yield ()
     } else {
       debug("mem-term");
       (for ((z, di) <- r;
 	    if entityHasBinders(di.cls).fresh(z);
             _ <- debug("mem-term restriction ok");
-	    _ <- di.cls===d.cls) yield ()) ++
+	    _ <- check(di.cls===d.cls)) yield ()) ++
       (for ((z, di) <- r;
 	    if !entityHasBinders(di.cls).fresh(z);
 	    _:Unit <- fail("mem-term restriction fails for " + z + " in " + di)) yield ())
@@ -173,12 +171,10 @@ trait DotTyper extends StandardTyperMonad with DotTyperSyntax with DotNominalSyn
     }
   }
 
-  def sub(tp1: Type, tp2: Type): TyperMonad[Unit] = {
-    if (tp2 === Top) ()
-    else if (tp1 === Bottom) ()
-    else fail(tp1 + " is not a subtype of " + tp2)
-    // TODO
-  }
+  def sub(tp1: Type, tp2: Type): TyperMonad[Unit] = first(List(
+    (for (_ <- check(tp2===Top); _ <- wfe(tp1)) yield ()),
+    (for (_ <- check(tp1===Bottom); _ <- wfe(tp2)) yield ()),
+    fail(tp1 + " is not a subtype of " + tp2)))
 
   def wf(tp: Type): TyperMonad[Unit] = tp match {
     case Top => ()
@@ -212,6 +208,11 @@ trait DotTyper extends StandardTyperMonad with DotTyperSyntax with DotNominalSyn
     case ValueDecl(l, u) => for(
                   _ <- wf(u)) yield ()
   }
+
[…]

### 2012-11-19 src/scala/src/main/scala/dot/TyperMonad.scala

@@ -47,8 +47,21 @@ trait TyperMonad extends AbstractBindingSyntax with Debug {
     def exactlyOne[T](xs: Traversable[T], err: String="required exactly 1 element"): TyperMonad[T] =
       if(xs.size == 1) result(xs.head)
       else fail(err)
-  }
 
+
+    // returns the first computation that succeeds
+    def first[A](ms: List[TyperMonad[A]]): TyperMonad[A] = ms match {
+      case Nil => fail("empty list of alternatives")
+      case m::Nil => m
+      case m::ms =>  TyperMonad{x =>
+	val mx = m(x)
+	mx mapStRes {
+	  case (_, r:Success[A]) => debug("first: success"); mx
+	  case (_, r:Failure[A]) => debug("first: failure"); first(ms)(x)
+	}
+      }
+    }
+  }
   
   // state that is carried along
   type State

### 2012-11-19 src/scala/src/test/scala/dot/TestTyperMonad.scala

@@ -6,7 +6,6 @@ import org.scalatest.junit.JUnitRunner
 
 trait LambdaTyper extends StandardTyperMonad with LambdaTyperSyntax with LambdaNominalSyntax {
   import TyperMonad._
-  override val debugMode = false
 
   type State = Option[Nothing]
   val initState = None

### 2012-11-19 src/scala/src/main/scala/dot/DotTyper.scala
sub except for rfn cases. refactored first into some...
... to give more implementation freedome.

@@ -171,10 +171,19 @@ trait DotTyper extends StandardTyperMonad with DotTyperSyntax with DotNominalSyn
     }
   }
 
-  def sub(tp1: Type, tp2: Type): TyperMonad[Unit] = first(List(
-    (for (_ <- check(tp2===Top); _ <- wfe(tp1)) yield ()),
-    (for (_ <- check(tp1===Bottom); _ <- wfe(tp2)) yield ()),
-    fail(tp1 + " is not a subtype of " + tp2)))
+  def sub(tp1: Type, tp2: Type): TyperMonad[Unit] = some(List(
+    /*refl*/    (for(_ <- check(tp1===tp2); _ <- wfe(tp1))                                                   yield()),
+    /*<:-top*/  (for(_ <- check(tp2===Top); _ <- wfe(tp1))                                                   yield()),
+    /*bot-<:*/  (for(_ <- check(tp1===Bottom); _ <- wfe(tp2))                                                yield()),
+    /*<:-and*/  (for(Intersect(a, b) <- tp2; _ <- sub(tp1, a); _ <- sub(tp1, b))                             yield()),
+    /*or-<:*/   (for(Union(a, b) <- tp1; _ <- sub(a, tp2); _ <- sub(b, tp2))                                 yield()),
+    /*and1-<:*/ (for(Intersect(a, b) <- tp1; _ <- sub(a, tp2); _ <- wfe(b))                                  yield()),
+    /*and2-<:*/ (for(Intersect(a, b) <- tp1; _ <- sub(b, tp2); _ <- wfe(a))                                  yield()),
+    /*<:-or1*/  (for(Union(a, b) <- tp2; _ <- sub(tp1, a); _ <- wfe(b))                                      yield()),
+    /*<:-or2*/  (for(Union(a, b) <- tp2; _ <- sub(tp1, b); _ <- wfe(a))                                      yield()),
+    /*<:-tsel*/ (for(Tsel(p, l) <- tp2; TypeBounds(s, u) <- memType(p, l); _ <- sub(s, u); _ <- sub(tp1, s)) yield()),
+    /*tsel-<:*/ (for(Tsel(p, l) <- tp1; TypeBounds(s, u) <- memType(p, l); _ <- sub(s, u); _ <- sub(u, tp2)) yield())),
+    err=tp1 + " is not a subtype of " + tp2)
 
   def wf(tp: Type): TyperMonad[Unit] = tp match {
     case Top => ()

### 2012-11-19 src/scala/src/main/scala/dot/TyperMonad.scala

@@ -49,15 +49,14 @@ trait TyperMonad extends AbstractBindingSyntax with Debug {
       else fail(err)
 
 
-    // returns the first computation that succeeds
-    def first[A](ms: List[TyperMonad[A]]): TyperMonad[A] = ms match {
-      case Nil => fail("empty list of alternatives")
-      case m::Nil => m
+    // returns some computation that succeeds or else fails with the given error
+    def some[A](ms: List[TyperMonad[A]], err: String = ""): TyperMonad[A] = ms match {
+      case Nil => fail(err)
       case m::ms =>  TyperMonad{x =>
 	val mx = m(x)
 	mx mapStRes {
-	  case (_, r:Success[A]) => debug("first: success"); mx
-	  case (_, r:Failure[A]) => debug("first: failure"); first(ms)(x)
+	  case (_, r:Success[A]) => mx
+	  case (_, r:Failure[A]) => some(ms, err)(x)
 	}
       }
     }

### 2012-12-03 src/scala/src/main/scala/dot/DotShell.scala
loose wf and expansion in Scala model.

@@ -87,7 +87,7 @@ trait DotShellWithSugar extends DotShell with DotSugarFunctions {
 // ===> y : ⊤ { y ⇒ l: ⊤ }
 // dot> y.l
 // ===> y.l : ⊤
-object sh extends DotShellWithSugar with REPL {
+object sh extends DotShellWithSugar with LooseDotTyper with REPL {
   override def prompt() = "dot> "
   override def processline(line: String): Unit = exec(line)
 }

### 2012-12-03 src/scala/src/main/scala/dot/DotTyper.scala

@@ -3,6 +3,59 @@ package dot
 import scala.util.parsing.input.Positional
 import scala.util.parsing.input.NoPosition
 
+trait LooseDotTyper extends DotTyper {
+  import Terms._
+  import Types._
+  import Members._
+  import TyperMonad._
+
+  override def wf(tp: Type): TyperMonad[Unit] = tp match {
+    case Tsel(p, l) => for (TypeBounds(s, u) <- memType(p, l)) yield ()
+    case _ => super.wf(tp)
+  }
+
+  override def expand(x: Name, tp: Type): TyperMonad[Dcls] = {
+    expandIter(Map(), x, tp)
+  }
+
+  def expandIter(m: Map[Tsel, Dcls], x: Name, tp: Type): TyperMonad[Dcls] = {
+    debug("loose expand_" + x + " of " + tp.pp)
+    tp match {
+      case Refine(parent, \\(z, ds)) =>
+        for (dsp <- expandIter(m, x, parent))
+        yield meet(dsp, ds swap(z, x))
+      case Intersect(a, b) =>
+        for (dsa <- expandIter(m, x, a);
+             dsb <- expandIter(m, x, b))
+        yield meet(dsa, dsb)
+      case Union(a, b) =>
+        for (dsa <- expandIter(m, x, a);
+             dsb <- expandIter(m, x, b))
+        yield join(dsa, dsb)
+      case selT@Tsel(p, l) =>
+        m.get(selT) match {
+          case Some(ds) => ds
+          case None =>
+            for(TypeBounds(s, u) <- memType(p, l);
+               ds <- expandFix(selT, Decls(List()), m, x, u))
+            yield ds
+        }
+      case Top => Decls(List())
+      case Bottom => BottomDecls
+    }
+  }
+
+  def expandFix(selT: Tsel, selDs: Dcls, m: Map[Tsel, Dcls], x: Name, tp: Type): TyperMonad[Dcls] = {
+    debug("fix expand_" + x + " of " + tp.pp)
+    some(List(
+    (for (ds <- expandIter(m + (selT -> selDs), x, tp);
+          if ds==selDs) yield ds),
+    (for (ds <- expandIter(m + (selT -> selDs), x, tp);
+          if ds!=selDs;
+          fixDs <- expandFix(selT, ds, m, x, tp)) yield fixDs)))
+  }
+}
+
[…]

### 2012-12-03 src/scala/src/test/scala/dot/TestDot.scala

@@ -5,6 +5,10 @@ import org.junit.runner.RunWith
 import org.scalatest.junit.JUnitRunner
 
 @RunWith(classOf[JUnitRunner])
+class TestLooseDot extends TestDot with LooseDotTyper {
+}
+
+@RunWith(classOf[JUnitRunner])
 class TestDot extends Suite with DotParsing with DotTyper {
   def parse(in: String): Term = phrase(Parser.term)(new lexical.Scanner(in)) match {
     case Success(tm, _) => tm

### 2012-12-03 src/scala/src/test/scala/dot/TestDotTyper.scala

@@ -5,6 +5,10 @@ import org.junit.runner.RunWith
 import org.scalatest.junit.JUnitRunner
 
 @RunWith(classOf[JUnitRunner])
+class TestLooseDotTyper extends TestDotTyper with LooseDotTyper {
+}
+
+@RunWith(classOf[JUnitRunner])
 class TestDotTyper extends Suite with DotTyper {
   def bad[A](r: Result[A]) {
     assert(r.isInstanceOf[TyperFailure[A]])

### 2012-12-04 src/scala/src/main/scala/dot/DotShell.scala
added wf query to dot.sh

@@ -8,18 +8,23 @@ trait DotShell extends DotParsing with DotTyper with DotPrettyPrint { shell =>
   sealed class Line extends Positional
   case class ValDef(x: String, tpe: Type, args: \\[Members.Defs]) extends Line
   case class LineTerm(tm: Term) extends Line
-  
+  case class LineQueryWf(tp: Type) extends Line
+
   var valDefs: List[ValDef] = List.empty
   var typerEnv: Env = initEnv
   var parsingEnv: Map[String, Name] = HashMap.empty
   
+  lexical.delimiters ++= List("?")
+  lexical.reserved ++= List("wf")
   override def BindingParser(envArg: Map[String, Name]) = new ShellParser { val env = envArg }
   trait ShellParser extends BindingParser {
     lazy val valdef: P[ValDef] = l(
       "val" ~> ident >> {xStr => bind(success(xStr)) >> {x => "=" ~> "new" ~> (concrete_typ ~
        under(x){p => p.defs <~ opt(";")}) ^^
        {case tyc~args => ValDef(xStr: String, tyc, args)}}}) ("val def")
-    def line: P[Line] = l(term ^^ LineTerm) ("line term") | valdef
+    lazy val query: P[Line] = wfQuery
+    lazy val wfQuery: P[LineQueryWf] = l("wf" ~> "?" ~> typ ^^ LineQueryWf) ("line query wf")
+    def line: P[Line] = query | l(term ^^ LineTerm) ("line term") | valdef
   }
 
   import Terms._
@@ -40,29 +45,36 @@ trait DotShell extends DotParsing with DotTyper with DotPrettyPrint { shell =>
       vy\\(Defs(List(MethodDef(funApply, Method(vx\\Var(vx))))),
            Msel(Var(vy), funApply, tm)))
   }
-  def toTerm(line: Line): Term = line match {
-    case ValDef(_, tpe, \\(x, args)) => New(tpe, x\\(args, cast(Var(x), Top)))
-    case LineTerm(tm) => tm
-  }
 
   def tc(in: String): String = {
     val line: Line = phrase(BindingParser(parsingEnv).line)(new lexical.Scanner(in)) match {
       case Success(line, _) => line
       case r@_ => return "parse error: " + r
     }
-    val r = typecheck(toTerm(line), Some(typerEnv))
-    val msg = r match {
-      case TyperSuccess(tp) => line match {
-        case ValDef(xStr, tpe, \\(x, args)) =>
[…]

### 2012-12-04 src/scala/src/main/scala/dot/DotTyper.scala

@@ -71,12 +71,10 @@ trait DotTyper extends StandardTyperMonad with DotTyperSyntax with DotNominalSyn
   import TyperMonad._
 
   def typecheck(tm: Term, env: Option[Env] = None): Result[Type] = {
-    val r = (for(
+    (for(
       ein <- Infer[Type]("in");
       _ <- ofT(tm, ein);
-      in <- !ein) yield in).findAll(env)
-    assert(r.length==1, "typecheck result is not unique: "+r.map(_.map(_.pp)).mkString(", "))
-    r.head
+      in <- !ein) yield in).findExactlyOne(env)
   }
 
   def pos(p: Positional, msg: String): String = {

### 2012-12-04 src/scala/src/main/scala/dot/TyperMonad.scala

@@ -93,6 +93,11 @@ trait TyperMonad extends AbstractBindingSyntax with Debug {
     def ++(alt: => TyperMonad[A]): TyperMonad[A] = TyperMonad{x => this(x) ++ alt(x) }
     
     def findAll(env: Option[Env] = None) = this(From(initState, env.getOrElse(initEnv))).tos map {case (st, r) => r} toList
+    def findExactlyOne(env: Option[Env] = None) = {
+      val r = findAll(env)
+      assert(r.length==1, "result is not unique")
+      r.head
+    }
     def run = this(From(initState, initEnv)).tos map {case (st, r) => r} find{case TyperSuccess(_) => true; case _ => false}
   } 
 

### 2012-12-04 src/scala/src/test/scala/dot/TestDotShell.scala

@@ -24,4 +24,7 @@ class TestDotShell extends Suite with DotShell { sh =>
       ("val x = 1", "parse error: [1.9] failure: ``new'' expected but 1 found\n\n" +
                     "val x = 1\n" +
                     "        ^")))
+
+  def test3() = ok(List(
+      ("wf? Top", "===> wf? ⊤ : yes")))
 }

### 2012-12-04 src/scala/src/main/scala/dot/TyperMonad.scala
hack: added stack overflow catcher

@@ -94,9 +94,13 @@ trait TyperMonad extends AbstractBindingSyntax with Debug {
     
     def findAll(env: Option[Env] = None) = this(From(initState, env.getOrElse(initEnv))).tos map {case (st, r) => r} toList
     def findExactlyOne(env: Option[Env] = None) = {
-      val r = findAll(env)
-      assert(r.length==1, "result is not unique")
-      r.head
+      try {
+        val r = findAll(env)
+        assert(r.length==1, "result is not unique")
+        r.head
+      } catch {
+        case eso: StackOverflowError => TyperFailure[A]("stuck in infinite derivation")
+      }
     }
     def run = this(From(initState, initEnv)).tos map {case (st, r) => r} find{case TyperSuccess(_) => true; case _ => false}
   } 

### 2012-12-04 src/scala/src/test/scala/dot/TestDot.scala

@@ -209,4 +209,8 @@ class TestDot extends Suite with DotParsing with DotTyper {
 (val x = new Top { x => L: Top..Top; l: x.L }(l=x); x).l
 """
   )}
+
+  def testTooRecursive() = expect("stuck in infinite derivation"){checkfail(
+      "val x = new Top {x => l: x.L; L: Bot .. x.l.L} (l=x); x")
+  }
 }

### 2012-12-04 src/scala/src/main/scala/dot/DotShell.scala
added expand query to dot.sh

@@ -9,21 +9,25 @@ trait DotShell extends DotParsing with DotTyper with DotPrettyPrint { shell =>
   case class ValDef(x: String, tpe: Type, args: \\[Members.Defs]) extends Line
   case class LineTerm(tm: Term) extends Line
   case class LineQueryWf(tp: Type) extends Line
+  case class LineQueryExpand(x: String, tp: Type) extends Line
 
   var valDefs: List[ValDef] = List.empty
   var typerEnv: Env = initEnv
   var parsingEnv: Map[String, Name] = HashMap.empty
   
   lexical.delimiters ++= List("?")
-  lexical.reserved ++= List("wf")
+  lexical.reserved ++= List("wf", "expand")
   override def BindingParser(envArg: Map[String, Name]) = new ShellParser { val env = envArg }
   trait ShellParser extends BindingParser {
     lazy val valdef: P[ValDef] = l(
       "val" ~> ident >> {xStr => bind(success(xStr)) >> {x => "=" ~> "new" ~> (concrete_typ ~
        under(x){p => p.defs <~ opt(";")}) ^^
        {case tyc~args => ValDef(xStr: String, tyc, args)}}}) ("val def")
-    lazy val query: P[Line] = wfQuery
+    lazy val query: P[Line] = wfQuery | expandQuery
     lazy val wfQuery: P[LineQueryWf] = l("wf" ~> "?" ~> typ ^^ LineQueryWf) ("line query wf")
+    lazy val expandQuery: P[LineQueryExpand] = l(
+        "expand" ~> opt("(" ~> ident <~ ")") ~ ("?" ~> typ) ^^ {
+        case ox~tp => LineQueryExpand(ox.getOrElse("self"), tp)}) ("line query expand")
     def line: P[Line] = query | l(term ^^ LineTerm) ("line term") | valdef
   }
 
@@ -73,6 +77,16 @@ trait DotShell extends DotParsing with DotTyper with DotPrettyPrint { shell =>
             "===> wf? " + tp.pp + " : yes"
           case TyperFailure(msg) =>
             "===> wf? " + tp.pp + " : no, " + msg
+        }
+      case LineQueryExpand(x, tp) =>
+        (for (dcls <- expand(Name(x), tp)) yield (dcls)).findExactlyOne(Some(typerEnv)) match {
+          case TyperSuccess(dcls) =>
+            "===> expand("+x+")? " + tp.pp + " : yes, " + (dcls match {
[…]

### 2012-12-14 src/scala/src/main/scala/dot/DotSugarFunctions.scala
fix lambda return type inference to avoid dependence on parameter causing unbound variables error later

@@ -120,6 +120,8 @@ trait DotSugarFunctions extends DotParsing with DotTyper with DotPrettyPrint {
          eu <- Infer[Type]("lambdaBodyTp");
          _ <- assume(x, s)(ofT(body, eu));
          u <- !eu;
+         _ <- some(List(check(u fresh(x))),
+                   pos(body, "cannot have return type (" + u.pp + ") of λ-body ("+ body.pp + ") path-dependent on λ-parameter " + x.pp));
          _ <- pt := LambdaType(s, u)) yield ()
      case _ => super.ofT(tm, pt)
    }

### 2013-05-29 src/scala/src/main/scala/dot/DotTyper.scala
Fix scoping issue with arg of method init in Scala model.
Reported by @samuelgruetter.

@@ -130,7 +130,7 @@ trait DotTyper extends StandardTyperMonad with DotTyperSyntax with DotNominalSyn
         etb <- Infer[Type]("bodyTp");
         _ <- assume(x, s)(ofT(b, etb));
         tb <- !etb;
-        _ <- sub(tb, u)) yield ()
+        _ <- assume(x, s)(sub(tb, u))) yield ()
     }
   }
 

### 2013-05-29 src/scala/src/test/scala/dot/TestDot.scala

@@ -213,4 +213,17 @@ class TestDot extends FunSuite with DotParsing with DotTyper {
   test("TooRecursive") { expectResult("stuck in infinite derivation"){checkfail(
       "val x = new Top {x => l: x.L; L: Bot .. x.l.L} (l=x); x")
   }}
+
+  test("MethodInitScope") { expectResult(Top){check(
+    """
+      val w = new Top { y =>
+        f: Top{a => S: Bot..Top; l: a.S} -> Top
+        id: Top -> Top
+      } (
+        f(a) = a.l
+        id(x) = x
+      );
+      w.id(w)
+    """
+)}}
 }

### 2015-12-04 doc/fool/scaladays-slides.tex
add old slides from Scala Days

\documentclass{beamer}

\input{dot_macros}
\usepackage{bcprules}

\usepackage{minted}
\usemintedstyle{eclipse}
% I added & and | as Scala keywords in Pygments:
%% diff -r 121c75491e0d pygments/lexers/jvm.py
%% --- a/pygments/lexers/jvm.py	Tue May 06 20:21:16 2014 -0700
%% +++ b/pygments/lexers/jvm.py	Sat May 10 19:21:51 2014 +0200
%% @@ -259,7 +259,7 @@
%%               u'lazy|match|new|override|pr(?:ivate|otected)'
%%               u'|re(?:quires|turn)|s(?:ealed|uper)|'
%%               u't(?:h(?:is|row)|ry)|va[lr]|w(?:hile|ith)|yield)\\b|'
%% -             u'(<[%:-]|=>|>:|[#=@_\u21D2\u2190])(\\b|(?=\\s)|$)', Keyword),
%% +             u'(<[%:-]|\\&|(\\|)|=>|>:|[#=@_\u21D2\u2190])(\\b|(?=\\s)|$)', Keyword),
%%              (u':(?!%s)' % op, Keyword, 'type'),
%%              (u'%s%s\\b' % (upper, idrest), Name.Class),
%%              (r'(true|false|null)\b', Keyword.Constant),

\usepackage{verbatim}
\usepackage{hyperref}

\useoutertheme{infolines}
\setbeamertemplate{headline}{}
\setbeamertemplate{footline}{
  \hfill
  \usebeamercolor[fg]{page number in head/foot}
  \usebeamerfont{page number in head/foot}
  \insertpagenumber\kern1em\vskip10pt
}
\setbeamertemplate{navigation symbols}{}

\title{The {\bf DOT} Calculus}
\subtitle{({\bf D}ependent {\bf O}bject {\bf T}ypes)}
\author{Nada Amin}
\institute{Scala Days}
\date{June 18, 2014}

\AtBeginSection{
  \begin{frame}
    \begin{center}
      \structure{\Huge \insertsection}
    \end{center}
  \end{frame}
}

\begin{document}

\frame{\titlepage}

\begin{frame}[fragile]{DOT: Dependent Object Types}
\begin{itemize}
\item DOT is a core calculus for path-dependent types.
\item Goals
\begin{itemize}
\item simplify Scala's type system by desugaring to DOT
\item simplify Scala's type inference by relying on DOT
\item prove that DOT is type-safe
\end{itemize}
\end{itemize}
\end{frame}

\begin{comment} (abstract)
The DOT (Dependent Object Types) calculus attempts to ground Scala's
type system in fewer, but powerful, constructs. In this talk I will
describe what these constructs are, and how they relate to Scala's
current type system. I will also show how DOT simplifies type
inference. Finally, I will touch upon the challenges in the
meta-theory.
\end{comment}

\section{Types in Scala and DOT}

\begin{frame}[fragile]{Types in Scala}
\begin{description}[functional]
\item[modular]\begin{description}[higher-kinded types]
\item[named type]\mint{scala}|scala.collection.BitSet|
\item[compound type]\mint{scala}|Channel with Logged|
\item[refined type]\mint{scala}|Channel { def close(): Unit } |
\end{description}
\item[functional]\begin{description}[higher-kinded types]
\item[parameterized type]\mint{scala}|List[String]|
\item[existential type]\mint{scala}|List[T] forSome { type T }|
\item[higher-kinded type]\mint{scala}|List|
\end{description}
\end{description}
\begin{comment}
too many orthogonal concepts?
can we simplify by keeping only the modular features?
\end{comment}
\end{frame}

\begin{frame}[fragile]{Reducing Functional to Modular?}
\begin{itemize}
\item type parameter to type member\\
\mint{scala}|class List[Elem] {} /*vs*/ class List { type Elem }|
\item parameterized type to refined type\\
\mint{scala}|List[String] /*vs*/ List { type Elem = String }|
\item existential type?\\
\mint{scala}|List[_] /*or*/ List[T] forSome { type T } /*~vs~*/ List|
\item higher-kinded type?\\
\mint{scala}|List /*~vs~*/ List|
\end{itemize}
\end{frame}

\begin{frame}[fragile]{Bonus: {\bf D}on't {\bf R}epeat {\bf Y}ourself}
\begin{minted}{scala}
         new HashMap[String, List[Int]]
with SynchronizedMap[String, List[Int]]

           /*vs*/

         new HashMap[String, List[Int]]
with SynchronizedMap
\end{minted}
\end{frame}

\begin{frame}[fragile]{Type Members and Modularity (Example)}
\begin{minted}{scala}
trait Symbols {
  type Type
  // a symbol has a type ...
  trait Symbol { def tpe: Type }
}
trait Types {
  type Symbol
  // ... and a type has a symbol
  trait Type { def sym: Symbol }
}
// ... but the references are not hard-coded!
// putting the cake together:
trait SymbolTable extends Symbols with Types
\end{minted}
\end{frame}

\begin{frame}[fragile]{Type Members and Data Abstraction (Example)}
\begin{minted}{scala}
trait HeapModule {
  type Heap
  type Elem
  def ord: Ordering[Elem]
  def empty: Heap
  def isEmpty(h: Heap): Boolean
  def insert(x: Elem, h: Heap): Heap
  def findMin(h: Heap): Elem
  def deleteMin(h: Heap): Heap
  def merge(h1: Heap, h2: Heap): Heap
}
trait BinomialHeapModule extends HeapModule {
  type Rank = Int
  case class Node(x: Elem, r: Rank, c: List[Node])
  override type Heap = List[Node]
  // ... implementation ...
}
\end{minted}
\end{frame}

\begin{frame}[fragile]{Path-Dependent Types and Binary Methods (Example)}
\begin{minted}{scala}
val m1: HeapModule = ...
val m2: HeapModule = ...
// in Scala REPL
> m1.merge(m1.empty, m2.empty)
// error: type mismatch;
// found   : m2.Heap
// required: m1.Heap
//              m1.merge(m1.empty, m2.empty)
//                                    ^
\end{minted}
\end{frame}

\begin{frame}[fragile]{Types in DOT}
\begin{description}[declarations]
\item[types]\mint{scala}|S, T, U|
\begin{description}[path-dependent type]
\item[path-dependent type]\mint{scala}|p.L|
\item[refined type]\mint{scala}|T { z => D }|
\item[intersection]\mint{scala}|T & T|
\item[union]
\begin{minted}{scala}
T | T
\end{minted}
\item[top]\mint{scala}|Any|
\item[bottom]\mint{scala}|Nothing|
\end{description}
\item[declarations]\mint{scala}|D|
\begin{description}[path-dependent type]
\item[type declaration]\mint{scala}|type L >: S <: U|
\item[field declaration]\mint{scala}|val l: U|
\item[method declaration]\mint{scala}|def m(x: S): U|
\end{description}
\end{description}
\end{frame}

\section{Type Inference}

\begin{frame}[fragile]{Type Inference in Scala (Least Upper Bound?)}
\begin{minted}{scala}
trait A { type T <: A }
trait B { type T <: B }
trait C extends A with B { type T <: C }
trait D extends A with B { type T <: D }
// in Scala, lub(C, D) is an infinite sequence
A with B { type T <: A with B { type T <: ... } }

// in Scala REPL
> val o = if (true) (new C{}) else (new D{})
o: A with B{type T <: A with B} = ...
> val o:A with B{type T<:A with B {type T<:A with B}} =
          if (true) (new C{}) else (new D{})
o: A with B{type T <: A with B{type T <: A with B}} = ...
\end{minted}
\end{frame}

\begin{frame}[fragile]{Type Inference in Scala (Working too Hard too Soon?)}
\includegraphics[width=\textwidth]{lub1.png}
\end{frame}

\begin{frame}[fragile]{Type Inference in Scala (Working too Hard too Soon?)}
\includegraphics[width=\textwidth]{lub2.png}
\end{frame}

\begin{frame}[fragile]{Type Inference in Scala (Working too Hard too Soon?)}
\begin{minted}{scala}
import scala.collection.mutable.{Map => MMap, Set => MSet}
val ms: MMap[Int, MSet[Int]] = MMap.empty

// in Scala REPL
> if (!ms.contains(1)) ms += 1 -> MSet(1) else ms(1) += 1
res0: ... (796 characters) ...
> :t res0
    : ... (21481 characters) ...
\end{minted}
\begin{itemize}
\item Inspired by a bug report\\
\href{https://issues.scala-lang.org/browse/SI-5862}{SI-5862}: very slow compilation due to humonguous LUB.
\item The character lengths reported are for Scala 2.11 ({\em after} the fix).
\item In Dotty, type inference can be lazy thanks to the native unions (for least upper bounds) and intersections (for greatest lower bounds) of the core calculus (DOT).
\end{itemize}
\end{frame}

\section{Meta-Theory}

\begin{frame}[fragile]{Type Safety: ``Well-typed programs can't go wrong.''}
\begin{minted}{javascript}
// ... at least in some ways ...
// JavaScript REPL
> var x = {foo: 'hello'}
undefined
> x.foo
"hello"
> x.bar
undefined
> x.bar.boo
// TypeError: Cannot read property 'boo' of undefined
> x.foo('hello')
// TypeError: string is not a function
\end{minted}
\end{frame}

\begin{frame}[fragile]{Type Safety: Syntactic Recipe (Concepts)}
\begin{description}[preservation]
\item[small-step] operational semantics (``show all your steps'')
\item[irreducible] terms, stuck terms vs values
\item[progress] theorem: if a term type-checks, then it can take a step or it is a value (but not stuck!)
\item[preservation] theorem: if a term type-checks and steps, then the new term also type-checks with the same type (or subtype)
\item[type safety] = progress + preservation
\end{description}
\end{frame}

\begin{frame}[fragile]{Preservation Challenge: Branding}
\begin{minted}{scala}
trait Brand {
  type Name
  def id(x: Any): Name
}
// in REPL
val brand: Brand = new Brand {
  type Name = Any
  def id(x: Any): Name = x
}
brand.id("hi"): brand.Name // ok
"hi": brand.Name // error but probably sound
val brandi: Brand = new Brand {
  type Name = Int
  def id(x: Any): Name = 0
}
brandi.id("hi"): brandi.Name // ok
"hi": brandi.Name // error and probably unsound
\end{minted}
\end{frame}

\begin{frame}[fragile]{Theory: Subtyping of Path-Dependent Types}
\begin{itemize}
\item \mint{scala}|/*If*/ p.L /*has lower bound*/ Sp /*and upper bound*/ Up|
\begin{itemize}
\item \mint{scala}|S   <: p.L /*if*/ S   <: Sp|
\item \mint{scala}|p.L <: U   /*if*/ Up  <: U|
\end{itemize}
\end{itemize}
\end{frame}

\begin{frame}[fragile]{Preservation Challenge: Path equality}
\begin{minted}{scala}
trait B { type X; val l: X }
val b1: B = new B { type X = String; val l: X = "hi" }
val b2: B = new B { type X = Int;    val l: X = 0    }
trait A { val i: B }
val a = new A { val i: B = b1 }
println(a.i.l : a.i.X) // ok
println(b1.l  : b1.X)  // ok
println(b1.l  : a.i.X) // error: type mismatch;
// found   : b1.l.type (with underlying type b1.X)
// required: a.i.X
// abstractly, would need to show
Any  <: Nothing // lattice collapse!
// to show
b1.X <: a.i.X
// because
U of b1.X = Any <: Nothing = S of a.i.X
println(b2.l  : a.i.X) // error and probably unsound
\end{minted}
\end{frame}

\begin{frame}[fragile]{Challenge: Subtyping Transitivity}
\begin{itemize}
\item \mint{scala}|S <: T <: U /*imply*/ S <: U|
\end{itemize}
\end{frame}

\begin{frame}[fragile]{Subtyping Transitivity and Path-Dependent Types}
\begin{itemize}
\item \mint{scala}|S <: p.L <: U|
\item \mint{scala}|p.L /*has lower bound*/ Sp /*and upper bound*/ Up|
\item \mint{scala}|S  <: Sp|
\item \mint{scala}|Up <: U|
\end{itemize}
\end{frame}

\begin{frame}[fragile]{Subtyping Transitivity and Path-Dependent Types with Bad Bounds}
\begin{itemize}
\item \mint{scala}|/*Suppose*/ p.L /*has bad bounds, e.g.*/|
\begin{itemize}
\item \mint{scala}|p.L /*has lower bound*/ Any /*and upper bound*/ Nothing|
\item \mint{scala}|S  <: Any     /*for any*/ S|
\item \mint{scala}|U  <: Nothing /*for any*/ U|
\item \mint{scala}|S <: p.L <: U /*imply*/ S <: U /*for any*/ S, U|
\end{itemize}
\item Lattice collapse!
\end{itemize}
\end{frame}

\begin{frame}[fragile]{Path-Dependent Types (Recap Example)}
\begin{minted}{scala}
trait Animal { type Food; def gets: Food
               def eats(food: Food) {}; }
trait Grass; trait Meat
trait Cow extends Animal with Meat {
  type Food = Grass; def gets = new Grass {} }
trait Lion extends Animal {
  type Food = Meat;  def gets = new Meat {} }
val leo = new Lion {}
val milka = new Cow {}
leo.eats(milka) // ok
val lambda: Animal = milka
lambda.eats(milka) // type mismatch
// found : Cow
// required: lambda.Food
lambda.eats(lambda.gets) // ok
\end{minted}
\end{frame}

\begin{frame}[fragile]{Path-Dependent Types (Recap Example Continued)}
\begin{minted}{scala}
def share(a1: Animal)(a2: Animal)
         (bite: a1.Food with a2.Food) {
  a1.eats(bite)
  a2.eats(bite)
}
val simba = new Lion {}
share(leo)(simba)(leo.gets) // ok
share(leo)(lambda)(leo.gets) // error: type mismatch
// found : Meat
// required: leo.Food with lambda.Food

// Observation:
// We don't know whether the parameter type of
share(lambda1)(lambda2)_
// is realizable until run-time.
\end{minted}
\end{frame}

\begin{frame}[fragile]{Realizability of Intersection Types at Run-Time}
\begin{minted}{scala}
val lambda1: Animal = new Lion {}
val lambda2: Animal = new Cow  {}
lazy val p: lambda1.Food & lambda2.Food = ???
// for illustration, say we re-defined the following:
trait Food { type T }
trait Meat extends Food  { type T = Nothing }
trait Grass extends Food { type T = Any }
// statically
p.T /*has fully abstract bounds*/
// at runtime
lambda1.Food /*is*/ Meat /*&*/ lambda2.Food /*is*/ Grass
p /*has type*/ Meat & Grass
// lower bound is union of lower bounds
p.T /*has lower bound*/ Nothing | Any /*is*/ Any
// upper bound is intersection of upper bounds
p.T /*has upper bound*/ Nothing & Any /*is*/ Nothing
p.T /*has bad bounds at run-time!*/
\end{minted}
\end{frame}

\begin{frame}[fragile]{Summary}
\begin{itemize}
\item DOT is a core calculus for path-dependent types.
\item Benefits for Scala
\begin{itemize}
\item simpler type system
\item better type inference
\item (hopefully) easier reasoning
\end{itemize}
\item Challenges in Meta-Theory
\begin{itemize}
\item type preservation
\begin{itemize}
\item run-time path equality
\item inlining branding
\end{itemize}
\item path-dependent types with ``bad bounds''
\begin{itemize}
\item subtyping transitivity
\item run-time ``bad bounds'' via intersection types
\end{itemize}
\end{itemize}
\end{itemize}
\end{frame}

\section{Thank You!}

\end{document}
